### 2021-03-25 reports/20210325Manifolds.Rmd
Summary on videos about manifolds, topology

---
title: "Manifolds, topological space, and tangent space"
# author:
# - familyname: Cheng
#   othernames: Fan
#   address: Monash University
#   email: Fan.Cheng@monash.edu
blind: true
cover: false
toc: false  
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache=FALSE, messages=FALSE, warning=FALSE, eval=TRUE, fig.pos = 'H')
# options(tinytex.verbose = TRUE)
```

```{r packages, include=FALSE, eval=FALSE}
library(tidyverse)
library(ggplot2)
library(knitr)
```

# Online Tutorials

YouTube tutorials on topology, manifolds & differential geometry by [*WHYBmaths*](https://www.youtube.com/playlist?list=PLxBAVPVHJPcrNrcEBKbqC_ykiVqfxZgNl)

- Preliminary: Sets & Maps

- Topology: Introduction, Constructing Torus, Topological Spaces, Continuity, Metric Spaces, Hausdorff property

- Manifolds: Introduction, Topological Manifolds, Transition Functions, Tangent Space

- Other: Vector Fields, Differential Forms

**Manifolds** are **topological spaces** that are **locally homeomorphic** to Euclidean space.

# Topological spaces

Topology space: a set $M$ with *topology* $\{\mathcal{T}_M\}$.

## Topology

Topology is the mathematical study of the properties of a geometric object that are preserved under *continuous deformations*, such as stretching, twisting, crumpling and bending. [*Wikipedia*](https://en.wikipedia.org/wiki/Topology)

Topological invariant: *homeomorphic* in the property of the space.

Homeomorphism: a continuous function between topological spaces that has a continuous inverse function.

For example: a circle is homeomorphic to a rectangle/square by certain stretching.

Constructing torus in 2-D.

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1342.jpg")
```

## Topological space: $(M, \{\mathcal{T}_M\})$

Topology properties:

1) $M,\phi \in \{\mathcal{T}_M\}$;

2) For all topological element $\tau_i \in \{\mathcal{T}_M\}$, the finite intersection set is also in the topology,
$$
\bigcap_i \tau_i \in \{\mathcal{T}_M\};
$$

3) Similarly, the infinite union set is also in the topology,
$$
\bigcup_i \tau_i \in \{\mathcal{T}_M\};
$$

Two extreme cases:

- Chaotic topology: $\mathcal{T}_M = \{M,\phi\}$.

- Discrete topology: $\mathcal{T}_M = \{P(M),\phi\}$, where $P(M)$ is the power set containing all possible subset of M.

Standard topology $T_{std}$ of a set $U \subset \mathbb{R}^2$:

1) Consider constructing a ball $B_r(x_i)$;

2) For any open set $U \in T_{std}$, there exists a ball $B_r(U) \subset U$.

1)&2) give a 2-d topology space $(U, B_r(U))$. [Example]

# Continuity of maps

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1343.jpg")
```

# Metric Spaces

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1344.jpg")
```

# Manifolds

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1345.jpg")
knitr::include_graphics("figures/manifold/IMG_1346.jpg")
```

# Transition functions

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1347.jpg")
```

## Example
```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1348.jpg")
```

# Tangent spaces

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1349.jpg")
```

# Others

## Vector fields

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1350.jpg")
```

## Differential forms

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/manifold/IMG_1351.jpg")
```

### 2021-05-26 docs/idea.Rmd
Rename reports and docs

---
title: "Some ideas about anomalies on manifolds"
author: "Anastasios Panagiotelis"
date: "05/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dimRed)
library(rgl)
library(hdrcde)
knitr::knit_hooks$set(webgl = hook_webgl)
set.seed(1983)
```

## Background

Consider the case of points on a unit-radius sphere with three coordinates $x$, $y$ and $z$.  Consider only the positive quadrant, i.e. points where $x>0$,$y>0$,$z>0$.  This is not just a toy example, for instance consider a statistical manifold for discrete distributions of a random variable $U$ with domain 0,1,2.  If we set $x=\sqrt{\textrm{Pr}(U=0)}$, $y=\sqrt{\textrm{Pr}(U=1)}$, $z=\sqrt{\textrm{Pr}(U=2)}$ then each distribution is a point that lies in the positive quadrant of unit sphere.

There are three coordinates since the ambient space is $\mathbb{R}^3$ but the surface of the sphere itself is 2-dimensional.  There are a number of ways we can reduce the dimension down to two coordinates.  For instance we can use the polar coordinates given by

$$
\theta=arccos(z)\\
\phi=arctan(y/x)
$$

Here, $\theta$ is an angle downwards from the north pole and $\phi$ is an angle around the equator.  For our purposes, think of computing polar coordinates as a way of doing dimension reduction, i.e. you input $(x,y,z)$ and it returns $(\theta,\phi)$.  Although an actual manifold learning algorithm is not going to return the polar coordinates, it serves as a nice illustration of some of the issues around identifying anomalies on manifolds.

## Uniform in sphere $\neq$ uniform in polar coordinates

For a more extensive discussion of parts of this section see [this blog post](http://corysimon.github.io/articles/uniformdistn-on-sphere/) noting that they use $\phi$ and $\theta$ the other way around.  Suppose we want to generate data uniformly on the sphere.  Without getting into too much detail as to why, the following will work:

```{r unifsphere, echo=TRUE, eval=TRUE}
n<-1000 #number of observations to be simulated

#Generate from independent standard normals with constant variance (the fact that constant variance and zero covariance is called sphericity is not a coincidence)

x<-rnorm(n)
y<-rnorm(n)
z<-rnorm(n)

#Norm of randomly generated vectors
norm_xyz<-sqrt(x^2+y^2+z^2)

#Normalise each coordinate to guarantee they lie on unit sphere
#Taking the absolute value puts everything into the upper quadrant
x<-abs(x)/norm_xyz 
y<-abs(y)/norm_xyz
z<-abs(z)/norm_xyz

```

This can be plotted in three dimensions and looked at interactively by clicking and dragging

```{r unisphereplot, webgl=TRUE, echo=TRUE}

plot3d(x,y,z)
```

Suppose we want to understand something about the density of these points.  Strictly speaking anomaly detection (Robs colorful plots) will break down a bit because these are uniform data and there are no anomalies.  However the following exercise does give intuition.

Suppose our 'manifold learning' algorithm is to apply the functions that give us spherical coordinates

```{r sphcoord, echo=TRUE, eval=TRUE}
theta<-acos(z)
phi<-atan(y/x)
```

below is a plot of the spherical coordinates with HDRs.

```{r hdeplot}
plot(hdrscatterplot(theta,phi))
```

Notice that the region with low values of $\theta$ is sparse and the data is not uniform.  This is the region near the north pole that gets stretched out when we transform from three dimensional coordinates to polar coordinates.  The anomalies we have detected are a consequence of the embedding and NOT a feature of the data.  More generally looking at the distribution of the low-dimensional data does not always tell us the correct things about the distribution of the high dimensional data

## Using dimension reduction algorithms

Now we can try to see what happens when we use an actual dimension reduction algorithm, for instance ISOMAP

```{r isomap, echo=TRUE,message=FALSE}
dat<-dimRedData(cbind(x,y,z)) #Prepare data to use in dimRed
emb<-embed(dat,"Isomap") #Isomap embedding
plot(hdrscatterplot(emb@data@data[,1],emb@data@data[,2])) #HDR plot
```

Which seems to not distort things too much.  The HDR plot still identifies outliers (because it has to) but they are just scattered around the corners in a very non-systematic way.

For something that leads to a lot of distortion have a look t t-SNE.

```{r tsneplot, echo=TRUE,message=FALSE}
emb_tsne<-embed(dat,"tSNE") #tSNE embedding
plot(hdrscatterplot(emb_tsne@data@data[,1],emb_tsne@data@data[,2])) #HDE plot
```

The aim of the project is not to evaluate different manifold learning algorithms.  Although ISOMAP worked OK here it is still distorted and in real examples you will never 'know' that the data are uniform on the sphere.  Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.

## Two dimensional kernel density estimation

In general a multivariate kernel density estimate looks something like this

$$
\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)
$$

where if a Gaussian kernel is used

$$
K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
$$

The matrix $\mathbf{H}$ is called the bandwidth matrix and is very important.  You will often read that the bandwidth matrix is about smoothing and it is.  But an alternative way to think of kernel estimation is that kernel densities 'borrow strength' from nearby points and the bandwidth determines what is "nearby".  If the bandwidth is large then all points are "nearby" and we get an overly smooth kernel density estimate.  The interesting thing about a bandwidth matrix is that it allows for different notions of what is "nearby" along different coordinates and even along diagonal directions.

## Using the Riemannian

The Riemannian estimated using the method of Perrault Joncas and Meila gives some idea of the distortion of an embedding (or so they claim). Mapping the points through a non-linear function "stretchs" some regions of space and "shrinks" others. The Riemannian gives us an idea of the direction and angle of this stretching.  The Riemannian is quite a technical concept but an important thing to understand is that the estimate that comes out of Perrault Joncas algorithm is a square matrix.

We saw how points that are far apart in the embedding may not have been so far apart on the original manifold.  The Riemannian gives us some way of correcting this.  Similarly a bandwidth matrix in a kernel density estimate is all about determining the "directions" in which there should be more or less "closeness".  So the basic idea is to replace the kernel density estimate with

$$
\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
$$

where $H_i$ is either the Riemannian or the inverse of the Riemannian (I am not totally sure which one).  Notice that the bandwidth matrix is different for each point.  This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.  There is a paper by Terrel and Scott from Annals of Statistics in 1992 where they talk about this (it is called "variable kernel density estimation"). Some other search terms to find relevant papers might be "adaptive" or "varying bandwidth". I don't know if anyone has had the idea of bringing all these ideas together with manifold learning and the estimate of the Riemannian - we should do a thorough literature search.  However, if no one has already done this I think it is a very interesting way to do anomaly detection for very high dimensional data.

### 2021-05-26 reports/20210513_riem_plot.Rmd
Rename reports and docs

---
title: "Riemannian metric from manifold learning"
author: "Fan Cheng"
date: "13/05/2021"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE}
library(tidyverse)
library(dimRed)
library(dtplyr)
library(reticulate)
library(viridis)
# library(ks)
library(hdrcde)
library(igraph)
# library(seurat)
library(matrixcalc)
Jmisc::sourceAll(here::here("R/sources"))
```

# Smart meter data for one household

```{r data}
# load(here::here("data/spdemand_3639id336tow.rda"))
# nid <- 1
# ntow <- 336
# 
# train <- spdemand %>%
#   lazy_dt() %>%
#   filter(tow <= ntow,
#          # id <= sort(unique(spdemand[,id]))[nid],
#          id == 1003) %>%
#   dplyr::select(-id, -tow) %>%
#   data.table::as.data.table()
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
N <- nrow(train)
```

# Metric learning

```{r parameters}
# Parameters fixed
x <- train
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4
```

## Radius searching with k-d trees

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype)
summary(metric_isomap)
```
The function `metricML()` returns a list of

- the embedding coordinates `embedding` of $N \times N$
- the embedding metric `rmetric` for each point $p \in x$ as an array
- the weighted neighborhood graph as an `igraph` object `weighted_graph`
- the sparse adjacency matrix from the graph `adj_matrix`

## Embedding plots

```{r}
# Comparison of Isomap embedding plot
par(mfrow = c(1, 2))
plot(metric_isomap$embedding, col = viridis::viridis(24)) # metricML
# plot(e@data@data, col = viridis::viridis(24)) # embedding from dimRed, same as metricML()
emb_isomap <-
  feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather"))
plot(-emb_isomap$`0`, emb_isomap$`1`, col = viridis::viridis(24)) # embedding from megaman, radius = 0.4
```

## Riemannian metric

```{r}
# Use adjacency matrix as input for metricML and dimRed, then the embedding should be close, as well as the Riemannian metric
metric_isomap$adj_matrix[1:10, 1:10]# renormalized adjacency matrix
metric_isomap$weighted_graph %>% is.connected()
riem_isomap <- metric_isomap$rmetric
riem_isomap[,,1] %>% 
  # isSymmetric()
  matrixcalc::is.positive.definite() # FALSE
```

**However, the Riemannian metric from the `metricML()` function is not positive definite for all points. **

Now compare it with the megaman output.

```{r, eval=TRUE}
np = import("numpy")
H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy"))
H_isomap[1,,, drop=TRUE] %>% 
  matrixcalc::is.positive.definite() # TRUE

pd <- rep(NA, N)
for (i in 1:N) {
  pd[i] <- riem_isomap[,,i] %>% 
    # isSymmetric()
    matrixcalc::is.positive.definite() # FALSE
}
pd %>% sum

determinant(riem_isomap[,,i])
mat <- riem_isomap[,,i]
eigen(mat)
```

## Bug: riemmanien metric not positive definite

```{r, eval=FALSE}
# Function for graph Laplacian
# Input: W: N*N weight matrix for the neighborhood graph, radius: bandwidth parameter
# Output: L: N*N graph Laplacian matrix
Laplacian <- function(W, radius){

  Tn <- Matrix::Diagonal(x = 1 / rowSums(W)) # inverse of a diagonal matrix
  # temp <- solve(Tn)
  W1 <- Tn %*% W %*% Tn
  Tn1 <- Matrix::Diagonal(x = 1 / rowSums(W1)) # inverse of Tn1
  L <- (Matrix::Diagonal(nrow(W)) - Tn1 %*% W1) / radius

  return(L)
}
```

```{r}
# Function for Riemannian metric for each point
# The Riemannian metric and its dual associated with an embedding Y. 
# The Riemannian metric is currently denoted by G, its dual by H, and the Laplacian by L. 
# G at each point is the matrix inverse of H.
riemann_metric <- function(Y, laplacian, ndim, invert.h = TRUE){

  H <- array(NA, dim = c(ndim, ndim, nrow(Y)))

  for (i in 1:ndim) {
    for (j in i:ndim) {
      yij <- Y[, i] * Y[, j]
      H[i, j, ] <- as.vector(0.5 * (laplacian %*% yij - Y[, i] * (laplacian %*% Y[, i]) - Y[, j] * (laplacian %*% Y[, j])))
      H[j, i, ] <- H[i, j, ] # symmetric matrix
    }
  }

  # Array H corresponds to \tilde{H} in Step 4(a)
  # The embedding metric H is the pseudo inverse of \tilde{H}
  # for (i in 1:nrow(Y)) {
  #   Hsvals <- eigen(H[,,i])
  #   Huu <- Hsvals$vectors
  #   Hvv <- Hsvals$values[1:ndim]
  #   Hvv1 <- diag(x = 1 / Hvv)
  #   H[,,i] <- Huu %*% Hvv1 %*% t(Huu)
  #   H[, , i] <- 0.5 * (H[, , i] + t(H[, , i]))
  # }

  if(invert.h){
    for (i in 1:nrow(Y)) {
      H[, , i] <- solve(H[, , i])
      H[, , i] <- 0.5 * (H[, , i] + t(H[, , i])) # fix not symmetric issue
    }
  }

  return(H)
}
```

# Plotting

# Plot ellipse to show distortion

Using the Riemannian metric for each point as the covariance and the 2-d embeddings as the mean to plot an ellipse for each point.

The underlying requirement is that the Riemannian metric needs to be a square positive definite matrix, i.e. `matrixcalc::is.positive.definite()` returns `TRUE`.

```{r}
emb <- metric_isomap$embedding
i = 1
mat <- riem_isomap[,,i] # not positive definite for metricML()
center <- c(emb[i,1], emb[i,2])

mat <- H_isomap[i,,, drop=TRUE] # is.positive.definite for megaman

mat <- matrix(c(2.2, 0.4, 0.4, 2.8), nrow=2) # covariance matrix -> cov(dataMatrix)

library(ellipse)
# plotcorr(mat)
# plotcorr(solve(mat))
plot(ellipse::ellipse(mat, center = center, scale = c(1,0.5)), type = "l", col = "blue", asp = 1)

car::ellipse(center, shape=mat, radius=1, col="red", lty=2)
# library(mixtools)
mixtools::ellipse(mu = center, sigma = mat, type = "l")
```

# Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by
$$\hat{f}(x) = n^{-1} \sum_i K_H (x - X_i).$$
```{r}
# 2-d kde
mkde <- function (x, h) {
  # Data is a matrix of n*d
  # H is an array of dimension (d,d,n)
  start <- Sys.time()

  n <- nrow(x)
  if (dim(x)[2] < 2)
    stop("Data matrix has only one variable.")
  if (any(!is.finite(x))) 
    stop("Missing or infinite values in the data are not allowed! ")
  if (!all.equal(nrow(h), ncol(h), dim(x)[2]))
    stop("The bandwidth matrix is of wrong dimension. ")

  s <- rep(0, n)
  y <- rep(0, n)
  for (i in 1:n) {
    for (j in 1:n){
      s[i] <- s[i] + abs(det(h[,,i]))^(-1/2) * exp(- 1/2 * t(x[i,] - x[j,]) %*% solve(h[,,i]) %*% (x[i,] - x[j,]))
    }
    y[i] <- s[i] / (n * 2 * pi)
  }
  print(Sys.time() - start)

  return(y)
}
x = metric_isomap$embedding
f <- mkde(x = metric_isomap$embedding, h = riem_isomap)
summary(f)
```

# Anomaly detection plots

```{r}
# Plotting
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na() 
embed_den_list <- list(x = embed_den$x, y = embed_den$y, z = embed_den$z)
embed_den
hdrscatterplot(embed_den$x, embed_den$y)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99)) 
plot.hdr2d(hdrinfo)
# If density values are specified
hdrinfo1 <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99), den = embed_den_list)
# plot.hdr2d(hdrinfo)
# filled.contour(x = embed_den$x, y = embed_den$y, z = embed_den$z)
```

# Contour plot after interpolation

```{r}
library(akima)
# ?interp()
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=100, ny=100, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax),10)
# colors <- heat.colors(length(breaks)-1)
# hdrscatterplot(embed_den$x, embed_den$y)
plot.hdr2d(hdrinfo)
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth  interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))})

```

### 2021-09-02 R/sources/hdrscatterplot.R
Source code modified for metricML

@@ -30,7 +30,7 @@
 #' hdrscatterplot(x, y)
 #' hdrscatterplot(x, y, label = paste0("p", 1:length(x)))
 #' @export hdrscatterplot
-hdrscatterplot <- function(x, y, levels = c(1, 50, 99), kde.package = c("ash", "ks"), noutliers = NULL, label = NULL) {
+hdrscatterplot <- function(x, y, levels = c(1, 50, 99), kde.package = c("ash", "ks"), noutliers = NULL, label = NULL, den = NULL) {
   levels <- sort(levels)
   if (missing(y)) {
     data <- x
@@ -41,7 +41,7 @@ hdrscatterplot <- function(x, y, levels = c(1, 50, 99), kde.package = c("ash", "
 
   vnames <- names(data)
 
-  den <- hdr.2d(data[, 1], data[, 2], prob = levels, kde.package = kde.package)
+  if(is.null(den)) den <- hdr.2d(data[, 1], data[, 2], prob = levels, kde.package = kde.package)
 
   region <- numeric(NROW(data)) + 100
   for (i in seq_along(levels))

### 2021-09-02 R/sources/nn2dist.R

@@ -19,7 +19,7 @@ nn2dist <- function(nn, N = NULL) { # , k = NULL, sparse = TRUE, keep_graph = FA
   
   Kn <- matrix(1e+5, N, N)
   for(i in 1:N) {
-    Kn[i, index[i,]] <- distances[i, ]
+    Kn[i, index[i,]] <- distances[i, index[i,] != 0]
   }
   
   return(Kn)

### 2021-09-02 R/sources/metricML.R
Source code modified for metricML

@@ -28,7 +28,8 @@
 #' 
 metricML <- function(x, s, k = min(10, nrow(x)), radius = 0, 
                      adjacency = NULL, affinity = NULL,
-                     method, annmethod = c("kdtree", "annoy", "hnsw"), 
+                     method, 
+                     annmethod = c("kdtree", "annoy", "hnsw"),
                      eps = 0, nt = 50, nlinks = 16, ef.construction = 200, ef.search = 10,
                      distance = c("euclidean", "manhattan"), diag = FALSE,
                      treetype = c("kd", "bd"),
@@ -81,7 +82,8 @@ metricML <- function(x, s, k = min(10, nrow(x)), radius = 0,
     )
     names(nn2res) <- c("nn.idx", "nn.dists")
 
-    Kn <- nn2dist(nn2res, N = N, k = k, sparse = TRUE)
+    Kn <- nn2dist(nn2res, N = N)
+    Kn[Kn == 1e+05] <- 0
     g <- igraph::graph_from_adjacency_matrix(Kn)
     
   }
@@ -101,18 +103,18 @@ metricML <- function(x, s, k = min(10, nrow(x)), radius = 0,
                      .method = method,
                      knn = k,
                      ndim = s,
-                     annmethod = annmethod,
-                     radius = radius,
-                     eps = eps,
-                     nt = nt,
-                     nlinks = nlinks,
-                     ef.construction = ef.construction,
-                     ef.search = ef.search,
-                     distance = distance,
-                     treetype = treetype,
-                     searchtype = searchtype,
-                     perplexity = perplexity,
-                     theta = theta,
+                     # annmethod = annmethod,
+                     # radius = radius,
+                     # eps = eps,
+                     # nt = nt,
+                     # nlinks = nlinks,
+                     # ef.construction = ef.construction,
+                     # ef.search = ef.search,
+                     # distance = distance,
+                     # treetype = treetype,
+                     # searchtype = searchtype,
+                     # perplexity = perplexity,
+                     # theta = theta,
                      .mute = c("output"),
                      # ...
 

### 2021-09-02 reports/20210709_outliers.Rmd
Swiss roll mapping data from copula and gaussian mixture

---
title: "Plotting for metric learning output"
author: "Fan Cheng"
date: "2021-06-17"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(dimRed)
library(reticulate)
library(here)
library(viridis)
library(hdrcde)
library(igraph)
library(matrixcalc)
library(akima)
library(car)
library(ggforce)
library(ks)
library(patchwork)
Jmisc::sourceAll(here::here("R/sources"))
set.seed(123)
```

# Metric learning

## Setup

```{r parameters}
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
# Parameters fixed
x <- train
N <- nrow(train)
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm
```

## The metric learning algorithm

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, invert.h = TRUE)
summary(metric_isomap)
```

```{r ggellipse, include=FALSE, eval=FALSE}
plot_embedding(metric_isomap) +
  labs(x = "ISO1", y = "ISO2")
plot_ellipse(metric_isomap, add = F, n.plot = 50, scale = 20,
             color = blues9[5], fill = blues9[5], alpha = 0.2)
```

# Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by $$\hat{f}(x) = n^{-1} \sum_i K_{H_i} (x - X_i).$$

## Outlier plot based on density estimates

### Fixed bandwidth

```{r}
# fixed bandwidth
fn <- metric_isomap$embedding
E1 <- fn[,1] # rename as Ed to match the aesthetics in plot_ellipse()
E2 <- fn[,2]
prob <- c(1, 50, 99)
p_hdr_isomap <- hdrscatterplot(E1, E2, levels = prob, noutliers = 20, label = NULL)
p_hdr_isomap_p <- p_hdr_isomap + 
  plot_ellipse(metric_isomap, add = T, n.plot = 50, scale = 20, 
             color = blues9[5], fill = blues9[5], alpha = 0.2)
# p_hdr_isomap_p
```

### Pointwise adaptive bandwidth

```{r outliers}
Rn <- metric_isomap$rmetric # array
n.grid <- 10
f <- vkde2d(x = fn[,1], y = fn[,2], h = Rn, n = n.grid)
plot_contour(metric_isomap, n.grid = n.grid, scale = 1/10)
```

```{r hdroutliers}
source(here::here("R/sources/hdrplotting.R"))
p_isomap <- plot_outlier(x = metric_isomap, n.grid = 50, prob = prob, scale = 1/8)
```

```{r compoutlier}
(p_hdr_isomap_p + p_isomap$p ) + coord_fixed()
```

# t-SNE

```{r tsne, message=FALSE, warning=FALSE, eval=TRUE}
x <- train
method <- "anntSNE"
perplexity <- round(k / 3) # 30 by default
theta <- 0 # for exact tSNE in the C++ tSNE Barnes-Hut implementation
# tictoc::tic()
metric_tsne <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = perplexity, theta = theta, invert.h = TRUE)
# summary(metric_tsne)
# tictoc::toc()
```

```{r, message=FALSE, eval=TRUE}
ftsne <- metric_tsne$embedding
E1 <- ftsne[,1]; E2 <- ftsne[,2]
plot_embedding(metric_tsne)
plot_ellipse(metric_tsne, n.plot = 50)
plot_contour(metric_tsne, n.grid = 50, scale = 1/10)
```

```{r}
p_tsne <- plot_outlier(x = metric_tsne, n.grid = 50, prob = prob, noutliers = 20, scale = 1/10)
p_hdr_tsne <- hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20)
p_hdr_tsne_p <- p_hdr_tsne$p + 
  plot_ellipse(metric_tsne, n.plot = 50, add = T)
(p_hdr_tsne_p + p_tsne$p) + coord_fixed()
```

# Compare outliers from Isomap and t-SNE

To compare two manifold learning methods, we calculate the overlap among outliers. To do so, we first rank the outliers from highest to lowest densities. Then calculate the correlation of ranks for both methods.

To show our methods performs better, we expect a higher correlation of ranks. This shows robustness in finding outliers using different manifold learning methods.

```{r}
(p_isomap$p + p_tsne$p) + coord_fixed()
```

## Outlier ranking comaprison using variable bandwidth

The ranking of outliers can be compared as follows.

```{r}
p_isomap_all <- plot_outlier(x = metric_isomap, n.grid = 50, prob = prob, noutliers = N, scale = 1/8)
p_tsne_all <- plot_outlier(x = metric_tsne, n.grid = 50, prob = prob, noutliers =N, scale = 1/10)

head(p_isomap_all$outlier, n = 20)
head(p_tsne_all$outlier, n = 20)
outlier_isomap <- order(as.numeric(p_isomap_all$outlier))
outlier_tsne <- order(as.numeric(p_tsne_all$outlier))
# outliers <- cbind(outlier_isomap, outlier_tsne)
plot(outlier_isomap, outlier_tsne)

```

```{r, include=FALSE, eval=FALSE}
# ggcorrplot::ggcorrplot(r, hc.order = TRUE, type = "lower", lab = TRUE)
vec1 <- p_isomap$outlier
vec2 <- p_tsne$outlier
topconfects::rank_rank_plot(vec1, vec2, n=20)

library(topconfects)
a <- sample(letters)
b <- sample(letters)
rank_rank_plot(a,b, n=20)
```

## Outlier ranking comaprison using fixed bandwidth

```{r}
head(p_hdr_isomap$outlier, n = 20)
head(p_hdr_tsne$outlier, n = 20)
hdroutlier_isomap <- order(as.numeric(p_hdr_isomap$outlier))
hdroutlier_tsne <- order(as.numeric(p_hdr_tsne$outlier))
# hdroutliers <- cbind(hdroutlier_isomap, hdroutlier_tsne)
cor(hdroutlier_isomap, hdroutlier_tsne)
cor(hdroutlier_isomap %>% head(50), hdroutlier_tsne %>% head(50))
```

# Compare high density regions

Similarly, we compare the highest density regions from both methods and expect a similar output in detecing most typical observations.

## Variable bandwidth

```{r}
tail(p_isomap_all$outlier, n = 20)
tail(p_tsne_all$outlier, n = 20)
cor(outlier_isomap %>% tail(50), outlier_tsne %>% tail(50))
```

## Fixed bandwidth

```{r}
tail(p_hdr_isomap$outlier, n = 20)
tail(p_hdr_tsne$outlier, n = 20)
cor(hdroutlier_isomap %>% tail(50), hdroutlier_tsne %>% tail(50))
```

### 2021-09-02 reports/Simulation_swissroll.R

@@ -1,0 +1,362 @@
+## ----setup, include=FALSE-------------------------------------------------------
+knitr::opts_chunk$set(echo = TRUE)
+
+
+## ----libraries, message=FALSE, echo=TRUE, results='hide'------------------------
+library(tidyverse)
+library(dimRed)
+library(reticulate)
+library(here)
+library(viridis)
+library(hdrcde)
+library(igraph)
+library(matrixcalc)
+library(akima)
+library(car)
+library(ggforce)
+library(ks)
+library(patchwork)
+library(copula)
+library(plotly)
+Jmisc::sourceAll(here::here("R/sources"))
+set.seed(1)
+
+# data size
+N <- 2000
+p <- 2
+
+# # ----Swiss roll dataset with a sparse area------------------------
+# height <- 30
+# x <- (3 * pi / 2) * (1 + 2 * runif(N*1.05, 0, 1))  
+# y <- height * runif(N*1.05, 0 , 1)
+# sr <- cbind(x * cos(x), y, x * sin(x))
+# colnames(sr) <- c("x", "y", "z")
+# sr <- cbind(as_tibble(sr), col = "grey")
+# lowden <- x >= 13 & y >= 18
+# sr$col[lowden] <- "red"
+# removed <- sample(which(lowden), size = N*0.05, replace = FALSE)
+# lowdenpoints <- sr[setdiff(which(lowden), removed), ]
+# sr <- sr[-removed, ]
+# scatterplot3d::scatterplot3d(sr, color = sr$col)
+# # rgl::plot3d(sr, col = sr$col)
+# 
+
+
+# ----Swiss roll from copula-----------------------------------------
+# ## a 3-dimensional normal copula
+# norm.cop <- normalCopula(0.5, dim = p)
+# u <- rCopula(N, norm.cop)
+# den <- dCopula(u, norm.cop)
+
+# # 3-d t copula
+# t.cop <- tCopula(c(0.5, 0.3), dim = p, dispstr = "toep",
+#                  df = 2, df.fixed = TRUE)
+# u <- rCopula(N, t.cop)
+# den <- dCopula(u, t.cop)
+# 
+## a 2-dimensional Clayton copula
+cl3 <- claytonCopula(2, dim = p)
+u <- rCopula(N, cl3)
+# pairs(u)
+claytonden <- dCopula(u, cl3) # true density
+# Swiss roll mapping
[…]

### 2022-01-18 paper/paper.Rmd
Paper setup

--- 2021-05-26 docs/idea.Rmd
@@
 ---
-title: "Some ideas about anomalies on manifolds"
-author: "Anastasios Panagiotelis"
-date: "05/02/2021"
-output: html_document
+title: "Anomaly detection with kernel density estimation on manifolds"
+author:
+- familyname: Cheng
+  othernames: Fan
+  address: Monash University
+  email: Fan.Cheng@monash.edu
+  # correspondingauthor: true
+- familyname: Hyndman
+  othernames: Rob J
+  address: Monash University
+  email: Rob.Hyndman@monash.edu
+- familyname: Panagiotelis
+  othernames: Anastasios
+  address: University of Sydney
+  email: Anastasios.Panagiotelis@sydney.edu.au
+abstract: ""
+# keywords: "manifold learning"
+# wpnumber: 03/2021
+# jelcodes: C55, C65, C80
+blind: true
+cover: false
+toc: false
+lot: false
+bibliography: references.bib
+biblio-style: authoryear-comp
+output:
+  MonashEBSTemplates::workingpaper:
+    fig_caption: yes
+    fig_height: 5
+    fig_width: 8
+    includes:
+      in_header: preamble.tex
+      # after_body: appendix.tex
+    keep_tex: yes
+    number_sections: yes
+    citation_package: biblatex
+    latex_engine: pdflatex
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-
+knitr::opts_chunk$set(
+  echo = TRUE,
+  cache = FALSE,
+  messages = FALSE,
+  warning = FALSE,
+  eval = TRUE
+  # include = FALSE
+)
+options(tinytex.verbose = TRUE)
 library(dimRed)
-library(rgl)
-library(hdrcde)
-knitr::knit_hooks$set(webgl = hook_webgl)
-set.seed(1983)
+library(tidyverse)
+library(ggplot2)
+library(reshape2)
+library(knitr)
+library(kableExtra)
+library(tibble)
+library(patchwork)
+library(data.table)
+library(dtplyr)
+library(igraph)
+library(ggraph)
+# set.seed(1)
+Jmisc::sourceAll(here::here("R/sources"))
 ```
 
-## Background
-
-Consider the case of points on a unit-radius sphere with three coordinates $x$, $y$ and $z$.
-Consider only the positive quadrant, i.e.
-points where $x>0$,$y>0$,$z>0$.
-This is not just a toy example, for instance consider a statistical manifold for discrete distributions of a random variable $U$ with domain 0,1,2.
-If we set $x=\sqrt{\textrm{Pr}(U=0)}$, $y=\sqrt{\textrm{Pr}(U=1)}$, $z=\sqrt{\textrm{Pr}(U=2)}$ then each distribution is a point that lies in the positive quadrant of unit sphere.
-
-There are three coordinates since the ambient space is $\mathbb{R}^3$ but the surface of the sphere itself is 2-dimensional.
-There are a number of ways we can reduce the dimension down to two coordinates.
-For instance we can use the polar coordinates given by
-
-$$
-\theta=arccos(z)\\
-\phi=arctan(y/x)
-$$
-
-Here, $\theta$ is an angle downwards from the north pole and $\phi$ is an angle around the equator.
-For our purposes, think of computing polar coordinates as a way of doing dimension reduction, i.e.
-you input $(x,y,z)$ and it returns $(\theta,\phi)$.
-Although an actual manifold learning algorithm is not going to return the polar coordinates, it serves as a nice illustration of some of the issues around identifying anomalies on manifolds.
-
-## Uniform in sphere $\neq$ uniform in polar coordinates
-
-For a more extensive discussion of parts of this section see [this blog post](http://corysimon.github.io/articles/uniformdistn-on-sphere/) noting that they use $\phi$ and $\theta$ the other way around.
-Suppose we want to generate data uniformly on the sphere.
-Without getting into too much detail as to why, the following will work:
-
-```{r unifsphere, echo=TRUE, eval=TRUE}
-n<-1000 #number of observations to be simulated
-
-#Generate from independent standard normals with constant variance (the fact that constant variance and zero covariance is called sphericity is not a coincidence)
-
-x<-rnorm(n)
-y<-rnorm(n)
-z<-rnorm(n)
-
-#Norm of randomly generated vectors
-norm_xyz<-sqrt(x^2+y^2+z^2)
-
-#Normalise each coordinate to guarantee they lie on unit sphere
-#Taking the absolute value puts everything into the upper quadrant
-x<-abs(x)/norm_xyz 
-y<-abs(y)/norm_xyz
-z<-abs(z)/norm_xyz
-
-```
-
-This can be plotted in three dimensions and looked at interactively by clicking and dragging
-
-```{r unisphereplot, webgl=TRUE, echo=TRUE}
-
-plot3d(x,y,z)
-```
-
-Suppose we want to understand something about the density of these points.
-Strictly speaking anomaly detection (Robs colorful plots) will break down a bit because these are uniform data and there are no anomalies.
-However the following exercise does give intuition.
-
-Suppose our 'manifold learning' algorithm is to apply the functions that give us spherical coordinates
-
-```{r sphcoord, echo=TRUE, eval=TRUE}
-theta<-acos(z)
-phi<-atan(y/x)
-```
-
-below is a plot of the spherical coordinates with HDRs.
-
-```{r hdeplot}
-plot(hdrscatterplot(theta,phi))
-```
-
-Notice that the region with low values of $\theta$ is sparse and the data is not uniform.
-This is the region near the north pole that gets stretched out when we transform from three dimensional coordinates to polar coordinates.
-The anomalies we have detected are a consequence of the embedding and NOT a feature of the data.
-More generally looking at the distribution of the low-dimensional data does not always tell us the correct things about the distribution of the high dimensional data
+# Introduction
 

### 2022-01-18 paper/paper.tex
Paper setup

\documentclass[11pt,a4paper,]{article}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Anomaly detection with kernel density estimation on manifolds},
            pdfkeywords={manifold learning},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Anomaly detection with kernel density estimation on manifolds}

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\RequirePackage{bera}
\RequirePackage{mathpazo}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*5}{*1}
\titlespacing{\subsection}{0pt}{*2}{*0.2}

%% TITLE PAGE
\def\Date{\number\day}
\def\Month{\ifcase\month\or
 January\or February\or March\or April\or May\or June\or
 July\or August\or September\or October\or November\or December\fi}
\def\Year{\number\year}

\makeatletter
\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
\def\nojel{\def\showjel{}}
\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
\def\cover{{\sffamily\setcounter{page}{0}
        \thispagestyle{empty}
        \placefig{2}{1.5}{width=5cm}{monash2}
        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
        \begin{textblock}{7}(12.7,27.9)\hfill
        \includegraphics[height=0.7cm]{AACSB}~~~
        \includegraphics[height=0.7cm]{EQUIS}~~~
        \includegraphics[height=0.7cm]{AMBA}
        \end{textblock}
        \vspace*{2cm}
        \begin{center}\Large
        Department of Econometrics and Business Statistics\\[.5cm]
        \footnotesize http://monash.edu/business/ebs/research/publications
        \end{center}\vspace{2cm}
        \begin{center}
        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
                \LARGE\@author\end{onehalfspace}
        }}
        \end{center}
        \vfill
                \begin{center}\Large
                \Month~\Year\\[1cm]
                Working Paper \@wp
        \end{center}\vspace*{2cm}}}
\def\pageone{{\sffamily\setstretch{1}%
        \thispagestyle{empty}%
        \vbox to \textheight{%
        \raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
        \vspace{2cm}\par
        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
        \hspace{1cm}\showjel\vss}}}
\def\blindtitle{{\sffamily
     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
        }}
\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}

\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
\let\maketitle\titlepage
\makeatother

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

%% KEYWORDS
\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}

%% ABSTRACT
\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
  {\end{minipage}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[showonlyrefs]{mathtools}
\usepackage[no-weekday]{eukdate}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

\nojel

\RequirePackage[absolute,overlay]{textpos}
\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}
\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}

\nocover
\blind

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf\@date}
\makeatother

%% Any special functions or other packages can be loaded here.
\usepackage[utf8]{inputenc}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}
%% \usepackage{makecell,interfaces-makecell} %latex error
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage{xcolor}
% \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{bm}
\usepackage{etoolbox}
\usepackage{tabu}
\usepackage{placeins}
\setlength {\marginparwidth }{2cm}
\usepackage{todonotes}
\urlstyle{tt}  % don't use monospace font for urls
\allowdisplaybreaks

\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{amsmath, amsfonts, amssymb, bm, mathtools, todonotes, booktabs, microtype}

\makeatletter
%If bera loaded, this is monash wp.
\@ifpackageloaded{bera}{}{\usepackage{mathptmx}}
\makeatother
\usepackage[cal = txupr]{mathalpha}

\makeatletter
\@ifpackageloaded{geometry}{}{\usepackage{geometry}}
\makeatother
\geometry{margin=2.2cm}

\mathtoolsset{showonlyrefs}
\newcommand{\argmax}{\mathop{\text{arg\,max}}}
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\usepackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.9}

\usepackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

% theorem in latex
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle
\begin{keywords}
manifold learning
\end{keywords}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

In a famous paper, \textcite{BC64} introduced a family of transformations \dots

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{qplot}\NormalTok{(}\FunctionTok{exp}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\DecValTok{200}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/histogram-1.pdf}
\caption{\label{fig:histogram}Nice histogram}
\end{figure}

Consider the logNormal data plotted in Figure \ref{fig:histogram}.

\[s^2 = \sum_{i=1}^n (x_i-\bar{x})^2\]

\printbibliography

\end{document}

### 2022-02-09 paper/paper.tex
Simulation - 2d

@@
             pdftitle={Anomaly detection with kernel density estimation on manifolds},
-            pdfkeywords={manifold learning},
             pdfborder={0 0 0},
@@
 \addbibresource{references.bib}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs}
@@
 \IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \IfFileExists{parskip.sty}{%
@@
 %% Any special functions or other packages can be loaded here.
-\usepackage[utf8]{inputenc}
-\usepackage[flushleft]{threeparttable}
-\usepackage{multirow}
-\usepackage[normalem]{ulem}
-\useunder{\uline}{\ul}{}
-%% \usepackage{makecell,interfaces-makecell} %latex error
-\usepackage[]{graphicx}
-\usepackage[]{color}
-\usepackage{xcolor}
-% \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
-\usepackage{algorithmic}
-\usepackage{bm}
-\usepackage{etoolbox}
-\usepackage{tabu}
-\usepackage{placeins}
-\setlength {\marginparwidth }{2cm}
-\usepackage{todonotes}
-\urlstyle{tt}  % don't use monospace font for urls
-\allowdisplaybreaks
-
+\usepackage{tabu,placeins,algorithmic}
 \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
-\usepackage{amsmath, amsfonts, amssymb, bm, mathtools, todonotes, booktabs, microtype}
-
-\makeatletter
-%If bera loaded, this is monash wp.
-\@ifpackageloaded{bera}{}{\usepackage{mathptmx}}
-\makeatother
+\urlstyle{tt}  % use monospace font for urls
 \usepackage[cal = txupr]{mathalpha}
-
-\makeatletter
-\@ifpackageloaded{geometry}{}{\usepackage{geometry}}
-\makeatother
 \geometry{margin=2.2cm}
-
 \mathtoolsset{showonlyrefs}
-\newcommand{\argmax}{\mathop{\text{arg\,max}}}
-\clubpenalty = 10000
-\widowpenalty = 10000
-\brokenpenalty = 10000
-\usepackage{graphicx}
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.9}
-\renewcommand{\bottomfraction}{0.9}
-\renewcommand{\textfraction}{0.1}
-\renewcommand{\floatpagefraction}{0.9}
-
-\usepackage{caption}
-\DeclareCaptionStyle{italic}[justification=centering]
- {labelfont={bf},textfont={it},labelsep=colon}
-\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
-\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
-
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-
-% theorem in latex
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}{Lemma}[section]
-\newtheorem{corollary}{Corollary}[section]
-\newtheorem{proposition}{Proposition}[section]
-\newtheorem{conjecture}{Conjecture}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Exercise}[section]
-\theoremstyle{definition}
-\newtheorem{hypothesis}{Hypothesis}[section]
-\theoremstyle{remark}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
+\def\sectionautorefname{Section}
+\def\subsectionautorefname{Section}
 
@@
 \maketitle
-\begin{keywords}
-manifold learning
-\end{keywords}
 
@@
 
-In a famous paper, \textcite{BC64} introduced a family of transformations \dots
+\begin{verbatim}
+Background (anomaly detection with density) with summarized contribution
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{qplot}\NormalTok{(}\FunctionTok{exp}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\DecValTok{200}\NormalTok{)))}
-\end{Highlighting}
-\end{Shaded}
+    Anomaly detection on very high-dimensional data. Applications. Main contribution.
+
+Anomaly detection using densities, other methods for anomaly detection. Based manifold learning embeddings, density estimates
+
+Density estimation methods, KDE with distortion, 2-D variable kernel denstiy estimate with Riemannian matrix
+
+- Main contributions: better anomaly detection
+
+    fix distortion with Riemmannian matrix
+
+    anomaly detection using density estimates
+
+    kernel density estimate is not accurate
+
+    improve density estimation with distortion in the embedding plot
+\end{verbatim}
+
+The rest of the paper is organized as follows.
+In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
+In this section, we provide justification for the use of Riemannien matrix as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
+\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data and the second with a 100-dimensional meta data.
+\autoref{application} contains the application to visualize and identify anomalies in the {[}TODO{]} data.
+Conclusions and discussions are presented in \autoref{conclusion}.
+
+\hypertarget{vkde}{%
+\section{Variable kernel density estimation with manifold embeddings}\label{vkde}}
+
+\begin{verbatim}
+Idea: whole process for the main contribution
+
+Kernel density estimate with variable bandwidth
+\end{verbatim}
+
+\hypertarget{two-dimensional-kernel-density-estimation}{%
+\subsection{Two dimensional kernel density estimation}\label{two-dimensional-kernel-density-estimation}}
+
+In general a multivariate kernel density estimate looks something like this
+
+\[
+\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)
+\]
+
+where if a Gaussian kernel is used
+
+\[
+K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
+\]
+
+The matrix \(\mathbf{H}\) is called the bandwidth matrix and is very important.
+You will often read that the bandwidth matrix is about smoothing and it is.
+But an alternative way to think of kernel estimation is that kernel densities `borrow strength' from nearby points and the bandwidth determines what is ``nearby''.
+If the bandwidth is large then all points are ``nearby'' and we get an overly smooth kernel density estimate.
+The interesting thing about a bandwidth matrix is that it allows for different notions of what is ``nearby'' along different coordinates and even along diagonal directions.
+
+\begin{verbatim}
+Riemannian matrix could be used to measure distortion
+
+Use riemannian matrix as variable bandwidth
+    Bandwidth is a symmetric positive-definite square matrix, similar to riemannian matrix
+\end{verbatim}
+
+\hypertarget{use-riemannian-matrix-as-variable-bandwidth}{%
+\subsection{Use riemannian matrix as variable bandwidth}\label{use-riemannian-matrix-as-variable-bandwidth}}
+
+The Riemannian estimated using the method of Perrault Joncas and Meila gives some idea of the distortion of an embedding (or so they claim).
+Mapping the points through a non-linear function ``stretchs'' some regions of space and ``shrinks'' others.
+The Riemannian gives us an idea of the direction and angle of this stretching.
+The Riemannian is quite a technical concept but an important thing to understand is that the estimate that comes out of Perrault Joncas algorithm is a square matrix.
+
+We saw how points that are far apart in the embedding may not have been so far apart on the original manifold.
+The Riemannian gives us some way of correcting this.
+Similarly a bandwidth matrix in a kernel density estimate is all about determining the ``directions'' in which there should be more or less ``closeness''.
+So the basic idea is to replace the kernel density estimate with
+
+\[
+\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
+K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
+\]
+
+where \(H_i\) is either the Riemannian or the inverse of the Riemannian (I am not totally sure which one).
+Notice that the bandwidth matrix is different for each point.
+This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.
+There is a paper by Terrel and Scott from Annals of Statistics in 1992 where they talk about this (it is called ``variable kernel density estimation'').
+Some other search terms to find relevant papers might be ``adaptive'' or ``varying bandwidth''.
+I don't know if anyone has had the idea of bringing all these ideas together with manifold learning and the estimate of the Riemannian - we should do a thorough literature search.
+However, if no one has already done this I think it is a very interesting way to do anomaly detection for very high dimensional data.
+
+\begin{verbatim}
+    - Manifold learning distortion
+
+        dimension reduction using manifold learning
+        Riemannian matrix to measure topological distortion
+        distortion is informative for true embedding
+\end{verbatim}
+
+The aim of the project is not to evaluate different manifold learning algorithms.
+Although ISOMAP worked OK here it is still distorted and in real examples you will never `know' that the data are uniform on the sphere.
+Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+
+\hypertarget{simulation}{%
+\section{Simulation}\label{simulation}}
+
+To examine our proposed algorithm, we now run two scenarios for both low and high dimensions.
+For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
+We first simulated the data of size \(N=2000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
+Different mapping functions are then applied to the 2-D meta data so that they now lie on a 3-D feature space, which gives the input data for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+The embedded dimension is set as \(D=2\), same as the meta data dimension.
+This enables us to compare the manifold learning embedding with the true meta data.
+We could now apply the algorithm described in {[}TODO{]}.
+Similarly, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is \(D=5\).
+
+\hypertarget{twodgaussian}{%
+\subsection{2-D Meta data from a Gaussian Mixture Model}\label{twodgaussian}}
+
+We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model of four kernels.
+The means of four kernels are \((0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)\) and the variance-covariance matrix is \(\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)\).
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points, with brighter colors showing high densities and dark colors showing low densities.
+We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
+Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
+This is the baseline to be compared with outliers from variable kernel density estimates.
 
 \begin{figure}
-\centering
-\includegraphics{paper_files/figure-latex/histogram-1.pdf}
-\caption{\label{fig:histogram}Nice histogram}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_4kernels}
+
+}
+
+\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and variance-covariance matrix $\big(\begin{smallmatrix} 0.02&0\ 0&0.02 \end{smallmatrix}\big)$}\label{fig:metadensity}
 \end{figure}
 
-Consider the logNormal data plotted in Figure \ref{fig:histogram}.
+Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
+One of the most famous example in manifold learning is the swiss roll data, with the mapping function as below.
+\[
+\left\{ 
+\begin{array}{lcl}
+X = X_1 \times \cos{X_1}, \\
+Y = X_2, \\
+Z = X_1 \times \sin{X_1}.
+\end{array}
+\right.
+\]
+Now we are able to apply different manifold learning algorithms to reduce the dimension back to 2, and further estimate the density of the embedding to detect anomalies.
+According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
+1\%, 50\%, 99\%, \textgreater99\%.
+The top outliers with lowest densities are also indexed in blue, which we call \emph{true outliers}.
+For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents in one column the 2-D embedding plot, with the colors indicating the densities, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
+Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
+As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as ISOMAP and UMAP.
+Instead, we could roughly compare the location of outliers.
+For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
 
-\[s^2 = \sum_{i=1}^n (x_i-\bar{x})^2\]
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/outliers_comparison_4ml_3cases}
+
+}
+
+\caption{Highest density region plots of four manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from top: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and bottom: KDE with fixed bandwidth.
+Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE.
+Both methods are not detecting outliers accurately when using UMAP.}\label{fig:sroutliers}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/mappings_sr_tp}
+
+}
+
+\caption{3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
+\end{figure}
+
+For comparison, we use the same meta data with different mapping function, twin peak mapping, with the corresponding 3-D data is shown in \autoref{fig:mappings}.
+The four colors in both mappings represents the four gaussian kernels in the meta data.
+Similar to \autoref{fig:sroutliers}, four manifold learnings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/twin_peak_outliers_comparison_4ml_3cases}
+
+}
+
+\caption{Highest density region plots of four manifold learning embeddings of the twin peak data.
+Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.}\label{fig:tpoutliers}
+\end{figure}
+
+\hypertarget{fivedgaussian}{%
+\subsection{5-D data embedding in 100-D space}\label{fivedgaussian}}
+
+4-D from mixture Gaussian kernels (99\% from N(0,1), 1\% from N(0, 5) to add outliers), 1-D from semi-sphere, all other 95-D as 0, then rotate using the R matrix from QR decomposition to get the 100-D input for manifold learning)
+
+\hypertarget{application}{%
+\section{Application}\label{application}}
+
+\begin{verbatim}
+Dataset
+
+Goal and process
+
+Results (Plots and tables)
+
+Analysis and summary
+\end{verbatim}
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+\begin{verbatim}
+Summary of the contribution and conclusion
+
+Details about the process
+
+Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate)
+\end{verbatim}
 

### 2022-02-09 paper/paper.Rmd
Simulation - 2d

@@
 options(tinytex.verbose = TRUE)
+library(tidyverse)
 library(dimRed)
-library(tidyverse)
-library(ggplot2)
-library(reshape2)
-library(knitr)
-library(kableExtra)
-library(tibble)
+library(reticulate)
+library(here)
+library(viridis)
+library(hdrcde)
+library(igraph)
+library(matrixcalc)
+library(akima)
+library(car)
+library(ggforce)
+library(ks)
 library(patchwork)
-library(data.table)
-library(dtplyr)
-library(igraph)
-library(ggraph)
-# set.seed(1)
+library(copula)
+library(plotly)
 Jmisc::sourceAll(here::here("R/sources"))
+# set.seed(123)
 ```
@@
 
-## Using dimension reduction algorithms
+	Background (anomaly detection with density) with summarized contribution
 
-Now we can try to see what happens when we use an actual dimension reduction algorithm, for instance ISOMAP
+		Anomaly detection on very high-dimensional data. Applications. Main contribution.
 
-```{r isomap, echo=TRUE,message=FALSE}
-dat<-dimRedData(cbind(x,y,z)) #Prepare data to use in dimRed
-emb<-embed(dat,"Isomap") #Isomap embedding
-plot(hdrscatterplot(emb@data@data[,1],emb@data@data[,2])) #HDR plot
-```
+	Anomaly detection using densities, other methods for anomaly detection. Based manifold learning embeddings, density estimates
 
-Which seems to not distort things too much.  The HDR plot still identifies outliers (because it has to) but they are just scattered around the corners in a very non-systematic way.
+	Density estimation methods, KDE with distortion, 2-D variable kernel denstiy estimate with Riemannian matrix
 
-For something that leads to a lot of distortion have a look t t-SNE.
+	- Main contributions: better anomaly detection
 
-```{r tsneplot, echo=TRUE,message=FALSE}
-emb_tsne<-embed(dat,"tSNE") #tSNE embedding
-plot(hdrscatterplot(emb_tsne@data@data[,1],emb_tsne@data@data[,2])) #HDE plot
-```
+		fix distortion with Riemmannian matrix
 
-The aim of the project is not to evaluate different manifold learning algorithms.
-Although ISOMAP worked OK here it is still distorted and in real examples you will never 'know' that the data are uniform on the sphere.
-Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+		anomaly detection using density estimates
+
+		kernel density estimate is not accurate
+
+		improve density estimation with distortion in the embedding plot
+
+The rest of the paper is organized as follows.
+In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
+In this section, we provide justification for the use of Riemannien matrix as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
+\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data and the second with a 100-dimensional meta data.
+\autoref{application} contains the application to visualize and identify anomalies in the [TODO] data.
+Conclusions and discussions are presented in \autoref{conclusion}.
+
+# Variable kernel density estimation with manifold embeddings {#vkde}
+
+	Idea: whole process for the main contribution
+
+	Kernel density estimate with variable bandwidth
+
+## Notations
 
@@
 
-## Using the Riemannian
+    Riemannian matrix could be used to measure distortion
+
+    Use riemannian matrix as variable bandwidth
+        Bandwidth is a symmetric positive-definite square matrix, similar to riemannian matrix
+
+## Use riemannian matrix as variable bandwidth
 
@@
 This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.
-There is a paper by Terrel and Scott from Annals of Statistics in 1992 where they talk about this (it is called "variable kernel density estimation").
+<!-- There is a paper by Terrel and Scott from Annals of Statistics in 1992 where they talk about this (it is called "variable kernel density estimation").
 Some other search terms to find relevant papers might be "adaptive" or "varying bandwidth".
@@
 However, if no one has already done this I think it is a very interesting way to do anomaly detection for very high dimensional data.
+-->
+
+    	- Manifold learning distortion
+
+    		dimension reduction using manifold learning
+    		Riemannian matrix to measure topological distortion
+    		distortion is informative for true embedding
+
+The aim of the project is not to evaluate different manifold learning algorithms.
+Although ISOMAP worked OK here it is still distorted and in real examples you will never 'know' that the data are uniform on the sphere.
+Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+
+# Simulation {#simulation}
+
+To examine our proposed algorithm, we now run two scenarios for both low and high dimensions.
+For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
+We first simulated the data of size $N=2000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
+Different mapping functions are then applied to the 2-D meta data so that they now lie on a 3-D feature space, which gives the input data for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+The embedded dimension is set as $D=2$, same as the meta data dimension.
+This enables us to compare the manifold learning embedding with the true meta data.
+We could now apply the algorithm described in [TODO].
+Similarly, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is $D=5$.
+
+## 2-D Meta data from a Gaussian Mixture Model {#twodgaussian}
+
+We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model of four kernels.
+The means of four kernels are $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the variance-covariance matrix is $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$.
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points, with brighter colors showing high densities and dark colors showing low densities.
+We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
+Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
+This is the baseline to be compared with outliers from variable kernel density estimates.
+
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and variance-covariance matrix $\\big(\\begin{smallmatrix} 0.02&0\\ 0&0.02 \\end{smallmatrix}\\big)$", echo = FALSE}
+knitr::include_graphics("figures/truedensity_4kernels.png")
+```
+
+Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
+One of the most famous example in manifold learning is the swiss roll data, with the mapping function as below.
+$$
+\left\{ 
+\begin{array}{lcl}
+X = X_1 \times \cos{X_1}, \\
+Y = X_2, \\
+Z = X_1 \times \sin{X_1}.
+\end{array}
+\right.
+$$
+Now we are able to apply different manifold learning algorithms to reduce the dimension back to 2, and further estimate the density of the embedding to detect anomalies.
+According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
+1%, 50%, 99%, >99%.
+The top outliers with lowest densities are also indexed in blue, which we call *true outliers*.
+For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents in one column the 2-D embedding plot, with the colors indicating the densities, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
+Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
+As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as ISOMAP and UMAP.
+Instead, we could roughly compare the location of outliers.
+For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+
+```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of four manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from top: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and bottom: KDE with fixed bandwidth.
+Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE.
+Both methods are not detecting outliers accurately when using UMAP.", echo = FALSE}
+knitr::include_graphics("figures/outliers_comparison_4ml_3cases.png")
+```
+
+```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.", echo = FALSE}
+knitr::include_graphics("figures/mappings_sr_tp.png")
+```
+
+For comparison, we use the same meta data with different mapping function, twin peak mapping, with the corresponding 3-D data is shown in \autoref{fig:mappings}.
+The four colors in both mappings represents the four gaussian kernels in the meta data.
+Similar to \autoref{fig:sroutliers}, four manifold learnings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+
+```{r tpoutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of four manifold learning embeddings of the twin peak data.
+Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.", echo = FALSE}
+knitr::include_graphics("figures/twin_peak_outliers_comparison_4ml_3cases.png")
+```
+
+<!-- Multiple mapping functions: swiss roll, twin peak -->
+
+<!-- Manifold learning results for the mapped 3-D data -->
+
+<!--     ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, UMAP (useful ones, and the one with too severe distortion) -->
+
+<!-- Outlier plot based on density estimates -->
+
+<!--     Fixed bandwidth -->
+
+<!--     Pointwise adaptive bandwidth using Riemmanien -->
+
+<!--     Comparison -->
+
+<!--         Rank of densities with the true density (f=f_1/4 + f_2/4 + f_3/4 + f_4/4) -->
+<!--         Sum of squared errors in the density estimates -->
+
+## 5-D data embedding in 100-D space {#fivedgaussian}
+
+4-D from mixture Gaussian kernels (99% from N(0,1), 1% from N(0, 5) to add outliers), 1-D from semi-sphere, all other 95-D as 0, then rotate using the R matrix from QR decomposition to get the 100-D input for manifold learning)
+
+# Application {#application}
+
+	Dataset
+
+	Goal and process
+
+	Results (Plots and tables)
+
+	Analysis and summary
+
+# Conclusion {#conclusion}
+
+	Summary of the contribution and conclusion
+
+	Details about the process
+
+	Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate)
+
+<!-- ## Using dimension reduction algorithms -->
+
+<!-- Now we can try to see what happens when we use an actual dimension reduction algorithm, for instance ISOMAP -->
+
+<!-- ```{r isomap, echo=TRUE,message=FALSE, eval=FALSE} -->
+<!-- dat<-dimRedData(cbind(x,y,z)) #Prepare data to use in dimRed -->
+<!-- emb<-embed(dat,"Isomap") #Isomap embedding -->
+<!-- plot(hdrscatterplot(emb@data@data[,1],emb@data@data[,2])) #HDR plot -->
+<!-- ``` -->
+
+<!-- Which seems to not distort things too much.  The HDR plot still identifies outliers (because it has to) but they are just scattered around the corners in a very non-systematic way.  -->
+
+<!-- For something that leads to a lot of distortion have a look t t-SNE. -->
+
+<!-- ```{r tsneplot, echo=TRUE,message=FALSE, eval=FALSE} -->
+<!-- emb_tsne<-embed(dat,"tSNE") #tSNE embedding -->
+<!-- plot(hdrscatterplot(emb_tsne@data@data[,1],emb_tsne@data@data[,2])) #HDE plot -->
+<!-- ``` -->

### 2022-02-09 reports/Simulation_swissroll.R
Code for 2d simulation

@@ -25,21 +25,21 @@ set.seed(1)
 N <- 2000
 p <- 2
 
-# # ----Swiss roll dataset with a sparse area------------------------
-# height <- 30
-# x <- (3 * pi / 2) * (1 + 2 * runif(N*1.05, 0, 1))  
-# y <- height * runif(N*1.05, 0 , 1)
-# sr <- cbind(x * cos(x), y, x * sin(x))
-# colnames(sr) <- c("x", "y", "z")
-# sr <- cbind(as_tibble(sr), col = "grey")
-# lowden <- x >= 13 & y >= 18
-# sr$col[lowden] <- "red"
-# removed <- sample(which(lowden), size = N*0.05, replace = FALSE)
-# lowdenpoints <- sr[setdiff(which(lowden), removed), ]
-# sr <- sr[-removed, ]
-# scatterplot3d::scatterplot3d(sr, color = sr$col)
-# # rgl::plot3d(sr, col = sr$col)
-# 
+# ----Swiss roll dataset with a sparse area------------------------
+height <- 30
+x <- (3 * pi / 2) * (1 + 2 * runif(N*1.05, 0, 1))
+y <- height * runif(N*1.05, 0 , 1)
+sr <- cbind(x * cos(x), y, x * sin(x))
+colnames(sr) <- c("x", "y", "z")
+sr <- cbind(as_tibble(sr), col = "grey")
+lowden <- x >= 13 & y >= 18
+sr$col[lowden] <- "red"
+removed <- sample(which(lowden), size = N*0.05, replace = FALSE)
+lowdenpoints <- sr[setdiff(which(lowden), removed), ]
+sr <- sr[-removed, ]
+scatterplot3d::scatterplot3d(sr, color = sr$col)
+# rgl::plot3d(sr, col = sr$col)
+
 
 
 # ----Swiss roll from copula-----------------------------------------
@@ -117,20 +117,36 @@ preswissroll <- preswissroll %>%
 a <- preswissroll$X1
 swissroll <- tibble(x = a * cos(a), y = preswissroll$X2, z = a * sin(a))
 
-# plotting
-preswissroll %>%
+
+
+###--------------------------------------------------------
+## Start from here! 
+###--------------------------------------------------------
+set.seed(1234)
+mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3]
+sr <- mldata(N = 2000, meta = "gaussian", mapping = mapping)
+swissroll <- sr$data %>% as.data.frame()
[…]

### 2022-02-09 R/sources/hdrplotting.R
Functions for kde to be improved

@@ -76,7 +76,7 @@ hdrscatterplot_new <- function(x, y, levels = c(1, 50, 99), kde.package = c("ash
   
   if (noutliers > 0) {
     outlier_rank <- order(den$fxy[ord])  # order den$fxy as well to match the new data
-    outliers <- outlier_rank[seq(noutliers)] # take top noutliers labels for annotation
+    outliers <- outlier_rank[seq_len(noutliers)] # take top noutliers labels for annotation
   }
   
   p <- ggplot2::ggplot(data, ggplot2::aes_string(vnames[1], vnames[2])) +
@@ -104,7 +104,8 @@ hdrscatterplot_new <- function(x, y, levels = c(1, 50, 99), kde.package = c("ash
   return(list(p=p, 
               # outlier= rownames(data)[outlier_rank], densities = den$fxy[ord]
               outlier = order(den$fxy, decreasing = FALSE),  # densities ascending, top anomalous to typical
-              densities = den$fxy # density order matching the data inpput
+              densities = den$fxy, # density order matching the data input
+              den = den # list of x, y, z(matrix)
               )
          ) ## TODO: check outlier/density order
 }
@@ -137,4 +138,39 @@ hdr.2d <- function(x, y, prob = c(50, 95, 99), den=NULL, kde.package=c("ash","ks
   index <- which.max(fxy)
   mode <- c(x[index],y[index])
   return(structure(list(mode=mode,falpha=falpha,fxy=fxy, den=den, alpha=alpha, x=x, y=y), class="hdr2d"))
+}
+
+
+
+# multivariate density estimate
+# Modified from hdrcde package https://github.com/robjhyndman/hdrcde/blob/master/R/hdr.boxplot.2d.R
+
+den.estimate.5d <- function(x, kde.package=c("ks", "ash"), h=NULL, xextend=0.15) {
+  kde.package <- match.arg(kde.package)
+  # Find ranges for estimates
+  xr <- apply(x, 2, function(x) diff(range(x, na.rm = TRUE)))
+  # xr <- diff(range(x,na.rm=TRUE))
+  # yr <- diff(range(y,na.rm=TRUE))
+  xr <- c(min(x)-xr*xextend,max(x)+xr*xextend)
+  # yr <- c(min(y)-yr*yextend,max(y)+yr*yextend)
[…]

### 2022-02-09 R/sources/vkde.R

@@ -114,7 +114,7 @@ vkde2d <- function(x, y, h, n = 25, lims = c(range(x), range(y)) ){
     z[,,k] <- det(hk) ^ (-1/2) * matrix(mvtnorm::dmvnorm(x = g, mean = xi[k,], sigma = hk), nrow = n[1], byrow = FALSE) # scalar
   }
   
-  z <- rowMeans(z, dims = 2)
+  z <- rowMeans(z, dims = 2, na.rm = TRUE)
 
     # hkd <- 1 / (2 * pi) * det(h[,,k])^(-1/2)
     # hks <- solve(h[,,k])

### 2022-02-09 R/sources/mldata.R
Functions for kde to be improved

@@ -11,11 +11,15 @@
 #' @export
 #'
 #' @examples
+#' library(copula)
 #' a <- mldata(N = 100, meta = "copula", mapping = "Swiss Roll")
 #' library(ggplot2)
 #' ggplot(as_tibble(a$data), aes(x, y)) + geom_point(aes(color = a$colors))
 #' b <- mldata(meta = "gaussian", mapping = "S Curve")
 #' ggplot(as_tibble(b$data), aes(x, y)) + geom_point(aes(color = b$colors))
+#' dat <- as.data.frame(b$data)
+#' library(scatterplot3d)
+#' scatterplot3d(x = dat$x, y = dat$y, z = dat$z, color = b$colors)
 mldata <- function(N = 2000, p = 2, meta = c("uniform", "copula", "gaussian"), 
                    mapping = c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")) {
   switch (meta,
@@ -31,34 +35,40 @@ mldata <- function(N = 2000, p = 2, meta = c("uniform", "copula", "gaussian"),
           },
           "gaussian" = {
             n <- round(N/4)
-            R <- matrix(c(1, 0,
-                          0, 1), 
+            R <- matrix(c(.02, 0,
+                          0, .02), # variance-covariance matrix
                         nrow = 2, ncol = 2)
-            # mu1 <- c(7.5, 7.5)
-            # mu2 <- c(7.5, 12.5)
-            # mu3 <- c(12.5, 7.5)
-            # mu4 <- c(12.5, 12.5)
-            mu <- tribble(~mu1, ~mu2,
-                          7.5, 7.5,
-                          7.5, 12.5,
-                          12.5, 7.5,
-                          12.5, 12.5)
+            # mu <- matrix(#~mu1, ~mu2,
+            #               c(7.5, 7.5,
+            #               7.5, 12.5,
+            #               12.5, 7.5,
+            #               12.5, 12.5), 
+            #               4, 2, byrow = TRUE)
+            mu <- matrix(#~mu1, ~mu2,
+              c(0.25, 0.25,
+                0.25, 0.75,
+                0.75, 0.25,
+                0.75, 0.75), 
+              4, 2, byrow = TRUE)
             co <- NULL
             for(i in 1:4) {
-              # mui <- get(paste0("mu", i))
-              mui <- as.matrix(mu[i,])
-              a <- mvtnorm::rmvnorm(n, mean = mui, sigma = R)
[…]

### 2022-02-09 R/app.R
Add true density plot in second tab

@@ -12,7 +12,7 @@ library(ggforce)
 library(patchwork)
 library(copula)
 Jmisc::sourceAll(here::here("R/sources"))
-# set.seed(123)
+set.seed(1)
 
 ml_outlier <- function(x, s = 2, k = min(10, nrow(x)), radius = 0, 
                        adjacency = NULL, affinity = NULL,
@@ -55,22 +55,27 @@ ml_outlier <- function(x, s = 2, k = min(10, nrow(x)), radius = 0,
   # prob <- c(1, 50, 99)
   f_vkde <- vkde2d(x = E1, y = E2, h = Rn, n = n.grid) # estimated densities with variable bandwidth
   # fxy <- hdrcde:::interp.2d(f_vkde$x, f_vkde$y, f_vkde$z, x0 = E1, y0 = E2) # linear interpolation
+  den <- hdrcde:::den.estimate.2d(x = E1, y = E2, kde.package = "ks", xextend=0.15, yextend = 0.15)
   
   p_vkde <- plot_outlier(x = metriclearn, n.grid = n.grid, prob = prob, noutliers = noutliers, scales = scales, ell_size = ell_size)
   p_hdr <- hdrscatterplot_new(E1, E2, kde.package = "ks", noutliers = noutliers)
   p_hdr$p <- p_hdr$p +
     plot_ellipse(metriclearn, n.plot = n.plot, add = TRUE, ell_size = ell_size)
 
-  p_contour <- filled.contour(f_vkde, color.palette = viridis, # p_vkde$hdr2d_info$den
-                      plot.axes = { axis(1); axis(2);
-                        points(fn, pch = 3, col= hcl(c=20, l = 8))})
-  p_contour_hdr <- filled.contour(p_hdr$densities, color.palette = viridis, # p_vkde$hdr2d_info$den
-                      plot.axes = { axis(1); axis(2);
-                        points(fn, pch = 3, col= hcl(c=20, l = 8))})
+  # p_contour <- filled.contour(f_vkde, color.palette = viridis, # p_vkde$hdr2d_info$den
+  #                     plot.axes = { axis(1); axis(2);
+  #                       points(fn, pch = 3, col= hcl(c=20, l = 8))})
+  # p_contour_hdr <- filled.contour(den, color.palette = viridis, # p_vkde$hdr2d_info$den
+  #                     plot.axes = { axis(1); axis(2);
[…]

### 2022-02-13 R/multidimension.R
Simulation

@@ -34,16 +34,16 @@ x1 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
 x2 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
 x3 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
 x4 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
-x <- cbind(x1, x2, x3, x4)
-head(x)
-range(x)
+X <- cbind(x1, x2, x3, x4)
+head(X)
+range(X)
 
 # GMM density
-den_calc <- function(x) {
+gmmdensity <- function(x) {
   p1 * dmvnorm(x, rep(mu1, 4), diag(s1, 4)) + 
     p2 *  dmvnorm(x, rep(mu2, 4), diag(s2, 4))
 }
-den <- apply(x, 1, den_calc)
+den <- apply(X, 1, gmmdensity)
 length(den)
 range(den)
 summary(den)
@@ -54,19 +54,19 @@ summary(den)
 # scales::rescale(x, c(0,1))
 r <- ceiling(max(sqrt(x1^2 + x2^2 + x3^2 + x4^2)))
 r
-range(x)
+range(X)
 
 # x_new <- cbind(x/r, 
 #            x5 = sqrt(1 - (x1^2 + x2^2 + x3^2 + x4^2)/r^2),
 #            matrix(0, N, 95)
 #            )
-x_new <- cbind(x, 
+X_new <- cbind(X, 
                x5 = sqrt(r^2 - (x1^2 + x2^2 + x3^2 + x4^2)),
                matrix(0, N, 95)
 )
 # x0 <- x
-head(x_new)
-x_new %>% as_tibble() %>% summarise(a = x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
+head(X_new)
+X_new %>% as_tibble() %>% summarise(a = x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
 
 # den1 <- apply(x_new[,1:4], 1, den_calc)
 # summary(den1)
@@ -105,10 +105,10 @@ Q[1:10,1:10]
 # A rotation preserves, not just ratios, but distances themselves.
 # x'x = (Qx)'(Qx) = x'Q'Qx = x'Ix = x'x
 
-train <- t(Q %*% t(x_new))
[…]

### 2022-02-13 reports/Simulation_swissroll.R
Simulation

@@ -25,97 +25,97 @@ set.seed(1)
 N <- 2000
 p <- 2
 
-# ----Swiss roll dataset with a sparse area------------------------
-height <- 30
-x <- (3 * pi / 2) * (1 + 2 * runif(N*1.05, 0, 1))
-y <- height * runif(N*1.05, 0 , 1)
-sr <- cbind(x * cos(x), y, x * sin(x))
-colnames(sr) <- c("x", "y", "z")
-sr <- cbind(as_tibble(sr), col = "grey")
-lowden <- x >= 13 & y >= 18
-sr$col[lowden] <- "red"
-removed <- sample(which(lowden), size = N*0.05, replace = FALSE)
-lowdenpoints <- sr[setdiff(which(lowden), removed), ]
-sr <- sr[-removed, ]
-scatterplot3d::scatterplot3d(sr, color = sr$col)
-# rgl::plot3d(sr, col = sr$col)
-
-
-
-# ----Swiss roll from copula-----------------------------------------
-# ## a 3-dimensional normal copula
-# norm.cop <- normalCopula(0.5, dim = p)
-# u <- rCopula(N, norm.cop)
-# den <- dCopula(u, norm.cop)
-
-# # 3-d t copula
-# t.cop <- tCopula(c(0.5, 0.3), dim = p, dispstr = "toep",
-#                  df = 2, df.fixed = TRUE)
-# u <- rCopula(N, t.cop)
-# den <- dCopula(u, t.cop)
+# # ----Swiss roll dataset with a sparse area------------------------
+# height <- 30
+# x <- (3 * pi / 2) * (1 + 2 * runif(N*1.05, 0, 1))
+# y <- height * runif(N*1.05, 0 , 1)
+# sr <- cbind(x * cos(x), y, x * sin(x))
+# colnames(sr) <- c("x", "y", "z")
+# sr <- cbind(as_tibble(sr), col = "grey")
+# lowden <- x >= 13 & y >= 18
+# sr$col[lowden] <- "red"
+# removed <- sample(which(lowden), size = N*0.05, replace = FALSE)
+# lowdenpoints <- sr[setdiff(which(lowden), removed), ]
+# sr <- sr[-removed, ]
+# scatterplot3d::scatterplot3d(sr, color = sr$col)
+# # rgl::plot3d(sr, col = sr$col)
 # 
-## a 2-dimensional Clayton copula
-cl3 <- claytonCopula(2, dim = p)
[…]

### 2022-02-13 paper/paper.Rmd
DGP done with unified gmm notations

@@
 
-To examine our proposed algorithm, we now run two scenarios for both low and high dimensions.
+In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm.
 For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulated the data of size $N=2000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
+We first simulate the data of size $N=2000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
 Different mapping functions are then applied to the 2-D meta data so that they now lie on a 3-D feature space, which gives the input data for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
@@
 
-We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model of four kernels.
-The means of four kernels are $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the variance-covariance matrix is $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$.
+We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with means $\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75), \pmb{\mu_3}=(0.75, 0.25), \pmb{\mu_4}(0.75, 0.75)$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$.
+The mixture proportions are equally set as $\pi_i=0.25, i=1,2,3,4$.
+<!-- $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$. -->
+Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
+\begin{equation}
+\label{eq:gmm}
+P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
+\end{equation}
 \autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points, with brighter colors showing high densities and dark colors showing low densities.
@@
 
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and variance-covariance matrix $\\big(\\begin{smallmatrix} 0.02&0\\ 0&0.02 \\end{smallmatrix}\\big)$", echo = FALSE}
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$", echo = FALSE}
 knitr::include_graphics("figures/truedensity_4kernels.png")
@@
 
-4-D from mixture Gaussian kernels (99% from N(0,1), 1% from N(0, 5) to add outliers), 1-D from semi-sphere, all other 95-D as 0, then rotate using the R matrix from QR decomposition to get the 100-D input for manifold learning)
+For the high-dimensional case, we generate the meta data from a 5-dimensional semi-hypersphere.
+First, we simulate $N=2000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(25,25,25,25)$.
+In order to manually generate anomalies, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
+The fifth dimension is calculated to satisfy $x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2$ where $r$ is set as 15.
+Similarly, the Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size $N$ (denote the transpose of the data matrix as $X$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+The rotation matrix is derived from the QR decomposition of a $100\times N$ matrix $A$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
+For any real matrix $A$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $Q$ and $R$ so that $A = QR$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $Q^\prime Q = I$, and $R$ is an upper triangular matrix.
+Matrix $Q$ is used as the rotation matrix and it satisfies
+$X^\prime X=(QX)^\prime(QX)$, where the rotated data matrix $(QX)^\prime$ is now the input for the manifold learning algorithms.
 

### 2022-02-13 paper/paper.tex
DGP done with unified gmm notations

@@
 \def\subsectionautorefname{Section}
+\usepackage{bm}
 
@@
 \end{verbatim}
+
+\hypertarget{notations}{%
+\subsection{Notations}\label{notations}}
 
@@
 Notice that the bandwidth matrix is different for each point.
-This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.
-There is a paper by Terrel and Scott from Annals of Statistics in 1992 where they talk about this (it is called ``variable kernel density estimation'').
-Some other search terms to find relevant papers might be ``adaptive'' or ``varying bandwidth''.
-I don't know if anyone has had the idea of bringing all these ideas together with manifold learning and the estimate of the Riemannian - we should do a thorough literature search.
-However, if no one has already done this I think it is a very interesting way to do anomaly detection for very high dimensional data.
+This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.\\
 
@@
 
-To examine our proposed algorithm, we now run two scenarios for both low and high dimensions.
+In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm.
 For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulated the data of size \(N=2000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
+We first simulate the data of size \(N=2000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
 Different mapping functions are then applied to the 2-D meta data so that they now lie on a 3-D feature space, which gives the input data for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
@@
 
-We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model of four kernels.
-The means of four kernels are \((0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)\) and the variance-covariance matrix is \(\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)\).
+We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75), \pmb{\mu_3}=(0.75, 0.25), \pmb{\mu_4}(0.75, 0.75)\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
+The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
+Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
+\begin{equation}
+\label{eq:gmm}
+P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
+\end{equation}
 \autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points, with brighter colors showing high densities and dark colors showing low densities.
@@
 
-\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and variance-covariance matrix $\big(\begin{smallmatrix} 0.02&0\ 0&0.02 \end{smallmatrix}\big)$}\label{fig:metadensity}
+\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$}\label{fig:metadensity}
 \end{figure}
@@
 
-4-D from mixture Gaussian kernels (99\% from N(0,1), 1\% from N(0, 5) to add outliers), 1-D from semi-sphere, all other 95-D as 0, then rotate using the R matrix from QR decomposition to get the 100-D input for manifold learning)
+For the high-dimensional case, we generate the meta data from a 5-dimensional semi-hypersphere.
+First, we simulate \(N=2000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(25,25,25,25)\).
+In order to manually generate anomalies, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
+The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as 15.
+Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size \(N\) (denote the transpose of the data matrix as \(X\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+The rotation matrix is derived from the QR decomposition of a \(100\times N\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
+For any real matrix \(A\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(Q\) and \(R\) so that \(A = QR\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(Q^\prime Q = I\), and \(R\) is an upper triangular matrix.
+Matrix \(Q\) is used as the rotation matrix and it satisfies
+\(X^\prime X=(QX)^\prime(QX)\), where the rotated data matrix \((QX)^\prime\) is now the input for the manifold learning algorithms.
 

### 2022-02-20 paper/paper.Rmd
Add proposed algorithm and learn metric alg

@@
 library(patchwork)
-library(copula)
+# library(copula)
 library(plotly)
@@
 
+In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space.
+One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
+These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co].
+To tackle this problem and measure the distortion incurred in manifold learning, @Perrault-Joncas2013-pq propose the Metric Learning algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+By applying the Metric Learning algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
+
+The Riemannian metric defined at each point of the manifold is used to compute the geometric quantities, including angle, length, and volume, of the low-dimensional manifold embedding in any coordinate system, and be further applied to correct the distortion caused by the manifold learning algorithms.
+In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point. 
+Therefore, if we could replace the bandwidth matrix with the Riemannian metric, we could further get the kernel density estimation of the manifold $\mathcal{M}$.
+This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for anomaly detection.
+The proposed schematic is shown in Figure \@ref(fig:vkde).
+
+(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
+```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=FALSE}
+knitr::include_graphics(here::here("report/vkde.png"))
+```
+
+Now I briefly describe different steps involved in this main chapter.
+By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_\mathcal{M}$ is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_\mathcal{M}$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.
+The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $H_i$ with the Riemannian metric $g_i$ for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\mathbb{x})$, for anomaly detection.
+
 The rest of the paper is organized as follows.
@@
 
+In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
+For a high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get a low-dimensional embedding.
+The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others.
+@Perrault2009 gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point.
+To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
+The outliers could then be defined as the points with lowest density estimates.
+
+To start with, we introduce the notations in this manuscript.
+Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metrix.
+Finally, we propose our novel method to detect anomalies for high-dimensional data set.
+
 ## Notations
@@
 
-In general a multivariate kernel density estimate looks something like this
+<!-- In general a multivariate kernel density estimate looks something like this -->
 
+<!-- $$ -->
+<!-- \hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i) -->
+<!-- $$ -->
+
+<!-- where if a Gaussian kernel is used -->
+
+<!-- $$ -->
+<!-- K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right] -->
+<!-- $$ -->
+
+<!-- The matrix $\mathbf{H}$ is called the bandwidth matrix and is very important.  -->
+
+For a bivariate random sample $\pmb X_1, \pmb X_2, \ldots, \pmb X_n$ 
+drawn from a density $f$, 
+the kernel density estimate is defined by
 $$
-\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)
+\hat{f} (\pmb x; \pmb H) = n^{-1}\sum_{i=1}^n K_{\pmb H} ( \pmb x - \pmb{X}_i)
 $$
+where $\pmb x = (x_1, x_2)^T$ and $\pmb{X}_i = (X_{i1}, X_{i2})^T, i = 1, 2,  
+\ldots, n$.  Here 
+$K(\pmb x)$ is the kernel which is a symmetric probability density function, 
+$\pmb H$ is the bandwidth matrix which is symmetric and positive-definite,  
+and $K_{\pmb H}(\pmb x) = |\pmb H|^{-1/2} K( \pmb H^{-1/2} \pmb x)$. 
+The choice of $K$ is not crucial: we take 
+$K(\pmb x) = (2\pi)^{-1} \exp(-\tfrac{1}{2} \pmb x^T \pmb x)$ the standard normal
+throughout.  
+In contrast, the choice of $\pmb H$ is crucial in determining the performance of $\hat f$. 
+The most common parameterizations of the bandwidth matrix
+are the diagonal and the 
+general or unconstrained which has no restrictions on $\pmb H$
+provided that $\pmb H$ remains positive definite and symmetric, that is 
+$$
+\pmb H = \begin{bmatrix}h_1^2 & 0 \\0 & h_2^2 \end{bmatrix}
+\ \mathrm{or} \ 
+\pmb H = \begin{bmatrix}h_1^2 & h_{12} \\ h_{12}  & h_2^2 \end{bmatrix}.
+$$
+This latter parameterization allows kernels to have an arbitrary orientation
+whereas the former only allows kernels which are oriented to the
+co-ordinate axes.
 
-where if a Gaussian kernel is used
-
-$$
-K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
-$$
-
-The matrix $\mathbf{H}$ is called the bandwidth matrix and is very important.
-You will often read that the bandwidth matrix is about smoothing and it is.
-But an alternative way to think of kernel estimation is that kernel densities 'borrow strength' from nearby points and the bandwidth determines what is "nearby".
+<!-- The bandwidth matrix is about smoothing and it is.  -->
+An alternative way to think of kernel estimation is that kernel densities 'borrow strength' from nearby points and the bandwidth determines what is "nearby".
 If the bandwidth is large then all points are "nearby" and we get an overly smooth kernel density estimate.
@@
 
-The Riemannian estimated using the method of Perrault Joncas and Meila gives some idea of the distortion of an embedding (or so they claim).
-Mapping the points through a non-linear function "stretchs" some regions of space and "shrinks" others.
+The Riemannian estimated using the method of Perrault Joncas and Meila (2009) gives some idea of the distortion of an embedding (or so they claim).
+Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
 The Riemannian gives us an idea of the direction and angle of this stretching.
@@
 
+The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_p\mathcal{M}$ for every point $p \in \mathcal{M}$. 
+If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of manifold.
+The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(\mathcal{M}, g)$.
+
+To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding. 
+The inner product between two vectors $u,v \in T_p\mathcal{M}$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors $\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
+Therefore, for each point $p\in \mathcal{M}$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
+
+[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$
+
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in \mathcal{M}$ is defined as 
+$$
+l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
+$$
+where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $\mathcal{M}$. While the volume of $V\subset \mathcal{M}$ is computed by
+$$
+Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
+$$
+Both the concepts of distance and volume are relevant to kernel density estimation.
+
     	- Manifold learning distortion
@@
 Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+
+### Metric Learning algorithm
+
+@Perrault-Joncas2013-pq propose the Metric Learning algorithm which mainly involves four main steps.
+
+As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension $s$ is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$ will be returned.
+This algorithm is also implemented in a Python library *megaman* [@McQueen2016-xz].
+It is designed to apply the manifold learning methods to large-scale data sets, as well as computing the Riemannian metric of the manifold.
+
+Learn metric algorithm
+
+\begin{algorithm}[!htb]
+  \caption{Learn metric algorithm}
+  \label{alg:learnmetric}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \Input{ high-dimensional data $x_i \in \pmb{R}^p$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $y_i \in \pmb{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
+  \OptParameter{ manifold learning parameters }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^p$;
+
+  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
+  $$
+  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
+  $$
+  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
+
+  \STATE Embed each data point $x\in \mathbb{R}^p$ to embedding coordinates $y(x)=(y^1(x),\dots,y^d(x))$ by any existing manifold learning algorithm;
+
+  \STATE Obtain the matrix $\tilde{h}(x)$ at each point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates $y$ with each element being
+  $$
+    \tilde{h}^{i j}=\frac{1}{2}\left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y_i \cdot y_j\right)-y_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} y_j\right)-y_j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} y_i\right)\right];
+  $$
+
+  \STATE Calculate the Riemannian metric $h(x)$ as the rank $d$ pseudo inverse of $\tilde{h}(x)$ with 
+  $$
+    h(x) = U diag{1/(\Gamma[1:d])} U^\prime,
+  $$
+  where $[U, \Gamma]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Gamma$ in descending order.
+
+  \end{algorithmic}
+\end{algorithm}
+
+## Proposed algorithm
+
+Now we present our proposed method for anomaly detection based on variable kernel density estimates.
+
+\begin{algorithm}[!htb]
+  \caption{Variable kernel density estimates with Riemannian metric}
+  \label{alg:vkderm}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
+  \Output{ outliers embedding coordinates $y_1, \dots, y_{\textit{n_outliers}}$ with their estimated densities $f_1, \dots, f_{\textit{n_outliers}}$ }
+  \Parameter{ number of outliers \textit{n_outliers}, embedding dimension $d$ }
+  \OptParameter{ }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE for all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using Algorithm&nbsp;\@ref(alg:learnmetric) with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+
+  \STATE Set the variable bandwidth for each point as $h_i$;
+
+  \STATE Compute the kernel density estimates for each point as
+  $$
+  \hat{f}(\mathbf{y})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
+  K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right];
+  $$
+
+  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top $\textit{n_outliers}}$ as the outliers.
+
+  \end{algorithmic}
+\end{algorithm}
+
+Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots [HDR plots; @Hyndman1996-lk] which are also computed using kernel density estimate of the embedding.
+Similar to Figure \@ref(fig:hdr) in Chapter \ref{mlann}, we can use the HDR plots to do anomaly detection and apply it to very high dimensional data sets.
 
@@
 \label{eq:gmm}
-P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
+P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
 \end{equation}
@@
 Similarly, the Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package *tourr* [REFERENCE].
 Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size $N$ (denote the transpose of the data matrix as $X$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a $100\times N$ matrix $A$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
+The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $A$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
 For any real matrix $A$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $Q$ and $R$ so that $A = QR$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $Q^\prime Q = I$, and $R$ is an upper triangular matrix.
@@
 $X^\prime X=(QX)^\prime(QX)$, where the rotated data matrix $(QX)^\prime$ is now the input for the manifold learning algorithms.
+
+```{r fivedmeta, fig.align = 'center', out.width = "60%", fig.cap="Scatterplot display of the animation of a 5-D tour path.", echo = FALSE}
+knitr::include_graphics("figures/tourr_5d_semisphere.png")
+```
 

### 2022-02-21 paper/paper.tex
Fix latex issue with n\_outlier

@@
             pdftitle={Anomaly detection with kernel density estimation on manifolds},
+            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
             pdfborder={0 0 0},
@@
 \makeatother
-\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}
+\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}
 
@@
 
+% Adjust headwidth in case user has changed geometry in header-includes
+\renewcommand{\headwidth}{\textwidth}
+
 \begin{document}
 \maketitle
+\begin{keywords}
+manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
+\end{keywords}
 
@@
 
+In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space.
+One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
+These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}.
+To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Metric Learning algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+By applying the Metric Learning algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
+
+The Riemannian metric defined at each point of the manifold is used to compute the geometric quantities, including angle, length, and volume, of the low-dimensional manifold embedding in any coordinate system, and be further applied to correct the distortion caused by the manifold learning algorithms.
+In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
+Therefore, if we could replace the bandwidth matrix with the Riemannian metric, we could further get the kernel density estimation of the manifold \(\mathcal{M}\).
+This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for anomaly detection.
+The proposed schematic is shown in Figure \ref{fig:vkde}.
+
+Now I briefly describe different steps involved in this main chapter.
+By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_\mathcal{M}\) is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_\mathcal{M}\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
+The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(H_i\) with the Riemannian metric \(g_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\mathbb{x})\), for anomaly detection.
+
 The rest of the paper is organized as follows.
@@
 
+In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
+For a high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get a low-dimensional embedding.
+The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others.
+\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point.
+To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
+The outliers could then be defined as the points with lowest density estimates.
+
+To start with, we introduce the notations in this manuscript.
+Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metrix.
+Finally, we propose our novel method to detect anomalies for high-dimensional data set.
+
 \hypertarget{notations}{%
@@
 
-In general a multivariate kernel density estimate looks something like this
+For a bivariate random sample \(\pmb X_1, \pmb X_2, \ldots, \pmb X_n\)
+drawn from a density \(f\),
+the kernel density estimate is defined by
+\[
+\hat{f} (\pmb x; \pmb H) = n^{-1}\sum_{i=1}^n K_{\pmb H} ( \pmb x - \pmb{X}_i)
+\]
+where \(\pmb x = (x_1, x_2)^T\) and \(\pmb{X}_i = (X_{i1}, X_{i2})^T, i = 1, 2, \ldots, n\). Here
+\(K(\pmb x)\) is the kernel which is a symmetric probability density function,
+\(\pmb H\) is the bandwidth matrix which is symmetric and positive-definite,\\
+and \(K_{\pmb H}(\pmb x) = |\pmb H|^{-1/2} K( \pmb H^{-1/2} \pmb x)\).
+The choice of \(K\) is not crucial: we take
+\(K(\pmb x) = (2\pi)^{-1} \exp(-\tfrac{1}{2} \pmb x^T \pmb x)\) the standard normal
+throughout.\\
+In contrast, the choice of \(\pmb H\) is crucial in determining the performance of \(\hat f\).
+The most common parameterizations of the bandwidth matrix
+are the diagonal and the
+general or unconstrained which has no restrictions on \(\pmb H\)
+provided that \(\pmb H\) remains positive definite and symmetric, that is
+\[
+\pmb H = \begin{bmatrix}h_1^2 & 0 \\0 & h_2^2 \end{bmatrix}
+\ \mathrm{or} \ 
+\pmb H = \begin{bmatrix}h_1^2 & h_{12} \\ h_{12}  & h_2^2 \end{bmatrix}.
+\]
+This latter parameterization allows kernels to have an arbitrary orientation
+whereas the former only allows kernels which are oriented to the
+co-ordinate axes.
 
-\[
-\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)
-\]
-
-where if a Gaussian kernel is used
-
-\[
-K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
-\]
-
-The matrix \(\mathbf{H}\) is called the bandwidth matrix and is very important.
-You will often read that the bandwidth matrix is about smoothing and it is.
-But an alternative way to think of kernel estimation is that kernel densities `borrow strength' from nearby points and the bandwidth determines what is ``nearby''.
+An alternative way to think of kernel estimation is that kernel densities `borrow strength' from nearby points and the bandwidth determines what is ``nearby''.
 If the bandwidth is large then all points are ``nearby'' and we get an overly smooth kernel density estimate.
@@
 
-The Riemannian estimated using the method of Perrault Joncas and Meila gives some idea of the distortion of an embedding (or so they claim).
-Mapping the points through a non-linear function ``stretchs'' some regions of space and ``shrinks'' others.
+The Riemannian estimated using the method of Perrault Joncas and Meila (2009) gives some idea of the distortion of an embedding (or so they claim).
+Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
 The Riemannian gives us an idea of the direction and angle of this stretching.
@@
 \[
-\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
-K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
+\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)\\
+K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{x}_i)\right]
 \]
@@
 This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.\\
+
+The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_p\mathcal{M}\) for every point \(p \in \mathcal{M}\).
+If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of manifold.
+The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((\mathcal{M}, g)\).
+
+To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
+The inner product between two vectors \(u,v \in T_p\mathcal{M}\), \(<u,v>_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{<u,v>_g}\) and the angle between two vectors \(\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}\) in the tangent space.
+Therefore, for each point \(p\in \mathcal{M}\) in any coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
+
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\in \mathcal{M}\) is defined as
+\[
+l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
+\]
+where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function mapping \([a,b]\) to \(\mathcal{M}\). While the volume of \(V\subset \mathcal{M}\) is computed by
+\[
+Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
+\]
+Both the concepts of distance and volume are relevant to kernel density estimation.
 
@@
 Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+
+\hypertarget{metric-learning-algorithm}{%
+\subsubsection{Metric Learning algorithm}\label{metric-learning-algorithm}}
+
+\textcite{Perrault-Joncas2013-pq} propose the Metric Learning algorithm which mainly involves four main steps.
+
+As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
+This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
+It is designed to apply the manifold learning methods to large-scale data sets, as well as computing the Riemannian metric of the manifold.
+
+Learn metric algorithm
+
+\begin{algorithm}[!htb]
+  \caption{Learn metric algorithm}
+  \label{alg:learnmetric}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \Input{ high-dimensional data $x_i \in \pmb{R}^p$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $y_i \in \pmb{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
+  \OptParameter{ manifold learning parameters }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^p$;
+
+  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
+  $$
+  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
+  $$
+  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
+
+  \STATE Embed each data point $x\in \mathbb{R}^p$ to embedding coordinates $y(x)=(y^1(x),\dots,y^d(x))$ by any existing manifold learning algorithm;
+
+  \STATE Obtain the matrix $\tilde{h}(x)$ at each point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates $y$ with each element being
+  $$
+    \tilde{h}^{i j}=\frac{1}{2}\left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y_i \cdot y_j\right)-y_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} y_j\right)-y_j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} y_i\right)\right];
+  $$
+
+  \STATE Calculate the Riemannian metric $h(x)$ as the rank $d$ pseudo inverse of $\tilde{h}(x)$ with 
+  $$
+    h(x) = U diag{1/(\Lambda[1:d])} U^\prime,
+  $$
+  where $[U, \Lambda]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+
+  \end{algorithmic}
+\end{algorithm}
+
+\hypertarget{proposed-algorithm}{%
+\subsection{Proposed algorithm}\label{proposed-algorithm}}
+
+Now we present our proposed method for anomaly detection based on variable kernel density estimates.
+
+\begin{algorithm}[!htb]
+  \caption{Variable kernel density estimates with Riemannian metric}
+  \label{alg:vkderm}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
+  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
+  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
+  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+
+  \STATE Set the variable bandwidth for each point as $\pmb{H}_i = h_i$;
+
+  \STATE Compute the kernel density estimates for each point as
+  $$
+  \hat{f}(\pmb{y})=\sum\limits_{i=1}^N (2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{y}_i)\right];
+  $$
+
+  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top \textit{n\_outliers} as the outliers.
+
+  \end{algorithmic}
+\end{algorithm}
+
+Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots {[}HDR plots; \textcite{Hyndman1996-lk}{]} which are also computed using kernel density estimate of the embedding.
 
@@
 Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
 Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size \(N\) (denote the transpose of the data matrix as \(X\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a \(100\times N\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
+The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
 For any real matrix \(A\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(Q\) and \(R\) so that \(A = QR\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(Q^\prime Q = I\), and \(R\) is an upper triangular matrix.
@@
 \(X^\prime X=(QX)^\prime(QX)\), where the rotated data matrix \((QX)^\prime\) is now the input for the manifold learning algorithms.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.6\linewidth]{figures/tourr_5d_semisphere}
+
+}
+
+\caption{Scatterplot display of the animation of a 5-D tour path.}\label{fig:fivedmeta}
+\end{figure}
 

### 2022-02-21 paper/paper.Rmd
Fix latex issue with n\_outlier

@@
 abstract: ""
-# keywords: "manifold learning"
+keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
 # wpnumber: 03/2021
@@
 output:
-  MonashEBSTemplates::workingpaper:
+  monash::workingpaper:
     fig_caption: yes
@@
 The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others.
-@Perrault2009 gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point.
+@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point.
 To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
@@
 <!-- $$ -->
-<!-- \hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i) -->
+<!-- \hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{x}_i) -->
 <!-- $$ -->
@@
 <!-- $$ -->
-<!-- K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right] -->
+<!-- K_{\pmb{H}}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{x}_i)\right] -->
 <!-- $$ -->
 
-<!-- The matrix $\mathbf{H}$ is called the bandwidth matrix and is very important.  -->
+<!-- The matrix $\pmb{H}$ is called the bandwidth matrix and is very important.  -->
 
@@
 $$
-\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
-K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
+\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)\\
+K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{x}_i)\right]
 $$
@@
   $$
-    h(x) = U diag{1/(\Gamma[1:d])} U^\prime,
+    h(x) = U diag{1/(\Lambda[1:d])} U^\prime,
   $$
-  where $[U, \Gamma]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Gamma$ in descending order.
+  where $[U, \Lambda]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
 
@@
   \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
   \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
-  \Output{ outliers embedding coordinates $y_1, \dots, y_{\textit{n_outliers}}$ with their estimated densities $f_1, \dots, f_{\textit{n_outliers}}$ }
-  \Parameter{ number of outliers \textit{n_outliers}, embedding dimension $d$ }
-  \OptParameter{ }
+  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
+  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
   \BlankLine
@@
 
-  \STATE for all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using Algorithm&nbsp;\@ref(alg:learnmetric) with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
 
-  \STATE Set the variable bandwidth for each point as $h_i$;
+  \STATE Set the variable bandwidth for each point as $\pmb{H}_i = h_i$;
 
@@
   $$
-  \hat{f}(\mathbf{y})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
-  K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right];
+  \hat{f}(\pmb{y})=\sum\limits_{i=1}^N (2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{y}_i)\right];
   $$
 
-  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top $\textit{n_outliers}}$ as the outliers.
+  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top \textit{n\_outliers} as the outliers.
 
@@
 Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots [HDR plots; @Hyndman1996-lk] which are also computed using kernel density estimate of the embedding.
-Similar to Figure \@ref(fig:hdr) in Chapter \ref{mlann}, we can use the HDR plots to do anomaly detection and apply it to very high dimensional data sets.
+<!-- Similar to Figure \@ref(fig:hdr) in Chapter \@ref{mlann}, we can use the HDR plots to do anomaly detection and apply it to very high dimensional data sets. -->
 

### 2022-02-21 R/sources/hdrplotting.R
Fix shiny app density scale by adding vkde2d(h=Rn*riem*scale)

@@ -152,7 +152,6 @@ den.estimate.5d <- function(x, kde.package=c("ks", "ash"), h=NULL, xextend=0.15)
   # xr <- diff(range(x,na.rm=TRUE))
   # yr <- diff(range(y,na.rm=TRUE))
   xr <- c(min(x)-xr*xextend,max(x)+xr*xextend)
-  # yr <- c(min(y)-yr*yextend,max(y)+yr*yextend)
   if(kde.package=="ash")
   {
     if(is.null(h))

### 2022-02-21 R/sources/plot_outlier.R

@@ -1,0 +1,74 @@
+
+plot_outlier <- function(x, gridsize = 20, f = NULL, prob = c(1, 50, 99), noutliers = 20, label = NULL, riem.scale = 1, ell.size = 1, ...){
+  
+  fn <- x$embedding
+  Rn <- x$rmetric
+  if(is.null(f)) f <- vkde2d(x = fn[,1], y = fn[,2], h = Rn * riem.scale, gridsize = gridsize)
+  
+  den <- f
+  # x <- E1; y <- E2
+  E1 <- fn[,1]
+  E2 <- fn[,2]
+  # Convert prob to coverage percentage if necessary
+  if(max(prob) > 50) {# Assume prob is coverage percentage
+    alpha <- (100-prob)/100
+  } else {# prob is tail probability (for backward compatibility)
+    alpha <- prob}
+  alpha <- sort(alpha)
+  # Calculates falpha needed to compute HDR of bivariate density den.
+  
+  # Also finds approximate mode.
+  fxy <- hdrcde:::interp.2d(den$x,den$y,den$z,E1,E2) # where to estimate density, stored in den
+  # fxy <- f$z, change eval.points = data in vkde2d
+  falpha <- quantile(fxy, alpha)
+  index <- which.max(fxy) 
+  
+  mode <- c(E1[index],E2[index]) # maximam density point
+  hdr2d_info <- structure(list(mode=mode, falpha=falpha, fxy=fxy, den=den, alpha=alpha, x=E1, y=E2), class="hdr2d") # list for hdr.2d() output
+  # plot.hdr2d(hdr2d_info, show.points = T, outside.points = T, pointcol = grey(0.5), xlim = round(range(E1)), ylim = round(range(E2)))
+  
+  p_hdr <- hdrscatterplot_new(E1, E2, levels = prob, noutliers = noutliers, label = label, den = hdr2d_info)
+  p_outlier_vkde <- p_hdr$p + 
+    plot_ellipse(x, add = T, ell.size = ell.size, ...) 
+                 # color = blues9[5], fill = blues9[5], alpha = 0.2, ...)
+  
+  return(list(p = p_outlier_vkde, outlier = p_hdr$outlier, densities = p_hdr$densities, hdr2d_info = hdr2d_info))
+}
+
+# Example
+# x <- dimRed::loadDataSet("Swiss Roll")
+# N <- nrow(x)
+# s <- 2
+# k <- 20
[…]

### 2022-02-21 R/sources/vkde.R

@@ -1,165 +1,100 @@
-# Variable kernel density estimate
-# \hat{f}(x) = n^{-1} \sum_i K_{H_i} (x - X_i)
+# input x is a multivariate data/matrix
+# takes care of the computation when bandwidth is full matrix instead of fixed vector or diagonal matrix
+# lims is a 2-column matrix of ranges(xmin, xmax) for each column in x
+vkde <- function(x, h, gridsize = 25, xmin = apply(x, 2, min), xmax = apply(x, 2, max), eval.points = NULL){
 
-# calculate the density estimate for each point in emb_isomap with each H matrix, H_isomap[i,,]
-
-ker.density <- function(x, h) {
-  x <- sort(x)
-  n <- length(x)
-  s <- 0
-  t <- 0
-  y <- 0
-  for (i in 2:n)
-    s[i] <- 0
-  for (i in 1:n) {
-    for (j in 1:n)
-      s[i] <- s[i] + exp(-((x[i] - x[j]) ^ 2) / (2 * h * h)) # h^2?    
-      t[i] <- s[i]
-  }
-  for (i in 1:n)
-    y[i] <- t[i] / (n * h * sqrt(2 * pi))
-  z <- complex(re = x, im = y)
-  hist(x, freq = FALSE)
-  lines(z)
-}
-
-# # univariate
-# dfn1 <- function(x) {
-#   0.5 * dnorm(x, 3, 1) + 0.5 * dnorm(x, -3, 1)
-# }
-# par(mfrow = c(2, 2))
-# curve(dfn1(x), from = -6, to = 6)
-# data <- c(rnorm(200, 3, 1), rnorm(200, -3, 1))
-# plot(density(data,bw=8))
-# plot(density(data,bw=0.8))
-# plot(density(data,bw=0.08))
-# 
-# par(mfrow = c(1,1))
-# ker.density(data, 0.8)
-
-
-# # 2-d kde
-# mkde <- function (x, h) {
-#   # Data is a matrix of n*d
-#   # H is an array of dimension (d,d,n)
-#   # start <- Sys.time()
-# 
-#   n <- nrow(x)
-#   if (dim(x)[2] < 2)
-#     stop("Data matrix has only one variable.")
-#   if (any(!is.finite(x)))
-#     stop("Missing or infinite values in the data are not allowed! ")
[…]

### 2022-02-21 R/app.R
Fix shiny app density scale by adding vkde2d(h=Rn*riem*scale)

@@ -53,7 +53,7 @@ ml_outlier <- function(x, s = 2, k = min(10, nrow(x)), radius = 0,
   Rn <- metriclearn$rmetric
   E1 <- fn[,1]; E2 <- fn[,2]
   # prob <- c(1, 50, 99)
-  f_vkde <- vkde2d(x = E1, y = E2, h = Rn, gridsize = gridsize) # estimated densities with variable bandwidth
+  f_vkde <- vkde2d(x = E1, y = E2, h = Rn*riem.scale, gridsize = gridsize) # estimated densities with variable bandwidth
   # fxy <- hdrcde:::interp.2d(f_vkde$x, f_vkde$y, f_vkde$z, x0 = E1, y0 = E2) # linear interpolation
   den <- hdrcde:::den.estimate.2d(x = E1, y = E2, kde.package = "ks", xextend=0.15, yextend = 0.15)
   
@@ -169,7 +169,7 @@ ui <- dashboardPage(
                                         # c("Swiss Roll", "Swiss Hole", "Corner Planes",   
                                         #           "Punctured Sphere", "Twin Peaks", "Clusters",
                                         #           "Toroidal Helix", "Gaussian"),
-                                        selected = "Swiss Roll")
+                                        selected = "Twin Peak")
                      ),
                      column(3, numericInput("num_pts", "No. of Points", 200)),
                      # column(3, uiOutput("data_par")),
@@ -218,7 +218,7 @@ ui <- dashboardPage(
                    style = "background-color: #ff6666;",
                    h4("Embedding and outlier plot parameters"),
                    fluidRow(
-                     column(4, numericInput("ell.no", "Number of ellipses", 10, min = 0, step = 1)),
+                     column(4, numericInput("ell.no", "Number of ellipses", 10, min = 1, step = 1)),
                      column(4, numericInput("ell.size", "Ellipse size", 0, min = 0, max = 100)),
                      column(4, numericInput("noutliers", "Number of outliers", 10, min = 1, max = 1e8, step = 1)),
                    ),
@@ -452,7 +452,7 @@ server <- shinyServer(function(input, output, session) {
         output[[paste0("c_plot_", local_i)]] <-
           renderPlot({
             if ((local_i) %in% seq_along(algor_list)) {
-              p3 <- res$p_emb # + coord_fixed()  ## TOOD: not showing with p3 + p6
[…]

### 2022-02-21 R/sources/vkde2d.R

@@ -1,0 +1,202 @@
+# Variable kernel density estimate
+# \hat{f}(x) = n^{-1} \sum_i K_{H_i} (x - X_i)
+
+# calculate the density estimate for each point in emb_isomap with each H matrix, H_isomap[i,,]
+
+ker.density <- function(x, h) {
+  x <- sort(x)
+  n <- length(x)
+  s <- 0
+  t <- 0
+  y <- 0
+  for (i in 2:n)
+    s[i] <- 0
+  for (i in 1:n) {
+    for (j in 1:n)
+      s[i] <- s[i] + exp(-((x[i] - x[j]) ^ 2) / (2 * h * h)) # h^2?    
+      t[i] <- s[i]
+  }
+  for (i in 1:n)
+    y[i] <- t[i] / (n * h * sqrt(2 * pi))
+  z <- complex(re = x, im = y)
+  hist(x, freq = FALSE)
+  lines(z)
+}
+
+# # univariate
+# dfn1 <- function(x) {
+#   0.5 * dnorm(x, 3, 1) + 0.5 * dnorm(x, -3, 1)
+# }
+# par(mfrow = c(2, 2))
+# curve(dfn1(x), from = -6, to = 6)
+# data <- c(rnorm(200, 3, 1), rnorm(200, -3, 1))
+# plot(density(data,bw=8))
+# plot(density(data,bw=0.8))
+# plot(density(data,bw=0.08))
+# 
+# par(mfrow = c(1,1))
+# ker.density(data, 0.8)
+
+
+# # 2-d kde
+# mkde <- function (x, h) {
+#   # Data is a matrix of n*d
+#   # H is an array of dimension (d,d,n)
+#   # start <- Sys.time()
+# 
+#   n <- nrow(x)
+#   if (dim(x)[2] < 2)
+#     stop("Data matrix has only one variable.")
+#   if (any(!is.finite(x)))
+#     stop("Missing or infinite values in the data are not allowed! ")
+#   if (!all.equal(nrow(h), ncol(h), dim(x)[2]))
+#     stop("The bandwidth matrix is of wrong dimension. ")
+# 
+#   s <- rep(0, n)
+#   y <- rep(0, n)
+#   for (i in 1:n) {
+#     for (j in 1:n){
+#       s[i] <- s[i] + det(h[,,i])^(-1/2) * exp(- 1/2 * t(x[i,] - x[j,]) %*% solve(h[,,i]) %*% (x[i,] - x[j,]))
[…]

### 2022-02-24 R/multidimension.R
5d plots and table

@@ -70,18 +70,18 @@ X_new %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
 # summary(den1)
 # all.equal(den, den1) # TRUE
 
-## Plot 5-d semi-sphere
-library(tourr)
-# col <- RColorBrewer::brewer.pal(3, "Dark2") # unique(flea$species)
-# animate_xy(flea[, 1:6], col = flea$species)
-
-animate_xy(X_new[,1:5], col = viridis(length(den), option = "magma"))
-# catogorize density levels and color them accordingly
-
-library(geozoo)
-sphere <- sphere.hollow(p = 5)
-sphere$points <- X_new
-sphere
+# ## Plot 5-d semi-sphere
+# library(tourr)
+# # col <- RColorBrewer::brewer.pal(3, "Dark2") # unique(flea$species)
+# # animate_xy(flea[, 1:6], col = flea$species)
+# 
+# animate_xy(X_new[,1:5], col = viridis(length(den), option = "magma"))
+# # catogorize density levels and color them accordingly
+# 
+# library(geozoo)
+# sphere <- sphere.hollow(p = 5)
+# sphere$points <- X_new
+# sphere
 
 
 
@@ -140,8 +140,8 @@ treetype <- "kd"
 searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
 radius <- 10 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
 
-riem.scale <- 10 # tune parameter
-gridsize <- 10
+riem.scale <- .1 # tune parameter
+gridsize <- 20
 
 ## ----isomap-------------------------------------------------------------
 metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
@@ -166,14 +166,14 @@ dim(fn)
 
 
 ## ----Fixed bandwidth density estimates with ks::kde-------------------------
-fixden_isomap <- vkde(x = fn, h = NULL, gridsize = gridsize, eval.points = fn)
[…]

### 2022-02-24 R/sources/vkde.R

@@ -4,7 +4,7 @@
 # eval.points by default make grids and generate estimates of grid points; if provided, eg. x, can evaluate density at certain points
 # output long list of x, eval.points (grid points or data points), estimates for density, bandwidth matrix, whether estimates are for grids
 
-vkde <- function(x, h = NULL, gridsize = 20, xmin = apply(x, 2, min), xmax = apply(x, 2, max), eval.points, xextend = 0.15){
+vkde <- function(x, h = NULL, gridsize = 20, xmin = apply(x, 2, min), xmax = apply(x, 2, max), eval.points, xextend = 0.15, ...){
 
   n <- nrow(x)
   d <- ncol(x)
@@ -16,7 +16,7 @@ vkde <- function(x, h = NULL, gridsize = 20, xmin = apply(x, 2, min), xmax = app
   if(gridsize <= 0) stop("Gridsize must be positive.")
   if(is_scalar_atomic(gridsize)) gridsize <- rep(gridsize[1], d) # vector of number of grid points
 
-  if(is.null(h) | !is.array(h)) return(ks::kde(x, h = h, gridsize = gridsize, xmin = xmin, xmax = xmax, eval.points = eval.points)) # fixed diagonal bandwidth # return(MASS::kde2d(x, y, h, n, lims)) # only works for 2d
+  if(is.null(h) | !is.array(h)) return(ks::kde(x, h = h, gridsize = gridsize, xmin = xmin, xmax = xmax, eval.points = eval.points, ...)) # fixed diagonal bandwidth # return(MASS::kde2d(x, y, h, n, lims)) # only works for 2d
   
   xr <- apply(x, 2, function(x) diff(range(x, na.rm = TRUE)))
   xmin <- xmin - xr * xextend

### 2022-02-24 reports/Simulation_swissroll.R
Swiss roll and S curve output

@@ -18,6 +18,7 @@ library(ks)
 library(patchwork)
 # library(copula)
 library(plotly)
+library(kableExtra)
 Jmisc::sourceAll(here::here("R/sources"))
 # set.seed(1)
 
@@ -123,7 +124,7 @@ p <- 2
 ## Start from here! 
 ###--------------------------------------------------------
 set.seed(1234)
-mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3]
+mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[1]
 sr <- mldata(N = 2000, meta = "gaussian", mapping = mapping)
 swissroll <- sr$data %>% as.data.frame()
 preswissroll <- sr$metadata %>% as_tibble() %>% mutate(den = sr$den, label = c(rep(1:4, each = 500)))
@@ -164,9 +165,9 @@ annmethod <- "kdtree"
 distance <- "euclidean"
 treetype <- "kd"
 searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
-radius <- 5 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
+radius <- 8 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
 
-riem.scale <- 0.1
+riem.scale <- 1
 gridsize <- 20
 
 ## ---- message=FALSE-------------------------------------------------------------
@@ -194,7 +195,7 @@ p_hdr_isomap <- hdrscatterplot_new(E1, E2, levels = prob, noutliers = 20, label 
 # p_hdr_isomap_p <- p_hdr_isomap$p +
 #   plot_ellipse(metric_isomap, add = T, ell.no = 50, ell.size = .5,
 #              color = blues9[5], fill = blues9[1], alpha = 0.2)
-p_hdr_isomap
+# p_hdr_isomap
 
 
 ## ----outliers-------------------------------------------------------------------
@@ -206,6 +207,7 @@ plot_contour(metric_isomap, gridsize = gridsize, riem.scale = riem.scale) # esti
[…]

### 2022-02-24 paper/paper.Rmd
Add all plots and tables

@@
 library(plotly)
+library(kableExtra)
 Jmisc::sourceAll(here::here("R/sources"))
@@
 
-	Background (anomaly detection with density) with summarized contribution
+<!-- Background (anomaly detection with density) with summarized contribution -->
 
-		Anomaly detection on very high-dimensional data. Applications. Main contribution.
+<!-- 	Anomaly detection on very high-dimensional data. Applications. Main contribution. -->
 
-	Anomaly detection using densities, other methods for anomaly detection. Based manifold learning embeddings, density estimates
+<!-- Anomaly detection using densities, other methods for anomaly detection. Based manifold learning embeddings, density estimates -->
 
-	Density estimation methods, KDE with distortion, 2-D variable kernel denstiy estimate with Riemannian matrix
+<!-- Density estimation methods, KDE with distortion, 2-D variable kernel denstiy estimate with Riemannian matrix -->
 
-	- Main contributions: better anomaly detection
+<!-- - Main contributions: better anomaly detection -->
 
-		fix distortion with Riemmannian matrix
+<!-- 	fix distortion with Riemmannian matrix -->
 
-		anomaly detection using density estimates
+<!-- 	anomaly detection using density estimates -->
 
-		kernel density estimate is not accurate
+<!-- 	kernel density estimate is not accurate -->
 
-		improve density estimation with distortion in the embedding plot
+<!-- 	improve density estimation with distortion in the embedding plot -->
 
@@
 This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for anomaly detection.
-The proposed schematic is shown in Figure \@ref(fig:vkde).
 
-(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
-```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=FALSE}
-knitr::include_graphics(here::here("report/vkde.png"))
-```
-
-Now I briefly describe different steps involved in this main chapter.
+<!-- Now I briefly describe different steps involved in this main chapter.  -->
 By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
@@
 In this section, we provide justification for the use of Riemannien matrix as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data and the second with a 100-dimensional meta data.
-\autoref{application} contains the application to visualize and identify anomalies in the [TODO] data.
+\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data embedded in 3-D and the second with a 100-dimensional meta data.
+\autoref{application} contains the application to visualize and identify anomalies in the [TODO] dataset.
 Conclusions and discussions are presented in \autoref{conclusion}.
@@
 # Variable kernel density estimation with manifold embeddings {#vkde}
-
-	Idea: whole process for the main contribution
-
-	Kernel density estimate with variable bandwidth
 
@@
 The outliers could then be defined as the points with lowest density estimates.
+The proposed schematic is shown in Figure \@ref(fig:vkde).
+
+(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
+```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
+knitr::include_graphics("figures/vkde.png")
+```
 
@@
 ## Notations
+
+Manifold learning finds a $d$-dimensional representation of data that lie on a manifold $\mathcal{M}$ embedded in a $p$-dimensional ambient space with $d \ll p$.
+We will denote the original data (or 'input' points) as $x_i$, $i=1,\dots,N$, where $x_i\in\mathbb{R}^p$, while the low-dimensional representation (or 'output' points) will be denoted as $y_i$, $i=1,\dots,N$, where $y_i\in\mathbb{R}^d$.
+Where two subscripts are used (e.g.
+$x_{ih}$ or $y_{ih}$), the second subscript refers to the $h^{th}$ coordinate or dimension of the data.
+<!-- Pairwise distances between input points $x_i$ and $x_j$ are denoted $\delta_{ij}$ while pairwise distances between output points $y_i$ and $y_j$ are denoted $d_{ij}$, where $i,j=1,\dots,N$.
+Unless otherwise stated, these distances are assumed to be Euclidean.
+-->
 
@@
 
-    Riemannian matrix could be used to measure distortion
+  <!-- Riemannian matrix could be used to measure distortion -->
 
-    Use riemannian matrix as variable bandwidth
-        Bandwidth is a symmetric positive-definite square matrix, similar to riemannian matrix
+  <!-- Use riemannian matrix as variable bandwidth -->
+  <!--     Bandwidth is a symmetric positive-definite square matrix, similar to riemannian matrix -->
 
@@
 
-    	- Manifold learning distortion
+<!-- - Manifold learning distortion -->
 
-    		dimension reduction using manifold learning
-    		Riemannian matrix to measure topological distortion
-    		distortion is informative for true embedding
+<!-- 	dimension reduction using manifold learning -->
+<!-- 	Riemannian matrix to measure topological distortion -->
+<!-- 	distortion is informative for true embedding -->
 
@@
 We first simulate the data of size $N=2000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
-Different mapping functions are then applied to the 2-D meta data so that they now lie on a 3-D feature space, which gives the input data for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as $D=2$, same as the meta data dimension.
+Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+The embedded dimension is set as $d=2$, same as the meta data dimension.
 This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply the algorithm described in [TODO].
-Similarly, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is $D=5$.
+We could now apply the algorithm described in \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
 
@@
 
-We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with means $\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75), \pmb{\mu_3}=(0.75, 0.25), \pmb{\mu_4}(0.75, 0.75)$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$.
+We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with different means $\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$.
 The mixture proportions are equally set as $\pi_i=0.25, i=1,2,3,4$.
@@
 \end{equation}
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points, with brighter colors showing high densities and dark colors showing low densities.
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities.
 We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
 Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-This is the baseline to be compared with outliers from variable kernel density estimates.
+This is the baseline, which we call *true densities*, to be compared with outliers from the kernel density estimates.
 
@@
 
+<!-- ### Swiss roll mapping -->
+
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous example in manifold learning is the swiss roll data, with the mapping function as below.
-$$
+One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
+The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})$, shown in the left plot of \autoref{fig:mappings}.
+\begin{equation}
+\label{eq:swissroll}
 \left\{ 
 \begin{array}{lcl}
-X = X_1 \times \cos{X_1}, \\
+X = X_1 \cos{X_1}, \\
 Y = X_2, \\
-Z = X_1 \times \sin{X_1}.
+Z = X_1 \sin{X_1}.
 \end{array}
 \right.
-$$
-Now we are able to apply different manifold learning algorithms to reduce the dimension back to 2, and further estimate the density of the embedding to detect anomalies.
-According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
-1%, 50%, 99%, >99%.
-The top outliers with lowest densities are also indexed in blue, which we call *true outliers*.
-For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents in one column the 2-D embedding plot, with the colors indicating the densities, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
-Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as ISOMAP and UMAP.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+\end{equation}
+
+```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.", echo = FALSE}
+knitr::include_graphics("figures/mappings_sr_tp.png")
+```
 
@@
 Both methods are not detecting outliers accurately when using UMAP.", echo = FALSE}
-knitr::include_graphics("figures/outliers_comparison_4ml_3cases.png")
+knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem001.png")
 ```
 
-```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.", echo = FALSE}
-knitr::include_graphics("figures/mappings_sr_tp.png")
+Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
+According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
+1%, 50%, 99%, >99%.
+For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same column, with the colors indicating the densities levels, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE method, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
+The top ten outliers with lowest densities are also highlighted in black with point indexes in blue.
+Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
+As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
+Instead, we could roughly compare the location of outliers.
+For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+
+```{r srcors, echo=FALSE, message=FALSE, eval=F}
+load("figures/CorrelationTable_Swiss Roll_4ml_riem01.rda")
+col.max <- apply(cors, 2, max)
+max.match <- rep(col.max, each = 2) == cors
+data.frame(cors) %>% 
+  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
+  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kable_styling(latex_options = "scale_down") %>%
+  # kable_paper(full_width = TRUE) %>%
+  column_spec(2, bold = max.match[,1]) %>%
+  column_spec(3, bold = max.match[,2]) %>%
+  column_spec(4, bold = max.match[,3]) %>%
+  column_spec(5, bold = max.match[,4]) %>% 
+  column_spec(1, width = "6cm")
 ```
 
-For comparison, we use the same meta data with different mapping function, twin peak mapping, with the corresponding 3-D data is shown in \autoref{fig:mappings}.
-The four colors in both mappings represents the four gaussian kernels in the meta data.
-Similar to \autoref{fig:sroutliers}, four manifold learnings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+### Twin peaks mapping
+
+For comparison, we use the same meta data with different mapping function, twin peak mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
+The four colors in both mappings represents the four gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)$.
+\begin{equation}
+\label{eq:twinpeak}
+\left\{ 
+\begin{array}{lcl}
+X = X_1, \\
+Y = X_2, \\
+Z = \sin(\pi X_1) \tanh (3 X_2).
+\end{array}
+\right.
+\end{equation}
 
@@
 Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.", echo = FALSE}
-knitr::include_graphics("figures/twin_peak_outliers_comparison_4ml_3cases.png")
+knitr::include_graphics("figures/Twin Peak_outliers_comparison_4ml_3cases_riem01.png")
 ```
 
-<!-- Multiple mapping functions: swiss roll, twin peak -->
+Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+[ANALYSE TWIN PEAK PLOT]
 
-<!-- Manifold learning results for the mapped 3-D data -->
+```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+The four colors represents the four gaussian kernels in the 2-D meta data.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem01.png")
+```
 
-<!--     ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, UMAP (useful ones, and the one with too severe distortion) -->
+With severe distortion in the t-SNE and UMAP embeddings, both variable and fixed bandwidth methods can not produce accurate density estimates.
+It would then be very difficult to uncover the Gaussian mixture model structure and
 
-<!-- Outlier plot based on density estimates -->
-
-<!--     Fixed bandwidth -->
-
-<!--     Pointwise adaptive bandwidth using Riemmanien -->
-
-<!--     Comparison -->
-
-<!--         Rank of densities with the true density (f=f_1/4 + f_2/4 + f_3/4 + f_4/4) -->
-<!--         Sum of squared errors in the density estimates -->
+```{r tpcors, echo=FALSE, message=FALSE, eval=T}
+load("figures/CorrelationTable_twinpeak_4ml_riem01.rda")
+col.max <- apply(cors, 2, max)
+max.match <- rep(col.max, each = 2) == cors
+data.frame(cors) %>% 
+  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
+  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kable_styling(latex_options = "scale_down") %>%
+  # kable_paper(full_width = TRUE) %>%
+  column_spec(2, bold = max.match[,1]) %>%
+  column_spec(3, bold = max.match[,2]) %>%
+  column_spec(4, bold = max.match[,3]) %>%
+  column_spec(5, bold = max.match[,4]) %>% 
+  column_spec(1, width = "6cm")
+```
 
@@
 
-```{r fivedmeta, fig.align = 'center', out.width = "60%", fig.cap="Scatterplot display of the animation of a 5-D tour path.", echo = FALSE}
+```{r fivedmeta, fig.align = 'center', out.width = "80%", fig.cap="Scatterplot display of the animation of a 5-D tour path.", echo = FALSE}
 knitr::include_graphics("figures/tourr_5d_semisphere.png")
+```
+
+Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+[ANALYSE TWIN PEAK PLOT]
+
+```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
+Top: ISOMAP; middle: LLE; bottom: UMAP.", echo = FALSE}
+knitr::include_graphics("figures/fived_density_comparison_isomap_riem20.png")
+```
+
+```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
+load("figures/CorrelationTable_fived_4ml_riem20.rda")
+col.max <- apply(cors, 2, max)
+max.match <- rep(col.max, each = 2) == cors
+data.frame(cors) %>% 
+  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
+  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kable_styling(latex_options = "scale_down") %>%
+  # kable_paper(full_width = TRUE) %>%
+  column_spec(2, bold = max.match[,1]) %>%
+  column_spec(3, bold = max.match[,2]) %>%
+  column_spec(4, bold = max.match[,3]) %>%
+  # column_spec(5, bold = max.match[,4]) %>% 
+  column_spec(1, width = "6cm")
 ```
@@
 
-	Dataset
+<!-- Dataset -->
 
-	Goal and process
+<!-- Goal and process -->
 
-	Results (Plots and tables)
+<!-- Results (Plots and tables) -->
 
-	Analysis and summary
+<!-- Analysis and summary -->
+
+## Victoria smart meter dataset
+
+[TODO]
+
+Next, we consider smart-meter data for residential and non-profiled meter consumers, collected in the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data].
+Electricity smart meters record consumption, on a near real-time basis, at the level of individual commercial and residential properties.
+The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] does not include energy for heating systems since it is either metered separately, or households use a different source of energy, such as oil or gas.
+In this study, the installed cooling systems are also not reported.
+
+```{r smartmeter, fig.align = 'center', out.width = "80%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=FALSE}
+knitr::include_graphics("figures/smartmeter.png")
+```
+
+We use measurements of half-hourly electricity consumption gathered from 3,639 residential consumers over 535 consecutive days.
+Every meter provides electricity consumption between 14 July 2009 and 31 December 2010.
+Demand data from two smart meters (ID 1003 and 1539) are shown in \autoref{fig:smartmeter} as time series plots.
+It is obvious that these meters have relatively different patterns.
+Meter 1539 (bottom of \autoref{fig:smartmeter}) has a period of around 150 days with lower (approximately half) the electricity usage of remaining days and is otherwise relatively stable.
+In contrast, Meter 1003 (top of \autoref{fig:smartmeter}) exhibits regular spikes on weekends.
+
+For electricity demand data, one particular problem of interest is to visualize and identify households or periods of the week with anomalous usage patterns.
+For this reason, the object of interest in comparing households or periods of the week is the distribution of electricity demand rather than the raw data itself [@Hyndman2018-ia].
+An additional advantage of looking at distributions rather than raw data is that it provides a convenient mechanism for handling missing data which can be a significant problem in smart meter data applications.
+In the rest of this section, we first describe how discrete approximations to the distributions of interest are computed.
+We then approximate the distances between vectors containing probabilities with the consistent estimator in \autoref{estimator}, which allows us to apply manifold learning techniques on statistical manifolds.
+
+<!-- ## Data processing {#dataprocessing} -->
+
+Let $d_{i,t}$ denote the electricity demand for observation $i$ and for time period $t$ (subsequently we will see that $i$ can either index the household, time of the week, or both, while $t$ may index the week or half-hour period).
+The objective is to estimate the distances between the distribution of electricity demand for observation $i$, $F_i$, over time.
+Since the raw data could be thought of as samples generated from the true electricity usage distributions for different households and different periods, we could apply the consistent Hellinger distance estimator derived in \autoref{estimator} to the smart meter data.
+The first step is to group all data together regardless of $i$ and $t$, and take the data as samples from the reference probability distribution $R$.
+The sample size $r=93,616,914$.
+Next, we construct $T_r=100$ intervals with about $1\%$ samples from $R$ within each interval (except for the last interval).
+The segments $\{I_g^r\}_{g=1,2,\dots,100}$ can then be used to construct discrete distribution approximation to $F_i$ by counting the ratio of samples points from $F_i$ fallen into segment $I_g^r$.
+This is found by computing $\pi_{i,g}=(1/m)\sum_t I(d_{i,t} \in I_g^r)$ where $m$ is the sample size from $F_i$. 
+Then the Hellinger distance between $F_i$ and $F_j$ could be estimated using Equation \@ref(eq:hellingerest1).
+In this way, we make it possible to search nearest neighbors more efficiently using ANN methods and further apply manifold learning algorithms.
 
@@
 
-	Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate)
+Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
+
+Discuss simulation with non-uniform meta data.
 

### 2022-02-24 paper/paper.tex
Add all plots and tables

@@
 \usepackage{bm}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 
-\begin{verbatim}
-Background (anomaly detection with density) with summarized contribution
-
-    Anomaly detection on very high-dimensional data. Applications. Main contribution.
-
-Anomaly detection using densities, other methods for anomaly detection. Based manifold learning embeddings, density estimates
-
-Density estimation methods, KDE with distortion, 2-D variable kernel denstiy estimate with Riemannian matrix
-
-- Main contributions: better anomaly detection
-
-    fix distortion with Riemmannian matrix
-
-    anomaly detection using density estimates
-
-    kernel density estimate is not accurate
-
-    improve density estimation with distortion in the embedding plot
-\end{verbatim}
-
 In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space.
@@
 This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for anomaly detection.
-The proposed schematic is shown in Figure \ref{fig:vkde}.
 
-Now I briefly describe different steps involved in this main chapter.
 By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
@@
 In this section, we provide justification for the use of Riemannien matrix as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data and the second with a 100-dimensional meta data.
-\autoref{application} contains the application to visualize and identify anomalies in the {[}TODO{]} data.
+\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data embedded in 3-D and the second with a 100-dimensional meta data.
+\autoref{application} contains the application to visualize and identify anomalies in the {[}TODO{]} dataset.
 Conclusions and discussions are presented in \autoref{conclusion}.
@@
 \section{Variable kernel density estimation with manifold embeddings}\label{vkde}}
-
-\begin{verbatim}
-Idea: whole process for the main contribution
-
-Kernel density estimate with variable bandwidth
-\end{verbatim}
 
@@
 The outliers could then be defined as the points with lowest density estimates.
+The proposed schematic is shown in Figure \ref{fig:vkde}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/vkde}
+
+}
+
+\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
+\end{figure}
 
@@
 \subsection{Notations}\label{notations}}
+
+Manifold learning finds a \(d\)-dimensional representation of data that lie on a manifold \(\mathcal{M}\) embedded in a \(p\)-dimensional ambient space with \(d \ll p\).
+We will denote the original data (or `input' points) as \(x_i\), \(i=1,\dots,N\), where \(x_i\in\mathbb{R}^p\), while the low-dimensional representation (or `output' points) will be denoted as \(y_i\), \(i=1,\dots,N\), where \(y_i\in\mathbb{R}^d\).
+Where two subscripts are used (e.g.~\(x_{ih}\) or \(y_{ih}\)), the second subscript refers to the \(h^{th}\) coordinate or dimension of the data.
 
@@
 The interesting thing about a bandwidth matrix is that it allows for different notions of what is ``nearby'' along different coordinates and even along diagonal directions.
-
-\begin{verbatim}
-Riemannian matrix could be used to measure distortion
-
-Use riemannian matrix as variable bandwidth
-    Bandwidth is a symmetric positive-definite square matrix, similar to riemannian matrix
-\end{verbatim}
 
@@
 Both the concepts of distance and volume are relevant to kernel density estimation.
-
-\begin{verbatim}
-    - Manifold learning distortion
-
-        dimension reduction using manifold learning
-        Riemannian matrix to measure topological distortion
-        distortion is informative for true embedding
-\end{verbatim}
 
@@
 We first simulate the data of size \(N=2000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
-Different mapping functions are then applied to the 2-D meta data so that they now lie on a 3-D feature space, which gives the input data for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as \(D=2\), same as the meta data dimension.
+Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+The embedded dimension is set as \(d=2\), same as the meta data dimension.
 This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply the algorithm described in {[}TODO{]}.
-Similarly, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is \(D=5\).
+We could now apply the algorithm described in \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
 
@@
 
-We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75), \pmb{\mu_3}=(0.75, 0.25), \pmb{\mu_4}(0.75, 0.75)\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
+We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
 The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
@@
 \end{equation}
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points, with brighter colors showing high densities and dark colors showing low densities.
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
 We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
 Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-This is the baseline to be compared with outliers from variable kernel density estimates.
+This is the baseline, which we call \emph{true densities}, to be compared with outliers from the kernel density estimates.
 
@@
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous example in manifold learning is the swiss roll data, with the mapping function as below.
-\[
+One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
+The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})\), shown in the left plot of \autoref{fig:mappings}.
+\begin{equation}
+\label{eq:swissroll}
 \left\{ 
 \begin{array}{lcl}
-X = X_1 \times \cos{X_1}, \\
+X = X_1 \cos{X_1}, \\
 Y = X_2, \\
-Z = X_1 \times \sin{X_1}.
+Z = X_1 \sin{X_1}.
 \end{array}
 \right.
-\]
-Now we are able to apply different manifold learning algorithms to reduce the dimension back to 2, and further estimate the density of the embedding to detect anomalies.
-According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
-1\%, 50\%, 99\%, \textgreater99\%.
-The top outliers with lowest densities are also indexed in blue, which we call \emph{true outliers}.
-For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents in one column the 2-D embedding plot, with the colors indicating the densities, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
-Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as ISOMAP and UMAP.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/outliers_comparison_4ml_3cases}
-
-}
-
-\caption{Highest density region plots of four manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from top: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and bottom: KDE with fixed bandwidth.
-Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE.
-Both methods are not detecting outliers accurately when using UMAP.}\label{fig:sroutliers}
-\end{figure}
+\end{equation}
 
@@
 
-For comparison, we use the same meta data with different mapping function, twin peak mapping, with the corresponding 3-D data is shown in \autoref{fig:mappings}.
-The four colors in both mappings represents the four gaussian kernels in the meta data.
-Similar to \autoref{fig:sroutliers}, four manifold learnings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem001}
+
+}
+
+\caption{Highest density region plots of four manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from top: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and bottom: KDE with fixed bandwidth.
+Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE.
+Both methods are not detecting outliers accurately when using UMAP.}\label{fig:sroutliers}
+\end{figure}
+
+Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
+According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
+1\%, 50\%, 99\%, \textgreater99\%.
+For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same column, with the colors indicating the densities levels, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE method, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
+The top ten outliers with lowest densities are also highlighted in black with point indexes in blue.
+Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
+As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
+Instead, we could roughly compare the location of outliers.
+For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+
+\hypertarget{twin-peaks-mapping}{%
+\subsubsection{Twin peaks mapping}\label{twin-peaks-mapping}}
+
+For comparison, we use the same meta data with different mapping function, twin peak mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
+The four colors in both mappings represents the four gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)\).
+\begin{equation}
+\label{eq:twinpeak}
+\left\{ 
+\begin{array}{lcl}
+X = X_1, \\
+Y = X_2, \\
+Z = \sin(\pi X_1) \tanh (3 X_2).
+\end{array}
+\right.
+\end{equation}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/twin_peak_outliers_comparison_4ml_3cases}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_outliers_comparison_4ml_3cases_riem01}
 
@@
 \end{figure}
+
+Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+{[}ANALYSE TWIN PEAK PLOT{]}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem01}
+
+}
+
+\caption{Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+The four colors represents the four gaussian kernels in the 2-D meta data.}\label{fig:tpisomapden}
+\end{figure}
+
+With severe distortion in the t-SNE and UMAP embeddings, both variable and fixed bandwidth methods can not produce accurate density estimates.
+It would then be very difficult to uncover the Gaussian mixture model structure and
+
+\begin{table}
+
+\caption{\label{tab:tpcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{6cm}>{}l>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & t.SNE & UMAP\\
+\midrule
+Variable bandwidth & \textbf{0.940} & \textbf{0.407} & 0.520 & 0.259\\
+Fixed bandwidth & 0.626 & 0.399 & \textbf{0.660} & \textbf{0.655}\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{figures/tourr_5d_semisphere}
+{\centering \includegraphics[width=0.8\linewidth]{figures/tourr_5d_semisphere}
 
@@
 
+Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
+{[}ANALYSE TWIN PEAK PLOT{]}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_isomap_riem20}
+
+}
+
+\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth. Top: ISOMAP; middle: LLE; bottom: UMAP.}\label{fig:fivedisomapden}
+\end{figure}
+
+\begin{table}
+
+\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{6cm}>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & UMAP\\
+\midrule
+Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{-0.130}\\
+Fixed bandwidth & 0.806 & 0.940 & -0.341\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
 \hypertarget{application}{%
@@
 
-\begin{verbatim}
-Dataset
+\hypertarget{victoria-smart-meter-dataset}{%
+\subsection{Victoria smart meter dataset}\label{victoria-smart-meter-dataset}}
 
-Goal and process
+{[}TODO{]}
 
-Results (Plots and tables)
+Next, we consider smart-meter data for residential and non-profiled meter consumers, collected in the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data}.
+Electricity smart meters record consumption, on a near real-time basis, at the level of individual commercial and residential properties.
+The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} does not include energy for heating systems since it is either metered separately, or households use a different source of energy, such as oil or gas.
+In this study, the installed cooling systems are also not reported.
 
-Analysis and summary
-\end{verbatim}
+We use measurements of half-hourly electricity consumption gathered from 3,639 residential consumers over 535 consecutive days.
+Every meter provides electricity consumption between 14 July 2009 and 31 December 2010.
+Demand data from two smart meters (ID 1003 and 1539) are shown in \autoref{fig:smartmeter} as time series plots.
+It is obvious that these meters have relatively different patterns.
+Meter 1539 (bottom of \autoref{fig:smartmeter}) has a period of around 150 days with lower (approximately half) the electricity usage of remaining days and is otherwise relatively stable.
+In contrast, Meter 1003 (top of \autoref{fig:smartmeter}) exhibits regular spikes on weekends.
+
+For electricity demand data, one particular problem of interest is to visualize and identify households or periods of the week with anomalous usage patterns.
+For this reason, the object of interest in comparing households or periods of the week is the distribution of electricity demand rather than the raw data itself \autocite{Hyndman2018-ia}.
+An additional advantage of looking at distributions rather than raw data is that it provides a convenient mechanism for handling missing data which can be a significant problem in smart meter data applications.
+In the rest of this section, we first describe how discrete approximations to the distributions of interest are computed.
+We then approximate the distances between vectors containing probabilities with the consistent estimator in \autoref{estimator}, which allows us to apply manifold learning techniques on statistical manifolds.
+
+Let \(d_{i,t}\) denote the electricity demand for observation \(i\) and for time period \(t\) (subsequently we will see that \(i\) can either index the household, time of the week, or both, while \(t\) may index the week or half-hour period).
+The objective is to estimate the distances between the distribution of electricity demand for observation \(i\), \(F_i\), over time.
+Since the raw data could be thought of as samples generated from the true electricity usage distributions for different households and different periods, we could apply the consistent Hellinger distance estimator derived in \autoref{estimator} to the smart meter data.
+The first step is to group all data together regardless of \(i\) and \(t\), and take the data as samples from the reference probability distribution \(R\).
+The sample size \(r=93,616,914\).
+Next, we construct \(T_r=100\) intervals with about \(1\%\) samples from \(R\) within each interval (except for the last interval).
+The segments \(\{I_g^r\}_{g=1,2,\dots,100}\) can then be used to construct discrete distribution approximation to \(F_i\) by counting the ratio of samples points from \(F_i\) fallen into segment \(I_g^r\).
+This is found by computing \(\pi_{i,g}=(1/m)\sum_t I(d_{i,t} \in I_g^r)\) where \(m\) is the sample size from \(F_i\).
+Then the Hellinger distance between \(F_i\) and \(F_j\) could be estimated using Equation \eqref{eq:hellingerest1}.
+In this way, we make it possible to search nearest neighbors more efficiently using ANN methods and further apply manifold learning algorithms.
 
@@
 
-\begin{verbatim}
 Summary of the contribution and conclusion
@@
 
-Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate)
-\end{verbatim}
+Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
+
+Discuss simulation with non-uniform meta data.
 

### 2022-02-25 R/sources/plot_contour.R
Remove fixed coordinates

@@ -4,7 +4,7 @@
 #'
 #' @param x List object from `metricML()` function containing the 2-d embeddings and the Riemannian metric
 #' @param gridsize Grid size for estimating densities
-#' @param f Pre-computed densities
+#' @param f Pre-computed densities over grid points
 #' @param riem.scale Scaling parameter for all Riemannien metric in variable KDE
 #'
 #' @return A full page display of the filled contour with scatterplot of embeddings

### 2022-02-25 R/sources/plot_embedding.R
Remove fixed coordinates

@@ -42,7 +42,7 @@ plot_embedding <- function(x, embedding = FALSE, color = NULL, alpha = NULL) {
     as_tibble() %>% 
     ggplot(aes(x = E1, y = E2, color = color, alpha = alpha)) + 
     geom_point() + 
-    coord_fixed(ratio = 1) + 
+    # coord_fixed(ratio = 1) + 
     # color + 
     theme(legend.position = 'bottom')
   

### 2022-02-25 reports/Simulation_swissroll.R
Swiss roll riem.scale=0.08

@@ -167,7 +167,7 @@ treetype <- "kd"
 searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
 radius <- 8 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
 
-riem.scale <- 1
+riem.scale <- .08 # scale Riemmanian
 gridsize <- 20
 
 ## ---- message=FALSE-------------------------------------------------------------
@@ -255,7 +255,39 @@ p_hdr_lle <- hdrscatterplot_new(E1, E2, kde.package = "ks", noutliers = 20, leve
 #   plot_ellipse(metric_lle, ell.no = 50, add = T)
 (p_lle$p + p_hdr_lle$p) + coord_fixed() + 
   plot_annotation(title = "Left: variable bandwidth; Right: fixed bandwidth", theme = theme(plot.title = element_text(hjust = 0.5)))
+cor(preswissroll$den, p_lle$densities)
+cor(preswissroll$den, p_hdr_lle$densities)
 
+
+
+# Laplacian Eigemaps
+
+## ----le, message=FALSE, warning=FALSE, eval=TRUE------------------------------
+x <- train
+method <- "LaplacianEigenmaps"
+metric_le <- metricML(x, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+                       # annmethod = annmethod, distance = distance, treetype = treetype, 
+                       searchtype = searchtype
+)
+fn <- metric_le$embedding
+Rn <- metric_le$rmetric
+E1 <- fn[,1]; E2 <- fn[,2]
+# fle <- vkde2d(x = E1, y = E2, h = Rn*riem.scale, gridsize = gridsize)
+# fxy_le <- hdrcde:::interp.2d(fle$x, fle$y, fle$z, x0 = E1, y0 = E2)
+# plot_embedding(metric_le)
+# plot_ellipse(metric_le, ell.no = 50)
+# plot_contour(metric_le, gridsize = gridsize, riem.scale = 1/20)
+
+fle <- vkde(x = fn, h = Rn*riem.scale, gridsize = gridsize, eval.points = fn)
+## ---- echo = F------------------------------------------------------------------
[…]

### 2022-02-25 paper/paper.Rmd
Update Swissroll plot, table full.width=F

@@
 Jmisc::sourceAll(here::here("R/sources"))
-# set.seed(123)
+# set.seed(1234)
 ```
@@
 
-<!-- ### Swiss roll mapping -->
+### Swiss roll mapping
 
@@
 Both methods are not detecting outliers accurately when using UMAP.", echo = FALSE}
-knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem001.png")
+knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
 ```
@@
 
-```{r srcors, echo=FALSE, message=FALSE, eval=F}
-load("figures/CorrelationTable_Swiss Roll_4ml_riem01.rda")
+```{r srcors, echo=FALSE, message=FALSE, eval=T}
+load("figures/CorrelationTable_Swiss Roll_4ml_riem0_08.rda")
 col.max <- apply(cors, 2, max)
@@
   kable_styling(latex_options = "scale_down") %>%
-  # kable_paper(full_width = TRUE) %>%
+  # kable_paper(full_width = FALSE) %>%
+  # column_spec(1, width = "6cm") %>% 
   column_spec(2, bold = max.match[,1]) %>%
@@
   column_spec(5, bold = max.match[,4]) %>%
-  column_spec(1, width = "6cm")
+  column_spec(6, bold = max.match[,5])
 ```
@@
   kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  kable_styling(latex_options = "scale_down") %>%
-  # kable_paper(full_width = TRUE) %>%
+  # kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = FALSE) %>%
+  # column_spec(1, width = "6cm") %>% 
   column_spec(2, bold = max.match[,1]) %>%
@@
   column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4]) %>% 
-  column_spec(1, width = "6cm")
+  column_spec(5, bold = max.match[,4]) #%>%
+  # column_spec(6, bold = max.match[,5])
 ```
@@
   kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  kable_styling(latex_options = "scale_down") %>%
-  # kable_paper(full_width = TRUE) %>%
+  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
+  kable_paper(full_width = FALSE) %>%
+  # column_spec(1, width = "6cm") %>% 
   column_spec(2, bold = max.match[,1]) %>%
   column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) %>%
-  # column_spec(5, bold = max.match[,4]) %>% 
-  column_spec(1, width = "6cm")
+  column_spec(4, bold = max.match[,3]) #%>%
+  # column_spec(5, bold = max.match[,4]) 
 ```

### 2022-02-25 paper/paper.tex
Update Swissroll plot, table full.width=F

@@
 
+\hypertarget{swiss-roll-mapping}{%
+\subsubsection{Swiss roll mapping}\label{swiss-roll-mapping}}
+
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem001}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
 
@@
 For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+
+\begin{table}
+
+\caption{\label{tab:srcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
+\midrule
+Variable bandwidth & 0.0696 & \textbf{0.400} & -0.2357 & 0.023 & \textbf{0.0138}\\
+Fixed bandwidth & \textbf{0.2798} & 0.351 & \textbf{0.0141} & \textbf{0.367} & -0.0110\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
@@
 \centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{6cm}>{}l>{}l>{}l>{}l}
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
 \toprule
@@
 \bottomrule
-\end{tabular}}
+\end{tabular}
 \end{table}
@@
 \centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{6cm}>{}l>{}l>{}l}
+\begin{tabular}[t]{l>{}l>{}l>{}l}
 \toprule

### 2022-02-25 paper/paper.tex
Swiss roll and tp plots update

@@
 For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulate the data of size \(N=2000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
+We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
 Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
@@
 This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply the algorithm described in \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
+We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=10,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
 
@@
 Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-This is the baseline, which we call \emph{true densities}, to be compared with outliers from the kernel density estimates.
+These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
 
@@
 
-\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$}\label{fig:metadensity}
+\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.}\label{fig:metadensity}
 \end{figure}
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/mappings_sr_tp}
+
+}
+
+\caption{3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
+\end{figure}
+
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
 One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
-The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})\), shown in the left plot of \autoref{fig:mappings}.
+The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
+The four colors in the mappings represents the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
 \begin{equation}
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{figures/mappings_sr_tp}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll5levels_outliers_comparison_4ml_3cases_riem0_08}
 
@@
 
-\caption{3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
+\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
+Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+Both methods are not detecting outliers accurately when using t-SNE and UMAP.}\label{fig:sroutliers}
 \end{figure}
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
-
-}
-
-\caption{Highest density region plots of four manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from top: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and bottom: KDE with fixed bandwidth.
-Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE.
-Both methods are not detecting outliers accurately when using UMAP.}\label{fig:sroutliers}
-\end{figure}
-
-Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
+Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
 According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
-1\%, 50\%, 99\%, \textgreater99\%.
-For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same column, with the colors indicating the densities levels, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE method, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
-The top ten outliers with lowest densities are also highlighted in black with point indexes in blue.
-Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+1\%, 5\%, 50\%, 99\%, \textgreater99\%.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+The top twenty outliers with lowest densities are highlighted in black with point indexes in blue.
+From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
+However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwith for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detect true outliers accurately.
+For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
 
@@
 
-\caption{\label{tab:srcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings.}
 \centering
-\resizebox{\linewidth}{!}{
 \begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
@@
 \bottomrule
-\end{tabular}}
+\end{tabular}
 \end{table}
+
+To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
+It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it is closest to rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are the very high in yellow, leading to a negative correlation.
+The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
+As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
+In conclusion, our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg.
+ISOMAP, or when the embedding are discontinuous, eg.
+t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
 
@@
 
-For comparison, we use the same meta data with different mapping function, twin peak mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
-The four colors in both mappings represents the four gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)\).
+For comparison, we use the same meta data in \autoref{fig:metadensity} with different mapping function, twin peak mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_outliers_comparison_4ml_3cases_riem01}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak5levels_outliers_comparison_4ml_3cases_riem0_1}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem01}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1}
 
@@
 \centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
 \toprule
-  & ISOMAP & LLE & t.SNE & UMAP\\
+  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
 \midrule
-Variable bandwidth & \textbf{0.940} & \textbf{0.407} & 0.520 & 0.259\\
-Fixed bandwidth & 0.626 & 0.399 & \textbf{0.660} & \textbf{0.655}\\
+Variable bandwidth & \textbf{0.899} & 0.385 & 0.620 & 0.259 & \textbf{0.659}\\
+Fixed bandwidth & 0.626 & \textbf{0.399} & \textbf{0.622} & \textbf{0.663} & 0.653\\
 \bottomrule

### 2022-02-25 paper/paper.Rmd
Swiss roll and tp plots update

@@
 For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulate the data of size $N=2000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
+We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
 Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
@@
 This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply the algorithm described in \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
+We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=10,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
 
@@
 Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-This is the baseline, which we call *true densities*, to be compared with outliers from the kernel density estimates.
+These are *true outliers* to be compared with outliers from the kernel density estimates.
 
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$", echo = FALSE}
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.", echo = FALSE}
 knitr::include_graphics("figures/truedensity_4kernels.png")
@@
 
+```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.", echo = FALSE}
+knitr::include_graphics("figures/mappings_sr_tp.png")
+```
+
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
 One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
-The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})$, shown in the left plot of \autoref{fig:mappings}.
+The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
+The four colors in the mappings represents the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
 \begin{equation}
@@
 
-```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.", echo = FALSE}
-knitr::include_graphics("figures/mappings_sr_tp.png")
+```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
+Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+Both methods are not detecting outliers accurately when using t-SNE and UMAP.", echo = FALSE}
+knitr::include_graphics("figures/Swiss Roll5levels_outliers_comparison_4ml_3cases_riem0_08.png")
 ```
 
-```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of four manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from top: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and bottom: KDE with fixed bandwidth.
-Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE.
-Both methods are not detecting outliers accurately when using UMAP.", echo = FALSE}
-knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
-```
-
-Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
+Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
 According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
-1%, 50%, 99%, >99%.
-For each of the four manifold learning methods, namely ISOMAP, LLE, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same column, with the colors indicating the densities levels, the top row for true densities from the Gaussian mixture model, the middle row for highest density region plots with densities from our proposed variable KDE method, and the bottom for similar HDR plots with densities from KDE with fixed bandwidth.
-The top ten outliers with lowest densities are also highlighted in black with point indexes in blue.
-Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+1%, 5%, 50%, 99%, >99%.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+The top twenty outliers with lowest densities are highlighted in black with point indexes in blue.
+From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
+However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwith for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detect true outliers accurately.
+For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
 
@@
 data.frame(cors) %>% 
-  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  kable_styling(latex_options = "scale_down") %>%
-  # kable_paper(full_width = FALSE) %>%
+  mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
+  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  # kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = FALSE) %>%
   # column_spec(1, width = "6cm") %>% 
@@
 
+To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
+It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it is closest to rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are the very high in yellow, leading to a negative correlation.
+The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
+As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
+In conclusion, our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg.
+ISOMAP, or when the embedding are discontinuous, eg.
+t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
+
+<!-- Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
+As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
+Instead, we could roughly compare the location of outliers.
+For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+-->
+
 ### Twin peaks mapping
 
-For comparison, we use the same meta data with different mapping function, twin peak mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
-The four colors in both mappings represents the four gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)$.
+For comparison, we use the same meta data in \autoref{fig:metadensity} with different mapping function, twin peak mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak_outliers_comparison_4ml_3cases_riem01.png")
+knitr::include_graphics("figures/Twin Peak5levels_outliers_comparison_4ml_3cases_riem0_1.png")
 ```
@@
 The four colors represents the four gaussian kernels in the 2-D meta data.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem01.png")
+knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1.png")
 ```
@@
 ```{r tpcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_twinpeak_4ml_riem01.rda")
+load("figures/CorrelationTable_Twin Peak_4ml_riem0_1.rda")
 col.max <- apply(cors, 2, max)
@@
   column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4]) #%>%
-  # column_spec(6, bold = max.match[,5])
+  column_spec(5, bold = max.match[,4]) %>%
+  column_spec(6, bold = max.match[,5])
 ```

### 2022-02-25 reports/Simulation_swissroll.R
2d done

@@ -124,7 +124,7 @@ p <- 2
 ## Start from here! 
 ###--------------------------------------------------------
 set.seed(1234)
-mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[1]
+mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3]
 sr <- mldata(N = 2000, meta = "gaussian", mapping = mapping)
 swissroll <- sr$data %>% as.data.frame()
 preswissroll <- sr$metadata %>% as_tibble() %>% mutate(den = sr$den, label = c(rep(1:4, each = 500)))
@@ -133,8 +133,51 @@ colnames(preswissroll) <- c("X1", "X2", "den", "label")
 #         type = "scatter3d", mode = "markers", size = 1)
 plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = sr$den,
         type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
-scatterplot3d::scatterplot3d(swissroll$x, swissroll$y, swissroll$z, color = preswissroll$label, xlab = "X", ylab = "Y", zlab = "Z")
 
+# # mappings 3D plot, plot once
+# par(mfrow=c(1,2))
+# # twinpeak <- swissroll # first run with mapping index is 3
+# # pretwinpeak <- preswissroll
+# mypalette <- scales::hue_pal()(4)
+# scatterplot3d::scatterplot3d(swissroll$x, swissroll$y, swissroll$z, color = mypalette[as.numeric(preswissroll$label)], xlab = "X", ylab = "Y", zlab = "Z")
+# "paper/figures/mappings_sr_tp.png"
+
+# # Add small dots on basal plane and on the depth plane 
+# # http://www.sthda.com/english/wiki/impressive-package-for-3d-and-4d-graph-r-software-and-data-visualization#:~:text=18%2C%20CI%20%3D%20CI)-,3D%20fancy%20Scatter%20plot%20with%20small%20dots%20on%20basal%20plane,-A%20helper%20function
+# library("plot3D")
+# scatter3D_fancy <- function(x, y, z,..., colvar = z)
+# {
+#   panelfirst <- function(pmat) {
+#     XY <- trans3D(x, y, z = rep(min(z), length(z)), pmat = pmat)
+#     scatter2D(XY$x, XY$y, colvar = colvar, pch = ".", 
[…]

### 2022-02-26 paper/paper.Rmd
Writing for twin peak; Add point shapes

@@
 
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.", echo = FALSE}
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.
+The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
+The shapes indicate the four kernels.", echo = FALSE}
 knitr::include_graphics("figures/truedensity_4kernels.png")
@@
 
-```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.", echo = FALSE}
+```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors and shapes indicating the four kernels.
+Left: swiss roll mapping.
+Right: twin peak mapping.", echo = FALSE}
 knitr::include_graphics("figures/mappings_sr_tp.png")
@@
 Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-Both methods are not detecting outliers accurately when using t-SNE and UMAP.", echo = FALSE}
+The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
 knitr::include_graphics("figures/Swiss Roll5levels_outliers_comparison_4ml_3cases_riem0_08.png")
@@
   mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>%
@@
 As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-In conclusion, our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
 However, if the distortion is too severe, eg.
-ISOMAP, or when the embedding are discontinuous, eg.
+ISOMAP, or when the embedding is discontinuous, eg.
 t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
@@
 
-For comparison, we use the same meta data in \autoref{fig:metadensity} with different mapping function, twin peak mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
+For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 
-Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
-[ANALYSE TWIN PEAK PLOT]
-
-```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-The four colors represents the four gaussian kernels in the 2-D meta data.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1.png")
-```
-
-With severe distortion in the t-SNE and UMAP embeddings, both variable and fixed bandwidth methods can not produce accurate density estimates.
-It would then be very difficult to uncover the Gaussian mixture model structure and
+Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
+In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
+For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
+In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
+Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
+For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
+Also due to the clusters in the embedding, the outliers found in UMAP are also clustered.
 
@@
   mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>%
@@
 ```
+
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from varaible and fixed bandwidth KDE by \autoref{tab:tpcors}. 
+Again the highest correlations appears from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
+The rank correlations between variable and fixed bandwidth is equivalent to the thrid decimal place in Laplacian Eigemaps and UMAP.
+As for t-SNE, the four clusters in the embedding are less seperated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
+Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being `r round(min(cors), 3)`.
+This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+
+```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Variable bandwidth shows a strong linear positive relationship.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1.png")
+```
+
+In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
+The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
+Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
 

### 2022-02-26 paper/paper.tex
Writing for twin peak; Add point shapes

@@
 
-\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.}\label{fig:metadensity}
+\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.
+The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
+The shapes indicate the four kernels.}\label{fig:metadensity}
 \end{figure}
@@
 
-\caption{3-D Mappings of the meta data with colors indicating four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
+\caption{3-D Mappings of the meta data with colors and shapes indicating the four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
 \end{figure}
@@
 Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-Both methods are not detecting outliers accurately when using t-SNE and UMAP.}\label{fig:sroutliers}
+The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
 \end{figure}
@@
 
-\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings.}
+\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
 \centering
@@
 
-For comparison, we use the same meta data in \autoref{fig:metadensity} with different mapping function, twin peak mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D data shown in the right plot of \autoref{fig:mappings}.
+For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 
-Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
-{[}ANALYSE TWIN PEAK PLOT{]}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1}
-
-}
-
-\caption{Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-The four colors represents the four gaussian kernels in the 2-D meta data.}\label{fig:tpisomapden}
-\end{figure}
-
-With severe distortion in the t-SNE and UMAP embeddings, both variable and fixed bandwidth methods can not produce accurate density estimates.
-It would then be very difficult to uncover the Gaussian mixture model structure and
+Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
+In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
+For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
+In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
+Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
+For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
+Also due to the clusters in the embedding, the outliers found in UMAP are also clustered.
 
@@
 
-\caption{\label{tab:tpcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
 \centering
@@
 \end{table}
+
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from varaible and fixed bandwidth KDE by \autoref{tab:tpcors}.
+Again the highest correlations appears from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
+The rank correlations between variable and fixed bandwidth is equivalent to the thrid decimal place in Laplacian Eigemaps and UMAP.
+As for t-SNE, the four clusters in the embedding are less seperated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
+Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259.
+This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1}
+
+}
+
+\caption{Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Variable bandwidth shows a strong linear positive relationship.}\label{fig:tpisomapden}
+\end{figure}
+
+In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
+The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
+Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
 

### 2022-02-26 R/multidimension.R

@@ -70,6 +70,8 @@ X_new %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
 # summary(den1)
 # all.equal(den, den1) # TRUE
 
+# distance to (0,0,0,0,0) is the same, = radius = 7
+
 # ## Plot 5-d semi-sphere
 # library(tourr)
 # # col <- RColorBrewer::brewer.pal(3, "Dark2") # unique(flea$species)
@@ -152,7 +154,6 @@ summary(metric_isomap)
 
 fn <- metric_isomap$embedding
 dim(fn)
-
 # E1 <- fn[,1] # rename as Ed to match the aesthetics in plot_ellipse()
 # E2 <- fn[,2]
 # prob <- c(1, 10, 50, 95, 99)
@@ -233,66 +234,67 @@ par(mfrow=c(1,2))
 plot(den, flle$estimate, main = paste("Variable bandwidth correlation:", round(cor(den, flle$estimate), 3)))
 plot(den, fixden_lle$estimate, main = paste("Fixed bandwidth correlation:", round(cor(den, fixden_lle$estimate), 3)))
 cor(den, flle$estimate)
-# # [1] 0.903419
 cor(den, fixden_lle$estimate)
-# # [1] 0.844388
 
 
 
+
+
+
+# Laplacian eigenmaps
+
+## ----tsne, message=FALSE, warning=FALSE, eval=TRUE------------------------------
+x <- train
+method <- "LaplacianEigenmaps"
+# perplexity <- 30 # round(k / 3) # 30 by default
+# theta <- 0 # for exact tSNE in the C++ tSNE Barnes-Hut implementation
+# # tictoc::tic()
+metric_le <- metricML(x, s = s, k = k, radius = radius, method = method,
+                        # annmethod = annmethod, eps = 0, distance = distance, treetype = treetype,
+                        searchtype = searchtype,
+                        # perplexity = perplexity, theta = theta, 
+                      invert.h = TRUE)
+fn <- metric_le$embedding
+Rn <- metric_le$rmetric
+# E1 <- fn[,1]; E2 <- fn[,2]
+# fle <- vkde2d(x = E1, y = E2, h = Rn, binned = FALSE, eval.points = fn)
+# fxy_le <- hdrcde:::interp.2d(fle$x, fle$y, fle$z, x0 = E1, y0 = E2)
+# # plot_embedding(metric_le)
[…]

### 2022-02-26 reports/Simulation_swissroll.R

@@ -125,7 +125,7 @@ p <- 2
 ###--------------------------------------------------------
 set.seed(1234)
 mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3]
-sr <- mldata(N = 2000, meta = "gaussian", mapping = mapping)
+sr <- mldata(N = N, meta = "gaussian", mapping = mapping)
 swissroll <- sr$data %>% as.data.frame()
 preswissroll <- sr$metadata %>% as_tibble() %>% mutate(den = sr$den, label = c(rep(1:4, each = 500)))
 colnames(preswissroll) <- c("X1", "X2", "den", "label")
@@ -134,12 +134,15 @@ colnames(preswissroll) <- c("X1", "X2", "den", "label")
 plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = sr$den,
         type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
 
-# # mappings 3D plot, plot once
-# par(mfrow=c(1,2))
-# # twinpeak <- swissroll # first run with mapping index is 3
-# # pretwinpeak <- preswissroll
-# mypalette <- scales::hue_pal()(4)
-# scatterplot3d::scatterplot3d(swissroll$x, swissroll$y, swissroll$z, color = mypalette[as.numeric(preswissroll$label)], xlab = "X", ylab = "Y", zlab = "Z")
+# mappings 3D plot, plot once
+par(mfrow=c(1,2))
+# twinpeak <- swissroll # first run with mapping index is 3
+# pretwinpeak <- preswissroll
+mypalette <- scales::hue_pal()(4)
+scatterplot3d::scatterplot3d(swissroll$x, swissroll$y, swissroll$z, 
+                             color = mypalette[as.numeric(preswissroll$label)], 
+                             pch = rep(c(16, 17, 15, 3), each = N/4), # 1,2,0,3
+                             xlab = "X", ylab = "Y", zlab = "Z")
 # "paper/figures/mappings_sr_tp.png"
 
 # # Add small dots on basal plane and on the depth plane 
@@ -164,7 +167,7 @@ plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = sr$den,
[…]

### 2022-02-26 paper/paper.tex
5d plots and tables done; add colors as distance to center

@@
 
-For the high-dimensional case, we generate the meta data from a 5-dimensional semi-hypersphere.
-First, we simulate \(N=2000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(25,25,25,25)\).
+As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space.
+First, we simulate \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
 In order to manually generate anomalies, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as 15.
+The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as \(7\).
 Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_isomap_riem20}
+{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_4ml_riem20}
 
@@
 \centering
-\begin{tabular}[t]{l>{}l>{}l>{}l}
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
 \toprule
-  & ISOMAP & LLE & UMAP\\
+  & ISOMAP & LLE & Laplacian.Eigenmaps & UMAP\\
 \midrule
-Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{-0.130}\\
-Fixed bandwidth & 0.806 & 0.940 & -0.341\\
+Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{0.662} & \textbf{-0.130}\\
+Fixed bandwidth & 0.806 & 0.940 & -0.181 & -0.341\\
 \bottomrule

### 2022-02-26 R/multidimension.R

@@ -23,7 +23,7 @@ library(plotly)
 library(mvtnorm)
 Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1234)
-N <- 2000
+N <- 10000
 p <- 100
 mu1 = mu2 = 0
 s1 <- 1#/(15^2)
@@ -37,6 +37,7 @@ x4 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
 X <- cbind(x1, x2, x3, x4)
 head(X)
 range(X)
+label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
 
 # GMM density as true meta data density
 gmmdensity <- function(x) {
@@ -71,20 +72,23 @@ X_new %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
 # all.equal(den, den1) # TRUE
 
 # distance to (0,0,0,0,0) is the same, = radius = 7
+# calculate distance to (0,0,0,0) and color them
+label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
+dist2center <- sqrt(r^2 - X_new[,5]^2)
 
-# ## Plot 5-d semi-sphere
-# library(tourr)
-# # col <- RColorBrewer::brewer.pal(3, "Dark2") # unique(flea$species)
-# # animate_xy(flea[, 1:6], col = flea$species)
-# 
-# animate_xy(X_new[,1:5], col = viridis(length(den), option = "magma"))
-# # catogorize density levels and color them accordingly
-# 
+## Plot 5-d semi-sphere, run once
+library(tourr)
+colors <- colourvalues::colour_values(- dist2center, palette = "magma") # generate colors for locations
+pchs <- c(rep(16, 0.99*N), rep(17, 0.01*N)) # point shapes for kernels
+animate_xy(X_new[,1:5], col = colors, pch = pchs, cex = 0.8, 
+           axes = "bottomleft", fps=15
+           )
+# "figures/tourr_5d_semisphere.png"
+
 # library(geozoo)
 # sphere <- sphere.hollow(p = 5)
 # sphere$points <- X_new
 # sphere
-
 
 
 ## QR decomposition
@@ -142,7 +146,7 @@ treetype <- "kd"
 searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
[…]

### 2022-02-26 paper/paper.Rmd
5d plots and tables done; add colors as distance to center

@@
 
-For the high-dimensional case, we generate the meta data from a 5-dimensional semi-hypersphere.
-First, we simulate $N=2000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(25,25,25,25)$.
+As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space.
+First, we simulate $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
 In order to manually generate anomalies, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-The fifth dimension is calculated to satisfy $x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2$ where $r$ is set as 15.
+The fifth dimension is calculated to satisfy $x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2$ where $r$ is set as $7$.
 Similarly, the Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
@@
 Top: ISOMAP; middle: LLE; bottom: UMAP.", echo = FALSE}
-knitr::include_graphics("figures/fived_density_comparison_isomap_riem20.png")
+knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
 ```
@@
   column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) #%>%
-  # column_spec(5, bold = max.match[,4]) 
+  column_spec(4, bold = max.match[,3]) %>%
+  column_spec(5, bold = max.match[,4])
 ```

### 2022-02-26 R/hpc_5d.R
100d simulation process writing

@@ -1,0 +1,133 @@
+# Run on HPC for N=10000, Isomap
+library(tidyverse)
+library(dimRed)
+# library(reticulate)
+library(here)
+library(viridis)
+library(hdrcde)
+library(igraph)
+# library(matrixcalc)
+# library(akima)
+# library(car)
+library(ggforce)
+library(ks)
+library(patchwork)
+library(plotly)
+library(mvtnorm)
+Jmisc::sourceAll(here::here("R/sources"))
+set.seed(1234)
+# scen <- as.numeric(commandArgs()[[6]])
+
+paste("Start at:", Sys.time())
+
+N <- 10000
+p <- 100
+mu1 = mu2 = 0
+s1 <- 1#/(15^2)
+s2 <- 2#/(15^2)
+p1 <- 0.99
+p2 <- 0.01
+x1 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+x2 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+x3 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+x4 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+X <- cbind(x1, x2, x3, x4)
+
+# GMM density as true meta data density
+gmmdensity <- function(x) {
+  p1 * dmvnorm(x, rep(mu1, 4), diag(s1, 4)) + 
+    p2 *  dmvnorm(x, rep(mu2, 4), diag(s2, 4))
+}
+den <- apply(X, 1, gmmdensity)
+
+# semi-sphere radius
+r <- ceiling(max(sqrt(x1^2 + x2^2 + x3^2 + x4^2)))
+r # radius = 15 for sigma=5
+range(X)
+
+X_new <- cbind(X, 
+               x5 = sqrt(r^2 - (x1^2 + x2^2 + x3^2 + x4^2)),
+               matrix(0, N, p - 5)
+)
+# head(X_new)
+# X_new %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
+
+# distance to (0,0,0,0,0) is the same, = radius = 7
+# calculate distance to (0,0,0,0) and color them
+label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
+dist2center <- sqrt(r^2 - X_new[,5]^2)
+
[…]

### 2022-02-26 paper/paper.Rmd
100d simulation process writing

@@
 We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=10,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
 
-## 2-D Meta data from a Gaussian Mixture Model {#twodgaussian}
+## 3-D mapping from a 2-D Gaussian Mixture Model {#twodgaussian}
 
@@
 
-## 5-D data embedding in 100-D space {#fivedgaussian}
+## 100-D mapping from a 5-D semi-hypersphere {#fivedgaussian}
 
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space.
-First, we simulate $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
-In order to manually generate anomalies, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-The fifth dimension is calculated to satisfy $x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2$ where $r$ is set as $7$.
-Similarly, the Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
-\autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package *tourr* [REFERENCE].
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size $N$ (denote the transpose of the data matrix as $X$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $A$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
-For any real matrix $A$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $Q$ and $R$ so that $A = QR$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $Q^\prime Q = I$, and $R$ is an upper triangular matrix.
-Matrix $Q$ is used as the rotation matrix and it satisfies
-$X^\prime X=(QX)^\prime(QX)$, where the rotated data matrix $(QX)^\prime$ is now the input for the manifold learning algorithms.
+As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in $d=5$ with manifold learning.
+First, we simulate $N=2,000$ points, $(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
+In order to manually add anomalies to be distant points from the means, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
+The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation $X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. 
+The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE].
+The round and triangular point shapes index the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the distance between the simulated $4-D$ data point from Gaussian mixture model and the kernel means $(0, 0, 0, 0)^\prime$.
+The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
+It can be seen that the most distant points are in triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
+The dark colors also indicate that they are the true outliers because of their low densities.
 
-```{r fivedmeta, fig.align = 'center', out.width = "80%", fig.cap="Scatterplot display of the animation of a 5-D tour path.", echo = FALSE}
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the $100$-dimensional data of size $N$ (denote the transpose of the data matrix as $\pmb{X}_0$ with dimension $100 \times N$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
+For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix.
+Matrix $\pmb{Q}$ satisfies
+$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
+Therefore, we use matrix $Q$ as the rotation matrix for  where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms.
+Again, we set the embedding dimension to be equal to the meta data dimension $d=5$.
+
+```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.", echo = FALSE}
 knitr::include_graphics("figures/tourr_5d_semisphere.png")

### 2022-02-26 paper/paper.tex
100d simulation process writing

@@
 We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=10,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
 
 \hypertarget{twodgaussian}{%
-\subsection{2-D Meta data from a Gaussian Mixture Model}\label{twodgaussian}}
+\subsection{3-D mapping from a 2-D Gaussian Mixture Model}\label{twodgaussian}}
 
@@
 \hypertarget{fivedgaussian}{%
-\subsection{5-D data embedding in 100-D space}\label{fivedgaussian}}
+\subsection{100-D mapping from a 5-D semi-hypersphere}\label{fivedgaussian}}
 
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space.
-First, we simulate \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
-In order to manually generate anomalies, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as \(7\).
-Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
-\autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size \(N\) (denote the transpose of the data matrix as \(X\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
-For any real matrix \(A\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(Q\) and \(R\) so that \(A = QR\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(Q^\prime Q = I\), and \(R\) is an upper triangular matrix.
-Matrix \(Q\) is used as the rotation matrix and it satisfies
-\(X^\prime X=(QX)^\prime(QX)\), where the rotated data matrix \((QX)^\prime\) is now the input for the manifold learning algorithms.
+As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
+First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
+In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
+The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
+The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
+The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
+The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
+It can be seen that the most distant points are in triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
+The dark colors also indicate that they are the true outliers because of their low densities.
+
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
+For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix.
+Matrix \(\pmb{Q}\) satisfies
+\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
+Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
+Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).
 
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{figures/tourr_5d_semisphere}
+{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
 
@@
 
-\caption{Scatterplot display of the animation of a 5-D tour path.}\label{fig:fivedmeta}
+\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
 \end{figure}

### 2022-02-26 R/multidimension.R

@@ -76,19 +76,19 @@ X_new %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
 label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
 dist2center <- sqrt(r^2 - X_new[,5]^2)
 
-## Plot 5-d semi-sphere, run once
-library(tourr)
-colors <- colourvalues::colour_values(- dist2center, palette = "magma") # generate colors for locations
-pchs <- c(rep(16, 0.99*N), rep(17, 0.01*N)) # point shapes for kernels
-animate_xy(X_new[,1:5], col = colors, pch = pchs, cex = 0.8, 
-           axes = "bottomleft", fps=15
-           )
-# "figures/tourr_5d_semisphere.png"
-
-# library(geozoo)
-# sphere <- sphere.hollow(p = 5)
-# sphere$points <- X_new
-# sphere
+# ## Plot 5-d semi-sphere, run once
+# library(tourr)
+# colors <- colourvalues::colour_values(- dist2center, palette = "magma") # generate colors for locations
+# pchs <- c(rep(16, 0.99*N), rep(17, 0.01*N)) # point shapes for kernels
+# animate_xy(X_new[,1:5], col = colors, pch = pchs, cex = 0.8, 
+#            axes = "bottomleft", fps=15
+#            )
+# # "figures/tourr_5d_semisphere.png"
+# 
+# # library(geozoo)
+# # sphere <- sphere.hollow(p = 5)
+# # sphere$points <- X_new
+# # sphere
 
 
 ## QR decomposition
@@ -431,7 +431,7 @@ result <- (p/p1/p2/p3) + plot_layout(guides = 'collect') &
 gt <- patchwork::patchworkGrob(result)
 gt <- gridExtra::grid.arrange(gt, left = "Estimated density", bottom = "True density")
 gt
-ggsave(paste0("paper/figures/", "fived", "_density_comparison_4ml_riem", format(riem.scale, decimal.mark = "_"), ".png"), gt, width = 8, height = 10, dpi = 300)
+ggsave(paste0("paper/figures/", "fived", N, "_density_comparison_4ml_riem", format(riem.scale, decimal.mark = "_"), ".png"), gt, width = 8, height = 10, dpi = 300)
 
 
 
@@ -453,4 +453,4 @@ cors %>%
   kable_styling(latex_options = "scale_down") %>%
[…]

### 2022-02-26 paper/paper.Rmd
100d result analysis done; Simulation section done

@@
 
+```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.", echo = FALSE}
+knitr::include_graphics("figures/tourr_5d_semisphere.png")
+```
+
 As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in $d=5$ with manifold learning.
@@
 Matrix $\pmb{Q}$ satisfies
-$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
+$\pmb{X}_0^\prime \pmb{X}_0=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
 Therefore, we use matrix $Q$ as the rotation matrix for  where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms.
@@
 
-```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.", echo = FALSE}
-knitr::include_graphics("figures/tourr_5d_semisphere.png")
+```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
+The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.", echo = FALSE}
+knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
 ```
 
-Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
-[ANALYSE TWIN PEAK PLOT]
-
-```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
-Top: ISOMAP; middle: LLE; bottom: UMAP.", echo = FALSE}
-knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
-```
+In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only application to embedding dimension within three.
+Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
+For well estimated densities, the true outliers with low true densities will also have low estimated densities, which suggest that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
+This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
+For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
+The rank correlations are also presented in \autoref{tab:fivedcors} with variable bandwidth always have a higher correlation of density ranking.
+This again suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
+In KDE with a fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
+By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
 
@@
   mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.
+Variable bandwidth always outperfoms fixed bandwidth although they both have negative correlations in UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width

### 2022-02-26 paper/paper.tex
100d result analysis done; Simulation section done

@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
+
+}
+
+\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
+\end{figure}
+
 As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
@@
 Matrix \(\pmb{Q}\) satisfies
-\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
+\(\pmb{X}_0^\prime \pmb{X}_0=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
 Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
 Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
-
-}
-
-\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
-\end{figure}
-
-Similar to \autoref{fig:sroutliers}, four manifold learning embeddings are obtained and used to detect outliers with two bandwidth selection methods in \autoref{fig:tpoutliers}.
-{[}ANALYSE TWIN PEAK PLOT{]}
 
@@
 
-\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth. Top: ISOMAP; middle: LLE; bottom: UMAP.}\label{fig:fivedisomapden}
+\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
+The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.}\label{fig:fivedisomapden}
 \end{figure}
+
+In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only application to embedding dimension within three.
+Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
+For well estimated densities, the true outliers with low true densities will also have low estimated densities, which suggest that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
+This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
+For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
+The rank correlations are also presented in \autoref{tab:fivedcors} with variable bandwidth always have a higher correlation of density ranking.
+This again suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
+In KDE with a fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
+By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
 
@@
 
-\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.
+Variable bandwidth always outperfoms fixed bandwidth although they both have negative correlations in UMAP.}
 \centering

### 2022-02-26 paper/paper.Rmd
Fix kable column names with as_tibble(cors, rownames = " ")

@@
 max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
+as_tibble(cors, rownames = " ") %>% 
   mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
@@
 max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
+as_tibble(cors, rownames = " ") %>% 
   mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 

### 2022-02-26 paper/paper.Rmd
Remove old figures

--- 2022-02-26 paper/paper.Rmd
@@
 The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
-knitr::include_graphics("figures/Swiss Roll5levels_outliers_comparison_4ml_3cases_riem0_08.png")
+knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
 ```
@@
 Matrix $\pmb{Q}$ satisfies
-$\pmb{X}_0^\prime \pmb{X}_0=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
+$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
 Therefore, we use matrix $Q$ as the rotation matrix for  where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms.
@@
 For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
-The rank correlations are also presented in \autoref{tab:fivedcors} with variable bandwidth always have a higher correlation of density ranking.
-This again suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
+This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
 In KDE with a fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
@@
   mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.
-Variable bandwidth always outperfoms fixed bandwidth although they both have negative correlations in UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
@@
 
-## Victoria smart meter dataset
+## Irish smart meter dataset
 

### 2022-02-26 paper/paper.tex
Fix kable column names with as_tibble(cors, rownames = " ")
Application data introduction and steps

@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll5levels_outliers_comparison_4ml_3cases_riem0_08}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
 
@@
 Matrix \(\pmb{Q}\) satisfies
-\(\pmb{X}_0^\prime \pmb{X}_0=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
+\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
 Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
@@
 
-\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.
-Variable bandwidth always outperfoms fixed bandwidth although they both have negative correlations in UMAP.}
+\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
 \centering
@@
 
-\hypertarget{victoria-smart-meter-dataset}{%
-\subsection{Victoria smart meter dataset}\label{victoria-smart-meter-dataset}}
+\hypertarget{irish-smart-meter-dataset}{%
+\subsection{Irish smart meter dataset}\label{irish-smart-meter-dataset}}
 
-{[}TODO{]}
+In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
+The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
+We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
 
-Next, we consider smart-meter data for residential and non-profiled meter consumers, collected in the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data}.
-Electricity smart meters record consumption, on a near real-time basis, at the level of individual commercial and residential properties.
-The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} does not include energy for heating systems since it is either metered separately, or households use a different source of energy, such as oil or gas.
-In this study, the installed cooling systems are also not reported.
+For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
+\textcite{Cheng2021-ex} propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
+\textcite{Cheng2021-ex} use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
+By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
 
-We use measurements of half-hourly electricity consumption gathered from 3,639 residential consumers over 535 consecutive days.
-Every meter provides electricity consumption between 14 July 2009 and 31 December 2010.
-Demand data from two smart meters (ID 1003 and 1539) are shown in \autoref{fig:smartmeter} as time series plots.
-It is obvious that these meters have relatively different patterns.
-Meter 1539 (bottom of \autoref{fig:smartmeter}) has a period of around 150 days with lower (approximately half) the electricity usage of remaining days and is otherwise relatively stable.
-In contrast, Meter 1003 (top of \autoref{fig:smartmeter}) exhibits regular spikes on weekends.
-
-For electricity demand data, one particular problem of interest is to visualize and identify households or periods of the week with anomalous usage patterns.
-For this reason, the object of interest in comparing households or periods of the week is the distribution of electricity demand rather than the raw data itself \autocite{Hyndman2018-ia}.
-An additional advantage of looking at distributions rather than raw data is that it provides a convenient mechanism for handling missing data which can be a significant problem in smart meter data applications.
-In the rest of this section, we first describe how discrete approximations to the distributions of interest are computed.
-We then approximate the distances between vectors containing probabilities with the consistent estimator in \autoref{estimator}, which allows us to apply manifold learning techniques on statistical manifolds.
-
-Let \(d_{i,t}\) denote the electricity demand for observation \(i\) and for time period \(t\) (subsequently we will see that \(i\) can either index the household, time of the week, or both, while \(t\) may index the week or half-hour period).
-The objective is to estimate the distances between the distribution of electricity demand for observation \(i\), \(F_i\), over time.
-Since the raw data could be thought of as samples generated from the true electricity usage distributions for different households and different periods, we could apply the consistent Hellinger distance estimator derived in \autoref{estimator} to the smart meter data.
-The first step is to group all data together regardless of \(i\) and \(t\), and take the data as samples from the reference probability distribution \(R\).
-The sample size \(r=93,616,914\).
-Next, we construct \(T_r=100\) intervals with about \(1\%\) samples from \(R\) within each interval (except for the last interval).
-The segments \(\{I_g^r\}_{g=1,2,\dots,100}\) can then be used to construct discrete distribution approximation to \(F_i\) by counting the ratio of samples points from \(F_i\) fallen into segment \(I_g^r\).
-This is found by computing \(\pi_{i,g}=(1/m)\sum_t I(d_{i,t} \in I_g^r)\) where \(m\) is the sample size from \(F_i\).
-Then the Hellinger distance between \(F_i\) and \(F_j\) could be estimated using Equation \eqref{eq:hellingerest1}.
-In this way, we make it possible to search nearest neighbors more efficiently using ANN methods and further apply manifold learning algorithms.
+In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
+Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+The data processing steps has been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
+Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
+However, we could generate all the density estimates with the exsiting KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
 

### 2022-02-26 R/Simulation2d_swissroll_twinpeak.R

@@ -1,0 +1,626 @@
+## ----setup, include=FALSE-------------------------------------------------------
+# knitr::opts_chunk$set(echo = TRUE)
+
+
+## ----libraries, message=FALSE, echo=TRUE, results='hide'------------------------
+library(tidyverse)
+library(dimRed)
+library(reticulate)
+library(here)
+library(viridis)
+library(hdrcde)
+library(igraph)
+library(matrixcalc)
+library(akima)
+library(car)
+library(ggforce)
+library(ks)
+library(patchwork)
+# library(copula)
+library(plotly)
+library(kableExtra)
+Jmisc::sourceAll(here::here("R/sources"))
+# set.seed(1)
+
+# data size
+N <- 2000
+p <- 2
+
+# # ----Swiss roll dataset with a sparse area------------------------
+# height <- 30
+# x <- (3 * pi / 2) * (1 + 2 * runif(N*1.05, 0, 1))
+# y <- height * runif(N*1.05, 0 , 1)
+# sr <- cbind(x * cos(x), y, x * sin(x))
+# colnames(sr) <- c("x", "y", "z")
+# sr <- cbind(as_tibble(sr), col = "grey")
+# lowden <- x >= 13 & y >= 18
+# sr$col[lowden] <- "red"
+# removed <- sample(which(lowden), size = N*0.05, replace = FALSE)
+# lowdenpoints <- sr[setdiff(which(lowden), removed), ]
+# sr <- sr[-removed, ]
+# scatterplot3d::scatterplot3d(sr, color = sr$col)
+# # rgl::plot3d(sr, col = sr$col)
+# 
+# 
+# 
+# # ----Swiss roll from copula-----------------------------------------
+# # ## a 3-dimensional normal copula
+# # norm.cop <- normalCopula(0.5, dim = p)
+# # u <- rCopula(N, norm.cop)
+# # den <- dCopula(u, norm.cop)
+# 
+# # # 3-d t copula
+# # t.cop <- tCopula(c(0.5, 0.3), dim = p, dispstr = "toep",
+# #                  df = 2, df.fixed = TRUE)
+# # u <- rCopula(N, t.cop)
+# # den <- dCopula(u, t.cop)
+# # 
+# ## a 2-dimensional Clayton copula
+# cl3 <- claytonCopula(2, dim = p)
+# u <- rCopula(N, cl3)
+# # pairs(u)
[…]

### 2022-02-26 paper/paper.Rmd
Application data introduction and steps

@@
 
-<!-- Dataset -->
-
-<!-- Goal and process -->
-
-<!-- Results (Plots and tables) -->
-
-<!-- Analysis and summary -->
-
 ## Irish smart meter dataset
 
-[TODO]
+In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010.
+The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
+We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.
 
-Next, we consider smart-meter data for residential and non-profiled meter consumers, collected in the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data].
-Electricity smart meters record consumption, on a near real-time basis, at the level of individual commercial and residential properties.
-The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] does not include energy for heating systems since it is either metered separately, or households use a different source of energy, such as oil or gas.
-In this study, the installed cooling systems are also not reported.
+For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
+@Cheng2021-ex propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
+@Cheng2021-ex use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
+By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
+
+ In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
+Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+The data processing steps has been clearly stated in the application section of @Cheng2021-ex and they are skipped here.
+Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
+However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
 
@@
 
-We use measurements of half-hourly electricity consumption gathered from 3,639 residential consumers over 535 consecutive days.
-Every meter provides electricity consumption between 14 July 2009 and 31 December 2010.
-Demand data from two smart meters (ID 1003 and 1539) are shown in \autoref{fig:smartmeter} as time series plots.
-It is obvious that these meters have relatively different patterns.
-Meter 1539 (bottom of \autoref{fig:smartmeter}) has a period of around 150 days with lower (approximately half) the electricity usage of remaining days and is otherwise relatively stable.
-In contrast, Meter 1003 (top of \autoref{fig:smartmeter}) exhibits regular spikes on weekends.
-
-For electricity demand data, one particular problem of interest is to visualize and identify households or periods of the week with anomalous usage patterns.
-For this reason, the object of interest in comparing households or periods of the week is the distribution of electricity demand rather than the raw data itself [@Hyndman2018-ia].
-An additional advantage of looking at distributions rather than raw data is that it provides a convenient mechanism for handling missing data which can be a significant problem in smart meter data applications.
-In the rest of this section, we first describe how discrete approximations to the distributions of interest are computed.
-We then approximate the distances between vectors containing probabilities with the consistent estimator in \autoref{estimator}, which allows us to apply manifold learning techniques on statistical manifolds.
-
-<!-- ## Data processing {#dataprocessing} -->
-
-Let $d_{i,t}$ denote the electricity demand for observation $i$ and for time period $t$ (subsequently we will see that $i$ can either index the household, time of the week, or both, while $t$ may index the week or half-hour period).
-The objective is to estimate the distances between the distribution of electricity demand for observation $i$, $F_i$, over time.
-Since the raw data could be thought of as samples generated from the true electricity usage distributions for different households and different periods, we could apply the consistent Hellinger distance estimator derived in \autoref{estimator} to the smart meter data.
-The first step is to group all data together regardless of $i$ and $t$, and take the data as samples from the reference probability distribution $R$.
-The sample size $r=93,616,914$.
-Next, we construct $T_r=100$ intervals with about $1\%$ samples from $R$ within each interval (except for the last interval).
-The segments $\{I_g^r\}_{g=1,2,\dots,100}$ can then be used to construct discrete distribution approximation to $F_i$ by counting the ratio of samples points from $F_i$ fallen into segment $I_g^r$.
-This is found by computing $\pi_{i,g}=(1/m)\sum_t I(d_{i,t} \in I_g^r)$ where $m$ is the sample size from $F_i$. 
-Then the Hellinger distance between $F_i$ and $F_j$ could be estimated using Equation \@ref(eq:hellingerest1).
-In this way, we make it possible to search nearest neighbors more efficiently using ANN methods and further apply manifold learning algorithms.
-
 # Conclusion {#conclusion}
 
-Summary of the contribution and conclusion
+<!-- Summary of the contribution and conclusion -->
 
-Details about the process
+<!-- Details about the process -->
 
-Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
+<!-- Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
+-->
 
-Discuss simulation with non-uniform meta data.
+<!-- Discuss simulation with non-uniform meta data. -->
 

### 2022-02-26 paper/paper.Rmd
Add conclusions; application results to be updated

@@
 
+<!-- 5x2 HDRs plot of vkde and hdr, 5levels, 5 ML methods, kdtrees, eps=1 -->
+<!-- index top 20 outliers -->
+<!-- compare density ranking of VKDE and HDR,  -->
+<!-- scatterplot no colors, table with one row for 5 ML methods -->
+
+<!-- top 1 typical points + top 2 outlier, HDR -->
+<!-- top 1 typical points + top 2 outlier, VKDE -->
+<!-- raw data plot side by side -->
+<!-- density plot in two figures -->
+
 # Conclusion {#conclusion}
-
-<!-- Summary of the contribution and conclusion -->
-
-<!-- Details about the process -->
 
@@
 
-<!-- ## Using dimension reduction algorithms -->
+In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
+The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
+By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
+We compare our propsed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that varibale bandwidth could improve the density estimation given a good manifold learning embedding.
 
-<!-- Now we can try to see what happens when we use an actual dimension reduction algorithm, for instance ISOMAP -->
+As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
+We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in differnt ways.
 
-<!-- ```{r isomap, echo=TRUE,message=FALSE, eval=FALSE} -->
-<!-- dat<-dimRedData(cbind(x,y,z)) #Prepare data to use in dimRed -->
-<!-- emb<-embed(dat,"Isomap") #Isomap embedding -->
-<!-- plot(hdrscatterplot(emb@data@data[,1],emb@data@data[,2])) #HDR plot -->
-<!-- ``` -->
+There are several open questions to be explored.
+The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of @Cheng2021-ex.
+The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
+The scale of distortion in each point in manifold learning could vary a lot.
+If we could smooth across all data points, eg.
+multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
+The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
+The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more close to real-life datasets.
 
-<!-- Which seems to not distort things too much.  The HDR plot still identifies outliers (because it has to) but they are just scattered around the corners in a very non-systematic way.  -->
+# Acknowledgment {-}
 
-<!-- For something that leads to a lot of distortion have a look t t-SNE. -->
-
-<!-- ```{r tsneplot, echo=TRUE,message=FALSE, eval=FALSE} -->
-<!-- emb_tsne<-embed(dat,"tSNE") #tSNE embedding -->
-<!-- plot(hdrscatterplot(emb_tsne@data@data[,1],emb_tsne@data@data[,2])) #HDE plot -->
-<!-- ``` -->
+This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.

### 2022-02-26 paper/paper.tex
Add conclusions; application results to be updated

@@
 Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
-However, we could generate all the density estimates with the exsiting KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
+However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
 
@@
 
-Summary of the contribution and conclusion
+In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
+The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
+By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
+We compare our propsed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that varibale bandwidth could improve the density estimation given a good manifold learning embedding.
 
-Details about the process
+As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
+We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in differnt ways.
 
-Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
+There are several open questions to be explored.
+The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-ex}.
+The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
+The scale of distortion in each point in manifold learning could vary a lot.
+If we could smooth across all data points, eg.
+multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
+The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
+The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more close to real-life datasets.
 
-Discuss simulation with non-uniform meta data.
+\hypertarget{acknowledgment}{%
+\section*{Acknowledgment}\label{acknowledgment}}
+\addcontentsline{toc}{section}{Acknowledgment}
+
+This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
 

### 2022-02-27 paper/paper.Rmd
Section 2 theory done

@@
   email: Anastasios.Panagiotelis@sydney.edu.au
-abstract: ""
+abstract: "In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
+The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
+By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
+We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways."
 keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
@@
 
+\newpage
+
 # Introduction
@@
 
-<!-- Now I briefly describe different steps involved in this main chapter.  -->
 By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
@@
 In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
-In this section, we provide justification for the use of Riemannien matrix as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data embedded in 3-D and the second with a 100-dimensional meta data.
-\autoref{application} contains the application to visualize and identify anomalies in the [TODO] dataset.
+In this section, we provide justification for the use of Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
+\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
 Conclusions and discussions are presented in \autoref{conclusion}.
@@
 
-To start with, we introduce the notations in this manuscript.
-Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metrix.
-Finally, we propose our novel method to detect anomalies for high-dimensional data set.
+To start with, we introduce the notations used in this manuscript.
+Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
+Readers familiar with these topics could skip the corresponding subsections.
+Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
 
@@
 
-Manifold learning finds a $d$-dimensional representation of data that lie on a manifold $\mathcal{M}$ embedded in a $p$-dimensional ambient space with $d \ll p$.
-We will denote the original data (or 'input' points) as $x_i$, $i=1,\dots,N$, where $x_i\in\mathbb{R}^p$, while the low-dimensional representation (or 'output' points) will be denoted as $y_i$, $i=1,\dots,N$, where $y_i\in\mathbb{R}^d$.
-Where two subscripts are used (e.g.
-$x_{ih}$ or $y_{ih}$), the second subscript refers to the $h^{th}$ coordinate or dimension of the data.
-<!-- Pairwise distances between input points $x_i$ and $x_j$ are denoted $\delta_{ij}$ while pairwise distances between output points $y_i$ and $y_j$ are denoted $d_{ij}$, where $i,j=1,\dots,N$.
-Unless otherwise stated, these distances are assumed to be Euclidean.
--->
+Denote the $s$-dimensional input data with observations $x_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$.
+The underlying manifold $\mathcal{M}$ has the true density $f$.
+Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $\mathcal{M}$, with output data points $y_i, i = 1, 2, \dots, N$, where $y_i \in \mathcal{R}^d$ and $d \ll s$.
+The Riemannian metric of the embedding is denoted as $\pmb{g_i}, i = 1, 2, \dots, N$.
+In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$.
 
-## Two dimensional kernel density estimation
+## Kernel density estimator on Riemannian manifolds
 
-<!-- In general a multivariate kernel density estimate looks something like this -->
+In the case of non-Euclidean sample space, the observations often lie on a manifold with the differentiable structure called the Riemannian manifold. 
+Consider a compact Riemannian manifold $(\mathcal{M}, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
+Assume $(\mathcal{M}, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by $g$ and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold.
+Denote $A_i, i = 1, 2, \dots, N$, where $A_i \in \mathcal{R}^d$, as i.i.d.
+random objects on $\mathcal{M}$ with density $f$.
+For each $p\in \mathcal{M}$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
+\begin{equation}
+\label{eq:denriem}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
+\end{equation}
+where $\theta_p(x)$ is the volume density function on the manifold, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative function, and $h$ is the bandwidth.
+@Pelletier2005-vu discussed the volume density function $\theta_p(x)$ defined for $p$ in a neighborhood of $x$ on the manifold, with the geodesic normal coordinates at $x$, $\theta_p(x)$ is equal to the determinant of the Riemannian metric $g$ at $\exp_p^{-1}(x)$.
+The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
+\begin{equation}
+\label{eq:denconsist}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
+\end{equation}
+and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].
 
-<!-- $$ -->
-<!-- \hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{x}_i) -->
-<!-- $$ -->
+However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
+Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
+Therefore, based on the estimator in \@ref(eq:denriem), we propose to a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
+\begin{equation}
+\label{eq:denestimator}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\end{equation}
+where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.
 
-<!-- where if a Gaussian kernel is used -->
+An important goal of manifold learning is to recover the local or global features of the data.
+In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric $g$.
+This could also be used to fix the geometric distortion in the embedding.
+It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for each data point.
+Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \@ref(eq:denestimator).
+In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
 
-<!-- $$ -->
-<!-- K_{\pmb{H}}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{x}_i)\right] -->
-<!-- $$ -->
+## Multivariate kernel density estimation
 
-<!-- The matrix $\pmb{H}$ is called the bandwidth matrix and is very important.  -->
+In general, a multivariate variable kernel density estimate of $f$ at point $\pmb{x} \in \mathcal{R}^d$ is defined as
+$$
+\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i),
+$$
+where $a_i, i = 1, 2, \dots, N$ is the random sample drawn from a density $f$, $\pmb{h}_i$ is a $d\times d$ symmetric positive definite matrix that is variable at $i$, and $K_{\pmb{h}_i}$ is the kernel function
+$$
+K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i)=(2\pi)^{-d/2}|\pmb{h}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{a}_i)'\pmb{h}_i^{-1}(\pmb{x}-\pmb{a}_i)\right],
+$$
+with $|\pmb{h}_i|$ being the determinant of matrix $\pmb{h}_i$.
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$.
+An adaptive bandwidth could be used to control the amount of smoothing on the location $\pmb{x}$ [balloon estimator; @Terrell1992-ef] or the data point $\pmb{a}_i$ [sample smoothing estimator; @Terrell1992-ef].
+We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
+@Duong2007-up present an R package *ks* that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and is not applicable to full bandwidth matrix.
 
-For a bivariate random sample $\pmb X_1, \pmb X_2, \ldots, \pmb X_n$ 
-drawn from a density $f$, 
-the kernel density estimate is defined by
-$$
-\hat{f} (\pmb x; \pmb H) = n^{-1}\sum_{i=1}^n K_{\pmb H} ( \pmb x - \pmb{X}_i)
-$$
-where $\pmb x = (x_1, x_2)^T$ and $\pmb{X}_i = (X_{i1}, X_{i2})^T, i = 1, 2,  
-\ldots, n$.  Here 
-$K(\pmb x)$ is the kernel which is a symmetric probability density function, 
-$\pmb H$ is the bandwidth matrix which is symmetric and positive-definite,  
-and $K_{\pmb H}(\pmb x) = |\pmb H|^{-1/2} K( \pmb H^{-1/2} \pmb x)$. 
-The choice of $K$ is not crucial: we take 
-$K(\pmb x) = (2\pi)^{-1} \exp(-\tfrac{1}{2} \pmb x^T \pmb x)$ the standard normal
-throughout.  
-In contrast, the choice of $\pmb H$ is crucial in determining the performance of $\hat f$. 
-The most common parameterizations of the bandwidth matrix
-are the diagonal and the 
-general or unconstrained which has no restrictions on $\pmb H$
-provided that $\pmb H$ remains positive definite and symmetric, that is 
-$$
-\pmb H = \begin{bmatrix}h_1^2 & 0 \\0 & h_2^2 \end{bmatrix}
-\ \mathrm{or} \ 
-\pmb H = \begin{bmatrix}h_1^2 & h_{12} \\ h_{12}  & h_2^2 \end{bmatrix}.
-$$
-This latter parameterization allows kernels to have an arbitrary orientation
-whereas the former only allows kernels which are oriented to the
-co-ordinate axes.
-
-<!-- The bandwidth matrix is about smoothing and it is.  -->
-An alternative way to think of kernel estimation is that kernel densities 'borrow strength' from nearby points and the bandwidth determines what is "nearby".
-If the bandwidth is large then all points are "nearby" and we get an overly smooth kernel density estimate.
-The interesting thing about a bandwidth matrix is that it allows for different notions of what is "nearby" along different coordinates and even along diagonal directions.
-
-  <!-- Riemannian matrix could be used to measure distortion -->
-
-  <!-- Use riemannian matrix as variable bandwidth -->
-  <!--     Bandwidth is a symmetric positive-definite square matrix, similar to riemannian matrix -->
-
-## Use riemannian matrix as variable bandwidth
-
-The Riemannian estimated using the method of Perrault Joncas and Meila (2009) gives some idea of the distortion of an embedding (or so they claim).
-Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
-The Riemannian gives us an idea of the direction and angle of this stretching.
-The Riemannian is quite a technical concept but an important thing to understand is that the estimate that comes out of Perrault Joncas algorithm is a square matrix.
-
-We saw how points that are far apart in the embedding may not have been so far apart on the original manifold.
-The Riemannian gives us some way of correcting this.
-Similarly a bandwidth matrix in a kernel density estimate is all about determining the "directions" in which there should be more or less "closeness".
-So the basic idea is to replace the kernel density estimate with
-
-$$
-\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)\\
-K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{x}_i)\right]
-$$
-
-where $H_i$ is either the Riemannian or the inverse of the Riemannian (I am not totally sure which one).
-Notice that the bandwidth matrix is different for each point.
-This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.
-<!-- There is a paper by Terrel and Scott from Annals of Statistics in 1992 where they talk about this (it is called "variable kernel density estimation").
-Some other search terms to find relevant papers might be "adaptive" or "varying bandwidth".
-I don't know if anyone has had the idea of bringing all these ideas together with manifold learning and the estimate of the Riemannian - we should do a thorough literature search.
-However, if no one has already done this I think it is a very interesting way to do anomaly detection for very high dimensional data.
--->
+## Riemannian metric as variable bandwidth
 
@@
 
-<!-- - Manifold learning distortion -->
-
-<!-- 	dimension reduction using manifold learning -->
-<!-- 	Riemannian matrix to measure topological distortion -->
-<!-- 	distortion is informative for true embedding -->
-
-The aim of the project is not to evaluate different manifold learning algorithms.
-Although ISOMAP worked OK here it is still distorted and in real examples you will never 'know' that the data are uniform on the sphere.
-Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
-
-### Metric Learning algorithm
-
-@Perrault-Joncas2013-pq propose the Metric Learning algorithm which mainly involves four main steps.
+The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
+Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
+The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
+@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
+We restate these four steps in Algorithm \@ref(alg:learnmetric).
 
@@
 
-Learn metric algorithm
-
 \begin{algorithm}[!htb]
-  \caption{Learn metric algorithm}
+  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
   \label{alg:learnmetric}
@@
   \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $x_i \in \pmb{R}^p$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $y_i \in \pmb{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
+  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
   \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
@@
 
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^p$;
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^s$;
 
@@
 
-  \STATE Embed each data point $x\in \mathbb{R}^p$ to embedding coordinates $y(x)=(y^1(x),\dots,y^d(x))$ by any existing manifold learning algorithm;
+  \STATE Embed each data point $x_i\in \mathcal{R}^s$ to embedding coordinates $\pmb{y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm;
 
-  \STATE Obtain the matrix $\tilde{h}(x)$ at each point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates $y$ with each element being
+  \STATE Obtain the matrix $\pmb{\tilde{h}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{y}$ with each element vector in $\pmb{\tilde{h}}$ being
   $$
-    \tilde{h}^{i j}=\frac{1}{2}\left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y_i \cdot y_j\right)-y_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} y_j\right)-y_j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} y_i\right)\right];
+    \pmb{\tilde{h}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
   $$
+  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
 
-  \STATE Calculate the Riemannian metric $h(x)$ as the rank $d$ pseudo inverse of $\tilde{h}(x)$ with 
+  \STATE Calculate the Riemannian metric $\pmb{h}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{    h}}$ with 
   $$
-    h(x) = U diag{1/(\Lambda[1:d])} U^\prime,
+    \pmb{h} = U diag{1/(\Lambda[1:d])} U^\prime,
   $$
-  where $[U, \Lambda]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+  where $[U, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
 
@@
 
-## Proposed algorithm
+## Algorithm for outlier detection on manifold learning embedding
 
-Now we present our proposed method for anomaly detection based on variable kernel density estimates.
+Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \@ref(alg:vkderm).
+There are mainly four steps involved in the algorithm.
+The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation.
+Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels.
+Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator).
+The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f_N(\pmb{y}_i)$.
 
@@
 
-  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
 
-  \STATE Set the variable bandwidth for each point as $\pmb{H}_i = h_i$;
+  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
 
-  \STATE Compute the kernel density estimates for each point as
-  $$
-  \hat{f}(\pmb{y})=\sum\limits_{i=1}^N (2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{y}_i)\right];
-  $$
+  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \@ref(equ:denestimator);
 
-  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top \textit{n\_outliers} as the outliers.
+  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
 
@@
 
-Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots [HDR plots; @Hyndman1996-lk] which are also computed using kernel density estimate of the embedding.
-<!-- Similar to Figure \@ref(fig:hdr) in Chapter \@ref{mlann}, we can use the HDR plots to do anomaly detection and apply it to very high dimensional data sets. -->
+If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in a highest density region of specified coverage, eg.
+1%, 50%, 99%, >99%.
 
@@
 Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
+According to the density estimates, we could rank the data points and then identify which observations lie in a highest density region of specified coverage, eg.
 1%, 5%, 50%, 99%, >99%.
@@
 
-<!-- Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
--->
-
-<!-- Discuss simulation with non-uniform meta data. -->
-
 In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
@@
 By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our propsed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that varibale bandwidth could improve the density estimation given a good manifold learning embedding.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
 
@@
 We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in differnt ways.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
 
@@
 The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more close to real-life datasets.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+
+<!-- Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
+-->
+
+<!-- Discuss simulation with non-uniform meta data. -->
+<!-- The aim of the project is not to evaluate different manifold learning algorithms.
+Although ISOMAP worked OK here it is still distorted and in real examples you will never 'know' that the data are uniform on the sphere.
+Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+ -->
 

### 2022-02-27 paper/paper.tex
Section 2 theory done

@@
 \maketitle
+\begin{abstract}
+In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
+The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
+By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
+We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
+\end{abstract}
 \begin{keywords}
@@
 \end{keywords}
+
+\newpage
 
@@
 In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
-In this section, we provide justification for the use of Riemannien matrix as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data embedded in 3-D and the second with a 100-dimensional meta data.
-\autoref{application} contains the application to visualize and identify anomalies in the {[}TODO{]} dataset.
+In this section, we provide justification for the use of Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
+\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
 Conclusions and discussions are presented in \autoref{conclusion}.
@@
 
-Manifold learning finds a \(d\)-dimensional representation of data that lie on a manifold \(\mathcal{M}\) embedded in a \(p\)-dimensional ambient space with \(d \ll p\).
-We will denote the original data (or `input' points) as \(x_i\), \(i=1,\dots,N\), where \(x_i\in\mathbb{R}^p\), while the low-dimensional representation (or `output' points) will be denoted as \(y_i\), \(i=1,\dots,N\), where \(y_i\in\mathbb{R}^d\).
-Where two subscripts are used (e.g.~\(x_{ih}\) or \(y_{ih}\)), the second subscript refers to the \(h^{th}\) coordinate or dimension of the data.
+Denote the \(s\)-dimensional input data with observations \(x_i, i = 1, 2, \dots, N\), where \(x_i \in \mathcal{R}^s\).
+The underlying manifold \(\mathcal{M}\) has the true density \(f\).
+Manifold learning takes the input data and finds its \(d\)-dimensional representation of data that lie on the manifold \(\mathcal{M}\), with output data points \(y_i, i = 1, 2, \dots, N\), where \(y_i \in \mathcal{R}^d\) and \(d \ll s\).
+The Riemannian metric of the embedding is denoted as \(\pmb{g_i}, i = 1, 2, \dots, N\).
+In kernel density estimate, the pointwise variable bandwidth matrix \(H_i\) is used to estimate the density of the embedding, \(\hat{f}\).
 
-\hypertarget{two-dimensional-kernel-density-estimation}{%
-\subsection{Two dimensional kernel density estimation}\label{two-dimensional-kernel-density-estimation}}
+\hypertarget{kernel-density-estimator-on-riemannian-manifolds}{%
+\subsection{Kernel density estimator on Riemannian manifolds}\label{kernel-density-estimator-on-riemannian-manifolds}}
 
-For a bivariate random sample \(\pmb X_1, \pmb X_2, \ldots, \pmb X_n\)
-drawn from a density \(f\),
-the kernel density estimate is defined by
+In the case of non-Euclidean sample space, the observations often lie on a manifold with the differentiable structure called the Riemannian manifold.
+Consider a compact Riemannian manifold \((\mathcal{M}, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
+Assume \((\mathcal{M}, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by \(g\) and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold.
+Denote \(A_i, i = 1, 2, \dots, N\), where \(A_i \in \mathcal{R}^d\), as i.i.d.
+random objects on \(\mathcal{M}\) with density \(f\).
+For each \(p\in \mathcal{M}\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
+\begin{equation}
+\label{eq:denriem}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
+\end{equation}
+where \(\theta_p(x)\) is the volume density function on the manifold, \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative function, and \(h\) is the bandwidth.
+\textcite{Pelletier2005-vu} discussed the volume density function \(\theta_p(x)\) defined for \(p\) in a neighborhood of \(x\) on the manifold, with the geodesic normal coordinates at \(x\), \(\theta_p(x)\) is equal to the determinant of the Riemannian metric \(g\) at \(\exp_p^{-1}(x)\).
+The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
+\begin{equation}
+\label{eq:denconsist}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
+\end{equation}
+and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.
+
+However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
+Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
+Therefore, based on the estimator in \eqref{eq:denriem}, we propose to a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
+\begin{equation}
+\label{eq:denestimator}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\end{equation}
+where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
+
+An important goal of manifold learning is to recover the local or global features of the data.
+In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric \(g\).
+This could also be used to fix the geometric distortion in the embedding.
+It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix for each data point.
+Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \eqref{eq:denestimator}.
+In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
+
+\hypertarget{multivariate-kernel-density-estimation}{%
+\subsection{Multivariate kernel density estimation}\label{multivariate-kernel-density-estimation}}
+
+In general, a multivariate variable kernel density estimate of \(f\) at point \(\pmb{x} \in \mathcal{R}^d\) is defined as
 \[
-\hat{f} (\pmb x; \pmb H) = n^{-1}\sum_{i=1}^n K_{\pmb H} ( \pmb x - \pmb{X}_i)
+\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i),
 \]
-where \(\pmb x = (x_1, x_2)^T\) and \(\pmb{X}_i = (X_{i1}, X_{i2})^T, i = 1, 2, \ldots, n\). Here
-\(K(\pmb x)\) is the kernel which is a symmetric probability density function,
-\(\pmb H\) is the bandwidth matrix which is symmetric and positive-definite,\\
-and \(K_{\pmb H}(\pmb x) = |\pmb H|^{-1/2} K( \pmb H^{-1/2} \pmb x)\).
-The choice of \(K\) is not crucial: we take
-\(K(\pmb x) = (2\pi)^{-1} \exp(-\tfrac{1}{2} \pmb x^T \pmb x)\) the standard normal
-throughout.\\
-In contrast, the choice of \(\pmb H\) is crucial in determining the performance of \(\hat f\).
-The most common parameterizations of the bandwidth matrix
-are the diagonal and the
-general or unconstrained which has no restrictions on \(\pmb H\)
-provided that \(\pmb H\) remains positive definite and symmetric, that is
+where \(a_i, i = 1, 2, \dots, N\) is the random sample drawn from a density \(f\), \(\pmb{h}_i\) is a \(d\times d\) symmetric positive definite matrix that is variable at \(i\), and \(K_{\pmb{h}_i}\) is the kernel function
 \[
-\pmb H = \begin{bmatrix}h_1^2 & 0 \\0 & h_2^2 \end{bmatrix}
-\ \mathrm{or} \ 
-\pmb H = \begin{bmatrix}h_1^2 & h_{12} \\ h_{12}  & h_2^2 \end{bmatrix}.
+K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i)=(2\pi)^{-d/2}|\pmb{h}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{a}_i)'\pmb{h}_i^{-1}(\pmb{x}-\pmb{a}_i)\right],
 \]
-This latter parameterization allows kernels to have an arbitrary orientation
-whereas the former only allows kernels which are oriented to the
-co-ordinate axes.
+with \(|\pmb{h}_i|\) being the determinant of matrix \(\pmb{h}_i\).
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\).
+An adaptive bandwidth could be used to control the amount of smoothing on the location \(\pmb{x}\) {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point \(\pmb{a}_i\) {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
+We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
+\textcite{Duong2007-up} present an R package \emph{ks} that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and is not applicable to full bandwidth matrix.
 
-An alternative way to think of kernel estimation is that kernel densities `borrow strength' from nearby points and the bandwidth determines what is ``nearby''.
-If the bandwidth is large then all points are ``nearby'' and we get an overly smooth kernel density estimate.
-The interesting thing about a bandwidth matrix is that it allows for different notions of what is ``nearby'' along different coordinates and even along diagonal directions.
-
-\hypertarget{use-riemannian-matrix-as-variable-bandwidth}{%
-\subsection{Use riemannian matrix as variable bandwidth}\label{use-riemannian-matrix-as-variable-bandwidth}}
-
-The Riemannian estimated using the method of Perrault Joncas and Meila (2009) gives some idea of the distortion of an embedding (or so they claim).
-Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
-The Riemannian gives us an idea of the direction and angle of this stretching.
-The Riemannian is quite a technical concept but an important thing to understand is that the estimate that comes out of Perrault Joncas algorithm is a square matrix.
-
-We saw how points that are far apart in the embedding may not have been so far apart on the original manifold.
-The Riemannian gives us some way of correcting this.
-Similarly a bandwidth matrix in a kernel density estimate is all about determining the ``directions'' in which there should be more or less ``closeness''.
-So the basic idea is to replace the kernel density estimate with
-
-\[
-\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)\\
-K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{x}_i)\right]
-\]
-
-where \(H_i\) is either the Riemannian or the inverse of the Riemannian (I am not totally sure which one).
-Notice that the bandwidth matrix is different for each point.
-This makes it a kernel density estimate with local smoothing, which is quite interesting, but we should take care to understand the properties of such things.\\
+\hypertarget{riemannian-metric-as-variable-bandwidth}{%
+\subsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}
 
@@
 
-The aim of the project is not to evaluate different manifold learning algorithms.
-Although ISOMAP worked OK here it is still distorted and in real examples you will never `know' that the data are uniform on the sphere.
-Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
-
-\hypertarget{metric-learning-algorithm}{%
-\subsubsection{Metric Learning algorithm}\label{metric-learning-algorithm}}
-
-\textcite{Perrault-Joncas2013-pq} propose the Metric Learning algorithm which mainly involves four main steps.
+The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding.
+Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
+The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
+\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
+We restate these four steps in Algorithm \ref{alg:learnmetric}.
 
@@
 
-\hypertarget{proposed-algorithm}{%
-\subsection{Proposed algorithm}\label{proposed-algorithm}}
+\hypertarget{algorithm-for-outlier-detection-on-manifold-learning-embedding}{%
+\subsection{Algorithm for outlier detection on manifold learning embedding}\label{algorithm-for-outlier-detection-on-manifold-learning-embedding}}
 
-Now we present our proposed method for anomaly detection based on variable kernel density estimates.
+Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \ref{alg:vkderm}.
+There are mainly four steps involved in the algorithm.
+The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
+Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels.
+Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}.
+The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).
 
@@
 
-  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
 
-  \STATE Set the variable bandwidth for each point as $\pmb{H}_i = h_i$;
+  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
 
-  \STATE Compute the kernel density estimates for each point as
-  $$
-  \hat{f}(\pmb{y})=\sum\limits_{i=1}^N (2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{y}_i)\right];
-  $$
+  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};
 
-  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top \textit{n\_outliers} as the outliers.
+  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
 
@@
 
-Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots {[}HDR plots; \textcite{Hyndman1996-lk}{]} which are also computed using kernel density estimate of the embedding.
+If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in a highest density region of specified coverage, eg.
+1\%, 50\%, 99\%, \textgreater99\%.
 
@@
 Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in an highest density region of specified coverage, eg.
+According to the density estimates, we could rank the data points and then identify which observations lie in a highest density region of specified coverage, eg.
 1\%, 5\%, 50\%, 99\%, \textgreater99\%.
@@
 The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more close to real-life datasets.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
 

### 2022-02-27 paper/paper.Rmd
Introduction and abstract

@@
   email: Anastasios.Panagiotelis@sydney.edu.au
-abstract: "In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways."
+abstract: "Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection.
+For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation.
+We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities.
+The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates.
+To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
+For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week.
+Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
 keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
@@
 library(tidyverse)
-library(dimRed)
-library(reticulate)
+# library(dimRed)
+# library(reticulate)
 library(here)
-library(viridis)
+# library(viridis)
 library(hdrcde)
-library(igraph)
-library(matrixcalc)
-library(akima)
+# library(igraph)
+# library(matrixcalc)
+# library(akima)
 library(car)
-library(ggforce)
-library(ks)
+# library(ggforce)
+# library(ks)
 library(patchwork)
@@
 library(kableExtra)
-Jmisc::sourceAll(here::here("R/sources"))
+# Jmisc::sourceAll(here::here("R/sources"))
 # set.seed(1234)
@@
 
-<!-- Background (anomaly detection with density) with summarized contribution -->
+Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
+It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
+In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which make it computationally expensive or impossible to detect anomalies.
+To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
 
-<!-- 	Anomaly detection on very high-dimensional data. Applications. Main contribution. -->
+Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular method to calculate the probability density function from a sample dataset.
+KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
+There are extensive research on bandwidth selection, two main categories of which are the cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
+For multivariate data, variable kernel density estimator [VKDE;  @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ef] or the data point [sample smoothing estimator; @Terrell1992-ef].
+However, these density estimators are based on random samples in the Euclidean space.
 
-<!-- Anomaly detection using densities, other methods for anomaly detection. Based manifold learning embeddings, density estimates -->
-
-<!-- Density estimation methods, KDE with distortion, 2-D variable kernel denstiy estimate with Riemannian matrix -->
-
-<!-- - Main contributions: better anomaly detection -->
-
-<!-- 	fix distortion with Riemmannian matrix -->
-
-<!-- 	anomaly detection using density estimates -->
-
-<!-- 	kernel density estimate is not accurate -->
-
-<!-- 	improve density estimation with distortion in the embedding plot -->
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed the a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
+The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
+[@Henry2009-ll] In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from @Pelletier2005-vu is not directly applicable.
+This is when we introduce manifold learning to reduce the input data dimension.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), are applied to get a low-dimensional embedding.
+We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with lowest densities.
 
@@
 These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co].
-To tackle this problem and measure the distortion incurred in manifold learning, @Perrault-Joncas2013-pq propose the Metric Learning algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying the Metric Learning algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
+To tackle this problem and measure the distortion incurred in manifold learning, @Perrault-Joncas2013-pq propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
 
-The Riemannian metric defined at each point of the manifold is used to compute the geometric quantities, including angle, length, and volume, of the low-dimensional manifold embedding in any coordinate system, and be further applied to correct the distortion caused by the manifold learning algorithms.
+The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
+Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
+The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
 In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point. 
-Therefore, if we could replace the bandwidth matrix with the Riemannian metric, we could further get the kernel density estimation of the manifold $\mathcal{M}$.
-This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for anomaly detection.
-
-By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_\mathcal{M}$ is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_\mathcal{M}$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.
-The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $H_i$ with the Riemannian metric $g_i$ for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\mathbb{x})$, for anomaly detection.
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in , we could further get the kernel density estimation of the manifold $\mathcal{M}$.
+This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for outlier visualization.
 
@@
 In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-For a high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get a low-dimensional embedding.
-The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others.
-@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point.
+@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive definite square matrix for each data point.
 To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
@@
 The proposed schematic is shown in Figure \@ref(fig:vkde).
+The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{h}_i$ with the Riemannian metric $\pmb{g}_i$ for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
 
@@
 
-In the case of non-Euclidean sample space, the observations often lie on a manifold with the differentiable structure called the Riemannian manifold. 
 Consider a compact Riemannian manifold $(\mathcal{M}, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
@@
 The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$.
-An adaptive bandwidth could be used to control the amount of smoothing on the location $\pmb{x}$ [balloon estimator; @Terrell1992-ef] or the data point $\pmb{a}_i$ [sample smoothing estimator; @Terrell1992-ef].
 We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
@@
 
-The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
-The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
 @Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
 We restate these four steps in Algorithm \@ref(alg:learnmetric).
+By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_\mathcal{M}$ is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_\mathcal{M}$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.
 

### 2022-02-27 paper/paper.tex
Introduction and abstract

@@
 \begin{abstract}
-In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
+Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection.
+For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation.
+We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities.
+The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates.
+To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
+For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week.
+Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
 \end{abstract}
@@
 
+Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
+It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
+In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which make it computationally expensive or impossible to detect anomalies.
+To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
+
+Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular method to calculate the probability density function from a sample dataset.
+KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
+There are extensive research on bandwidth selection, two main categories of which are the cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
+For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
+However, these density estimators are based on random samples in the Euclidean space.
+
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed the a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
+The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
+\autocite{Henry2009-ll} In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from \textcite{Pelletier2005-vu} is not directly applicable.
+This is when we introduce manifold learning to reduce the input data dimension.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied to get a low-dimensional embedding.
+We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with lowest densities.
+
 In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space.
@@
 These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}.
-To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Metric Learning algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying the Metric Learning algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
+To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
 
-The Riemannian metric defined at each point of the manifold is used to compute the geometric quantities, including angle, length, and volume, of the low-dimensional manifold embedding in any coordinate system, and be further applied to correct the distortion caused by the manifold learning algorithms.
+The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding.
+Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
+The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
 In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
-Therefore, if we could replace the bandwidth matrix with the Riemannian metric, we could further get the kernel density estimation of the manifold \(\mathcal{M}\).
-This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for anomaly detection.
-
-By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_\mathcal{M}\) is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_\mathcal{M}\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
-The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(H_i\) with the Riemannian metric \(g_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\mathbb{x})\), for anomaly detection.
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in , we could further get the kernel density estimation of the manifold \(\mathcal{M}\).
+This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.
 
@@
 In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-For a high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get a low-dimensional embedding.
-The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others.
-\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point.
+\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive definite square matrix for each data point.
 To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
@@
 The proposed schematic is shown in Figure \ref{fig:vkde}.
+The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{h}_i\) with the Riemannian metric \(\pmb{g}_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
 
@@
 
-In the case of non-Euclidean sample space, the observations often lie on a manifold with the differentiable structure called the Riemannian manifold.
 Consider a compact Riemannian manifold \((\mathcal{M}, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
@@
 The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\).
-An adaptive bandwidth could be used to control the amount of smoothing on the location \(\pmb{x}\) {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point \(\pmb{a}_i\) {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
 We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
@@
 
-The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
-The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
 \textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
 We restate these four steps in Algorithm \ref{alg:learnmetric}.
+By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_\mathcal{M}\) is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_\mathcal{M}\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
 

### 2022-02-27 paper/paper.tex
Application done

@@
 \nocover
-\blind
+
+\author{Fan~Cheng, Rob J~Hyndman, Anastasios~Panagiotelis}
+\addresses{\textbf{Fan Cheng}\newline
+Monash University
+\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
+\textbf{Rob J Hyndman}\newline
+Monash University
+\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
+\textbf{Anastasios Panagiotelis}\newline
+University of Sydney
+\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
+}
 
@@
 \makeatletter
- \lfoot{\sf\@date}
+ \lfoot{\sf Cheng, Hyndman, Panagiotelis: \@date}
 \makeatother
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/kde_2densities_raw_electricity}
+
+}
+
+\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.}\label{fig:electricityoutliers}
+\end{figure}
+
+\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
+The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
+As to the anomalies, variable bandwidth tend to capture the unusual electricity demand volume when the usage is very low or high.
+It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time period in ID 2838.
+In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happens in certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/hdr_electricity_hdrbox_3id_7dow}
+
+}
+
+\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsfixed}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/electricity_hdrbox_3id_7dow}
+
+}
+
+\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsvkde}
+\end{figure}
+
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
+Again the distribution of both typical household in the top panel have shown a repeated period-of-the-week usage pattern, with higher usage during meal time on all seven days of the week and slightly higher usage for weekends.
+However, this repeated pattern in a week window is clearly for the typical household ID 1321.
+As for the distributions for outliers, the middle row outliers from variable bandwidth has spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
+The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
+These finding show the difference in finding typical and anomalous households with different bandwidth selection.
+
 \hypertarget{conclusion}{%
@@
 
+\newpage
+
 \printbibliography

### 2022-02-27 paper/paper.Rmd
Application done

@@
 # jelcodes: C55, C65, C80
-blind: true
+blind: false
 cover: false
@@
 
-```{r smartmeter, fig.align = 'center', out.width = "80%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=FALSE}
-knitr::include_graphics("figures/smartmeter.png")
+```{r electricityoutliers, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.", echo = FALSE, eval=T}
+knitr::include_graphics("figures/kde_2densities_raw_electricity.png")
 ```
 
-<!-- 5x2 HDRs plot of vkde and hdr, 5levels, 5 ML methods, kdtrees, eps=1 -->
-<!-- index top 20 outliers -->
-<!-- compare density ranking of VKDE and HDR,  -->
-<!-- scatterplot no colors, table with one row for 5 ML methods -->
+\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
+The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
+As to the anomalies, variable bandwidth tend to capture the unusual electricity demand volume when the usage is very low or high.
+It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time period in ID 2838.
+In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happens in certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
 
-<!-- top 1 typical points + top 2 outlier, HDR -->
-<!-- top 1 typical points + top 2 outlier, VKDE -->
-<!-- raw data plot side by side -->
-<!-- density plot in two figures -->
+```{r electricityhdrsfixed, fig.align = 'center', out.width = "90%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=T}
+knitr::include_graphics("figures/hdr_electricity_hdrbox_3id_7dow.png")
+```
+
+```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=T}
+knitr::include_graphics("figures/electricity_hdrbox_3id_7dow.png")
+```
+
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
+Again the distribution of both typical household in the top panel have shown a repeated period-of-the-week usage pattern, with higher usage during meal time on all seven days of the week and slightly higher usage for weekends.
+However, this repeated pattern in a week window is clearly for the typical household ID 1321.
+As for the distributions for outliers, the middle row outliers from variable bandwidth has spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
+The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
+These finding show the difference in finding typical and anomalous households with different bandwidth selection.
 
@@
 This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
+
+\newpage

### 2022-02-27 paper/paper.Rmd
Typo

@@
   email: Anastasios.Panagiotelis@sydney.edu.au
-abstract: "Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection.
-For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation.
-We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities.
+abstract: "Kernel density estimation based on random samples often involves a crucial bandwidth selection.
+For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation.
+We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities.
 The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates.
-To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
-For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week.
+To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
+For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week.
 Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
@@
 It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which make it computationally expensive or impossible to detect anomalies.
+In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
 To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
 
-Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular method to calculate the probability density function from a sample dataset.
+Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset.
 KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
-There are extensive research on bandwidth selection, two main categories of which are the cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
+There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
 For multivariate data, variable kernel density estimator [VKDE;  @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ef] or the data point [sample smoothing estimator; @Terrell1992-ef].
@@
 
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed the a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
 The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
@@
 For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), are applied to get a low-dimensional embedding.
-We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with lowest densities.
+We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
 
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space.
+In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
 One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
@@
 In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point. 
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in , we could further get the kernel density estimation of the manifold $\mathcal{M}$.
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $\mathcal{M}$.
 This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for outlier visualization.
@@
 In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
-In this section, we provide justification for the use of Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
+\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
 \autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
@@
 To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with lowest density estimates.
+The outliers could then be defined as the points with the lowest density estimates.
 The proposed schematic is shown in Figure \@ref(fig:vkde).
@@
 The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$.
-We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
-@Duong2007-up present an R package *ks* that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and is not applicable to full bandwidth matrix.
+We use the sample smooth estimator for each data point, which makes a kernel density estimate with local smoothing.
+@Duong2007-up present an R package *ks* that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and does not apply to full bandwidth matrix.
 
@@
 The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_p\mathcal{M}$ for every point $p \in \mathcal{M}$. 
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of manifold.
+If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
 The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(\mathcal{M}, g)$.
@@
 This algorithm is also implemented in a Python library *megaman* [@McQueen2016-xz].
-It is designed to apply the manifold learning methods to large-scale data sets, as well as computing the Riemannian metric of the manifold.
+It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
 
@@
 
-If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in a highest density region of specified coverage, eg.
+If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
 1%, 50%, 99%, >99%.
@@
 
-In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
+In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
+For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
 We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
 Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as $d=2$, same as the meta data dimension.
+The embedded dimension is set as $d=2$, the same as the meta data dimension.
 This enables us to compare the manifold learning embedding with the true meta data.
@@
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
 The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represents the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
+The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
 \begin{equation}
@@
 Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
 The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
@@
 Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in a highest density region of specified coverage, eg.
+According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
 1%, 5%, 50%, 99%, >99%.
 For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with lowest densities are highlighted in black with point indexes in blue.
+The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
 From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
 However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwith for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detect true outliers accurately.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
 For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
@@
 It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it is closest to rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are the very high in yellow, leading to a negative correlation.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative correlation.
 The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
@@
 Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot have indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along the curved embedding.
+For t-SNE, the middle plot has indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along with the curved embedding.
 -->
@@
 In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
-For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
+For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with the lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
 In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
@@
 For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
-Also due to the clusters in the embedding, the outliers found in UMAP are also clustered.
+Also due to the clusters in the embedding, the outliers found in UMAP are clustered.
 
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from varaible and fixed bandwidth KDE by \autoref{tab:tpcors}. 
-Again the highest correlations appears from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth is equivalent to the thrid decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less seperated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}. 
+Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
+The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
+As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
 Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being `r round(min(cors), 3)`.
@@
 The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
+It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
 The dark colors also indicate that they are the true outliers because of their low densities.
@@
 In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only application to embedding dimension within three.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
 Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
-For well estimated densities, the true outliers with low true densities will also have low estimated densities, which suggest that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
+For well-estimated densities, the true outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
 This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
@@
 This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
-In KDE with a fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
+In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
 By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
@@
 Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps has been clearly stated in the application section of @Cheng2021-ex and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
+The data processing steps have been clearly stated in the application section of @Cheng2021-ex and they are skipped here.
+Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
 However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
@@
 The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
-As to the anomalies, variable bandwidth tend to capture the unusual electricity demand volume when the usage is very low or high.
-It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time period in ID 2838.
-In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happens in certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
+It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
+In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
 
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
-Again the distribution of both typical household in the top panel have shown a repeated period-of-the-week usage pattern, with higher usage during meal time on all seven days of the week and slightly higher usage for weekends.
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
+Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
 However, this repeated pattern in a week window is clearly for the typical household ID 1321.
-As for the distributions for outliers, the middle row outliers from variable bandwidth has spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
+As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
 The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
-These finding show the difference in finding typical and anomalous households with different bandwidth selection.
+These findings show the difference in finding typical and anomalous households with different bandwidth selections.
 
@@
 By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
 
@@
 Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
+We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
 Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
@@
 <!-- The aim of the project is not to evaluate different manifold learning algorithms.
-Although ISOMAP worked OK here it is still distorted and in real examples you will never 'know' that the data are uniform on the sphere.
-Instead the idea of this project is to do kernel density estimation in a way that takes distortion into account.
+Although ISOMAP worked OK here it is still distorted and in real examples, you will never 'know' that the data are uniform on the sphere.
+Instead, the idea of this project is to do kernel density estimation in a way that takes distortion into account.
  -->

### 2022-02-27 paper/paper.tex
Typo

@@
 \begin{abstract}
-Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection.
-For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation.
-We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities.
+Kernel density estimation based on random samples often involves a crucial bandwidth selection.
+For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation.
+We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities.
 The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates.
-To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
-For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week.
+To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
+For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week.
 Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
@@
 It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which make it computationally expensive or impossible to detect anomalies.
+In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
 To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
 
-Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular method to calculate the probability density function from a sample dataset.
+Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular methods to calculate the probability density function from a sample dataset.
 KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
-There are extensive research on bandwidth selection, two main categories of which are the cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
+There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
 For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
@@
 
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed the a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
 The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
@@
 For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied to get a low-dimensional embedding.
-We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with lowest densities.
+We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
 
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space.
+In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
 One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
@@
 In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in , we could further get the kernel density estimation of the manifold \(\mathcal{M}\).
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(\mathcal{M}\).
 This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.
@@
 To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with lowest density estimates.
+The outliers could then be defined as the points with the lowest density estimates.
 The proposed schematic is shown in Figure \ref{fig:vkde}.
@@
 The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\).
-We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
-\textcite{Duong2007-up} present an R package \emph{ks} that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and is not applicable to full bandwidth matrix.
+We use the sample smooth estimator for each data point, which makes a kernel density estimate with local smoothing.
+\textcite{Duong2007-up} present an R package \emph{ks} that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and does not apply to full bandwidth matrix.
 
@@
 The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_p\mathcal{M}\) for every point \(p \in \mathcal{M}\).
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of manifold.
+If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
 The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((\mathcal{M}, g)\).
@@
 This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
-It is designed to apply the manifold learning methods to large-scale data sets, as well as computing the Riemannian metric of the manifold.
+It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
 
@@
 
-If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in a highest density region of specified coverage, eg.
+If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
 1\%, 50\%, 99\%, \textgreater99\%.
@@
 
-In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example.
+In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
+For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
 We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
 Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as \(d=2\), same as the meta data dimension.
+The embedded dimension is set as \(d=2\), the same as the meta data dimension.
 This enables us to compare the manifold learning embedding with the true meta data.
@@
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
 The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represents the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
+The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
 \begin{equation}
@@
 Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
 The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
@@
 Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in a highest density region of specified coverage, eg.
+According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
 1\%, 5\%, 50\%, 99\%, \textgreater99\%.
 For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with lowest densities are highlighted in black with point indexes in blue.
+The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
 From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
 However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwith for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detect true outliers accurately.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
 For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from varaible and fixed bandwidth KDE by \autoref{tab:tpcors}.
-Again the highest correlations appears from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth is equivalent to the thrid decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less seperated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
+Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
+The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
+As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
 Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259.
@@
 The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
+It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
 The dark colors also indicate that they are the true outliers because of their low densities.
@@
 Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps has been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
+The data processing steps have been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
+Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
 However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.

### 2022-03-21 paper/paper.Rmd
Definitions for differentiable manifolds and tangent space

@@
 
-# Variable kernel density estimation with manifold embeddings {#vkde}
+# Notions about Riemannian geometry {#riemgeo}
+
+In this section, we present some notions about the Riemannian geometry used in this paper.
+
+## Differentiable manifolds
+
+In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions. 
+A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
+A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
+Such a homeomorphism $\varphi: U \rightarrow V$ and $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
+A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
+The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
+$$
+\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
+$$
+are differentiable of class $C^\infty$ (smooth).
+
+## Tangent vector and tangent space
+
+The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through the point $p$ or as the directional derivatives at $p$.
+Here we define the tangent vector via the velocity of curves.
+
+For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e.
+$\gamma_1(0) = \gamma_2(0) = p$.
+$\gamma_1$ and $\gamma_2$ are equivalent if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
+$$
+(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
+$$
+A *tangent vector* to $M$ at $p$ is any equivalent class of the differentiable curves initialized at $p$. 
+The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
+The tangent space is vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the charts $\varphi$ locally at $p$.
+The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
+Tangent vectors can also be seen as the directionnal derivatives at $p$.
+For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space $T_pM$.
+
+The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the *cotangent space* with corresponding *cotangent vector* $z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
+
+## Pushforward and pullback metric
+
+embedding pushforward metric [@Perrault-Joncas2013-pq]
+
+## Riemannian metrics
+
+Let $(\mathcal{M}, g)$ be a $d$-dimensional Riemannian manifold without boundary.
+Denote $d_g$ the Riemannian distance.
+Assume that $(\mathcal{M}, g)$ is complete, i.e.
+$(\mathcal{M}, d_g)$ is a complete metric space, and that the injectivity radius of $(\mathcal{M}, g)$, $inj_g\mathcal{M}$, is strictly positive.
+
+Denote $B_s(p)$ the normal ball in $(\mathcal{M}, g)$ centered at point $p$ with raidus $s$.
+
+$d_g$ is induced by $g$ and by the injectivity radius of $(\mathcal{M}, g)$.
+
+normal ball
+
+A geodesic is a smooth curve that locally joins their points along the shortest path. All the geodesics passing through p are called radial geodesics.
+
+## Exponential map and logarithmic map
+
+## Canonical metric
+
+For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional.
+
+## Isometry
+
+## Laplace-Beltrami operator
+
+## Volumn density function
+
+# Kernel density estimation on manifolds {#vkde}
 
@@
 
-## Algorithm for outlier detection on manifold learning embedding
+# Outlier detection on manifold learning embedding
 
@@
 
-# Simulation {#simulation}
+# Simulations {#simulation}
 

### 2022-03-21 paper/paper.tex
Definitions for differentiable manifolds and tangent space

@@
 
+\hypertarget{riemgeo}{%
+\section{Notions about Riemannian geometry}\label{riemgeo}}
+
+In this section, we present some notions about the Riemannian geometry used in this paper.
+
+A geodesic is a smooth
+curve that locally joins their points along the shortest path. All
+the geodesics passing through p are called radial geodesics. The
+local coordinates defined by the chart ðU ; exp1 Þ are called p
+(Riemannian) normal coordinates with center p.~Note that normal coordinates preserve the distances on radial geode- sics.
+For example, unfolding a sphere onto a plane in normal coordinates can preserve the distances on great circles, like paring an orange.
+Note that for a manifold with complex topology, for example, a torus, multiple normal coordinate charts, rather than one single chart, are needed to cover the whole manifold.
+
+\hypertarget{differentiable-manifolds}{%
+\subsection{Differentiable manifolds}\label{differentiable-manifolds}}
+
+In topology, a \emph{homeomorphism} is a bijective map between two topological spaces that is continuous in both directions.
+A \emph{Hausdorff space} is a topological space where any two distinct points can be separated by disjoint neighborhoods.
+A \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
+Such a homeomorphism \(\varphi: U \rightarrow V\) and \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
+A collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
+The manifold \(M\) is a \emph{differentiable manifold} if there exists an atlas of \(M\), \(\{U_\alpha, \varphi_\alpha\}\), such that the \emph{transition maps} between any two charts,
+\[
+\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
+\]
+are differentiable of class \(C^\infty\) (smooth).
+
+\hypertarget{tangent-vector-and-tangent-space}{%
+\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}}
+
+The tangent vector at point \(p\) can be intuitively viewed as the velocity of a curve passing through the point \(p\) or as the directional derivatives at \(p\).
+Here we define the tangent vector via the velocity of curves.
+
+For any \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
+\(\gamma_1\) and \(\gamma_2\) are equivalent if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
+\[
+(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
+\]
+A \emph{tangent vector} to \(M\) at \(p\) is any equivalent class of the differentiable curves initialized at \(p\).
+The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\).
+The tangent space is vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the charts \(\varphi\) locally at \(p\).
+The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
+Tangent vectors can also be seen as the directionnal derivatives at \(p\).
+For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a basis of the tangent space \(T_pM\).
+
+The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the cotangent space with corresponding cotangent vector \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
+
+Define the directional derivative \(D_v\) in the direction of \(v\) as
+\[
+\forall f \in C^\infty(M): D_v(f)
+\]
+
+\hypertarget{riemannian-metrics}{%
+\subsection{Riemannian metrics}\label{riemannian-metrics}}
+
+Let \((\mathcal{M}, g)\) be a \(d\)-dimensional Riemannian manifold without boundary.
+Denote \(d_g\) the Riemannian distance.
+Assume that \((\mathcal{M}, g)\) is complete, i.e.~\((\mathcal{M}, d_g)\) is a complete metric space, and that the injectivity radius of \((\mathcal{M}, g)\), \(inj_g\mathcal{M}\), is strictly positive.
+
+Denote \(B_s(p)\) the normal ball in \((\mathcal{M}, g)\) centered at point \(p\) with raidus \(s\).
+
+\(d_g\) is induced by \(g\) and by the injectivity radius of \((\mathcal{M}, g)\).
+
+normal ball
+
+\hypertarget{exponential-map-and-logarithmic-map}{%
+\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}}
+
+\hypertarget{pushforward-and-pullback-metric}{%
+\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
+
+embedding pushforward metric \autocite{Perrault-Joncas2013-pq}
+
+\hypertarget{canonical-metric}{%
+\subsection{Canonical metric}\label{canonical-metric}}
+
+For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional.
+
+\hypertarget{isometry}{%
+\subsection{Isometry}\label{isometry}}
+
+\hypertarget{laplace-beltrami-operator}{%
+\subsection{Laplace-Beltrami operator}\label{laplace-beltrami-operator}}
+
+\hypertarget{volumn-density-function}{%
+\subsection{Volumn density function}\label{volumn-density-function}}
+
 \hypertarget{vkde}{%
-\section{Variable kernel density estimation with manifold embeddings}\label{vkde}}
+\section{Kernel density estimation on manifolds}\label{vkde}}
 
@@
 
-\hypertarget{algorithm-for-outlier-detection-on-manifold-learning-embedding}{%
-\subsection{Algorithm for outlier detection on manifold learning embedding}\label{algorithm-for-outlier-detection-on-manifold-learning-embedding}}
+\hypertarget{outlier-detection-on-manifold-learning-embedding}{%
+\section{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}
 
@@
 \hypertarget{simulation}{%
-\section{Simulation}\label{simulation}}
+\section{Simulations}\label{simulation}}
 

### 2022-03-21 R/Simulation2d_swissroll_twinpeak.R
Update rank correlation and plot

@@ -537,21 +537,30 @@ methods <- c("isomap", "lle", "le", "tsne", "umap")
 
 ## scatterplot to compare f_xy for ISOMAP
 f <- tibble(fxy = fxy, # true densities
-            fxy_vkde = p_isomap$densities, fxy_hdr = p_hdr_isomap$densities, # estimated densities with ISOMAP, best
-            fxy_vkde_tsne = p_tsne$densities, fxy_hdr_tsne = p_hdr_tsne$densities, # estimated densities with tSNE, worst
-            )
+            fxy_vkde = p_isomap$densities, fxy_hdr = p_hdr_isomap$densities#, # estimated densities with ISOMAP, best
+            # fxy_vkde_tsne = p_tsne$densities, fxy_hdr_tsne = p_hdr_tsne$densities, # estimated densities with tSNE, worst
+            ) %>% summarise_all(rank) # Comment this part if plotting the density estimates instead of ranks
 f
 cor(f$fxy_vkde, f$fxy)
 cor(f$fxy_hdr, f$fxy)
+cor(f$fxy_vkde, f$fxy, method = "spearman")
+cor(f$fxy_hdr, f$fxy, method = "spearman")
+
+# # Plot ranks instead of densities
+# uncomment when plotting ranks instead of densities
+f <- f %>% summarise_all(rank)
+# cor(af$fxy_vkde, af$fxy, method = "spearman")
+# cor(af$fxy_hdr, af$fxy, method = "spearman")
+
 pf_vkde <- f %>% 
   ggplot(aes(x = fxy, y = fxy_vkde, col = factor(preswissroll$label), shape = factor(preswissroll$label))) + 
   geom_point() + 
-  labs(x = "", y = "", color = "Kernels", shape = "Kernels", title = paste("Variable bandwidth correlation", round(cor(f$fxy_vkde, f$fxy), 3)) ) +
+  labs(x = "", y = "", color = "Kernels", shape = "Kernels", title = paste("Variable bandwidth correlation", round(cor(f$fxy_vkde, f$fxy, method = "spearman"), 3)) ) +
   scale_y_continuous(n.breaks = 6)
 pf_hdr <- f %>% 
   ggplot(aes(x = fxy, y = fxy_hdr, col = factor(preswissroll$label), shape = factor(preswissroll$label))) + 
   geom_point() + 
[…]

### 2022-03-21 R/app.R

@@ -30,7 +30,7 @@ ml_outlier <- function(x, s = 2, k = min(10, nrow(x)), radius = 0,
                        invert.h = TRUE,
                        ell.no = 10, ell.size = 1, gridsize = 20, noutliers = 10,
                        riem.scale = 1, # scaling Riemmanien matrix
-                       prob = c(1, 50, 99), # currently not an input of the app
+                       prob = c(1, 5, 50, 99), # currently not an input of the app
                        ...) {
   
   if (is.null(x) | is.null(method)) return(NULL)

### 2022-03-22 paper/paper.Rmd
Riemannian metric, geodesics, exp and log maps

@@
 
+Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$.
+Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism.
+$\varphi(N)$ is called an embedded submanifold of $M$ with the subspace topology.
+
 ## Tangent vector and tangent space
@@
 $\gamma_1(0) = \gamma_2(0) = p$.
-$\gamma_1$ and $\gamma_2$ are equivalent if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
+$\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
 $$
@@
 The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
-Tangent vectors can also be seen as the directionnal derivatives at $p$.
+Tangent vectors can also be seen as the directional derivatives at $p$.
 For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space $T_pM$.
@@
 
-embedding pushforward metric [@Perrault-Joncas2013-pq]
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
+Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
+The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
+$$
+<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+$$
+where $df_p^{-1}$ is the Jacobian of $f^{-1}$. 
+The tangent vectors  $f_*u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
+Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
 
-## Riemannian metrics
+<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->
 
-Let $(\mathcal{M}, g)$ be a $d$-dimensional Riemannian manifold without boundary.
-Denote $d_g$ the Riemannian distance.
-Assume that $(\mathcal{M}, g)$ is complete, i.e.
-$(\mathcal{M}, d_g)$ is a complete metric space, and that the injectivity radius of $(\mathcal{M}, g)$, $inj_g\mathcal{M}$, is strictly positive.
+## Riemannian metric and geodesic distance
 
-Denote $B_s(p)$ the normal ball in $(\mathcal{M}, g)$ centered at point $p$ with raidus $s$.
+A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
+Generally, we omit the subscript $p$ and refer to $g$ as the Riemannian metric.
+The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. 
+A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
 
-$d_g$ is induced by $g$ and by the injectivity radius of $(\mathcal{M}, g)$.
+The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
+$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
+For a curve $\gamma: I \rightarrow M$, the length of the curve is 
+$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
+where $\gamma(I) \subset U$. The volume $W \subset U$ is defined as
+$$
+Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
+$$
+which is also called the *Riemannian measure* on $M$.
 
-normal ball
+The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
+the second-order ordinary differential equation is satisfied,
+$$
+\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
+$$
+where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
+$$
+\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
++ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
+$$
+The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \int_0^1 \| \gamma^\prime(t) \| dt$ when the curves are defined over the interval $[0,1]$.
+The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
+For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
+The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
 
-A geodesic is a smooth curve that locally joins their points along the shortest path. All the geodesics passing through p are called radial geodesics.
+<!-- ### Isometry -->
 
@@
 
+Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$.
+Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e.
+*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$.
+The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit. 
+The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$.
+Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
+Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
+
 ## Canonical metric
 
-For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional.
+<!-- For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional. -->
 
-## Isometry
+## Volumn density function
+
+For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
+Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
+$$
+\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+$$
+<!-- $$ -->
+<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
+<!-- $$ -->
+Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
+The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{vol}$ on $T_pM$, and
+$\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
+$$
+d\exp_p^*vol(u) = \theta_p(u)du.
+$$
+
+<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->
 
 ## Laplace-Beltrami operator
-
-## Volumn density function
 

### 2022-03-22 paper/paper.tex
Riemannian metric, geodesics, exp and log maps

@@
 
-A geodesic is a smooth
-curve that locally joins their points along the shortest path. All
-the geodesics passing through p are called radial geodesics. The
-local coordinates defined by the chart ðU ; exp1 Þ are called p
-(Riemannian) normal coordinates with center p.~Note that normal coordinates preserve the distances on radial geode- sics.
-For example, unfolding a sphere onto a plane in normal coordinates can preserve the distances on great circles, like paring an orange.
-Note that for a manifold with complex topology, for example, a torus, multiple normal coordinate charts, rather than one single chart, are needed to cover the whole manifold.
-
 \hypertarget{differentiable-manifolds}{%
@@
 
+Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\).
+Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism.
+\(\varphi(N)\) is called an embedded submanifold of \(M\) with the subspace topology.
+
 \hypertarget{tangent-vector-and-tangent-space}{%
@@
 For any \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
-\(\gamma_1\) and \(\gamma_2\) are equivalent if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
+\(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
 \[
@@
 The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
-Tangent vectors can also be seen as the directionnal derivatives at \(p\).
-For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a basis of the tangent space \(T_pM\).
-
-The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the cotangent space with corresponding cotangent vector \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
-
-Define the directional derivative \(D_v\) in the direction of \(v\) as
-\[
-\forall f \in C^\infty(M): D_v(f)
-\]
-
-\hypertarget{riemannian-metrics}{%
-\subsection{Riemannian metrics}\label{riemannian-metrics}}
-
-Let \((\mathcal{M}, g)\) be a \(d\)-dimensional Riemannian manifold without boundary.
-Denote \(d_g\) the Riemannian distance.
-Assume that \((\mathcal{M}, g)\) is complete, i.e.~\((\mathcal{M}, d_g)\) is a complete metric space, and that the injectivity radius of \((\mathcal{M}, g)\), \(inj_g\mathcal{M}\), is strictly positive.
-
-Denote \(B_s(p)\) the normal ball in \((\mathcal{M}, g)\) centered at point \(p\) with raidus \(s\).
-
-\(d_g\) is induced by \(g\) and by the injectivity radius of \((\mathcal{M}, g)\).
-
-normal ball
+Tangent vectors can also be seen as the directional derivatives at \(p\).
+For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space \(T_pM\).
+
+The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector} \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
+
+\hypertarget{pushforward-and-pullback-metric}{%
+\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
+
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
+Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
+The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
+\[
+<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+\]
+where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
+The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
+Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
+
+\hypertarget{riemannian-metric-and-geodesic-distance}{%
+\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}}
+
+A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is \(d\times d\) symmetric positive definite and varies smoothly at \(p\).
+Generally, we omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
+The inner product for two vectors \(u, v \in T_pM\) is written as \(<u, v>_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
+A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
+
+The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
+\(g\) could also be used to define the line element \(dl^2 = g_{ij}dx_idx_j\) and the volume element \(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the local coordinates of the chart \((U, \varphi)\).
+For a curve \(\gamma: I \rightarrow M\), the length of the curve is
+\[l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}\],
+where \(\gamma(I) \subset U\). The volume \(W \subset U\) is defined as
+\[
+Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
+\]
+which is also called the \emph{Riemannian measure} on \(M\).
+
+The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold.
+A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\),
+the second-order ordinary differential equation is satisfied,
+\[
+\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
+\]
+where \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
+\[
+\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
++ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
+\]
+The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc length functional \(l:\gamma \rightarrow \int_0^1 \| \gamma^\prime(t) \| dt\) when the curves are defined over the interval \([0,1]\).
+The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
+For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the \emph{injectivity radius} at \(p \in M\) is the distance to the cut locus.
+The injectivity radius of \((M,g)\), \(\textit{inj}_gM\) is the infimum of the injectivity radii over all points on the manifold.
 
@@
 
-\hypertarget{pushforward-and-pullback-metric}{%
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
-
-embedding pushforward metric \autocite{Perrault-Joncas2013-pq}
+Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with raidus \(r\).
+Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\).
+Define \(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a differentiable, bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\).
+The exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint after covering the length of \(\|u\|\) along the geodesic in one time unit.
+The inverse of the exponential map is called the \emph{logarithm map}, \(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to \(q\).
+Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
+Then we could interpolate a geodesic \(\gamma\) between two points \(p\) and \(q\) with the exponential map and the logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).
 
@@
 
-For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional.
-
-\hypertarget{isometry}{%
-\subsection{Isometry}\label{isometry}}
+\hypertarget{volumn-density-function}{%
+\subsection{Volumn density function}\label{volumn-density-function}}
+
+For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
+Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
+\[
+\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+\]
+Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
+The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{vol}\) on \(T_pM\), and
+\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
+\[
+d\exp_p^*vol(u) = \theta_p(u)du.
+\]
 
@@
 \subsection{Laplace-Beltrami operator}\label{laplace-beltrami-operator}}
-
-\hypertarget{volumn-density-function}{%
-\subsection{Volumn density function}\label{volumn-density-function}}
 

### 2022-03-29 paper/paper.tex
Corrections

@@
 A \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
-Such a homeomorphism \(\varphi: U \rightarrow V\) and \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
+Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
 A collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
@@
 Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\).
-Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism.
-\(\varphi(N)\) is called an embedded submanifold of \(M\) with the subspace topology.
+Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
 
@@
 \]
-A \emph{tangent vector} to \(M\) at \(p\) is any equivalent class of the differentiable curves initialized at \(p\).
+A \emph{tangent vector} to a manifold \(M\) at point \(p\), \(v_p\), is any equivalent class of the differentiable curves initialized at \(p\).
 The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\).
-The tangent space is vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the charts \(\varphi\) locally at \(p\).
+The tangent space is a vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the chart \(\varphi\) locally at \(p\).
 The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
 Tangent vectors can also be seen as the directional derivatives at \(p\).
-For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space \(T_pM\).
+For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space.
 
 The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector} \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
-
-\hypertarget{pushforward-and-pullback-metric}{%
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
-The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
-\[
-<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-\]
-where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
-The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
-Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
 
@@
 \[l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}\],
-where \(\gamma(I) \subset U\). The volume \(W \subset U\) is defined as
+where \(\gamma(I) \subset U\). The volume of \(W \subset U\) is defined as
 \[
@@
 \[
-\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
+\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
 \]
@@
 \]
-The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc length functional \(l:\gamma \rightarrow \int_0^1 \| \gamma^\prime(t) \| dt\) when the curves are defined over the interval \([0,1]\).
+The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc length functional \(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\) when the curves are defined over the interval \([0,1]\).
 The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
@@
 The injectivity radius of \((M,g)\), \(\textit{inj}_gM\) is the infimum of the injectivity radii over all points on the manifold.
+
+\hypertarget{pushforward-and-pullback-metric}{%
+\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
+
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
+Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
+The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
+\[
+<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+\]
+where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
+The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
+Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
 

### 2022-03-29 paper/paper.Rmd
Corrections

@@
 A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
-Such a homeomorphism $\varphi: U \rightarrow V$ and $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
+Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
 A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
@@
 Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$.
-Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism.
-$\varphi(N)$ is called an embedded submanifold of $M$ with the subspace topology.
+Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
 
@@
 $$
-A *tangent vector* to $M$ at $p$ is any equivalent class of the differentiable curves initialized at $p$. 
+A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$. 
 The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
-The tangent space is vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the charts $\varphi$ locally at $p$.
+The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
 The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
 Tangent vectors can also be seen as the directional derivatives at $p$.
-For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space $T_pM$.
+For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.
 
 The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the *cotangent space* with corresponding *cotangent vector* $z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
+
+## Riemannian metric and geodesic distance
+
+A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
+Generally, we omit the subscript $p$ and refer to $g$ as the Riemannian metric.
+The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. 
+A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
+
+The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
+$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
+For a curve $\gamma: I \rightarrow M$, the length of the curve is 
+$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
+where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
+$$
+Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
+$$
+which is also called the *Riemannian measure* on $M$.
+
+The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
+the second-order ordinary differential equation is satisfied,
+$$
+\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
+$$
+where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
+$$
+\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
++ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
+$$
+The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$.
+The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
+For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
+The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
+
+<!-- ### Isometry -->
 
@@
 <!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->                                                                                            
-
-## Riemannian metric and geodesic distance
-
-A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
-Generally, we omit the subscript $p$ and refer to $g$ as the Riemannian metric.
-The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. 
-A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
-
-The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
-$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
-For a curve $\gamma: I \rightarrow M$, the length of the curve is 
-$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
-where $\gamma(I) \subset U$. The volume $W \subset U$ is defined as
-$$
-Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
-$$
-which is also called the *Riemannian measure* on $M$.
-
-The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
-the second-order ordinary differential equation is satisfied,
-$$
-\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
-$$
-where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
-$$
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
-+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
-$$
-The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \int_0^1 \| \gamma^\prime(t) \| dt$ when the curves are defined over the interval $[0,1]$.
-The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
-For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
-The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
-
-<!-- ### Isometry -->
-
 ## Exponential map and logarithmic map

### 2022-04-11 paper/paper.Rmd
Move Riemmanian geometry to appendix

@@
 Conclusions and discussions are presented in \autoref{conclusion}.
-
-# Notions about Riemannian geometry {#riemgeo}
-
-In this section, we present some notions about the Riemannian geometry used in this paper.
-
-## Differentiable manifolds
-
-In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions. 
-A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
-Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
-A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
-The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
-$$
-\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-$$
-are differentiable of class $C^\infty$ (smooth).
-
-Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$.
-Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
-
-## Tangent vector and tangent space
-
-The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through the point $p$ or as the directional derivatives at $p$.
-Here we define the tangent vector via the velocity of curves.
-
-For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e.
-$\gamma_1(0) = \gamma_2(0) = p$.
-$\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
-$$
-(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-$$
-A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$. 
-The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
-The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
-The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
-Tangent vectors can also be seen as the directional derivatives at $p$.
-For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.
-
-The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the *cotangent space* with corresponding *cotangent vector* $z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
-
-## Riemannian metric and geodesic distance
-
-A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
-Generally, we omit the subscript $p$ and refer to $g$ as the Riemannian metric.
-The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. 
-A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
-
-The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
-$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
-For a curve $\gamma: I \rightarrow M$, the length of the curve is 
-$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
-where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
-$$
-Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
-$$
-which is also called the *Riemannian measure* on $M$.
-
-The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
-the second-order ordinary differential equation is satisfied,
-$$
-\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
-$$
-where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
-$$
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
-+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
-$$
-The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$.
-The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
-For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
-The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
-
-<!-- ### Isometry -->
-
-## Pushforward and pullback metric
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
-Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
-The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
-$$
-<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-$$
-where $df_p^{-1}$ is the Jacobian of $f^{-1}$. 
-The tangent vectors  $f_*u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
-Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
-
-<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->                                                                                            
-## Exponential map and logarithmic map
-
-Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$.
-Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e.
-*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$.
-The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit. 
-The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$.
-Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
-Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
-
-## Canonical metric
-
-<!-- For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional. -->
-
-## Volumn density function
-
-For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
-Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
-$$
-\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-$$
-<!-- $$ -->
-<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
-<!-- $$ -->
-Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
-The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{vol}$ on $T_pM$, and
-$\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
-$$
-d\exp_p^*vol(u) = \theta_p(u)du.
-$$
-
-<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->
-
-## Laplace-Beltrami operator
 
@@
 
-\newpage
+\clearpage\appendix
+\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}
+<!-- %\pagenumbering{arabic} -->
+<!-- %\def\thepage{A\arabic{page}} -->
+
+<!-- # Notions about Riemannian geometry {#riemgeo} -->
+
+In this appendix, we present some notions about the Riemannian geometry used in this paper.
+
+## Differentiable manifolds
+
+In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions. 
+A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
+A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
+Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
+A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
+The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
+$$
+\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
+$$
+are differentiable of class $C^\infty$ (smooth).
+
+Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$.
+Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
+
+## Tangent vector and tangent space
+
+The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through the point $p$ or as the directional derivatives at $p$.
+Here we define the tangent vector via the velocity of curves.
+
+For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e.
+$\gamma_1(0) = \gamma_2(0) = p$.
+$\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
+$$
+(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
+$$
+A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$. 
+The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
+The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
+The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
+Tangent vectors can also be seen as the directional derivatives at $p$.
+For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.
+
+The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the *cotangent space* with corresponding *cotangent vector* $z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
+
+## Riemannian metric and geodesic distance
+
+A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
+Generally, we omit the subscript $p$ and refer to $g$ as the Riemannian metric.
+The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. 
+A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
+
+The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
+$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
+For a curve $\gamma: I \rightarrow M$, the length of the curve is 
+$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
+where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
+$$
+Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
+$$
+which is also called the *Riemannian measure* on $M$.
+
+The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
+the second-order ordinary differential equation is satisfied,
+$$
+\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
+$$
+where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
+$$
+\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
++ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
+$$
+The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$.
+The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
+For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
+The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
+
+<!-- ### Isometry -->
+
+## Pushforward and pullback metric
+
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
+Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
+The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
+$$
+<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+$$
+where $df_p^{-1}$ is the Jacobian of $f^{-1}$. 
+The tangent vectors  $f_*u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
+Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
+
+<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->                                                                                            
+## Exponential map and logarithmic map
+
+Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$.
+Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e.
+*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$.
+The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit. 
+The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$.
+Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
+Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
+
+## Canonical metric
+
+<!-- For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional. -->
+
+## Volumn density function
+
+For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
+Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
+$$
+\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+$$
+<!-- $$ -->
+<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
+<!-- $$ -->
+Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
+The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{vol}$ on $T_pM$, and
+$\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
+$$
+d\exp_p^*vol(u) = \theta_p(u)du.
+$$
+
+<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->
+
+## Laplace-Beltrami operator

### 2022-04-11 paper/paper.Rmd
Move KDE to Sec 2.2

@@
 In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point. 
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $\mathcal{M}$.
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $M$.
 This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for outlier visualization.
@@
 In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive definite square matrix for each data point.
+@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
 To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
@@
 The proposed schematic is shown in Figure \@ref(fig:vkde).
-The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{h}_i$ with the Riemannian metric $\pmb{g}_i$ for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
+The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
 
@@
 
-Denote the $s$-dimensional input data with observations $x_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$.
-The underlying manifold $\mathcal{M}$ has the true density $f$.
-Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $\mathcal{M}$, with output data points $y_i, i = 1, 2, \dots, N$, where $y_i \in \mathcal{R}^d$ and $d \ll s$.
-The Riemannian metric of the embedding is denoted as $\pmb{g_i}, i = 1, 2, \dots, N$.
+Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$.
+The underlying manifold $M$ has the true density $f$.
+Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$.
+The Riemannian metric of the embedding is denoted as $g_i, i = 1, 2, \dots, N$.
 In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$.
 
-## Kernel density estimator on Riemannian manifolds
+## Multivariate kernel density estimation
 
-Consider a compact Riemannian manifold $(\mathcal{M}, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
-Assume $(\mathcal{M}, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by $g$ and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold.
+In general, given a multivariate random sample $\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
+\begin{equation}
+\label{eq:kde}
+\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
+\end{equation}
+where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{y})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
+When using the standard gaussian kernel $K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
+\begin{equation}
+\label{eq:kernel}
+K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
+\end{equation}
+with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
+
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{y})$.
+We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
+@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
+Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean.
+In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
+
+## Kernel density estimator on Riemannian manifolds {#riemest}
+
+Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
+Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by $g$ and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold.
 Denote $A_i, i = 1, 2, \dots, N$, where $A_i \in \mathcal{R}^d$, as i.i.d.
-random objects on $\mathcal{M}$ with density $f$.
-For each $p\in \mathcal{M}$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
+random objects on $M$ with density $f$.
+For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
 \begin{equation}
@@
 
-## Multivariate kernel density estimation
-
-In general, a multivariate variable kernel density estimate of $f$ at point $\pmb{x} \in \mathcal{R}^d$ is defined as
-$$
-\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i),
-$$
-where $a_i, i = 1, 2, \dots, N$ is the random sample drawn from a density $f$, $\pmb{h}_i$ is a $d\times d$ symmetric positive definite matrix that is variable at $i$, and $K_{\pmb{h}_i}$ is the kernel function
-$$
-K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i)=(2\pi)^{-d/2}|\pmb{h}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{a}_i)'\pmb{h}_i^{-1}(\pmb{x}-\pmb{a}_i)\right],
-$$
-with $|\pmb{h}_i|$ being the determinant of matrix $\pmb{h}_i$.
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$.
-We use the sample smooth estimator for each data point, which makes a kernel density estimate with local smoothing.
-@Duong2007-up present an R package *ks* that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and does not apply to full bandwidth matrix.
-
 ## Riemannian metric as variable bandwidth
 
-The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_p\mathcal{M}$ for every point $p \in \mathcal{M}$. 
+The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_pM$ for every point $p \in M$. 
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(\mathcal{M}, g)$.
+The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(M, g)$.
 
 To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding. 
-The inner product between two vectors $u,v \in T_p\mathcal{M}$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors $\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
-Therefore, for each point $p\in \mathcal{M}$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
+The inner product between two vectors $u,v \in T_pM$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors $\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
+Therefore, for each point $p\in M$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
 
@@
 
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in \mathcal{M}$ is defined as 
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in M$ is defined as 
 $$
@@
 $$
-where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $\mathcal{M}$. While the volume of $V\subset \mathcal{M}$ is computed by
+where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by
 $$
@@
 By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_\mathcal{M}$ is built.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built.
 The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_\mathcal{M}$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
 The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.

### 2022-04-11 paper/paper.tex
Move Riemmanian geometry to appendix
Move KDE to Sec 2.2

@@
 In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(\mathcal{M}\).
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(M\).
 This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.
@@
 
-\hypertarget{riemgeo}{%
-\section{Notions about Riemannian geometry}\label{riemgeo}}
-
-In this section, we present some notions about the Riemannian geometry used in this paper.
+\hypertarget{vkde}{%
+\section{Kernel density estimation on manifolds}\label{vkde}}
+
+In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
+\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
+To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
+The outliers could then be defined as the points with the lowest density estimates.
+The proposed schematic is shown in Figure \ref{fig:vkde}.
+The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/vkde}
+
+}
+
+\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
+\end{figure}
+
+To start with, we introduce the notations used in this manuscript.
+Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
+Readers familiar with these topics could skip the corresponding subsections.
+Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
+
+\hypertarget{notations}{%
+\subsection{Notations}\label{notations}}
+
+Denote the \(s\)-dimensional input data with observations \(X_i, i = 1, 2, \dots, N\), where \(x_i \in \mathcal{R}^s\).
+The underlying manifold \(M\) has the true density \(f\).
+Manifold learning takes the input data and finds its \(d\)-dimensional representation of data that lie on the manifold \(M\), with output data points \(Y_i, i = 1, 2, \dots, N\), where \(Y_i \in \mathcal{R}^d\) and \(d \ll s\).
+The Riemannian metric of the embedding is denoted as \(g_i, i = 1, 2, \dots, N\).
+In kernel density estimate, the pointwise variable bandwidth matrix \(H_i\) is used to estimate the density of the embedding, \(\hat{f}\).
+
+\hypertarget{multivariate-kernel-density-estimation}{%
+\subsection{Multivariate kernel density estimation}\label{multivariate-kernel-density-estimation}}
+
+In general, given a multivariate random sample \(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
+\begin{equation}
+\label{eq:kde}
+\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
+\end{equation}
+where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
+When using the standard gaussian kernel \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
+\begin{equation}
+\label{eq:kernel}
+K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
+\end{equation}
+with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
+
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{y})\).
+We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
+\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
+Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean.
+In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
+
+\hypertarget{riemest}{%
+\subsection{Kernel density estimator on Riemannian manifolds}\label{riemest}}
+
+Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
+Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by \(g\) and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold.
+Denote \(A_i, i = 1, 2, \dots, N\), where \(A_i \in \mathcal{R}^d\), as i.i.d.
+random objects on \(M\) with density \(f\).
+For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
+\begin{equation}
+\label{eq:denriem}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
+\end{equation}
+where \(\theta_p(x)\) is the volume density function on the manifold, \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative function, and \(h\) is the bandwidth.
+\textcite{Pelletier2005-vu} discussed the volume density function \(\theta_p(x)\) defined for \(p\) in a neighborhood of \(x\) on the manifold, with the geodesic normal coordinates at \(x\), \(\theta_p(x)\) is equal to the determinant of the Riemannian metric \(g\) at \(\exp_p^{-1}(x)\).
+The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
+\begin{equation}
+\label{eq:denconsist}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
+\end{equation}
+and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.
+
+However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
+Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
+Therefore, based on the estimator in \eqref{eq:denriem}, we propose to a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
+\begin{equation}
+\label{eq:denestimator}
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\end{equation}
+where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
+
+An important goal of manifold learning is to recover the local or global features of the data.
+In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric \(g\).
+This could also be used to fix the geometric distortion in the embedding.
+It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix for each data point.
+Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \eqref{eq:denestimator}.
+In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
+
+\hypertarget{riemannian-metric-as-variable-bandwidth}{%
+\subsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}
+
+The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_pM\) for every point \(p \in M\).
+If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
+The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((M, g)\).
+
+To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
+The inner product between two vectors \(u,v \in T_pM\), \(<u,v>_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{<u,v>_g}\) and the angle between two vectors \(\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}\) in the tangent space.
+Therefore, for each point \(p\in M\) in any coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
+
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\in M\) is defined as
+\[
+l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
+\]
+where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function mapping \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
+\[
+Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
+\]
+Both the concepts of distance and volume are relevant to kernel density estimation.
+
+\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
+We restate these four steps in Algorithm \ref{alg:learnmetric}.
+By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_M\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
+
+As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
+This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
+It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
+
+\begin{algorithm}[!htb]
+  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
+  \label{alg:learnmetric}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
+  \OptParameter{ manifold learning parameters }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^s$;
+
+  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
+  $$
+  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
+  $$
+  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
+
+  \STATE Embed each data point $x_i\in \mathcal{R}^s$ to embedding coordinates $\pmb{y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm;
+
+  \STATE Obtain the matrix $\pmb{\tilde{h}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{y}$ with each element vector in $\pmb{\tilde{h}}$ being
+  $$
+    \pmb{\tilde{h}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+  $$
+  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
+
+  \STATE Calculate the Riemannian metric $\pmb{h}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{    h}}$ with 
+  $$
+    \pmb{h} = U diag{1/(\Lambda[1:d])} U^\prime,
+  $$
+  where $[U, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+
+  \end{algorithmic}
+\end{algorithm}
+
+\hypertarget{outlier-detection-on-manifold-learning-embedding}{%
+\section{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}
+
+Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \ref{alg:vkderm}.
+There are mainly four steps involved in the algorithm.
+The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
+Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels.
+Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}.
+The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).
+
+\begin{algorithm}[!htb]
+  \caption{Variable kernel density estimates with Riemannian metric}
+  \label{alg:vkderm}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
+  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
+  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
+  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+
+  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
+
+  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};
+
+  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
+
+  \end{algorithmic}
+\end{algorithm}
+
+If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
+1\%, 50\%, 99\%, \textgreater99\%.
+
+\hypertarget{simulation}{%
+\section{Simulations}\label{simulation}}
+
+In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
+For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
+We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
+Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+The embedded dimension is set as \(d=2\), the same as the meta data dimension.
+This enables us to compare the manifold learning embedding with the true meta data.
+We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
+
+\hypertarget{twodgaussian}{%
+\subsection{3-D mapping from a 2-D Gaussian Mixture Model}\label{twodgaussian}}
+
+We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
+The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
+Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
+\begin{equation}
+\label{eq:gmm}
+P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
+\end{equation}
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
+We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
+Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
+These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_4kernels}
+
+}
+
+\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.
+The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
+The shapes indicate the four kernels.}\label{fig:metadensity}
+\end{figure}
+
+\hypertarget{swiss-roll-mapping}{%
+\subsubsection{Swiss roll mapping}\label{swiss-roll-mapping}}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/mappings_sr_tp}
+
+}
+
+\caption{3-D Mappings of the meta data with colors and shapes indicating the four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
+\end{figure}
+
+Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
+The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
+The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
+\begin{equation}
+\label{eq:swissroll}
+\left\{
+\begin{array}{lcl}
+X = X_1 \cos{X_1}, \\
+Y = X_2, \\
+Z = X_1 \sin{X_1}.
+\end{array}
+\right.
+\end{equation}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
+
+}
+
+\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
+Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
+\end{figure}
+
+Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
+According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
+1\%, 5\%, 50\%, 99\%, \textgreater99\%.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
+From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
+However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
+For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
+
+\begin{table}
+
+\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
+\centering
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
+\midrule
+Variable bandwidth & 0.0696 & \textbf{0.400} & -0.2357 & 0.023 & \textbf{0.0138}\\
+Fixed bandwidth & \textbf{0.2798} & 0.351 & \textbf{0.0141} & \textbf{0.367} & -0.0110\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
+It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative correlation.
+The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
+As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
+This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg.
+ISOMAP, or when the embedding is discontinuous, eg.
+t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
+
+\hypertarget{twin-peaks-mapping}{%
+\subsubsection{Twin peaks mapping}\label{twin-peaks-mapping}}
+
+For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
+\begin{equation}
+\label{eq:twinpeak}
+\left\{ 
+\begin{array}{lcl}
+X = X_1, \\
+Y = X_2, \\
+Z = \sin(\pi X_1) \tanh (3 X_2).
+\end{array}
+\right.
+\end{equation}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak5levels_outliers_comparison_4ml_3cases_riem0_1}
+
+}
+
+\caption{Highest density region plots of four manifold learning embeddings of the twin peak data.
+Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.}\label{fig:tpoutliers}
+\end{figure}
+
+Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
+In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
+For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with the lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
+In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
+Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
+For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
+Also due to the clusters in the embedding, the outliers found in UMAP are clustered.
+
+\begin{table}
+
+\caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
+\centering
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
+\midrule
+Variable bandwidth & \textbf{0.899} & 0.385 & 0.620 & 0.259 & \textbf{0.659}\\
+Fixed bandwidth & 0.626 & \textbf{0.399} & \textbf{0.622} & \textbf{0.663} & 0.653\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
+Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
+The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
+As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
+Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259.
+This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1}
+
+}
+
+\caption{Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Variable bandwidth shows a strong linear positive relationship.}\label{fig:tpisomapden}
+\end{figure}
+
+In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
+The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
+Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+
+\hypertarget{fivedgaussian}{%
+\subsection{100-D mapping from a 5-D semi-hypersphere}\label{fivedgaussian}}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
+
+}
+
+\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
+\end{figure}
+
+As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
+First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
+In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
+The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
+The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
+The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
+The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
+It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
+The dark colors also indicate that they are the true outliers because of their low densities.
+
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
+For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix.
+Matrix \(\pmb{Q}\) satisfies
+\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
+Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
+Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_4ml_riem20}
+
+}
+
+\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
+The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.}\label{fig:fivedisomapden}
+\end{figure}
+
+In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
+Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
+For well-estimated densities, the true outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
+This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
+For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
+This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
+In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
+By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
+
+\begin{table}
+
+\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\centering
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & Laplacian.Eigenmaps & UMAP\\
+\midrule
+Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{0.662} & \textbf{-0.130}\\
+Fixed bandwidth & 0.806 & 0.940 & -0.181 & -0.341\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\hypertarget{application}{%
+\section{Application}\label{application}}
+
+\hypertarget{irish-smart-meter-dataset}{%
+\subsection{Irish smart meter dataset}\label{irish-smart-meter-dataset}}
+
+In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
+The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
+We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
+
+For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
+\textcite{Cheng2021-ex} propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
+\textcite{Cheng2021-ex} use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
+By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
+
+In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
+Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+The data processing steps have been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
+Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
+However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/kde_2densities_raw_electricity}
+
+}
+
+\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.}\label{fig:electricityoutliers}
+\end{figure}
+
+\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
+The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
+As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
+It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
+In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/hdr_electricity_hdrbox_3id_7dow}
+
+}
+
+\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsfixed}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/electricity_hdrbox_3id_7dow}
+
+}
+
+\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsvkde}
+\end{figure}
+
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
+Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
+However, this repeated pattern in a week window is clearly for the typical household ID 1321.
+As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
+The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
+These findings show the difference in finding typical and anomalous households with different bandwidth selections.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
+The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
+By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+
+As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
+We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
+
+There are several open questions to be explored.
+The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-ex}.
+The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
+The scale of distortion in each point in manifold learning could vary a lot.
+If we could smooth across all data points, eg.
+multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
+The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
+The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+
+\hypertarget{acknowledgment}{%
+\section*{Acknowledgment}\label{acknowledgment}}
+\addcontentsline{toc}{section}{Acknowledgment}
+
+This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
+
+\clearpage\appendix
+\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}
+
+In this appendix, we present some notions about the Riemannian geometry used in this paper.
 
@@
 
-\hypertarget{vkde}{%
-\section{Kernel density estimation on manifolds}\label{vkde}}
-
-In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive definite square matrix for each data point.
-To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with the lowest density estimates.
-The proposed schematic is shown in Figure \ref{fig:vkde}.
-The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{h}_i\) with the Riemannian metric \(\pmb{g}_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/vkde}
-
-}
-
-\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
-\end{figure}
-
-To start with, we introduce the notations used in this manuscript.
-Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
-Readers familiar with these topics could skip the corresponding subsections.
-Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
-
-\hypertarget{notations}{%
-\subsection{Notations}\label{notations}}
-
-Denote the \(s\)-dimensional input data with observations \(x_i, i = 1, 2, \dots, N\), where \(x_i \in \mathcal{R}^s\).
-The underlying manifold \(\mathcal{M}\) has the true density \(f\).
-Manifold learning takes the input data and finds its \(d\)-dimensional representation of data that lie on the manifold \(\mathcal{M}\), with output data points \(y_i, i = 1, 2, \dots, N\), where \(y_i \in \mathcal{R}^d\) and \(d \ll s\).
-The Riemannian metric of the embedding is denoted as \(\pmb{g_i}, i = 1, 2, \dots, N\).
-In kernel density estimate, the pointwise variable bandwidth matrix \(H_i\) is used to estimate the density of the embedding, \(\hat{f}\).
-
-\hypertarget{kernel-density-estimator-on-riemannian-manifolds}{%
-\subsection{Kernel density estimator on Riemannian manifolds}\label{kernel-density-estimator-on-riemannian-manifolds}}
-
-Consider a compact Riemannian manifold \((\mathcal{M}, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
-Assume \((\mathcal{M}, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by \(g\) and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold.
-Denote \(A_i, i = 1, 2, \dots, N\), where \(A_i \in \mathcal{R}^d\), as i.i.d.
-random objects on \(\mathcal{M}\) with density \(f\).
-For each \(p\in \mathcal{M}\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
-\begin{equation}
-\label{eq:denriem}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
-\end{equation}
-where \(\theta_p(x)\) is the volume density function on the manifold, \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative function, and \(h\) is the bandwidth.
-\textcite{Pelletier2005-vu} discussed the volume density function \(\theta_p(x)\) defined for \(p\) in a neighborhood of \(x\) on the manifold, with the geodesic normal coordinates at \(x\), \(\theta_p(x)\) is equal to the determinant of the Riemannian metric \(g\) at \(\exp_p^{-1}(x)\).
-The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
-\begin{equation}
-\label{eq:denconsist}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
-\end{equation}
-and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.
-
-However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
-Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
-Therefore, based on the estimator in \eqref{eq:denriem}, we propose to a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
-\begin{equation}
-\label{eq:denestimator}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
-\end{equation}
-where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
-
-An important goal of manifold learning is to recover the local or global features of the data.
-In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric \(g\).
-This could also be used to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix for each data point.
-Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \eqref{eq:denestimator}.
-In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
-
-\hypertarget{multivariate-kernel-density-estimation}{%
-\subsection{Multivariate kernel density estimation}\label{multivariate-kernel-density-estimation}}
-
-In general, a multivariate variable kernel density estimate of \(f\) at point \(\pmb{x} \in \mathcal{R}^d\) is defined as
-\[
-\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i),
-\]
-where \(a_i, i = 1, 2, \dots, N\) is the random sample drawn from a density \(f\), \(\pmb{h}_i\) is a \(d\times d\) symmetric positive definite matrix that is variable at \(i\), and \(K_{\pmb{h}_i}\) is the kernel function
-\[
-K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i)=(2\pi)^{-d/2}|\pmb{h}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{a}_i)'\pmb{h}_i^{-1}(\pmb{x}-\pmb{a}_i)\right],
-\]
-with \(|\pmb{h}_i|\) being the determinant of matrix \(\pmb{h}_i\).
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\).
-We use the sample smooth estimator for each data point, which makes a kernel density estimate with local smoothing.
-\textcite{Duong2007-up} present an R package \emph{ks} that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and does not apply to full bandwidth matrix.
-
-\hypertarget{riemannian-metric-as-variable-bandwidth}{%
-\subsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}
-
-The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_p\mathcal{M}\) for every point \(p \in \mathcal{M}\).
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((\mathcal{M}, g)\).
-
-To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
-The inner product between two vectors \(u,v \in T_p\mathcal{M}\), \(<u,v>_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{<u,v>_g}\) and the angle between two vectors \(\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}\) in the tangent space.
-Therefore, for each point \(p\in \mathcal{M}\) in any coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
-
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\in \mathcal{M}\) is defined as
-\[
-l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-\]
-where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function mapping \([a,b]\) to \(\mathcal{M}\). While the volume of \(V\subset \mathcal{M}\) is computed by
-\[
-Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
-\]
-Both the concepts of distance and volume are relevant to kernel density estimation.
-
-\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
-We restate these four steps in Algorithm \ref{alg:learnmetric}.
-By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_\mathcal{M}\) is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_\mathcal{M}\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
-
-As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
-This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
-It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
-
-\begin{algorithm}[!htb]
-  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
-  \label{alg:learnmetric}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters }
-  \BlankLine
-  \begin{algorithmic}[1]
-
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^s$;
-
-  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
-  $$
-  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
-  $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
-
-  \STATE Embed each data point $x_i\in \mathcal{R}^s$ to embedding coordinates $\pmb{y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm;
-
-  \STATE Obtain the matrix $\pmb{\tilde{h}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{y}$ with each element vector in $\pmb{\tilde{h}}$ being
-  $$
-    \pmb{\tilde{h}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
-  $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{h}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{    h}}$ with 
-  $$
-    \pmb{h} = U diag{1/(\Lambda[1:d])} U^\prime,
-  $$
-  where $[U, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
-
-  \end{algorithmic}
-\end{algorithm}
-
-\hypertarget{outlier-detection-on-manifold-learning-embedding}{%
-\section{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}
-
-Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \ref{alg:vkderm}.
-There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
-Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels.
-Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}.
-The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).
-
-\begin{algorithm}[!htb]
-  \caption{Variable kernel density estimates with Riemannian metric}
-  \label{alg:vkderm}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
-  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
-  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
-  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
-  \BlankLine
-  \begin{algorithmic}[1]
-
-  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-
-  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
-
-  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};
-
-  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
-
-  \end{algorithmic}
-\end{algorithm}
-
-If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
-1\%, 50\%, 99\%, \textgreater99\%.
-
-\hypertarget{simulation}{%
-\section{Simulations}\label{simulation}}
-
-In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
-Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as \(d=2\), the same as the meta data dimension.
-This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
-
-\hypertarget{twodgaussian}{%
-\subsection{3-D mapping from a 2-D Gaussian Mixture Model}\label{twodgaussian}}
-
-We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
-The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
-Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
-\begin{equation}
-\label{eq:gmm}
-P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
-\end{equation}
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_4kernels}
-
-}
-
-\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.
-The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
-The shapes indicate the four kernels.}\label{fig:metadensity}
-\end{figure}
-
-\hypertarget{swiss-roll-mapping}{%
-\subsubsection{Swiss roll mapping}\label{swiss-roll-mapping}}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/mappings_sr_tp}
-
-}
-
-\caption{3-D Mappings of the meta data with colors and shapes indicating the four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
-\end{figure}
-
-Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
-The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
-\begin{equation}
-\label{eq:swissroll}
-\left\{
-\begin{array}{lcl}
-X = X_1 \cos{X_1}, \\
-Y = X_2, \\
-Z = X_1 \sin{X_1}.
-\end{array}
-\right.
-\end{equation}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
-
-}
-
-\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
-\end{figure}
-
-Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1\%, 5\%, 50\%, 99\%, \textgreater99\%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
-
-\begin{table}
-
-\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
-\midrule
-Variable bandwidth & 0.0696 & \textbf{0.400} & -0.2357 & 0.023 & \textbf{0.0138}\\
-Fixed bandwidth & \textbf{0.2798} & 0.351 & \textbf{0.0141} & \textbf{0.367} & -0.0110\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg.
-ISOMAP, or when the embedding is discontinuous, eg.
-t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
-
-\hypertarget{twin-peaks-mapping}{%
-\subsubsection{Twin peaks mapping}\label{twin-peaks-mapping}}
-
-For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
-\begin{equation}
-\label{eq:twinpeak}
-\left\{ 
-\begin{array}{lcl}
-X = X_1, \\
-Y = X_2, \\
-Z = \sin(\pi X_1) \tanh (3 X_2).
-\end{array}
-\right.
-\end{equation}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak5levels_outliers_comparison_4ml_3cases_riem0_1}
-
-}
-
-\caption{Highest density region plots of four manifold learning embeddings of the twin peak data.
-Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.}\label{fig:tpoutliers}
-\end{figure}
-
-Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
-In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
-For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with the lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
-In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
-Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
-For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
-Also due to the clusters in the embedding, the outliers found in UMAP are clustered.
-
-\begin{table}
-
-\caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
-\midrule
-Variable bandwidth & \textbf{0.899} & 0.385 & 0.620 & 0.259 & \textbf{0.659}\\
-Fixed bandwidth & 0.626 & \textbf{0.399} & \textbf{0.622} & \textbf{0.663} & 0.653\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
-Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
-Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259.
-This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1}
-
-}
-
-\caption{Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
-Variable bandwidth shows a strong linear positive relationship.}\label{fig:tpisomapden}
-\end{figure}
-
-In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
-The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
-Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
-
-\hypertarget{fivedgaussian}{%
-\subsection{100-D mapping from a 5-D semi-hypersphere}\label{fivedgaussian}}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
-
-}
-
-\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
-\end{figure}
-
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
-First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
-In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
-The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
-The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
-The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
-The dark colors also indicate that they are the true outliers because of their low densities.
-
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
-For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix.
-Matrix \(\pmb{Q}\) satisfies
-\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
-Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
-Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_4ml_riem20}
-
-}
-
-\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
-The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.}\label{fig:fivedisomapden}
-\end{figure}
-
-In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
-Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
-For well-estimated densities, the true outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
-This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
-For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
-This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
-In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
-By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
-
-\begin{table}
-
-\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & UMAP\\
-\midrule
-Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{0.662} & \textbf{-0.130}\\
-Fixed bandwidth & 0.806 & 0.940 & -0.181 & -0.341\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{application}{%
-\section{Application}\label{application}}
-
-\hypertarget{irish-smart-meter-dataset}{%
-\subsection{Irish smart meter dataset}\label{irish-smart-meter-dataset}}
-
-In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
-The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
-We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
-
-For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
-\textcite{Cheng2021-ex} propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
-\textcite{Cheng2021-ex} use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
-By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
-
-In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps have been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
-However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/kde_2densities_raw_electricity}
-
-}
-
-\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.}\label{fig:electricityoutliers}
-\end{figure}
-
-\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
-As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
-It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
-In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/hdr_electricity_hdrbox_3id_7dow}
-
-}
-
-\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsfixed}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/electricity_hdrbox_3id_7dow}
-
-}
-
-\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsvkde}
-\end{figure}
-
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
-Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-However, this repeated pattern in a week window is clearly for the typical household ID 1321.
-As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
-The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
-These findings show the difference in finding typical and anomalous households with different bandwidth selections.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
-
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
-
-There are several open questions to be explored.
-The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-ex}.
-The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
-The scale of distortion in each point in manifold learning could vary a lot.
-If we could smooth across all data points, eg.
-multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
-The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
-The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
-
-\hypertarget{acknowledgment}{%
-\section*{Acknowledgment}\label{acknowledgment}}
-\addcontentsline{toc}{section}{Acknowledgment}
-
-This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
-
-\newpage
-
 \printbibliography

### 2022-04-24 paper/paper.Rmd
Appendix

@@
 Conclusions and discussions are presented in \autoref{conclusion}. 
+Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
 
@@
 
-## Notations
+<!-- ## Notations -->
 
-Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$.
+<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$.
 The underlying manifold $M$ has the true density $f$.
@@
 In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$.
+-->
 
@@
 Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
-Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by $g$ and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold.
-Denote $A_i, i = 1, 2, \dots, N$, where $A_i \in \mathcal{R}^d$, as i.i.d.
+Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$.
+Denote $A_i, i = 1, 2, \dots, N$, where $\pmb{A}_i \in \mathcal{R}^d$, as i.i.d.
 random objects on $M$ with density $f$.
@@
 \label{eq:denriem}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
 \end{equation}
-where $\theta_p(x)$ is the volume density function on the manifold, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative function, and $h$ is the bandwidth.
-@Pelletier2005-vu discussed the volume density function $\theta_p(x)$ defined for $p$ in a neighborhood of $x$ on the manifold, with the geodesic normal coordinates at $x$, $\theta_p(x)$ is equal to the determinant of the Riemannian metric $g$ at $\exp_p^{-1}(x)$.
+where $h$ is the bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
+@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$.
 The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
@@
 \clearpage\appendix
-\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}
+<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
 <!-- %\pagenumbering{arabic} -->
@@
 
-<!-- # Notions about Riemannian geometry {#riemgeo} -->
+# Appendix: Notions about Riemannian geometry {#riemgeo}
 
@@
 
-## Canonical metric
+<!-- ## Canonical metric -->
 
@@
 Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
-The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{vol}$ on $T_pM$, and
+The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and
 $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
 $$
-d\exp_p^*vol(u) = \theta_p(u)du.
+d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
 $$
@@
 
-## Laplace-Beltrami operator
+<!-- ## Laplace-Beltrami operator -->
+
+#### Example
+
+Manifold: S2
+
+$\theta \sim U(0, \pi/2)$
+
+$X = \cos(\theta)$
+
+$$

### 2022-04-24 paper/paper.tex
Appendix

@@
 Conclusions and discussions are presented in \autoref{conclusion}.
+Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
 
@@
 Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
-
-\hypertarget{notations}{%
-\subsection{Notations}\label{notations}}
-
-Denote the \(s\)-dimensional input data with observations \(X_i, i = 1, 2, \dots, N\), where \(x_i \in \mathcal{R}^s\).
-The underlying manifold \(M\) has the true density \(f\).
-Manifold learning takes the input data and finds its \(d\)-dimensional representation of data that lie on the manifold \(M\), with output data points \(Y_i, i = 1, 2, \dots, N\), where \(Y_i \in \mathcal{R}^d\) and \(d \ll s\).
-The Riemannian metric of the embedding is denoted as \(g_i, i = 1, 2, \dots, N\).
-In kernel density estimate, the pointwise variable bandwidth matrix \(H_i\) is used to estimate the density of the embedding, \(\hat{f}\).
 
@@
 Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
-Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by \(g\) and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold.
-Denote \(A_i, i = 1, 2, \dots, N\), where \(A_i \in \mathcal{R}^d\), as i.i.d.
+Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\).
+Denote \(A_i, i = 1, 2, \dots, N\), where \(\pmb{A}_i \in \mathcal{R}^d\), as i.i.d.
 random objects on \(M\) with density \(f\).
@@
 \label{eq:denriem}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_p(A_i)} K(\frac{d_g(p, A_i)}{h}),
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
 \end{equation}
-where \(\theta_p(x)\) is the volume density function on the manifold, \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative function, and \(h\) is the bandwidth.
-\textcite{Pelletier2005-vu} discussed the volume density function \(\theta_p(x)\) defined for \(p\) in a neighborhood of \(x\) on the manifold, with the geodesic normal coordinates at \(x\), \(\theta_p(x)\) is equal to the determinant of the Riemannian metric \(g\) at \(\exp_p^{-1}(x)\).
+where \(h\) is the bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
+\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q \(\exp_q^{-1}(p)\).
 The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
@@
 \clearpage\appendix
-\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}
+
+\hypertarget{riemgeo}{%
+\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}}
 
@@
 
-\hypertarget{canonical-metric}{%
-\subsection{Canonical metric}\label{canonical-metric}}
-
 \hypertarget{volumn-density-function}{%
@@
 
-\hypertarget{laplace-beltrami-operator}{%
-\subsection{Laplace-Beltrami operator}\label{laplace-beltrami-operator}}
+\hypertarget{example}{%
+\paragraph{Example}\label{example}}
+
+Manifold: S2
+
+\(\theta \sim U(0, \pi/2)\)
+
+\(X = \cos(\theta)\)
+
+\$\$
 

### 2022-04-24 R/Simulation2d_swissroll_twinpeak.R
Example S1

@@ -253,7 +253,7 @@ Rn <- metric_isomap$rmetric # array
 # fxy_isomap <- hdrcde:::interp.2d(fisomap$x, fisomap$y, fisomap$z, x0 = E1, y0 = E2)
 # plot_contour(metric_isomap, gridsize = gridsize, riem.scale = riem.scale) # estimate grid densities with vkde()
 
-opt.method <- c("AMISE", "MEAN", "SCALED")[1]
+opt.method <- c("AMISE", "MEAN", "SCALED")[2]
 riem.scale <- .1
 fisomap <- vkde(x = fn, h = Rn, gridsize = gridsize, eval.points = fn, opt.method = opt.method, riem.scale = riem.scale) # 0.923
 # fisomap <- vkde(x = fn, h = Rn*riem.scale, gridsize = gridsize, eval.points = fn) # 0.64

### 2022-04-24 R/riemgeom_s2.R
Example S1

@@ -1,0 +1,120 @@
+set.seed(1)
+N <- 100
+d <- 2
+theta <- runif(N, 0, pi/2)
+# density
+dentheta <- dunif(theta, 0, pi/2) # density = 1/(MAX-MIN) = 0.6366198
+x <- cos(theta)
+fx <- 0
+# r <- ??
+for(i in 1:N){
+  fi <- 1 / r ^ d / sqrt(1 - x[i] ^ 2) * dmvnorm(x = eval.points, mean = x[k,], sigma = hk)
+  fx <- sum(fx, )
+}
+
+
+# evaluate density on grid points
+fhat <- NULL
+for (k in 1:N) {
+  # hk <- h[,,k]
+  z <-  abind::abind(z, array(mvtnorm::dmvnorm(x = eval.points, mean = x[k,], sigma = hk), dim = gridsize), along = d + 1)  # stack array of dimension (gridsize*gridsize) with abind
+}
+z <- rowMeans(z, dims = 2, na.rm = TRUE)
+
+
+# # x ~ U(0,1), theta = acos(x)
+# par(mfrow=c(1,2))
+# plot(x, sqrt(1-x^2), main = "Cartesian coordinates", ylim = c(0,1))
+# points(x, rep(0, N), col = grey(0.5), type = "p")
+# plot(x, denx, main = "U(0,1) CDF")
+# 
+# # polar coordinates
+# # theta <- acos(x)
+# dentheta <- 1 - cos(theta)
+# plot(cos(theta), sin(theta), main = "Polar coordinates", ylim = c(0,1))
+# points(theta, rep(0, N), col = grey(0.5), type = "p")
+# plot(theta, dentheta, main = "Polar CDF")
+
+
+train <- swissroll
+# Parameters fixed
+x <- train
+N <- nrow(x)
+s <- 2
+k <- 20
+method <- "Isomap"
+annmethod <- "kdtree"
+distance <- "euclidean"
+treetype <- "kd"
+searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
+radius <- 8 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
+
+gridsize <- 20
+noutliers <- 20
+# riem.scale <- .1 # .1 # scale Riemmanian
+
+# ISOMAP
+
+## ---- message=FALSE-------------------------------------------------------------
[…]

### 2022-04-24 R/sources/vkde.R

@@ -39,8 +39,8 @@ vkde <- function(x, h = NULL, gridsize = 20, xmin = apply(x, 2, min), xmax = app
   if (d > 1 & !positive) 
     H <- ks::Hpi(x = x, nstage = 2, binned = ks:::default.bflag(d = d, n = n), deriv.order = 0)
   hidet <- apply(h, 3, det)
-  h <- h * ((de(H) / mean(hidet)))^(1/d)
-  # h <- sweep(h, 3, det(H) / hidet, "*")
+  # h <- h * ((det(H) / mean(hidet)))#^(1/d)
+  h <- sweep(h, 3, det(H) / hidet, "*")
   }
   # Option 3: scale by a input constant
   if(opt.method == "SCALED") h <- h * riem.scale

### 2022-04-27 R/riemgeom_s2.R
Math to be corrected for volume density function

@@ -1,28 +1,32 @@
+rm(list= ls())
+library(tidyverse)
+library(dimRed)
+# library(viridis)
+# library(hdrcde)
+library(ks)
+# library(patchwork)
+# # library(copula)
+# library(plotly)
+# library(kableExtra)
+Jmisc::sourceAll(here::here("R/sources"))
+
 set.seed(1)
-N <- 100
+N <- 10000
 d <- 2
 theta <- runif(N, 0, pi/2)
+# theta <- rnorm(N, pi / 4, pi / 20)
 # density
 dentheta <- dunif(theta, 0, pi/2) # density = 1/(MAX-MIN) = 0.6366198
 x <- cos(theta)
-fx <- 0
-# r <- ??
-for(i in 1:N){
-  fi <- 1 / r ^ d / sqrt(1 - x[i] ^ 2) * dmvnorm(x = eval.points, mean = x[k,], sigma = hk)
-  fx <- sum(fx, )
-}
+y <- sin(theta)
+plot(x, y, asp = 1)
+# range(x) # [0, 1]
+denx <- 2 / pi * (1 / sqrt(1 - x ^ 2))
+plot(x, denx)
+range(denx)
 
 
-# evaluate density on grid points
-fhat <- NULL
-for (k in 1:N) {
-  # hk <- h[,,k]
-  z <-  abind::abind(z, array(mvtnorm::dmvnorm(x = eval.points, mean = x[k,], sigma = hk), dim = gridsize), along = d + 1)  # stack array of dimension (gridsize*gridsize) with abind
-}
-z <- rowMeans(z, dims = 2, na.rm = TRUE)
-
-
-# # x ~ U(0,1), theta = acos(x)
+# # If x ~ U(0,1), theta = acos(x)
 # par(mfrow=c(1,2))
 # plot(x, sqrt(1-x^2), main = "Cartesian coordinates", ylim = c(0,1))
 # points(x, rep(0, N), col = grey(0.5), type = "p")
@@ -36,12 +40,12 @@ z <- rowMeans(z, dims = 2, na.rm = TRUE)
 # plot(theta, dentheta, main = "Polar CDF")
 
 
-train <- swissroll
+train <- data.frame(x = x, y = y)
 # Parameters fixed
-x <- train
-N <- nrow(x)
-s <- 2
-k <- 20
+# x <- train
+N <- nrow(train)
+s <- 1
+k <- 10
 method <- "Isomap"
[…]

### 2022-04-27 paper/paper.Rmd
Add example of S1 in appendix

@@
 
-#### Example
+#### Example: $S^1$
 
-Manifold: S2
+Manifold: $S^1$ with center 0 and radius $r=1$
 
-$\theta \sim U(0, \pi/2)$
+Polar coordinates: $(r, \theta)$
 
-$X = \cos(\theta)$
+Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
 
+Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$
+
+Density of $\theta$: $f(\theta) = 2 / \pi$
+
+Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$
+
+The CDF of X is given by
 $$
+P(X \leq x) = P(\cos(\theta) \leq x) \\
+= P(\theta \geq \arccos(x)) \\
+= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+= 1 - \frac{2}{\pi} \arccos(x).
+$$
+Then the density of X is the derivative of the CDF w.r.t. $x$
+$$
+f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+= \frac{2}{\pi \sqrt{1 - x^2}}.
+$$
+Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ and compare it with its true density $f(x)$.
+
+<!-- we could reduce the dimension to $d = 1$ and estimate the density of the 1-dimensional embedding$. -->
+
+<!-- First, we use the kernel density estimator in \@ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->
+
+We use the density estimator in @Pelletier2005-vu. 
+<!-- \begin{equation} -->
+<!-- \label{eq:pelletier} -->
+$$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
+$$
+<!-- \end{equation} -->
+where $d=2$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
+
+Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$.
+Therefore, the geodesic distance is
+$$
+d_g(p, X_i) = \arccos(p) - \arccos(X_i)
+$$
+As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
+$|g(\exp_p^{-1}(x_i))|$.
+
+[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
+
+The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
+$$
+g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
+$$
+is smooth in a local neightborhood $U$ of $M$.
+
+Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
+$$
+\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
+$$
+Therefore, the estimator for $p \in S^1$ is given by
+$$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
+^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
+$$
+The bandwidth $h$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.
+
+\newpage

### 2022-04-27 paper/paper.tex
Add example of S1 in appendix

@@
 Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
-The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{vol}\) on \(T_pM\), and
+The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and
 \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
 \[
-d\exp_p^*vol(u) = \theta_p(u)du.
-\]
-
-\hypertarget{example}{%
-\paragraph{Example}\label{example}}
-
-Manifold: S2
-
-\(\theta \sim U(0, \pi/2)\)
-
-\(X = \cos(\theta)\)
-
-\$\$
+d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
+\]
+
+\hypertarget{example-s1}{%
+\paragraph{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}
+
+Manifold: \(S^1\) with center 0 and radius \(r=1\)
+
+Polar coordinates: \((r, \theta)\)
+
+Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).
+
+Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)
+
+Density of \(\theta\): \(f(\theta) = 2 / \pi\)
+
+Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)
+
+The CDF of X is given by
+\[
+P(X \leq x) = P(\cos(\theta) \leq x) \\
+= P(\theta \geq \arccos(x)) \\
+= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+= 1 - \frac{2}{\pi} \arccos(x).
+\]
+Then the density of X is the derivative of the CDF w.r.t. \(x\)
+\[
+f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+= \frac{2}{\pi \sqrt{1 - x^2}}.
+\]
+Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\) and compare it with its true density \(f(x)\).
+
+We use the density estimator in \textcite{Pelletier2005-vu}.
+\[
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
+\]
+where \(d=2\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
+
+Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\).
+Therefore, the geodesic distance is
+\[
+d_g(p, X_i) = \arccos(p) - \arccos(X_i)
+\]
+As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).
+
+{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}
+
+The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
+\[
+g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
+\]
+is smooth in a local neightborhood \(U\) of \(M\).
+
+Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
+\[
+\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
+\]
+Therefore, the estimator for \(p \in S^1\) is given by
+\[
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
+^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
+\]
+The bandwidth \(h\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.
+
+\newpage
 

### 2022-05-03 R/riemgeom_s2.R
S1 example done

@@ -1,75 +1,63 @@
 rm(list= ls())
 library(tidyverse)
 library(dimRed)
-# library(viridis)
-# library(hdrcde)
 library(ks)
-# library(patchwork)
-# # library(copula)
-# library(plotly)
-# library(kableExtra)
 Jmisc::sourceAll(here::here("R/sources"))
 
 set.seed(1)
 N <- 10000
-d <- 2
+d <- 1
 theta <- runif(N, 0, pi/2)
-# theta <- rnorm(N, pi / 4, pi / 20)
 # density
 dentheta <- dunif(theta, 0, pi/2) # density = 1/(MAX-MIN) = 0.6366198
+# theta <- rnorm(N, pi / 4, pi / 20)
+# dentheta <- dnorm(theta, pi / 4, pi / 20)
+# denx <- 1 / sqrt(1 - x ^ 2) * dnorm(theta, pi / 4, pi / 20)
 x <- cos(theta)
 y <- sin(theta)
 plot(x, y, asp = 1)
 # range(x) # [0, 1]
+# True density of x
 denx <- 2 / pi * (1 / sqrt(1 - x ^ 2))
 plot(x, denx)
 range(denx)
 
-
+# KDE with fixed bandwidth
 h <- ks::hpi(x, binned = TRUE) # 0.034
 # h <- 0.05
 fxkde <- kde(x, eval.points = x)$estimate
 plot(x, fxkde)
-
+# KDE with Riemannian metric
 d <- 1
 fx <- rep(0, N)
 # eval.points <- x
 for(i in 1:N){
-  # fi <- sqrt(1 - x[i] ^ 2) * dmvnorm(x = eval.points, mean = x[i], sigma = diag(h, N))
-  # fx <- sum(fx, fi)
   fi <- 1 / sqrt(1 - x[i] ^ 2) * dnorm(x = acos(x), mean = acos(x[i]), sd = h)
   fx <- fx + fi
 }
 fx <- fx / N
 
+# Plotting
 par(mfrow=c(2,2))
-plot(x, denx)
-plot(x, fx)
-plot(x, fxkde)
-plot(x, dentheta)
+plot(x, denx, main = "True density of x")
+plot(x, fxkde, main = paste("Estimated f(x) with KDE, h = ", round(h, 3)))
+plot(x, fx, main = "Estimated f(x) with riemannian metric")
+abline(h = dentheta, col = "red", lty = "dashed")
 
[…]

### 2022-05-03 R/sources/metricML.R
S1 example done

@@ -182,7 +182,7 @@ riemann_metric <- function(Y, laplacian, d, invert.h = FALSE){
     Hsvals <- eigen(H[ , ,i])
     Huu <- Hsvals$vectors
     Hvv <- Hsvals$values[1:d] # top d largest eigenvalues, already sorted in decreasing order
-    Hvv1 <- diag(x = 1 / Hvv)
+    Hvv1 <- diag(x = 1 / Hvv, nrow = d)
     H[ , ,i] <- Huu %*% Hvv1 %*% t(Huu)
     H[, , i] <- 0.5 * (H[, , i] + t(H[, , i])) # fix not symmetric issue
   }

### 2022-05-03 paper/paper.Rmd
Add S2 example reference

@@
     latex_engine: pdflatex
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
-It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
-To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
-
-Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset.
-KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
-There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
-For multivariate data, variable kernel density estimator [VKDE;  @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ef] or the data point [sample smoothing estimator; @Terrell1992-ef].
-However, these density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
-[@Henry2009-ll] In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from @Pelletier2005-vu is not directly applicable.
-This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), are applied to get a low-dimensional embedding.
-We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
-
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
-One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co].
-To tackle this problem and measure the distortion incurred in manifold learning, @Perrault-Joncas2013-pq propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-
-The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
-The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
-In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point. 
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $M$.
-This kernel density estimate can then be used to produce the highest density region plots [@Hyndman1996-lk] for outlier visualization.
-
-The rest of the paper is organized as follows.
-In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
-In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
-\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
-Conclusions and discussions are presented in \autoref{conclusion}. 
-Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
+Anomaly detection has been an important and diverse area where anomalies
+or outliers are detected in a given data. It often involves a data
+analysis process to uncover the unusual patterns and has been widely
+applied to machine learning [@Omar2013-lz], network intrusions
+identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging
+[@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security
+[@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which
+explored detecting households with anomalous electricity usage
+distributions instead of raw data. In the case of non-Euclidean sample
+space, the observations lie on a Riemannian manifold embedded in a very
+high-dimensional ambient space, which makes it computationally expensive
+or impossible to detect anomalies. To address this problem, we propose a
+kernel density estimator of the low-dimensional statistical manifold
+embedding and find outliers as the data with the lowest density
+estimates.
+
+Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of
+the most popular methods to calculate the probability density function
+from a sample dataset. KDE is flexible to learn the shape of the
+underlying density from the data controlled by the bandwidth and the
+selection of bandwidth is crucial in KDE [@Jones1990-oe;
+@Terrell1992-ut]. There is extensive research on bandwidth selection,
+two main categories of which are cross-validation [@Jones1992-ta;
+@Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp]. For
+multivariate data, variable kernel density estimator [VKDE;
+@Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive
+bandwidth matrix to control the amount of smoothing on the location of
+the estimated point [balloon estimator; @Terrell1992-ef] or the data
+point [sample smoothing estimator; @Terrell1992-ef]. However, these
+density estimators are based on random samples in the Euclidean space.
+
+For samples points lying on a manifold with the differentiable structure
+called the Riemannian manifold, @Pelletier2005-vu proposed a kernel
+density estimator based on the kernel weight depending on the distance
+between the estimated points and the sample data points. The idea of the
+estimator is to use a strictly positive function of the geodesic
+distance on the manifold and then normalize it with the volume density
+function of the Riemannian manifold for curvature. [@Henry2009-ll] In
+many application scenarios, we tend to find that the sample points on a
+manifold are embedded in a much higher-dimensional space, and the kernel
+density estimator from @Pelletier2005-vu is not directly applicable.
+This is when we introduce manifold learning to reduce the input data
+dimension. For these high-dimensional data set, various manifold
+learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE,
+and UMAP (see details of these algorithms in @Cheng2021-ex), are applied
+to get a low-dimensional embedding. We propose a kernel density
+estimator to be applied to the low-dimensional embedding and define
+outliers as those with the lowest densities.
+
+In manifold learning, the underlying idea is that the data lies on a
+low-dimensional smooth manifold that is embedded in a high-dimensional
+space. One of the fundamental objectives of manifold learning is to
+explore the geometry of the dataset, including the distances between
+points and volumes of regions of data. These intrinsic geometric
+attributes of the data, such as distances, angles, and areas, however,
+can be distorted in the low-dimensional embedding, leading to failure to
+recover the geometry of the manifold [@Goldberg2008-co]. To tackle this
+problem and measure the distortion incurred in manifold learning,
+@Perrault-Joncas2013-pq propose the Learn metric algorithm to augment
+any existing embedding output with geometric information in the
+Riemannian metric of the manifold itself. By applying this algorithm,
+the outputs of different manifold learning methods can be unified and
+compared under the same framework, which would highly benefit in
+improving the effectiveness of the embedding.
+
+The Riemannian metric using the method of @Perrault-Joncas2013-pq gives
+some idea of the distortion of an embedding. Mapping the points through
+a non-linear function "stretches" some regions of space and "shrinks"
+others. The Riemannian gives us an idea of the direction and angle of
+this stretching, which is informative for learning the manifold. In
+variable kernel density estimate, the bandwidth matrix $H$ is also
+defined to control the amount of smoothing for each data point.
+Therefore, we propose to the use Riemannian metric as the bandwidth
+matrix in, we could further get the kernel density estimation of the
+manifold $M$. This kernel density estimate can then be used to produce
+the highest density region plots [@Hyndman1996-lk] for outlier
+visualization.
+
+The rest of the paper is organized as follows. In \autoref{vkde}, we
+present the proposed algorithm to detect anomalies based on variable
+kernel density estimates of manifold embeddings. In this section, we
+provide justification for the use of the Riemannian metric as the
+bandwidth of variable kernel density estimation, including the
+comparison with fixed bandwidth. \autoref{simulation} is composed of two
+simulations with the proposed algorithm; the first deals with
+2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data
+and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+\autoref{application} contains the application to visualize and identify
+anomalies in the Irish smart meter dataset. Conclusions and discussions
+are presented in \autoref{conclusion}. Readers interested in the the
+notions of Riemannian geometry mentioned in this paper could use the
+Appendix in \autoref{riemgeo} as a reference.
 
@@
 
-In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
-To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with the lowest density estimates.
-The proposed schematic is shown in Figure \@ref(fig:vkde).
-The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
-
-(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
+In this section, we introduce the proposed method to detect anomalies
+based on the kernel density estimates of manifold learning embeddings
+where the Riemannian matrix is used as the pointwise variable bandwidth
+to measure the direction and angle of the distortion of the
+low-dimensional embeddings. @Perrault-Joncas2013-pq gives us an idea of
+how to measure the direction and angle of the distortion using the
+Riemannian metric which is a positive definite square matrix for each
+data point. To learn the distribution of the low-dimensional embedding,
+we use the kernel density estimation with the bandwidth matrix being the
+Riemannian metric. The outliers could then be defined as the points with
+the lowest density estimates. The proposed schematic is shown in Figure
+\@ref(fig:vkde). The highlighted two steps in Figure \@ref(fig:vkde) are
+the main contributions of this main chapter, replacing the bandwidth
+matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
+variable kernel density estimate, and computing the highest density
+region plots based on the density estimates, $\hat{f}(\pmb{x})$, for
+anomaly detection.
+
+(ref:vkdecaption) The proposed schematic for variable kernel density
+estimation with recovered geometry.
+
 ```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
@@
 
-To start with, we introduce the notations used in this manuscript.
-Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
-Readers familiar with these topics could skip the corresponding subsections.
-Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
+To start with, we introduce the notations used in this manuscript. Then
+we introduce the multivariate kernel density estimation method with
+variable bandwidth matrix and the metric learning algorithm to derive
+the pointwise Riemannian metric. Readers familiar with these topics
+could skip the corresponding subsections. Finally, we propose our novel
+algorithm to estimate densities and detect anomalies for
+high-dimensional data set.
 
@@
 
-In general, given a multivariate random sample $\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
+In general, given a multivariate random sample
+$\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$,
+$i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel
+density estimate of $f$ at point
+$\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
 \begin{equation}
@@
 \hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
-\end{equation}
-where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{y})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
-When using the standard gaussian kernel $K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
+\end{equation} where $\pmb{H}_i$ is called the bandwidth matrix which is
+a $d\times d$ symmetric positive definite matrix variable at each $i$,
+$K(\pmb{y})$ is the kernel function, and
+$K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
+When using the standard gaussian kernel
+$K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
 \begin{equation}
@@
 K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
-\end{equation}
-with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
-
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{y})$.
-We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
-@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
-Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
+\end{equation} with $|\pmb{H}_i|$ being the determinant of matrix
+$\pmb{H}_i$.
+
+The bandwidth matrix is important in controlling the smoothing across
+all data points and the shape is determined by the kernel function
+$K(\pmb{y})$. We consider the sample smooth estimator for each data
+point, which gives a kernel density estimate with local smoothing.
+@Duong2007-up implements the kernel density estimates for data from 1 to
+6 dimensions in the R package *ks*. However, for variable kernel density
+estimation where the bandwidth is no longer a constant matrix, the
+implementation is limited to 2-dimension and the bandwidth can either be
+a vector or a diagonal matrix instead of a full bandwidth matrix.
+Further, the estimator in Equation \@ref(eq:kde) is only applicable when
+the sample space is Euclidean. In the cases of a Riemannian manifold,
+the sample space used for kernel density estimation is no longer
+Euclidean and we need to use the estimator described in
+\autoref{riemest}.
 
@@
 
-Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
-Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$.
-Denote $A_i, i = 1, 2, \dots, N$, where $\pmb{A}_i \in \mathcal{R}^d$, as i.i.d.
-random objects on $M$ with density $f$.
-For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
-\begin{equation}
+Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without
+boundary and a probability distribution with density $f$ on the
+manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is
+the Riemannian distance induced by the Riemannian, $g$, and the strictly
+positive injectivity radius [@Chavel2006-mp] of the manifold,
+$inj_g(M)$. Denote $A_i, i = 1, 2, \dots, N$, where
+$\pmb{A}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with
+density $f$. For each $p\in M$, @Pelletier2005-vu proposed the kernel
+density estimator of $f$ to be \begin{equation}
 \label{eq:denriem}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
-\end{equation}
-where $h$ is the bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
-@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$.
-The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
-\begin{equation}
+\end{equation} where $h$ is the bandwidth satisfying $h \leq h_0$ for
+fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$,
+$K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and
+$\theta_{q}(p)$ is the volume density function on the manifold.
+@Pelletier2005-vu mention that the volume density function $\theta_q(p)$
+defined for $p$ in a neighborhood of $q$ on the manifold, with the
+geodesic normal coordinates at $q$, is equal to the determinant of the
+Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$. The
+expression in \@ref(eq:denriem) is also proved to be consistent with the
+kernel density estimators in the Euclidean case as \begin{equation}
 \label{eq:denconsist}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
-\end{equation}
-and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].
-
-However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
-Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
-Therefore, based on the estimator in \@ref(eq:denriem), we propose to a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
-\begin{equation}
+\end{equation} and converges at the same rate as the Euclidean kernel
+density estimator [@Henry2009-ll].
+
+However, in application, we usually found the observations lie on a
+$d$-dimensional manifold embedded in a $s$-dimensional space, where
+$d \ll s$. Then manifold learning can be applies to find the
+$d$-dimensional representation of the manifold $M$. Therefore, based on
+the estimator in \@ref(eq:denriem), we propose to a kernel density
+estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to
+be \begin{equation}
 \label{eq:denestimator}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
-\end{equation}
-where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.
-
-An important goal of manifold learning is to recover the local or global features of the data.
-In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric $g$.
-This could also be used to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for each data point.
-Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \@ref(eq:denestimator).
-In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
+\end{equation} where $h_i, i = 1, 2, \dots, N$ is the pointwise variable
+bandwidth matrix and $h_i$ is positive definite with dimension
+$d\times d$.
+
+An important goal of manifold learning is to recover the local or global
+features of the data. In order to recover the geometry of the manifold,
+@Perrault-Joncas2013-pq propose a method to augment any existing
+reasonable embedding and allow for the computation of geometric values
+to be calculated with an estimation of the Riemannian metric $g$. This
+could also be used to fix the geometric distortion in the embedding. It
+is worth noticing that the Riemannian metric from
+@Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for
+each data point. Therefore, we propose to take the geometric distortion
+into consideration when estimating the kernel densities of the embedding
+with Equation \@ref(eq:denestimator). In this way, we present a
+consistent kernel density estimator with pointwise variable bandwidth as
+the Riemannian metric.
 
@@
 
-The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_pM$ for every point $p \in M$. 
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(M, g)$.
-
-To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding. 
-The inner product between two vectors $u,v \in T_pM$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors $\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
-Therefore, for each point $p\in M$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
-
-[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$
-
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in M$ is defined as 
-$$
+The Riemannian metric $g$ is a symmetric and positive definite tensor
+field which defines an inner product $<,>_g$ on the tangent space $T_pM$
+for every point $p \in M$. If the inner product of the tangent space is
+known for a given geometry, the Riemannian metric is a good measure to
+recover the geometry of the manifold. The Metric Learning algorithm
+[@Perrault-Joncas2013-pq] then augment the embedded manifold with the
+Riemannian metric and produce a Riemannian manifold $(M, g)$.
+
+To recover the original geometry of the manifold, we need to know what
+the inner product corresponds to in the embedding. The inner product
+between two vectors $u,v \in T_pM$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be
+used to define some geometric quantities, such as the vector norm
+$\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors
+$\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
+Therefore, for each point $p\in M$ in any coordinate system, the
+Riemannian metric $g$ is a $d\times d$ symmetric positive definite
+matrix, where $d$ is the dimension of the manifold.
+
+[^1]: Here the Einstein notation is used where superscripts denote
+    summation over $i$ and $j$
+
+The line element and volume element of the full manifold or a subset of
+the manifold can also be computed from $g$. The arc length of a curve
+$c\in M$ is defined as $$
 l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-$$
-where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by
-$$
+$$ where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and
+$c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of
+$V\subset M$ is computed by $$
 Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
-$$
-Both the concepts of distance and volume are relevant to kernel density estimation.
-
-@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
-We restate these four steps in Algorithm \@ref(alg:learnmetric).
-By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.
-
-As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension $s$ is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$ will be returned.
-This algorithm is also implemented in a Python library *megaman* [@McQueen2016-xz].
-It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
-
+$$ Both the concepts of distance and volume are relevant to kernel
+density estimation.
+
+@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly
+involves four steps: weighted neighborhood graph construction; geometric
+graph Laplacian calculation; manifold learning; and Riemannian metric
+calculation. We restate these four steps in Algorithm
+\@ref(alg:learnmetric). By applying an existing manifold learning
+algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a
+low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed. Most
+manifold learning methods involve the construction of the nearest
+neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$
+is built. The Laplacian is quite useful because it can be
+coordinate-free while containing all the important geometry.
+@Perrault-Joncas2013-pq have stated one way to compute the approximated
+$\Delta_M$ with a discrete consistent estimator, the geometric graph
+Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where
+$\varepsilon$ is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the
+Metric Learning algorithm to achieve the augmented embedding with the
+Riemannian metric $(f_n, g_n)$.
+
+As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension
+$s$ is larger than the manifold intrinsic dimension $d$, the rank of the
+embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$
+will be returned. This algorithm is also implemented in a Python library
+*megaman* [@McQueen2016-xz]. It is designed to apply the manifold
+learning methods to large-scale data sets, as well as compute the
+Riemannian metric of the manifold.
+
+```{=tex}
 \begin{algorithm}[!htb]
@@
 \end{algorithm}
-
+```
 # Outlier detection on manifold learning embedding
 
-Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \@ref(alg:vkderm).
-There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation.
-Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels.
-Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator).
-The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f_N(\pmb{y}_i)$.
-
+Now we present our proposed algorithm for anomaly detection based on
+variable kernel density estimates in \@ref(alg:vkderm). There are mainly
+four steps involved in the algorithm. The first step is to apply the
+Learn metric algorithm described in \@ref(alg:learnmetric) to the input
+high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to
+get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$
+and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for
+each observation. Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$
+are given for Gaussian kernels. Then we could use the pointwise
+Riemannian metric to calculate the kernel density estimate with our
+proposed estimator in Equation \@ref(eq:denestimator). The top outliers
+of size $n\_outliers$ are obtained by ordering the embedding points
+$y_i$ according to their density estimates $f_N(\pmb{y}_i)$.
+
+```{=tex}
 \begin{algorithm}[!htb]
@@
 \end{algorithm}
-
-If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
-1%, 50%, 99%, >99%.
+```
+If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could
+be used to identify the relative location of outliers in the embedding
+and find which observations lie in the highest density region of
+specified coverage, eg. 1%, 50%, 99%, \>99%.
 
@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
-Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as $d=2$, the same as the meta data dimension.
-This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
+In this section, we examine two scenarios for both low and high
+dimensions to test our proposed algorithm. For visualization purposes,
+\autoref{twodgaussian} presents a 2-D meta data example. We first
+simulate the data of size $N=2,000$ from a mixture of four Gaussian
+kernels with the same covariance but different means, each consisting of
+$500$ points. Different mapping functions are then applied to the 2-D
+meta data to be mapped in a 3-D feature space, which gives the
+higher-dimensional input for different manifold learning algorithms,
+including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The
+embedded dimension is set as $d=2$, the same as the meta data dimension.
+This enables us to compare the manifold learning embedding with the true
+meta data. We could now apply Algorithm \@ref(alg:vkderm) to get the
+density estimates of all data points and further detect anomalies. As a
+high-dimensional example, the second simulation in
+\autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$
+embedded in a 100-D space and the corresponding embedding dimension is
+$d=5$.
 
@@
 
-We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with different means $\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$.
-The mixture proportions are equally set as $\pi_i=0.25, i=1,2,3,4$.
+We first generate a 2-dimensional data of size $N=2000$ from a Gaussian
+mixture model with four components with different means
+$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$
+and the same variance-covariance matrix
+$\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$. The mixture proportions
+are equally set as $\pi_i=0.25, i=1,2,3,4$.
 <!-- $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$. -->
-Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
-\begin{equation}
+Then the mixture Gaussian mixture density function is a weighted linear
+combination of the four component Gaussian densities as \begin{equation}
 \label{eq:gmm}
 P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
-\end{equation}
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are *true outliers* to be compared with outliers from the kernel density estimates.
+\end{equation} \autoref{fig:metadensity} shows the 2-dimensional meta
+data and the colors indicate the true density of all data points
+calculated from \@ref(eq:gmm), with brighter colors showing higher
+densities and darker colors showing lower densities. We then define
+outliers as points with lowest densities shown in black and typical
+points with highest densities shown in yellow. Based on the true density
+plot, the outliers are scattered in the middle and the outer area of the
+whole structure, while typical points are near the means of four
+kernels. These are *true outliers* to be compared with outliers from the
+kernel density estimates.
 
@@
 
-Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
-The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
-\begin{equation}
+Given the 2-D meta data, multiple mapping functions could be applied to
+embed the data in a 3-D space. One of the most famous examples in
+manifold learning is the swiss roll data, with the mapping function in
+\@ref(eq:swissroll). The two-dimensional meta data
+$(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the
+three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in
+the left plot of \autoref{fig:mappings}. The four colors in the mappings
+represent the four Gaussian kernels used to generate the meta data
+$(\pmb{X}_1, \pmb{X}_2)^\prime$. \begin{equation}
 \label{eq:swissroll}
@@
 
-Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1%, 5%, 50%, 99%, >99%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
+Now we are able to apply different manifold learning algorithms to
+$(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to
+$d=2$, and further estimate the density of the 2-D embedding. According
+to the density estimates, we could rank the data points and then
+identify which observations lie in the highest density region of
+specified coverage, eg. 1%, 5%, 50%, 99%, \>99%. For each of the five
+manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps,
+t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding
+plot in the same row, with the colors indicating the densities levels,
+the left column for true densities from the Gaussian mixture model, the
+middle column for highest density region plots with densities from our
+proposed variable KDE method, and the right for similar HDR plots with
+densities from KDE with fixed bandwidth. The top twenty outliers with
+the lowest densities are highlighted in black with point indexes in
+blue. From \autoref{fig:metadensity} and the data generating process, we
+know that there are four highest density regions. However, in all
+manifold learning embeddings colored with true densities (left column in
+\autoref{fig:sroutliers} ), except for LLE, the number of highest
+density regions are not the same as the meta data. When comparing the
+number of HDRs for variable and fixed bandwidth (middle and right column
+in \autoref{fig:sroutliers} ), our proposed method with variable
+bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian
+Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the
+top 20 outliers found rowwise, variable bandwidth could find most
+outliers lying on the left area of the embedding in ISOMAP and UMAP, and
+both methods in LLE embedding could find the outliers in the outer area,
+but for the other methods, both variable and fixed bandwidth are not
+detected true outliers accurately. For t-SNE and UMAP embedding, the
+embedding structure is highly distorted and the points are clustered
+together in a discontinuous way, which is also shown in the clustered
+outliers.
 
@@
 
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg.
-ISOMAP, or when the embedding is discontinuous, eg.
-t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
+To further compare the accuracy of the estimated densities for all data
+points, we calculate the correlation between the rank of true densities
+and the estimated densities and present in \autoref{tab:srcors}. It can
+be seen that the rank correlation of our proposed method with variable
+bandwidth is higher for LLE and UMAP, although the correlation for UMAP
+is very close to zero. The highest correlation comes from our method in
+LLE embedding, which is mainly due to it being closest to the
+rectangular structure of the meta data shown in
+\autoref{fig:metadensity}. For Laplacian Eigenmaps, our method has
+wrongly estimated the left area with lower densities even though their
+true densities are very high in yellow, leading to a negative
+correlation. The negative correlation would occur typically when the
+highest or lowest true density areas are not well estimated. As for the
+estimates in highly distorted embedding, including ISOMAP, t-SNE, and
+UMAP, the rank correlations are quite low. This shows that our proposed
+method could improve the kernel density estimate of manifold learning
+embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg. ISOMAP, or when the
+embedding is discontinuous, eg. t-SNE and UMAP, the density estimates
+are not as reliable for outlier detection.
 
@@
 
-For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
+For comparison, we use the same 2-D meta data in
+\autoref{fig:metadensity} with a different mapping function, twin peaks
+mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D
+structure shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 
-Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
-In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
-For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with the lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
-In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
-Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
-For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
-Also due to the clusters in the embedding, the outliers found in UMAP are clustered.
+Similar to \autoref{fig:sroutliers}, different manifold learning
+embeddings are obtained and used to detect outliers with true densities
+and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
+In general, the four highest density regions in yellow are identified in
+almost all manifold learning embeddings except for ISOMAP with fixed
+bandwidth and t-SNE. For ISOMAP, our proposed variable KDE, compared
+with the true density, gives the most accurate mixture kernel structure
+with the lowest estimated densities (darker colored points) in the
+outside and the center of the embedding, and the kernel means (yellow
+points) with highest densities are also clearly identified. In contrast,
+the fixed bandwidth KDE failed to identify the lowest density area in
+the center. Both variable and fixed bandwidth KDE are quite close with
+LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the
+top outliers are indexed in the middle instead of the true outer areas
+due to the large distortion in the middle. For t-SNE and UMAP, there are
+four clusters in the embedding and UMAP does a better job in finding the
+HDRs than t-SNE. Also due to the clusters in the embedding, the outliers
+found in UMAP are clustered.
 
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}. 
-Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
-Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being `r round(min(cors), 3)`.
-This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+We can gain further insight by comparing the correlation between ranks
+of true densities and estimated densities from variable and fixed
+bandwidth KDE by \autoref{tab:tpcors}. Again the highest correlations
+appear from embedding with higher quality, including ISOMAP, Laplacian
+Eigenmaps, and UMAP. The rank correlations between variable and fixed
+bandwidth are equivalent to the third decimal place in Laplacian
+Eigemaps and UMAP. As for t-SNE, the four clusters in the embedding are
+less separated than in UMAP and our proposed method has misidentified
+the kernel cores, leading to a lower rank correlation in variable
+bandwidth. Since the embeddings from twin peak data generally capture
+the rectangular structure in the meta data than those from the swiss
+roll data, the rank correlations are much higher in \autoref{tab:tpcors}
+than in \autoref{tab:srcors}, with the lowest correlation being
+`r round(min(cors), 3)`. This again suggests that the accuracy of
+outlier detection is highly related to the quality of manifold learning
+embedding.
 
@@
 
-In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
-The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
-Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+In \autoref{fig:tpisomapden}, we plot the estimated density against the
+true density of the ISOMAP embedding for KDE with variable and fixed
+bandwidth, with colors and shapes showing the four kernels in the meta
+data. The linear positive relationship between the true densities and
+variable KDEs on the left handside is stronger than that of the fixed
+bandwidth KDEs. Combined with the top-right subplot in
+\autoref{fig:tpoutliers}, we could tell that most points are
+underestimated near the true kernel cores, which also suggests that the
+fixed bandwidth tries to smooth across all the data points and fails to
+fix the local distortions in the manifold learning process like the
+proposed pointwise variable bandwidth.
 
@@
 
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in $d=5$ with manifold learning.
-First, we simulate $N=2,000$ points, $(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
-In order to manually add anomalies to be distant points from the means, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation $X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. 
-The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE].
-The round and triangular point shapes index the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the distance between the simulated $4-D$ data point from Gaussian mixture model and the kernel means $(0, 0, 0, 0)^\prime$.
-The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
-The dark colors also indicate that they are the true outliers because of their low densities.
-
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the $100$-dimensional data of size $N$ (denote the transpose of the data matrix as $\pmb{X}_0$ with dimension $100 \times N$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
-For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix.
-Matrix $\pmb{Q}$ satisfies
-$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
-Therefore, we use matrix $Q$ as the rotation matrix for where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms.
-Again, we set the embedding dimension to be equal to the meta data dimension $d=5$.
+As a high-dimensional experiment, we generate the meta data from a
+5-dimensional semi-hypersphere, transform it into a 100-dimensional
+space, and then embed it in $d=5$ with manifold learning. First, we
+simulate $N=2,000$ points,
+$(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a
+4-dimensional Gaussian mixture model with two mixture components,
+$\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
+$\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where
+$\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$,
+$\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
+In order to manually add anomalies to be distant points from the means,
+the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$. The
+fifth dimension is calculated to satisfy the five-dimensional
+semi-hypersphere surface equation
+$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is
+set as $7$. The Gaussian mixture densities could be calculated using
+Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D
+tour path with the R package *tourr* [REFERENCE]. The round and
+triangular point shapes index the two mixture components
+$\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
+$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the
+distance between the simulated $4-D$ data point from Gaussian mixture
+model and the kernel means $(0, 0, 0, 0)^\prime$. The more distant from
+the point to the kernel cores, the lower the true densities, which shows
+in a darker color in \autoref{fig:fivedmeta}. It can be seen that the
+most distant points are in a triangular shape, meaning that they are
+simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$. The dark
+colors also indicate that they are the true outliers because of their
+low densities.
+
+Then we initial the other 95 dimensions in the high-dimensional space as
+zero columns and further rotate the $100$-dimensional data of size $N$
+(denote the transpose of the data matrix as $\pmb{X}_0$ with dimension
+$100 \times N$) to get rid of the zeros so that it could be passed to
+the manifold learning algorithms. The rotation matrix is derived from
+the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all
+components randomly generated from a uniform distribution
+$\mathcal{U}(0,1)$. For any real matrix $\pmb{A}$ of dimension
+$p\times q$, the QR decomposition could decompose the matrix into the
+multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that
+$\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit
+norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and
+$\pmb{R}$ is an upper triangular matrix. Matrix $\pmb{Q}$ satisfies
+$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that
+the pairwise Euclidean distances between data points in
+$\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
+Therefore, we use matrix $Q$ as the rotation matrix for where the
+rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension
+$N \times 100$ is now the input data for the manifold learning
+algorithms. Again, we set the embedding dimension to be equal to the
+meta data dimension $d=5$.
 
@@
 
-In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
-Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
-For well-estimated densities, the true outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
-This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
-For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
-This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
-In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
-By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
+In \autoref{fig:fivedisomapden}, the estimated densities are compared
+with the true density on the x-axis for four manifold learning
+embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP. Note that we
+exclude t-SNE algorithm in this section because it is designed mainly
+for low-dimensional visualization purposes, and it is only applicable to
+embedding dimensions within three. Similar to \autoref{fig:fivedmeta},
+the point shapes show the two mixture component in the meta data, and
+the colors represent the distance to the kernel means, with distant
+outliers shown in darker colors. For well-estimated densities, the true
+outliers with low true densities will also have low estimated densities,
+which suggests that darker-colored points should appear in the
+bottom-left corner in \autoref{fig:fivedisomapden}. This is true for
+both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in
+UMAP where these outliers have relatively high density estimates. For
+variable bandwidth KDE, there is a strong positive linear relationship
+with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and
+the relationship is stronger than the fixed bandwidth. This suggests
+that our proposed KDE with variable bandwidth is more accurate than the
+fixed bandwidth in estimating the manifold learning embedding densities.
+In KDE with fixed bandwidth, the bandwidth is often too large to smooth
+across all data points, especially when there is severe distortion in
+the embedding data. By introducing the pointwise variable Riemannian
+metric in kernel density estimation, it is reasonable to believe that it
+could fix the distortion introduced by these three manifold learning
+algorithms.
 
@@
 
-In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010.
-The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
-We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.
-
-For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
-@Cheng2021-ex propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
-@Cheng2021-ex use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
-By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
-
- In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps have been clearly stated in the application section of @Cheng2021-ex and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
-However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
+In this application, we use the smart meter data from the *CER Smart
+Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in
+Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010. The
+CER dataset[^2] records the half-hourly electricity consumption of
+individual residential and commercial properties, not including energy
+for cooling or heating systems. We selected the $3,639$ residential data
+with no missing values during the data collection period for a total of
+$535$ days.
+
+[^2]: accessed via the Irish Social Science Data Archive -
+    www.ucd.ie/issda.
+
+For the electricity consumption data of residential individuals, it
+would be worthwhile to explore the distribution of electricity demand
+rather than the raw consumption data to study the usage patterns of
+different households or different periods or the week [@Hyndman2018-ia].
+@Cheng2021-ex propose two non-Euclidean distance estimators to enable
+manifold learning algorithms in statistical manifolds with each
+observation as a distribution. @Cheng2021-ex use the same smart meter
+data for identifying outliers with kernel bandwidth estimation but fail
+to consider the distortion and information loss in the 2-dimensional
+embeddings given that the input data dimension is much higher. By
+introducing the Riemannian metric as the bandwidth matrix, we could take
+into account the distortion in the 2-D embedding and further improve the
+accuracy of the density estimation.
+
+In this section, we first calculated the same empirical distributions of
+the $336$ half-hourly periods of the week for each household, and apply
+the total variation distance estimator proposed in @Cheng2021-ex in the
+statistical manifold learning to get the 2-D embedding of all
+households. Algorithm \@ref(alg:vkderm) is then used to obtain density
+estimates with the pointwise variable Riemannian metric as the bandwidth
+matrix and detect outliers. The data processing steps have been clearly
+stated in the application section of @Cheng2021-ex and they are skipped
+here. Unlike the simulations in \autoref{simulation}, we know nothing
+about the true density of the electricity distributions for all periods
+of the week and all households, so it is impossible to compare the
+estimated densities with the true meta data density as in
+\autoref{fig:fivedisomapden}. However, we could generate all the density
+estimates with the existing KDE method with fixed bandwidth, which is an
+optimal method for density estimation, and compare the densities from
+our proposed method with them.
 
@@
 
-\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
-As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
-It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
-In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+\autoref{fig:electricityoutliers} shows the electricity usage data of
+three households for both density estimation methods respectively, with
+the top one being the most typical household with the highest density
+and the bottom two being the top two outliers with the lowest densities.
+The typical households in the top row are close except that there are a
+few spikes for the one with fixed bandwidth. As to the anomalies,
+variable bandwidth tends to capture the unusual electricity demand
+volume when the usage is very low or high. It could also capture the
+unusual usage pattern when there are sudden spikes in ID 3243 or very
+high base electricity usage for the middle time periods in ID 2838. In
+contrast, fixed bandwidth KDE is more sensitive to spikes even when the
+spikes happen in a certain time window in 7049, or when the usage has an
+obvious time-of-week pattern with a few low electricity usage periods.
 
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
-Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-However, this repeated pattern in a week window is clearly for the typical household ID 1321.
-As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
-The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
-These findings show the difference in finding typical and anomalous households with different bandwidth selections.
+Further insights could be gained by comparing the quantile region plots
+of electricity demand against the time of the week for the same typical
+or anomalous households in \autoref{fig:electricityhdrsfixed} and
+\autoref{fig:electricityhdrsvkde}. Again the distribution of both
+typical households in the top panel has shown a repeated
+period-of-the-week usage pattern, with higher usage during mealtime on
+all seven days of the week and slightly higher usage for weekends.
+However, this repeated pattern in a week window is clearly for the
+typical household ID 1321. As for the distributions for outliers, the
+middle row outliers from variable bandwidth have spikes only on Tuesday
+and Sunday noons, while the fixed bandwidth has an increasing
+electricity demand across the day of the week. The bottom row outliers
+both have a repeated time usage pattern, but the electricity usage
+amount is higher with the highest median above 3kWh. These findings show
+the difference in finding typical and anomalous households with
+different bandwidth selections.
 
@@
 
-In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
-
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
-
-There are several open questions to be explored.
-The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of @Cheng2021-ex.
-The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
-The scale of distortion in each point in manifold learning could vary a lot.
-If we could smooth across all data points, eg.
-multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
-The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
-The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+In this paper, we propose a new method to estimate the density of
+manifold learning embedding and further identify outliers based on the
+densities. The Riemannian metric measure the direction and angle of the
+distortion when mapping data points through a non-linear function in
+manifold learning algorithms. By introducing the Riemannian metric as
+the pointwise variable bandwidth matrix in kernel density estimation,
+the local distortion in the low-dimensional embedding could be used to
+estimate densities, leading to a more accurate description of the data
+distributions. We compare our proposed method with fixed bandwidth KDE
+by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or
+twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D
+space, and show that variable bandwidth could improve the density
+estimation given a good manifold learning embedding.
+
+As an empirical example, we explore the distributions of different
+households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian
+Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and
+KDE with both variable and fixed bandwidth are used to get the density
+estimates. We compare both density estimates by looking at the
+distributions of the most typical households with the highest densities
+and the most anomalous households with the lowest densities. Both
+methods could identify the typical households with certain usage
+patterns, while the outliers are anomalous in different ways.
+
+There are several open questions to be explored. The first involves the
+selection of tuning parameters for the manifold algorithm so that a
+maximal level of embedding quality is achieved, where embedding quality
+is measured using one of the metrics discussed in the online
+supplementary material of @Cheng2021-ex. The scaling of the Riemannian
+metric to get the closest range of the true densities is also worth
+exploring. The scale of distortion in each point in manifold learning
+could vary a lot. If we could smooth across all data points, eg.
+multiply the Riemannian matrix with the ratio between its determinant
+and the sum of all Riemannian matrix determinants, the global density
+estimates could be potentially smoothed. The density estimates on the
+edges of the whole data structure could be improved because most outer
+area points tend to be detected as outliers. The choice of manifold
+learning algorithms also has a large impact on the embedding accuracy,
+which we have seen will affect the density estimation and outlier
+detection. However, the outperformance of VKDE with Riemannian bandwidth
+than the fixed bandwidth has been shown in the higher dimensional
+simulation data and the electricity usage data, which are more related
+to real-life data sets.
 
@@
 <!-- Discuss simulation with non-uniform meta data. -->
+
+```{=html}
 <!-- The aim of the project is not to evaluate different manifold learning algorithms.
@@
  -->
-
-# Acknowledgment {-}
-
-This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
+```
+# Acknowledgment {.unnumbered}
+
+This research was supported in part by the Monash eResearch Centre and
+eSolutions-Research Support Services through the use of the MonARCH HPC
+Cluster.
 
 \clearpage\appendix
+
 <!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
+
 <!-- %\pagenumbering{arabic} -->
+
 <!-- %\def\thepage{A\arabic{page}} -->
@@
 
-In this appendix, we present some notions about the Riemannian geometry used in this paper.
+In this appendix, we present some notions about the Riemannian geometry
+used in this paper.
 
@@
 
-In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions. 
-A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
-Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
-A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
-The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
-$$
+In topology, a *homeomorphism* is a bijective map between two
+topological spaces that is continuous in both directions. A *Hausdorff
+space* is a topological space where any two distinct points can be
+separated by disjoint neighborhoods. A $d$-dimensional (topological)
+*manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where
+the neighborhood $U$ for each point $p$ is homeomorphic to an open
+subset $V$ of the Euclidean space $\mathbb{R}^d$. Such a homeomorphism
+$\varphi: U \rightarrow V$ together with $U$ gives a (coordinate)
+*chart*, denoted as $(U, \varphi)$, with the corresponding local
+coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$. A collection of
+charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is
+called an *atlas*, denoted as $\mathcal{A}$. The manifold $M$ is a
+*differentiable manifold* if there exists an atlas of $M$,
+$\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between
+any two charts, $$
 \varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-$$
-are differentiable of class $C^\infty$ (smooth).
-
-Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$.
-Then $\varphi$ is an embedding of $E$ into $M$ if and only if $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
+$$ are differentiable of class $C^\infty$ (smooth).
+
+Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$. Then
+$\varphi$ is an embedding of $E$ into $M$ if and only if
+$\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$
+is called an embedded submanifold of $M$ with the subspace topology.
 
@@
 
-The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through the point $p$ or as the directional derivatives at $p$.
-Here we define the tangent vector via the velocity of curves.
-
-For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e.
-$\gamma_1(0) = \gamma_2(0) = p$.
-$\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
-$$
+The tangent vector at point $p$ can be intuitively viewed as the
+velocity of a curve passing through the point $p$ or as the directional
+derivatives at $p$. Here we define the tangent vector via the velocity
+of curves.
+
+For any $p \in M$, let
+$\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and
+$\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve
+through $p$, i.e. $\gamma_1(0) = \gamma_2(0) = p$. $\gamma_1$ and
+$\gamma_2$ are *equivalent* if and only if there exists a chart,
+$(U,\varphi)$, at $p$ such that $$
 (\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-$$
-A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$. 
-The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
-The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
-The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
-Tangent vectors can also be seen as the directional derivatives at $p$.
-For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.
-
-The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the *cotangent space* with corresponding *cotangent vector* $z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
+$$ A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any
+equivalent class of the differentiable curves initialized at $p$. The
+set of all tangent vectors at $p$ defines the *tangent space* of $M$ at
+$p$, denoted as $T_pM$. The tangent space is a vector space of dimension
+$d$, equal to the dimension of $M$, and it does not depend on the chart
+$\varphi$ locally at $p$. The collection of all tangent spaces defines
+the *tangent bundle* $TM = \cup_{p \in M}T_pM$. Tangent vectors can also
+be seen as the directional derivatives at $p$. For a given coordinate
+chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial
+derivatives are denoted as
+$\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$,
+which define a *basis* of the tangent space.
+
+The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the
+*cotangent space* with corresponding *cotangent vector*
+$z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as
+$dx_1(p),\dots,dx_d(p)$.
 
@@
 
-A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
-Generally, we omit the subscript $p$ and refer to $g$ as the Riemannian metric.
-The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. 
-A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
-
-The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
-$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
-For a curve $\gamma: I \rightarrow M$, the length of the curve is 
+A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each
+point $p$ is a local inner product
+$T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$
+symmetric positive definite and varies smoothly at $p$. Generally, we
+omit the subscript $p$ and refer to $g$ as the Riemannian metric. The
+inner product for two vectors $u, v \in T_pM$ is written as
+$<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where
+implicit summation over all indices, $\sum_{i,j}$, is assumed. A
+differentiable manifold $M$ endowed with the Riemannian metric $g$ on
+each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
+
+The Riemannian metric $g$ can be used to define the norm of a vector
+$u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and
+$v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric
+quantities induced by $g$. $g$ could also be used to define the line
+element $dl^2 = g_{ij}dx_idx_j$ and the volume element
+$dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the
+local coordinates of the chart $(U, \varphi)$. For a curve
+$\gamma: I \rightarrow M$, the length of the curve is
 $$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
@@
 Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
-$$
-which is also called the *Riemannian measure* on $M$.
-
-The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$,
-the second-order ordinary differential equation is satisfied,
-$$
+$$ which is also called the *Riemannian measure* on $M$.
+
+The *geodesics* of $M$ are the smooth curves that locally joins the
+points along the shortest path on the manifold. A curve
+$\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the
+second-order ordinary differential equation is satisfied, $$
 \frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
-$$
-where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by
-$$
+$$ where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by $$
 \Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
 + \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
-$$
-The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$.
-The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
-For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
-The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
+$$ The geodesics have a constant speed with norm
+$\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc
+length functional
+$l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$
+when the curves are defined over the interval $[0,1]$. The *geodesic
+distance* $d_g$ is the length of the shortest geodesic between two
+points on the manifold. For a point $p \in M$, when the geodesic
+distance starting at $p$ is not minimized, we call such set of points
+the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the
+distance to the cut locus. The injectivity radius of $(M,g)$,
+$\textit{inj}_gM$ is the infimum of the injectivity radii over all
+points on the manifold.
 
@@
 
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
-Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
-The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
-$$
+Pushforward and pullback are two notions corresponding to the notions of
+tangent and cotangent vectors. Let $f$ be an embedding from the
+Riemannian manifold $(M,g)$ to another smooth manifold $E$. The
+pushforward $h=\varphi*g$ of the Riemannian metric $g$ along
+$\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which
+maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors
+$f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for
+$u, v \in T_f(p)E$, $$
 <u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-$$
-where $df_p^{-1}$ is the Jacobian of $f^{-1}$. 
-The tangent vectors  $f_*u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
-Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
+$$ where $df_p^{-1}$ is the Jacobian of $f^{-1}$. The tangent vectors
+$f_*u_p$ are equivalent to the velocity vector of a curve
+$\gamma: I\rightarrow M$ passing through point $p$ at time zero with a
+constant speed $\gamma^{-1}(0)=u_p$. Similarly, the pullback maps the
+cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at
+$p \in M$ acting on tangent vectors $u \in T_pM$.
 
 <!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->
+
 ## Exponential map and logarithmic map
 
-Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$.
-Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e.
-*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$.
-The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit. 
-The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$.
-Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
-Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
+Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with
+raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open
+neighborhood of $0_p$ in the tangent space at $p$, $T_pM$. Define
+$exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a
+differentiable, bijective map of differentiable inverse (i.e.
+*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the
+endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying
+$\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$. The
+exponential map moves point $p$ from $p$ at speed $u$ to an endpoint
+after covering the length of $\|u\|$ along the geodesic in one time
+unit. The inverse of the exponential map is called the *logarithm map*,
+$\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to
+$q$. Also define the *geodesic ball* centered at $p$ of radius $r > 0$
+as the image by the exponential map of $B(0_p, r) \subset T_pM$ with
+$r < \textit{inj}_gM$. Then we could interpolate a geodesic $\gamma$
+between two points $p$ and $q$ with the exponential map and the
+logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic
+distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
 
@@
 
-## Volumn density function
-
-For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
-Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
-$$
+## Volume density function
+
+For any $u \in T_pM$, let
+$\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let
+$w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$,
+for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an
+orthonomal basis of the tangent space $T_pM$. Define
+$\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function
+on $M$ as $$
 \theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-$$
-<!-- $$ -->
+$$ <!-- $$ -->
 <!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
-<!-- $$ -->
-Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
-The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and
-$\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
-$$
+<!-- $$ --> Rewrite $\theta_p(u) = \theta_p(\log_p(u))$. The exponential
+map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form
+$\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with
+respect to the Lebesgue measure of the Euclidean structure on $T_pM$,
+which gives in normal coordinates $$
 d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
@@
 
-#### Example: $S^1$
+## Example: $S^1$
 
@@
 
-Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
-
-Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$
+Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with
+$f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
+
+Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and
+$y = \sin(\theta) \in [0, 1]$
 
@@
 
-The CDF of X is given by
-$$
+The CDF of X is given by $$
 P(X \leq x) = P(\cos(\theta) \leq x) \\
@@
 = 1 - \frac{2}{\pi} \arccos(x).
-$$
-Then the density of X is the derivative of the CDF w.r.t. $x$
-$$
+$$ Then the density of X is the derivative of the CDF w.r.t. $x$ $$
 f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
 = \frac{2}{\pi \sqrt{1 - x^2}}.
-$$
-Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ and compare it with its true density $f(x)$.
+$$ Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
+estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
+and compare it with its true density $f(x)$.
 
@@
 
-<!-- First, we use the kernel density estimator in \@ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->
-
-We use the density estimator in @Pelletier2005-vu. 
-<!-- \begin{equation} -->
-<!-- \label{eq:pelletier} -->
-$$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
-$$
-<!-- \end{equation} -->
-where $d=2$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
-
-Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$.
-Therefore, the geodesic distance is
-$$
+<!-- First, we use the kernel density estimator in @ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->
+
+We use the density estimator in @Pelletier2005-vu. For point
+$p \in S^1$, the density at $p$ is estimated by
+<!-- \begin{equation} --> <!-- \label{eq:pelletier} --> $$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
+$$ <!-- \end{equation} --> where $d=1$, $h$ is the bandwidth,
+$\theta_{X_i}(p)$ is the volume density function at sample point $X_i$,
+and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along
+the manifold $S^1$.
+
+Since the manifold is known ($S^1$), we know the geodesic distance is
+the arc length between two points on the manifold, which is given by
+$L(\gamma) = r(\beta - \alpha)$ with $r=1$. Therefore, the geodesic
+distance is $$
 d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-$$
-As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
-$|g(\exp_p^{-1}(x_i))|$.
-
-[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
-
-The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
-$$
+$$ As stated in @Pelletier2005-vu, in terms of *geodesic normal
+coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$
+equals the determinant of the metric $g$ expressed in these coordinates
+at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$.
+
+[When using normal coordinates(orthogonal basis) for tangent vector in
+$T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
+
+The Riemannian metric $g$ for $S^1$ is given by the inner products on
+the tangent spaces of $S^1$. For each $X_i \in M$, $$
 g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-$$
-is smooth in a local neightborhood $U$ of $M$.
-
-Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
-$$
-\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
-$$
-Therefore, the estimator for $p \in S^1$ is given by
-$$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
-^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
-$$
-The bandwidth $h$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.
+$$ is smooth in a local neighborhood $U$ of $M$.
+
+Then the volume density function $\theta_{X_i}(p)$ is the inner product
+of the tangent vectors produced by the logarithm map from $X_i$ to $p$,
+which is given by $$
+\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
+$$ Therefore, the estimator for $p \in S^1$ is given by $$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+$$ where the bandwidth $r$ is selected by the plug-in bandwidth selector
+implemented by the R package *ks*.
+
+### How to derive volume density function $\theta_{X_i}(p)$
+
+Example of $S^2$ in @Henry2009-om.
+
+Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let
+$v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$
+is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential
+chart $(U, \Psi)$ induced by the parametrisation
+$\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by $$
+\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
+$$ if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where
+$r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
+Then we have $$
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
+$$ Then each component of the riemannian metric could be calculated as
+$g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$. We obtain
+that the volume density on the sphere is $$
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
+$$ for $q \neq p, -p$, and $\theta_p(p) = 1$.
+
+In $S^1$, polar coordinates: $\phi(\exp_p{vt})^{\prime}(0) = v$
+
+cartesian coordinates:
+$\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}$
+\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
 

### 2022-05-03 paper/paper.tex
Add S2 example reference

--- 2022-04-24 paper/paper.tex
@@
 
-Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
-It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
-To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
-
-Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular methods to calculate the probability density function from a sample dataset.
-KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
-There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
-For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
-However, these density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
-\autocite{Henry2009-ll} In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from \textcite{Pelletier2005-vu} is not directly applicable.
-This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied to get a low-dimensional embedding.
-We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
-
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
-One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}.
-To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-
-The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
-The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
-In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(M\).
-This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.
-
-The rest of the paper is organized as follows.
-In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
-In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
-\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
-Conclusions and discussions are presented in \autoref{conclusion}.
-Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
+Anomaly detection has been an important and diverse area where anomalies
+or outliers are detected in a given data. It often involves a data
+analysis process to uncover the unusual patterns and has been widely
+applied to machine learning \autocite{Omar2013-lz}, network intrusions
+identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging
+\autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security
+\autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which
+explored detecting households with anomalous electricity usage
+distributions instead of raw data. In the case of non-Euclidean sample
+space, the observations lie on a Riemannian manifold embedded in a very
+high-dimensional ambient space, which makes it computationally expensive
+or impossible to detect anomalies. To address this problem, we propose a
+kernel density estimator of the low-dimensional statistical manifold
+embedding and find outliers as the data with the lowest density
+estimates.
+
+Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of
+the most popular methods to calculate the probability density function
+from a sample dataset. KDE is flexible to learn the shape of the
+underlying density from the data controlled by the bandwidth and the
+selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}. There is extensive research on bandwidth selection,
+two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}. For
+multivariate data, variable kernel density estimator {[}VKDE;
+\textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive
+bandwidth matrix to control the amount of smoothing on the location of
+the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data
+point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}. However, these
+density estimators are based on random samples in the Euclidean space.
+
+For samples points lying on a manifold with the differentiable structure
+called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed a kernel
+density estimator based on the kernel weight depending on the distance
+between the estimated points and the sample data points. The idea of the
+estimator is to use a strictly positive function of the geodesic
+distance on the manifold and then normalize it with the volume density
+function of the Riemannian manifold for curvature. \autocite{Henry2009-ll} In
+many application scenarios, we tend to find that the sample points on a
+manifold are embedded in a much higher-dimensional space, and the kernel
+density estimator from \textcite{Pelletier2005-vu} is not directly applicable.
+This is when we introduce manifold learning to reduce the input data
+dimension. For these high-dimensional data set, various manifold
+learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE,
+and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied
+to get a low-dimensional embedding. We propose a kernel density
+estimator to be applied to the low-dimensional embedding and define
+outliers as those with the lowest densities.
+
+In manifold learning, the underlying idea is that the data lies on a
+low-dimensional smooth manifold that is embedded in a high-dimensional
+space. One of the fundamental objectives of manifold learning is to
+explore the geometry of the dataset, including the distances between
+points and volumes of regions of data. These intrinsic geometric
+attributes of the data, such as distances, angles, and areas, however,
+can be distorted in the low-dimensional embedding, leading to failure to
+recover the geometry of the manifold \autocite{Goldberg2008-co}. To tackle this
+problem and measure the distortion incurred in manifold learning,
+\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment
+any existing embedding output with geometric information in the
+Riemannian metric of the manifold itself. By applying this algorithm,
+the outputs of different manifold learning methods can be unified and
+compared under the same framework, which would highly benefit in
+improving the effectiveness of the embedding.
+
+The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives
+some idea of the distortion of an embedding. Mapping the points through
+a non-linear function ``stretches'' some regions of space and ``shrinks''
+others. The Riemannian gives us an idea of the direction and angle of
+this stretching, which is informative for learning the manifold. In
+variable kernel density estimate, the bandwidth matrix \(H\) is also
+defined to control the amount of smoothing for each data point.
+Therefore, we propose to the use Riemannian metric as the bandwidth
+matrix in, we could further get the kernel density estimation of the
+manifold \(M\). This kernel density estimate can then be used to produce
+the highest density region plots \autocite{Hyndman1996-lk} for outlier
+visualization.
+
+The rest of the paper is organized as follows. In \autoref{vkde}, we
+present the proposed algorithm to detect anomalies based on variable
+kernel density estimates of manifold embeddings. In this section, we
+provide justification for the use of the Riemannian metric as the
+bandwidth of variable kernel density estimation, including the
+comparison with fixed bandwidth. \autoref{simulation} is composed of two
+simulations with the proposed algorithm; the first deals with
+2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data
+and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+\autoref{application} contains the application to visualize and identify
+anomalies in the Irish smart meter dataset. Conclusions and discussions
+are presented in \autoref{conclusion}. Readers interested in the the
+notions of Riemannian geometry mentioned in this paper could use the
+Appendix in \autoref{riemgeo} as a reference.
 
@@
 
-In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
-To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with the lowest density estimates.
-The proposed schematic is shown in Figure \ref{fig:vkde}.
-The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
+In this section, we introduce the proposed method to detect anomalies
+based on the kernel density estimates of manifold learning embeddings
+where the Riemannian matrix is used as the pointwise variable bandwidth
+to measure the direction and angle of the distortion of the
+low-dimensional embeddings. \textcite{Perrault-Joncas2013-pq} gives us an idea of
+how to measure the direction and angle of the distortion using the
+Riemannian metric which is a positive definite square matrix for each
+data point. To learn the distribution of the low-dimensional embedding,
+we use the kernel density estimation with the bandwidth matrix being the
+Riemannian metric. The outliers could then be defined as the points with
+the lowest density estimates. The proposed schematic is shown in Figure
+\ref{fig:vkde}. The highlighted two steps in Figure \ref{fig:vkde} are
+the main contributions of this main chapter, replacing the bandwidth
+matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
+variable kernel density estimate, and computing the highest density
+region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for
+anomaly detection.
 
@@
 
-\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
+\caption{The proposed schematic for variable kernel density
+estimation with recovered geometry.}\label{fig:vkde}
 \end{figure}
 
-To start with, we introduce the notations used in this manuscript.
-Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
-Readers familiar with these topics could skip the corresponding subsections.
-Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
+To start with, we introduce the notations used in this manuscript. Then
+we introduce the multivariate kernel density estimation method with
+variable bandwidth matrix and the metric learning algorithm to derive
+the pointwise Riemannian metric. Readers familiar with these topics
+could skip the corresponding subsections. Finally, we propose our novel
+algorithm to estimate densities and detect anomalies for
+high-dimensional data set.
 
@@
 
-In general, given a multivariate random sample \(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
+In general, given a multivariate random sample
+\(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\),
+\(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel
+density estimate of \(f\) at point
+\(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
 \begin{equation}
@@
 \hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
-\end{equation}
-where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
-When using the standard gaussian kernel \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
+\end{equation} where \(\pmb{H}_i\) is called the bandwidth matrix which is
+a \(d\times d\) symmetric positive definite matrix variable at each \(i\),
+\(K(\pmb{y})\) is the kernel function, and
+\(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
+When using the standard gaussian kernel
+\(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
 \begin{equation}
@@
 K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
-\end{equation}
-with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
-
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{y})\).
-We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
-\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
-Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
+\end{equation} with \(|\pmb{H}_i|\) being the determinant of matrix
+\(\pmb{H}_i\).
+
+The bandwidth matrix is important in controlling the smoothing across
+all data points and the shape is determined by the kernel function
+\(K(\pmb{y})\). We consider the sample smooth estimator for each data
+point, which gives a kernel density estimate with local smoothing.
+\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to
+6 dimensions in the R package \emph{ks}. However, for variable kernel density
+estimation where the bandwidth is no longer a constant matrix, the
+implementation is limited to 2-dimension and the bandwidth can either be
+a vector or a diagonal matrix instead of a full bandwidth matrix.
+Further, the estimator in Equation \eqref{eq:kde} is only applicable when
+the sample space is Euclidean. In the cases of a Riemannian manifold,
+the sample space used for kernel density estimation is no longer
+Euclidean and we need to use the estimator described in
+\autoref{riemest}.
 
@@
 
-Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
-Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\).
-Denote \(A_i, i = 1, 2, \dots, N\), where \(\pmb{A}_i \in \mathcal{R}^d\), as i.i.d.
-random objects on \(M\) with density \(f\).
-For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
-\begin{equation}
+Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without
+boundary and a probability distribution with density \(f\) on the
+manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is
+the Riemannian distance induced by the Riemannian, \(g\), and the strictly
+positive injectivity radius \autocite{Chavel2006-mp} of the manifold,
+\(inj_g(M)\). Denote \(A_i, i = 1, 2, \dots, N\), where
+\(\pmb{A}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with
+density \(f\). For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel
+density estimator of \(f\) to be \begin{equation}
 \label{eq:denriem}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
-\end{equation}
-where \(h\) is the bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
-\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q \(\exp_q^{-1}(p)\).
-The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
-\begin{equation}
+\end{equation} where \(h\) is the bandwidth satisfying \(h \leq h_0\) for
+fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\),
+\(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and
+\(\theta_{q}(p)\) is the volume density function on the manifold.
+\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\)
+defined for \(p\) in a neighborhood of \(q\) on the manifold, with the
+geodesic normal coordinates at \(q\), is equal to the determinant of the
+Riemannian metric \(g\) at the logarithm map at q \(\exp_q^{-1}(p)\). The
+expression in \eqref{eq:denriem} is also proved to be consistent with the
+kernel density estimators in the Euclidean case as \begin{equation}
 \label{eq:denconsist}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
-\end{equation}
-and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.
-
-However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
-Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
-Therefore, based on the estimator in \eqref{eq:denriem}, we propose to a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
-\begin{equation}
+\end{equation} and converges at the same rate as the Euclidean kernel
+density estimator \autocite{Henry2009-ll}.
+
+However, in application, we usually found the observations lie on a
+\(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where
+\(d \ll s\). Then manifold learning can be applies to find the
+\(d\)-dimensional representation of the manifold \(M\). Therefore, based on
+the estimator in \eqref{eq:denriem}, we propose to a kernel density
+estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to
+be \begin{equation}
 \label{eq:denestimator}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
-\end{equation}
-where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
-
-An important goal of manifold learning is to recover the local or global features of the data.
-In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric \(g\).
-This could also be used to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix for each data point.
-Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \eqref{eq:denestimator}.
-In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
+\end{equation} where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable
+bandwidth matrix and \(h_i\) is positive definite with dimension
+\(d\times d\).
+
+An important goal of manifold learning is to recover the local or global
+features of the data. In order to recover the geometry of the manifold,
+\textcite{Perrault-Joncas2013-pq} propose a method to augment any existing
+reasonable embedding and allow for the computation of geometric values
+to be calculated with an estimation of the Riemannian metric \(g\). This
+could also be used to fix the geometric distortion in the embedding. It
+is worth noticing that the Riemannian metric from
+\textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix for
+each data point. Therefore, we propose to take the geometric distortion
+into consideration when estimating the kernel densities of the embedding
+with Equation \eqref{eq:denestimator}. In this way, we present a
+consistent kernel density estimator with pointwise variable bandwidth as
+the Riemannian metric.
 
@@
 
-The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_pM\) for every point \(p \in M\).
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((M, g)\).
-
-To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
-The inner product between two vectors \(u,v \in T_pM\), \(<u,v>_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{<u,v>_g}\) and the angle between two vectors \(\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}\) in the tangent space.
-Therefore, for each point \(p\in M\) in any coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
-
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\in M\) is defined as
-\[
+The Riemannian metric \(g\) is a symmetric and positive definite tensor
+field which defines an inner product \(<,>_g\) on the tangent space \(T_pM\)
+for every point \(p \in M\). If the inner product of the tangent space is
+known for a given geometry, the Riemannian metric is a good measure to
+recover the geometry of the manifold. The Metric Learning algorithm
+\autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the
+Riemannian metric and produce a Riemannian manifold \((M, g)\).
+
+To recover the original geometry of the manifold, we need to know what
+the inner product corresponds to in the embedding. The inner product
+between two vectors \(u,v \in T_pM\), \(<u,v>_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote
+  summation over \(i\) and \(j\)}, can be
+used to define some geometric quantities, such as the vector norm
+\(\|u\|=\sqrt{<u,v>_g}\) and the angle between two vectors
+\(\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}\) in the tangent space.
+Therefore, for each point \(p\in M\) in any coordinate system, the
+Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite
+matrix, where \(d\) is the dimension of the manifold.
+
+The line element and volume element of the full manifold or a subset of
+the manifold can also be computed from \(g\). The arc length of a curve
+\(c\in M\) is defined as \[
 l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-\]
-where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function mapping \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
-\[
+\] where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and
+\(c(t)\) is a function mapping \([a,b]\) to \(M\). While the volume of
+\(V\subset M\) is computed by \[
 Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
-\]
-Both the concepts of distance and volume are relevant to kernel density estimation.
-
-\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
-We restate these four steps in Algorithm \ref{alg:learnmetric}.
-By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_M\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
-
-As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
-This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
-It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
+\] Both the concepts of distance and volume are relevant to kernel
+density estimation.
+
+\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly
+involves four steps: weighted neighborhood graph construction; geometric
+graph Laplacian calculation; manifold learning; and Riemannian metric
+calculation. We restate these four steps in Algorithm
+\ref{alg:learnmetric}. By applying an existing manifold learning
+algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a
+low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed. Most
+manifold learning methods involve the construction of the nearest
+neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\)
+is built. The Laplacian is quite useful because it can be
+coordinate-free while containing all the important geometry.
+\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated
+\(\Delta_M\) with a discrete consistent estimator, the geometric graph
+Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where
+\(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the
+Metric Learning algorithm to achieve the augmented embedding with the
+Riemannian metric \((f_n, g_n)\).
+
+As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension
+\(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the
+embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\)
+will be returned. This algorithm is also implemented in a Python library
+\emph{megaman} \autocite{McQueen2016-xz}. It is designed to apply the manifold
+learning methods to large-scale data sets, as well as compute the
+Riemannian metric of the manifold.
 
@@
 
-Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \ref{alg:vkderm}.
-There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
-Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels.
-Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}.
-The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).
+Now we present our proposed algorithm for anomaly detection based on
+variable kernel density estimates in \ref{alg:vkderm}. There are mainly
+four steps involved in the algorithm. The first step is to apply the
+Learn metric algorithm described in \ref{alg:learnmetric} to the input
+high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to
+get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\)
+and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for
+each observation. Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\)
+are given for Gaussian kernels. Then we could use the pointwise
+Riemannian metric to calculate the kernel density estimate with our
+proposed estimator in Equation \eqref{eq:denestimator}. The top outliers
+of size \(n\_outliers\) are obtained by ordering the embedding points
+\(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).
 
@@
 
-If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
-1\%, 50\%, 99\%, \textgreater99\%.
+If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could
+be used to identify the relative location of outliers in the embedding
+and find which observations lie in the highest density region of
+specified coverage, eg. 1\%, 50\%, 99\%, \textgreater99\%.
 
@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
-We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
-Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as \(d=2\), the same as the meta data dimension.
-This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
+In this section, we examine two scenarios for both low and high
+dimensions to test our proposed algorithm. For visualization purposes,
+\autoref{twodgaussian} presents a 2-D meta data example. We first
+simulate the data of size \(N=2,000\) from a mixture of four Gaussian
+kernels with the same covariance but different means, each consisting of
+\(500\) points. Different mapping functions are then applied to the 2-D
+meta data to be mapped in a 3-D feature space, which gives the
+higher-dimensional input for different manifold learning algorithms,
+including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The
+embedded dimension is set as \(d=2\), the same as the meta data dimension.
+This enables us to compare the manifold learning embedding with the true
+meta data. We could now apply Algorithm \ref{alg:vkderm} to get the
+density estimates of all data points and further detect anomalies. As a
+high-dimensional example, the second simulation in
+\autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\)
+embedded in a 100-D space and the corresponding embedding dimension is
+\(d=5\).
 
@@
 
-We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
-The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
-Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
-\begin{equation}
+We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian
+mixture model with four components with different means
+\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\)
+and the same variance-covariance matrix
+\(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\). The mixture proportions
+are equally set as \(\pi_i=0.25, i=1,2,3,4\).
+Then the mixture Gaussian mixture density function is a weighted linear
+combination of the four component Gaussian densities as \begin{equation}
 \label{eq:gmm}
 P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
-\end{equation}
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
+\end{equation} \autoref{fig:metadensity} shows the 2-dimensional meta
+data and the colors indicate the true density of all data points
+calculated from \eqref{eq:gmm}, with brighter colors showing higher
+densities and darker colors showing lower densities. We then define
+outliers as points with lowest densities shown in black and typical
+points with highest densities shown in yellow. Based on the true density
+plot, the outliers are scattered in the middle and the outer area of the
+whole structure, while typical points are near the means of four
+kernels. These are \emph{true outliers} to be compared with outliers from the
+kernel density estimates.
 
@@
 
-Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
-The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
-\begin{equation}
+Given the 2-D meta data, multiple mapping functions could be applied to
+embed the data in a 3-D space. One of the most famous examples in
+manifold learning is the swiss roll data, with the mapping function in
+\eqref{eq:swissroll}. The two-dimensional meta data
+\((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the
+three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in
+the left plot of \autoref{fig:mappings}. The four colors in the mappings
+represent the four Gaussian kernels used to generate the meta data
+\((\pmb{X}_1, \pmb{X}_2)^\prime\). \begin{equation}
 \label{eq:swissroll}
@@
 
-Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1\%, 5\%, 50\%, 99\%, \textgreater99\%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
+Now we are able to apply different manifold learning algorithms to
+\((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to
+\(d=2\), and further estimate the density of the 2-D embedding. According
+to the density estimates, we could rank the data points and then
+identify which observations lie in the highest density region of
+specified coverage, eg. 1\%, 5\%, 50\%, 99\%, \textgreater99\%. For each of the five
+manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps,
+t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding
+plot in the same row, with the colors indicating the densities levels,
+the left column for true densities from the Gaussian mixture model, the
+middle column for highest density region plots with densities from our
+proposed variable KDE method, and the right for similar HDR plots with
+densities from KDE with fixed bandwidth. The top twenty outliers with
+the lowest densities are highlighted in black with point indexes in
+blue. From \autoref{fig:metadensity} and the data generating process, we
+know that there are four highest density regions. However, in all
+manifold learning embeddings colored with true densities (left column in
+\autoref{fig:sroutliers} ), except for LLE, the number of highest
+density regions are not the same as the meta data. When comparing the
+number of HDRs for variable and fixed bandwidth (middle and right column
+in \autoref{fig:sroutliers} ), our proposed method with variable
+bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian
+Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the
+top 20 outliers found rowwise, variable bandwidth could find most
+outliers lying on the left area of the embedding in ISOMAP and UMAP, and
+both methods in LLE embedding could find the outliers in the outer area,
+but for the other methods, both variable and fixed bandwidth are not
+detected true outliers accurately. For t-SNE and UMAP embedding, the
+embedding structure is highly distorted and the points are clustered
+together in a discontinuous way, which is also shown in the clustered
+outliers.
 
@@
 
-For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
+For comparison, we use the same 2-D meta data in
+\autoref{fig:metadensity} with a different mapping function, twin peaks
+mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D
+structure shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
-Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
-Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259.
-This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+We can gain further insight by comparing the correlation between ranks
+of true densities and estimated densities from variable and fixed
+bandwidth KDE by \autoref{tab:tpcors}. Again the highest correlations
+appear from embedding with higher quality, including ISOMAP, Laplacian
+Eigenmaps, and UMAP. The rank correlations between variable and fixed
+bandwidth are equivalent to the third decimal place in Laplacian
+Eigemaps and UMAP. As for t-SNE, the four clusters in the embedding are
+less separated than in UMAP and our proposed method has misidentified
+the kernel cores, leading to a lower rank correlation in variable
+bandwidth. Since the embeddings from twin peak data generally capture
+the rectangular structure in the meta data than those from the swiss
+roll data, the rank correlations are much higher in \autoref{tab:tpcors}
+than in \autoref{tab:srcors}, with the lowest correlation being
+0.259. This again suggests that the accuracy of
+outlier detection is highly related to the quality of manifold learning
+embedding.
 
@@
 
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
-First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
-In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
-The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
-The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
-The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
-The dark colors also indicate that they are the true outliers because of their low densities.
-
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
-For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix.
-Matrix \(\pmb{Q}\) satisfies
-\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
-Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
-Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).
+As a high-dimensional experiment, we generate the meta data from a
+5-dimensional semi-hypersphere, transform it into a 100-dimensional
+space, and then embed it in \(d=5\) with manifold learning. First, we
+simulate \(N=2,000\) points,
+\((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a
+4-dimensional Gaussian mixture model with two mixture components,
+\(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
+\(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where
+\(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\),
+\(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
+In order to manually add anomalies to be distant points from the means,
+the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\). The
+fifth dimension is calculated to satisfy the five-dimensional
+semi-hypersphere surface equation
+\(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is
+set as \(7\). The Gaussian mixture densities could be calculated using
+Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D
+tour path with the R package \emph{tourr} {[}REFERENCE{]}. The round and
+triangular point shapes index the two mixture components
+\(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
+\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the
+distance between the simulated \(4-D\) data point from Gaussian mixture
+model and the kernel means \((0, 0, 0, 0)^\prime\). The more distant from
+the point to the kernel cores, the lower the true densities, which shows
+in a darker color in \autoref{fig:fivedmeta}. It can be seen that the
+most distant points are in a triangular shape, meaning that they are
+simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\). The dark
+colors also indicate that they are the true outliers because of their
+low densities.
+
+Then we initial the other 95 dimensions in the high-dimensional space as
+zero columns and further rotate the \(100\)-dimensional data of size \(N\)
+(denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension
+\(100 \times N\)) to get rid of the zeros so that it could be passed to
+the manifold learning algorithms. The rotation matrix is derived from
+the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all
+components randomly generated from a uniform distribution
+\(\mathcal{U}(0,1)\). For any real matrix \(\pmb{A}\) of dimension
+\(p\times q\), the QR decomposition could decompose the matrix into the
+multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that
+\(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit
+norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and
+\(\pmb{R}\) is an upper triangular matrix. Matrix \(\pmb{Q}\) satisfies
+\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that
+the pairwise Euclidean distances between data points in
+\(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
+Therefore, we use matrix \(Q\) as the rotation matrix for where the
+rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension
+\(N \times 100\) is now the input data for the manifold learning
+algorithms. Again, we set the embedding dimension to be equal to the
+meta data dimension \(d=5\).
 
@@
 
-In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
-The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
-We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
-
-For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
-\textcite{Cheng2021-ex} propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
-\textcite{Cheng2021-ex} use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
-By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
-
-In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps have been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
-However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
+In this application, we use the smart meter data from the \emph{CER Smart
+Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in
+Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010. The
+CER dataset\footnote{accessed via the Irish Social Science Data Archive -
+  www.ucd.ie/issda.} records the half-hourly electricity consumption of
+individual residential and commercial properties, not including energy
+for cooling or heating systems. We selected the \(3,639\) residential data
+with no missing values during the data collection period for a total of
+\(535\) days.
+
+For the electricity consumption data of residential individuals, it
+would be worthwhile to explore the distribution of electricity demand
+rather than the raw consumption data to study the usage patterns of
+different households or different periods or the week \autocite{Hyndman2018-ia}.
+\textcite{Cheng2021-ex} propose two non-Euclidean distance estimators to enable
+manifold learning algorithms in statistical manifolds with each
+observation as a distribution. \textcite{Cheng2021-ex} use the same smart meter
+data for identifying outliers with kernel bandwidth estimation but fail
+to consider the distortion and information loss in the 2-dimensional
+embeddings given that the input data dimension is much higher. By
+introducing the Riemannian metric as the bandwidth matrix, we could take
+into account the distortion in the 2-D embedding and further improve the
+accuracy of the density estimation.
+
+In this section, we first calculated the same empirical distributions of
+the \(336\) half-hourly periods of the week for each household, and apply
+the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the
+statistical manifold learning to get the 2-D embedding of all
+households. Algorithm \ref{alg:vkderm} is then used to obtain density
+estimates with the pointwise variable Riemannian metric as the bandwidth
+matrix and detect outliers. The data processing steps have been clearly
+stated in the application section of \textcite{Cheng2021-ex} and they are skipped
+here. Unlike the simulations in \autoref{simulation}, we know nothing
+about the true density of the electricity distributions for all periods
+of the week and all households, so it is impossible to compare the
+estimated densities with the true meta data density as in
+\autoref{fig:fivedisomapden}. However, we could generate all the density
+estimates with the existing KDE method with fixed bandwidth, which is an
+optimal method for density estimation, and compare the densities from
+our proposed method with them.
 
@@
 
-In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
-
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
-
-There are several open questions to be explored.
-The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-ex}.
-The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
-The scale of distortion in each point in manifold learning could vary a lot.
-If we could smooth across all data points, eg.
-multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
-The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
-The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+In this paper, we propose a new method to estimate the density of
+manifold learning embedding and further identify outliers based on the
+densities. The Riemannian metric measure the direction and angle of the
+distortion when mapping data points through a non-linear function in
+manifold learning algorithms. By introducing the Riemannian metric as
+the pointwise variable bandwidth matrix in kernel density estimation,
+the local distortion in the low-dimensional embedding could be used to
+estimate densities, leading to a more accurate description of the data
+distributions. We compare our proposed method with fixed bandwidth KDE
+by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or
+twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D
+space, and show that variable bandwidth could improve the density
+estimation given a good manifold learning embedding.
+
+As an empirical example, we explore the distributions of different
+households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian
+Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and
+KDE with both variable and fixed bandwidth are used to get the density
+estimates. We compare both density estimates by looking at the
+distributions of the most typical households with the highest densities
+and the most anomalous households with the lowest densities. Both
+methods could identify the typical households with certain usage
+patterns, while the outliers are anomalous in different ways.
+
+There are several open questions to be explored. The first involves the
+selection of tuning parameters for the manifold algorithm so that a
+maximal level of embedding quality is achieved, where embedding quality
+is measured using one of the metrics discussed in the online
+supplementary material of \textcite{Cheng2021-ex}. The scaling of the Riemannian
+metric to get the closest range of the true densities is also worth
+exploring. The scale of distortion in each point in manifold learning
+could vary a lot. If we could smooth across all data points, eg.
+multiply the Riemannian matrix with the ratio between its determinant
+and the sum of all Riemannian matrix determinants, the global density
+estimates could be potentially smoothed. The density estimates on the
+edges of the whole data structure could be improved because most outer
+area points tend to be detected as outliers. The choice of manifold
+learning algorithms also has a large impact on the embedding accuracy,
+which we have seen will affect the density estimation and outlier
+detection. However, the outperformance of VKDE with Riemannian bandwidth
+than the fixed bandwidth has been shown in the higher dimensional
+simulation data and the electricity usage data, which are more related
+to real-life data sets.
 
@@
 
-This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
+This research was supported in part by the Monash eResearch Centre and
+eSolutions-Research Support Services through the use of the MonARCH HPC
+Cluster.
 
@@
 
-In topology, a \emph{homeomorphism} is a bijective map between two topological spaces that is continuous in both directions.
-A \emph{Hausdorff space} is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-A \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
-Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
-A collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
-The manifold \(M\) is a \emph{differentiable manifold} if there exists an atlas of \(M\), \(\{U_\alpha, \varphi_\alpha\}\), such that the \emph{transition maps} between any two charts,
-\[
+In topology, a \emph{homeomorphism} is a bijective map between two
+topological spaces that is continuous in both directions. A \emph{Hausdorff
+space} is a topological space where any two distinct points can be
+separated by disjoint neighborhoods. A \(d\)-dimensional (topological)
+\emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where
+the neighborhood \(U\) for each point \(p\) is homeomorphic to an open
+subset \(V\) of the Euclidean space \(\mathbb{R}^d\). Such a homeomorphism
+\(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate)
+\emph{chart}, denoted as \((U, \varphi)\), with the corresponding local
+coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\). A collection of
+charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is
+called an \emph{atlas}, denoted as \(\mathcal{A}\). The manifold \(M\) is a
+\emph{differentiable manifold} if there exists an atlas of \(M\),
+\(\{U_\alpha, \varphi_\alpha\}\), such that the \emph{transition maps} between
+any two charts, \[
 \varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-\]
-are differentiable of class \(C^\infty\) (smooth).
-
-Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\).
-Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
+\] are differentiable of class \(C^\infty\) (smooth).
+
+Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\). Then
+\(\varphi\) is an embedding of \(E\) into \(M\) if and only if
+\(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\)
+is called an embedded submanifold of \(M\) with the subspace topology.
 
@@
 
-The tangent vector at point \(p\) can be intuitively viewed as the velocity of a curve passing through the point \(p\) or as the directional derivatives at \(p\).
-Here we define the tangent vector via the velocity of curves.
-
-For any \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
-\(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
-\[
+The tangent vector at point \(p\) can be intuitively viewed as the
+velocity of a curve passing through the point \(p\) or as the directional
+derivatives at \(p\). Here we define the tangent vector via the velocity
+of curves.
+
+For any \(p \in M\), let
+\(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and
+\(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve
+through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\). \(\gamma_1\) and
+\(\gamma_2\) are \emph{equivalent} if and only if there exists a chart,
+\((U,\varphi)\), at \(p\) such that \[
 (\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-\]
-A \emph{tangent vector} to a manifold \(M\) at point \(p\), \(v_p\), is any equivalent class of the differentiable curves initialized at \(p\).
-The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\).
-The tangent space is a vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the chart \(\varphi\) locally at \(p\).
-The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
-Tangent vectors can also be seen as the directional derivatives at \(p\).
-For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space.
-
-The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector} \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
+\] A \emph{tangent vector} to a manifold \(M\) at point \(p\), \(v_p\), is any
+equivalent class of the differentiable curves initialized at \(p\). The
+set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at
+\(p\), denoted as \(T_pM\). The tangent space is a vector space of dimension
+\(d\), equal to the dimension of \(M\), and it does not depend on the chart
+\(\varphi\) locally at \(p\). The collection of all tangent spaces defines
+the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\). Tangent vectors can also
+be seen as the directional derivatives at \(p\). For a given coordinate
+chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial
+derivatives are denoted as
+\(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\),
+which define a \emph{basis} of the tangent space.
+
+The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the
+\emph{cotangent space} with corresponding \emph{cotangent vector}
+\(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as
+\(dx_1(p),\dots,dx_d(p)\).
 
@@
 
-A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is \(d\times d\) symmetric positive definite and varies smoothly at \(p\).
-Generally, we omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
-The inner product for two vectors \(u, v \in T_pM\) is written as \(<u, v>_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
-A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
-
-The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
-\(g\) could also be used to define the line element \(dl^2 = g_{ij}dx_idx_j\) and the volume element \(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the local coordinates of the chart \((U, \varphi)\).
-For a curve \(\gamma: I \rightarrow M\), the length of the curve is
+A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each
+point \(p\) is a local inner product
+\(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is \(d\times d\)
+symmetric positive definite and varies smoothly at \(p\). Generally, we
+omit the subscript \(p\) and refer to \(g\) as the Riemannian metric. The
+inner product for two vectors \(u, v \in T_pM\) is written as
+\(<u, v>_g = g_{ij}u^iv^j\) using the Einstein summation convention where
+implicit summation over all indices, \(\sum_{i,j}\), is assumed. A
+differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on
+each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
+
+The Riemannian metric \(g\) can be used to define the norm of a vector
+\(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and
+\(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric
+quantities induced by \(g\). \(g\) could also be used to define the line
+element \(dl^2 = g_{ij}dx_idx_j\) and the volume element
+\(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the
+local coordinates of the chart \((U, \varphi)\). For a curve
+\(\gamma: I \rightarrow M\), the length of the curve is
 \[l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}\],
@@
 Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
-\]
-which is also called the \emph{Riemannian measure} on \(M\).
-
-The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold.
-A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\),
-the second-order ordinary differential equation is satisfied,
-\[
+\] which is also called the \emph{Riemannian measure} on \(M\).
+
+The \emph{geodesics} of \(M\) are the smooth curves that locally joins the
+points along the shortest path on the manifold. A curve
+\(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the
+second-order ordinary differential equation is satisfied, \[
 \frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
-\]
-where \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
-\[
+\] where \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by \[
 \Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
 + \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
-\]
-The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc length functional \(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\) when the curves are defined over the interval \([0,1]\).
-The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
-For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the \emph{injectivity radius} at \(p \in M\) is the distance to the cut locus.
-The injectivity radius of \((M,g)\), \(\textit{inj}_gM\) is the infimum of the injectivity radii over all points on the manifold.
+\] The geodesics have a constant speed with norm
+\(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc
+length functional
+\(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\)
+when the curves are defined over the interval \([0,1]\). The \emph{geodesic
+distance} \(d_g\) is the length of the shortest geodesic between two
+points on the manifold. For a point \(p \in M\), when the geodesic
+distance starting at \(p\) is not minimized, we call such set of points
+the \emph{cut locus} of \(p\), and the \emph{injectivity radius} at \(p \in M\) is the
+distance to the cut locus. The injectivity radius of \((M,g)\),
+\(\textit{inj}_gM\) is the infimum of the injectivity radii over all
+points on the manifold.
 
@@
 
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
-The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
-\[
+Pushforward and pullback are two notions corresponding to the notions of
+tangent and cotangent vectors. Let \(f\) be an embedding from the
+Riemannian manifold \((M,g)\) to another smooth manifold \(E\). The
+pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along
+\(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which
+maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors
+\(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for
+\(u, v \in T_f(p)E\), \[
 <u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-\]
-where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
-The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
-Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
+\] where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\). The tangent vectors
+\(f_*u_p\) are equivalent to the velocity vector of a curve
+\(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a
+constant speed \(\gamma^{-1}(0)=u_p\). Similarly, the pullback maps the
+cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at
+\(p \in M\) acting on tangent vectors \(u \in T_pM\).
 
@@
 
-Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with raidus \(r\).
-Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\).
-Define \(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a differentiable, bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\).
-The exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint after covering the length of \(\|u\|\) along the geodesic in one time unit.
-The inverse of the exponential map is called the \emph{logarithm map}, \(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to \(q\).
-Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
-Then we could interpolate a geodesic \(\gamma\) between two points \(p\) and \(q\) with the exponential map and the logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).
-
-\hypertarget{volumn-density-function}{%
-\subsection{Volumn density function}\label{volumn-density-function}}
-
-For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
-Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
-\[
+Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with
+raidus \(r\). Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open
+neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\). Define
+\(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a
+differentiable, bijective map of differentiable inverse (i.e.
+\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the
+endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying
+\(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\). The
+exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint
+after covering the length of \(\|u\|\) along the geodesic in one time
+unit. The inverse of the exponential map is called the \emph{logarithm map},
+\(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to
+\(q\). Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\)
+as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with
+\(r < \textit{inj}_gM\). Then we could interpolate a geodesic \(\gamma\)
+between two points \(p\) and \(q\) with the exponential map and the
+logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic
+distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).
+
+\hypertarget{volume-density-function}{%
+\subsection{Volume density function}\label{volume-density-function}}
+
+For any \(u \in T_pM\), let
+\(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let
+\(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\),
+for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an
+orthonomal basis of the tangent space \(T_pM\). Define
+\(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function
+on \(M\) as \[
 \theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
 \]
-Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
-The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{vol}\) on \(T_pM\), and
-\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
+Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\). The exponential
+map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form
+\(\exp_p^*\textit{Vol}\) on \(T_pM\), and \(\theta_p\) is its density with
+respect to the Lebesgue measure of the Euclidean structure on \(T_pM\),
+which gives in normal coordinates \[
+d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
+\]
+
+\hypertarget{example-s1}{%
+\subsection{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}
+
+Manifold: \(S^1\) with center 0 and radius \(r=1\)
+
+Polar coordinates: \((r, \theta)\)
+
+Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with
+\(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).
+
+Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and
+\(y = \sin(\theta) \in [0, 1]\)
+
+Density of \(\theta\): \(f(\theta) = 2 / \pi\)
+
+Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)
+
+The CDF of X is given by \[
+P(X \leq x) = P(\cos(\theta) \leq x) \\
+= P(\theta \geq \arccos(x)) \\
+= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+= 1 - \frac{2}{\pi} \arccos(x).
+\] Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
+f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+= \frac{2}{\pi \sqrt{1 - x^2}}.
+\] Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
+estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
+and compare it with its true density \(f(x)\).
+
+We use the density estimator in \textcite{Pelletier2005-vu}. For point
+\(p \in S^1\), the density at \(p\) is estimated by
 \[
-d\exp_p^*vol(u) = \theta_p(u)du.
-\]
-
-\hypertarget{example}{%
-\paragraph{Example}\label{example}}
-
-Manifold: S2
-
-\(\theta \sim U(0, \pi/2)\)
-
-\(X = \cos(\theta)\)
-
-\$\$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
+\] where \(d=1\), \(h\) is the bandwidth,
+\(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\),
+and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along
+the manifold \(S^1\).
+
+Since the manifold is known (\(S^1\)), we know the geodesic distance is
+the arc length between two points on the manifold, which is given by
+\(L(\gamma) = r(\beta - \alpha)\) with \(r=1\). Therefore, the geodesic
+distance is \[
+d_g(p, X_i) = \arccos(p) - \arccos(X_i)
+\] As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal
+coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\)
+equals the determinant of the metric \(g\) expressed in these coordinates
+at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).
+
+{[}When using normal coordinates(orthogonal basis) for tangent vector in
+\(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}
+
+The Riemannian metric \(g\) for \(S^1\) is given by the inner products on
+the tangent spaces of \(S^1\). For each \(X_i \in M\), \[
+g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
+\] is smooth in a local neighborhood \(U\) of \(M\).
+
+Then the volume density function \(\theta_{X_i}(p)\) is the inner product
+of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\),
+which is given by \[
+\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
+\] Therefore, the estimator for \(p \in S^1\) is given by \[
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+\] where the bandwidth \(r\) is selected by the plug-in bandwidth selector
+implemented by the R package \emph{ks}.
+
+\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
+\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}
+
+Example of \(S^2\) in \textcite{Henry2009-om}.
+
+Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let
+\(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\)
+is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential
+chart \((U, \Psi)\) induced by the parametrisation
+\(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by \[
+\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
+\] if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where
+\(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
+Then we have \[
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
+\] Then each component of the riemannian metric could be calculated as
+\(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain
+that the volume density on the sphere is \[
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
+\] for \(q \neq p, -p\), and \(\theta_p(p) = 1\).
+
+In \(S^1\), polar coordinates: \(\phi(\exp_p{vt})^{\prime}(0) = v\)
+
+cartesian coordinates:
+\(\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}\)
+\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
+
+\newpage
 

### 2022-05-05 paper/paper.Rmd
True volume density function to be used, also try small N, optimize h with AMSE

@@
 $r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
-Then we have $$
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
+Then we take the partial derivatives of \$\\Psi\^{-1}(s,t)\$ with
+respect to \$s\$ and \$t\$ and have that$$
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, \\
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
 $$ Then each component of the riemannian metric could be calculated as
@@
 
-In $S^1$, polar coordinates: $\phi(\exp_p{vt})^{\prime}(0) = v$
-
-cartesian coordinates:
-$\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}$
-\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
+In $S^1$, we could let \$\\{v, p\\}\$ be the basis of
+\$\\mathbb{R}\^2\$. Then the exponential chart induced
+
+Polar coordinates: $\phi(\exp_p{vt})^{\prime}(0) = v$
+
+Cartesian coordinates:
+$\psi(\exp_p{vt})^{\prime}(0) = (x_p \cos{v} - \sqrt{1-x_p^2} \sin{v})^\prime  \bigg|_{t=0} = - \sqrt{1 - x_p^2}v$
 

### 2022-05-05 R/riemgeom_s2.R
True volume density function to be used, also try small N, optimize h with AMSE

@@ -5,7 +5,7 @@ library(ks)
 Jmisc::sourceAll(here::here("R/sources"))
 
 set.seed(1)
-N <- 10000
+N <- 1000
 d <- 1
 theta <- runif(N, 0, pi/2)
 # density
@@ -137,6 +137,7 @@ fhat <- rep(0, N)
 h <- h
 for(i in 1:N){
   hk <- rmetric[,,k]
+  # fi <- hk ^ (-1/2) * dnorm(x = acos(fn), mean = acos(fn[i]), sd = h) # true geodesic distance, TODO: true volume density function
   # fi <- hk ^ (-1/2) * dnorm(x = acos(fn), mean = acos(fn[i]), sd = h) # true geodesic distance
   fi <- dnorm(x = fn, mean = fn[i], sd = h * sqrt(hk)) # approximate geodesic distance with inner product adjusted by Riem metric
   # fi <- hk ^ (-1/2) * h ^ (-1) * (2 * pi) ^ (-1/2) * exp(- adj_matrix[,i]^2 / (2 * h ^ 2)) # approximate geodesic distance with shortest path graph distance
@@ -165,7 +166,8 @@ plot(rank(denx), rank(ffixed), main = paste("Rank correlation:", round(cor(rank(
 mean((fhat - dentheta)^2)
 mean((ffixed - dentheta)^2)
 
-
+## reduce N to test
+## optimize h using AMSE for our estimator
 
 
 

### 2022-05-10 paper/paper.Rmd
Finalise example

--- 2022-05-03 paper/paper.Rmd
@@
 $r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
-Then we have $$
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
-$$ Then each component of the riemannian metric could be calculated as
+Then we take the partial derivatives of \$\\Psi\^{-1}(s,t)\$ with
+respect to \$s\$ and \$t\$ and have that
+$$
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
+$$
+$$
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
+$$ 
+Then each component of the riemannian metric could be calculated as
 $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$. We obtain
@@
 
-In $S^1$, polar coordinates: $\phi(\exp_p{vt})^{\prime}(0) = v$
-
-cartesian coordinates:
-$\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}$
-\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
-
-\newpage
+Similarly in $S^1$, we could let $\{v, p\}$ be the basis of
+$\mathbb{R}^2$. Then the exponential chart induced
+
+Polar coordinates:
+$\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v$.
+
+Cartesian coordinates:
+$$
+\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
+= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
+= - \sqrt{1 - p^2}v
+$$
+Then the volume density function is
+$$
+\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
+$$
+and the geodesic distance is
+$$
+d_g(p,q) = \arccos(p) - \arccos(q).
+$$

### 2022-05-10 R/riemgeom_s2.R
Add nn output

@@ -4,15 +4,17 @@ library(dimRed)
 library(ks)
 Jmisc::sourceAll(here::here("R/sources"))
 
-set.seed(1)
-N <- 1000
+set.seed(123)
+N <- 100
 d <- 1
 theta <- runif(N, 0, pi/2)
 # density
 dentheta <- dunif(theta, 0, pi/2) # density = 1/(MAX-MIN) = 0.6366198
-# theta <- rnorm(N, pi / 4, pi / 20)
-# dentheta <- dnorm(theta, pi / 4, pi / 20)
-# denx <- 1 / sqrt(1 - x ^ 2) * dnorm(theta, pi / 4, pi / 20)
+theta <- rnorm(N, 0, pi / 20)
+theta <- exp(theta)
+hist(theta)
+dentheta <- dnorm(theta, pi / 4, pi / 20)
+denx <- 1 / sqrt(1 - x ^ 2) * dnorm(theta, pi / 4, pi / 20)
 x <- cos(theta)
 y <- sin(theta)
 plot(x, y, asp = 1)
@@ -127,6 +129,7 @@ riem.scale <- 1
 adj_matrix <- metric_isomap$adj_matrix
 dim(adj_matrix)
 summary(metric_isomap)
+nn.idx <- metric_isomap$nn2res$nn.idx
 
 # geodesic distance estimation
 # Option 1: true geodesic distance d_g
@@ -134,17 +137,27 @@ summary(metric_isomap)
 # Option 3: approximate geodesic distance with shortest path graph distance d_G
 d <- 1
 fhat <- rep(0, N)
-h <- h
+h <- h # N=100, h=0.147
+# h <- 0.35
+hidet <- apply(rmetric, 3, det) %>% mean()
+# hidetmean <- mean(hidet)
 for(i in 1:N){
-  hk <- rmetric[,,k]
-  # fi <- hk ^ (-1/2) * dnorm(x = acos(fn), mean = acos(fn[i]), sd = h) # true geodesic distance, TODO: true volume density function
-  # fi <- hk ^ (-1/2) * dnorm(x = acos(fn), mean = acos(fn[i]), sd = h) # true geodesic distance
-  fi <- dnorm(x = fn, mean = fn[i], sd = h * sqrt(hk)) # approximate geodesic distance with inner product adjusted by Riem metric
[…]

### 2022-05-10 R/sources/metricML.R
Add nn output

@@ -135,7 +135,8 @@ metricML <- function(x, s = 2, k = min(10, nrow(x)), radius = 0,
               rmetric=hn, 
               weighted_graph=g,
               adj_matrix=Kn,
-              laplacian=Ln))
+              laplacian=Ln,
+              nn2res = nn2res))
 }
 
 

### 2022-05-15 paper/paper.Rmd
Submit abstract

@@
   email: Anastasios.Panagiotelis@sydney.edu.au
-abstract: "Kernel density estimation based on random samples often involves a crucial bandwidth selection.
-For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation.
-We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities.
-The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates.
-To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
-For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week.
-Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
+abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
+However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
+We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
+The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density.
+A shiny app in R is also developed for various simulation scenarios.
+The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
+This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
 keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"

### 2022-05-15 paper/paper.tex
True volume density function to be used, also try small N, optimize h with AMSE
Finalise example
Submit abstract

@@
 \begin{abstract}
-Kernel density estimation based on random samples often involves a crucial bandwidth selection.
-For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation.
-We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities.
-The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates.
-To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density.
-For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week.
-Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
+Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
+However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
+We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
+The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density.
+A shiny app in R is also developed for various simulation scenarios.
+The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
+This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data.
 \end{abstract}
@@
 \(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
-Then we have \[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
-\] Then each component of the riemannian metric could be calculated as
+Then we take the partial derivatives of \$\textbackslash Psi\^{}\{-1\}(s,t)\$ with
+respect to \$s\$ and \$t\$ and have that
+\[
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
+\]
+\[
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
+\]
+Then each component of the riemannian metric could be calculated as
 \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain
@@
 
-In \(S^1\), polar coordinates: \(\phi(\exp_p{vt})^{\prime}(0) = v\)
-
-cartesian coordinates:
-\(\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}\)
-\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
-
-\newpage
+Similarly in \(S^1\), we could let \(\{v, p\}\) be the basis of
+\(\mathbb{R}^2\). Then the exponential chart induced
+
+Polar coordinates:
+\(\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v\).
+
+Cartesian coordinates:
+\[
+\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
+= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
+= - \sqrt{1 - p^2}v
+\]
+Then the volume density function is
+\[
+\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
+\]
+and the geodesic distance is
+\[
+d_g(p,q) = \arccos(p) - \arccos(q).
+\]
 

### 2022-05-22 paper/paper.Rmd
Remove spaces

@@
 
-Anomaly detection has been an important and diverse area where anomalies
-or outliers are detected in a given data. It often involves a data
-analysis process to uncover the unusual patterns and has been widely
-applied to machine learning [@Omar2013-lz], network intrusions
-identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging
-[@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security
-[@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which
-explored detecting households with anomalous electricity usage
-distributions instead of raw data. In the case of non-Euclidean sample
-space, the observations lie on a Riemannian manifold embedded in a very
-high-dimensional ambient space, which makes it computationally expensive
-or impossible to detect anomalies. To address this problem, we propose a
-kernel density estimator of the low-dimensional statistical manifold
-embedding and find outliers as the data with the lowest density
-estimates.
-
-Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of
-the most popular methods to calculate the probability density function
-from a sample dataset. KDE is flexible to learn the shape of the
-underlying density from the data controlled by the bandwidth and the
-selection of bandwidth is crucial in KDE [@Jones1990-oe;
-@Terrell1992-ut]. There is extensive research on bandwidth selection,
-two main categories of which are cross-validation [@Jones1992-ta;
-@Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp]. For
-multivariate data, variable kernel density estimator [VKDE;
-@Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive
-bandwidth matrix to control the amount of smoothing on the location of
-the estimated point [balloon estimator; @Terrell1992-ef] or the data
-point [sample smoothing estimator; @Terrell1992-ef]. However, these
-density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure
-called the Riemannian manifold, @Pelletier2005-vu proposed a kernel
-density estimator based on the kernel weight depending on the distance
-between the estimated points and the sample data points. The idea of the
-estimator is to use a strictly positive function of the geodesic
-distance on the manifold and then normalize it with the volume density
-function of the Riemannian manifold for curvature. [@Henry2009-ll] In
-many application scenarios, we tend to find that the sample points on a
-manifold are embedded in a much higher-dimensional space, and the kernel
-density estimator from @Pelletier2005-vu is not directly applicable.
-This is when we introduce manifold learning to reduce the input data
-dimension. For these high-dimensional data set, various manifold
-learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE,
-and UMAP (see details of these algorithms in @Cheng2021-ex), are applied
-to get a low-dimensional embedding. We propose a kernel density
-estimator to be applied to the low-dimensional embedding and define
-outliers as those with the lowest densities.
-
-In manifold learning, the underlying idea is that the data lies on a
-low-dimensional smooth manifold that is embedded in a high-dimensional
-space. One of the fundamental objectives of manifold learning is to
-explore the geometry of the dataset, including the distances between
-points and volumes of regions of data. These intrinsic geometric
-attributes of the data, such as distances, angles, and areas, however,
-can be distorted in the low-dimensional embedding, leading to failure to
-recover the geometry of the manifold [@Goldberg2008-co]. To tackle this
-problem and measure the distortion incurred in manifold learning,
-@Perrault-Joncas2013-pq propose the Learn metric algorithm to augment
-any existing embedding output with geometric information in the
-Riemannian metric of the manifold itself. By applying this algorithm,
-the outputs of different manifold learning methods can be unified and
-compared under the same framework, which would highly benefit in
-improving the effectiveness of the embedding.
-
-The Riemannian metric using the method of @Perrault-Joncas2013-pq gives
-some idea of the distortion of an embedding. Mapping the points through
-a non-linear function "stretches" some regions of space and "shrinks"
-others. The Riemannian gives us an idea of the direction and angle of
-this stretching, which is informative for learning the manifold. In
-variable kernel density estimate, the bandwidth matrix $H$ is also
-defined to control the amount of smoothing for each data point.
-Therefore, we propose to the use Riemannian metric as the bandwidth
-matrix in, we could further get the kernel density estimation of the
-manifold $M$. This kernel density estimate can then be used to produce
-the highest density region plots [@Hyndman1996-lk] for outlier
-visualization.
-
-The rest of the paper is organized as follows. In \autoref{vkde}, we
-present the proposed algorithm to detect anomalies based on variable
-kernel density estimates of manifold embeddings. In this section, we
-provide justification for the use of the Riemannian metric as the
-bandwidth of variable kernel density estimation, including the
-comparison with fixed bandwidth. \autoref{simulation} is composed of two
-simulations with the proposed algorithm; the first deals with
-2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data
-and the second with a 5-D semi-hypersphere mapped in a 100-D space.
-\autoref{application} contains the application to visualize and identify
-anomalies in the Irish smart meter dataset. Conclusions and discussions
-are presented in \autoref{conclusion}. Readers interested in the the
-notions of Riemannian geometry mentioned in this paper could use the
-Appendix in \autoref{riemgeo} as a reference.
+Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
+It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
+In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
+To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
+
+Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset.
+KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
+There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
+For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of
+the estimated point [balloon estimator; @Terrell1992-ef] or the data point [sample smoothing estimator; @Terrell1992-ef].
+However, these density estimators are based on random samples in the Euclidean space.
+
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
+The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
+[@Henry2009-ll] In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from @Pelletier2005-vu is not directly applicable.
+This is when we introduce manifold learning to reduce the input data dimension.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), are applied to get a low-dimensional embedding.
+We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
+
+In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
+One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
+These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co].
+To tackle this problem and measure the distortion incurred in manifold learning,
+@Perrault-Joncas2013-pq propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
+
+The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
+Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
+The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
+In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point.
+Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $M$.
+This kernel density estimate can then be used to produce the highest density region plots[@Hyndman1996-lk] for outlier visualization.
+
+The rest of the paper is organized as follows.
+In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
+In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
+\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
+Conclusions and discussions are presented in \autoref{conclusion}.
+Readers interested in the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
 
@@
 
-In this section, we introduce the proposed method to detect anomalies
-based on the kernel density estimates of manifold learning embeddings
-where the Riemannian matrix is used as the pointwise variable bandwidth
-to measure the direction and angle of the distortion of the
-low-dimensional embeddings. @Perrault-Joncas2013-pq gives us an idea of
-how to measure the direction and angle of the distortion using the
-Riemannian metric which is a positive definite square matrix for each
-data point. To learn the distribution of the low-dimensional embedding,
-we use the kernel density estimation with the bandwidth matrix being the
-Riemannian metric. The outliers could then be defined as the points with
-the lowest density estimates. The proposed schematic is shown in Figure
-\@ref(fig:vkde). The highlighted two steps in Figure \@ref(fig:vkde) are
-the main contributions of this main chapter, replacing the bandwidth
-matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
-variable kernel density estimate, and computing the highest density
-region plots based on the density estimates, $\hat{f}(\pmb{x})$, for
-anomaly detection.
-
-(ref:vkdecaption) The proposed schematic for variable kernel density
-estimation with recovered geometry.
+In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
+@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
+To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
+The outliers could then be defined as the points with the lowest density estimates.
+The proposed schematic is shown in Figure \@ref(fig:vkde).
+The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
+variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
+
+(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
 
@@
 
-To start with, we introduce the notations used in this manuscript. Then
-we introduce the multivariate kernel density estimation method with
-variable bandwidth matrix and the metric learning algorithm to derive
-the pointwise Riemannian metric. Readers familiar with these topics
-could skip the corresponding subsections. Finally, we propose our novel
-algorithm to estimate densities and detect anomalies for
-high-dimensional data set.
+To start with, we introduce the notations used in this manuscript.
+Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
+Readers familiar with these topics could skip the corresponding subsections.
+Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
 
@@
 
-In general, given a multivariate random sample
-$\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$,
-$i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel
-density estimate of $f$ at point
-$\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
+In general, given a multivariate random sample $\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
 \begin{equation}
@@
 \hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
-\end{equation} where $\pmb{H}_i$ is called the bandwidth matrix which is
-a $d\times d$ symmetric positive definite matrix variable at each $i$,
-$K(\pmb{y})$ is the kernel function, and
-$K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
-When using the standard gaussian kernel
-$K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
+\end{equation} 
+where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{y})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
+When using the standard gaussian kernel, $K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
 \begin{equation}
@@
 K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
-\end{equation} with $|\pmb{H}_i|$ being the determinant of matrix
-$\pmb{H}_i$.
-
-The bandwidth matrix is important in controlling the smoothing across
-all data points and the shape is determined by the kernel function
-$K(\pmb{y})$. We consider the sample smooth estimator for each data
-point, which gives a kernel density estimate with local smoothing.
-@Duong2007-up implements the kernel density estimates for data from 1 to
-6 dimensions in the R package *ks*. However, for variable kernel density
-estimation where the bandwidth is no longer a constant matrix, the
-implementation is limited to 2-dimension and the bandwidth can either be
-a vector or a diagonal matrix instead of a full bandwidth matrix.
-Further, the estimator in Equation \@ref(eq:kde) is only applicable when
-the sample space is Euclidean. In the cases of a Riemannian manifold,
-the sample space used for kernel density estimation is no longer
-Euclidean and we need to use the estimator described in
-\autoref{riemest}.
+\end{equation} 
+with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
+
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{y})$.
+We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
+@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
+Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean.
+In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
 
@@
 
-Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without
-boundary and a probability distribution with density $f$ on the
-manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is
-the Riemannian distance induced by the Riemannian, $g$, and the strictly
-positive injectivity radius [@Chavel2006-mp] of the manifold,
-$inj_g(M)$. Denote $A_i, i = 1, 2, \dots, N$, where
-$\pmb{A}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with
-density $f$. For each $p\in M$, @Pelletier2005-vu proposed the kernel
-density estimator of $f$ to be \begin{equation}
+Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
+Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$.
+Denote $A_i, i = 1, 2, \dots, N$, where $\pmb{A}_i \in \mathcal{R}^d$, as i.i.d.
+random objects on $M$ with density $f$.
+For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
+\begin{equation}
 \label{eq:denriem}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
-\end{equation} where $h$ is the bandwidth satisfying $h \leq h_0$ for
-fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$,
-$K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and
-$\theta_{q}(p)$ is the volume density function on the manifold.
-@Pelletier2005-vu mention that the volume density function $\theta_q(p)$
-defined for $p$ in a neighborhood of $q$ on the manifold, with the
-geodesic normal coordinates at $q$, is equal to the determinant of the
-Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$. The
-expression in \@ref(eq:denriem) is also proved to be consistent with the
-kernel density estimators in the Euclidean case as \begin{equation}
+\end{equation} 
+where $h$ is the bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
+@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$.
+The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
+\begin{equation}
 \label{eq:denconsist}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
-\end{equation} and converges at the same rate as the Euclidean kernel
-density estimator [@Henry2009-ll].
-
-However, in application, we usually found the observations lie on a
-$d$-dimensional manifold embedded in a $s$-dimensional space, where
-$d \ll s$. Then manifold learning can be applies to find the
-$d$-dimensional representation of the manifold $M$. Therefore, based on
-the estimator in \@ref(eq:denriem), we propose to a kernel density
-estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to
-be \begin{equation}
+\end{equation} 
+and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].
+
+However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
+Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
+Therefore, based on the estimator in \@ref(eq:denriem), we propose to a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
+\begin{equation}
 \label{eq:denestimator}
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
-\end{equation} where $h_i, i = 1, 2, \dots, N$ is the pointwise variable
-bandwidth matrix and $h_i$ is positive definite with dimension
+\end{equation} 
+where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension
 $d\times d$.
 
-An important goal of manifold learning is to recover the local or global
-features of the data. In order to recover the geometry of the manifold,
-@Perrault-Joncas2013-pq propose a method to augment any existing
-reasonable embedding and allow for the computation of geometric values
-to be calculated with an estimation of the Riemannian metric $g$. This
-could also be used to fix the geometric distortion in the embedding. It
-is worth noticing that the Riemannian metric from
-@Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for
-each data point. Therefore, we propose to take the geometric distortion
-into consideration when estimating the kernel densities of the embedding
-with Equation \@ref(eq:denestimator). In this way, we present a
-consistent kernel density estimator with pointwise variable bandwidth as
-the Riemannian metric.
+An important goal of manifold learning is to recover the local or global features of the data.
+In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric $g$.
+This could also be used to fix the geometric distortion in the embedding.
+It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for each data point.
+Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \@ref(eq:denestimator).
+In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
 
@@
 
-The Riemannian metric $g$ is a symmetric and positive definite tensor
-field which defines an inner product $<,>_g$ on the tangent space $T_pM$
-for every point $p \in M$. If the inner product of the tangent space is
-known for a given geometry, the Riemannian metric is a good measure to
-recover the geometry of the manifold. The Metric Learning algorithm
-[@Perrault-Joncas2013-pq] then augment the embedded manifold with the
-Riemannian metric and produce a Riemannian manifold $(M, g)$.
-
-To recover the original geometry of the manifold, we need to know what
-the inner product corresponds to in the embedding. The inner product
-between two vectors $u,v \in T_pM$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be
-used to define some geometric quantities, such as the vector norm
-$\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors
-$\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
-Therefore, for each point $p\in M$ in any coordinate system, the
-Riemannian metric $g$ is a $d\times d$ symmetric positive definite
-matrix, where $d$ is the dimension of the manifold.
-
-[^1]: Here the Einstein notation is used where superscripts denote
-    summation over $i$ and $j$
-
-The line element and volume element of the full manifold or a subset of
-the manifold can also be computed from $g$. The arc length of a curve
-$c\in M$ is defined as $$
+The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_pM$ for every point $p \in M$.
+If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
+The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(M, g)$.
+
+To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
+The inner product between two vectors $u,v \in T_pM$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors $\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
+Therefore, for each point $p\in M$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
+
+[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$
+
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in M$ is defined as 
+$$
 l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-$$ where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and
-$c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of
-$V\subset M$ is computed by $$
+$$ 
+where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
+$$
 Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
-$$ Both the concepts of distance and volume are relevant to kernel
-density estimation.
-
-@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly
-involves four steps: weighted neighborhood graph construction; geometric
-graph Laplacian calculation; manifold learning; and Riemannian metric
-calculation. We restate these four steps in Algorithm
-\@ref(alg:learnmetric). By applying an existing manifold learning
-algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a
-low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed. Most
-manifold learning methods involve the construction of the nearest
-neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$
-is built. The Laplacian is quite useful because it can be
-coordinate-free while containing all the important geometry.
-@Perrault-Joncas2013-pq have stated one way to compute the approximated
-$\Delta_M$ with a discrete consistent estimator, the geometric graph
-Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where
-$\varepsilon$ is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the
-Metric Learning algorithm to achieve the augmented embedding with the
-Riemannian metric $(f_n, g_n)$.
-
-As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension
-$s$ is larger than the manifold intrinsic dimension $d$, the rank of the
-embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$
-will be returned. This algorithm is also implemented in a Python library
-*megaman* [@McQueen2016-xz]. It is designed to apply the manifold
-learning methods to large-scale data sets, as well as compute the
-Riemannian metric of the manifold.
-
-```{=tex}
+$$ 
+Both the concepts of distance and volume are relevant to kernel density estimation.
+
+@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
+We restate these four steps in Algorithm \@ref(alg:learnmetric).
+By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
+Most manifold learning methods involve the construction of the nearest
+neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built. The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.
+
+As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension $s$ is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$ will be returned.
+This algorithm is also implemented in a Python library *megaman* [@McQueen2016-xz].
+It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
+
+<!-- ```{=tex} -->
 \begin{algorithm}[!htb]
@@
 \end{algorithm}
-```
+
 # Outlier detection on manifold learning embedding
 
-Now we present our proposed algorithm for anomaly detection based on
-variable kernel density estimates in \@ref(alg:vkderm). There are mainly
-four steps involved in the algorithm. The first step is to apply the
-Learn metric algorithm described in \@ref(alg:learnmetric) to the input
-high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to
-get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$
-and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for
-each observation. Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$
-are given for Gaussian kernels. Then we could use the pointwise
-Riemannian metric to calculate the kernel density estimate with our
-proposed estimator in Equation \@ref(eq:denestimator). The top outliers
-of size $n\_outliers$ are obtained by ordering the embedding points
-$y_i$ according to their density estimates $f_N(\pmb{y}_i)$.
-
-```{=tex}
+Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \@ref(alg:vkderm).
+There are mainly four steps involved in the algorithm.
+The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation.
+Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels.
+Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator).
+The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f_N(\pmb{y}_i)$.
+
 \begin{algorithm}[!htb]
@@
   \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-
   \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
-
   \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \@ref(equ:denestimator);
-
   \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
-
   \end{algorithmic}
 \end{algorithm}
-```
-If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could
-be used to identify the relative location of outliers in the embedding
-and find which observations lie in the highest density region of
-specified coverage, eg. 1%, 50%, 99%, \>99%.
+
+If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
+1%, 50%, 99%, \>99%.
 
@@
 
-In this section, we examine two scenarios for both low and high
-dimensions to test our proposed algorithm. For visualization purposes,
-\autoref{twodgaussian} presents a 2-D meta data example. We first
-simulate the data of size $N=2,000$ from a mixture of four Gaussian
-kernels with the same covariance but different means, each consisting of
-$500$ points. Different mapping functions are then applied to the 2-D
-meta data to be mapped in a 3-D feature space, which gives the
-higher-dimensional input for different manifold learning algorithms,
-including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The
-embedded dimension is set as $d=2$, the same as the meta data dimension.
-This enables us to compare the manifold learning embedding with the true
-meta data. We could now apply Algorithm \@ref(alg:vkderm) to get the
-density estimates of all data points and further detect anomalies. As a
-high-dimensional example, the second simulation in
-\autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$
-embedded in a 100-D space and the corresponding embedding dimension is
-$d=5$.
+In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
+For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
+We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
+Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+The embedded dimension is set as $d=2$, the same as the meta data dimension.
+This enables us to compare the manifold learning embedding with the true meta data.
+We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
 
@@
 
-We first generate a 2-dimensional data of size $N=2000$ from a Gaussian
-mixture model with four components with different means
-$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$
-and the same variance-covariance matrix
-$\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$. The mixture proportions
+We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with different means
+$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$.
+The mixture proportions
 are equally set as $\pi_i=0.25, i=1,2,3,4$.
 <!-- $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$. -->
-Then the mixture Gaussian mixture density function is a weighted linear
-combination of the four component Gaussian densities as \begin{equation}
+Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as 
+\begin{equation}
 \label{eq:gmm}
 P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
-\end{equation} \autoref{fig:metadensity} shows the 2-dimensional meta
-data and the colors indicate the true density of all data points
-calculated from \@ref(eq:gmm), with brighter colors showing higher
-densities and darker colors showing lower densities. We then define
-outliers as points with lowest densities shown in black and typical
-points with highest densities shown in yellow. Based on the true density
-plot, the outliers are scattered in the middle and the outer area of the
-whole structure, while typical points are near the means of four
-kernels. These are *true outliers* to be compared with outliers from the
+\end{equation}
+
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities.
+We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
+Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
+These are *true outliers* to be compared with outliers from the
 kernel density estimates.
@@
 
-Given the 2-D meta data, multiple mapping functions could be applied to
-embed the data in a 3-D space. One of the most famous examples in
-manifold learning is the swiss roll data, with the mapping function in
-\@ref(eq:swissroll). The two-dimensional meta data
-$(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the
-three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in
-the left plot of \autoref{fig:mappings}. The four colors in the mappings
-represent the four Gaussian kernels used to generate the meta data
-$(\pmb{X}_1, \pmb{X}_2)^\prime$. \begin{equation}
+Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
+The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
+The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
+\begin{equation}
 \label{eq:swissroll}
@@
 
-Now we are able to apply different manifold learning algorithms to
-$(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to
-$d=2$, and further estimate the density of the 2-D embedding. According
-to the density estimates, we could rank the data points and then
-identify which observations lie in the highest density region of
-specified coverage, eg. 1%, 5%, 50%, 99%, \>99%. For each of the five
-manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps,
-t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding
-plot in the same row, with the colors indicating the densities levels,
-the left column for true densities from the Gaussian mixture model, the
-middle column for highest density region plots with densities from our
-proposed variable KDE method, and the right for similar HDR plots with
-densities from KDE with fixed bandwidth. The top twenty outliers with
-the lowest densities are highlighted in black with point indexes in
-blue. From \autoref{fig:metadensity} and the data generating process, we
-know that there are four highest density regions. However, in all
-manifold learning embeddings colored with true densities (left column in
-\autoref{fig:sroutliers} ), except for LLE, the number of highest
-density regions are not the same as the meta data. When comparing the
-number of HDRs for variable and fixed bandwidth (middle and right column
-in \autoref{fig:sroutliers} ), our proposed method with variable
-bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian
-Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the
-top 20 outliers found rowwise, variable bandwidth could find most
-outliers lying on the left area of the embedding in ISOMAP and UMAP, and
-both methods in LLE embedding could find the outliers in the outer area,
-but for the other methods, both variable and fixed bandwidth are not
-detected true outliers accurately. For t-SNE and UMAP embedding, the
-embedding structure is highly distorted and the points are clustered
-together in a discontinuous way, which is also shown in the clustered
-outliers.
+Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
+According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
+1%, 5%, 50%, 99%, \>99%.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
+the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
+From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
+However, in all
+manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
+For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
 
@@
 
-To further compare the accuracy of the estimated densities for all data
-points, we calculate the correlation between the rank of true densities
-and the estimated densities and present in \autoref{tab:srcors}. It can
-be seen that the rank correlation of our proposed method with variable
-bandwidth is higher for LLE and UMAP, although the correlation for UMAP
-is very close to zero. The highest correlation comes from our method in
-LLE embedding, which is mainly due to it being closest to the
-rectangular structure of the meta data shown in
-\autoref{fig:metadensity}. For Laplacian Eigenmaps, our method has
-wrongly estimated the left area with lower densities even though their
-true densities are very high in yellow, leading to a negative
-correlation. The negative correlation would occur typically when the
-highest or lowest true density areas are not well estimated. As for the
-estimates in highly distorted embedding, including ISOMAP, t-SNE, and
-UMAP, the rank correlations are quite low. This shows that our proposed
-method could improve the kernel density estimate of manifold learning
-embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg. ISOMAP, or when the
-embedding is discontinuous, eg. t-SNE and UMAP, the density estimates
-are not as reliable for outlier detection.
+To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
+It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
+rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
+correlation.
+The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
+As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
+This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
 
@@
 
-For comparison, we use the same 2-D meta data in
-\autoref{fig:metadensity} with a different mapping function, twin peaks
-mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D
-structure shown in the right plot of \autoref{fig:mappings}.
+For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
+mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 
-Similar to \autoref{fig:sroutliers}, different manifold learning
-embeddings are obtained and used to detect outliers with true densities
-and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
-In general, the four highest density regions in yellow are identified in
-almost all manifold learning embeddings except for ISOMAP with fixed
-bandwidth and t-SNE. For ISOMAP, our proposed variable KDE, compared
-with the true density, gives the most accurate mixture kernel structure
-with the lowest estimated densities (darker colored points) in the
-outside and the center of the embedding, and the kernel means (yellow
-points) with highest densities are also clearly identified. In contrast,
-the fixed bandwidth KDE failed to identify the lowest density area in
-the center. Both variable and fixed bandwidth KDE are quite close with
-LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the
-top outliers are indexed in the middle instead of the true outer areas
-due to the large distortion in the middle. For t-SNE and UMAP, there are
-four clusters in the embedding and UMAP does a better job in finding the
-HDRs than t-SNE. Also due to the clusters in the embedding, the outliers
+Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
+In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
+For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with the lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
+In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
+Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
+For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
+Also due to the clusters in the embedding, the outliers
 found in UMAP are clustered.
@@
 
-We can gain further insight by comparing the correlation between ranks
-of true densities and estimated densities from variable and fixed
-bandwidth KDE by \autoref{tab:tpcors}. Again the highest correlations
-appear from embedding with higher quality, including ISOMAP, Laplacian
-Eigenmaps, and UMAP. The rank correlations between variable and fixed
-bandwidth are equivalent to the third decimal place in Laplacian
-Eigemaps and UMAP. As for t-SNE, the four clusters in the embedding are
-less separated than in UMAP and our proposed method has misidentified
-the kernel cores, leading to a lower rank correlation in variable
-bandwidth. Since the embeddings from twin peak data generally capture
-the rectangular structure in the meta data than those from the swiss
-roll data, the rank correlations are much higher in \autoref{tab:tpcors}
-than in \autoref{tab:srcors}, with the lowest correlation being
-`r round(min(cors), 3)`. This again suggests that the accuracy of
-outlier detection is highly related to the quality of manifold learning
-embedding.
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
+Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
+The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
+As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
+Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being `r round(min(cors), 3)`.
+This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
 
@@
 
-In \autoref{fig:tpisomapden}, we plot the estimated density against the
-true density of the ISOMAP embedding for KDE with variable and fixed
-bandwidth, with colors and shapes showing the four kernels in the meta
-data. The linear positive relationship between the true densities and
-variable KDEs on the left handside is stronger than that of the fixed
-bandwidth KDEs. Combined with the top-right subplot in
-\autoref{fig:tpoutliers}, we could tell that most points are
-underestimated near the true kernel cores, which also suggests that the
-fixed bandwidth tries to smooth across all the data points and fails to
-fix the local distortions in the manifold learning process like the
-proposed pointwise variable bandwidth.
+In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
+The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
+Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to
+fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
 
@@
 
-As a high-dimensional experiment, we generate the meta data from a
-5-dimensional semi-hypersphere, transform it into a 100-dimensional
-space, and then embed it in $d=5$ with manifold learning. First, we
-simulate $N=2,000$ points,
-$(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a
-4-dimensional Gaussian mixture model with two mixture components,
-$\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where
-$\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$,
-$\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
-In order to manually add anomalies to be distant points from the means,
-the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$. The
-fifth dimension is calculated to satisfy the five-dimensional
-semi-hypersphere surface equation
-$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is
-set as $7$. The Gaussian mixture densities could be calculated using
-Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D
-tour path with the R package *tourr* [REFERENCE]. The round and
-triangular point shapes index the two mixture components
-$\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the
-distance between the simulated $4-D$ data point from Gaussian mixture
-model and the kernel means $(0, 0, 0, 0)^\prime$. The more distant from
-the point to the kernel cores, the lower the true densities, which shows
-in a darker color in \autoref{fig:fivedmeta}. It can be seen that the
-most distant points are in a triangular shape, meaning that they are
-simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$. The dark
-colors also indicate that they are the true outliers because of their
-low densities.
-
-Then we initial the other 95 dimensions in the high-dimensional space as
-zero columns and further rotate the $100$-dimensional data of size $N$
-(denote the transpose of the data matrix as $\pmb{X}_0$ with dimension
-$100 \times N$) to get rid of the zeros so that it could be passed to
-the manifold learning algorithms. The rotation matrix is derived from
-the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all
-components randomly generated from a uniform distribution
-$\mathcal{U}(0,1)$. For any real matrix $\pmb{A}$ of dimension
-$p\times q$, the QR decomposition could decompose the matrix into the
-multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that
-$\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit
-norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and
-$\pmb{R}$ is an upper triangular matrix. Matrix $\pmb{Q}$ satisfies
-$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that
-the pairwise Euclidean distances between data points in
+As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in $d=5$ with manifold learning.
+First, we simulate $N=2,000$ points,
+$(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a 4-dimensional Gaussian mixture model with two mixture components,
+$\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
+In order to manually add anomalies to be distant points from the means, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
+The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation
+$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$.
+The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE].
+The round and triangular point shapes index the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
+$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the distance between the simulated $4-D$ data point from Gaussian mixture model and the kernel means $(0, 0, 0, 0)^\prime$.
+The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
+It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
+The dark colors also indicate that they are the true outliers because of their low densities.
+
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the $100$-dimensional data of size $N$(denote the transpose of the data matrix as $\pmb{X}_0$ with dimension $100 \times N$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$.
+For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix.
+Matrix $\pmb{Q}$ satisfies
+$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in
 $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
-Therefore, we use matrix $Q$ as the rotation matrix for where the
-rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension
-$N \times 100$ is now the input data for the manifold learning
-algorithms. Again, we set the embedding dimension to be equal to the
-meta data dimension $d=5$.
+Therefore, we use matrix $Q$ as the rotation matrix for where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms.
+Again, we set the embedding dimension to be equal to the meta data dimension $d=5$.
 
@@
 
-In \autoref{fig:fivedisomapden}, the estimated densities are compared
-with the true density on the x-axis for four manifold learning
-embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP. Note that we
-exclude t-SNE algorithm in this section because it is designed mainly
-for low-dimensional visualization purposes, and it is only applicable to
-embedding dimensions within three. Similar to \autoref{fig:fivedmeta},
-the point shapes show the two mixture component in the meta data, and
-the colors represent the distance to the kernel means, with distant
-outliers shown in darker colors. For well-estimated densities, the true
-outliers with low true densities will also have low estimated densities,
-which suggests that darker-colored points should appear in the
-bottom-left corner in \autoref{fig:fivedisomapden}. This is true for
-both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in
-UMAP where these outliers have relatively high density estimates. For
-variable bandwidth KDE, there is a strong positive linear relationship
-with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and
-the relationship is stronger than the fixed bandwidth. This suggests
-that our proposed KDE with variable bandwidth is more accurate than the
-fixed bandwidth in estimating the manifold learning embedding densities.
-In KDE with fixed bandwidth, the bandwidth is often too large to smooth
-across all data points, especially when there is severe distortion in
-the embedding data. By introducing the pointwise variable Riemannian
-metric in kernel density estimation, it is reasonable to believe that it
-could fix the distortion introduced by these three manifold learning
-algorithms.
+In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
+Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
+For well-estimated densities, the true
+outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
+This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
+For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and
+the relationship is stronger than the fixed bandwidth.
+This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
+In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
+By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
 
@@
 
-In this application, we use the smart meter data from the *CER Smart
-Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in
-Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010. The
-CER dataset[^2] records the half-hourly electricity consumption of
-individual residential and commercial properties, not including energy
-for cooling or heating systems. We selected the $3,639$ residential data
-with no missing values during the data collection period for a total of
+In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010.
+The CER dataset[^2] records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
+We selected the $3,639$ residential data with no missing values during the data collection period for a total of
 $535$ days.
 
-[^2]: accessed via the Irish Social Science Data Archive -
-    www.ucd.ie/issda.
-
-For the electricity consumption data of residential individuals, it
-would be worthwhile to explore the distribution of electricity demand
-rather than the raw consumption data to study the usage patterns of
-different households or different periods or the week [@Hyndman2018-ia].
-@Cheng2021-ex propose two non-Euclidean distance estimators to enable
-manifold learning algorithms in statistical manifolds with each
-observation as a distribution. @Cheng2021-ex use the same smart meter
-data for identifying outliers with kernel bandwidth estimation but fail
-to consider the distortion and information loss in the 2-dimensional
-embeddings given that the input data dimension is much higher. By
-introducing the Riemannian metric as the bandwidth matrix, we could take
-into account the distortion in the 2-D embedding and further improve the
-accuracy of the density estimation.
-
-In this section, we first calculated the same empirical distributions of
-the $336$ half-hourly periods of the week for each household, and apply
-the total variation distance estimator proposed in @Cheng2021-ex in the
-statistical manifold learning to get the 2-D embedding of all
-households. Algorithm \@ref(alg:vkderm) is then used to obtain density
-estimates with the pointwise variable Riemannian metric as the bandwidth
-matrix and detect outliers. The data processing steps have been clearly
-stated in the application section of @Cheng2021-ex and they are skipped
-here. Unlike the simulations in \autoref{simulation}, we know nothing
-about the true density of the electricity distributions for all periods
-of the week and all households, so it is impossible to compare the
-estimated densities with the true meta data density as in
-\autoref{fig:fivedisomapden}. However, we could generate all the density
-estimates with the existing KDE method with fixed bandwidth, which is an
-optimal method for density estimation, and compare the densities from
-our proposed method with them.
+[^2]: accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.
+
+For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
+@Cheng2021-ex propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
+@Cheng2021-ex use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
+By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
+
+In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
+Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+The data processing steps have been clearly
+stated in the application section of @Cheng2021-ex and they are skipped here.
+Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
+However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
 
@@
 
-\autoref{fig:electricityoutliers} shows the electricity usage data of
-three households for both density estimation methods respectively, with
-the top one being the most typical household with the highest density
-and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are close except that there are a
-few spikes for the one with fixed bandwidth. As to the anomalies,
-variable bandwidth tends to capture the unusual electricity demand
-volume when the usage is very low or high. It could also capture the
-unusual usage pattern when there are sudden spikes in ID 3243 or very
-high base electricity usage for the middle time periods in ID 2838. In
-contrast, fixed bandwidth KDE is more sensitive to spikes even when the
-spikes happen in a certain time window in 7049, or when the usage has an
-obvious time-of-week pattern with a few low electricity usage periods.
+\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
+The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
+As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
+It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
+In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
 
@@
 
-Further insights could be gained by comparing the quantile region plots
-of electricity demand against the time of the week for the same typical
-or anomalous households in \autoref{fig:electricityhdrsfixed} and
-\autoref{fig:electricityhdrsvkde}. Again the distribution of both
-typical households in the top panel has shown a repeated
-period-of-the-week usage pattern, with higher usage during mealtime on
-all seven days of the week and slightly higher usage for weekends.
-However, this repeated pattern in a week window is clearly for the
-typical household ID 1321. As for the distributions for outliers, the
-middle row outliers from variable bandwidth have spikes only on Tuesday
-and Sunday noons, while the fixed bandwidth has an increasing
-electricity demand across the day of the week. The bottom row outliers
-both have a repeated time usage pattern, but the electricity usage
-amount is higher with the highest median above 3kWh. These findings show
-the difference in finding typical and anomalous households with
-different bandwidth selections.
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
+Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
+However, this repeated pattern in a week window is clearly for the typical household ID 1321.
+As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
+The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
+These findings show the difference in finding typical and anomalous households with different bandwidth selections.
 
@@
 
-In this paper, we propose a new method to estimate the density of
-manifold learning embedding and further identify outliers based on the
-densities. The Riemannian metric measure the direction and angle of the
-distortion when mapping data points through a non-linear function in
-manifold learning algorithms. By introducing the Riemannian metric as
-the pointwise variable bandwidth matrix in kernel density estimation,
-the local distortion in the low-dimensional embedding could be used to
-estimate densities, leading to a more accurate description of the data
-distributions. We compare our proposed method with fixed bandwidth KDE
-by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or
-twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D
-space, and show that variable bandwidth could improve the density
-estimation given a good manifold learning embedding.
-
-As an empirical example, we explore the distributions of different
-households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian
-Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and
-KDE with both variable and fixed bandwidth are used to get the density
-estimates. We compare both density estimates by looking at the
-distributions of the most typical households with the highest densities
-and the most anomalous households with the lowest densities. Both
-methods could identify the typical households with certain usage
-patterns, while the outliers are anomalous in different ways.
-
-There are several open questions to be explored. The first involves the
-selection of tuning parameters for the manifold algorithm so that a
-maximal level of embedding quality is achieved, where embedding quality
-is measured using one of the metrics discussed in the online
-supplementary material of @Cheng2021-ex. The scaling of the Riemannian
-metric to get the closest range of the true densities is also worth
-exploring. The scale of distortion in each point in manifold learning
-could vary a lot. If we could smooth across all data points, eg.
+In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
+The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
+By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+
+As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
+We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
+Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
+
+There are several open questions to be explored.
+The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of @Cheng2021-ex.
+The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
+The scale of distortion in each point in manifold learning could vary a lot.
+If we could smooth across all data points, eg.
 multiply the Riemannian matrix with the ratio between its determinant
-and the sum of all Riemannian matrix determinants, the global density
-estimates could be potentially smoothed. The density estimates on the
-edges of the whole data structure could be improved because most outer
-area points tend to be detected as outliers. The choice of manifold
-learning algorithms also has a large impact on the embedding accuracy,
-which we have seen will affect the density estimation and outlier
-detection. However, the outperformance of VKDE with Riemannian bandwidth
-than the fixed bandwidth has been shown in the higher dimensional
-simulation data and the electricity usage data, which are more related
+and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
+The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
+The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related
 to real-life data sets.
@@
 
-```{=html}
 <!-- The aim of the project is not to evaluate different manifold learning algorithms.
@@
 -->
-```
+
 # Acknowledgment {.unnumbered}
 
-This research was supported in part by the Monash eResearch Centre and
-eSolutions-Research Support Services through the use of the MonARCH HPC
-Cluster.
+This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
 

### 2022-05-22 paper/paper.tex
Format

--- 2022-04-27 paper/paper.tex
@@
 There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
-For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
+For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of
+the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
 However, these density estimators are based on random samples in the Euclidean space.
@@
 These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}.
-To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+To tackle this problem and measure the distortion incurred in manifold learning,
+\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
 By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
@@
 Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(M\).
-This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.
+This kernel density estimate can then be used to produce the highest density region plots\autocite{Hyndman1996-lk} for outlier visualization.
 
@@
 Conclusions and discussions are presented in \autoref{conclusion}.
-Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
+Readers interested in the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
 
@@
 The proposed schematic is shown in Figure \ref{fig:vkde}.
-The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
+The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
+variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
 
@@
 where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
-When using the standard gaussian kernel \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
+When using the standard gaussian kernel, \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
 \begin{equation}
@@
 \end{equation}
-where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
+where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension
+\(d\times d\).
 
@@
 By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+Most manifold learning methods involve the construction of the nearest
+neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built. The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
 \textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_M\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
@@
   \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-
   \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
-
   \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};
-
   \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
-
   \end{algorithmic}
@@
 
-We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
-The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
+We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means
+\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
+The mixture proportions
+are equally set as \(\pi_i=0.25, i=1,2,3,4\).
 Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
@@
 \end{equation}
+
 \autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
@@
 Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
+These are \emph{true outliers} to be compared with outliers from the
+kernel density estimates.
 
@@
 1\%, 5\%, 50\%, 99\%, \textgreater99\%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
+the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
 The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
 From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+However, in all
+manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
 When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately.
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
 For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
@@
 It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative correlation.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
+rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
+correlation.
 The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
@@
 This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg.
-ISOMAP, or when the embedding is discontinuous, eg.
-t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
+However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
 
@@
 
-For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
+For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
+mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
 \begin{equation}
@@
 For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
-Also due to the clusters in the embedding, the outliers found in UMAP are clustered.
+Also due to the clusters in the embedding, the outliers
+found in UMAP are clustered.
 
@@
 The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
-Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to
+fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
 
@@
 As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
-First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
+First, we simulate \(N=2,000\) points,
+\((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components,
+\(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
 In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
+The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation
+\(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
 The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
 \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
-The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
+The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
+\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
 The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
@@
 
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\)(denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
 The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
@@
 Matrix \(\pmb{Q}\) satisfies
-\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
+\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in
+\(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
 Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
@@
 Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
-For well-estimated densities, the true outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
+For well-estimated densities, the true
+outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
 This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
-For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and the relationship is stronger than the fixed bandwidth.
+For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and
+the relationship is stronger than the fixed bandwidth.
 This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
@@
 The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
-We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
+We selected the \(3,639\) residential data with no missing values during the data collection period for a total of
+\(535\) days.
 
@@
 Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps have been clearly stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
+The data processing steps have been clearly
+stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
 Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
@@
 By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
 
@@
 If we could smooth across all data points, eg.
-multiply the Riemannian matrix with the ratio between its determinant and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
+multiply the Riemannian matrix with the ratio between its determinant
+and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
 The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
 The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related
+to real-life data sets.
 
@@
 A \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
-Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
+Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local
+coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
 A collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
@@
 \varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-\]
-are differentiable of class \(C^\infty\) (smooth).
-
-Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\).
-Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
+\] are differentiable of class \(C^\infty\) (smooth).
+
+Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\). Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if
+\(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
 
@@
 Tangent vectors can also be seen as the directional derivatives at \(p\).
-For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space.
-
-The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector} \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
+For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial
+derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space.
+
+The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector}
+\(z_p: T^\star_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
 
@@
 A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is \(d\times d\) symmetric positive definite and varies smoothly at \(p\).
-Generally, we omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
+Generally, we
+omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
 The inner product for two vectors \(u, v \in T_pM\) is written as \(<u, v>_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
@@
 For a curve \(\gamma: I \rightarrow M\), the length of the curve is
-\[l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}\],
+\[
+l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt},
+\]
 where \(\gamma(I) \subset U\). The volume of \(W \subset U\) is defined as
@@
 
-The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold.
-A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\),
-the second-order ordinary differential equation is satisfied,
+The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold. A curve
+\(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
 \[
@@
 \]
-The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc length functional \(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\) when the curves are defined over the interval \([0,1]\).
+The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc
+length functional \(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\) when the curves are defined over the interval \([0,1]\).
 The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
@@
 where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
-The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
+The tangent vectors \(f_\star u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
 Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
@@
 
-\hypertarget{volumn-density-function}{%
-\subsection{Volumn density function}\label{volumn-density-function}}
+\hypertarget{volume-density-function}{%
+\subsection{Volume density function}\label{volume-density-function}}
 
@@
 Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
-The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and
-\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
+The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
 \[
@@
 \hypertarget{example-s1}{%
-\paragraph{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}
+\subsection{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}
 
@@
 \]
-Then the density of X is the derivative of the CDF w.r.t. \(x\)
-\[
+Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
 f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
@@
 \]
-Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\) and compare it with its true density \(f(x)\).
-
-We use the density estimator in \textcite{Pelletier2005-vu}.
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
-\]
-where \(d=2\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
-
-Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\).
+Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
+estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
+and compare it with its true density \(f(x)\).
+
+We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
+\[
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
+\]
+where \(d=1\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
+
+Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\).
 Therefore, the geodesic distance is
@@
 \]
-is smooth in a local neightborhood \(U\) of \(M\).
+is smooth in a local neighborhood \(U\) of \(M\).
 
@@
 \[
-\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
+\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
 \]
@@
 \[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
-^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
-\]
-The bandwidth \(h\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.
-
-\newpage
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+\]
+where the bandwidth \(r\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.
+
+\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
+\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}
+
+Example of \(S^2\) in \textcite{Henry2009-om}.
+
+Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\).
+Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\).
+Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
+\[
+\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
+\]
+if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
+Then we take the partial derivatives of \$\textbackslash Psi\^{}\{-1\}(s,t)\$ with respect to \$s\$ and \$t\$ and have that
+\[
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
+\]
+\[
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
+\]
+Then each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain that the volume density on the sphere is
+\[
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
+\]
+for \(q \neq p, -p\), and \(\theta_p(p) = 1\).
+
+Similarly in \(S^1\), we could let \(\{v, p\}\) be the basis of \(\mathbb{R}^2\). Then the exponential chart induced
+
+Polar coordinates:
+\(\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v\).
+
+Cartesian coordinates:
+\[
+\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
+= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
+= - \sqrt{1 - p^2}v
+\]
+Then the volume density function is
+\[
+\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
+\]
+and the geodesic distance is
+\[
+d_g(p,q) = \arccos(p) - \arccos(q).
+\]
 

### 2022-05-22 paper/paper.Rmd
Format

@@
 
-# Acknowledgment {.unnumbered}
+# Acknowledgment {-}
 
@@
 
-In this appendix, we present some notions about the Riemannian geometry
-used in this paper.
+In this appendix, we present some notions about the Riemannian geometry used in this paper.
 
@@
 
-In topology, a *homeomorphism* is a bijective map between two
-topological spaces that is continuous in both directions. A *Hausdorff
-space* is a topological space where any two distinct points can be
-separated by disjoint neighborhoods. A $d$-dimensional (topological)
-*manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where
-the neighborhood $U$ for each point $p$ is homeomorphic to an open
-subset $V$ of the Euclidean space $\mathbb{R}^d$. Such a homeomorphism
-$\varphi: U \rightarrow V$ together with $U$ gives a (coordinate)
-*chart*, denoted as $(U, \varphi)$, with the corresponding local
-coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$. A collection of
-charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is
-called an *atlas*, denoted as $\mathcal{A}$. The manifold $M$ is a
-*differentiable manifold* if there exists an atlas of $M$,
-$\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between
-any two charts, $$
+In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions.
+A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
+A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
+Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local
+coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
+A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
+The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts,
+$$
 \varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
@@
 
-Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$. Then
-$\varphi$ is an embedding of $E$ into $M$ if and only if
-$\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$
-is called an embedded submanifold of $M$ with the subspace topology.
+Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$. Then $\varphi$ is an embedding of $E$ into $M$ if and only if
+$\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
 
@@
 
-The tangent vector at point $p$ can be intuitively viewed as the
-velocity of a curve passing through the point $p$ or as the directional
-derivatives at $p$. Here we define the tangent vector via the velocity
-of curves.
+The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through the point $p$ or as the directional derivatives at $p$.
+Here we define the tangent vector via the velocity of curves.
 
-For any $p \in M$, let
-$\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and
-$\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve
-through $p$, i.e. $\gamma_1(0) = \gamma_2(0) = p$. $\gamma_1$ and
-$\gamma_2$ are *equivalent* if and only if there exists a chart,
-$(U,\varphi)$, at $p$ such that $$
+For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e.
+$\gamma_1(0) = \gamma_2(0) = p$.
+$\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
+$$
 (\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-$$ A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any
-equivalent class of the differentiable curves initialized at $p$. The
-set of all tangent vectors at $p$ defines the *tangent space* of $M$ at
-$p$, denoted as $T_pM$. The tangent space is a vector space of dimension
-$d$, equal to the dimension of $M$, and it does not depend on the chart
-$\varphi$ locally at $p$. The collection of all tangent spaces defines
-the *tangent bundle* $TM = \cup_{p \in M}T_pM$. Tangent vectors can also
-be seen as the directional derivatives at $p$. For a given coordinate
-chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial
-derivatives are denoted as
-$\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$,
-which define a *basis* of the tangent space.
+$$ 
+A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$.
+The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
+The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
+The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
+Tangent vectors can also be seen as the directional derivatives at $p$.
+For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial
+derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.
 
-The tangent space $T_pM$ also admits a dual space $T^*_pM$ called the
-*cotangent space* with corresponding *cotangent vector*
-$z_p: T^*_pM \rightarrow \mathbb{R}^d$, and basis denoted as
-$dx_1(p),\dots,dx_d(p)$.
+The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with corresponding *cotangent vector*
+$z_p: T^\star_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
 
@@
 
-A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each
-point $p$ is a local inner product
-$T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$
-symmetric positive definite and varies smoothly at $p$. Generally, we
-omit the subscript $p$ and refer to $g$ as the Riemannian metric. The
-inner product for two vectors $u, v \in T_pM$ is written as
-$<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where
-implicit summation over all indices, $\sum_{i,j}$, is assumed. A
-differentiable manifold $M$ endowed with the Riemannian metric $g$ on
-each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
+A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
+Generally, we
+omit the subscript $p$ and refer to $g$ as the Riemannian metric.
+The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed.
+A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
 
-The Riemannian metric $g$ can be used to define the norm of a vector
-$u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and
-$v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric
-quantities induced by $g$. $g$ could also be used to define the line
-element $dl^2 = g_{ij}dx_idx_j$ and the volume element
-$dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the
-local coordinates of the chart $(U, \varphi)$. For a curve
-$\gamma: I \rightarrow M$, the length of the curve is
-$$l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}$$,
+The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
+$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
+For a curve $\gamma: I \rightarrow M$, the length of the curve is
+$$
+l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt},
+$$
 where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
@@
 Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
-$$ which is also called the *Riemannian measure* on $M$.
+$$
+which is also called the *Riemannian measure* on $M$.
 
-The *geodesics* of $M$ are the smooth curves that locally joins the
-points along the shortest path on the manifold. A curve
-$\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the
-second-order ordinary differential equation is satisfied, $$
+The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve
+$\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the second-order ordinary differential equation is satisfied, 
+$$
 \frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
-$$ where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by $$
+$$ 
+where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
+$$
 \Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
 + \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
-$$ The geodesics have a constant speed with norm
-$\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc
-length functional
-$l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$
-when the curves are defined over the interval $[0,1]$. The *geodesic
-distance* $d_g$ is the length of the shortest geodesic between two
-points on the manifold. For a point $p \in M$, when the geodesic
-distance starting at $p$ is not minimized, we call such set of points
-the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the
-distance to the cut locus. The injectivity radius of $(M,g)$,
-$\textit{inj}_gM$ is the infimum of the injectivity radii over all
-points on the manifold.
+$$ 
+The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc
+length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$. 
+The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
+For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
+The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
 
@@
 
-Pushforward and pullback are two notions corresponding to the notions of
-tangent and cotangent vectors. Let $f$ be an embedding from the
-Riemannian manifold $(M,g)$ to another smooth manifold $E$. The
-pushforward $h=\varphi*g$ of the Riemannian metric $g$ along
-$\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which
-maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors
-$f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for
-$u, v \in T_f(p)E$, $$
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
+Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
+The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
+$$
 <u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-$$ where $df_p^{-1}$ is the Jacobian of $f^{-1}$. The tangent vectors
-$f_*u_p$ are equivalent to the velocity vector of a curve
-$\gamma: I\rightarrow M$ passing through point $p$ at time zero with a
-constant speed $\gamma^{-1}(0)=u_p$. Similarly, the pullback maps the
-cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at
-$p \in M$ acting on tangent vectors $u \in T_pM$.
+$$ 
+where $df_p^{-1}$ is the Jacobian of $f^{-1}$.
+The tangent vectors $f_\star u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
+Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
 
@@
 
-Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with
-raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open
-neighborhood of $0_p$ in the tangent space at $p$, $T_pM$. Define
-$exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a
-differentiable, bijective map of differentiable inverse (i.e.
-*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the
-endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying
-$\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$. The
-exponential map moves point $p$ from $p$ at speed $u$ to an endpoint
-after covering the length of $\|u\|$ along the geodesic in one time
-unit. The inverse of the exponential map is called the *logarithm map*,
-$\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to
-$q$. Also define the *geodesic ball* centered at $p$ of radius $r > 0$
-as the image by the exponential map of $B(0_p, r) \subset T_pM$ with
-$r < \textit{inj}_gM$. Then we could interpolate a geodesic $\gamma$
-between two points $p$ and $q$ with the exponential map and the
-logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic
-distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
+Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$.
+Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$.
+Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e.
+*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$.
+The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit.
+The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$.
+Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
+Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
 
@@
 
-For any $u \in T_pM$, let
-$\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let
-$w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$,
-for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an
-orthonomal basis of the tangent space $T_pM$. Define
-$\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function
-on $M$ as $$
+For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
+Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
+$$
 \theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-$$ <!-- $$ -->
+$$ 
+<!-- $$ -->
 <!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
-<!-- $$ --> Rewrite $\theta_p(u) = \theta_p(\log_p(u))$. The exponential
-map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form
-$\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with
-respect to the Lebesgue measure of the Euclidean structure on $T_pM$,
-which gives in normal coordinates $$
+<!-- $$ --> 
+Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
+The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
+$$
 d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
@@
 
-Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with
-$f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
+Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
 
-Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and
-$y = \sin(\theta) \in [0, 1]$
+Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$
 
@@
 
-The CDF of X is given by $$
+The CDF of X is given by 
+$$
 P(X \leq x) = P(\cos(\theta) \leq x) \\
@@
 = 1 - \frac{2}{\pi} \arccos(x).
-$$ Then the density of X is the derivative of the CDF w.r.t. $x$ $$
+$$ 
+Then the density of X is the derivative of the CDF w.r.t. $x$ $$
 f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
 = \frac{2}{\pi \sqrt{1 - x^2}}.
-$$ Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
+$$ 
+Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
 estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
@@
 
-We use the density estimator in @Pelletier2005-vu. For point
-$p \in S^1$, the density at $p$ is estimated by
-<!-- \begin{equation} --> <!-- \label{eq:pelletier} --> $$
+We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
+<!-- \begin{equation} --> 
+<!-- \label{eq:pelletier} --> 
+$$
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
-$$ <!-- \end{equation} --> where $d=1$, $h$ is the bandwidth,
-$\theta_{X_i}(p)$ is the volume density function at sample point $X_i$,
-and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along
-the manifold $S^1$.
+$$ 
+<!-- \end{equation} --> 
+where $d=1$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
 
-Since the manifold is known ($S^1$), we know the geodesic distance is
-the arc length between two points on the manifold, which is given by
-$L(\gamma) = r(\beta - \alpha)$ with $r=1$. Therefore, the geodesic
-distance is $$
+Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
+Therefore, the geodesic distance is
+$$
 d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-$$ As stated in @Pelletier2005-vu, in terms of *geodesic normal
-coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$
-equals the determinant of the metric $g$ expressed in these coordinates
-at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$.
+$$ 
+As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
+$|g(\exp_p^{-1}(x_i))|$.
 
-[When using normal coordinates(orthogonal basis) for tangent vector in
-$T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
+[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
 
-The Riemannian metric $g$ for $S^1$ is given by the inner products on
-the tangent spaces of $S^1$. For each $X_i \in M$, $$
+The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$, 
+$$
 g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-$$ is smooth in a local neighborhood $U$ of $M$.
+$$ 
+is smooth in a local neighborhood $U$ of $M$.
 
-Then the volume density function $\theta_{X_i}(p)$ is the inner product
-of the tangent vectors produced by the logarithm map from $X_i$ to $p$,
-which is given by $$
+Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by 
+$$
 \theta_{X_i}(p) = \sqrt{1 - X_i^2}.
-$$ Therefore, the estimator for $p \in S^1$ is given by $$
+$$ 
+Therefore, the estimator for $p \in S^1$ is given by 
+$$
 f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-$$ where the bandwidth $r$ is selected by the plug-in bandwidth selector
-implemented by the R package *ks*.
+$$ 
+where the bandwidth $r$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.
 
@@
 
-Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let
-$v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$
-is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential
-chart $(U, \Psi)$ induced by the parametrisation
-$\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by $$
+Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
+Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$.
+Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
+$$
 \Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
-$$ if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where
-$r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
-Then we take the partial derivatives of \$\\Psi\^{-1}(s,t)\$ with
-respect to \$s\$ and \$t\$ and have that
+$$ 
+if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
+Then we take the partial derivatives of \$\\Psi\^{-1}(s,t)\$ with respect to \$s\$ and \$t\$ and have that
 $$
@@
 $$ 
-Then each component of the riemannian metric could be calculated as
-$g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$. We obtain
-that the volume density on the sphere is $$
+Then each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$. We obtain that the volume density on the sphere is 
+$$
 \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
 = \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
-$$ for $q \neq p, -p$, and $\theta_p(p) = 1$.
+$$ 
+for $q \neq p, -p$, and $\theta_p(p) = 1$.
 
-Similarly in $S^1$, we could let $\{v, p\}$ be the basis of
-$\mathbb{R}^2$. Then the exponential chart induced
+Similarly in $S^1$, we could let $\{v, p\}$ be the basis of $\mathbb{R}^2$. Then the exponential chart induced
 

### 2022-05-25 paper/paper.Rmd
Test for S1 works

@@
   # correspondingauthor: true
+- familyname: Panagiotelis
+  othernames: Anastasios
+  address: University of Sydney
+  email: Anastasios.Panagiotelis@sydney.edu.au
 - familyname: Hyndman
@@
   email: Rob.Hyndman@monash.edu
-- familyname: Panagiotelis
-  othernames: Anastasios
-  address: University of Sydney
-  email: Anastasios.Panagiotelis@sydney.edu.au
 abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
@@
 
-Example of $S^2$ in @Henry2009-om.
+#### Example of $S^2$ in @Henry2009-om.
 
@@
 $$
-\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
+\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
 $$ 
-if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
-Then we take the partial derivatives of \$\\Psi\^{-1}(s,t)\$ with respect to \$s\$ and \$t\$ and have that
+if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. 
+Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
+Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
 $$
@@
 $$ 
-Then each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$. We obtain that the volume density on the sphere is 
+The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
+Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
+Therefore, we obtain that the volume density on the sphere is 
 $$
@@
 
-Similarly in $S^1$, we could let $\{v, p\}$ be the basis of $\mathbb{R}^2$. Then the exponential chart induced
+#### $S^1$
 
-Polar coordinates:
-$\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v$.
+Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization 
+$$
+\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
+$$
+and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have
+$$
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
+$$
+So the riemannian metric is $g_q = \cos(r)$ and further, 
+$$
+\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
+$$
+where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$.
 
-Cartesian coordinates:
-$$
-\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
-= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
-= - \sqrt{1 - p^2}v
-$$
-Then the volume density function is
-$$
-\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
-$$
-and the geodesic distance is
-$$
-d_g(p,q) = \arccos(p) - \arccos(q).
-$$
+<!-- Polar coordinates: -->
+<!-- $\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v$. -->
+
+<!-- Cartesian coordinates: -->
+<!-- $$ -->
+<!-- \psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\ -->
+<!-- = (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\ -->
+<!-- = - \sqrt{1 - p^2}v -->
+<!-- $$ -->
+<!-- Then the volume density function is -->
+<!-- $$ -->
+<!-- \theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2}, -->
+<!-- $$ -->
+<!-- and the geodesic distance is -->
+<!-- $$ -->
+<!-- d_g(p,q) = \arccos(p) - \arccos(q). -->
+<!-- $$ -->

### 2022-05-25 paper/paper.tex
Test for S1 works

@@
 
-\author{Fan~Cheng, Rob J~Hyndman, Anastasios~Panagiotelis}
+\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
 \addresses{\textbf{Fan Cheng}\newline
@@
 \newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
+\textbf{Anastasios Panagiotelis}\newline
+University of Sydney
+\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
 \textbf{Rob J Hyndman}\newline
@@
 \newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
-\textbf{Anastasios Panagiotelis}\newline
-University of Sydney
-\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
 }
@@
 \makeatletter
- \lfoot{\sf Cheng, Hyndman, Panagiotelis: \@date}
+ \lfoot{\sf Cheng, Panagiotelis, Hyndman: \@date}
 \makeatother
@@
 
-Example of \(S^2\) in \textcite{Henry2009-om}.
+\hypertarget{example-of-s2-in-henry2009-om.}{%
+\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}
 
@@
 \[
-\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
-\]
-if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
-Then we take the partial derivatives of \$\textbackslash Psi\^{}\{-1\}(s,t)\$ with respect to \$s\$ and \$t\$ and have that
+\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
+\]
+if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
+Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
+Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
 \[
@@
 \]
-Then each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain that the volume density on the sphere is
+The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
+Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
+Therefore, we obtain that the volume density on the sphere is
 \[
@@
 
-Similarly in \(S^1\), we could let \(\{v, p\}\) be the basis of \(\mathbb{R}^2\). Then the exponential chart induced
-
-Polar coordinates:
-\(\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v\).
-
-Cartesian coordinates:
-\[
-\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
-= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
-= - \sqrt{1 - p^2}v
-\]
-Then the volume density function is
-\[
-\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
-\]
-and the geodesic distance is
-\[
-d_g(p,q) = \arccos(p) - \arccos(q).
-\]
+\hypertarget{s1}{%
+\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}
+
+Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
+\[
+\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
+\]
+and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
+\[
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
+\]
+So the riemannian metric is \(g_q = \cos(r)\) and further,
+\[
+\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
+\]
+where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).
 

### 2022-05-25 R/riemgeom_s2.R

@@ -1,3 +1,12 @@
+#' ---
+#' title: "Kernel density estimation for S1 using fixed and variable bandwidths"
+#' author: "Fan Cheng"
+#' output:
+#'   html_document:
+#'     keep_tex: true
+#' ---
+#+ echo=T
+
 rm(list= ls())
 library(tidyverse)
 library(dimRed)
@@ -28,10 +37,10 @@ switch(distribution,
        },
        "beta" = {
          # 3. theta ~ Beta(a, b) \in [0,1]; (2,5), (5,1)
-         shape1 <- .5
-         shape2 <- .5
-         # shape1 <- 2
-         # shape2 <- 5
+         # shape1 <- .5
+         # shape2 <- .5
+         shape1 <- 2
+         shape2 <- 5
          theta <- rbeta(N, shape1, shape2)
          dentheta <- dbeta(theta, shape1, shape2)
          theta <- theta * (pi / 2)
@@ -136,7 +145,7 @@ plot(dentheta, fx); abline(0, 1, lty = "dashed")
 plot(dentheta, fxkde); abline(0, 1, lty = "dashed")
 
 
-
+# Summary: Pelletier's estimator gives a similar result as the fixed bandwidth KDE.
 
 
 # Now reduce the dimension and estimate the density of the embedding
@@ -159,14 +168,14 @@ noutliers <- 20
 
 # ISOMAP
 ## ---- message=FALSE-------------------------------------------------------------
-metric_isomap <- metricML(x = train, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+suppressWarnings(metric_isomap <- metricML(x = train, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
                           # annmethod = annmethod, distance = distance, treetype = treetype, 
                           searchtype = searchtype
-)
+))
 # summary(metric_isomap)
 
 ## -------------------------------------------------------------------------------
-# fixed bandwidth
+# KDE with fixed bandwidth
 fn <- metric_isomap$embedding
 # fn <- fn+0.45
 # E1 <- fn[,1] # rename as Ed to match the aesthetics in plot_ellipse()
[…]

### 2022-06-06 paper/paper.Rmd
Add kde reference

@@
 
-In general, given a multivariate random sample $\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
+Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
+It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
+
+In general, given a multivariate random sample $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
 \begin{equation}
 \label{eq:kde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
+\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
 \end{equation} 
-where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{y})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
-When using the standard gaussian kernel, $K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
+where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})$.
+When using the standard gaussian kernel, $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$, we have
 \begin{equation}
 \label{eq:kernel}
-K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
+K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
 \end{equation} 
@@
 
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{y})$.
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$.
 We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.

### 2022-06-06 paper/paper.tex
Add kde reference

@@
 
-In general, given a multivariate random sample \(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
+In general, given a multivariate random sample \(\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
 \begin{equation}
 \label{eq:kde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
+\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
 \end{equation}
-where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
-When using the standard gaussian kernel, \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
+where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})\).
+When using the standard gaussian kernel, \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\), we have
 \begin{equation}
 \label{eq:kernel}
-K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
+K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
 \end{equation}
@@
 
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{y})\).
+The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\).
 We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.

### 2022-06-07 paper/paper.Rmd
More on KDE in Sec2

@@
 There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
-For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of
-the estimated point [balloon estimator; @Terrell1992-ef] or the data point [sample smoothing estimator; @Terrell1992-ef].
+For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ut] or the data point [sample smoothing estimator; @Terrell1992-ut].
 However, these density estimators are based on random samples in the Euclidean space.
@@
 It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
-
-In general, given a multivariate random sample $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
+@Scott1992-cj and @Scott2015-vl discussed a number of techniques for estimating the kernel densities of multivariate data.
+In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
+<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
+drawn from a density $f$, the kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
 \begin{equation}
 \label{eq:kde}
-\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
+\hat{f}(\pmb{x};\pmb{H})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{X}_i),
 \end{equation} 
-where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})$.
-When using the standard gaussian kernel, $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$, we have
+where $\pmb{H}$ is called the bandwidth matrix and it is a $d\times d$ symmetric positive definite matrix, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})$.
+When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
 \begin{equation}
 \label{eq:kernel}
-K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
+K_{\pmb{H}}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{X}_i)\right],
 \end{equation} 
-with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
+with $|\pmb{H}|$ being the determinant of matrix $\pmb{H}$.
+Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function.
+And an average over all data points is taken as the estimated density $\hat{f}$.
+KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point $\pmb{x}$.
 
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function $K(\pmb{x})$.
-We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
-@Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
+The choice of the the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$.
+It controls the smoothing across all sample data points.
+For a given kernel function $K(\pmb{x})$, the shape of the local smoothing at each point is determined.
+Then the performance of the density estimate $\hat{f}$ depends on the choice of $\pmb{H}$.
+When $\pmb{H}$ is too small, the density estimates will apprear to be wiggly.
+While with $\pmb{H}$ being too large, the density estimates will smooth out some local feathers of $f$.
+So the accuracy of $\hat{f}$ depends strongly on the bandwidth matrix.
+Many bandwidth selection methods has been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
+The generalization to multivariate case could mainly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
+
+Note that the bandwidth matrix in \@ref(eq:kde) is a global smoothing parameter for all data points.
+However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
+This is ususally referred to as variable kernel density estimation [VKDE; @Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
+In VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $i$ instead of a fixed $\pmb{H}$ as below,
+\begin{equation}
+\label{eq:vkde}
+\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
+\end{equation}
+
+For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix, which limits the full flexibility of the kernel function.
 Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
+In the cases of a Riemannian manifold, the underlying density is supported on a single manifold, the fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion.
+Therefore, there is a need for an estimator for Riemannian manifolds as described in \autoref{riemest}.
 

### 2022-06-07 paper/paper.tex
More on KDE in Sec2

@@
 There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
-For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of
-the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
+For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ut}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]}.
 However, these density estimators are based on random samples in the Euclidean space.
@@
 
-In general, given a multivariate random sample \(\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
+Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
+It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
+\textcite{Scott1992-cj} and \textcite{Scott2015-vl} discussed a number of techniques for estimating the kernel densities of multivariate data.
+In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
+drawn from a density \(f\), the kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
 \begin{equation}
 \label{eq:kde}
-\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
+\hat{f}(\pmb{x};\pmb{H})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{X}_i),
 \end{equation}
-where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{x})\).
-When using the standard gaussian kernel, \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\), we have
+where \(\pmb{H}\) is called the bandwidth matrix and it is a \(d\times d\) symmetric positive definite matrix, \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})\).
+When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
 \begin{equation}
 \label{eq:kernel}
-K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
+K_{\pmb{H}}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{X}_i)\right],
 \end{equation}
-with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
-
-The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\).
-We consider the sample smooth estimator for each data point, which gives a kernel density estimate with local smoothing.
-\textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix.
+with \(|\pmb{H}|\) being the determinant of matrix \(\pmb{H}\).
+Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function.
+And an average over all data points is taken as the estimated density \(\hat{f}\).
+KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point \(\pmb{x}\).
+
+The choice of the the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\).
+It controls the smoothing across all sample data points.
+For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined.
+Then the performance of the density estimate \(\hat{f}\) depends on the choice of \(\pmb{H}\).
+When \(\pmb{H}\) is too small, the density estimates will apprear to be wiggly.
+While with \(\pmb{H}\) being too large, the density estimates will smooth out some local feathers of \(f\).
+So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
+Many bandwidth selection methods has been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
+The generalization to multivariate case could mainly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
+
+Note that the bandwidth matrix in \eqref{eq:kde} is a global smoothing parameter for all data points.
+However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
+This is ususally referred to as variable kernel density estimation {[}VKDE; \textcite{Breiman1977-qc}; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]}.
+In VKDE, the estimator in Equation \eqref{eq:vkde} is similar to that in \eqref{eq:kde} except that the bandwidth is a variable matrix \(\pmb{H}_i\) for each \(i\) instead of a fixed \(\pmb{H}\) as below,
+\begin{equation}
+\label{eq:vkde}
+\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
+\end{equation}
+
+For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix, which limits the full flexibility of the kernel function.
 Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the sample space used for kernel density estimation is no longer Euclidean and we need to use the estimator described in \autoref{riemest}.
+In the cases of a Riemannian manifold, the underlying density is supported on a single manifold, the fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion.
+Therefore, there is a need for an estimator for Riemannian manifolds as described in \autoref{riemest}.
 

### 2022-06-07 paper/paper.tex
Change notation for \hat{f}

@@
 Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\).
-Denote \(A_i, i = 1, 2, \dots, N\), where \(\pmb{A}_i \in \mathcal{R}^d\), as i.i.d.
+Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d.
 random objects on \(M\) with density \(f\).
-For each \(p\in M\), \textcite{Pelletier2005-vu} proposed the kernel density estimator of \(f\) to be
+For each point \(p\in M\), \textcite{Pelletier2005-vu} proposed a kernel density estimator of \(f\) to be
 \begin{equation}
 \label{eq:denriem}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i}{h}),
 \end{equation}
-where \(h\) is the bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
-\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q \(\exp_q^{-1}(p)\).
+where \(h\) is the global bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
+\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q, \(\exp_q^{-1}(p)\).
 The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
@@
 \label{eq:denconsist}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
 \end{equation}
@@
 
+\hypertarget{volume-density-function}{%
+\subsubsection{Volume density function}\label{volume-density-function}}
+
 However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
 Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
-Therefore, based on the estimator in \eqref{eq:denriem}, we propose to a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
+Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
 \begin{equation}
 \label{eq:denestimator}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
 \end{equation}
-where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension
-\(d\times d\).
+where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
 
@@
 \hypertarget{riemannian-metric-as-variable-bandwidth}{%
-\subsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}
+\subsubsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}
 
@@
 
-\hypertarget{volume-density-function}{%
-\subsection{Volume density function}\label{volume-density-function}}
+\hypertarget{volume-density-function-1}{%
+\subsection{Volume density function}\label{volume-density-function-1}}
 

### 2022-06-07 paper/paper.Rmd
Change notation for \hat{f}

@@
 Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$.
-Denote $A_i, i = 1, 2, \dots, N$, where $\pmb{A}_i \in \mathcal{R}^d$, as i.i.d.
+Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathcal{R}^d$, as i.i.d.
 random objects on $M$ with density $f$.
-For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be
+For each point $p\in M$, @Pelletier2005-vu proposed a kernel density estimator of $f$ to be
 \begin{equation}
 \label{eq:denriem}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i}{h}),
 \end{equation} 
-where $h$ is the bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
-@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$.
+where $h$ is the global bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
+@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q, $\exp_q^{-1}(p)$.
 The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
@@
 \label{eq:denconsist}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
 \end{equation} 
@@
 
+### Volume density function
+
 However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
 Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
-Therefore, based on the estimator in \@ref(eq:denriem), we propose to a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
+Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
 \begin{equation}
 \label{eq:denestimator}
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
 \end{equation} 
-where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension
-$d\times d$.
+where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.
 
@@
 
-## Riemannian metric as variable bandwidth
+### Riemannian metric as variable bandwidth
 

### 2022-06-13 paper/paper.Rmd
VKDE done

@@
 It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
-@Scott1992-cj and @Scott2015-vl discussed a number of techniques for estimating the kernel densities of multivariate data.
+@Scott1992-cj and @Scott2015-vl covered a wide range of techniques for estimating the kernel densities of multivariate data.
 In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
@@
 Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function.
-And an average over all data points is taken as the estimated density $\hat{f}$.
+Then an average over all data points is taken as the estimated density $\hat{f}$.
 KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point $\pmb{x}$.
 
-The choice of the the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$.
-It controls the smoothing across all sample data points.
+The choice of the the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$ because it controls the smoothing across the sample data points.
 For a given kernel function $K(\pmb{x})$, the shape of the local smoothing at each point is determined.
@@
 When $\pmb{H}$ is too small, the density estimates will apprear to be wiggly.
-While with $\pmb{H}$ being too large, the density estimates will smooth out some local feathers of $f$.
+On the contrary, when $\pmb{H}$ is too large, the density estimates might smooth out some local feathers of $f$.
 So the accuracy of $\hat{f}$ depends strongly on the bandwidth matrix.
@@
 For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
-The generalization to multivariate case could mainly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
+The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
+In this paper, we consider the multivariate kernel density estimation.
 
@@
 However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-This is ususally referred to as variable kernel density estimation [VKDE; @Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
-In VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $i$ instead of a fixed $\pmb{H}$ as below,
+This is ususally referred to as variable kernel density estimation [VKDE; @Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut], where the bandwidth is varied depending on either the location of the sample points or that of the estimated points [Section 6.6 of @Scott2015-vl].
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point $\pmb{X}_i$ and will refer to this as the variable/adaptive kernel density estimation unless otherwise stated.
+
+In multivariate VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $\pmb{X}_i$,
 \begin{equation}
@@
 \end{equation}
+Compared to a global fixed $\pmb{H}$, a variable bandwidth matrix is able to account for the pointwise adaptive local structure at each point and generate an estimate more applicable to the case where the distortion at each point is different at each point in the manifold learning embedding.
 
 For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix, which limits the full flexibility of the kernel function.
-Further, the estimator in Equation \@ref(eq:kde) is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the underlying density is supported on a single manifold, the fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion.
-Therefore, there is a need for an estimator for Riemannian manifolds as described in \autoref{riemest}.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector with the diagonal elements, which limits the full flexibility of the kernel function.
+More importantly, the estimator in Equation \@ref(eq:vkde) is only applicable when the sample space is Euclidean.
+In the cases of a Riemannian manifold, the underlying density is supported on a single manifold.
+The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
+Therefore, there is a need for an adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
 
@@
 Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
-Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
+Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
 \begin{equation}

### 2022-06-13 paper/paper.tex
VKDE done

@@
 It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
-\textcite{Scott1992-cj} and \textcite{Scott2015-vl} discussed a number of techniques for estimating the kernel densities of multivariate data.
+\textcite{Scott1992-cj} and \textcite{Scott2015-vl} covered a wide range of techniques for estimating the kernel densities of multivariate data.
 In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
@@
 Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function.
-And an average over all data points is taken as the estimated density \(\hat{f}\).
+Then an average over all data points is taken as the estimated density \(\hat{f}\).
 KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point \(\pmb{x}\).
 
-The choice of the the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\).
-It controls the smoothing across all sample data points.
+The choice of the the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\) because it controls the smoothing across the sample data points.
 For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined.
@@
 When \(\pmb{H}\) is too small, the density estimates will apprear to be wiggly.
-While with \(\pmb{H}\) being too large, the density estimates will smooth out some local feathers of \(f\).
+On the contrary, when \(\pmb{H}\) is too large, the density estimates might smooth out some local feathers of \(f\).
 So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
@@
 For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
-The generalization to multivariate case could mainly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
+The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
+In this paper, we consider the multivariate kernel density estimation.
 
@@
 However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-This is ususally referred to as variable kernel density estimation {[}VKDE; \textcite{Breiman1977-qc}; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]}.
-In VKDE, the estimator in Equation \eqref{eq:vkde} is similar to that in \eqref{eq:kde} except that the bandwidth is a variable matrix \(\pmb{H}_i\) for each \(i\) instead of a fixed \(\pmb{H}\) as below,
+This is ususally referred to as variable kernel density estimation {[}VKDE; \textcite{Breiman1977-qc}; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]}, where the bandwidth is varied depending on either the location of the sample points or that of the estimated points \autocite[Section 6.6 of][]{Scott2015-vl}.
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point \(\pmb{X}_i\) and will refer to this as the variable/adaptive kernel density estimation unless otherwise stated.
+
+In multivariate VKDE, the estimator in Equation \eqref{eq:vkde} is similar to that in \eqref{eq:kde} except that the bandwidth is a variable matrix \(\pmb{H}_i\) for each \(\pmb{X}_i\),
 \begin{equation}
@@
 \end{equation}
+Compared to a global fixed \(\pmb{H}\), a variable bandwidth matrix is able to account for the pointwise adaptive local structure at each point and generate an estimate more applicable to the case where the distortion at each point is different at each point in the manifold learning embedding.
 
 For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a vector or a diagonal matrix instead of a full bandwidth matrix, which limits the full flexibility of the kernel function.
-Further, the estimator in Equation \eqref{eq:kde} is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the underlying density is supported on a single manifold, the fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion.
-Therefore, there is a need for an estimator for Riemannian manifolds as described in \autoref{riemest}.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector with the diagonal elements, which limits the full flexibility of the kernel function.
+More importantly, the estimator in Equation \eqref{eq:vkde} is only applicable when the sample space is Euclidean.
+In the cases of a Riemannian manifold, the underlying density is supported on a single manifold.
+The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
+Therefore, there is a need for an adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
 
@@
 Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
-Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of f with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
+Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
 \begin{equation}

### 2022-06-13 R/riemgeom_s2.R
VKDE done

@@ -106,7 +106,7 @@ fx <- rep(0, N)
 for(i in 1:N){
   fi <- rep(0, N)
   bindex <- which((abs(acos(x) - acos(x[i])) / h) <= 1) # use only neighbors within radius r
-  fi[bindex] <- 1 / sqrt((cos(acos(x[bindex]) - acos(x[i])))) * dnorm(x = acos(x[5]), mean = acos(x[i]), sd = h) / (pnorm(1) - pnorm(-1))
+  fi[bindex] <- 1 / sqrt((cos(acos(x[bindex]) - acos(x[i])))) * dnorm(x = acos(x[bindex]), mean = acos(x[i]), sd = h) / (pnorm(1) - pnorm(-1))
   # OR use the true theta and true dg and use all data points for a weighted estimation
   # fi <- 1 / sqrt((cos(acos(x) - acos(x[i])))) * dnorm(x = acos(x), mean = acos(x[i]), sd = h)
   fx <- fx + fi

### 2022-06-14 R/riemgeom_s2.R
Add scaling with AUC

@@ -16,7 +16,7 @@ Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1)
 N <- 1000
 d <- 1
-distribution <- c("uniform", "normal", "beta", "lognormal", "mixed")[4] # CHANGE DISTRIBUTION OF THETA
+distribution <- c("uniform", "normal", "beta", "lognormal", "mixed")[3] # CHANGE DISTRIBUTION OF THETA
 switch(distribution,
        "uniform" = {
          # 1. theta ~ U(0, pi/2)
@@ -88,7 +88,8 @@ y <- sin(theta)
 # range(x) # [0, 1]
 # plot(x, denx)
 # range(denx)
-
+idx <- order(x)
+sum(diff(x[idx])*rollmean(denx[idx],2)) # 0.9984189
 
 # KDE with fixed bandwidth
 h <- ks::hpi(theta, binned = TRUE) # 0.034
@@ -117,7 +118,7 @@ fx <- fx / N
 # Plotting
 # par(mfrow=c(1,1))
 plot(theta, dentheta, main = "True density of theta")
-# plot(x, denx, main = "True density of x")
+# points(x, fxkde, main = "True density of x")
 points(theta, fxkde, main = paste("Estimated f(x) with KDE, h = ", round(h, 3)), col = "green")
 points(theta, fx, main = "Estimated f(x) with riemannian metric", col = "red")
 # points(x, dentheta, col = "red", lty = "dashed")
@@ -130,21 +131,48 @@ summary(dentheta)
 summary(fxkde)
 summary(fx)
 
+
+
+
+
+## Check area under the curve
+# trapz(theta, dentheta)
+id <- order(theta)
+sum(diff(theta[id])*rollmean(dentheta[id],2)) # 0.99899 for true densities
+sum(diff(theta[id])*rollmean(fxkde[id],2)) # 0.98569 for kde
+AUC <- sum(diff(theta[id])*rollmean(fx[id],2)) # 0.98690 for dc-kde
+
+# # using sintegral in Bolstad2
+# require(Bolstad2)
+# sintegral(theta[id],dentheta[id])
+# # using auc in MESS
+# require(MESS)
+# auc(theta[id],dentheta[id], type = 'spline')
+# # using integrate.xy in sfsmisc
+# require(sfsmisc)
+# integrate.xy(theta[id],dentheta[id])
[…]

### 2022-06-14 R/riemgeom_s2.R
HTML with distortion correction for S1

@@ -16,7 +16,7 @@ Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1)
 N <- 1000
 d <- 1
-distribution <- c("uniform", "normal", "beta", "lognormal", "mixed")[3] # CHANGE DISTRIBUTION OF THETA
+distribution <- c("uniform", "normal", "beta", "lognormal", "mixed")[4] # CHANGE DISTRIBUTION OF THETA
 switch(distribution,
        "uniform" = {
          # 1. theta ~ U(0, pi/2)
@@ -140,7 +140,7 @@ summary(fx)
 id <- order(theta)
 sum(diff(theta[id])*rollmean(dentheta[id],2)) # 0.99899 for true densities
 sum(diff(theta[id])*rollmean(fxkde[id],2)) # 0.98569 for kde
-AUC <- sum(diff(theta[id])*rollmean(fx[id],2)) # 0.98690 for dc-kde
+(AUC <- sum(diff(theta[id])*rollmean(fx[id],2))) # 0.98690 for dc-kde
 
 # # using sintegral in Bolstad2
 # require(Bolstad2)
@@ -151,7 +151,26 @@ AUC <- sum(diff(theta[id])*rollmean(fx[id],2)) # 0.98690 for dc-kde
 # # using integrate.xy in sfsmisc
 # require(sfsmisc)
 # integrate.xy(theta[id],dentheta[id])
-# # The trapezoidal method is less accurate than the spline method, so MESS::auc (uses spline method) or Bolstad2::sintegral (uses Simpson's rule) should probably be preferred.
+## # The trapezoidal method is less accurate than the spline method, so MESS::auc (uses spline method) or Bolstad2::sintegral (uses Simpson's rule) should probably be preferred.
+# library(bayestestR)
+# posterior <- distribution_normal(1000)
+# dens <- estimate_density(posterior)
+# dens <- dens[dens$x > 0, ]
+# x <- dens$x
+# y <- dens$y
+# plot(x, y)
+# area_under_curve(x, y, method = "trapezoid")
+# #> [1] 0.498307
+# area_under_curve(x, y, method = "step")
+# #> [1] 0.499534
+# area_under_curve(x, y, method = "spline")
+# #> [1] 0.498307
+library(bayestestR)
+area_under_curve(theta[id], fx[id], method = "trapezoid")
[…]

### 2022-06-26 paper/paper.Rmd
HTML with distortion correction for S1
Check MKDE

@@
 Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
-@Scott1992-cj and @Scott2015-vl covered a wide range of techniques for estimating the kernel densities of multivariate data.
+It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
+@Scott1992-cj and @Scott2015-vl cover a wide range of techniques for estimating the kernel densities of multivariate data.
 In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
@@
 \end{equation} 
-where $\pmb{H}$ is called the bandwidth matrix and it is a $d\times d$ symmetric positive definite matrix, $K(\pmb{x})$ is the kernel function, and $K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})$.
+where $K(\pmb{x})$ is the kernel function, $\pmb{H}$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})$.
 When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
@@
 Then an average over all data points is taken as the estimated density $\hat{f}$.
-KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point $\pmb{x}$.
+KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point $\pmb{x}$.
 
-The choice of the the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$ because it controls the smoothing across the sample data points.
+The choice of the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$ because it controls the smoothing across the sample data.
 For a given kernel function $K(\pmb{x})$, the shape of the local smoothing at each point is determined.
@@
 When $\pmb{H}$ is too small, the density estimates will apprear to be wiggly.
-On the contrary, when $\pmb{H}$ is too large, the density estimates might smooth out some local feathers of $f$.
+On the contrary, when $\pmb{H}$ is too large, the density estimates might smooth out some local features of $f$.
 So the accuracy of $\hat{f}$ depends strongly on the bandwidth matrix.
-Many bandwidth selection methods has been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
+Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
 For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
 The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
-In this paper, we consider the multivariate kernel density estimation.
+In this paper, we focus on the multivariate kernel density estimation.
 
@@
 This is ususally referred to as variable kernel density estimation [VKDE; @Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut], where the bandwidth is varied depending on either the location of the sample points or that of the estimated points [Section 6.6 of @Scott2015-vl].
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point $\pmb{X}_i$ and will refer to this as the variable/adaptive kernel density estimation unless otherwise stated.
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point $\pmb{X}_i$ and will refer to this as the *variable/adaptive kernel density estimation(VKDE)* unless otherwise stated.
 
@@
 \label{eq:vkde}
-\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
+\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
 \end{equation}
-Compared to a global fixed $\pmb{H}$, a variable bandwidth matrix is able to account for the pointwise adaptive local structure at each point and generate an estimate more applicable to the case where the distortion at each point is different at each point in the manifold learning embedding.
+Compared to a global fixed $\pmb{H}$, a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
 
 For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector with the diagonal elements, which limits the full flexibility of the kernel function.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
 More importantly, the estimator in Equation \@ref(eq:vkde) is only applicable when the sample space is Euclidean.
@@
 The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
-Therefore, there is a need for an adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
+This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
 
@@
 \label{eq:denriem}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i}{h}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i)}{h}),
 \end{equation} 

### 2022-06-27 paper/paper.Rmd
Correct notations in apped

@@
 where $h$ is the global bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
-@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q, $\exp_q^{-1}(p)$.
+@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q, $\exp_q^{-1}(p)$.
 The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
@@
 To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
-The inner product between two vectors $u,v \in T_pM$, $<u,v>_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{<u,v>_g}$ and the angle between two vectors $\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}$ in the tangent space.
+The inner product between two vectors $u,v \in T_pM$, $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{\langle u,v \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{\|u\|\|v\|}$ in the tangent space.
 Therefore, for each point $p\in M$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
@@
 This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
+The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at the Monash University.
 
@@
 A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-A $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
-Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local
-coordinates $(x_1(p),\dots, x_d(p)) := \varphi(p)$.
-A collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
+And a $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
+Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x^1(p),\dots, x^d(p)) := \varphi(p)$.
+Further, a collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
+
 The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts, 
@@
 $$ are differentiable of class $C^\infty$ (smooth).
-
-Let $\varphi$ be an injective map $E \rightarrow \varphi(E)$. Then $\varphi$ is an embedding of $E$ into $M$ if and only if
+Let $\varphi$ be an injective map: $E \rightarrow \varphi(E)$. Then $\varphi$ is an *embedding* of $E$ into $M$ if and only if
 $\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
@@
 
-The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through the point $p$ or as the directional derivatives at $p$.
+The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through point $p$ or as the directional derivatives at $p$.
 Here we define the tangent vector via the velocity of curves.
 
-For any $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curve through $p$, i.e.
+For any point $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curves passing through $p$, i.e.
 $\gamma_1(0) = \gamma_2(0) = p$.
-$\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
+We say $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
 $$
@@
 $$ 
-A *tangent vector* to a manifold $M$ at point $p$, $v_p$, is any equivalent class of the differentiable curves initialized at $p$.
+A *tangent vector* to a manifold $M$ at point $p$, denoted as $v_p$, is any equivalent class of the differentiable curves initialized at $p$.
 The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
 The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
-The collection of all tangent spaces defines the *tangent bundle* $TM = \cup_{p \in M}T_pM$.
-Tangent vectors can also be seen as the directional derivatives at $p$.
-For a given coordinate chart $\varphi=(x_1,\dots,x_d)$, the tangent vectors defining partial
-derivatives are denoted as $\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)$, which define a *basis* of the tangent space.
+The collection of all tangent spaces defines the *tangent bundle*, $TM = \cup_{p \in M}T_pM$.
 
-The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with corresponding *cotangent vector*
-$z_p: T^\star_pM \rightarrow \mathbb{R}^d$, and basis denoted as $dx_1(p),\dots,dx_d(p)$.
+Tangent vectors can also be seen as the directional derivatives at $p$. For a given coordinate chart $\varphi=(x^1,\dots,x^d)$, the tangent vectors defining partial
+derivatives are denoted as $\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)$, which defines a *basis* of the tangent space.
+The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with corresponding *cotangent vectors*
+$z_p: T^\star_pM \rightarrow \mathbb{R}^d$, and a basis denoted as $dx^1(p),\dots,dx^d(p)$.
 
@@
 
-A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is $d\times d$ symmetric positive definite and varies smoothly at $p$.
+A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is a $d\times d$ symmetric positive definite matrix and varies smoothly at $p$.
 Generally, we
 omit the subscript $p$ and refer to $g$ as the Riemannian metric.
-The inner product for two vectors $u, v \in T_pM$ is written as $<u, v>_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed.
+The inner product between two vectors $u, v \in T_pM$ is written as $\langle u, v \rangle_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed.
 A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
 
-The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{<u,v>_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
-$g$ could also be used to define the line element $dl^2 = g_{ij}dx_idx_j$ and the volume element $dV_g = \sqrt{\det(g)}dx_1\dots dx_d$, where $(x_1,\dots,x_d)$ are the local coordinates of the chart $(U, \varphi)$.
+The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{\langle u,v \rangle_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
+It could also be used to define the line element $dl^2 = g_{ij}dx^i dx^j$ and the volume element $dV_g = \sqrt{\det(g)}dx^1 \dots dx^d$, where $(x^1,\dots,x^d)$ are the local coordinates of the chart $(U, \varphi)$.
 For a curve $\gamma: I \rightarrow M$, the length of the curve is
 $$
-l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt},
+l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
 $$
@@
 $$
-Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
+Vol(W) = \int_W \sqrt{\det(g)}dx^1 \dots dx^d,
 $$
@@
 
-The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. A curve
-$\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the second-order ordinary differential equation is satisfied, 
+The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold.
+Intuitively, geodesics are the *straightest possible curves* in a Riemannian manifold [Section 7.2.3 of @Nakahara2018-zs].
+A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the second-order ordinary differential equation is satisfied, 
 $$
-\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
+\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
 $$ 
-where $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
+where ${x^i}$ are the coordinates of the curve $\gamma$ and $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
 $$
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
-+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
+\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k} 
++ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
 $$ 
@@
 The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
-For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the *injectivity radius* at $p \in M$ is the distance to the cut locus.
-The injectivity radius of $(M,g)$, $\textit{inj}_gM$ is the infimum of the injectivity radii over all points on the manifold.
+For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the distance to the cut locus is the *injectivity radius* at $p \in M$.
+Therefore, the injectivity radius of the Riemannian manifold $(M,g)$, $\textit{inj}_gM$, is the infimum of the injectivity radii over all points on the manifold.
 
 <!-- ### Isometry -->
-
-## Pushforward and pullback metric
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
-The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
-$$
-<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-$$ 
-where $df_p^{-1}$ is the Jacobian of $f^{-1}$.
-The tangent vectors $f_\star u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
-Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
-
-<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->
 
@@
 Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
+
+## Pushforward and pullback metric
+
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
+Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
+The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
+$$
+\langle u,v \rangle_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+$$ 
+where $df_p^{-1}$ is the Jacobian of $f^{-1}$.
+The tangent vectors $f_\star u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
+Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
+
+<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->
 

### 2022-06-27 paper/paper.tex
Correct notations in apped

@@
 Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the density of the data based on weighted kernels centered at the data.
-\textcite{Scott1992-cj} and \textcite{Scott2015-vl} covered a wide range of techniques for estimating the kernel densities of multivariate data.
+It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
+\textcite{Scott1992-cj} and \textcite{Scott2015-vl} cover a wide range of techniques for estimating the kernel densities of multivariate data.
 In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
@@
 \end{equation}
-where \(\pmb{H}\) is called the bandwidth matrix and it is a \(d\times d\) symmetric positive definite matrix, \(K(\pmb{x})\) is the kernel function, and \(K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})\).
+where \(K(\pmb{x})\) is the kernel function, \(\pmb{H}\) is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension \(d\times d\), and \(K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})\).
 When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
@@
 Then an average over all data points is taken as the estimated density \(\hat{f}\).
-KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points that are far away from the estimated data point \(\pmb{x}\).
-
-The choice of the the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\) because it controls the smoothing across the sample data points.
+KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point \(\pmb{x}\).
+
+The choice of the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\) because it controls the smoothing across the sample data.
 For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined.
@@
 When \(\pmb{H}\) is too small, the density estimates will apprear to be wiggly.
-On the contrary, when \(\pmb{H}\) is too large, the density estimates might smooth out some local feathers of \(f\).
+On the contrary, when \(\pmb{H}\) is too large, the density estimates might smooth out some local features of \(f\).
 So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
-Many bandwidth selection methods has been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
+Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
 For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
 The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
-In this paper, we consider the multivariate kernel density estimation.
+In this paper, we focus on the multivariate kernel density estimation.
 
@@
 This is ususally referred to as variable kernel density estimation {[}VKDE; \textcite{Breiman1977-qc}; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]}, where the bandwidth is varied depending on either the location of the sample points or that of the estimated points \autocite[Section 6.6 of][]{Scott2015-vl}.
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point \(\pmb{X}_i\) and will refer to this as the variable/adaptive kernel density estimation unless otherwise stated.
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point \(\pmb{X}_i\) and will refer to this as the \emph{variable/adaptive kernel density estimation(VKDE)} unless otherwise stated.
 
@@
 \label{eq:vkde}
-\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i),
+\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
 \end{equation}
-Compared to a global fixed \(\pmb{H}\), a variable bandwidth matrix is able to account for the pointwise adaptive local structure at each point and generate an estimate more applicable to the case where the distortion at each point is different at each point in the manifold learning embedding.
+Compared to a global fixed \(\pmb{H}\), a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
 
 For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector with the diagonal elements, which limits the full flexibility of the kernel function.
+However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
 More importantly, the estimator in Equation \eqref{eq:vkde} is only applicable when the sample space is Euclidean.
@@
 \label{eq:denriem}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i}{h}),
+\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i)}{h}),
 \end{equation}
 where \(h\) is the global bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
-\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q, \(\exp_q^{-1}(p)\).
+\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q, \(\exp_q^{-1}(p)\).
 The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
@@
 To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
-The inner product between two vectors \(u,v \in T_pM\), \(<u,v>_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{<u,v>_g}\) and the angle between two vectors \(\cos{\theta}=\frac{<u,v>_g}{\|u\|\|v\|}\) in the tangent space.
+The inner product between two vectors \(u,v \in T_pM\), \(\langle u,v \rangle_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{\langle u,v \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{\|u\|\|v\|}\) in the tangent space.
 Therefore, for each point \(p\in M\) in any coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
@@
 A \emph{Hausdorff space} is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-A \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
-Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local
-coordinates \((x_1(p),\dots, x_d(p)) := \varphi(p)\).
-A collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
+And a \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
+Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x^1(p),\dots, x^d(p)) := \varphi(p)\).
+Further, a collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
+
 The manifold \(M\) is a \emph{differentiable manifold} if there exists an atlas of \(M\), \(\{U_\alpha, \varphi_\alpha\}\), such that the \emph{transition maps} between any two charts,
@@
 \] are differentiable of class \(C^\infty\) (smooth).
-
-Let \(\varphi\) be an injective map \(E \rightarrow \varphi(E)\). Then \(\varphi\) is an embedding of \(E\) into \(M\) if and only if
+Let \(\varphi\) be an injective map: \(E \rightarrow \varphi(E)\). Then \(\varphi\) is an \emph{embedding} of \(E\) into \(M\) if and only if
 \(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
@@
 
-The tangent vector at point \(p\) can be intuitively viewed as the velocity of a curve passing through the point \(p\) or as the directional derivatives at \(p\).
+The tangent vector at point \(p\) can be intuitively viewed as the velocity of a curve passing through point \(p\) or as the directional derivatives at \(p\).
 Here we define the tangent vector via the velocity of curves.
 
-For any \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curve through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
-\(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
+For any point \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curves passing through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
+We say \(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
 \[
@@
 \]
-A \emph{tangent vector} to a manifold \(M\) at point \(p\), \(v_p\), is any equivalent class of the differentiable curves initialized at \(p\).
+A \emph{tangent vector} to a manifold \(M\) at point \(p\), denoted as \(v_p\), is any equivalent class of the differentiable curves initialized at \(p\).
 The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\).
 The tangent space is a vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the chart \(\varphi\) locally at \(p\).
-The collection of all tangent spaces defines the \emph{tangent bundle} \(TM = \cup_{p \in M}T_pM\).
-Tangent vectors can also be seen as the directional derivatives at \(p\).
-For a given coordinate chart \(\varphi=(x_1,\dots,x_d)\), the tangent vectors defining partial
-derivatives are denoted as \(\frac{\partial}{\partial x_1}(p),\dots,\frac{\partial}{\partial x_d}(p)\), which define a \emph{basis} of the tangent space.
-
-The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector}
-\(z_p: T^\star_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).
+The collection of all tangent spaces defines the \emph{tangent bundle}, \(TM = \cup_{p \in M}T_pM\).
+
+Tangent vectors can also be seen as the directional derivatives at \(p\). For a given coordinate chart \(\varphi=(x^1,\dots,x^d)\), the tangent vectors defining partial
+derivatives are denoted as \(\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)\), which defines a \emph{basis} of the tangent space.
+The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vectors}
+\(z_p: T^\star_pM \rightarrow \mathbb{R}^d\), and a basis denoted as \(dx^1(p),\dots,dx^d(p)\).
 
@@
 
-A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is \(d\times d\) symmetric positive definite and varies smoothly at \(p\).
+A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is a \(d\times d\) symmetric positive definite matrix and varies smoothly at \(p\).
 Generally, we
 omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
-The inner product for two vectors \(u, v \in T_pM\) is written as \(<u, v>_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
+The inner product between two vectors \(u, v \in T_pM\) is written as \(\langle u, v \rangle_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
 A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
 
-The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
-\(g\) could also be used to define the line element \(dl^2 = g_{ij}dx_idx_j\) and the volume element \(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the local coordinates of the chart \((U, \varphi)\).
+The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{\langle u,v \rangle_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
+It could also be used to define the line element \(dl^2 = g_{ij}dx^i dx^j\) and the volume element \(dV_g = \sqrt{\det(g)}dx^1 \dots dx^d\), where \((x^1,\dots,x^d)\) are the local coordinates of the chart \((U, \varphi)\).
 For a curve \(\gamma: I \rightarrow M\), the length of the curve is
 \[
-l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt},
+l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
 \]
@@
 \[
-Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
+Vol(W) = \int_W \sqrt{\det(g)}dx^1 \dots dx^d,
 \]
@@
 
-The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold. A curve
-\(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
-\[
-\frac{d^2 x_i}{dt^2} + \Gamma^i_{jk} \frac{d x_j}{dt} \frac{dx_k}{dt} = 0,
-\]
-where \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
-\[
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g_{il} (\frac{\partial g_{il}}{\partial x_k} 
-+ \frac{\partial g_{kl}}{\partial x_j} - \frac{\partial g_{jk}}{\partial x_l}).
+The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold.
+Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold.
+\autocite[Section 7.2.3 of][]{Nakahara2018-zs}
+A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
+\[
+\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
+\]
+where \({x^i}\) are the coordinates of the curve \(\gamma\) and \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
+\[
+\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k} 
++ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
 \]
@@
 The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
-For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the \emph{injectivity radius} at \(p \in M\) is the distance to the cut locus.
-The injectivity radius of \((M,g)\), \(\textit{inj}_gM\) is the infimum of the injectivity radii over all points on the manifold.
-
-\hypertarget{pushforward-and-pullback-metric}{%
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
-The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
-\[
-<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-\]
-where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
-The tangent vectors \(f_\star u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
-Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
+For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the distance to the cut locus is the \emph{injectivity radius} at \(p \in M\).
+Therefore, the injectivity radius of the Riemannian manifold \((M,g)\), \(\textit{inj}_gM\), is the infimum of the injectivity radii over all points on the manifold.
 
@@
 
+\hypertarget{pushforward-and-pullback-metric}{%
+\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
+
+Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
+Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
+The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
+\[
+\langle u,v \rangle_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+\]
+where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
+The tangent vectors \(f_\star u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
+Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
+
 \hypertarget{volume-density-function-1}{%

### 2022-06-28 paper/paper.Rmd
Rewrite pushforward and pullback

@@
 $\gamma_1(0) = \gamma_2(0) = p$.
-We say $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that
+We say $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart $(U,\varphi)$ at $p$ such that
 $$
@@
 derivatives are denoted as $\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)$, which defines a *basis* of the tangent space.
-The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with corresponding *cotangent vectors*
-$z_p: T^\star_pM \rightarrow \mathbb{R}^d$, and a basis denoted as $dx^1(p),\dots,dx^d(p)$.
+The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with the corresponding *cotangent vectors* $z_p: T^\star_pM \rightarrow \mathbb{R}^d$ and a basis denoted as $dx^1(p),\dots,dx^d(p)$.
 
@@
 omit the subscript $p$ and refer to $g$ as the Riemannian metric.
-The inner product between two vectors $u, v \in T_pM$ is written as $\langle u, v \rangle_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed.
+The inner product between two vectors $u, v \in T_pM$ is written as $\langle u, v \rangle_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j} g_{ij}u^iv^j$, is assumed.
 A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
 
-The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{\langle u,v \rangle_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
+The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{\langle u,v \rangle_g}$, and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
 It could also be used to define the line element $dl^2 = g_{ij}dx^i dx^j$ and the volume element $dV_g = \sqrt{\det(g)}dx^1 \dots dx^d$, where $(x^1,\dots,x^d)$ are the local coordinates of the chart $(U, \varphi)$.
@@
 $$
-l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
+l(\gamma) = \sqrt{\int_0^1 \|\gamma^\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
 $$
@@
 $$
-Vol(W) = \int_W \sqrt{\det(g)}dx^1 \dots dx^d,
+Vol(W) = \int_W \sqrt{\det(g)}dx^1 \cdots dx^d,
 $$
@@
 $$ 
-where ${x^i}$ are the coordinates of the curve $\gamma$ and $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
+where $\{x^i\}$ are the coordinates of the curve $\gamma$ and $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
 $$
@@
 Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$.
-Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$.
-Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e.
-*diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$.
-The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit.
-The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$.
+Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$.
+The exponential map maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{\prime}(0)=u$, and $\gamma(1)=exp_p(u)$.
+It is a differentiable bijective map of differentiable inverse (i.e.
+*diffeomorphism*).
+Intuitively, the exponential map moves point $p$ to an endpoint  at speed $u$ after covering the length of $\|u\|$ along the geodesic in one time unit.
+
+The inverse of the exponential map is called the *logarithm map*, denoted as $\log_p(q):= \exp^{-1}_p(q)$, which gives the tangent vector to get from point $p$ to $q$ in one unit time.
 Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
 Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
+<!-- [TODO: reference in appendix]. -->
 
@@
 Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
-Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
-The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$,
+Let $\phi: M \rightarrow E$ be a smooth map between the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
+Then the differential of $\phi$ at point $p$ is a linear map $d\phi_p: T_pM \rightarrow T_{\phi (p)}E$, which pushes the tangent vector $u \in T_pM$ at point $p$ forward to the tangent vector $\phi_*u \in T_{\phi (p)}E$ at the mapping point $\phi(p)$.
+The image of the tangent vector $u \in T_pM$ under the differential $d\phi_p$, denoted as $d\phi_p u$ is called the pushforward of $u$ by the map $\phi$.
+Then pushforward metric $h=\varphi_*g$ of the Riemannian metric $g$ along $\varphi$ is given by the inner product
 $$
-\langle u,v \rangle_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
+\langle \phi_*u,\phi_*v \rangle_{\varphi*g} = \langle d\phi_p \phi_*u, d\phi_p \phi_*v \rangle_{g}.
 $$
-where $df_p^{-1}$ is the Jacobian of $f^{-1}$.
-The tangent vectors $f_\star u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$.
+
+<!--  is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u$ at point $p \in M$ to the tangent vectors $f_*u$ at the mapping point $f(p) \in E$.
+For two tangent vectors $u, v \in T_{f(p)}E$, we have
+$$
+\langle u,v \rangle_{\varphi*g} = \langle df_p^{-1}u, df_p^{-1}v \rangle_{g},
+$$ 
+where $df_p^{-1}$ is the Jacobian of $f^{-1}$.  -->
+
+The tangent vectors $\phi_*u$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{\prime}(0)=u$,
+$$
+d\phi_p(\gamma^{\prime}(0)) = (\phi \circ \gamma)^\prime (0).
+$$
 Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
+The linear map is called the pullback by $\phi$ and is often denoted as $\phi^*$.
 
@@
 
-<!-- ## Canonical metric -->
-
-<!-- For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional. -->
-
 ## Volume density function
 
-For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
-Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as
+For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$.
+Then $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
+Define the volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$, as
 $$
@@
 <!-- $$ --> 
-Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.
+<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
 The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
@@
 $$
+The volume density function is motivated by how a measure $du$ changes under the change of coordinates.
+It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
 
@@
 
-<!-- ## Laplace-Beltrami operator -->
-
-## Example: $S^1$
+## Example of $S^1$ for manifold density estimation
 
@@
 $$
+
 Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
@@
 and compare it with its true density $f(x)$.
-
-<!-- we could reduce the dimension to $d = 1$ and estimate the density of the 1-dimensional embedding$. -->
-
-<!-- First, we use the kernel density estimator in @ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->
-
 We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
@@
 <!-- \end{equation} --> 
-where $d=1$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
+where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
+
+<!-- we could reduce the dimension to $d = 1$ and estimate the density of the 1-dimensional embedding$. -->
+
+<!-- First, we use the kernel density estimator in @ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->
 
@@
 $$
-\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
+\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
 $$ 
@@
 $$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
 $$ 
-where the bandwidth $r$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.
+where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*.
 

### 2022-06-28 paper/paper.tex
Rewrite pushforward and pullback

@@
 For any point \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curves passing through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
-We say \(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart, \((U,\varphi)\), at \(p\) such that
+We say \(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart \((U,\varphi)\) at \(p\) such that
 \[
@@
 derivatives are denoted as \(\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)\), which defines a \emph{basis} of the tangent space.
-The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vectors}
-\(z_p: T^\star_pM \rightarrow \mathbb{R}^d\), and a basis denoted as \(dx^1(p),\dots,dx^d(p)\).
+The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with the corresponding \emph{cotangent vectors} \(z_p: T^\star_pM \rightarrow \mathbb{R}^d\) and a basis denoted as \(dx^1(p),\dots,dx^d(p)\).
 
@@
 omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
-The inner product between two vectors \(u, v \in T_pM\) is written as \(\langle u, v \rangle_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j}\), is assumed.
+The inner product between two vectors \(u, v \in T_pM\) is written as \(\langle u, v \rangle_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j} g_{ij}u^iv^j\), is assumed.
 A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
 
-The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{\langle u,v \rangle_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
+The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{\langle u,v \rangle_g}\), and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
 It could also be used to define the line element \(dl^2 = g_{ij}dx^i dx^j\) and the volume element \(dV_g = \sqrt{\det(g)}dx^1 \dots dx^d\), where \((x^1,\dots,x^d)\) are the local coordinates of the chart \((U, \varphi)\).
@@
 \[
-l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
+l(\gamma) = \sqrt{\int_0^1 \|\gamma^\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
 \]
@@
 \[
-Vol(W) = \int_W \sqrt{\det(g)}dx^1 \dots dx^d,
+Vol(W) = \int_W \sqrt{\det(g)}dx^1 \cdots dx^d,
 \]
@@
 The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold.
-Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold.
-\autocite[Section 7.2.3 of][]{Nakahara2018-zs}
+Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold \autocite[Section 7.2.3 of][]{Nakahara2018-zs}.
 A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
@@
 \]
-where \({x^i}\) are the coordinates of the curve \(\gamma\) and \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
+where \(\{x^i\}\) are the coordinates of the curve \(\gamma\) and \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
 \[
@@
 Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with raidus \(r\).
-Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\).
-Define \(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a differentiable, bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\).
-The exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint after covering the length of \(\|u\|\) along the geodesic in one time unit.
-The inverse of the exponential map is called the \emph{logarithm map}, \(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to \(q\).
+Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\), where \(exp_p\) is the \emph{exponential map} at point \(p\).
+The exponential map maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{\prime}(0)=u\), and \(\gamma(1)=exp_p(u)\).
+It is a differentiable bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}).
+Intuitively, the exponential map moves point \(p\) to an endpoint at speed \(u\) after covering the length of \(\|u\|\) along the geodesic in one time unit.
+
+The inverse of the exponential map is called the \emph{logarithm map}, denoted as \(\log_p(q):= \exp^{-1}_p(q)\), which gives the tangent vector to get from point \(p\) to \(q\) in one unit time.
 Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
@@
 Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
-The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
-\[
-\langle u,v \rangle_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
-\]
-where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
-The tangent vectors \(f_\star u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
+Let \(\phi: M \rightarrow E\) be a smooth map between the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
+Then the differential of \(\phi\) at point \(p\) is a linear map \(d\phi_p: T_pM \rightarrow T_{\phi (p)}E\), which pushes the tangent vector \(u \in T_pM\) at point \(p\) forward to the tangent vector \(\phi_*u \in T_{\phi (p)}E\) at the mapping point \(\phi(p)\).
+The image of the tangent vector \(u \in T_pM\) under the differential \(d\phi_p\), denoted as \(d\phi_p u\) is called the pushforward of \(u\) by the map \(\phi\).
+Then pushforward metric \(h=\varphi_*g\) of the Riemannian metric \(g\) along \(\varphi\) is given by the inner product
+\[
+\langle \phi_*u,\phi_*v \rangle_{\varphi*g} = \langle d\phi_p \phi_*u, d\phi_p \phi_*v \rangle_{g}.
+\]
+
+The tangent vectors \(\phi_*u\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{\prime}(0)=u\),
+\[
+d\phi_p(\gamma^{\prime}(0)) = (\phi \circ \gamma)^\prime (0).
+\]
 Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
+The linear map is called the pullback by \(\phi\) and is often denoted as \(\phi^*\).
 
@@
 
-For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
-Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
+For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\).
+Then \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
+Define the volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\), as
 \[
@@
 \]
-Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
 The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
@@
 \]
-
-\hypertarget{example-s1}{%
-\subsection{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}
+The volume density function is motivated by how a measure \(du\) changes under the change of coordinates.
+It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
+
+\hypertarget{example-of-s1-for-manifold-density-estimation}{%
+\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation}{Example of S\^{}1 for manifold density estimation}}\label{example-of-s1-for-manifold-density-estimation}}
 
@@
 \]
+
 Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
@@
 and compare it with its true density \(f(x)\).
-
 We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
@@
 \]
-where \(d=1\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
+where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
 
@@
 \[
-\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
+\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
 \]
@@
 \[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-\]
-where the bandwidth \(r\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+\]
+where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.
 

### 2022-07-05 paper/paper.Rmd
Rewrite kde for riem manifolds

@@
 abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
+However, kernel density estimates of the low-dimensional embe
+dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
 We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
@@
 
-# Kernel density estimation on manifolds {#vkde}
+# Kernel density estimation on manifold learning embeddings {#vkde}
 
@@
 
-## Multivariate kernel density estimation
+## Multivariate kernel density estimation {#mvkde}
 
@@
 More importantly, the estimator in Equation \@ref(eq:vkde) is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the underlying density is supported on a single manifold.
-The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
-This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
+In the cases of a Riemannian manifold, the manifold has a differentiable structure, which has to be taken into account when considering the neighborhood of a sample point.
+Therefore, we need to consider the probability density estimation on a Riemannian manifold and adapts the techiniques on Euclidean samples to a non-Euclidean setting.
+This is discussed in \autoref{riemest}.
 
@@
 Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
-Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$.
+Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the geodesic distance induced by the Riemannian metric $g$ and the strictly positive injectivity radius of the manifold $inj_g(M)$ [@Chavel2006-mp].
 Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathcal{R}^d$, as i.i.d.
 random objects on $M$ with density $f$.
-For each point $p\in M$, @Pelletier2005-vu proposed a kernel density estimator of $f$ to be
+To estimate the density of the manifold, @Pelletier2005-vu proposed a kernel density estimator for $f$ at point $\pmb{p}\in M$ as
 \begin{equation}
 \label{eq:denriem}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i)}{h}),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
 \end{equation} 
-where $h$ is the global bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
-@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q, $\exp_q^{-1}(p)$.
-The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as
-\begin{equation}
-\label{eq:denconsist}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
-\end{equation} 
-and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].
+where $r$ is the global bandwidth satisfying $r \leq r_0$ for fixed $r_0$ such that $0\leq r_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{\pmb{X}_i}(\pmb{p})$ is the volume density function on the manifold for point $\pmb{X}_i$.
+The estimator in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case in Equation \@ref(eq:kde) and converges at the same rate as the Euclidean estimator [@Henry2009-ll].
 
-### Volume density function
+Denote $\lambda$ as the Lebesque measure of $\mathbb{R}^d$.
+The kernels $K$ in \@ref(eq:denriem) are functions of the geodesic distance of the manifold $d_g$ and they satisfy the following conditions [@Pelletier2005-vu]:
+\begin{align}
+& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
+(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
+(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
+& (iv) \supp K=[0; 1];
+(v) \sup K(x)=K(0).
+\end{align}
+Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, $r\leq inj_g(M)$.
+Under such condition, for each $\pmb{p} \in M$, there exists a normal coordinate neighborhood of $\pmb{p}$ containing an open ball $B(\pmb{p}, r) \in T_pM$ centered at $\pmb{p}$ with radius $r$.
+
+Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
+For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denotes a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
+Then $\frac{Dw_1}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$.
+The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined as 
+$$
+\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+$$
+
+For the exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives
+$$
+d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
+$$
+in normal coordinates.
+The volume density function is motivated by how a measure $du$ changes under the change of coordinates.
+It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
+@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.
+
+<!-- $$ -->
+<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
+<!-- $$ --> 
+<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
+<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->
+
+# Distotion-corrected kernel density estimator (DC-KDE) {#dckde}
 
@@
 where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.
+
+<!-- The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
+This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
+-->
 
@@
 
-## Volume density function
+## Volume density function [MOVE TO SECTION 2.2.1]
 

### 2022-07-05 paper/paper.tex
Rewrite kde for riem manifolds

@@
 \makeatother
-\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}
+\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}
 
@@
 \usepackage{bm}
+\DeclareMathOperator\supp{supp}
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
-
-% Adjust headwidth in case user has changed geometry in header-includes
-\renewcommand{\headwidth}{\textwidth}
 
@@
 \hypertarget{vkde}{%
-\section{Kernel density estimation on manifolds}\label{vkde}}
+\section{Kernel density estimation on manifold learning embeddings}\label{vkde}}
 
@@
 
-\hypertarget{multivariate-kernel-density-estimation}{%
-\subsection{Multivariate kernel density estimation}\label{multivariate-kernel-density-estimation}}
+\hypertarget{mvkde}{%
+\subsection{Multivariate kernel density estimation}\label{mvkde}}
 
@@
 More importantly, the estimator in Equation \eqref{eq:vkde} is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the underlying density is supported on a single manifold.
-The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
-This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
+In the cases of a Riemannian manifold, the manifold has a differentiable structure, which has to be taken into account when considering the neighborhood of a sample point.
+Therefore, we need to consider the probability density estimation on a Riemannian manifold and adapts the techiniques on Euclidean samples to a non-Euclidean setting.
+This is discussed in \autoref{riemest}.
 
@@
 Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
-Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\).
+Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}.
 Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d.
 random objects on \(M\) with density \(f\).
-For each point \(p\in M\), \textcite{Pelletier2005-vu} proposed a kernel density estimator of \(f\) to be
+To estimate the density of the manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
 \begin{equation}
 \label{eq:denriem}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i)}{h}),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
 \end{equation}
-where \(h\) is the global bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
-\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q, \(\exp_q^{-1}(p)\).
-The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
-\begin{equation}
-\label{eq:denconsist}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
-\end{equation}
-and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.
-
-\hypertarget{volume-density-function}{%
-\subsubsection{Volume density function}\label{volume-density-function}}
+where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\).
+The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case in Equation \eqref{eq:kde} and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.
+
+Denote \(\lambda\) as the Lebesque measure of \(\mathbb{R}^d\).
+The kernels \(K\) in \eqref{eq:denriem} are functions of the geodesic distance of the manifold \(d_g\) and they satisfy the following conditions \autocite{Pelletier2005-vu}:
+\begin{align}
+& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
+(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
+(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
+& (iv) \supp K=[0; 1];
+(v) \sup K(x)=K(0).
+\end{align}
+Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, \(r\leq inj_g(M)\).
+Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_pM\) centered at \(\pmb{p}\) with radius \(r\).
+
+Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
+For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denotes a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
+Then \(\frac{Dw_1}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
+The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined as
+\[
+\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+\]
+
+For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
+\[
+d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
+\]
+in normal coordinates.
+The volume density function is motivated by how a measure \(du\) changes under the change of coordinates.
+It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
+\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
+
+\hypertarget{dckde}{%
+\section{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}
 
@@
 
-\hypertarget{volume-density-function-1}{%
-\subsection{Volume density function}\label{volume-density-function-1}}
+\hypertarget{volume-density-function-move-to-section-2.2.1}{%
+\subsection{Volume density function {[}MOVE TO SECTION 2.2.1{]}}\label{volume-density-function-move-to-section-2.2.1}}
 

### 2022-07-05 paper/paper.Rmd
Move swiss roll to appendix b

@@
 
-# Kernel density estimation on manifold learning embeddings {#vkde}
+# Kernel density estimate on Riemannian manifold embeddings {#vkde}
 
@@
 
-# Distotion-corrected kernel density estimator (DC-KDE) {#dckde}
+## Distotion-corrected kernel density estimator (DC-KDE) {#dckde}
 
-However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
+However, in application, we usually find the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
 Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
@@
 
-### Riemannian metric as variable bandwidth
+## Riemannian metric estimation
 
@@
 
-# Outlier detection on manifold learning embedding
+## Outlier detection on manifold learning embedding
 
@@
 
-### Swiss roll mapping
-
 ```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors and shapes indicating the four kernels.
@@
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
-The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
-\begin{equation}
-\label{eq:swissroll}
-\left\{
-\begin{array}{lcl}
-X = X_1 \cos{X_1}, \\
-Y = X_2, \\
-Z = X_1 \sin{X_1}.
-\end{array}
-\right.
-\end{equation}
-
-```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
-knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
-```
-
-Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1%, 5%, 50%, 99%, \>99%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
-the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all
-manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
-
-```{r srcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_Swiss Roll_4ml_riem0_08.rda")
-col.max <- apply(cors, 2, max)
-max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
-  mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  # kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = FALSE) %>%
-  # column_spec(1, width = "6cm") %>% 
-  column_spec(2, bold = max.match[,1]) %>%
-  column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4]) %>%
-  column_spec(6, bold = max.match[,5])
-```
-
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
-rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
-correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
-
-<!-- Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot has indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along with the curved embedding.
--->
-
-### Twin peaks mapping
-
-For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
+We use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
 mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
+For simplicity, the simulation results for the swiss roll mapping in te left plot is shown in the Appendix \autoref{swissrollappe}.
 \begin{equation}
@@
 
-\clearpage\appendix
+<!-- \clearpage -->
 
@@
 <!-- %\def\thepage{A\arabic{page}} -->
+
+\appendix
 
@@
 
-<!-- ### Isometry -->
+<!-- ## Isometry -->
 
@@
 
-## Volume density function [MOVE TO SECTION 2.2.1]
+<!-- ## Volume density function [MOVED TO SECTION 2.2]
 
@@
 $$
-<!-- $$ -->
-<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
-<!-- $$ --> 
-<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
+
 The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
@@
 It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
+-->
 
@@
 
-## Example of $S^1$ for manifold density estimation
+## Example of $S^1$ for manifold density estimation [TODO: rewrite as appendix B with plots?]
 
@@
 
-<!-- Polar coordinates: -->
-<!-- $\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v$. -->
+\clearpage
 
-<!-- Cartesian coordinates: -->
-<!-- $$ -->
-<!-- \psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\ -->
-<!-- = (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\ -->
-<!-- = - \sqrt{1 - p^2}v -->
-<!-- $$ -->
-<!-- Then the volume density function is -->
-<!-- $$ -->
-<!-- \theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2}, -->
-<!-- $$ -->
-<!-- and the geodesic distance is -->
-<!-- $$ -->
-<!-- d_g(p,q) = \arccos(p) - \arccos(q). -->
-<!-- $$ -->
+# Appendix: Simulation with swiss roll mapping {#swissrollappe}
+
+In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.
+
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
+The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
+The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
+\begin{equation}
+\label{eq:swissroll}
+\left\{
+\begin{array}{lcl}
+X = X_1 \cos{X_1}, \\
+Y = X_2, \\
+Z = X_1 \sin{X_1}.
+\end{array}
+\right.
+\end{equation}
+
+```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
+Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
+knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
+```
+
+Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
+According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
+1%, 5%, 50%, 99%, \>99%.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
+the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
+From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
+However, in all
+manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
+For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
+
+```{r srcors, echo=FALSE, message=FALSE, eval=T}
+load("figures/CorrelationTable_Swiss Roll_4ml_riem0_08.rda")
+col.max <- apply(cors, 2, max)
+max.match <- rep(col.max, each = 2) == cors
+data.frame(cors) %>% 
+  mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
+  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  # kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = FALSE) %>%
+  # column_spec(1, width = "6cm") %>% 
+  column_spec(2, bold = max.match[,1]) %>%
+  column_spec(3, bold = max.match[,2]) %>%
+  column_spec(4, bold = max.match[,3]) %>%
+  column_spec(5, bold = max.match[,4]) %>%
+  column_spec(6, bold = max.match[,5])
+```
+
+To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
+It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
+rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
+correlation.
+The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
+As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
+This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
+
+<!-- Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
+As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
+Instead, we could roughly compare the location of outliers.
+For t-SNE, the middle plot has indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along with the curved embedding.
+-->

### 2022-07-05 paper/paper.tex
Move swiss roll to appendix b

@@
 Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
+However, kernel density estimates of the low-dimensional embe dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
 We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
@@
 \hypertarget{vkde}{%
-\section{Kernel density estimation on manifold learning embeddings}\label{vkde}}
+\section{Kernel density estimate on Riemannian manifold embeddings}\label{vkde}}
 
@@
 \hypertarget{dckde}{%
-\section{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}
-
-However, in application, we usually found the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
+\subsection{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}
+
+However, in application, we usually find the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
 Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
@@
 
-\hypertarget{riemannian-metric-as-variable-bandwidth}{%
-\subsubsection{Riemannian metric as variable bandwidth}\label{riemannian-metric-as-variable-bandwidth}}
+\hypertarget{riemannian-metric-estimation}{%
+\subsection{Riemannian metric estimation}\label{riemannian-metric-estimation}}
 
@@
 \hypertarget{outlier-detection-on-manifold-learning-embedding}{%
-\section{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}
+\subsection{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}
 
@@
 
-\hypertarget{swiss-roll-mapping}{%
-\subsubsection{Swiss roll mapping}\label{swiss-roll-mapping}}
-
 \begin{figure}
@@
 Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
-The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
-\begin{equation}
-\label{eq:swissroll}
-\left\{
-\begin{array}{lcl}
-X = X_1 \cos{X_1}, \\
-Y = X_2, \\
-Z = X_1 \sin{X_1}.
-\end{array}
-\right.
-\end{equation}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
-
-}
-
-\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
-\end{figure}
-
-Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1\%, 5\%, 50\%, 99\%, \textgreater99\%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
-the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all
-manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
-
-\begin{table}
-
-\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
-\midrule
-Variable bandwidth & 0.0696 & \textbf{0.400} & -0.2357 & 0.023 & \textbf{0.0138}\\
-Fixed bandwidth & \textbf{0.2798} & 0.351 & \textbf{0.0141} & \textbf{0.367} & -0.0110\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
-rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
-correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
-
-\hypertarget{twin-peaks-mapping}{%
-\subsubsection{Twin peaks mapping}\label{twin-peaks-mapping}}
-
-For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
+We use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
 mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
+For simplicity, the simulation results for the swiss roll mapping in te left plot is shown in the Appendix \autoref{swissrollappe}.
 \begin{equation}
@@
 
-\clearpage\appendix
+\appendix
 
@@
 
-\hypertarget{volume-density-function-move-to-section-2.2.1}{%
-\subsection{Volume density function {[}MOVE TO SECTION 2.2.1{]}}\label{volume-density-function-move-to-section-2.2.1}}
-
-For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\).
-Then \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
-Define the volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\), as
-\[
-\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-\]
-The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
-\[
-d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
-\]
-The volume density function is motivated by how a measure \(du\) changes under the change of coordinates.
-It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
-
-\hypertarget{example-of-s1-for-manifold-density-estimation}{%
-\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation}{Example of S\^{}1 for manifold density estimation}}\label{example-of-s1-for-manifold-density-estimation}}
+\hypertarget{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}{%
+\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}{Example of S\^{}1 for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}}\label{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}}
 
@@
 
+\clearpage
+
+\hypertarget{swissrollappe}{%
+\section{Appendix: Simulation with swiss roll mapping}\label{swissrollappe}}
+
+In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.
+
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
+The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
+The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
+\begin{equation}
+\label{eq:swissroll}
+\left\{
+\begin{array}{lcl}
+X = X_1 \cos{X_1}, \\
+Y = X_2, \\
+Z = X_1 \sin{X_1}.
+\end{array}
+\right.
+\end{equation}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
+
+}
+
+\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data.
+Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
+Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
+The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
+\end{figure}
+
+Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
+According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
+1\%, 5\%, 50\%, 99\%, \textgreater99\%.
+For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
+the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
+The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
+From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
+However, in all
+manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
+When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
+In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
+For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
+
+\begin{table}
+
+\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
+Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
+\centering
+\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
+\toprule
+  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
+\midrule
+Variable bandwidth & 0.0696 & \textbf{0.400} & -0.2357 & 0.023 & \textbf{0.0138}\\
+Fixed bandwidth & \textbf{0.2798} & 0.351 & \textbf{0.0141} & \textbf{0.367} & -0.0110\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
+It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
+The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
+rectangular structure of the meta data shown in \autoref{fig:metadensity}.
+For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
+correlation.
+The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
+As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
+This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
+However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
+
 \printbibliography

### 2022-07-05 paper/paper.Rmd
Change title, replace Figure 5 with density ranks

@@
 ---
-title: "Anomaly detection with kernel density estimation on manifolds"
+title: "Kernel density estimate for embedding of Riemannian manifolds"
 author:
@@
 abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embe
-dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
+However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
 We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
@@
 # library(copula)
-library(plotly)
+# library(plotly)
 library(kableExtra)
@@
 
-```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
 The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
 Variable bandwidth shows a strong linear positive relationship.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1.png")
+knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1_rank.png")
 ```

### 2022-07-05 paper/paper.tex
Change title, replace Figure 5 with density ranks

@@
 \hypersetup{
-            pdftitle={Anomaly detection with kernel density estimation on manifolds},
+            pdftitle={Kernel density estimate for embedding of Riemannian manifolds},
             pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
@@
 
-\title{Anomaly detection with kernel density estimation on manifolds}
+\title{Kernel density estimate for embedding of Riemannian manifolds}
 
@@
 Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embe dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
+However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
 We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1}
-
-}
-
-\caption{Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth.
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1_rank}
+
+}
+
+\caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
 The four colors and shapes represents the four gaussian kernels in the 2-D meta data.

### 2022-07-06 paper/paper.tex
Change title

@@
 \hypersetup{
-            pdftitle={Kernel density estimate for embedding of Riemannian manifolds},
+            pdftitle={Distortion-corrected kernel density estimate for Riemannian manifolds},
             pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
@@
 
-\title{Kernel density estimate for embedding of Riemannian manifolds}
+\title{Distortion-corrected kernel density estimate for Riemannian manifolds}
 
@@
 The outliers could then be defined as the points with the lowest density estimates.
-The proposed schematic is shown in Figure \ref{fig:vkde}.
-The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
-variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/vkde}
-
-}
-
-\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
-\end{figure}
 
@@
 \textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
+Note that this is true only when using the geodesic normal coordinates.
+If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise.
+Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
 
 \hypertarget{dckde}{%
-\subsection{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}
-
-However, in application, we usually find the observations lie on a \(d\)-dimensional manifold embedded in a \(s\)-dimensional space, where \(d \ll s\).
-Then manifold learning can be applies to find the \(d\)-dimensional representation of the manifold \(M\).
-Therefore, based on the estimator in \eqref{eq:denriem}, we propose a kernel density estimator of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
+\subsection{Distortion-corrected kernel density estimator for manifold embedding}\label{dckde}}
+
+However, in application, we usually find the observations lying on a \(d\)-dimensional manifold embedded in a much higher \(s\)-dimensional space, where \(d \ll s\).
+Instead of sampling from the undering manifold \(M\), we apply manifold learning algorithms to reduce dimension and find a \(d\)-dimensional representation of the manifold \(M\).
+Notice that this representation is not necessarily isometry to \(M\), and the underlying manifold is usually unknown in real data sets.
+Then the density estimator in Equation \eqref{eq:denriem} proposed by \textcite{Pelletier2005-vu} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
+Motivated on the estimator in \eqref{eq:denriem}, we propose a distortion-corrected kernel density estimator (DC-KDE) of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r} \frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|} K\bigg( \frac{(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})}{r} \bigg),
 \end{equation}
-where \(h_i, i = 1, 2, \dots, N\) is the pointwise variable bandwidth matrix and \(h_i\) is positive definite with dimension \(d\times d\).
+where \(h(\pmb{y_i}), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and \(h(x)\) is positive definite with dimension \(d\times d\), \(\frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})\) is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point \(\pmb{y_i} \ \mathbb{R}^d\).
+Then the only problem left is the estimation of the Riemannian metric at each embedding point.
 
@@
 1\%, 50\%, 99\%, \textgreater99\%.
+
+The proposed schematic is shown in Figure \ref{fig:vkde}.
+The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
+variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/vkde}
+
+}
+
+\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
+\end{figure}
 

### 2022-07-06 paper/paper.Rmd
Change title

@@
 ---
-title: "Kernel density estimate for embedding of Riemannian manifolds"
+title: "Distortion-corrected kernel density estimate for Riemannian manifolds"
 author:
@@
 The outliers could then be defined as the points with the lowest density estimates.
-The proposed schematic is shown in Figure \@ref(fig:vkde).
-The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
-variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
-
-(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
-
-```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
-knitr::include_graphics("figures/vkde.png")
-```
 
@@
 @Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.
+Note that this is true only when using the geodesic normal coordinates.
+If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise.
+Therefore, the Riemannian manifold density estimator in @Pelletier2005-vu only works when the true manifold $M$ is known.
 
@@
 
-## Distotion-corrected kernel density estimator (DC-KDE) {#dckde}
+## Distortion-corrected kernel density estimator for manifold embedding {#dckde}
 
-However, in application, we usually find the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$.
-Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$.
-Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be
+However, in application, we usually find the observations lying on a $d$-dimensional manifold embedded in a much higher $s$-dimensional space, where $d \ll s$.
+Instead of sampling from the undering manifold $M$, we apply manifold learning algorithms to reduce dimension and find a $d$-dimensional representation of the manifold $M$.
+Notice that this representation is not necessarily isometry to $M$, and the underlying manifold is usually unknown in real data sets.
+Then the density estimator in Equation \@ref(eq:denriem) proposed by @Pelletier2005-vu no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
+Motivated on the estimator in \@ref(eq:denriem), we propose a distortion-corrected kernel density estimator (DC-KDE) of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r} \frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|} K\bigg( \frac{(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})}{r} \bigg),
 \end{equation} 
-where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.
+where $h(\pmb{y_i}), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and $h(x)$ is positive definite with dimension $d\times d$, $\frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})$ is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point $\pmb{y_i} \ \mathbb{R}^d$.
+Then the only problem left is the estimation of the Riemannian metric at each embedding point.
 
@@
 1%, 50%, 99%, \>99%.
+
+The proposed schematic is shown in Figure \@ref(fig:vkde).
+The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
+variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
+
+(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
+
+```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
+knitr::include_graphics("figures/vkde.png")
+```
 

### 2022-07-12 paper/paper.Rmd
Sec 2 restructure

@@
 ---
-title: "Distortion-corrected kernel density estimate for Riemannian manifolds"
+title: "Distortion-corrected kernel density estimate on Riemannian manifolds"
 author:
@@
 
-Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
+<!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.  -->
+Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
+It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
 It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
@@
 The rest of the paper is organized as follows.
-In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings.
-In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
-\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
-\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
+In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
+We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds.
+Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator to detect anomalies based on the embedding.
+\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
+\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset.
 Conclusions and discussions are presented in \autoref{conclusion}.
@@
 
-# Kernel density estimate on Riemannian manifold embeddings {#vkde}
+# Kernel density estimate on Riemannian manifold embedding {#kderm}
 
-In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
+<!-- In this section, we introduce the proposed kernel density estimator on the embedding of Riemannian manifolds.  -->
+For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data.
+@Pelletier2005-vu generalize the multivariate kernel density esitmator to Riemnnian manifolds.
+However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space.
+Therefore, we propose a distortion-corrected kernel density estimator as in Equation \@ref(eq:denestimator0) based on the low-dimensional embedding, $\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, of the high-dimensional data with $N$ data points,
+\begin{equation}
+\label{eq:denestimator0}
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\end{equation} 
+where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise adaptive Riemannian matrix of dimension $d \times d$.
+The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions.
+These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
+
+<!-- to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
 @Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
@@
 The outliers could then be defined as the points with the lowest density estimates.
+-->
 
-To start with, we introduce the notations used in this manuscript.
-Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
-Readers familiar with these topics could skip the corresponding subsections.
-Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
+To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's estimator for Riemannian manifolds.
+Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm in @Perrault-Joncas2013-pq.
+Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on a Riemannian manifold densities.
 
@@
 It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
-@Scott1992-cj and @Scott2015-vl cover a wide range of techniques for estimating the kernel densities of multivariate data.
+@Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
 In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
 <!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
-drawn from a density $f$, the kernel density estimate of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
+drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
 \begin{equation}
-\label{eq:kde}
-\hat{f}(\pmb{x};\pmb{H})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{X}_i),
+\label{eq:vkde}
+\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
 \end{equation}
-where $K(\pmb{x})$ is the kernel function, $\pmb{H}$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})$.
+where $K(\pmb{x})$ is the kernel function, $\pmb{H}_i$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})$.
 When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
@@
 \label{eq:kernel}
-K_{\pmb{H}}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{X}_i)\right],
+K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
 \end{equation} 
-with $|\pmb{H}|$ being the determinant of matrix $\pmb{H}$.
+with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
 Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function.
@@
 
-The choice of the bandwidth matrix $\pmb{H}$ is crucial in the accuracy of $\hat{f}$ because it controls the smoothing across the sample data.
+The choice of the bandwidth matrix $\pmb{H}_i$ is crucial in the accuracy of $\hat{f}$ because it controls the smoothing across the sample data.
 For a given kernel function $K(\pmb{x})$, the shape of the local smoothing at each point is determined.
-Then the performance of the density estimate $\hat{f}$ depends on the choice of $\pmb{H}$.
-When $\pmb{H}$ is too small, the density estimates will apprear to be wiggly.
-On the contrary, when $\pmb{H}$ is too large, the density estimates might smooth out some local features of $f$.
+Then the performance of the density estimate $\hat{f}$ depends on the choice of $\pmb{H}_i$.
+When $\pmb{H}_i$ is too small, the density estimates will apprear to be wiggly.
+On the contrary, when $\pmb{H}_i$ is too large, the density estimates might smooth out some local features of $f$.
 So the accuracy of $\hat{f}$ depends strongly on the bandwidth matrix.
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
+Compared to a global fixed $\pmb{H}$, a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
+
+<!-- Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
 For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
@@
 In this paper, we focus on the multivariate kernel density estimation.
+ -->
 
-Note that the bandwidth matrix in \@ref(eq:kde) is a global smoothing parameter for all data points.
+<!-- Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points.
 However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-This is ususally referred to as variable kernel density estimation [VKDE; @Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut], where the bandwidth is varied depending on either the location of the sample points or that of the estimated points [Section 6.6 of @Scott2015-vl].
+The bandwidth is varied depending on either the location of the sample points or that of the estimated points [Section 6.6 of @Scott2015-vl].
 In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point $\pmb{X}_i$ and will refer to this as the *variable/adaptive kernel density estimation(VKDE)* unless otherwise stated.
-
-In multivariate VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $\pmb{X}_i$,
+ -->
+<!-- In multivariate VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $\pmb{X}_i$,
 \begin{equation}
@@
 \hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
-\end{equation}
-Compared to a global fixed $\pmb{H}$, a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
+\end{equation} -->
 
@@
 where $r$ is the global bandwidth satisfying $r \leq r_0$ for fixed $r_0$ such that $0\leq r_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{\pmb{X}_i}(\pmb{p})$ is the volume density function on the manifold for point $\pmb{X}_i$.
-The estimator in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case in Equation \@ref(eq:kde) and converges at the same rate as the Euclidean estimator [@Henry2009-ll].
+The estimator in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator [@Henry2009-ll].
 
@@
 Then the density estimator in Equation \@ref(eq:denriem) proposed by @Pelletier2005-vu no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
-Motivated on the estimator in \@ref(eq:denriem), we propose a distortion-corrected kernel density estimator (DC-KDE) of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
+Motivated on the estimator in \@ref(eq:denriem), we propose a distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), with the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$ to be
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r} \frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|} K\bigg( \frac{(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation} 
-where $h(\pmb{y_i}), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and $h(x)$ is positive definite with dimension $d\times d$, $\frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})$ is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point $\pmb{y_i} \ \mathbb{R}^d$.
+where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point $\pmb{y}_i \in \mathbb{R}^d$.
 Then the only problem left is the estimation of the Riemannian metric at each embedding point.
@@
 
-The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $<,>_g$ on the tangent space $T_pM$ for every point $p \in M$.
+The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_pM$ for every point $p \in M$.
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.

### 2022-07-12 paper/paper.tex
Sec 2 restructure

@@
 \hypersetup{
-            pdftitle={Distortion-corrected kernel density estimate for Riemannian manifolds},
+            pdftitle={Distortion-corrected kernel density estimate on Riemannian manifolds},
             pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
@@
 
-\title{Distortion-corrected kernel density estimate for Riemannian manifolds}
+\title{Distortion-corrected kernel density estimate on Riemannian manifolds}
 
@@
 
-Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
+Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
+It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
 It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
@@
 
-\hypertarget{vkde}{%
-\section{Kernel density estimate on Riemannian manifold embeddings}\label{vkde}}
-
-In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-\textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
-To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with the lowest density estimates.
-
-To start with, we introduce the notations used in this manuscript.
-Then we introduce the multivariate kernel density estimation method with variable bandwidth matrix and the metric learning algorithm to derive the pointwise Riemannian metric.
-Readers familiar with these topics could skip the corresponding subsections.
-Finally, we propose our novel algorithm to estimate densities and detect anomalies for high-dimensional data set.
+\hypertarget{kderm}{%
+\section{Kernel density estimate on Riemannian manifold embedding}\label{kderm}}
+
+For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data.
+\textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds.
+However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space.
+Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data with \(N\) data points,
+\begin{equation}
+\label{eq:denestimator0}
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\end{equation}
+where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
+The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions.
+These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
+
+To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's estimator for Riemannian manifolds.
+Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm in \textcite{Perrault-Joncas2013-pq}.
+Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on a Riemannian manifold densities.
 
@@
 It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
-\textcite{Scott1992-cj} and \textcite{Scott2015-vl} cover a wide range of techniques for estimating the kernel densities of multivariate data.
+\textcite{Scott2015-vl} covers a wide range of techniques for estimating the kernel densities of multivariate data.
 In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
-drawn from a density \(f\), the kernel density estimate of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
-\begin{equation}
-\label{eq:kde}
-\hat{f}(\pmb{x};\pmb{H})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{X}_i),
-\end{equation}
-where \(K(\pmb{x})\) is the kernel function, \(\pmb{H}\) is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension \(d\times d\), and \(K_{\pmb{H}}(\pmb{x}) = |\pmb{H}|^{-1/2} K(\pmb{H}^{-1/2}\pmb{x})\).
-When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
-\begin{equation}
-\label{eq:kernel}
-K_{\pmb{H}}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{X}_i)\right],
-\end{equation}
-with \(|\pmb{H}|\) being the determinant of matrix \(\pmb{H}\).
-Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function.
-Then an average over all data points is taken as the estimated density \(\hat{f}\).
-KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point \(\pmb{x}\).
-
-The choice of the bandwidth matrix \(\pmb{H}\) is crucial in the accuracy of \(\hat{f}\) because it controls the smoothing across the sample data.
-For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined.
-Then the performance of the density estimate \(\hat{f}\) depends on the choice of \(\pmb{H}\).
-When \(\pmb{H}\) is too small, the density estimates will apprear to be wiggly.
-On the contrary, when \(\pmb{H}\) is too large, the density estimates might smooth out some local features of \(f\).
-So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
-The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
-In this paper, we focus on the multivariate kernel density estimation.
-
-Note that the bandwidth matrix in \eqref{eq:kde} is a global smoothing parameter for all data points.
-However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-This is ususally referred to as variable kernel density estimation {[}VKDE; \textcite{Breiman1977-qc}; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]}, where the bandwidth is varied depending on either the location of the sample points or that of the estimated points \autocite[Section 6.6 of][]{Scott2015-vl}.
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point \(\pmb{X}_i\) and will refer to this as the \emph{variable/adaptive kernel density estimation(VKDE)} unless otherwise stated.
-
-In multivariate VKDE, the estimator in Equation \eqref{eq:vkde} is similar to that in \eqref{eq:kde} except that the bandwidth is a variable matrix \(\pmb{H}_i\) for each \(\pmb{X}_i\),
+drawn from a density \(f\), the variable kernel density estimate \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut} of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
 \begin{equation}
@@
 \end{equation}
+where \(K(\pmb{x})\) is the kernel function, \(\pmb{H}_i\) is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension \(d\times d\), and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})\).
+When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
+\begin{equation}
+\label{eq:kernel}
+K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
+\end{equation}
+with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
+Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function.
+Then an average over all data points is taken as the estimated density \(\hat{f}\).
+KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point \(\pmb{x}\).
+
+The choice of the bandwidth matrix \(\pmb{H}_i\) is crucial in the accuracy of \(\hat{f}\) because it controls the smoothing across the sample data.
+For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined.
+Then the performance of the density estimate \(\hat{f}\) depends on the choice of \(\pmb{H}_i\).
+When \(\pmb{H}_i\) is too small, the density estimates will apprear to be wiggly.
+On the contrary, when \(\pmb{H}_i\) is too large, the density estimates might smooth out some local features of \(f\).
+So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
 Compared to a global fixed \(\pmb{H}\), a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
@@
 where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\).
-The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case in Equation \eqref{eq:kde} and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.
+The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.
 
@@
 Then the density estimator in Equation \eqref{eq:denriem} proposed by \textcite{Pelletier2005-vu} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
-Motivated on the estimator in \eqref{eq:denriem}, we propose a distortion-corrected kernel density estimator (DC-KDE) of \(f\) with the manifold embedding \(y_i, i = 1, 2, \dots, N\) to be
+Motivated on the estimator in \eqref{eq:denriem}, we propose a distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, with the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\) to be
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r} \frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|} K\bigg( \frac{(\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation}
-where \(h(\pmb{y_i}), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and \(h(x)\) is positive definite with dimension \(d\times d\), \(\frac{|h(\pmb{p})|}{|h(\pmb{y_i}))|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y_i})^\prime h^{-1}(\pmb{y_i}) (\pmb{p} - \pmb{y_i})\) is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point \(\pmb{y_i} \ \mathbb{R}^d\).
+where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point \(\pmb{y}_i \in \mathbb{R}^d\).
 Then the only problem left is the estimation of the Riemannian metric at each embedding point.
@@
 
-The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_pM\) for every point \(p \in M\).
+The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_pM\) for every point \(p \in M\).
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.

### 2022-07-15 paper/paper.tex
More on DC_KDE

@@
 However, in application, we usually find the observations lying on a \(d\)-dimensional manifold embedded in a much higher \(s\)-dimensional space, where \(d \ll s\).
-Instead of sampling from the undering manifold \(M\), we apply manifold learning algorithms to reduce dimension and find a \(d\)-dimensional representation of the manifold \(M\).
+Instead of sampling from the underlying manifold \(M\), we could only sample from the high-dimensional data and apply dimension reduction techniques such as manifold learning to find a \(d\)-dimensional representation of \(M\).
 Notice that this representation is not necessarily isometry to \(M\), and the underlying manifold is usually unknown in real data sets.
-Then the density estimator in Equation \eqref{eq:denriem} proposed by \textcite{Pelletier2005-vu} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
-Motivated on the estimator in \eqref{eq:denriem}, we propose a distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, with the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\) to be
+We could use the low-dimensional embedding as an approximation of the true manifold \(M\).
+However, the pointwise distortion induced by the manifold learning process needs to be taken care of to accurately estimate the density of the manifold.
+Therefore, the density estimator in Equation \eqref{eq:denriem} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
+Further, the bandwidth matrix should be adaptive to each sample point to take into account the distortion.
+Motivated by the estimator in \eqref{eq:denriem}, we restate the proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
 \begin{equation}

### 2022-07-15 paper/paper.Rmd
More on DC_KDE

@@
 However, in application, we usually find the observations lying on a $d$-dimensional manifold embedded in a much higher $s$-dimensional space, where $d \ll s$.
-Instead of sampling from the undering manifold $M$, we apply manifold learning algorithms to reduce dimension and find a $d$-dimensional representation of the manifold $M$.
+Instead of sampling from the underlying manifold $M$, we could only sample from the high-dimensional data and apply dimension reduction techniques such as manifold learning to find a $d$-dimensional representation of $M$.
 Notice that this representation is not necessarily isometry to $M$, and the underlying manifold is usually unknown in real data sets.
-Then the density estimator in Equation \@ref(eq:denriem) proposed by @Pelletier2005-vu no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
-Motivated on the estimator in \@ref(eq:denriem), we propose a distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), with the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$ to be
+We could use the low-dimensional embedding as an approximation of the true manifold $M$.
+However, the pointwise distortion induced by the manifold learning process needs to be taken care of to accurately estimate the density of the manifold.
+Therefore, the density estimator in Equation \@ref(eq:denriem) no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
+Further, the bandwidth matrix should be adaptive to each sample point to take into account the distortion.
+Motivated by the estimator in \@ref(eq:denriem), we restate the proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
 \begin{equation}

### 2022-07-16 paper/paper.Rmd
More on DC_KDE

@@
 
-However, in application, we usually find the observations lying on a $d$-dimensional manifold embedded in a much higher $s$-dimensional space, where $d \ll s$.
-Instead of sampling from the underlying manifold $M$, we could only sample from the high-dimensional data and apply dimension reduction techniques such as manifold learning to find a $d$-dimensional representation of $M$.
-Notice that this representation is not necessarily isometry to $M$, and the underlying manifold is usually unknown in real data sets.
+In application, we usually find the observations lying on a $d$-dimensional manifold embedded in a much higher $s$-dimensional space, where $d \ll s$.
+Instead of sampling from the underlying manifold $M$, we could sample from the high-dimensional data only and apply dimension reduction techniques such as manifold learning to find a $d$-dimensional representation of $M$.
+Notice that this representation is not necessarily isometry to $M$, and the underlying manifold tends to be unknown in real data sets.
 We could use the low-dimensional embedding as an approximation of the true manifold $M$.
 However, the pointwise distortion induced by the manifold learning process needs to be taken care of to accurately estimate the density of the manifold.
-Therefore, the density estimator in Equation \@ref(eq:denriem) no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
-Further, the bandwidth matrix should be adaptive to each sample point to take into account the distortion.
-Motivated by the estimator in \@ref(eq:denriem), we restate the proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
+Therefore, the kernel density estimator in Equation \@ref(eq:denriem) no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
+Further, the bandwidth matrix should be adaptive to each sample point to take the distortion into account.
+
+Motivated by the estimator in \@ref(eq:denriem), we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
 \begin{equation}
@@
 \end{equation}
-where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point $\pmb{y}_i \in \mathbb{R}^d$.
-Then the only problem left is the estimation of the Riemannian metric at each embedding point.
+where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}^{-1}(\pmb{y}_i)$.
+Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{#riemmetric}.
+
+[TODO: How to derive the volume density function?]
+
+[TODO: Further fix the approximation induced by the estimated d_g so that f will integral to 1.]
 
@@
 -->
+
+## Riemannian metric estimation {#riemmetric}
 
@@
 In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
-
-## Riemannian metric estimation
 

### 2022-07-18 paper/paper.Rmd
Derive volume density function as ratio

@@
 Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
-For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denotes a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
+For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denote a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
 Then $\frac{Dw_1}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$.
 The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined as 
-$$
+\begin{equation}
+\label{eq:volden}
 \theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-$$
+\end{equation}
 
@@
 
+<!-- The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
+This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
+-->
+
 Motivated by the estimator in \@ref(eq:denriem), we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
@@
 where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}^{-1}(\pmb{y}_i)$.
-Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{#riemmetric}.
-
-[TODO: How to derive the volume density function?]
+Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{riemmetric}.
+
+<!-- [TODO: How to derive the volume density function?] -->
+By the definition of the volume density function in Equation \@ref(eq:volden), we could calculate how the volume change under the change of coordinates.
+When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.
+geodesic normal coordinates.
+In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_p(q)) |$.
+In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
+Since the logarithm map of the manifold is also unknown, the volume density function in @Pelletier2005-vu is no longer valid.
+Instead, we need a volume density function independent of the chart.
+
+For $p \in M$, denote an open ball centered at $p$ with radius $r$ as $B(p,r)$ and the open neighborhood of $0_p \in T_pM$ can be denoted as $B(0_p,r)=\exp_p^{-1}(B(p,r))$.
+Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_p,r)$, with $g^{\prime} = \exp^{\ast}_p(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_p$ in $B(0_p,r)$.
+Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\overline{U}, \overline{\psi})$ of $B(0_p,r)$, we have $u \in \overline{U}$. 
+The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\overline{\psi}_1|_u), \dots, (\partial / \partial\overline{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$.
+The quotient of these two volumes is then independent of the chart $(\overline{U}, \overline{\psi})$ [@Henry2009-om].
+Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\overline{U}, \overline{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
+$$
+\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} } \\ .
+$$
+The volume density function is only defined for points satisfying $d_g(p,q) < inj_gM$ so that there exists an open ball $B(p,r) \in T_pM$.
+And $\theta_p(q)=1$ for all $p,q \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.
+
+Now we consider the case with the manifold learning embedding.
+Consider the chart $(U, \psi)$ induced by an orthonormal basis of the tangent space of the embedding $T_{\pmb{y}_i}E$ and a normal neighborhood of $\pmb{y}_i$.
+Given a query point $\pmb{p} \in B(\pmb{y}_i,r)$, the volume of the parallelepiped is spanned by the same basis.
+Then the volume density function is given by 
+$$
+\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
+$$ 
+where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
+Again, the volume density function is not defined for all $\pmb{p}$ but for points that satisfies $\pmb{p} \in B(\pmb{y}_i,r)$.
 
@@
 
-<!-- The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
-This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
--->
+As for the geodesic distance $d_g$,
 
@@
 
-Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$.
+Denote $B(p, r) \subset M$ as an open ball centered at point $p$ with raidus $r$.
 Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$.

### 2022-07-18 paper/paper.tex
Derive volume density function as ratio

@@
 Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
-For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denotes a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
+For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
 Then \(\frac{Dw_1}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
 The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined as
-\[
+\begin{equation}
+\label{eq:volden}
 \theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-\]
+\end{equation}
 
@@
 
-However, in application, we usually find the observations lying on a \(d\)-dimensional manifold embedded in a much higher \(s\)-dimensional space, where \(d \ll s\).
-Instead of sampling from the underlying manifold \(M\), we could only sample from the high-dimensional data and apply dimension reduction techniques such as manifold learning to find a \(d\)-dimensional representation of \(M\).
-Notice that this representation is not necessarily isometry to \(M\), and the underlying manifold is usually unknown in real data sets.
+In application, we usually find the observations lying on a \(d\)-dimensional manifold embedded in a much higher \(s\)-dimensional space, where \(d \ll s\).
+Instead of sampling from the underlying manifold \(M\), we could sample from the high-dimensional data only and apply dimension reduction techniques such as manifold learning to find a \(d\)-dimensional representation of \(M\).
+Notice that this representation is not necessarily isometry to \(M\), and the underlying manifold tends to be unknown in real data sets.
 We could use the low-dimensional embedding as an approximation of the true manifold \(M\).
 However, the pointwise distortion induced by the manifold learning process needs to be taken care of to accurately estimate the density of the manifold.
-Therefore, the density estimator in Equation \eqref{eq:denriem} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
-Further, the bandwidth matrix should be adaptive to each sample point to take into account the distortion.
-Motivated by the estimator in \eqref{eq:denriem}, we restate the proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
+Therefore, the kernel density estimator in Equation \eqref{eq:denriem} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
+Further, the bandwidth matrix should be adaptive to each sample point to take the distortion into account.
+
+Motivated by the estimator in \eqref{eq:denriem}, we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
 \begin{equation}
@@
 \end{equation}
-where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the estimated geodesic distance with the distortion corrected by the Riemannian metric at embedding point \(\pmb{y}_i \in \mathbb{R}^d\).
-Then the only problem left is the estimation of the Riemannian metric at each embedding point.
+where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}^{-1}(\pmb{y}_i)\).
+Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{riemmetric}.
+
+By the definition of the volume density function in Equation \eqref{eq:volden}, we could calculate how the volume change under the change of coordinates.
+When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates.
+In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
+In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
+Since the logarithm map of the manifold is also unknown, the volume density function in \textcite{Pelletier2005-vu} is no longer valid.
+Instead, we need a volume density function independent of the chart.
+
+For \(p \in M\), denote an open ball centered at \(p\) with radius \(r\) as \(B(p,r)\) and the open neighborhood of \(0_p \in T_pM\) can be denoted as \(B(0_p,r)=\exp_p^{-1}(B(p,r))\).
+Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_p,r)\), with \(g^{\prime} = \exp^{\ast}_p(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_p\) in \(B(0_p,r)\).
+Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\overline{U}, \overline{\psi})\) of \(B(0_p,r)\), we have \(u \in \overline{U}\).
+The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\overline{\psi}_1|_u), \dots, (\partial / \partial\overline{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\).
+The quotient of these two volumes is then independent of the chart \((\overline{U}, \overline{\psi})\) \autocite{Henry2009-om}.
+Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\overline{U}, \overline{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
+\[
+\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} } \\ .
+\]
+The volume density function is only defined for points satisfying \(d_g(p,q) < inj_gM\) so that there exists an open ball \(B(p,r) \in T_pM\).
+And \(\theta_p(q)=1\) for all \(p,q \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.
+
+Now we consider the case with the manifold learning embedding.
+Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\).
+Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
+Then the volume density function is given by
+\[
+\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
+\]
+where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
+Again, the volume density function is not defined for all \(\pmb{p}\) but for points that satisfies \(\pmb{p} \in B(\pmb{y}_i,r)\).
+
+{[}TODO: Further fix the approximation induced by the estimated d\_g so that f will integral to 1.{]}
+
+As for the geodesic distance \(d_g\),
+
+\hypertarget{riemmetric}{%
+\subsection{Riemannian metric estimation}\label{riemmetric}}
 
@@
 In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
-
-\hypertarget{riemannian-metric-estimation}{%
-\subsection{Riemannian metric estimation}\label{riemannian-metric-estimation}}
 
@@
 
-Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with raidus \(r\).
+Denote \(B(p, r) \subset M\) as an open ball centered at point \(p\) with raidus \(r\).
 Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\), where \(exp_p\) is the \emph{exponential map} at point \(p\).

### 2022-07-18 paper/paper.Rmd
DC-KDE done

@@
 \begin{align}
+\label{eq:kernelcondition}
 & (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
@@
 For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denote a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
-Then $\frac{Dw_1}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$.
-The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined as 
+Then $\frac{Dw_i}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$.
+The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined in @Le_Brigant2019-lj as 
 \begin{equation}
@@
 $$
-in normal coordinates.
+in normal coordinates [@Le_Brigant2019-lj].
 The volume density function is motivated by how a measure $du$ changes under the change of coordinates.
-It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
+The volume form is variant to the exponential map except that the manifold is flat.
+Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1.
+It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
 @Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.
@@
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation}
@@
 where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all $\pmb{p}$ but for points that satisfies $\pmb{p} \in B(\pmb{y}_i,r)$.
-
-[TODO: Further fix the approximation induced by the estimated d_g so that f will integral to 1.]
-
-As for the geodesic distance $d_g$,
+Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in (v) of \@ref(eq:kernelcondition).
+
+<!-- [TODO: Further fix the approximation induced by the estimated d_g so that f will integral to 1.] -->
+As for the geodesic distance $d_g$ in \@ref(eq:denriem), we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
+\begin{equation}
+\label{eq:dgestimate}
+\hat{d}_g(\pmb{p}, \pmb{y}_i) = \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)},
+\end{equation}
+where $\pmb{H}(\pmb{y}_i)$ is the estimated Riemannian metric at embedding point $\pmb{y}_i$.
+
+For two points $p,q \in M$, the geodesic distance between $p$ and $q$ is defined as the length of the shortest curve from $p$ to $q$ along the manifold.
+The geodesic distance is intrinsic, which means it is a coordinate-independent quantity.
+The Riemannian metric functions as the correction term for different coordinates of manifold learning embedding.
+By using the first-order distance estimator, $\hat{d}_g(\pmb{p}, \pmb{y}_i)$ serves as a good approximation of the true geodesic distance $d_g$ with the distortion corrected at each embedding point $\pmb{y}_i$.
+Compared with the graph distance, the shortest path between point in the neighborhood graph $G$, our proposed $\hat{d}_g$ is better because we use the same graph $G$ for all manifold learning embeddings but the distances between different nodes in $G$ might not be the same due to the distortion.
+
+As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|p-q\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality.
+By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric.
+In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1.
+The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface* in higher dimension.
+We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
+$$
+C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
+$$
+where the distortion is corrected for the tangent plane of the embedding points.
 

### 2022-07-18 paper/paper.tex
DC-KDE done

@@
 For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
-Then \(\frac{Dw_1}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
-The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined as
+Then \(\frac{Dw_i}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
+The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined in \textcite{Le_Brigant2019-lj} as
 \begin{equation}
@@
 \]
-in normal coordinates.
+in normal coordinates \autocite{Le_Brigant2019-lj}.
 The volume density function is motivated by how a measure \(du\) changes under the change of coordinates.
-It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
+The volume form is variant to the exponential map except that the manifold is flat.
+Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1.
+It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
 \textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
@@
 where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all \(\pmb{p}\) but for points that satisfies \(\pmb{p} \in B(\pmb{y}_i,r)\).
-
-{[}TODO: Further fix the approximation induced by the estimated d\_g so that f will integral to 1.{]}
-
-As for the geodesic distance \(d_g\),
+Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in (v) of \eqref{eq:kernelcondition}.
+
+As for the geodesic distance \(d_g\) in \eqref{eq:denriem}, we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
+\begin{equation}
+\label{eq:dgestimate}
+\hat{d}_g(\pmb{p}, \pmb{y}_i) = \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)},
+\end{equation}
+where \(\pmb{H}(\pmb{y}_i)\) is the estimated Riemannian metric at embedding point \(\pmb{y}_i\).
+
+For two points \(p,q \in M\), the geodesic distance between \(p\) and \(q\) is defined as the length of the shortest curve from \(p\) to \(q\) along the manifold.
+The geodesic distance is intrinsic, which means it is a coordinate-independent quantity.
+The Riemannian metric functions as the correction term for different coordinates of manifold learning embedding.
+By using the first-order distance estimator, \(\hat{d}_g(\pmb{p}, \pmb{y}_i)\) serves as a good approximation of the true geodesic distance \(d_g\) with the distortion corrected at each embedding point \(\pmb{y}_i\).
+Compared with the graph distance, the shortest path between point in the neighborhood graph \(G\), our proposed \(\hat{d}_g\) is better because we use the same graph \(G\) for all manifold learning embeddings but the distances between different nodes in \(G\) might not be the same due to the distortion.
+
+As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|p-q\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality.
+By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric.
+In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1.
+The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface} in higher dimension.
+We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
+\[
+C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
+\]
+where the distortion is corrected for the tangent plane of the embedding points.
 

### 2022-07-18 paper/paper.Rmd
Sec 2.1-2.3 rewrite

@@
 \label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
 \end{equation} 
@@
 Therefore, the Riemannian manifold density estimator in @Pelletier2005-vu only works when the true manifold $M$ is known.
-
-<!-- $$ -->
-<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
-<!-- $$ --> 
-<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
-<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->
+Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.
 
@@
 
-<!-- The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point.
-This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.
--->
-
-Motivated by the estimator in \@ref(eq:denriem), we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
+Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
 \end{equation}
-where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}^{-1}(\pmb{y}_i)$.
-Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{riemmetric}.
-
-<!-- [TODO: How to derive the volume density function?] -->
+where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C$ is a correction term for the volume between the spaces so that the integral is 1.
+We will now state below how to calculate each term in the estimator.
+Then the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
+
 By the definition of the volume density function in Equation \@ref(eq:volden), we could calculate how the volume change under the change of coordinates.
@@
 geodesic normal coordinates.
-In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_p(q)) |$.
+In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_p(q)) |$.
 In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
@@
 Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_p,r)$, with $g^{\prime} = \exp^{\ast}_p(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_p$ in $B(0_p,r)$.
-Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\overline{U}, \overline{\psi})$ of $B(0_p,r)$, we have $u \in \overline{U}$. 
-The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\overline{\psi}_1|_u), \dots, (\partial / \partial\overline{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$.
-The quotient of these two volumes is then independent of the chart $(\overline{U}, \overline{\psi})$ [@Henry2009-om].
-Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\overline{U}, \overline{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
-$$
-\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} } \\ .
+Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$, we have $u \in \bar{U}$. 
+The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$.
+The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om].
+Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
+$$
+\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
 $$
@@
 where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in (v) of \@ref(eq:kernelcondition).
-
-<!-- [TODO: Further fix the approximation induced by the estimated d_g so that f will integral to 1.] -->
+Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in condition (v) of \@ref(eq:kernelcondition).
+
 As for the geodesic distance $d_g$ in \@ref(eq:denriem), we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
@@
 In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1.
-The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface* in higher dimension.
+The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension.
 We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
@@
 where the distortion is corrected for the tangent plane of the embedding points.
+
+Now we have all the components needed to estimate the density of manifold learning embedding in Equation \@ref(eq:denestimator) except for the Riemannian metric.
+The Riemannian metric is a key to correct the distortion from the manifold learning algorithms and we discuss its estimation in \autoref{riemmetric}.
 

### 2022-07-18 paper/paper.tex
Sec 2.1-2.3 rewrite

@@
 \label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime H^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
 \end{equation}
@@
 Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
+Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.
 
@@
 
-Motivated by the estimator in \eqref{eq:denriem}, we restate our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
+Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|} K\bigg( \frac{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
 \end{equation}
-where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\pmb{H}(\pmb{p})|}{|\pmb{H}(\pmb{y}_i)|}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}^{-1}(\pmb{y}_i)\).
-Then the only problem left is the estimation of the Riemannian metric at each embedding point, discussed in \autoref{riemmetric}.
+where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C\) is a correction term for the volume between the spaces so that the integral is 1.
+We will now state below how to calculate each term in the estimator.
+Then the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
 
@@
 When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates.
-In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
+In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
 In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
@@
 Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_p,r)\), with \(g^{\prime} = \exp^{\ast}_p(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_p\) in \(B(0_p,r)\).
-Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\overline{U}, \overline{\psi})\) of \(B(0_p,r)\), we have \(u \in \overline{U}\).
-The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\overline{\psi}_1|_u), \dots, (\partial / \partial\overline{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\).
-The quotient of these two volumes is then independent of the chart \((\overline{U}, \overline{\psi})\) \autocite{Henry2009-om}.
-Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\overline{U}, \overline{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
-\[
-\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\overline{\psi}_i|_u), (\partial / \partial\overline{\psi}_j|_u)) |^{1/2} } \\ .
+Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\), we have \(u \in \bar{U}\).
+The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\).
+The quotient of these two volumes is then independent of the chart \((\bar{U}, \bar{\psi})\) \autocite{Henry2009-om}.
+Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
+\[
+\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
 \]
@@
 where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in (v) of \eqref{eq:kernelcondition}.
+Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in condition (v) of \eqref{eq:kernelcondition}.
 
@@
 In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1.
-The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface} in higher dimension.
+The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension.
 We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
@@
 where the distortion is corrected for the tangent plane of the embedding points.
+
+Now we have all the components needed to estimate the density of manifold learning embedding in Equation \eqref{eq:denestimator} except for the Riemannian metric.
+The Riemannian metric is a key to correct the distortion from the manifold learning algorithms and we discuss its estimation in \autoref{riemmetric}.
 

### 2022-07-19 paper/paper.Rmd
Sec 2.4 rewrite

@@
 Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
-$$
+\begin{equation}
+\label{eq:volumedensityratio}
 \theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
-$$
+\end{equation}
 The volume density function is only defined for points satisfying $d_g(p,q) < inj_gM$ so that there exists an open ball $B(p,r) \in T_pM$.
@@
 Then the volume density function is given by 
-$$
+\begin{equation}
+\label{eq:volumedensityest}
 \theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
-$$ 
+\end{equation} 
 where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
@@
 We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
-$$
+\begin{equation}
+\label{eq:vbs}
 C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
-$$
+\end{equation}
 where the distortion is corrected for the tangent plane of the embedding points.
@@
 
-An important goal of manifold learning is to recover the local or global features of the data.
-In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric $g$.
-This could also be used to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for each data point.
-Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \@ref(eq:denestimator).
-In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
+An important goal of manifold learning is to recover the local or global features of the data. However, manifold learning algorithms usually fail to generate an isometric embedding.
+In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimated Riemannian metric.
+Instead of developing a new manifold learning method, they aim to fix the geometric distortion in the embedding.
+It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix variable at each data point.
+The estimated Riemannian metric completes our proposed kernel density estimation in \autoref{dckde} and enables us to take the geometric distortion into consideration when estimating the densities of the manifold.
 
@@
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(M, g)$.
-
-To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
-The inner product between two vectors $u,v \in T_pM$, $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{\langle u,v \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{\|u\|\|v\|}$ in the tangent space.
-Therefore, for each point $p\in M$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
-
-[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$
-
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in M$ is defined as 
+For each point $p\in M$ in any embedding coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
+
+The inner product between two vectors $u,v \in T_pM$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
+[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
 $$
@@
 $$ 
-where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
-$$
-Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
-$$ 
-Both the concepts of distance and volume are relevant to kernel density estimation.
-
-@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
+where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
+$$
+Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
+$$ 
+Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).
+
+@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any manifold learning embedding with the estimated Riemannian metric.
+There are mainly four steps in the algorithm: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
 We restate these four steps in Algorithm \@ref(alg:learnmetric).
-By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed.
-Most manifold learning methods involve the construction of the nearest
-neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built. The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.
-
-As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension $s$ is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$ will be returned.
-This algorithm is also implemented in a Python library *megaman* [@McQueen2016-xz].
+By applying an existing manifold learning algorithm to the high-dimensional input data $\pmb{X}\in \mathbb{R}^s$ with $N$ observations, a low-dimensional embedding $\pmb{Y} \in \mathbb{R}^d$ can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\hat{\mathcal{L}}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+The graph Laplacian estimator together with the embedding $\pmb{Y}$ could further be used to achieve the augmented embedding with the Riemannian metric, $(\pmb{Y}, \pmb{H})$.
+This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz].
 It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
-
-<!-- ```{=tex} -->
+We have re-implemented the algorithm in *R*.
+
+As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned.
+To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process.
+@Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl].
+By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
+An *R* library named *intRinsic* has implemented this algorithm and we are going to apply it to when $d$ is unknown.
+
 \begin{algorithm}[!htb]
@@
   \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
+  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
   \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters }
+  \OptParameter{ manifold learning parameters $MANILEARN$}
   \BlankLine
@@
 
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^s$;
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
 
@@
 
-  \STATE Embed each data point $x_i\in \mathcal{R}^s$ to embedding coordinates $\pmb{y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm;
-
-  \STATE Obtain the matrix $\pmb{\tilde{h}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{y}$ with each element vector in $\pmb{\tilde{h}}$ being
-  $$
-    \pmb{\tilde{h}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm $MANILEARN$;
+
+  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
+  $$
+    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
   $$
@@
 
-  \STATE Calculate the Riemannian metric $\pmb{h}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{    h}}$ with 
-  $$
-    \pmb{h} = U diag{1/(\Lambda[1:d])} U^\prime,
-  $$
-  where $[U, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
+  $$
+    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
+  $$
+  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
 
@@
 ## Outlier detection on manifold learning embedding
+
+Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \@ref(eq:denestimator).
 

### 2022-07-19 paper/paper.tex
Sec 2.4 rewrite

@@
 \end{equation}
-where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \((\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C\) is a correction term for the volume between the spaces so that the integral is 1.
+where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C\) is a correction term for the volume between the spaces so that the integral is 1.
 We will now state below how to calculate each term in the estimator.
@@
 Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
-\[
+\begin{equation}
+\label{eq:volumedensityratio}
 \theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
-\]
+\end{equation}
 The volume density function is only defined for points satisfying \(d_g(p,q) < inj_gM\) so that there exists an open ball \(B(p,r) \in T_pM\).
@@
 Then the volume density function is given by
-\[
+\begin{equation}
+\label{eq:volumedensityest}
 \theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
-\]
+\end{equation}
 where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
@@
 We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
-\[
+\begin{equation}
+\label{eq:vbs}
 C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
-\]
+\end{equation}
 where the distortion is corrected for the tangent plane of the embedding points.
@@
 
-An important goal of manifold learning is to recover the local or global features of the data.
-In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric \(g\).
-This could also be used to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix for each data point.
-Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \eqref{eq:denestimator}.
-In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.
+An important goal of manifold learning is to recover the local or global features of the data. However, manifold learning algorithms usually fail to generate an isometric embedding.
+In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimated Riemannian metric.
+Instead of developing a new manifold learning method, they aim to fix the geometric distortion in the embedding.
+It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix variable at each data point.
+The estimated Riemannian metric completes our proposed kernel density estimation in \autoref{dckde} and enables us to take the geometric distortion into consideration when estimating the densities of the manifold.
 
@@
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((M, g)\).
-
-To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding.
-The inner product between two vectors \(u,v \in T_pM\), \(\langle u,v \rangle_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\)}, can be used to define some geometric quantities, such as the vector norm \(\|u\|=\sqrt{\langle u,v \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{\|u\|\|v\|}\) in the tangent space.
-Therefore, for each point \(p\in M\) in any coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
-
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\in M\) is defined as
+For each point \(p\in M\) in any embedding coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
+
+The inner product between two vectors \(u,v \in T_pM\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
+{[}\^{}1{]}: Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
 \[
@@
 \]
-where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function mapping \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
-\[
-Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
-\]
-Both the concepts of distance and volume are relevant to kernel density estimation.
-
-\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
+where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function that maps \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
+\[
+Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
+\]
+Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \eqref{eq:volumedensityratio}, \eqref{eq:volumedensityest}, \eqref{eq:dgestimate}, and \eqref{eq:vbs}.
+
+\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any manifold learning embedding with the estimated Riemannian metric.
+There are mainly four steps in the algorithm: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
 We restate these four steps in Algorithm \ref{alg:learnmetric}.
-By applying an existing manifold learning algorithm to the data \(X\in \mathbb{R}^r\) with \(n\) observations, a low-dimensional embedding \(f_n \in \mathbb{R}^d\) can be computed.
-Most manifold learning methods involve the construction of the nearest
-neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built. The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_M\) with a discrete consistent estimator, the geometric graph Laplacian \(\mathcal{L}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric \((f_n, g_n)\).
-
-As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
-This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
+By applying an existing manifold learning algorithm to the high-dimensional input data \(\pmb{X}\in \mathbb{R}^s\) with \(N\) observations, a low-dimensional embedding \(\pmb{Y} \in \mathbb{R}^d\) can be computed.
+Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built.
+The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
+\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_M\) with a discrete consistent estimator, the geometric graph Laplacian \(\hat{\mathcal{L}}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+The graph Laplacian estimator together with the embedding \(\pmb{Y}\) could further be used to achieve the augmented embedding with the Riemannian metric, \((\pmb{Y}, \pmb{H})\).
+This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
 It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
+We have re-implemented the algorithm in \emph{R}.
+
+As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(\pmb{H}(p)\) is equal to \(d\); otherwise, the Riemannian metric will be returned.
+To ensure that the estimated Riemannian metric is returned as a \(d\times d\) matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} to derive id estimator from a Poisson point process.
+\textcite{Facco2017-rl} point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter \(1\) and shape parameter \(d\), the intrinsic dimension of the manifold \autocite{Facco2017-rl}.
+By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
+An \emph{R} library named \emph{intRinsic} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.
 
@@
 
-  \STATE Embed each data point $x_i\in \mathcal{R}^s$ to embedding coordinates $\pmb{y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm;
-
-  \STATE Obtain the matrix $\pmb{\tilde{h}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{y}$ with each element vector in $\pmb{\tilde{h}}$ being
+  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm $MANILEARN$;
+
+  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
   $$
-    \pmb{\tilde{h}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
   $$
@@
 
-  \STATE Calculate the Riemannian metric $\pmb{h}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{    h}}$ with 
+  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
   $$
-    \pmb{h} = U diag{1/(\Lambda[1:d])} U^\prime,
+    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
   $$
-  where $[U, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
 

### 2022-07-19 paper/paper.Rmd
Comment out Appendix A S1 example

@@
 
-## Example of $S^1$ for manifold density estimation [TODO: rewrite as appendix B with plots?]
-
-Manifold: $S^1$ with center 0 and radius $r=1$
-
-Polar coordinates: $(r, \theta)$
-
-Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
-
-Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$
-
-Density of $\theta$: $f(\theta) = 2 / \pi$
-
-Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$
-
-The CDF of X is given by 
-$$
-P(X \leq x) = P(\cos(\theta) \leq x) \\
-= P(\theta \geq \arccos(x)) \\
-= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-= 1 - \frac{2}{\pi} \arccos(x).
-$$ 
-Then the density of X is the derivative of the CDF w.r.t. $x$ $$
-f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-= \frac{2}{\pi \sqrt{1 - x^2}}.
-$$
-
-Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
-estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
-and compare it with its true density $f(x)$.
-We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
-<!-- \begin{equation} --> 
-<!-- \label{eq:pelletier} --> 
-$$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
-$$ 
-<!-- \end{equation} --> 
-where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
-
-<!-- we could reduce the dimension to $d = 1$ and estimate the density of the 1-dimensional embedding$. -->
-
-<!-- First, we use the kernel density estimator in @ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->
-
-Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
-Therefore, the geodesic distance is
-$$
-d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-$$ 
-As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
+<!-- ## Example of $S^1$ for manifold density estimation [TODO: rewrite as appendix B with plots?] -->
+
+<!-- Manifold: $S^1$ with center 0 and radius $r=1$ -->
+
+<!-- Polar coordinates: $(r, \theta)$ -->
+
+<!-- Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$. -->
+
+<!-- Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$ -->
+
+<!-- Density of $\theta$: $f(\theta) = 2 / \pi$ -->
+
+<!-- Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$ -->
+
+<!-- The CDF of X is given by  -->
+<!-- $$ -->
+<!-- P(X \leq x) = P(\cos(\theta) \leq x) \\ -->
+<!-- = P(\theta \geq \arccos(x)) \\ -->
+<!-- = \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\ -->
+<!-- =  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\ -->
+<!-- = 1 - \frac{2}{\pi} \arccos(x). -->
+<!-- $$  -->
+<!-- Then the density of X is the derivative of the CDF w.r.t. $x$ $$ -->
+<!-- f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\ -->
+<!-- = \frac{2}{\pi \sqrt{1 - x^2}}. -->
+<!-- $$ -->
+
+<!-- Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could -->
+<!-- estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ -->
+<!-- and compare it with its true density $f(x)$. -->
+<!-- We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by -->
+<!-- $$ -->
+<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}), -->
+<!-- $$  -->
+<!-- where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
+-->
+
+<!-- Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
+Therefore, the geodesic distance is  -->
+<!-- $$ -->
+<!-- d_g(p, X_i) = \arccos(p) - \arccos(X_i) -->
+<!-- $$  -->
+<!-- As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
 $|g(\exp_p^{-1}(x_i))|$.
-
-[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
-
-The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$, 
-$$
-g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-$$ 
-is smooth in a local neighborhood $U$ of $M$.
-
-Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by 
-$$
-\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
-$$ 
-Therefore, the estimator for $p \in S^1$ is given by 
-$$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-$$ 
-where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*.
-
-### How to derive volume density function $\theta_{X_i}(p)$
-
-#### Example of $S^2$ in @Henry2009-om.
-
-Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
+-->
+
+<!-- [When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.] -->
+
+<!-- The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,  -->
+<!-- $$ -->
+<!-- g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R}, -->
+<!-- $$  -->
+<!-- is smooth in a local neighborhood $U$ of $M$. -->
+
+<!-- Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by  -->
+<!-- $$ -->
+<!-- \theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}. -->
+<!-- $$  -->
+<!-- Therefore, the estimator for $p \in S^1$ is given by  -->
+<!-- $$ -->
+<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg), -->
+<!-- $$  -->
+<!-- where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*. -->
+
+<!-- ### How to derive volume density function $\theta_{X_i}(p)$ -->
+
+<!-- #### Example of $S^2$ in @Henry2009-om. -->
+
+<!-- Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
 Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$.
-Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
-$$
-\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
-$$ 
-if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. 
-Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
-Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
-$$
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
-$$
-$$
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
-$$ 
-The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
-Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
-Therefore, we obtain that the volume density on the sphere is 
-$$
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
-= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
-$$ 
-for $q \neq p, -p$, and $\theta_p(p) = 1$.
-
-#### $S^1$
-
-Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization 
-$$
-\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
-$$
-and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have
-$$
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
-$$
-So the riemannian metric is $g_q = \cos(r)$ and further, 
-$$
-\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
-$$
-where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$.
-
-\clearpage
+Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by  -->
+<!-- $$ -->
+<!-- \Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw), -->
+<!-- $$  -->
+<!-- if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.  -->
+<!-- Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$. -->
+<!-- Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that -->
+<!-- $$ -->
+<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w. -->
+<!-- $$  -->
+<!-- The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors. -->
+<!-- Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
+-->
+<!-- Therefore, we obtain that the volume density on the sphere is  -->
+<!-- $$ -->
+<!-- \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\ -->
+<!-- = \frac{|\sin(d_g(p,q))|}{d_g(p,g)}, -->
+<!-- $$  -->
+<!-- for $q \neq p, -p$, and $\theta_p(p) = 1$. -->
+
+<!-- #### $S^1$ -->
+
+<!-- Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization  -->
+<!-- $$ -->
+<!-- \Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv, -->
+<!-- $$ -->
+<!-- and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have -->
+<!-- $$ -->
+<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v. -->
+<!-- $$ -->
+<!-- So the riemannian metric is $g_q = \cos(r)$ and further,  -->
+<!-- $$ -->
+<!-- \theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|}, -->
+<!-- $$ -->
+<!-- where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$. -->
+
+<!-- \clearpage -->
 

### 2022-07-19 paper/paper.tex
Comment out Appendix A S1 example

@@
 
+Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}.
+
 Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \ref{alg:vkderm}.
@@
 
-\hypertarget{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}{%
-\subsection{\texorpdfstring{Example of \(S^1\) for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}{Example of S\^{}1 for manifold density estimation {[}TODO: rewrite as appendix B with plots?{]}}}\label{example-of-s1-for-manifold-density-estimation-todo-rewrite-as-appendix-b-with-plots}}
-
-Manifold: \(S^1\) with center 0 and radius \(r=1\)
-
-Polar coordinates: \((r, \theta)\)
-
-Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).
-
-Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)
-
-Density of \(\theta\): \(f(\theta) = 2 / \pi\)
-
-Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)
-
-The CDF of X is given by
-\[
-P(X \leq x) = P(\cos(\theta) \leq x) \\
-= P(\theta \geq \arccos(x)) \\
-= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-= 1 - \frac{2}{\pi} \arccos(x).
-\]
-Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
-f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-= \frac{2}{\pi \sqrt{1 - x^2}}.
-\]
-
-Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
-estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
-and compare it with its true density \(f(x)\).
-We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
-\]
-where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
-
-Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\).
-Therefore, the geodesic distance is
-\[
-d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-\]
-As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).
-
-{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}
-
-The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
-\[
-g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-\]
-is smooth in a local neighborhood \(U\) of \(M\).
-
-Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
-\[
-\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
-\]
-Therefore, the estimator for \(p \in S^1\) is given by
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-\]
-where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.
-
-\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
-\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}
-
-\hypertarget{example-of-s2-in-henry2009-om.}{%
-\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}
-
-Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\).
-Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\).
-Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
-\[
-\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
-\]
-if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
-Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
-Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
-\[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
-\]
-\[
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
-\]
-The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
-Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
-Therefore, we obtain that the volume density on the sphere is
-\[
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
-= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
-\]
-for \(q \neq p, -p\), and \(\theta_p(p) = 1\).
-
-\hypertarget{s1}{%
-\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}
-
-Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
-\[
-\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
-\]
-and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
-\[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
-\]
-So the riemannian metric is \(g_q = \cos(r)\) and further,
-\[
-\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
-\]
-where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).
-
-\clearpage
-
 \hypertarget{swissrollappe}{%

### 2022-07-19 paper/paper.Rmd
Update schematic plot

@@
   echo = TRUE,
-  cache = FALSE,
+  cache = T,
   messages = FALSE,
@@
 Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \@ref(eq:denestimator).
-
-Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \@ref(alg:vkderm).
+The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection.
+As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \@ref(alg:kdeanomaly).
 There are mainly four steps involved in the algorithm.
@@
 \begin{algorithm}[!htb]
-  \caption{Variable kernel density estimates with Riemannian metric}
-  \label{alg:vkderm}
+  \caption{Anomaly detection with distortion-corrected kernel density estimates}
+  \label{alg:kdeanomaly}
   \DontPrintSemicolon
@@
   \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
-  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
-  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
+  \Input{ high-dimensional data $\pmb{x}_i$ for all $i=1,\dots,N$ }
+  \Output{ outliers embedding coordinates $\pmb{y}_1, \dots, \pmb{y}_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
   \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
@@
   \begin{algorithmic}[1]
-
-  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
-  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \@ref(equ:denestimator);
-  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
+  \STATE Estimate the intrinsic dimension $d$ with the TWO-NN algorithm if $d$ is not given;
+  \STATE For all $i=1,\dots,N$, compute the $d$-dimensional embeddings $\pmb{y}_i$ with any existing manifold learning algorithms and the corresponding Riemannian metric $\pmb{H}(\pmb{y}_i)$ using the Learn Metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+  \STATE Compute the kernel density estimates $\hat{f}(\pmb{y}_i)$ for all $i=1,\dots,N$ using the DC\_KDE in \ref{eq:denestimator};
+  \STATE Reorder the embedding coordinates $\pmb{Y}$ from smallest to largest according to the density estimates $\hat{f}$ and subset the top \textit{n\_outliers} points as the outliers.
   \end{algorithmic}
@@
 
-If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
+If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding plot and find which observations lie in the highest density region of specified coverage, eg.
 1%, 50%, 99%, \>99%.
@@
 The proposed schematic is shown in Figure \@ref(fig:vkde).
-The highlighted two steps in Figure \@ref(fig:vkde) are the main contributions of this main chapter, replacing the bandwidth matrix $\pmb{H}_i$ with the Riemannian metric $g_i$ for each point in
-variable kernel density estimate, and computing the highest density region plots based on the density estimates, $\hat{f}(\pmb{x})$, for anomaly detection.
+The last two steps in Figure \@ref(fig:vkde) are the main contributions of Algorithm \@ref(alg:kdeanomaly), generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
+Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-ex, the changes are also highlighted in blue.
+With this anomaly detection algorithm, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
 
@@
 ```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
-knitr::include_graphics("figures/vkde.png")
+knitr::include_graphics("figures/kde_h-vkde.png")
 ```

### 2022-07-19 paper/paper.tex
Update schematic plot

@@
 Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}.
-
-Now we present our proposed algorithm for anomaly detection based on variable kernel density estimates in \ref{alg:vkderm}.
+The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection.
+As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \ref{alg:kdeanomaly}.
 There are mainly four steps involved in the algorithm.
@@
   \begin{algorithmic}[1]
-
-  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
-  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};
-  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
+  \STATE Estimate the intrinsic dimension $d$ with the TWO-NN algorithm if $d$ is not given;
+  \STATE For all $i=1,\dots,N$, compute the $d$-dimensional embeddings $\pmb{y}_i$ with any existing manifold learning algorithms and the corresponding Riemannian metric $\pmb{H}(\pmb{y}_i)$ using the Learn Metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
+  \STATE Compute the kernel density estimates $\hat{f}(\pmb{y}_i)$ for all $i=1,\dots,N$ using the DC\_KDE in \ref{eq:denestimator};
+  \STATE Reorder the embedding coordinates $\pmb{Y}$ from smallest to largest according to the density estimates $\hat{f}$ and subset the top \textit{n\_outliers} points as the outliers.
   \end{algorithmic}
@@
 
-If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg.
+If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding plot and find which observations lie in the highest density region of specified coverage, eg.
 1\%, 50\%, 99\%, \textgreater99\%.
@@
 The proposed schematic is shown in Figure \ref{fig:vkde}.
-The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{H}_i\) with the Riemannian metric \(g_i\) for each point in
-variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.
+The last two steps in Figure \ref{fig:vkde} are the main contributions of Algorithm \ref{alg:kdeanomaly}, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
+Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-ex}, the changes are also highlighted in blue.
+With this anomaly detection algorithm, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/vkde}
+{\centering \includegraphics[width=0.95\linewidth]{figures/kde_h-vkde}
 

### 2022-07-19 paper/paper.Rmd
Add reference for intRinsic

@@
 Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$.
-The Riemannian metric of the embedding is denoted as $g_i, i = 1, 2, \dots, N$.
+The Riemannian metric of the embedding is denoted as $H_i, i = 1, 2, \dots, N$.
 In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$.
@@
 @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
-In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
+In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
 <!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
@@
 By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
-An *R* library named *intRinsic* has implemented this algorithm and we are going to apply it to when $d$ is unknown.
+An *R* library named *intRinsic* [@Denti2021-qc] has implemented this algorithm and we are going to apply it to when $d$ is unknown.
 
@@
   \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters $MANILEARN$}
+  \OptParameter{ manifold learning parameters EMBED }
   \BlankLine
@@
 
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm $MANILEARN$;
+  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
 
@@
 Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator).
-The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f_N(\pmb{y}_i)$.
+The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f(\pmb{y}_i)$.
 

### 2022-07-19 R/sources/vkde.R

@@ -80,6 +80,7 @@ vkde <- function(x, h = NULL, vh = NULL, r = NULL, gridsize = 20, xmin = apply(x
   } else {
     
     # evaluate density on given data points
+    # ONLY USE THIS PART
     # z <- NULL
     z <- matrix(0, n, n)
     for (i in 1:n) {
@@ -89,7 +90,7 @@ vkde <- function(x, h = NULL, vh = NULL, r = NULL, gridsize = 20, xmin = apply(x
       hi <- hi / mean(vh[,,bindex])
       for(j in bindex) {
         hj <- vh[,,j]
-        z[i,j] <- (det(hj) / det(hi)) ^ (-0.5) * (2 * pi) ^ (-d / 2) * r ^ (-0.5) * exp( -0.5 / r * t(eval.points[i,] - eval.points[j,]) %*% solve(hi) %*% (eval.points[i,] - eval.points[j,]) ) / (pnorm(1) - pnorm(-1)) # suppK = [0,1], scale to integral to 1
+        z[i,j] <- (det(hj) / det(hi)) ^ (-0.5) * (2 * pi) ^ (-d / 2) * r ^ (-1/d) * exp( -0.5 / (r^2) * t(eval.points[i,] - eval.points[j,]) %*% solve(hi) %*% (eval.points[i,] - eval.points[j,]) ) / (pnorm(1) - pnorm(-1)) # suppK = [0,1], scale to integral to 1
       }
     }
     z <- colMeans(z, na.rm = TRUE)

### 2022-07-23 paper/paper.Rmd
Correct equations

@@
 @Pelletier2005-vu generalize the multivariate kernel density esitmator to Riemnnian manifolds.
-However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space.
-Therefore, we propose a distortion-corrected kernel density estimator as in Equation \@ref(eq:denestimator0) based on the low-dimensional embedding, $\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, of the high-dimensional data with $N$ data points,
+However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space.
+Therefore, we propose a distortion-corrected kernel density estimator as in Equation \@ref(eq:denestimator0) based on the low-dimensional manifold learning embedding, $\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, of the high-dimensional data,
 \begin{equation}
 \label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation}
 where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise adaptive Riemannian matrix of dimension $d \times d$.
-The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions.
+The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by volume density function at each point.
 These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
@@
 
-To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's estimator for Riemannian manifolds.
-Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm in @Perrault-Joncas2013-pq.
-Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on a Riemannian manifold densities.
+To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds.
+Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from @Perrault-Joncas2013-pq.
+Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
 
@@
 
-<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$.
+<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathbb{R}^s$.
 The underlying manifold $M$ has the true density $f$.
-Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$.
+Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathbb{R}^d$ and $d \ll s$.
 The Riemannian metric of the embedding is denoted as $H_i, i = 1, 2, \dots, N$.
@@
 @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
-In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
-<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
-drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
+In general, given multivariate random samples $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
+<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathbb{R}^d$, $i = 1, 2, \dots, N$  -->
+drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d$ is given by
 \begin{equation}
@@
 Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
-Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the geodesic distance induced by the Riemannian metric $g$ and the strictly positive injectivity radius of the manifold $inj_g(M)$ [@Chavel2006-mp].
-Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathcal{R}^d$, as i.i.d.
+Assume $(M, d_g)$ as a complete metric space, where $d_g$ is the geodesic distance induced by the Riemannian metric $g$ and the strictly positive injectivity radius of the manifold $inj_g(M)$ [@Chavel2006-mp].
+Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathbb{R}^d$, as i.i.d.
 random objects on $M$ with density $f$.
-To estimate the density of the manifold, @Pelletier2005-vu proposed a kernel density estimator for $f$ at point $\pmb{p}\in M$ as
+To estimate the density of the manifold, @Pelletier2005-vu propose a kernel density estimator for $f$ at point $\pmb{p}\in M$ as
 \begin{equation}
@@
 \end{equation}
-where $r$ is the global bandwidth satisfying $r \leq r_0$ for fixed $r_0$ such that $0\leq r_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{\pmb{X}_i}(\pmb{p})$ is the volume density function on the manifold for point $\pmb{X}_i$.
+where $r$ is the global bandwidth satisfying $r \leq r_0$ for fixed $r_0$ such that $0\leq r_0\leq inj_g(M)$, $K: \mathbb{R}_{+} \rightarrow \mathbb{R}$ is a non-negative map, and $\theta_{\pmb{X}_i}(\pmb{p})$ is the volume density function on the manifold for point $\pmb{X}_i$.
 The estimator in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator [@Henry2009-ll].
@@
 It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
-@Pelletier2005-vu mentions that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.
+
+@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.
 Note that this is true only when using the geodesic normal coordinates.
@@
 
-Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
+Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold learning embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
-\end{equation}
-where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C$ is a correction term for the volume between the spaces so that the integral is 1.
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\end{equation}
+where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the spaces so that the integral is 1.
 We will now state below how to calculate each term in the estimator.
-Then the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
+And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
 
@@
 \label{eq:volumedensityest}
-\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
+\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
 \end{equation} 
 where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in condition (v) of \@ref(eq:kernelcondition).
+Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in Condition (v) of \@ref(eq:kernelcondition).
 
@@
 \label{eq:dgestimate}
-\hat{d}_g(\pmb{p}, \pmb{y}_i) = \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)},
+\hat{d}_g(\pmb{p}, \pmb{y}_i) = \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i),
 \end{equation}
@@
 
-For two points $p,q \in M$, the geodesic distance between $p$ and $q$ is defined as the length of the shortest curve from $p$ to $q$ along the manifold.
+For two points $\pmb{p},\pmb{q} \in M$, the geodesic distance between $\pmb{p}$ and $\pmb{q}$ is defined as the length of the shortest curve from $\pmb{p}$ to $\pmb{q}$ along the manifold.
 The geodesic distance is intrinsic, which means it is a coordinate-independent quantity.
@@
 
-As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|p-q\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality.
+As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|\pmb{p}-\pmb{q}\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality.
 By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric.
@@
 
-The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_pM$ for every point $p \in M$.
+The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_\pmb{p} M$ for every point $\pmb{p}\in M$.
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-For each point $p\in M$ in any embedding coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
-
-The inner product between two vectors $u,v \in T_pM$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
+For each point $\pmb{p} \in M$ in any embedding coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
+
+The inner product between two vectors $u,v \in T_\pmb{p}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
 [^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.
+
 The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
@@
 ```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
-knitr::include_graphics("figures/kde_h-vkde.png")
+knitr::include_graphics("figures/DC-KDE.png")
 ```

### 2022-07-23 paper/paper.Rmd
Sec 2 done

@@
 
-The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_\pmb{p} M$ for every point $\pmb{p}\in M$.
+The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
@@
 
-The inner product between two vectors $u,v \in T_\pmb{p}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
+The inner product between two vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
 [^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.

### 2022-07-23 paper/paper.tex
Sec 2 done

@@
 \textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds.
-However, it is not applicable when the underlying manifold is not known or when the true manifold is mapped in a much higher dimensional space.
-Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data with \(N\) data points,
+However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space.
+Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional manifold learning embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data,
 \begin{equation}
 \label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation}
 where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
-The proposed estimator could be used as an estimate of the density of the true manifold, with the local distortion at each point corrected by volume density functions.
+The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by volume density function at each point.
 These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
 
-To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's estimator for Riemannian manifolds.
-Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm in \textcite{Perrault-Joncas2013-pq}.
-Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on a Riemannian manifold densities.
+To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds.
+Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from \textcite{Perrault-Joncas2013-pq}.
+Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
 
@@
 \textcite{Scott2015-vl} covers a wide range of techniques for estimating the kernel densities of multivariate data.
-In general, given a multivariate random sample \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d\)
-drawn from a density \(f\), the variable kernel density estimate \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut} of \(f\) at point \(\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d\) is given by
+In general, given multivariate random samples \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d\)
+drawn from a density \(f\), the variable kernel density estimate \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut} of \(f\) at point \(\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d\) is given by
 \begin{equation}
@@
 Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
-Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}.
-Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d.
+Assume \((M, d_g)\) as a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}.
+Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathbb{R}^d\), as i.i.d.
 random objects on \(M\) with density \(f\).
-To estimate the density of the manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
+To estimate the density of the manifold, \textcite{Pelletier2005-vu} propose a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
 \begin{equation}
@@
 \end{equation}
-where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\).
+where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathbb{R}_{+} \rightarrow \mathbb{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\).
 The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.
@@
 It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
-\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
+
+\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
 Note that this is true only when using the geodesic normal coordinates.
@@
 
-Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
+Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold learning embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}} K\bigg( \frac{ \sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)} }{r} \bigg),
-\end{equation}
-where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\sqrt{(\pmb{p} - \pmb{y}_i)^\prime \pmb{H}^{-1}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C\) is a correction term for the volume between the spaces so that the integral is 1.
+\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\end{equation}
+where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the spaces so that the integral is 1.
 We will now state below how to calculate each term in the estimator.
-Then the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
+And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
 
@@
 \label{eq:volumedensityest}
-\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
+\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
 \end{equation}
 where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in condition (v) of \eqref{eq:kernelcondition}.
+Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in Condition (v) of \eqref{eq:kernelcondition}.
 
@@
 
-For two points \(p,q \in M\), the geodesic distance between \(p\) and \(q\) is defined as the length of the shortest curve from \(p\) to \(q\) along the manifold.
+For two points \(\pmb{p},\pmb{q} \in M\), the geodesic distance between \(\pmb{p}\) and \(\pmb{q}\) is defined as the length of the shortest curve from \(\pmb{p}\) to \(\pmb{q}\) along the manifold.
 The geodesic distance is intrinsic, which means it is a coordinate-independent quantity.
@@
 
-As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|p-q\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality.
+As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|\pmb{p}-\pmb{q}\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality.
 By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric.
@@
 
-The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_pM\) for every point \(p \in M\).
+The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
 If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-For each point \(p\in M\) in any embedding coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
-
-The inner product between two vectors \(u,v \in T_pM\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
+For each point \(\pmb{p} \in M\) in any embedding coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
+
+The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
 {[}\^{}1{]}: Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).
+
 The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
@@
 By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
-An \emph{R} library named \emph{intRinsic} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.
+An \emph{R} library named \emph{intRinsic} \autocite{Denti2021-qc} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.
 
@@
 Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}.
-The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f_N(\pmb{y}_i)\).
+The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f(\pmb{y}_i)\).
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/kde_h-vkde}
+{\centering \includegraphics[width=0.95\linewidth]{figures/DC-KDE}
 

### 2022-07-24 paper/paper.Rmd
Fix footnote and alg name

@@
 The inner product between two vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
+The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
+$$
+l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
+$$ 
+where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
+$$
+Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
+$$ 
+Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).
+
 [^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.
-
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
-$$
-l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-$$ 
-where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
-$$
-Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
-$$ 
-Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).
 
@@
 This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies.
+We could now apply Algorithm \@ref(alg:kdeanomaly) to get the density estimates of all data points and further detect anomalies.
 As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
@@
 In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+Algorithm \@ref(alg:kdeanomaly) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly

### 2022-07-24 paper/paper.tex
Fix footnote and alg name

@@
 
-The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
-{[}\^{}1{]}: Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).
-
+The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
 The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
@@
 This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies.
+We could now apply Algorithm \ref{alg:kdeanomaly} to get the density estimates of all data points and further detect anomalies.
 As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
@@
 In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+Algorithm \ref{alg:kdeanomaly} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly

### 2022-07-26 paper/paper.tex
Intro updated

@@
             pdftitle={Distortion-corrected kernel density estimate on Riemannian manifolds},
-            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels},
+            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
             pdfborder={0 0 0},
@@
 However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
-We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
+We propose a novel distortion-corrected kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
 The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
@@
 \begin{keywords}
-manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
+manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
 \end{keywords}
@@
 Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
-It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
-To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
-
-Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular methods to calculate the probability density function from a sample dataset.
-KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
-There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}.
-For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ut}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]}.
-However, these density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
-\autocite{Henry2009-ll} In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from \textcite{Pelletier2005-vu} is not directly applicable.
+It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of \(\mathbb{R}^d\).
+Applications of kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on.
+However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than \(\mathbb{R}^d\), kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted.
+\textcite{Pelletier2005-vu} propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known.
+Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold.
+Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than \(\mathbb{R}^d\).
+Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function.
+This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.
+
+For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
+Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
+The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
+In this paper, we focus on the multivariate kernel density estimation.
+
+Note that a fixed bandwidth matrix \(\pmb{H}\) is a global smoothing parameter for all data points.
+However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
+The bandwidth is varied depending on either the location of the sample points {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]} or that of the estimated points {[}balloon estimator; \textcite{Terrell1992-ut}{]}.
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} {[}VKDE; Section 6.6 of \textcite{Scott2015-vl}{]} unless otherwise stated.
+However, these kernel density estimators are based on random samples in the Euclidean space.
+
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points.
+The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature \autocite{Henry2009-ll}.
+However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space.
+Therefore, the kernel density estimator from \textcite{Pelletier2005-vu} is not applicable because the geodesic distance and the volume density function are unknown.
 This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied to get a low-dimensional embedding.
-We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
 
@@
 To tackle this problem and measure the distortion incurred in manifold learning,
-\textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
 By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-
 The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding.
 Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
-The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
-In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(M\).
-This kernel density estimate can then be used to produce the highest density region plots\autocite{Hyndman1996-lk} for outlier visualization.
+The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.
+
+By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding.
+Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
+Then the DC\_KDE is used to estimate the density of the manifold and distortion induced by manifold learning methods are fixed with the estimated geometric information.
+Our distortion-corrected KDE is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
+These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
+Similar to \textcite{Cheng2021-ex}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
 
@@
 We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds.
-Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator to detect anomalies based on the embedding.
-\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere mapped in a 100-D space.
+Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection.
+\autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space.
 Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
@@
 Conclusions and discussions are presented in \autoref{conclusion}.
-Readers interested in the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.
+Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.
 
@@
 
-Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
+Kernel density estimation is one of the most popular methods to calculate the probability density function from a sample data.
 \textcite{Scott2015-vl} covers a wide range of techniques for estimating the kernel densities of multivariate data.
@@
 
-\caption{The proposed schematic for variable kernel density estimation with recovered geometry.}\label{fig:vkde}
+\caption{The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.}\label{fig:vkde}
 \end{figure}

### 2022-07-26 paper/paper.Rmd
Intro updated

@@
 This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
-keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
+keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
 # wpnumber: 03/2021
@@
 library(tidyverse)
-# library(dimRed)
-# library(reticulate)
 library(here)
-# library(viridis)
-library(hdrcde)
-# library(igraph)
-# library(matrixcalc)
-# library(akima)
-library(car)
-# library(ggforce)
-# library(ks)
+# library(hdrcde)
+# library(car)
 library(patchwork)
-# library(copula)
-# library(plotly)
 library(kableExtra)
@@
 
-<!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.  -->
 Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
+It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of $\mathbb{R}^d$.
+Applications of kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac], prime prediction via Twitter messages [@Gerber2014-tq], and so on.
+However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than $\mathbb{R}^d$, kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted.
+@Pelletier2005-vu propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known.
+Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold.
+Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than $\mathbb{R}^d$.
+Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function.
+This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.
+
+<!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
 It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
@@
 To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
-
-Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset.
-KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
-There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
-For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ut] or the data point [sample smoothing estimator; @Terrell1992-ut].
-However, these density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature.
-[@Henry2009-ll] In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from @Pelletier2005-vu is not directly applicable.
+-->
+
+For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
+<!-- There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].  -->
+Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
+The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
+In this paper, we focus on the multivariate kernel density estimation.
+
+Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points.
+However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
+The bandwidth is varied depending on either the location of the sample points [sample smoothing estimator; @Terrell1992-ut] or that of the estimated points [balloon estimator; @Terrell1992-ut].
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the *variable/adaptive kernel density estimation* [VKDE; Section 6.6 of @Scott2015-vl] unless otherwise stated.
+However, these kernel density estimators are based on random samples in the Euclidean space.
+
+For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points.
+The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature [@Henry2009-ll].
+However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space.
+Therefore, the kernel density estimator from @Pelletier2005-vu is not applicable because the geodesic distance and the volume density function are unknown.
 This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), are applied to get a low-dimensional embedding.
-We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
 
@@
 To tackle this problem and measure the distortion incurred in manifold learning,
-@Perrault-Joncas2013-pq propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
+@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
 By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-
 The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
 Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
-The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
-In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point.
-Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $M$.
-This kernel density estimate can then be used to produce the highest density region plots[@Hyndman1996-lk] for outlier visualization.
+The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.
+
+By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding.
+Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
+Then the DC_KDE is used to estimate the density of the manifold and distortion induced by manifold learning methods are fixed with the estimated geometric information.
+Our distortion-corrected KDE is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
+These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
+Similar to @Cheng2021-ex, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
 
@@
 
-Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
+Kernel density estimation is one of the most popular methods to calculate the probability density function from a sample data.
 @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
@@
 
-(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
+(ref:vkdecaption) The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.
 

### 2022-07-26 paper/paper.Rmd
Declare C as a function of p, but it is fixed for given y_i since p=y_i

@@
 \end{equation}
-where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the spaces so that the integral is 1.
+where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1.
 We will now state below how to calculate each term in the estimator.
@@
 The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension.
-We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
+We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and the estimated point $\pmb{p}$ so that for $\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}$, we have
 \begin{equation}
 \label{eq:vbs}
-C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
-\end{equation}
-where the distortion is corrected for the tangent plane of the embedding points.
+C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
+\end{equation}
+where $\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))$.
+The distortions are corrected for the tangent plane of the embedding points.
+Since we aim to estimate the density of the manifold for the embedding of sample points, $\pmb{p}$ is equal to $\pmb{y}_i$, for $i=1,\dots,N$.
+Therefore, we have
+\begin{equation}
+\label{eq:vbsyi}
+C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
+\end{equation}
 

### 2022-07-26 paper/paper.tex
Declare C as a function of p, but it is fixed for given y_i since p=y_i

@@
 \end{equation}
-where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the spaces so that the integral is 1.
+where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1.
 We will now state below how to calculate each term in the estimator.
@@
 The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension.
-We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
+We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and the estimated point \(\pmb{p}\) so that for \(\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}\), we have
 \begin{equation}
 \label{eq:vbs}
-C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
-\end{equation}
-where the distortion is corrected for the tangent plane of the embedding points.
+C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
+\end{equation}
+where \(\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))\).
+The distortions are corrected for the tangent plane of the embedding points.
+Since we aim to estimate the density of the manifold for the embedding of sample points, \(\pmb{p}\) is equal to \(\pmb{y}_i\), for \(i=1,\dots,N\).
+Therefore, we have
+\begin{equation}
+\label{eq:vbsyi}
+C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
+\end{equation}
 

### 2022-08-02 paper/paper.Rmd
p and q in vector format

@@
   email: Rob.Hyndman@monash.edu
-abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
+abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
+However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
+We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements.
 The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
+We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
 A shiny app in R is also developed for various simulation scenarios.
 The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
+This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
 keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
@@
 One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co].
+These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold [@Goldberg2008-co].
 To tackle this problem and measure the distortion incurred in manifold learning,
@@
 Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
-Then the DC_KDE is used to estimate the density of the manifold and distortion induced by manifold learning methods are fixed with the estimated geometric information.
-Our distortion-corrected KDE is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
+Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information.
+Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
 These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
@@
 Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection.
-\autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space.
+\autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space.
 Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
@@
 For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data.
-@Pelletier2005-vu generalize the multivariate kernel density esitmator to Riemnnian manifolds.
+@Pelletier2005-vu generalize the multivariate kernel density estimator to Riemannian manifolds.
 However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space.
@@
 where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise adaptive Riemannian matrix of dimension $d \times d$.
-The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by volume density function at each point.
+The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point.
 These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
@@
 Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from @Perrault-Joncas2013-pq.
-Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
+Finally, we show how to use these densities to carry out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
 
@@
 Assume $(M, d_g)$ as a complete metric space, where $d_g$ is the geodesic distance induced by the Riemannian metric $g$ and the strictly positive injectivity radius of the manifold $inj_g(M)$ [@Chavel2006-mp].
-Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathbb{R}^d$, as i.i.d.
+Denote $\pmb{X}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, as i.i.d.
 random objects on $M$ with density $f$.
@@
 Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, $r\leq inj_g(M)$.
-Under such condition, for each $\pmb{p} \in M$, there exists a normal coordinate neighborhood of $\pmb{p}$ containing an open ball $B(\pmb{p}, r) \in T_pM$ centered at $\pmb{p}$ with radius $r$.
+Under such condition, for each $\pmb{p} \in M$, there exists a normal coordinate neighborhood of $\pmb{p}$ containing an open ball $B(\pmb{p}, r) \in T_{\pmb{p}}M$ centered at $\pmb{p}$ with radius $r$.
 
 Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
-For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denote a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
-Then $\frac{Dw_i}{dt}(0)$ forms an orthonomal basis of the tangent space $T_pM$.
-The volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$ is defined in @Le_Brigant2019-lj as 
+For any tangent vector $u \in T_{\pmb{p}}M$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denote a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
+Then $\frac{Dw_i}{dt}(0)$ forms an orthonomal basis of the tangent space $T_{\pmb{p}}M$.
+The volume density function on $M$, $\theta_{\pmb{p}}(u): T_{\pmb{p}}M \rightarrow \mathbb{R}$ is defined in @Le_Brigant2019-lj as 
 \begin{equation}
 \label{eq:volden}
-\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-\end{equation}
-
-For the exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives
-$$
-d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
+\theta_{\pmb{p}}(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+\end{equation}
+
+For the exponential map $\exp_{\pmb{p}}: T_{\pmb{p}}M \rightarrow M$ induced by pullback of a volume form $\exp_{\pmb{p}}^*\textit{Vol}$ on $T_{\pmb{p}}M$, $\theta_{\pmb{p}}$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_{\pmb{p}}M$, which gives
+$$
+d\exp_{\pmb{p}}^*\textit{Vo}l(u) = \theta_{\pmb{p}}(u)du,
 $$
@@
 
-@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of q, $\exp_q^{-1}(p)$.
+@Pelletier2005-vu mention that the volume density function $\theta_{\pmb{p}}(\pmb{q})$ defined for $\pmb{q}$ in a neighborhood of $\pmb{p}$ on the manifold, with the geodesic normal coordinates at $\pmb{p}$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of $\pmb{q}$, $\exp_{\pmb{p}}^{-1}(\pmb{q})$.
 Note that this is true only when using the geodesic normal coordinates.
@@
 geodesic normal coordinates.
-In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_p(q)) |$.
+In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_{\pmb{p}}(\pmb{q})) |$.
 In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
@@
 
-For $p \in M$, denote an open ball centered at $p$ with radius $r$ as $B(p,r)$ and the open neighborhood of $0_p \in T_pM$ can be denoted as $B(0_p,r)=\exp_p^{-1}(B(p,r))$.
-Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_p,r)$, with $g^{\prime} = \exp^{\ast}_p(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_p$ in $B(0_p,r)$.
-Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$, we have $u \in \bar{U}$. 
+For $\pmb{p} \in M$, denote an open ball centered at $\pmb{p}$ with radius $r$ as $B(\pmb{p},r)$ and the open neighborhood of $0_{\pmb{p}} \in T_{\pmb{p}}M$ can be denoted as $B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))$.
+Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_{\pmb{p}},r)$, with $g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_{\pmb{p}}$ in $B(0_{\pmb{p}},r)$.
+Let $u \in B(0_{\pmb{p}},r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$, we have $u \in \bar{U}$. 
 The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$.
 The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om].
-Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
+Therefore, given a point $\pmb{q} \in B(\pmb{p},r)$ and a tangent vector $u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
 \begin{equation}
@@
 \end{equation}
-The volume density function is only defined for points satisfying $d_g(p,q) < inj_gM$ so that there exists an open ball $B(p,r) \in T_pM$.
-And $\theta_p(q)=1$ for all $p,q \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.
+The volume density function is only defined for points satisfying $d_g(\pmb{p},\pmb{q}) < inj_gM$ so that there exists an open ball $B(\pmb{p},r) \in T_{\pmb{p}}M$.
+And $\theta_{\pmb{p}}(\pmb{q})=1$ for all $\pmb{p},\pmb{q} \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.
 
@@
 There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation.
+The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathbb{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathbb{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation.
 Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels.

### 2022-08-02 paper/paper.tex
p and q in vector format

@@
 \begin{abstract}
-Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point.
+Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
+However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
+We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements.
 The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
+We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
 A shiny app in R is also developed for various simulation scenarios.
 The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data.
+This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
 \end{abstract}
@@
 One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}.
+These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold \autocite{Goldberg2008-co}.
 To tackle this problem and measure the distortion incurred in manifold learning,
@@
 Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
-Then the DC\_KDE is used to estimate the density of the manifold and distortion induced by manifold learning methods are fixed with the estimated geometric information.
-Our distortion-corrected KDE is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
+Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information.
+Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
 These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
@@
 For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data.
-\textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds.
+\textcite{Pelletier2005-vu} generalize the multivariate kernel density estimator to Riemannian manifolds.
 However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space.
@@
 where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
-The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by volume density function at each point.
+The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point.
 These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
@@
 Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from \textcite{Perrault-Joncas2013-pq}.
-Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
+Finally, we show how to use these densities to carry out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
 
@@
 Assume \((M, d_g)\) as a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}.
-Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathbb{R}^d\), as i.i.d.
+Denote \(\pmb{X}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), as i.i.d.
 random objects on \(M\) with density \(f\).
@@
 Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, \(r\leq inj_g(M)\).
-Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_pM\) centered at \(\pmb{p}\) with radius \(r\).
+Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_{\pmb{p}}M\) centered at \(\pmb{p}\) with radius \(r\).
 
 Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
-For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
-Then \(\frac{Dw_i}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
-The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined in \textcite{Le_Brigant2019-lj} as
+For any tangent vector \(u \in T_{\pmb{p}}M\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
+Then \(\frac{Dw_i}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_{\pmb{p}}M\).
+The volume density function on \(M\), \(\theta_{\pmb{p}}(u): T_{\pmb{p}}M \rightarrow \mathbb{R}\) is defined in \textcite{Le_Brigant2019-lj} as
 \begin{equation}
 \label{eq:volden}
-\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-\end{equation}
-
-For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
+\theta_{\pmb{p}}(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
+\end{equation}
+
+For the exponential map \(\exp_{\pmb{p}}: T_{\pmb{p}}M \rightarrow M\) induced by pullback of a volume form \(\exp_{\pmb{p}}^*\textit{Vol}\) on \(T_{\pmb{p}}M\), \(\theta_{\pmb{p}}\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_{\pmb{p}}M\), which gives
 \[
-d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
+d\exp_{\pmb{p}}^*\textit{Vo}l(u) = \theta_{\pmb{p}}(u)du,
 \]
@@
 
-\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
+\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_{\pmb{p}}(\pmb{q})\) defined for \(\pmb{q}\) in a neighborhood of \(\pmb{p}\) on the manifold, with the geodesic normal coordinates at \(\pmb{p}\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of \(\pmb{q}\), \(\exp_{\pmb{p}}^{-1}(\pmb{q})\).
 Note that this is true only when using the geodesic normal coordinates.
@@
 When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates.
-In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
+In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_{\pmb{p}}(\pmb{q})) |\).
 In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
@@
 
-For \(p \in M\), denote an open ball centered at \(p\) with radius \(r\) as \(B(p,r)\) and the open neighborhood of \(0_p \in T_pM\) can be denoted as \(B(0_p,r)=\exp_p^{-1}(B(p,r))\).
-Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_p,r)\), with \(g^{\prime} = \exp^{\ast}_p(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_p\) in \(B(0_p,r)\).
-Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\), we have \(u \in \bar{U}\).
+For \(\pmb{p} \in M\), denote an open ball centered at \(\pmb{p}\) with radius \(r\) as \(B(\pmb{p},r)\) and the open neighborhood of \(0_{\pmb{p}} \in T_{\pmb{p}}M\) can be denoted as \(B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))\).
+Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_{\pmb{p}},r)\), with \(g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_{\pmb{p}}\) in \(B(0_{\pmb{p}},r)\).
+Let \(u \in B(0_{\pmb{p}},r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_{\pmb{p}},r)\), we have \(u \in \bar{U}\).
 The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\).
 The quotient of these two volumes is then independent of the chart \((\bar{U}, \bar{\psi})\) \autocite{Henry2009-om}.
-Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
+Therefore, given a point \(\pmb{q} \in B(\pmb{p},r)\) and a tangent vector \(u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_{\pmb{p}},r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
 \begin{equation}
@@
 \end{equation}
-The volume density function is only defined for points satisfying \(d_g(p,q) < inj_gM\) so that there exists an open ball \(B(p,r) \in T_pM\).
-And \(\theta_p(q)=1\) for all \(p,q \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.
+The volume density function is only defined for points satisfying \(d_g(\pmb{p},\pmb{q}) < inj_gM\) so that there exists an open ball \(B(\pmb{p},r) \in T_{\pmb{p}}M\).
+And \(\theta_{\pmb{p}}(\pmb{q})=1\) for all \(\pmb{p},\pmb{q} \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.
 
@@
 Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\).
-Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
+Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.
 Then the volume density function is given by
@@
 There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
+The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathbb{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathbb{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
 Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels.

### 2022-09-08 R/sources/mldata.R
density after change of coordinates

@@ -89,6 +89,7 @@ mldata <- function(N = 2000, p = 2, meta = c("uniform", "copula", "gaussian"),
          "Twin Peak" = {
            u <- 2 * co[, 1:2] - 1
            data <- twinPeaksMapping(u[,1], u[,2])
+           den <- 1/2 * den # change of coordinates from co to u
          }, 
          "S Curve" = {
            u <- cbind(1.5 * pi * (2 * co[,1] - 1), 2 * co[,2])

### 2022-09-08 R/sources/dr_quality.R
density after change of coordinates

@@ -42,7 +42,9 @@ dr_quality <- function(X, Y, K = 20, nn.idx, treetype = "kd", searchtype = "stan
   # auc_lnK_Rnx <- dimRed::quality(e, "AUC_lnK_R_NX")
   # suppressMessages(procrustes <- proc.measure(X, Y, K, nn.idx, treetype = treetype, searchtype = searchtype)$R)
   
+  ## UNCOMMENT IF PROCRUSTES MEASURE IS NEEDED
   # procrustes <- proc.measure(X, Y, K, nn.idx, treetype = treetype, searchtype = searchtype)$R_N # standerdized procrustes measure R_N, unstanderdized $R
+  
   # K_optim <- as.numeric(names(Qglobal))
   
   return(list(Q = Q, 

### 2022-09-08 paper/paper.tex
Minor

@@
 \textbf{Anastasios Panagiotelis}\newline
-University of Sydney
+The University of Sydney
 \newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
@@
 
+% Adjust headwidth in case user has changed geometry in header-includes
+\renewcommand{\headwidth}{\textwidth}
+
 \begin{document}
@@
 For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
+Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][]{Heidenreich2013-bl}[ for details]{Scott2015-vl}.
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef,Cao1994-st,Jones1996-cb}[ for reviews]{Wand1994-xu}.
 The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
@@
 
+\hypertarget{s1simulation}{%
+\section{\texorpdfstring{Appendix: Simulation of \(S^1\) for manifold density estimation}{Appendix: Simulation of S\^{}1 for manifold density estimation}}\label{s1simulation}}
+
+To start with, we demonstrate the use of the proposed DC-KDE on a \(1\)-dimensional example.
+For a \(S^1\) curve centered at \(O\) with radius \(1\), the coordinated of points on the curve could be expressed in polar coordinates \((1, \alpha)\) or Cartesian coordinates \((x, \sqrt{1-x^2})\).
+
+The first simulation starts with a \(1\)-dimensional example.
+
+Manifold: \(S^1\) with center 0 and radius \(r=1\)
+
+Polar coordinates: \((r, \theta)\)
+
+Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).
+
+Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)
+
+Density of \(\theta\): \(f(\theta) = 2 / \pi\)
+
+Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)
+
+The CDF of X is given by
+\[
+P(X \leq x) = P(\cos(\theta) \leq x) \\
+= P(\theta \geq \arccos(x)) \\
+= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+= 1 - \frac{2}{\pi} \arccos(x).
+\]
+Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
+f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+= \frac{2}{\pi \sqrt{1 - x^2}}.
+\]
+
+Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
+estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
+and compare it with its true density \(f(x)\).
+We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
+\[
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
+\]
+where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
+
+Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\).
+Therefore, the geodesic distance is
+\[
+d_g(p, X_i) = \arccos(p) - \arccos(X_i)
+\]
+As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).
+
+{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}
+
+The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
+\[
+g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
+\]
+is smooth in a local neighborhood \(U\) of \(M\).
+
+Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
+\[
+\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
+\]
+Therefore, the estimator for \(p \in S^1\) is given by
+\[
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+\]
+where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.
+
+\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
+\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}
+
+\hypertarget{example-of-s2-in-henry2009-om.}{%
+\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}
+
+Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\).
+Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\).
+Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
+\[
+\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
+\]
+if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
+Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
+Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
+\[
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
+\]
+\[
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
+\]
+The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
+Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
+Therefore, we obtain that the volume density on the sphere is
+\[
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
+\]
+for \(q \neq p, -p\), and \(\theta_p(p) = 1\).
+
+\hypertarget{s1}{%
+\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}
+
+Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
+\[
+\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
+\]
+and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
+\[
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
+\]
+So the riemannian metric is \(g_q = \cos(r)\) and further,
+\[
+\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
+\]
+where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).
+
 \hypertarget{swissrollappe}{%

### 2022-09-08 paper/paper.Rmd
Minor

@@
   othernames: Anastasios
-  address: University of Sydney
+  address: The University of Sydney
   email: Anastasios.Panagiotelis@sydney.edu.au
@@
 
-<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->
-
-<!-- ## Volume density function [MOVED TO SECTION 2.2]
-
-For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$.
-Then $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$.
-Define the volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$, as
-$$
-\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-$$
-
-The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates
-$$
-d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
-$$
-The volume density function is motivated by how a measure $du$ changes under the change of coordinates.
-It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
--->
-
-<!-- the volume density function is the quotient of the canonical measure of the Riemannian metric $\exp_p{g}$ on $T_pM$ by the Lebesgue measure of the Euclidean structure $g_p$ on $T_pM$. -->
-
-<!-- ## Example of $S^1$ for manifold density estimation [TODO: rewrite as appendix B with plots?] -->
-
-<!-- Manifold: $S^1$ with center 0 and radius $r=1$ -->
-
-<!-- Polar coordinates: $(r, \theta)$ -->
-
-<!-- Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$. -->
-
-<!-- Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$ -->
-
-<!-- Density of $\theta$: $f(\theta) = 2 / \pi$ -->
-
-<!-- Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$ -->
-
-<!-- The CDF of X is given by  -->
-<!-- $$ -->
-<!-- P(X \leq x) = P(\cos(\theta) \leq x) \\ -->
-<!-- = P(\theta \geq \arccos(x)) \\ -->
-<!-- = \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\ -->
-<!-- =  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\ -->
-<!-- = 1 - \frac{2}{\pi} \arccos(x). -->
-<!-- $$  -->
-<!-- Then the density of X is the derivative of the CDF w.r.t. $x$ $$ -->
-<!-- f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\ -->
-<!-- = \frac{2}{\pi \sqrt{1 - x^2}}. -->
-<!-- $$ -->
-
-<!-- Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could -->
-<!-- estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ -->
-<!-- and compare it with its true density $f(x)$. -->
-<!-- We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by -->
-<!-- $$ -->
-<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}), -->
-<!-- $$  -->
-<!-- where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
--->
-
-<!-- Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
-Therefore, the geodesic distance is  -->
-<!-- $$ -->
-<!-- d_g(p, X_i) = \arccos(p) - \arccos(X_i) -->
-<!-- $$  -->
-<!-- As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
+# Appendix: Simulation of $S^1$ for manifold density estimation {#s1simulation}
+
+To start with, we demonstrate the use of the proposed DC-KDE on a $1$-dimensional example.
+For a $S^1$ curve centered at $O$ with radius $1$, the coordinated of points on the curve could be expressed in polar coordinates $(1, \alpha)$ or Cartesian coordinates $(x, \sqrt{1-x^2})$.
+
+The first simulation starts with a $1$-dimensional example.
+
+Manifold: $S^1$ with center 0 and radius $r=1$
+
+Polar coordinates: $(r, \theta)$
+
+Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
+
+Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$
+
+Density of $\theta$: $f(\theta) = 2 / \pi$
+
+Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$
+
+The CDF of X is given by
+$$
+P(X \leq x) = P(\cos(\theta) \leq x) \\
+= P(\theta \geq \arccos(x)) \\
+= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+= 1 - \frac{2}{\pi} \arccos(x).
+$$
+Then the density of X is the derivative of the CDF w.r.t. $x$ $$
+f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+= \frac{2}{\pi \sqrt{1 - x^2}}.
+$$
+
+Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
+estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
+and compare it with its true density $f(x)$.
+We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
+$$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
+$$
+where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
+
+Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
+Therefore, the geodesic distance is
+$$
+d_g(p, X_i) = \arccos(p) - \arccos(X_i)
+$$
+As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
 $|g(\exp_p^{-1}(x_i))|$.
--->
-
-<!-- [When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.] -->
-
-<!-- The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,  -->
-<!-- $$ -->
-<!-- g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R}, -->
-<!-- $$  -->
-<!-- is smooth in a local neighborhood $U$ of $M$. -->
-
-<!-- Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by  -->
-<!-- $$ -->
-<!-- \theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}. -->
-<!-- $$  -->
-<!-- Therefore, the estimator for $p \in S^1$ is given by  -->
-<!-- $$ -->
-<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg), -->
-<!-- $$  -->
-<!-- where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*. -->
-
-<!-- ### How to derive volume density function $\theta_{X_i}(p)$ -->
-
-<!-- #### Example of $S^2$ in @Henry2009-om. -->
-
-<!-- Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
+
+[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
+
+The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
+$$
+g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
+$$
+is smooth in a local neighborhood $U$ of $M$.
+
+Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
+$$
+\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
+$$
+Therefore, the estimator for $p \in S^1$ is given by
+$$
+f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
+$$
+where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*.
+
+### How to derive volume density function $\theta_{X_i}(p)$
+
+#### Example of $S^2$ in @Henry2009-om.
+
+Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
 Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$.
-Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by  -->
-<!-- $$ -->
-<!-- \Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw), -->
-<!-- $$  -->
-<!-- if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.  -->
-<!-- Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$. -->
-<!-- Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that -->
-<!-- $$ -->
-<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, -->
-<!-- $$ -->
-<!-- $$ -->
-<!-- B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w. -->
-<!-- $$  -->
-<!-- The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors. -->
-<!-- Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
--->
-<!-- Therefore, we obtain that the volume density on the sphere is  -->
-<!-- $$ -->
-<!-- \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\ -->
-<!-- = \frac{|\sin(d_g(p,q))|}{d_g(p,g)}, -->
-<!-- $$  -->
-<!-- for $q \neq p, -p$, and $\theta_p(p) = 1$. -->
-
-<!-- #### $S^1$ -->
-
-<!-- Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization  -->
-<!-- $$ -->
-<!-- \Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv, -->
-<!-- $$ -->
-<!-- and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have -->
-<!-- $$ -->
-<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v. -->
-<!-- $$ -->
-<!-- So the riemannian metric is $g_q = \cos(r)$ and further,  -->
-<!-- $$ -->
-<!-- \theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|}, -->
-<!-- $$ -->
-<!-- where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$. -->
+Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
+$$
+\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
+$$
+if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.
+Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
+Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
+$$
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
+$$
+$$
+B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
+$$
+The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
+Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
+Therefore, we obtain that the volume density on the sphere is
+$$
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
+$$
+for $q \neq p, -p$, and $\theta_p(p) = 1$.
+
+#### $S^1$
+
+Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization
+$$
+\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
+$$
+and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have
+$$
+A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
+$$
+So the riemannian metric is $g_q = \cos(r)$ and further,
+$$
+\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
+$$
+where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$.
 

### 2022-09-08 R/SphericalCap.R
Sample uniformly on S1 and S2 with change of coordinates

@@ -8,7 +8,7 @@ Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1)
 
 # Simulate uniformly from unit sphere
-n <- 2000
+n <- 1000
 
 x <- matrix(rnorm(n * 3), n, 3)
 x <- t(apply(x, 1, function(a) {a / sqrt(sum(a ^ 2))} ))
@@ -192,23 +192,50 @@ rmetric[,,N]
 # adj_matrix[N,]
 
 # Metric embedding of p: locally isometric embedding
-fn_metric <- det(rmetric[,,N]) ^ (-1/2) * fn
+# fn_metric = hn ^ (-1/2) %*% fn
+hp <- expm::sqrtm(solve(rmetric[,,N]))
+fn_metric <- t(apply(fn, 1, function(x) t(hp %*% x)))
 fn_metric[N,]
 
-plot(fn_metric, asp = 1, col = c(cc, "red"), xlim = c(-1,1), ylim=c(-1,1))
-plot(rbind(y,c(0,0)), asp = 1, col = c(cc, "red"), xlim = c(-1,1), ylim=c(-1,1))
-mean(sum((rbind(y, c(0,0)) - fn_metric)^2))
+par(mfrow=c(1,3))
+plot(rbind(y,c(0,0)), asp = 1, col = c(cc, "red"), xlim = c(-2, 2), ylim = c(-2, 2))
+plot(fn_metric, asp = 1, col = c(cc, "red"), xlim = c(-2, 2), ylim = c(-2, 2))
+plot(fn, asp = 1, col = c(cc, "red"), xlim = c(-2, 2), ylim = c(-2, 2))
 
-# Other forms
-det(rmetric[,,N]) ^ (-1/2) * fn[N,]
-sqrt(det(rmetric[,,N]) * mean(apply(rmetric, 3, det)[nn.idx[N,]]) ) * fn[N,]
-sqrt(det(rmetric[,,N]) / mean(apply(rmetric, 3, det)[nn.idx[N,]]) ) * fn[N,]
-sqrt(mean(apply(rmetric, 3, det)[nn.idx[N,]]) / det(rmetric[,,N]) ) * fn[N,]
+# mean(sum((rbind(y, c(0,0)) - fn_metric)[c(which(cc == "black"), N)]^2))
+# mean(sum((rbind(y, c(0,0)) - fn)[c(which(cc == "black"), N)]^2))
[…]

### 2022-09-08 R/uniform_sample_polar_coordinates.R
Sample uniformly on S1 and S2 with change of coordinates

@@ -1,0 +1,27 @@
+# sample uniformly from the unit disk
+# The unit disk is given by the polar coordinates in the set [0,1] x [0,2pi)
+# How to sample the polar coordinates so that the resulting point distribution is uniform on the disk
+# Reference: Section 3.3 of https://www.cs.cornell.edu/courses/cs6630/2015fa/notes/pdf-transform.pdf
+# (r, phi) -> (x, y)
+# x = r*cos(phi); y = r*sin(phi)
+# r := sqrt(e1); phi = 2 * pi * e2; where e1, e2 ~ U(0,1)
+set.seed(1234)
+N <- 1000
+e1 <- runif(N, 0, 1)
+e2 <- runif(N, 0, 1)
+r <- sqrt(e1)
+phi <- 2 * pi * e2
+# e1, e2 ~ U(0,1)
+# density of r: p_r(r) = p_{e1}(e1) * (2*r) = 2*r
+# density of phi: p_phi(phi) = p_{e2}(e2) * 1 / (2 * pi) = 1/(2 * pi)
+# CDF of r: P_r(r) = \int_0^r  p_r(x) dx = r^2
+# CDF of phi: P_phi(phi) = \int_0^phi (2*pi) dx = 2*pi*phi
+# Density of (r, phi): p_A(r, phi) = p_r(r) * p_phi(phi) = 2 * r * 1 / (2 * pi) = r / pi
+# Density of (x, y): p_B(x, y) = p_A(r, phi) / |J| = (r / pi) / r = 1 / pi
+# Therefore, (x, y) is uniformly distributed with density 1 / pi.
+x <- r * cos(phi)
+y <- r * sin(phi)
+plot(x, y, asp = 1)
+
+# True density of (x,y) is known as constant, 1 / pi
+# Test if (x, y) is uniformly distributed in the set [0, 1] x [0, 1] 

### 2022-09-15 R/sources/mldata.R
Tune parameters for mldata()

@@ -36,7 +36,7 @@ mldata <- function(N = 2000, p = 2, meta = c("uniform", "copula", "gaussian"),
           "gaussian" = {
             n <- round(N/4)
             R <- matrix(c(.02, 0,
-                          0, .02)*.5, # variance-covariance matrix
+                          0, .02) * .5, # variance-covariance matrix
                         nrow = 2, ncol = 2)
             # mu <- matrix(#~mu1, ~mu2,
             #               c(7.5, 7.5,

### 2022-09-15 R/sources/vkde.R
Tune parameters for mldata()

@@ -87,20 +87,21 @@ vkde <- function(x, h = NULL, vh = NULL, r = NULL, gridsize = 20, xmin = apply(x
       hi <- vh[,,i]
       # z <- cbind(z, mvtnorm::dmvnorm(x = eval.points, mean = x[k,], sigma = hk))  # stack vector of length n
       bindex <- which(adj_matrix[i,] <= r) # only smooth over points within the bandwidth radius ## ADD AN ARGUMENT AND CHECK DIMENSION, NEW RADIUS
-      # hi <- hi / mean(vh[,,bindex]) ## ???
+      # hi <- hi / mean(vh[,,bindex]) ## NOT used
       for(j in bindex) {
         hj <- vh[,,j]
         z[i,j] <- 
-          # (det(hj) / det(hi)) ^ (-0.5) * # volume density function !!! (1)
-          # (det(hj) / det(hi)) ^ (0.5) * # (2)
-          # (det(hj) * det(hi)) ^ (-0.5) * # (3)
-          (det(hj) * det(hi)) ^ (0.5) * # (4)
+          (det(hi) / det(hj)) ^ (-0.5) * # volume density function !!! (1)
+          # (det(hi) / det(hj)) ^ (0.5) * # (2) reversed theta
+          # (det(hi) * det(hj)) ^ (-0.5) * # (3)
+          # (det(hi) * det(hj)) ^ (0.5) * # (4)
           (2 * pi) ^ (-d / 2) * r ^ (-d) * 
           exp( -0.5 / (r^2) * t(eval.points[i,] - eval.points[j,]) %*% solve(hi) %*% (eval.points[i,] - eval.points[j,]) ) / 
           (pnorm(1) - pnorm(-1)) # suppK = [0,1], scale to integral to 1
       }
     }
-    z <- colMeans(z, na.rm = TRUE)
+    # Density for y_i
+    z <- rowMeans(z, na.rm = TRUE) # !!! rowMeans for hi; colMeans for hj, change d_g accordingly
     
     f <- list(x = x, eval.points = eval.points, estimate = z, H = vh, r = r, gridded = FALSE)
   }

### 2022-09-15 R/uniform_sample_polar_coordinates.R
Simulation for twin peaks

@@ -1,6 +1,8 @@
 # How to sample uniformly from 3-D manifolds (mapping function) and transform the density after change of coordinates
-
+library(tidyverse)
 library(scatterplot3d)
+library(plotly)
+source("R/sources/mldata.R")
 ## --------------------------------------------------------
 # 1. Sample uniformly from the unit disk
 ## --------------------------------------------------------

### 2022-09-15 R/Simulation2d_swissroll_twinpeak.R
Simulation for twin peaks

@@ -24,7 +24,7 @@ Jmisc::sourceAll(here::here("R/sources"))
 # set.seed(1)
 
 # data size
-N <- 2000
+N <- 1000
 p <- 2
 
 # # ----Swiss roll dataset with a sparse area------------------------

### 2022-09-15 R/SphericalCap.R

@@ -213,6 +213,10 @@ fn[N,]
 rmetric[,,N]
 # adj_matrix[N,]
 
+
+
+
+
 ## --------------------------------------------------------
 # Metric embedding of p: locally isometric embedding
 # fn_metric = hn ^ (-1/2) %*% fn # WRONG
@@ -259,6 +263,9 @@ points(f_metric, col = "green")
 
 
 
+
+
+
 ## --------------------------------------------------------
 ## Check local coordinate change 
 # g_q = h_p ^ {-.5} %*% h_q %*% h_p ^ {-.5}
@@ -271,19 +278,15 @@ for(i in 1:N){
   bindex <- which(adj_matrix[i,] <= r)
   # theta[i] <- sqrt(rmetric_det[i])
   # theta[i] <- sqrt(rmetric_det[i]) ^ (-1)
-  theta[i] <- sqrt(mean(rmetric_det[bindex]) * rmetric_det[i] )
+  # theta[i] <- sqrt(mean(rmetric_det[bindex]) * rmetric_det[i] )
   # theta[i] <- sqrt(mean(rmetric_det[bindex]) / rmetric_det[i] ) ^ (-1)
-  # theta[i] <- sqrt( mean(rmetric_det[bindex]) / rmetric_det[i] ) # used in the paper, (\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} )^{1/2}
-  fn_dc[i,] <- fn[i,] * theta[i]
+  theta[i] <- sqrt( rmetric_det[i] / mean(rmetric_det[bindex]) ) # used in the paper, (\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} )^{1/2}
+  fn_dc[i,] <- fn[i,] * theta[i] ^ (-1)
 }
 par(mfrow=c(1,3))
 plot(rbind(y, c(0,0)), asp = 1, col = c(cc, "red"), xlim = c(-2, 2), ylim=c(-2, 2))
 plot(fn_dc, asp = 1, col = c(cc, "red"), xlim = c(-2, 2), ylim=c(-2, 2))
 plot(fn, asp = 1, col = c(cc, "red"), xlim = c(-2, 2), ylim=c(-2, 2))
-
-
-
-
 
 
 ## --------------------------------------------------------
@@ -317,7 +320,6 @@ f_dc <- fn_dc[cap_index,]
 # # $quality
[…]

### 2022-09-15 R/TwinPeaks.R

@@ -19,18 +19,18 @@ library(ks)
 library(patchwork)
 # library(copula)
 library(plotly)
-# library(kableExtra)
+library(kableExtra)
 Jmisc::sourceAll(here::here("R/sources"))
 
 # data size
-N <- 2000
+N <- 1000
 p <- 2
 
 ###--------------------------------------------------------
 ## Start from here! 
 ###--------------------------------------------------------
-set.seed(1234)
-mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[1] # only run [1] and [3]
+set.seed(123)
+mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3] # only run [1] and [3]
 sr <- mldata(N = N, meta = "gaussian", mapping = mapping)
 # sr <- mldata(N = N, meta = "uniform", mapping = mapping)
 swissroll <- sr$data %>% as.data.frame()
@@ -46,7 +46,13 @@ summary(sr$den) # [0, 1]
 # P_A(u,v) = sr$den
 u <- preswissroll$X1
 v <- preswissroll$X2
-den_twinpeaks <- sr$den / sqrt(1 + (pi * cos(pi * u) * tan(3 * v)) ^ 2 + 9 * sin(pi * u) ^ 2 / (cos(3 * v)) ^ 4)
+if(mapping == "Swiss Roll") {
+  den_twinpeaks <- sr$den / sqrt(1 + u ^ 2)
+} else if(mapping == "Twin Peak") {
+  den_twinpeaks <- sr$den / sqrt(1 + (pi * cos(pi * u) * tan(3 * v)) ^ 2 + 9 * sin(pi * u) ^ 2 / (cos(3 * v)) ^ 4)
+} else {
+  print("Double check mldata() for the density after change of coordinates!")
+}
 summary(den_twinpeaks)
 plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = den_twinpeaks, # colored with 3d density
         type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
@@ -58,7 +64,7 @@ train <- swissroll
 # Parameters fixed
 x <- train
 y <- sr$metadata # given known embedding with known density
-y <- NULL
+# y <- NULL
 N <- nrow(x)
 s <- 2
 k <- 20
[…]

### 2022-09-15 paper/paper.Rmd
Remove C

@@
 \label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation}
@@
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
-\end{equation}
-where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1.
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\end{equation}
+where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$
+<!-- , and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1 -->
+.
 We will now state below how to calculate each term in the estimator.

### 2022-09-26 paper/paper.tex
Upload twin peaks plots

@@
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
 \end{equation}
-where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1.
+where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\)
+.
 We will now state below how to calculate each term in the estimator.
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_4kernels}
+{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_twinpeaks_dc}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak5levels_outliers_comparison_4ml_3cases_riem0_1}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
 
@@
 \midrule
-Variable bandwidth & \textbf{0.899} & 0.385 & 0.620 & 0.259 & \textbf{0.659}\\
-Fixed bandwidth & 0.626 & \textbf{0.399} & \textbf{0.622} & \textbf{0.663} & 0.653\\
+DC-KDE & \textbf{0.823} & \textbf{0.673} & \textbf{0.672} & \textbf{0.806} & \textbf{0.794}\\
+KDE & 0.798 & 0.500 & 0.606 & 0.451 & 0.469\\
 \bottomrule
@@
 As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
-Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259.
+Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.451.
 This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak_density_comparison_isomap_riem0_1_rank}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_isomap_r0_5_rank}
 
@@
 The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
-Variable bandwidth shows a strong linear positive relationship.}\label{fig:tpisomapden}
+DC-KDE shows a strong linear positive relationship.}\label{fig:tpisomapden}
 \end{figure}
@@
 fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_comparison_isomapvs4ml_r0_5_rank}
+
+}
+
+\caption{Comparison of outliers found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpisomapvs4ml}
+\end{figure}
+
+Finally, \autoref(fig:tpisomapvs4ml) is used to show the robustness of the density estimates using DC-KDE regardless of the embedding method.
+By plotting the density ranks of LLE, Laplacian Eigenmaps, t-SNE, and UMAP against ISOMAP, we could tell that the rank correlation from DC-KDE is higher that that from KDE, meaning that the rank of DC-KDE are more consistent with the manifold learning methods.
+Compared with KDE where the density ranks varies with the embedding methods even for the same manifold, DC-KDE is more robust when the distortions are fixed.
 

### 2022-09-26 paper/paper.Rmd
Upload twin peaks plots

@@
 The shapes indicate the four kernels.", echo = FALSE}
-knitr::include_graphics("figures/truedensity_4kernels.png")
+knitr::include_graphics("figures/truedensity_twinpeaks_dc.png")
 ```
@@
 Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak5levels_outliers_comparison_4ml_3cases_riem0_1.png")
+knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
 ```
@@
 ```{r tpcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_Twin Peak_4ml_riem0_1.rda")
+load("figures/CorrelationTable_Twin Peak2000_5ml_r0_5.rda")
 col.max <- apply(cors, 2, max)
@@
 The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
-Variable bandwidth shows a strong linear positive relationship.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1_rank.png")
+DC-KDE shows a strong linear positive relationship.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank.png")
 ```
@@
 fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+
+```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak2000_density_comparison_isomapvs4ml_r0_5_rank.png")
+```
+
+Finally, \autoref(fig:tpisomapvs4ml) is used to show the robustness of the density estimates using DC-KDE regardless of the embedding method.
+By plotting the density ranks of LLE, Laplacian Eigenmaps, t-SNE, and UMAP against ISOMAP, we could tell that the rank correlation from DC-KDE is higher that that from KDE, meaning that the rank of DC-KDE are more consistent with the manifold learning methods.
+Compared with KDE where the density ranks varies with the embedding methods even for the same manifold, DC-KDE is more robust when the distortions are fixed.
 

### 2022-09-26 R/TwinPeaks.R

@@ -69,7 +69,7 @@ gridsize <- 20
 noutliers <- 20
 
 metric_meta <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
-                          # annmethod = annmethod, distance = distance, treetype = treetype, 
+                          annmethod = annmethod, distance = distance, treetype = treetype,
                           searchtype = searchtype
 )
 # summary(metric_meta)
@@ -101,8 +101,9 @@ plotmanifold <- preswissroll %>%
   labs(color = "Density" # , shape = "Kernels"
        )
 plotmanifold
-# plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = den_2dmanifold, # colored with 3d density
-#         type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
+ggsave("paper/figures/truedensity_twinpeaks_dc.png", plotmanifold, width = 8, height = 6, dpi = 300)
+plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = den_2dmanifold, # colored with 3d density
+        type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
 trueden <- den_2dmanifold # sr$den
 
 
@@ -288,6 +289,7 @@ metric_lle <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method
 )
 fn <- metric_lle$embedding
 Rn <- metric_lle$rmetric
+adj_matrix <- metric_lle$adj_matrix
 E1 <- fn[,1]; E2 <- fn[,2]
 # flle <- vkde2d(x = E1, y = E2, h = Rn*riem.scale, gridsize = gridsize)
 # fxy_lle <- hdrcde:::interp.2d(flle$x, flle$y, flle$z, x0 = E1, y0 = E2)
@@ -341,6 +343,7 @@ metric_le <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method,
 )
 fn <- metric_le$embedding
 Rn <- metric_le$rmetric
+adj_matrix <- metric_le$adj_matrix
 E1 <- fn[,1]; E2 <- fn[,2]
 # fle <- vkde2d(x = E1, y = E2, h = Rn*riem.scale, gridsize = gridsize)
[…]

### 2022-09-26 paper/paper.tex
Remove C and colors for 4 kernels

@@
 
-As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|\pmb{p}-\pmb{q}\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality.
-By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric.
-In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1.
-The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension.
-We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and the estimated point \(\pmb{p}\) so that for \(\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}\), we have
-\begin{equation}
-\label{eq:vbs}
-C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
-\end{equation}
-where \(\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))\).
-The distortions are corrected for the tangent plane of the embedding points.
-Since we aim to estimate the density of the manifold for the embedding of sample points, \(\pmb{p}\) is equal to \(\pmb{y}_i\), for \(i=1,\dots,N\).
-Therefore, we have
-\begin{equation}
-\label{eq:vbsyi}
-C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
-\end{equation}
-
 Now we have all the components needed to estimate the density of manifold learning embedding in Equation \eqref{eq:denestimator} except for the Riemannian metric.
@@
 We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means
-\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\).
+\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\).
 The mixture proportions
@@
 \end{equation}
-
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are \emph{true outliers} to be compared with outliers from the
-kernel density estimates.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_twinpeaks_dc}
-
-}
-
-\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.
-The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
-The shapes indicate the four kernels.}\label{fig:metadensity}
-\end{figure}
 
@@
 
+{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_twinpeaks_dc}
+
+}
+
+\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
+The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
+The shapes indicate the four kernels.}\label{fig:metadensity}
+\end{figure}
+
+We use the multivariate KDE with a fixed plug-in bandwidth in \autoref{mvkde} as the optimal density estimates of the manifold to test the proposed DC-KDE.
+To compare both density estimation methods, it is essential to know the \emph{true density} of the 2-D manifold.
+The Gaussian mixture distribution is known from \eqref{eq:gmm}, however, it is not the density of the manifold because of the mapping function in \autoref{eq:twinpeak} distorts the distribution and the manifold of the density is no longer the Gaussian mixture density.
+Therefore, we apply DC-KDE to the 3-D twin peaks data and set the embedding as the 2-D meta data (true manifold) to first calculate the true density of the manifold.
+We use this technique to obtain the true density of the manifold throughout the simulation process.
+
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
+We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
+Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
+These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
+
+\begin{figure}
+
 {\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_isomap_r0_5_rank}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor}
 
@@
 \caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
 DC-KDE shows a strong linear positive relationship.}\label{fig:tpisomapden}

### 2022-09-26 paper/paper.tex
changes to enable  merge

--- 2022-09-08 paper/paper.tex
@@
 \textbf{Anastasios Panagiotelis}\newline
-The University of Sydney
+University of Sydney
 \newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
@@
 It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of \(\mathbb{R}^d\).
-Applications of kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on.
+Applications of kernel density estimation \autocites[KDE;][]{Parzen1962-gt}{Chen2017-dw} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on.
 However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than \(\mathbb{R}^d\), kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted.
@@
 Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][]{Heidenreich2013-bl}[ for details]{Scott2015-vl}.
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef,Cao1994-st,Jones1996-cb}[ for reviews]{Wand1994-xu}.
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef}{Cao1994-st}{Jones1996-cb}[ for reviews]{Wand1994-xu}.
 The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
@@
 However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-The bandwidth is varied depending on either the location of the sample points {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]} or that of the estimated points {[}balloon estimator; \textcite{Terrell1992-ut}{]}.
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} {[}VKDE; Section 6.6 of \textcite{Scott2015-vl}{]} unless otherwise stated.
+The bandwidth is varied depending on either the location of the sample points \autocite[sample smoothing estimator;][]{Terrell1992-ut} or that of the estimated points \autocite[balloon estimator;][]{Terrell1992-ut}.
+In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} \autocite[VKDE; Section 6.6 of][]{Scott2015-vl} unless otherwise stated.
 However, these kernel density estimators are based on random samples in the Euclidean space.
@@
 
-\hypertarget{s1simulation}{%
-\section{\texorpdfstring{Appendix: Simulation of \(S^1\) for manifold density estimation}{Appendix: Simulation of S\^{}1 for manifold density estimation}}\label{s1simulation}}
-
-To start with, we demonstrate the use of the proposed DC-KDE on a \(1\)-dimensional example.
-For a \(S^1\) curve centered at \(O\) with radius \(1\), the coordinated of points on the curve could be expressed in polar coordinates \((1, \alpha)\) or Cartesian coordinates \((x, \sqrt{1-x^2})\).
-
-The first simulation starts with a \(1\)-dimensional example.
-
-Manifold: \(S^1\) with center 0 and radius \(r=1\)
-
-Polar coordinates: \((r, \theta)\)
-
-Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).
-
-Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)
-
-Density of \(\theta\): \(f(\theta) = 2 / \pi\)
-
-Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)
-
-The CDF of X is given by
-\[
-P(X \leq x) = P(\cos(\theta) \leq x) \\
-= P(\theta \geq \arccos(x)) \\
-= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-= 1 - \frac{2}{\pi} \arccos(x).
-\]
-Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
-f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-= \frac{2}{\pi \sqrt{1 - x^2}}.
-\]
-
-Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
-estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
-and compare it with its true density \(f(x)\).
-We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
-\]
-where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
-
-Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\).
-Therefore, the geodesic distance is
-\[
-d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-\]
-As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).
-
-{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}
-
-The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
-\[
-g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-\]
-is smooth in a local neighborhood \(U\) of \(M\).
-
-Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
-\[
-\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
-\]
-Therefore, the estimator for \(p \in S^1\) is given by
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-\]
-where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.
-
-\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
-\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}
-
-\hypertarget{example-of-s2-in-henry2009-om.}{%
-\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}
-
-Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\).
-Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\).
-Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
-\[
-\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
-\]
-if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
-Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
-Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
-\[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
-\]
-\[
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
-\]
-The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
-Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
-Therefore, we obtain that the volume density on the sphere is
-\[
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
-= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
-\]
-for \(q \neq p, -p\), and \(\theta_p(p) = 1\).
-
-\hypertarget{s1}{%
-\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}
-
-Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
-\[
-\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
-\]
-and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
-\[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
-\]
-So the riemannian metric is \(g_q = \cos(r)\) and further,
-\[
-\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
-\]
-where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).
-
 \hypertarget{swissrollappe}{%

### 2022-09-26 paper/paper.Rmd

@@
 
-As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|\pmb{p}-\pmb{q}\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality.
+<!-- As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|\pmb{p}-\pmb{q}\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality.
 By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric.
@@
 The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension.
-We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and the estimated point $\pmb{p}$ so that for $\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}$, we have
-\begin{equation}
-\label{eq:vbs}
-C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
-\end{equation}
-where $\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))$.
+We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and the estimated point $\pmb{p}$ so that for $\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}$, we have -->
+<!-- \begin{equation} -->
+<!-- \label{eq:vbs} -->
+<!-- C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i), -->
+<!-- \end{equation} -->
+<!-- where $\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))$.
 The distortions are corrected for the tangent plane of the embedding points.
 Since we aim to estimate the density of the manifold for the embedding of sample points, $\pmb{p}$ is equal to $\pmb{y}_i$, for $i=1,\dots,N$.
-Therefore, we have
-\begin{equation}
-\label{eq:vbsyi}
-C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
-\end{equation}
+Therefore, we have -->
+<!-- \begin{equation} -->
+<!-- \label{eq:vbsyi} -->
+<!-- C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i). -->
+<!-- \end{equation} -->
 
@@
 We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with different means
-$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$.
+$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4$.
 The mixture proportions
@@
 \end{equation}
-
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are *true outliers* to be compared with outliers from the
-kernel density estimates.
-
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.02, 0.02)$.
-The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
-The shapes indicate the four kernels.", echo = FALSE}
-knitr::include_graphics("figures/truedensity_twinpeaks_dc.png")
-```
 
@@
 \end{equation}
+
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
+The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
+The shapes indicate the four kernels.", echo = FALSE}
+knitr::include_graphics("figures/truedensity_twinpeaks_dc.png")
+```
+
+We use the multivariate KDE with a fixed plug-in bandwidth in \autoref{mvkde} as the optimal density estimates of the manifold to test the proposed DC-KDE.
+To compare both density estimation methods, it is essential to know the *true density* of the 2-D manifold.
+The Gaussian mixture distribution is known from \@ref(eq:gmm), however, it is not the density of the manifold because of the mapping function in \autoref{eq:twinpeak} distorts the distribution and the manifold of the density is no longer the Gaussian mixture density.
+Therefore, we apply DC-KDE to the 3-D twin peaks data and set the embedding as the 2-D meta data (true manifold) to first calculate the true density of the manifold.
+We use this technique to obtain the true density of the manifold throughout the simulation process.
+
+\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities.
+We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
+Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
+These are *true outliers* to be compared with outliers from the kernel density estimates.
 
@@
 ```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
 DC-KDE shows a strong linear positive relationship.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
 ```

### 2022-09-26 paper/paper.tex

--- 2022-09-26 paper/paper.tex
@@
 For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][]{Heidenreich2013-bl}[ for details]{Scott2015-vl}.
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef,Cao1994-st,Jones1996-cb}[ for reviews]{Wand1994-xu}.
+Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
+For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
 The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.

### 2022-09-27 paper/paper.Rmd
started rewrite of methodology section

@@
 
+# Kernel density estimate on Riemannian manifolds - Tas {#kdermtas}
+
+In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of @Perrault-Joncas2013-pq.
+We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator @Pelletier2005-vu, where the data lie on some known manifold.
+Finally we describing how our own proposed method generalises the estimator of @Pelletier2005-vu to the case where the underlying manifold is unknown.
+Denoting $\lambda$ as the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
+\begin{align}
+\label{eq:kernelcondition}
+& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=1;
+(ii) \int_{\mathbb{R}^{d}} \pmb{x} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=0;
+(iii) \int_{\mathbb{R}^{d}}\|\pmb{x}\|^{2} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})<\infty; \\
+& (iv) \supp K=[0; 1];
+(v) \sup K(||\pmb{x}||)=K(0).
+\end{align}
+Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation.
+For instance condition (iv) requires the support of the kernel to be bounded.
+The reasons for this will become clearer when we discuss the manifold setting in more detail.
+Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise.
+In our empirical section more general kernel functions can be, and are, employed.
+
+Assume the data $\pmb{x}_i\in\mathbb{R}^d$ for $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{x}$ is given by
+
+\begin{equation}
+\label{eq:vkde}
+\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{x}-\pmb{x}_i||}{r}\right).
+\end{equation}
+
+where $||.||$ denotes the usual Euclidean norm.
+The intuition of this estimator is very clear for a uniform kernel.
+The density at a point $\pmb{x}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{x}$, times a term that ensures the density integrates to 1.
+In general, the bandwidth matrix need not be proportional to the identity matrix.
+However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{x}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
+For more on kernel density estimation in the Euclidean case see @Scott2015-vl and references therein.
+
+To extend this to the case where data are on a manifold $\pmb{p}_i\in\mathcal{M}$, for $i=1,\dots,N$, @Pelletier2005-vu propose the following estimator,
+
+\begin{equation}
+\label{eq:denriem}
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
+\end{equation}
+
+where $d_g(\pmb{p}, \pmb{p}_i)$ denotes the geodesic distance between two points on the manifold $\pmb{p}$ and $\pmb{p}_i$ and $\theta_{\pmb{p}_i}(\pmb{p})$ is known as the volume density function.
+The intuition behind the term $K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)$ is relatively clear.
+For example, for a uniform kernel the estimator at point $\pmb{p}$ will still be depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
+However in this case, the distance metric used is the geodesic distance on the manifold rather than Euclidean or Mahalanobis distance.
+An additional technical assumption is that $r$ is less than the injectivity radius of the manifold.
+A technical definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix.
+For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\pmb{p}$ is so large that some points `fall inside' the ball more than once.
+This could occur with a ball on the surface of a sphere with radius greater than half the circumference of a great circle on that sphere.
+This phenomenon also explains why the kernel function must be bounded fore density estimation on manifolds.
+
+The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore before providing formal details we will briefly discuss the intuition behind including this term.
+We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
+However what should also be taken into account the volume of a ball of radius $r$ around $\pmb{p}$.
+In Euclidean space, this is not relevant since a radius $r$ ball will always have the same volume regardless of its center.
+The same does not hold for manifolds and the volume density function corrects for this while ensuring that the density estimate integrates to one.
+
+Technical details discussion to come...
+
 # Kernel density estimate on Riemannian manifold embedding {#kderm}

### 2022-09-27 paper/paper.tex
started rewrite of methodology section

@@
 
+\hypertarget{kdermtas}{%
+\section{Kernel density estimate on Riemannian manifolds - Tas}\label{kdermtas}}
+
+In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}.
+We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator \textcite{Pelletier2005-vu}, where the data lie on some known manifold.
+Finally we describing how our own proposed method generalises the estimator of \textcite{Pelletier2005-vu} to the case where the underlying manifold is unknown.
+Denoting \(\lambda\) as the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
+\begin{align}
+\label{eq:kernelcondition}
+& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=1;
+(ii) \int_{\mathbb{R}^{d}} \pmb{x} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=0;
+(iii) \int_{\mathbb{R}^{d}}\|\pmb{x}\|^{2} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})<\infty; \\
+& (iv) \supp K=[0; 1];
+(v) \sup K(||\pmb{x}||)=K(0).
+\end{align}
+Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation.
+For instance condition (iv) requires the support of the kernel to be bounded.
+The reasons for this will become clearer when we discuss the manifold setting in more detail.
+Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise.
+In our empirical section more general kernel functions can be, and are, employed.
+
+Assume the data \(\pmb{x}_i\in\mathbb{R}^d\) for \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{x}\) is given by
+
+\begin{equation}
+\label{eq:vkde}
+\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{x}-\pmb{x}_i||}{r}\right).
+\end{equation}
+
+where \(||.||\) denotes the usual Euclidean norm.
+The intuition of this estimator is very clear for a uniform kernel.
+The density at a point \(\pmb{x}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{x}\), times a term that ensures the density integrates to 1.
+In general, the bandwidth matrix need not be proportional to the identity matrix.
+However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{x}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
+For more on kernel density estimation in the Euclidean case see \textcite{Scott2015-vl} and references therein.
+
+To extend this to the case where data are on a manifold \(\pmb{p}_i\in\mathcal{M}\), for \(i=1,\dots,N\), \textcite{Pelletier2005-vu} propose the following estimator,
+
+\begin{equation}
+\label{eq:denriem}
+\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
+\end{equation}
+
+where \(d_g(\pmb{p}, \pmb{p}_i)\) denotes the geodesic distance between two points on the manifold \(\pmb{p}\) and \(\pmb{p}_i\) and \(\theta_{\pmb{p}_i}(\pmb{p})\) is known as the volume density function.
+The intuition behind the term \(K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)\) is relatively clear.
+For example, for a uniform kernel the estimator at point \(\pmb{p}\) will still be depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
+However in this case, the distance metric used is the geodesic distance on the manifold rather than Euclidean or Mahalanobis distance.
+An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold.
+A technical definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
+For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once.
+This could occur with a ball on the surface of a sphere with radius greater than half the circumference of a great circle on that sphere.
+This phenomenon also explains why the kernel function must be bounded fore density estimation on manifolds.
+
+The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore before providing formal details we will briefly discuss the intuition behind including this term.
+We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
+However what should also be taken into account the volume of a ball of radius \(r\) around \(\pmb{p}\).
+In Euclidean space, this is not relevant since a radius \(r\) ball will always have the same volume regardless of its center.
+The same does not hold for manifolds and the volume density function corrects for this while ensuring that the density estimate integrates to one.
+
+Technical details discussion to come\ldots{}
+
 \hypertarget{kderm}{%

### 2022-09-29 paper/paper.Rmd
finished discussion of Pelletier

@@
 In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of @Perrault-Joncas2013-pq.
-We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator @Pelletier2005-vu, where the data lie on some known manifold.
-Finally we describing how our own proposed method generalises the estimator of @Pelletier2005-vu to the case where the underlying manifold is unknown.
+We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+Finally, we describe how our own proposed method generalises the estimator of @Pelletier2005-vu to the case where the underlying manifold is unknown.
 Denoting $\lambda$ as the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
@@
 
-where $||.||$ denotes the usual Euclidean norm.
-The intuition of this estimator is very clear for a uniform kernel.
+The intuition behind this estimator is very clear for a uniform kernel.
 The density at a point $\pmb{x}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{x}$, times a term that ensures the density integrates to 1.
@@
 For example, for a uniform kernel the estimator at point $\pmb{p}$ will still be depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
-However in this case, the distance metric used is the geodesic distance on the manifold rather than Euclidean or Mahalanobis distance.
+However in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
 An additional technical assumption is that $r$ is less than the injectivity radius of the manifold.
-A technical definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix.
+A definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix.
 For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\pmb{p}$ is so large that some points `fall inside' the ball more than once.
-This could occur with a ball on the surface of a sphere with radius greater than half the circumference of a great circle on that sphere.
-This phenomenon also explains why the kernel function must be bounded fore density estimation on manifolds.
+For example on a sphere, a ball with radius greater than half the circumference of a great circle will wrap back around the sphere.
+This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
 
@@
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
-However what should also be taken into account the volume of a ball of radius $r$ around $\pmb{p}$.
-In Euclidean space, this is not relevant since a radius $r$ ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and the volume density function corrects for this while ensuring that the density estimate integrates to one.
-
-Technical details discussion to come...
+However the volume of a ball of radius $r$ around $\pmb{p}$ must also be taken into account.
+In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center.
+The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.
+
+More formally, the volume density function can be explained as follows.
+Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space $\pmb{v}\in T_{\pmb{p}}M$ to points the manifold $\pmb{q}\in M$ .
+Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and along this geodesic at uniform speed $||\pmb{v}||$ takes place in one unit of time.
+Now, consider a chart $\varphi$ mapping points in the neighbourhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
+The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system.
+For more on the volume density function see @Le_Brigant2019-lj.
+
+## Distortion corrected KDE
+
+To come...
 

### 2022-09-29 R/sources/plot_embedding_color_levels.R
Function to plot categorized density colors for embedding input

@@ -1,0 +1,50 @@
+# Categorize density to `length(prob)` levels prob and plot the embedding colored by the density levels
+plot_embedding_color_levels <- function(x = trueden, N, prob = c(1, 50, 95, 99), noutliers, fn = NULL) {
+  if(max(prob) > 50){ # Assume prob is coverage percentage
+    alpha <- (100 - prob) / 100
+  } else {# prob is tail probability (for backward compatibility)
+    alpha <- prob
+  }
+  alpha <- sort(alpha) # 0.99 0.50 0.05 0.01
+  falpha <- quantile(x, alpha) # increasing
+  region <- numeric(N) + 100 # rep(100, N)
+  for (i in seq_along(prob)) region[x > falpha[i]] <- 100 - prob[i]
+  if (is.null(noutliers)) noutliers <- sum(region > max(prob))
+  noutliers <- min(noutliers, N)
+  
+  # Construct region factor
+  prob <- sort(unique(region[region < 100]), decreasing = TRUE)
+  prob <- c(prob, 100)
+  Region <- factor(region,
+                   levels = prob,
+                   labels = c(paste(head(prob, -1)), ">99") # "99"  "50"  "5"   "1"   ">99"
+  )
+  # Sort data so the larger regions go first (other than outliers)
+  # rr <- region
+  # rr[region == 100] <- 0
+  # ord <- order(rr, decreasing = TRUE)
+  ord <- order(x, decreasing = T)  # outliers at the end
+  # fn_dec <- fn[ord,] # data is reordered by ord
+  outliers <- tail(ord, noutliers)
+  
+  # if (noutliers > 0) {
+  #   outlier_rank <- order(x[ord])  # order density as well to match the new data
+  #   outliers <- outlier_rank[seq_len(noutliers)] # take top noutliers labels for annotation
+  # }
+  
+  p <- fn %>% 
+    as_tibble() %>% 
+    ggplot(aes(x = E1, y = E2)) + 
+    ggplot2::geom_point(ggplot2::aes_string(col = "Region")) + 
+    ggplot2::scale_colour_manual(
+      name = "HDRs",
+      breaks = c(paste(head(sort(prob), -1)), ">99"),
+      values = c(RColorBrewer::brewer.pal(length(prob), "YlOrRd")[-1], "#000000")) + 
+    ggplot2::annotate("text",
+                      x = fn[outliers, 1] + diff(range(fn[outliers, 1])) / 50, 
[…]

### 2022-09-30 paper/paper.Rmd
Add new figures

--- 2022-09-27 paper/paper.Rmd
@@
 
-```{r tpoutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of four manifold learning embeddings of the twin peak data.
-Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.", echo = FALSE}
+```{r tpoutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
+The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel).
+DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
@@
 
-```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-DC-KDE shows a strong linear positive relationship.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
+```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+# knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_rankdistcol.png")
 ```
@@
 The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.", echo = FALSE}
-knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
+# knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
+knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity.png")
 ```

### 2022-09-30 paper/paper.Rmd
update twinpeaks

@@
 # knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_rankdistcol.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob.png")
 ```

### 2022-09-30 R/TwinPeaks.R

@@ -17,7 +17,7 @@ N <- 2000
 p <- 2
 
 ###--------------------------------------------------------
-## Start from here! 
+## 1. Generate 2-D manifold and 3-D twin peaks
 ###--------------------------------------------------------
 set.seed(1234)
 mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3] # only run [1] and [3]
@@ -49,7 +49,9 @@ plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = den_twinpeaks, # co
 
 
 
-
+## -------------------------------------------------------------------------------
+# 2. TRUE density of 2-D manifold
+## -------------------------------------------------------------------------------
 train <- swissroll
 # Parameters fixed
 x <- train
@@ -67,7 +69,6 @@ radius <- 8 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note th
 
 gridsize <- 20
 noutliers <- 20
-
 metric_meta <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
                           annmethod = annmethod, distance = distance, treetype = treetype,
                           searchtype = searchtype
@@ -77,7 +78,7 @@ metric_meta <- metricML(x, fn = y, s = s, k = k, radius = radius, method = metho
 fn <- metric_meta$embedding
 Rn <- metric_meta$rmetric
 
-# ## ----ggellipse, include=FALSE, eval=FALSE---------------------------------------
+# ## ----ggellipse, include=FALSE, eval=FALSE-----------------------------------
 # plot_embedding(metric_meta) +
 #   labs(x = "ISO1", y = "ISO2")
 # plot_ellipse(metric_meta, add = F, ell.no = 500, ell.size = .1,
@@ -85,9 +86,6 @@ Rn <- metric_meta$rmetric
 
 
 # Run once with y <- sr$metadata
-## -------------------------------------------------------------------------------
-# TRUE density of manifold
-## -------------------------------------------------------------------------------
 # Transformed from sr$den
 den_2dmanifold <- sr$den * (apply(Rn, 3, det)) ^ (.5)
 summary(den_2dmanifold)
[…]

### 2022-09-30 paper/paper.Rmd

--- 2022-09-29 paper/paper.Rmd
@@
 
-```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-DC-KDE shows a strong linear positive relationship.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
+```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+# knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob.png")
 ```

### 2022-09-30 paper/paper.tex

@@
 In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}.
-We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator \textcite{Pelletier2005-vu}, where the data lie on some known manifold.
-Finally we describing how our own proposed method generalises the estimator of \textcite{Pelletier2005-vu} to the case where the underlying manifold is unknown.
+We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+Finally, we describe how our own proposed method generalises the estimator of \textcite{Pelletier2005-vu} to the case where the underlying manifold is unknown.
 Denoting \(\lambda\) as the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
@@
 
-where \(||.||\) denotes the usual Euclidean norm.
-The intuition of this estimator is very clear for a uniform kernel.
+The intuition behind this estimator is very clear for a uniform kernel.
 The density at a point \(\pmb{x}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{x}\), times a term that ensures the density integrates to 1.
@@
 For example, for a uniform kernel the estimator at point \(\pmb{p}\) will still be depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
-However in this case, the distance metric used is the geodesic distance on the manifold rather than Euclidean or Mahalanobis distance.
+However in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
 An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold.
-A technical definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
+A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
 For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once.
-This could occur with a ball on the surface of a sphere with radius greater than half the circumference of a great circle on that sphere.
-This phenomenon also explains why the kernel function must be bounded fore density estimation on manifolds.
+For example on a sphere, a ball with radius greater than half the circumference of a great circle will wrap back around the sphere.
+This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
 
@@
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
-However what should also be taken into account the volume of a ball of radius \(r\) around \(\pmb{p}\).
-In Euclidean space, this is not relevant since a radius \(r\) ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and the volume density function corrects for this while ensuring that the density estimate integrates to one.
-
-Technical details discussion to come\ldots{}
+However the volume of a ball of radius \(r\) around \(\pmb{p}\) must also be taken into account.
+In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center.
+The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.
+
+More formally, the volume density function can be explained as follows.
+Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space \(\pmb{v}\in T_{\pmb{p}}M\) to points the manifold \(\pmb{q}\in M\) .
+Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and along this geodesic at uniform speed \(||\pmb{v}||\) takes place in one unit of time.
+Now, consider a chart \(\varphi\) mapping points in the neighbourhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
+The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system.
+For more on the volume density function see \textcite{Le_Brigant2019-lj}.
+
+\hypertarget{distortion-corrected-kde}{%
+\subsection{Distortion corrected KDE}\label{distortion-corrected-kde}}
+
+To come\ldots{}
 

### 2022-09-30 R/Simulation4d_03plots.R
4d simulation hpc scripts

@@ -2,16 +2,21 @@
 ## Gather and plot the results from 4 senarios in HPC results
 ###--------------------------------------------------------
 # first load all .rda files
-load(paste0("data/compareden_4d_N", N, "_", method, "_radius", radius, "_r", format(r, decimal.mark = "_"), "_annIsomap.rda"))
-fxy <- den
+library(viridis)
+N <- 10000
+radius <- 10
+r <- 1
+# method <- c("isomap", "lle", "le", "tsne", "umap")[1]
+load("~/git/kderm/data/simdata_3d_N2000_01_trueden_k100.rda")
+fxy <- trueden
 label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
-dist2center <- sqrt(r^2 - X_new[,5]^2)
+dist2center <- sqrt(radius^2 - semisphere[,5]^2)
 methods <- c("isomap", "lle", "le", "umap")
 ## scatterplot to compare f_xy for ISOMAP
-f <- tibble(fxy = fxy, fxy_vkde = fisomap$estimate, fxy_hdr = fixden_isomap$estimate)
-f1 <- tibble(fxy = fxy, fxy_vkde = flle$estimate, fxy_hdr = fixden_lle$estimate)
-f2 <- tibble(fxy = fxy, fxy_vkde = fle$estimate, fxy_hdr = fixden_le$estimate)
-f3 <- tibble(fxy = fxy, fxy_vkde = fumap$estimate, fxy_hdr = fixden_umap$estimate)
+f <- tibble(fxy = fxy, fxy_vkde = fisomap$estimate, fxy_hdr = fixden_isomap$estimate) #%>% rank()
+f1 <- tibble(fxy = fxy, fxy_vkde = flle$estimate, fxy_hdr = fixden_lle$estimate) #%>% rank()
+f2 <- tibble(fxy = fxy, fxy_vkde = fle$estimate, fxy_hdr = fixden_le$estimate) #%>% rank()
+f3 <- tibble(fxy = fxy, fxy_vkde = fumap$estimate, fxy_hdr = fixden_umap$estimate) #%>% rank()
 
 summary(f)
 cor(f$fxy_vkde, f$fxy)
@@ -79,12 +84,12 @@ result <- (p/p1/p2/p3) + plot_layout(guides = 'collect') &
 gt <- patchwork::patchworkGrob(result)
 gt <- gridExtra::grid.arrange(gt, left = "Estimated density", bottom = "True density")
 gt
[…]

### 2022-09-30 R/multidimension.R

@@ -154,7 +154,7 @@ all.equal(train %*% t(train), X_new %*% t(X_new))
 # Parameters fixed
 x <- train
 N <- nrow(x)
-s <- 5 # embedded in 5-D
+s <- 4 # embedded in 5-D
 k <- 20
 method <- "Isomap"
 annmethod <- "kdtree"
@@ -371,7 +371,13 @@ cor(den, fixden_umap$estimate)
 
 
 
-## ----compareDensity---------------------------------------------------------------------------
+
+
+
+###--------------------------------------------------------
+## compareDensity
+###--------------------------------------------------------
+# first load all .rda files
 fxy <- den
 label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
 dist2center <- sqrt(r^2 - X_new[,5]^2)

### 2022-09-30 R/sim4d_isomap.R

@@ -12,7 +12,8 @@ Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1234)
 r <- 1
 # # r <- as.numeric(commandArgs()[[6]])
-load("data/simdata_100d_4dmanifold_N10000_trueden_k200.rda")
+k <- 500
+load(paste0("data/simdata_100d_4dmanifold_N10000_trueden_k", k, ".rda"))
 
 paste("Start at:", Sys.time())
 # ----parameters----------------------------------------------------------------
@@ -20,7 +21,7 @@ x <- train
 y <- NULL # without pre-computed embedding
 N <- nrow(x)
 s <- 4 # embedded in 4-D
-k <- N / 50
+# k <- N / 50
 method <- "annIsomap"
 annmethod <- "kdtree"
 distance <- "euclidean"
@@ -33,10 +34,14 @@ opt.method <- c("AMISE", "MEAN", "SCALED")[3] # 3 ONLY FOR NOW, no scaling for R
 riem.scale <- 1 # tune parameter
 
 ## ----isomap-------------------------------------------------------------------
-metric_isomap <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
-                          annmethod = annmethod, distance = distance, treetype = treetype,
-                          searchtype = searchtype
-)
+if(file.exists("data/metric_isomap_4d_N10000_radius10.rda")){
+  load("data/metric_isomap_4d_N10000_radius10.rda")
+} else {
+  metric_isomap <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+                            annmethod = annmethod, distance = distance, treetype = treetype,
+                            searchtype = searchtype
+  )
+}
 fn <- metric_isomap$embedding
 Rn <- metric_isomap$rmetric
 adj_matrix <- metric_isomap$adj_matrix
@@ -92,11 +97,11 @@ p <- f %>%
   labs(x = "True density rank", y = "Estimated density rank") +
   scale_y_continuous(n.breaks = 6)
 # p
-ggsave(paste0("figures/compareden_5d_N", N, "_", method, "_r", format(r, decimal.mark = "_"), ".png"), p, width = 10, height = 6, dpi = 300)
[…]

### 2022-09-30 R/sim4d_le.R

@@ -12,7 +12,8 @@ Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1234)
 r <- 1
 # # r <- as.numeric(commandArgs()[[6]])
-load("data/simdata_100d_4dmanifold_N10000_trueden_k200.rda")
+k <- 500
+load(paste0("data/simdata_100d_4dmanifold_N10000_trueden_k", k, ".rda"))
 
 paste("Start at:", Sys.time())
 # ----parameters----------------------------------------------------------------
@@ -20,7 +21,7 @@ x <- train
 y <- NULL # without pre-computed embedding
 N <- nrow(x)
 s <- 4 # embedded in 4-D
-k <- N / 50
+# k <- N / 50
 method <- "annLaplacianEigenmaps"
 annmethod <- "kdtree"
 distance <- "euclidean"
@@ -33,10 +34,14 @@ opt.method <- c("AMISE", "MEAN", "SCALED")[3] # 3 ONLY FOR NOW, no scaling for R
 riem.scale <- 1 # tune parameter
 
 ## ----le-------------------------------------------------------------------
-metric_le <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
-                          annmethod = annmethod, distance = distance, treetype = treetype,
-                          searchtype = searchtype
-)
+if(file.exists("data/metric_le_4d_N10000_radius10.rda")){
+  load("data/metric_le_4d_N10000_radius10.rda")
+} else {
+  metric_le <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+                            annmethod = annmethod, distance = distance, treetype = treetype,
+                            searchtype = searchtype
+  )
+}
 fn <- metric_le$embedding
 Rn <- metric_le$rmetric
 adj_matrix <- metric_le$adj_matrix
@@ -92,11 +97,11 @@ p <- f %>%
   labs(x = "True density rank", y = "Estimated density rank") +
   scale_y_continuous(n.breaks = 6)
 # p
-ggsave(paste0("figures/compareden_5d_N", N, "_", method, "_r", format(r, decimal.mark = "_"), ".png"), p, width = 10, height = 6, dpi = 300)
[…]

### 2022-09-30 R/sim4d_umap.R

@@ -12,7 +12,8 @@ Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1234)
 r <- 1
 # # r <- as.numeric(commandArgs()[[6]])
-load("data/simdata_100d_4dmanifold_N10000_trueden_k200.rda")
+k <- 500
+load(paste0("data/simdata_100d_4dmanifold_N10000_trueden_k", k, ".rda"))
 
 paste("Start at:", Sys.time())
 # ----parameters----------------------------------------------------------------
@@ -20,7 +21,7 @@ x <- train
 y <- NULL # without pre-computed embedding
 N <- nrow(x)
 s <- 4 # embedded in 4-D
-k <- N / 50
+# k <- N / 50
 method <- "annUMAP"
 annmethod <- "kdtree"
 distance <- "euclidean"
@@ -33,10 +34,14 @@ opt.method <- c("AMISE", "MEAN", "SCALED")[3] # 3 ONLY FOR NOW, no scaling for R
 riem.scale <- 1 # tune parameter
 
 ## ----umap-------------------------------------------------------------------
-metric_umap <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
-                          annmethod = annmethod, distance = distance, treetype = treetype,
-                          searchtype = searchtype
-)
+if(file.exists("data/metric_umap_4d_N10000_radius10.rda")){
+  load("data/metric_umap_4d_N10000_radius10.rda")
+} else {
+  metric_umap <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+                            annmethod = annmethod, distance = distance, treetype = treetype,
+                            searchtype = searchtype
+  )
+}
 fn <- metric_umap$embedding
 Rn <- metric_umap$rmetric
 adj_matrix <- metric_umap$adj_matrix
@@ -92,11 +97,11 @@ p <- f %>%
   labs(x = "True density rank", y = "Estimated density rank") +
   scale_y_continuous(n.breaks = 6)
 # p
-ggsave(paste0("figures/compareden_5d_N", N, "_", method, "_r", format(r, decimal.mark = "_"), ".png"), p, width = 10, height = 6, dpi = 300)
[…]

### 2022-09-30 R/Simulation4d_02dckde.R
4d simulation hpc scripts

@@ -5,6 +5,7 @@
 # `radius=10`
 # Tune `r`, the bandwidth parameter, for DC-KDE
 # One ggplot for rank plots, and two .rda files for ML and density estimates
+# This script is modified for HPC jobs for four ML methods
 rm(list = ls())
 library(tidyverse)
 library(dimRed)
@@ -15,7 +16,8 @@ library(patchwork)
 Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1234)
 r <- 1
-load("data/simdata_100d_4dmanifold_N10000_trueden_k200.rda")
+k <- 500
+load(paste0("data/simdata_100d_4dmanifold_N10000_trueden_k", k, ".rda"))
 
 paste("Start at:", Sys.time())
 # ----parameters----------------------------------------------------------------
@@ -36,10 +38,14 @@ opt.method <- c("AMISE", "MEAN", "SCALED")[3] # 3 ONLY FOR NOW, no scaling for R
 riem.scale <- 1 # tune parameter
 
 ## ----isomap-------------------------------------------------------------------
-metric_isomap <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
-                          annmethod = annmethod, distance = distance, treetype = treetype,
-                          searchtype = searchtype
-)
+if(file.exists("data/metric_isomap_4d_N10000_radius10.rda")){
+  load("data/metric_isomap_4d_N10000_radius10.rda")
+} else {
+  metric_isomap <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+                            annmethod = annmethod, distance = distance, treetype = treetype,
+                            searchtype = searchtype
+  )
+}
 fn <- metric_isomap$embedding
 Rn <- metric_isomap$rmetric
 adj_matrix <- metric_isomap$adj_matrix
@@ -100,6 +106,6 @@ ggsave(paste0("figures/compareden_5d_N", N, "_", method, "_r", format(r, decimal
 
 save(method, fixden_isomap, fisomap, train, trueden, cors,
      file = paste0("data/compareden_4d_N", N, "_", method, "_radius", radius, "_r", format(r, decimal.mark = "_"), "_annIsomap.rda"))
-save(metric_isomap, file = paste0("data/metric_isomap_4d_N10000_radius", radius, ".rda"))
[…]

### 2022-09-30 R/sim4d_lle.R

@@ -11,7 +11,8 @@ library(patchwork)
 Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1234)
 r <- 1
-load("data/simdata_100d_4dmanifold_N10000_trueden_k200.rda")
+k <- 500
+load(paste0("data/simdata_100d_4dmanifold_N10000_trueden_k", k, ".rda"))
 
 paste("Start at:", Sys.time())
 # ----parameters----------------------------------------------------------------
@@ -19,7 +20,7 @@ x <- train
 y <- NULL # without pre-computed embedding
 N <- nrow(x)
 s <- 4 # embedded in 4-D
-k <- N / 50
+# k <- N / 50
 method <- "annLLE"
 annmethod <- "kdtree"
 distance <- "euclidean"
@@ -32,10 +33,14 @@ opt.method <- c("AMISE", "MEAN", "SCALED")[3] # 3 ONLY FOR NOW, no scaling for R
 riem.scale <- 1 # tune parameter
 
 ## ----lle-------------------------------------------------------------------
-metric_lle <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
-                          annmethod = annmethod, distance = distance, treetype = treetype,
-                          searchtype = searchtype
-)
+if(file.exists("data/metric_lle_4d_N10000_radius10.rda")){
+  load("data/metric_lle_4d_N10000_radius10.rda")
+} else {
+  metric_lle <- metricML(x, fn = y, s = s, k = k, radius = radius, method = method, invert.h = TRUE, eps = 0,
+                            annmethod = annmethod, distance = distance, treetype = treetype,
+                            searchtype = searchtype
+  )
+}
 fn <- metric_lle$embedding
 Rn <- metric_lle$rmetric
 adj_matrix <- metric_lle$adj_matrix
@@ -91,11 +96,11 @@ p <- f %>%
   labs(x = "True density rank", y = "Estimated density rank") +
   scale_y_continuous(n.breaks = 6)
 # p
-ggsave(paste0("figures/compareden_5d_N", N, "_", method, "_r", format(r, decimal.mark = "_"), ".png"), p, width = 10, height = 6, dpi = 300)
+ggsave(paste0("figures/rankplot_4d_N", N, "_", method, "_radius", radius, "_r", format(r, decimal.mark = "_"), ".png"), p, width = 10, height = 6, dpi = 300)
 
 
[…]

### 2022-10-04 paper/paper.Rmd
Remove colors for rank comparison

@@
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_density_comparison_isomapvs4ml_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
 ```

### 2022-10-07 paper/paper.Rmd
Highlight false points

--- 2022-09-30 paper/paper.Rmd
@@
 # knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_circlefalse.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_nocolprob_circlefalse.png")
 ```
@@
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_density_comparison_isomapvs4ml_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
+# knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
+# knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
+# knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
+# knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
 ```

### 2022-10-07 R/TwinPeaks.R
Highlight false points

@@ -812,69 +812,116 @@ pf_hdr <- frank %>%
 # color the points with HDR, add rectangle and highlight points outside
 levels <- c(sort(100-prob) / 100, 1)
 q_rank <- quantile(1:N, levels)
-apply(frank[1:3], 2, function(x) cut(x, 
-                                     breaks = c(0, quantile(1:N, levels)), 
-                                     labels = c(paste(sort(100 - prob)), ">99"))) %>% head
-
-
-
-frank %>% 
-  mutate(col_trueden = )
-frank1 <- frank %>%
-  mutate(col_vkde = factor(p_isomap$p[["plot_env"]][["region"]],
-                           levels = c(paste(sort(100 - prob)), "100")),
-         col_kde = factor(p_hdr_isomap$p[["plot_env"]][["region"]],
-                          levels = c(paste(sort(100 - prob)), "100")),
-         col_truehdr = factor(p_meta$p[["plot_env"]][["region"]],
-                               levels = c(paste(sort(100 - prob)), "100"))
-         )
+# apply(frank[1:3], 2, function(x) cut(x, 
+#                                      breaks = c(0, quantile(1:N, levels)), 
+#                                      labels = c(paste(sort(100 - prob)), ">99"))) %>% head
+xname <- "isomap"
+frank1 <- frank %>% 
+  select(fxy:fxy_hdr) %>% # isomap
+  # rename(fxy_vkde_isomap = fxy_vkde, 
+  #        fxy_hdr_isomap = fxy_hdr) %>%
+  select(fxy, contains(paste0("_", xname))) %>% 
+  mutate(across(everything(), 
+                function(x) cut(x, 
+                                breaks = c(0, quantile(1:N, levels)), 
+                                labels = c(paste(sort(100 - prob)), ">99")))
+         ) %>% 
+  `colnames<-`(gsub("fxy_", "col_", colnames(.))) %>% 
+  # `colnames<-`(gsub(paste0("_", xname), "", colnames(.))) %>% 
+  bind_cols(frank[,1:3]) %>% 
+  mutate(dckde_false = (col_fxy == col_vkde),
+         kde_false = (col_fxy == col_hdr) )
[…]

### 2022-10-10 R/TwinPeaks.R
2D figures done

@@ -16,9 +16,9 @@ Jmisc::sourceAll(here::here("R/sources"))
 N <- 2000
 p <- 2
 
-###--------------------------------------------------------
+##----Data Generation-------------------------------------
 ## 1. Generate 2-D manifold and 3-D twin peaks
-###--------------------------------------------------------
+##--------------------------------------------------------
 set.seed(1234)
 mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3] # only run [1] and [3]
 sr <- mldata(N = N, meta = "gaussian", mapping = mapping)
@@ -324,7 +324,7 @@ E1 <- fn[,1]; E2 <- fn[,2]
 # opt.method <- c("AMISE", "MEAN", "SCALED")[2]
 # riem.scale <- .1
 flle <- vkde(x = fn, h = h_hdr_meta, vh = Rn, r = r, gridsize = gridsize, eval.points = fn, opt.method = opt.method, riem.scale = riem.scale, adj_matrix = adj_matrix)
-## ---- echo = F------------------------------------------------------------------
+## ----------------------------------------------------------------------
 p_lle <- plot_outlier(x = metric_lle, gridsize = gridsize, prob = prob, noutliers = noutliers, opt.method = opt.method, riem.scale = riem.scale, f = flle, ell.size = 0)
 p_hdr_lle <- hdrscatterplot_new(E1, E2, kde.package = "ks", noutliers = noutliers, levels = prob)
 # p_hdr_lle_p <- p_hdr_lle + 
@@ -373,7 +373,7 @@ E1 <- fn[,1]; E2 <- fn[,2]
 # opt.method <- c("AMISE", "MEAN", "SCALED")[3]
 # riem.scale <- 1
 fle <- vkde(x = fn, h = h_hdr_meta, vh = Rn, r = r, gridsize = gridsize, eval.points = fn, opt.method = opt.method, riem.scale = riem.scale, adj_matrix = adj_matrix)
-## ---- echo = F------------------------------------------------------------------
+## ---------------------------------------------------------------------
[…]

### 2022-10-10 paper/paper.Rmd
Update all twin peaks plots in Rmd, need to be knited

@@
 
-```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
 # knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_circlefalse.png")
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_nocolprob_circlefalse.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
+```
+
+```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of t-SNE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
 ```

### 2022-10-10 R/TwinPeaks.R
Update all twin peaks plots in Rmd, need to be knited

@@ -717,54 +717,128 @@ gt
 
 ## -----------------------------------------------------------------------------
 ## 6.3 Only isomap for the paper (Figure `tpisomapden`)
-
-# Not run
+# RUN this!!!
 ##---------------------------------------------------------------------
-# color the points with distance to y=x (Improvement: rotate 45 degrees for better comparison)
-# log scale of x and y axis, with colors indicating the absolute rank error weighted by the sum of true and estimated ranks
-p_isomap_logrank_distcol <- frank %>% 
-  select(fxy, fxy_vkde, fxy_hdr) %>%
-  # mutate(outlier = case_when((fxy <= noutliers) ~ "True outliers",
-  #                            (fxy_dckde <= noutliers) ~ "DC-KDE outliers",
-  #                            (fxy_kde <= noutliers) ~ "KDE outliers",
-  #                            TRUE ~ "Not outliers")) %>% 
-  pivot_longer(cols = -c(fxy), names_to = "kde", values_to = "densities") %>%
-  mutate(fcol = abs((fxy - densities) / (fxy + densities) ) ) %>% # colors for weighted rank error
-  ggplot(aes(fxy, densities)) + # , col = densities
-  geom_point(aes(col = fcol)) + 
-  facet_grid(~ factor(kde, levels = c("fxy_vkde", "fxy_hdr")),
-             labeller = as_labeller(c(
-               `fxy_vkde` = paste("DC-KDE correlation =", format(cor(frank$fxy, frank$fxy_vkde), digits = 3)),
-               `fxy_hdr` = paste("KDE correlation =", format(cor(frank$fxy, frank$fxy_hdr), digits = 3))))
-  ) + 
-  scale_x_continuous(trans = "log10", limits = c(1, 2000), breaks = c(1, 10, 100, 1000, 2000)) +
-  scale_y_continuous(trans = "log10", limits = c(1, 2000), breaks = c(1, 10, 100, 1000, 2000)) +
-  labs(x = "True density rank", y = "Estimated density rank") + 
-  scale_color_viridis(option = "D", direction = -1) +
-  guides(color = "none")
[…]

### 2022-10-10 R/TwinPeaks.R
Remove unused figures, add different versions for Figure 5&6

@@ -720,12 +720,12 @@ gt
 # RUN this!!!
 ##---------------------------------------------------------------------
 # color the points with HDR, add rectangle and highlight points outside
+xname <- c("isomap", "tsne")[2] # !!! Modify index for different ML methods comparison
 probs <- prob[1:2]
 levels <- c(probs / 100, 1)
 q_rank <- quantile(1:N, levels)
 rec <- log10(q_rank)
-greys <- scales::grey_pal()(10)[c(8,10)]
-xname <- c("isomap", "tsne")[1] # !!! Modify index for different ML methods comparison
+greys <- scales::grey_pal()(10)[c(6, 8, 10)]
 frank1 <- frank %>% 
   # select(fxy:fxy_hdr) %>% # isomap
   rename(fxy_vkde_isomap = fxy_vkde,
@@ -739,7 +739,7 @@ frank1 <- frank1 %>%
                                 labels = c(">99", paste(100 - probs))))
   ) %>% 
   `colnames<-`(gsub("fxy", "col", colnames(.))) %>% 
-  bind_cols(select(frank, fxy, contains(paste0("_", xname)))) %>% 
+  bind_cols(select(frank1, fxy, contains(paste0("_", xname)))) %>% 
   `colnames<-`(gsub(paste0("_", xname), "", colnames(.))) %>% 
   mutate(dckde_false = (col == col_vkde) & (col != last(probs)) & (col_vkde != last(probs)),
          kde_false = (col == col_hdr) & (col != last(probs)) & (col_hdr != last(probs)) ) 
@@ -747,8 +747,10 @@ frank1 <- frank1 %>%
 plogrec_isomap_vkde <-
   frank1 %>%
   ggplot(aes(x = fxy, y = fxy_vkde)) +
-  geom_rect(xmin = 0, xmax = rec[2], ymin = -Inf, ymax = rec[2], fill = greys[2], col = greys[2]) +
-  geom_rect(xmin = 0, xmax = rec[1], ymin = -Inf, ymax = rec[1], fill = greys[1], col = greys[1]) +
+  # geom_rect(xmin = 0, xmax = rec[2], ymin = 0, ymax = rec[2], fill = greys[2], col = greys[2]) +
+  # geom_rect(xmin = rec[2], xmax = rec[3], ymin = rec[2], ymax = rec[3], fill = greys[3], col = greys[3]) + # third blocks
[…]

### 2022-10-10 paper/paper.Rmd
Remove unused figures, add different versions for Figure 5&6

@@
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
-# knitr::include_graphics("figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor.png")
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
+# knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
+# knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks_crossfalse.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
 ```
@@
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
 ```

### 2022-10-10 paper/paper.Rmd
100D simulation rank plot

@@
 \label{eq:volumedensityratio}
-\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
+\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }  .
 \end{equation}
@@
 
-```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
-The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.", echo = FALSE}
-# knitr::include_graphics("figures/fived_density_comparison_4ml_riem20.png")
-knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity.png")
+```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE.
+Four manifold learning methods are used rowwise.
+The point shapes indicates whether they are the true outleirs, and the grey shading highlights the top 1\\% rank region.
+The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
+# knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity.png")
+knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec.png")
 ```
@@
 ```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_fived_4ml_riem20.rda")
+load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
 col.max <- apply(cors, 2, max)
@@
 $$
-P(X \leq x) = P(\cos(\theta) \leq x) \\
-= P(\theta \geq \arccos(x)) \\
-= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-= 1 - \frac{2}{\pi} \arccos(x).
-$$
-Then the density of X is the derivative of the CDF w.r.t. $x$ $$
-f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-= \frac{2}{\pi \sqrt{1 - x^2}}.
+\begin{aligned}
+P(X \leq x) =& P(\cos(\theta) \leq x) \\
+=& P(\theta \geq \arccos(x)) \\
+=& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+=& 1 - \frac{2}{\pi} \arccos(x).
+\end{aligned}
+$$
+Then the density of X is the derivative of the CDF w.r.t. $x$ 
+$$
+\begin{aligned}
+f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+=& \frac{2}{\pi \sqrt{1 - x^2}}.
+\end{aligned}
 $$
@@
 $$
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} 
 = \frac{|\sin(d_g(p,q))|}{d_g(p,g)},

### 2022-10-10 R/Simulation4d_03plots.R

@@ -1,96 +1,29 @@
 ###--------------------------------------------------------
-## Gather and plot the results from 4 senarios in HPC results
+## 1. Gather and plot the results from 4 scenarios in HPC results
 ###--------------------------------------------------------
+rm(list = ls())
 # first load all .rda files
+library(tidyverse)
 library(viridis)
 N <- 10000
 radius <- 10
 r <- 1
 # method <- c("isomap", "lle", "le", "tsne", "umap")[1]
-load("~/git/kderm/data/simdata_3d_N2000_01_trueden_k100.rda")
+load("data/simdata_100d_4dmanifold_N10000_trueden_k100.rda")
 fxy <- trueden
-label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
-dist2center <- sqrt(radius^2 - semisphere[,5]^2)
+label <- as.factor(c(rep(FALSE, 0.99*N), rep(TRUE, 0.01*N))) # where the points are manual outliers
+dist2center <- sqrt(radius^2 - semisphere[,5]^2) # Euclidean distance to the center O
 methods <- c("isomap", "lle", "le", "umap")
-## scatterplot to compare f_xy for ISOMAP
-f <- tibble(fxy = fxy, fxy_vkde = fisomap$estimate, fxy_hdr = fixden_isomap$estimate) #%>% rank()
-f1 <- tibble(fxy = fxy, fxy_vkde = flle$estimate, fxy_hdr = fixden_lle$estimate) #%>% rank()
-f2 <- tibble(fxy = fxy, fxy_vkde = fle$estimate, fxy_hdr = fixden_le$estimate) #%>% rank()
-f3 <- tibble(fxy = fxy, fxy_vkde = fumap$estimate, fxy_hdr = fixden_umap$estimate) #%>% rank()
-
-summary(f)
-cor(f$fxy_vkde, f$fxy)
-cor(f$fxy_hdr, f$fxy)
-f=f
-pf_vkde <- f %>% 
-  ggplot(aes(x = fxy, y = fxy_vkde, col = dist2center, shape = label)) + 
-  # scale_color_viridis(option = "A", direction = -1) + 
-  geom_point() + 
-  labs(x = "", y = "ISOMAP", color = "Distance", shape = "Kernels", title = "Variable bandwidth") +
-  scale_y_continuous(limits = c(0, max(f$fxy_vkde)), n.breaks = 6)
-pf_hdr <- f %>% 
[…]

### 2022-10-10 paper/paper.tex
100D simulation rank plot

@@
 
-\caption{Highest density region plots of four manifold learning embeddings of the twin peak data.
-Variable KDE preforms better in finding kernel structures with ISOMAP and LLE, and in locating outliers with t-SNE and UMAP.}\label{fig:tpoutliers}
+\caption{Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
+The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel).
+DC-KDE finds more true outliers than KDE in all five rows.}\label{fig:tpoutliers}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_isomap_r0_5_rank_nocolor}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth.
-DC-KDE shows a strong linear positive relationship.}\label{fig:tpisomapden}
+{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
+
+}
+
+\caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpisomapden}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
+
+}
+
+\caption{Scatterplot of true density and estimated density ranks of t-SNE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tptsneden}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_comparison_isomapvs4ml_r0_5_rank}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_4ml_riem20}
-
-}
-
-\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
-The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.}\label{fig:fivedisomapden}
+{\centering \includegraphics[width=0.95\linewidth]{figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec}
+
+}
+
+\caption{Rank comparison between the true density and estimated density from both DC-KDE and KDE.
+Four manifold learning methods are used rowwise.
+The point shapes indicates whether they are the true outleirs, and the grey shading highlights the top 1\% rank region.
+The colors show the distance to the center of the semisphere, with darker points being distant from the center.}\label{fig:fivedisomapden}
 \end{figure}
@@
 \midrule
-Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{0.662} & \textbf{-0.130}\\
-Fixed bandwidth & 0.806 & 0.940 & -0.181 & -0.341\\
+Variable bandwidth & 0.968 & 0.970 & \textbf{0.8674} & \textbf{0.782}\\
+Fixed bandwidth & \textbf{0.976} & \textbf{0.971} & 0.0328 & -0.181\\
 \bottomrule
@@
 \[
-P(X \leq x) = P(\cos(\theta) \leq x) \\
-= P(\theta \geq \arccos(x)) \\
-= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-= 1 - \frac{2}{\pi} \arccos(x).
-\]
-Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
-f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-= \frac{2}{\pi \sqrt{1 - x^2}}.
+\begin{aligned}
+P(X \leq x) =& P(\cos(\theta) \leq x) \\
+=& P(\theta \geq \arccos(x)) \\
+=& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+=& 1 - \frac{2}{\pi} \arccos(x).
+\end{aligned}
+\]
+Then the density of X is the derivative of the CDF w.r.t. \(x\)
+\[
+\begin{aligned}
+f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+=& \frac{2}{\pi \sqrt{1 - x^2}}.
+\end{aligned}
 \]
@@
 \[
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} 
 = \frac{|\sin(d_g(p,q))|}{d_g(p,g)},

### 2022-10-10 R/TwinPeaks.R

@@ -237,7 +237,7 @@ frank_long %>%
 #     as_tibble() %>% 
 #     ggplot(aes(x = E1, y = E2)) + 
 #     ggplot2::geom_point(ggplot2::aes_string(col = "Region")) + 
-#     ggplot2::scale_colour_manual(
+#     ggplot2::scale_color_manual(
 #       name = "HDRs",
 #       breaks = c(paste(head(sort(prob), -1)), ">99"),
 #       values = c(RColorBrewer::brewer.pal(length(prob), "YlOrRd")[-1], "#000000")) + 
@@ -630,7 +630,7 @@ frank_long <- frank %>%
   pivot_longer(cols = -c(fxy), names_to = "kde", values_to = "densities") %>% 
   separate(kde, c("KDE", "ML"), sep = "_") %>% 
   mutate(KDE = factor(KDE, levels = c("vkde", "hdr"), labels = c("DC-KDE", "KDE")),
-         ML = factor(ML, labels = c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"))
+         ML = factor(ML, levels = methods, labels = c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"))
          )
 gt <- frank_long %>%
   ggplot(aes(fxy, densities)) +
@@ -752,7 +752,7 @@ plogrec_isomap_vkde <-
   geom_rect(xmin = rec[1], xmax = rec[2], ymin = rec[1], ymax = rec[2], fill = greys[2], col = greys[2]) + # small blocks
   geom_rect(xmin = 0, xmax = rec[1], ymin = 0, ymax = rec[1], fill = greys[1], col = greys[1]) +
   geom_point(aes(col = col_vkde, shape = dckde_false, size = dckde_false, alpha = dckde_false)) + # if color HDRs
-  ggplot2::scale_colour_manual(
+  scale_color_manual(
     name = "HDRs",
     breaks = levels(frank1$col),
     values = rev(c(RColorBrewer::brewer.pal(length(prob)+1, "YlOrRd")[-1], "#000000"))[1:length(levels)],
@@ -775,7 +775,7 @@ plogrec_isomap_kde <-
   geom_rect(xmin = 0, xmax = rec[1], ymin = 0, ymax = rec[1], fill = greys[1], col = greys[1]) +
   # geom_point()+ # if not color HDRs
[…]

### 2022-10-10 paper/paper.Rmd
Plots all updated for simulations
Table 3 last line has empty line before when knitted from Rmd but not from tex

@@
 ```
+```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
+load("figures/HDRTable_4d_N10000_4ml_radius10_r1.rda")
+hdrtable %>% 
+  kableExtra::kbl(caption = "Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
+                  col.names = c(rep(c("DC-KDE", "KDE"), 4))
+                  ) %>%
+  # kable_classic("striped") %>% 
+  kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = FALSE) %>%
+  # add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
+  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "UMAP" = 2)) %>% 
+  # column_spec(1, bold = TRUE) %>% 
+  column_spec(2, bold = hdrmax[,1]) %>% 
+  column_spec(3, bold = hdrmax[,2]) %>% 
+  column_spec(4, bold = hdrmax[,3]) %>% 
+  column_spec(5, bold = hdrmax[,4]) %>% 
+  column_spec(6, bold = hdrmax[,5]) %>% 
+  column_spec(7, bold = hdrmax[,6]) %>% 
+  column_spec(8, bold = hdrmax[,7]) %>% 
+  column_spec(2*(1:4), border_left = T, background = grey(0.9))
+```
 
@@
 
-# Appendix: Simulation of $S^1$ for manifold density estimation {#s1simulation}
-
-To start with, we demonstrate the use of the proposed DC-KDE on a $1$-dimensional example.
+<!-- # Appendix: Simulation of $S^1$ for manifold density estimation {#s1simulation} -->
+
+<!-- To start with, we demonstrate the use of the proposed DC-KDE on a $1$-dimensional example.
 For a $S^1$ curve centered at $O$ with radius $1$, the coordinated of points on the curve could be expressed in polar coordinates $(1, \alpha)$ or Cartesian coordinates $(x, \sqrt{1-x^2})$.
-
-The first simulation starts with a $1$-dimensional example.
-
-Manifold: $S^1$ with center 0 and radius $r=1$
-
-Polar coordinates: $(r, \theta)$
-
-Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.
-
-Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$
-
-Density of $\theta$: $f(\theta) = 2 / \pi$
-
-Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$
-
-The CDF of X is given by
-$$
-\begin{aligned}
-P(X \leq x) =& P(\cos(\theta) \leq x) \\
-=& P(\theta \geq \arccos(x)) \\
-=& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-=& 1 - \frac{2}{\pi} \arccos(x).
-\end{aligned}
-$$
-Then the density of X is the derivative of the CDF w.r.t. $x$ 
-$$
-\begin{aligned}
-f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-=& \frac{2}{\pi \sqrt{1 - x^2}}.
-\end{aligned}
-$$
-
-Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
-estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
-and compare it with its true density $f(x)$.
-We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
-$$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
-$$
-where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
-
-Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
-Therefore, the geodesic distance is
-$$
-d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-$$
-As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
+-->
+
+<!-- The first simulation starts with a $1$-dimensional example.  -->
+
+<!-- Manifold: $S^1$ with center 0 and radius $r=1$ -->
+
+<!-- Polar coordinates: $(r, \theta)$ -->
+
+<!-- Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$. -->
+
+<!-- Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$ -->
+
+<!-- Density of $\theta$: $f(\theta) = 2 / \pi$ -->
+
+<!-- Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$ -->
+
+<!-- The CDF of X is given by -->
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!-- P(X \leq x) =& P(\cos(\theta) \leq x) \\ -->
+<!-- =& P(\theta \geq \arccos(x)) \\ -->
+<!-- =& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\ -->
+<!-- =&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\ -->
+<!-- =& 1 - \frac{2}{\pi} \arccos(x). -->
+<!-- \end{aligned} -->
+<!-- $$ -->
+<!-- Then the density of X is the derivative of the CDF w.r.t. $x$  -->
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!-- f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\ -->
+<!-- =& \frac{2}{\pi \sqrt{1 - x^2}}. -->
+<!-- \end{aligned} -->
+<!-- $$ -->
+
+<!-- Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could -->
+<!-- estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ -->
+<!-- and compare it with its true density $f(x)$. -->
+<!-- We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by -->
+<!-- $$ -->
+<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}), -->
+<!-- $$ -->
+<!-- where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
+-->
+
+<!-- Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
+Therefore, the geodesic distance is -->
+<!-- $$ -->
+<!-- d_g(p, X_i) = \arccos(p) - \arccos(X_i) -->
+<!-- $$ -->
+<!-- As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
 $|g(\exp_p^{-1}(x_i))|$.
-
-[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]
-
-The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
-$$
-g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-$$
-is smooth in a local neighborhood $U$ of $M$.
-
-Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
-$$
-\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
-$$
-Therefore, the estimator for $p \in S^1$ is given by
-$$
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-$$
-where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*.
-
-### How to derive volume density function $\theta_{X_i}(p)$
-
-#### Example of $S^2$ in @Henry2009-om.
-
-Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
+-->
+
+<!-- [When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.] -->
+
+<!-- The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$, -->
+<!-- $$ -->
+<!-- g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R}, -->
+<!-- $$ -->
+<!-- is smooth in a local neighborhood $U$ of $M$. -->
+
+<!-- Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by -->
+<!-- $$ -->
+<!-- \theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}. -->
+<!-- $$ -->
+<!-- Therefore, the estimator for $p \in S^1$ is given by -->
+<!-- $$ -->
+<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg), -->
+<!-- $$ -->
+<!-- where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*. -->
+
+<!-- ### How to derive volume density function $\theta_{X_i}(p)$ -->
+
+<!-- #### Example of $S^2$ in @Henry2009-om. -->
+
+<!-- Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
 Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$.
-Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by
-$$
-\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
-$$
-if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$.
-Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
-Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that
-$$
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
-$$
-$$
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
-$$
-The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors.
-Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
-Therefore, we obtain that the volume density on the sphere is
-$$
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} 
-= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
-$$
-for $q \neq p, -p$, and $\theta_p(p) = 1$.
-
-#### $S^1$
-
-Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization
-$$
-\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
-$$
-and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have
-$$
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
-$$
-So the riemannian metric is $g_q = \cos(r)$ and further,
-$$
-\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
-$$
-where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$.
+Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by -->
+<!-- $$ -->
+<!-- \Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw), -->
+<!-- $$ -->
+<!-- if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. -->
+<!-- Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$. -->
+<!-- Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that -->
+<!-- $$ -->
+<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w. -->
+<!-- $$ -->
+<!-- The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors. -->
+<!-- Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
+-->
+<!-- Therefore, we obtain that the volume density on the sphere is -->
+<!-- $$ -->
+<!-- \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2}  -->
+<!-- = \frac{|\sin(d_g(p,q))|}{d_g(p,g)}, -->
+<!-- $$ -->
+<!-- for $q \neq p, -p$, and $\theta_p(p) = 1$. -->
+
+<!-- #### $S^1$ -->
+
+<!-- Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization -->
+<!-- $$ -->
+<!-- \Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv, -->
+<!-- $$ -->
+<!-- and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have -->
+<!-- $$ -->
+<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v. -->
+<!-- $$ -->
+<!-- So the riemannian metric is $g_q = \cos(r)$ and further, -->
+<!-- $$ -->
+<!-- \theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|}, -->
+<!-- $$ -->
+<!-- where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$. -->
 

### 2022-10-10 R/TwinPeaks.R

@@ -612,7 +612,6 @@ cors %>%
   kable_styling(latex_options = "scale_down") %>%
   kable_paper(full_width = TRUE) %>%
   row_spec(1, bold = TRUE)
-
 # save(cors, file = paste0("paper/figures/CorrelationTable_", mapping, N,"_5ml_radius", radius, "_r", format(r, decimal.mark = "_"), ".rda"))
 
 
@@ -720,7 +719,7 @@ gt
 # RUN this!!!
 ##---------------------------------------------------------------------
 # color the points with HDR, add rectangle and highlight points outside
-xname <- c("isomap", "tsne")[2] # !!! Modify index for different ML methods comparison
+xname <- c("isomap", "tsne")[1] # !!! Modify index for different ML methods comparison
 probs <- prob[1:2]
 levels <- c(probs / 100, 1)
 q_rank <- quantile(1:N, levels)
@@ -748,7 +747,7 @@ plogrec_isomap_vkde <-
   frank1 %>%
   ggplot(aes(x = fxy, y = fxy_vkde)) +
   # geom_rect(xmin = 0, xmax = rec[2], ymin = 0, ymax = rec[2], fill = greys[2], col = greys[2]) +
-  # geom_rect(xmin = rec[2], xmax = rec[3], ymin = rec[2], ymax = rec[3], fill = greys[3], col = greys[3]) + # third blocks
+  geom_rect(xmin = rec[2], xmax = rec[3], ymin = rec[2], ymax = rec[3], fill = greys[3], col = greys[3]) + # third blocks
   geom_rect(xmin = rec[1], xmax = rec[2], ymin = rec[1], ymax = rec[2], fill = greys[2], col = greys[2]) + # small blocks
   geom_rect(xmin = 0, xmax = rec[1], ymin = 0, ymax = rec[1], fill = greys[1], col = greys[1]) +
   geom_point(aes(col = col_vkde, shape = dckde_false, size = dckde_false, alpha = dckde_false)) + # if color HDRs
@@ -764,13 +763,14 @@ plogrec_isomap_vkde <-
   scale_shape_manual(values = c(`TRUE` = 20, `FALSE` = 17 )) +
   scale_size_manual(values = c(`TRUE` = 1, `FALSE` = 2 )) +
   scale_alpha_manual(values = c(`TRUE` = .5, `FALSE` = 1) ) + 
[…]

### 2022-10-10 R/_test.R

@@ -1,0 +1,199 @@
+set.seed(1)
+N <- 2000
+mu1 = mu2 = 0
+s1 <- 1#/(15^2)
+s2 <- 2#/(15^2)
+p1 <- 0.99
+p2 <- 0.01
+x1 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+x2 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+x3 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+x4 <- c(rnorm(0.99*N, mu1, s1), rnorm(0.01*N, mu2, s2))
+X <- cbind(x1, x2, x3, x4)
+# semi-sphere radius
+# scales::rescale(x, c(0,1))
+r <- ceiling(max(sqrt(x1^2 + x2^2 + x3^2 + x4^2)))
+r
+range(X)
+
+# X <- matrix(rnorm(4*N,0,1), N, 4)
+X_new <- cbind(X, 
+               x5 = sqrt(6^2 - (x1^2 + x2^2 + x3^2 + x4^2))
+)
+X_new %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2) # radius is 15
+
+library(geozoo)
+sphere <- sphere.hollow(p = 5)
+sphere$points <- X_new
+sphere
+
+# GMM density as true meta data density
+gmmdensity <- function(x) {
+  p1 * dmvnorm(x, rep(mu1, 4), diag(s1, 4)) + 
+    p2 *  dmvnorm(x, rep(mu2, 4), diag(s2, 4))
+}
+den <- apply(X, 1, gmmdensity)
+length(den)
+range(den)
+summary(den)
+
+library(tourr)
+animate_xy(X_new[,1:5], col = viridis(length(den),option = "magma"))
+
+col <- RColorBrewer::brewer.pal(length(den), "Dark2") # unique(flea$species)
+animate_xy(flea[, 1:6], col = flea$species)
+
+
+
+
+library(ks)
+set.seed(8192)
+samp <- 200
+mus <- rbind(c(-2,2), c(0,0), c(2,-2))
+Sigmas <- rbind(diag(2), matrix(c(0.8, -0.72, -0.72, 0.8), nrow=2), diag(2))
+cwt <- 3/11
[…]

### 2022-10-10 paper/paper.tex
Plots all updated for simulations
Table 3 last line has empty line before when knitted from Rmd but not from tex

@@
 \midrule
-Variable bandwidth & 0.968 & 0.970 & \textbf{0.8674} & \textbf{0.782}\\
-Fixed bandwidth & \textbf{0.976} & \textbf{0.971} & 0.0328 & -0.181\\
+DC-KDE & 0.968 & 0.970 & \textbf{0.8674} & \textbf{0.782}\\
+KDE & \textbf{0.976} & \textbf{0.971} & 0.0328 & -0.181\\
 \bottomrule
 \end{tabular}
+\end{table}
+\begin{table}
+
+\caption{\label{tab:fourdhdrtable}Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
+  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
+\midrule
+>99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
+99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
+90 & \cellcolor[HTML]{E6E6E6}{0.919} & \textbf{0.926} & \cellcolor[HTML]{E6E6E6}{\textbf{0.921}} & 0.900 & \cellcolor[HTML]{E6E6E6}{\textbf{0.798}} & 0.440 & \cellcolor[HTML]{E6E6E6}{\textbf{0.779}} & 0.428\\
+50 & \cellcolor[HTML]{E6E6E6}{0.825} & \textbf{0.851} & \cellcolor[HTML]{E6E6E6}{0.831} & \textbf{0.848} & \cellcolor[HTML]{E6E6E6}{\textbf{0.659}} & 0.512 & \cellcolor[HTML]{E6E6E6}{\textbf{0.623}} & 0.448\\
+5 & \cellcolor[HTML]{E6E6E6}{0.508} & \textbf{0.556} & \cellcolor[HTML]{E6E6E6}{0.514} & \textbf{0.562} & \cellcolor[HTML]{E6E6E6}{\textbf{0.431}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.220}} & 0.000\\
+% \addlinespace
+1 & \cellcolor[HTML]{E6E6E6}{0.080} & \textbf{0.190} & \cellcolor[HTML]{E6E6E6}{0.090} & \textbf{0.230} & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000\\
+\bottomrule
+\end{tabular}}
 \end{table}
@@
 
-\hypertarget{s1simulation}{%
-\section{\texorpdfstring{Appendix: Simulation of \(S^1\) for manifold density estimation}{Appendix: Simulation of S\^{}1 for manifold density estimation}}\label{s1simulation}}
-
-To start with, we demonstrate the use of the proposed DC-KDE on a \(1\)-dimensional example.
-For a \(S^1\) curve centered at \(O\) with radius \(1\), the coordinated of points on the curve could be expressed in polar coordinates \((1, \alpha)\) or Cartesian coordinates \((x, \sqrt{1-x^2})\).
-
-The first simulation starts with a \(1\)-dimensional example.
-
-Manifold: \(S^1\) with center 0 and radius \(r=1\)
-
-Polar coordinates: \((r, \theta)\)
-
-Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).
-
-Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)
-
-Density of \(\theta\): \(f(\theta) = 2 / \pi\)
-
-Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)
-
-The CDF of X is given by
-\[
-\begin{aligned}
-P(X \leq x) =& P(\cos(\theta) \leq x) \\
-=& P(\theta \geq \arccos(x)) \\
-=& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-=& 1 - \frac{2}{\pi} \arccos(x).
-\end{aligned}
-\]
-Then the density of X is the derivative of the CDF w.r.t. \(x\)
-\[
-\begin{aligned}
-f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-=& \frac{2}{\pi \sqrt{1 - x^2}}.
-\end{aligned}
-\]
-
-Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
-estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
-and compare it with its true density \(f(x)\).
-We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
-\]
-where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).
-
-Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\).
-Therefore, the geodesic distance is
-\[
-d_g(p, X_i) = \arccos(p) - \arccos(X_i)
-\]
-As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).
-
-{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}
-
-The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
-\[
-g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
-\]
-is smooth in a local neighborhood \(U\) of \(M\).
-
-Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
-\[
-\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
-\]
-Therefore, the estimator for \(p \in S^1\) is given by
-\[
-f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
-\]
-where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.
-
-\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
-\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}
-
-\hypertarget{example-of-s2-in-henry2009-om.}{%
-\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}
-
-Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\).
-Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\).
-Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
-\[
-\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
-\]
-if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
-Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
-Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
-\[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
-\]
-\[
-B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
-\]
-The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
-Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
-Therefore, we obtain that the volume density on the sphere is
-\[
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} 
-= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
-\]
-for \(q \neq p, -p\), and \(\theta_p(p) = 1\).
-
-\hypertarget{s1}{%
-\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}
-
-Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
-\[
-\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
-\]
-and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
-\[
-A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
-\]
-So the riemannian metric is \(g_q = \cos(r)\) and further,
-\[
-\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
-\]
-where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).
-
 \hypertarget{swissrollappe}{%

### 2022-10-10 R/Simulation4d_03plots.R

@@ -17,7 +17,12 @@ methods <- c("isomap", "lle", "le", "umap")
 # load HPC results for four methods
 files <- list.files(path = "data/truedenk100", pattern = "*.rda", full.names = TRUE)
 lapply(files, load, .GlobalEnv)
-
+f <- tibble(fxy = fxy,
+            fxy_vkde_isomap = fisomap$estimate, fxy_hdr_isomap = fixden_isomap$estimate,
+            fxy_vkde_lle = flle$estimate, fxy_hdr_lle = fixden_lle$estimate,
+            fxy_vkde_le = fle$estimate, fxy_hdr_le = fixden_le$estimate,
+            fxy_vkde_umap = fumap$estimate, fxy_hdr_umap = fixden_umap$estimate
+)
 
 ###--------------------------------------------------------
 ## 2. Table for density correlations
@@ -42,23 +47,67 @@ cors %>%
 
 
 ###--------------------------------------------------------
-## 2. Table for correctlly identified outliers percentage
+## 2.2 Table for correctlly identified outliers percentage
 ###--------------------------------------------------------
-
-
+prob <- c(1, 10, 50, 95, 99)
+hdrbreaks <- c(0, quantile(1:N, c(sort(100 - prob) / 100)), N)
+hdrlabels <- c(">99", paste0(100 - prob)) #, "%", not working for latex, \\%
+frankhdr <- f %>% 
+  summarise_all(.fun = list(~ rank(x = ., ties.method = "average"))) %>% # using average ties
+  # mutate(outliers = factor(label, levels = c("TRUE", "FALSE")),
+  #        dist2center = dist2center) %>%
+  `colnames<-`(gsub("fxy_", "", colnames(.))) %>% 
+  pivot_longer(cols = -c(fxy), names_to = "kde", values_to = "densities") %>% 
+  separate(kde, c("KDE", "ML"), sep = "_") %>% 
+  mutate(KDE = factor(KDE, levels = c("vkde", "hdr"), labels = c("DC-KDE", "KDE")),
+         ML = factor(ML, levels = methods, labels = c("ISOMAP", "LLE", "Laplacian Eigenmaps", "UMAP"))
+  ) %>% 
+  mutate(truehdr = cut(fxy, 
+                       breaks = hdrbreaks, 
[…]

### 2022-10-10 paper/paper.tex
Add appendix plots for other ML methods

@@
 5 & \cellcolor[HTML]{E6E6E6}{0.508} & \textbf{0.556} & \cellcolor[HTML]{E6E6E6}{0.514} & \textbf{0.562} & \cellcolor[HTML]{E6E6E6}{\textbf{0.431}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.220}} & 0.000\\
-% \addlinespace
+\addlinespace
 1 & \cellcolor[HTML]{E6E6E6}{0.080} & \textbf{0.190} & \cellcolor[HTML]{E6E6E6}{0.090} & \textbf{0.230} & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000\\
@@
 
+\hypertarget{twinpeaksappe}{%
+\section{Appendix: Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
+
+This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
+By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into highest density regions more accurately than KDE.
+By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest density regions, which are usually used to detect anomalies.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_lle_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
+
+}
+
+\caption{Scatterplot of true density and estimated density ranks of LLE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tplleden}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_le_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
+
+}
+
+\caption{Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpleden}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_umap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
+
+}
+
+\caption{Scatterplot of true density and estimated density ranks of UMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpumapden}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank}
+
+}
+
+\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-1}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank}
+
+}
+
+\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-2}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank}
+
+}
+
+\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-3}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank}
+
+}
+
+\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-4}
+\end{figure}
+
 \hypertarget{swissrollappe}{%

### 2022-10-10 R/TwinPeaks.R

@@ -719,7 +719,7 @@ gt
 # RUN this!!!
 ##---------------------------------------------------------------------
 # color the points with HDR, add rectangle and highlight points outside
-xname <- c("isomap", "tsne")[1] # !!! Modify index for different ML methods comparison
+xname <- methods[2] # !!! Modify index for different ML methods comparison
 probs <- prob[1:2]
 levels <- c(probs / 100, 1)
 q_rank <- quantile(1:N, levels)

### 2022-10-10 paper/paper.Rmd
Add appendix plots for other ML methods

@@
 
+# Appendix: Rank comparison plots for twin peaks mapping {#twinpeaksappe}
+
+This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
+By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into highest density regions more accurately than KDE.
+By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest density regions, which are usually used to detect anomalies.
+
+```{r tplleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of LLE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_lle_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
+```
+
+```{r tpleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_le_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
+```
+
+```{r tpumapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of UMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_umap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
+```
+
+```{r tpllevs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
+# knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
+knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
+```
+
 <!-- \clearpage -->

### 2022-10-10 paper/paper.tex
minor

--- 2022-10-10 paper/paper.tex
@@
 \label{eq:volumedensityratio}
-\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }  .
+\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
 \end{equation}
@@
 
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob}
 
@@
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpisomapden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of t-SNE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tptsneden}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec}
+{\centering \includegraphics[width=0.95\linewidth]{figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity}
 
@@
 
-\caption{Rank comparison between the true density and estimated density from both DC-KDE and KDE.
-Four manifold learning methods are used rowwise.
-The point shapes indicates whether they are the true outleirs, and the grey shading highlights the top 1\% rank region.
-The colors show the distance to the center of the semisphere, with darker points being distant from the center.}\label{fig:fivedisomapden}
+\caption{Scatterplot of true density and estimated density of different embeddings for KDE with both variable and fixed bandwidth.
+The point shapes indicates the two Gaussian mixture components and the colors shows the distance to the kernel cores.}\label{fig:fivedisomapden}
 \end{figure}
@@
 \midrule
-Variable bandwidth & 0.968 & 0.970 & \textbf{0.8674} & \textbf{0.782}\\
-Fixed bandwidth & \textbf{0.976} & \textbf{0.971} & 0.0328 & -0.181\\
+Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{0.662} & \textbf{-0.130}\\
+Fixed bandwidth & 0.806 & 0.940 & -0.181 & -0.341\\
 \bottomrule
@@
 \[
-\begin{aligned}
-P(X \leq x) =& P(\cos(\theta) \leq x) \\
-=& P(\theta \geq \arccos(x)) \\
-=& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
-=&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
-=& 1 - \frac{2}{\pi} \arccos(x).
-\end{aligned}
-\]
-Then the density of X is the derivative of the CDF w.r.t. \(x\)
-\[
-\begin{aligned}
-f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
-=& \frac{2}{\pi \sqrt{1 - x^2}}.
-\end{aligned}
+P(X \leq x) = P(\cos(\theta) \leq x) \\
+= P(\theta \geq \arccos(x)) \\
+= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
+=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
+= 1 - \frac{2}{\pi} \arccos(x).
+\]
+Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
+f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
+= \frac{2}{\pi \sqrt{1 - x^2}}.
 \]
@@
 \[
-\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} 
+\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
 = \frac{|\sin(d_g(p,q))|}{d_g(p,g)},

### 2022-10-10 paper/paper.Rmd
Change load file name

@@
 ```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
-load("figures/HDRTable_4d_N10000_4ml_radius10_r1.rda")
+load("figures/hdrtable_4d_N10000_4ml_radius10_r1.rda")
 hdrtable %>% 

### 2022-10-11 paper/paper.tex

--- 2022-10-10 paper/paper.tex
@@
 \hypertarget{kdermtas}{%
-\section{Kernel density estimate on Riemannian manifolds - Tas}\label{kdermtas}}
-
-In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}.
-We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
-Finally, we describe how our own proposed method generalises the estimator of \textcite{Pelletier2005-vu} to the case where the underlying manifold is unknown.
-Denoting \(\lambda\) as the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
+\section{Distortion Corrected Kernel density estimate on Riemannian manifolds - Tas}\label{kdermtas}}
+
+In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
+Readers familiar with kernel density estimation may not be familiar with some of the nuances of manifolds.
+Therefore, before introducing our own estimator we first introduce kernel density estimation for data in Euclidean space, then illustrate in Section \ref{Pellet} how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+In Section \ref{MetLearn} we describe the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
+By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq} we derive our own novel distortion corrected kernel density estimate in Section \ref{DCKDE}.
+This allows for kernel density estimation on unknown manifolds using any popular dimension reduction algorithm.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers less familiar with differential geometry in Appendix \ref{riemgeo}.
+
+In the following we denote as \(M\) the \(d\)-dimensional manifold from which our data are sampled.
+Points on this manifold are denoted \(\pmb{p}\) in general with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
+Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m>>d\).
+However this need not be that case, for instance \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at \(\pmb{p}_i\) only require that there be available some sense of `input' distance \(d(\pmb{p}_i,\pmb{p}_j)\).
+We will apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\).
+Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
 \begin{align}
 \label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=1;
-(ii) \int_{\mathbb{R}^{d}} \pmb{x} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=0;
-(iii) \int_{\mathbb{R}^{d}}\|\pmb{x}\|^{2} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})<\infty; \\
+& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
+(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
+(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
 & (iv) \supp K=[0; 1];
-(v) \sup K(||\pmb{x}||)=K(0).
+(v) \sup K(||\pmb{y}||)=K(0).
 \end{align}
@@
 
-Assume the data \(\pmb{x}_i\in\mathbb{R}^d\) for \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{x}\) is given by
+For data \(\pmb{y}_i\in\mathbb{R}^d\) for \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{y}\) is given by
 
@@
 \label{eq:vkde}
-\hat{f}(\pmb{x})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{x}-\pmb{x}_i||}{r}\right).
+\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{y}-\pmb{y}_i||}{r}\right).
 \end{equation}
@@
 The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point \(\pmb{x}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{x}\), times a term that ensures the density integrates to 1.
+The density at a point \(\pmb{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{y}\), times a term that ensures the density integrates to 1.
 In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{x}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
+However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
 For more on kernel density estimation in the Euclidean case see \textcite{Scott2015-vl} and references therein.
+
+While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors \(\pmb{y}_i\) is different to the density on the manifold itself.
+As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
+Furthermore any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
+
+\hypertarget{Pellet}{%
+\subsection{Kernel Density estimation on manifolds}\label{Pellet}}
 
@@
 Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space \(\pmb{v}\in T_{\pmb{p}}M\) to points the manifold \(\pmb{q}\in M\) .
-Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and along this geodesic at uniform speed \(||\pmb{v}||\) takes place in one unit of time.
+Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and travel along this geodesic at uniform speed \(||\pmb{v}||\) takes place in one unit of time.
 Now, consider a chart \(\varphi\) mapping points in the neighbourhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
@@
 
-\hypertarget{distortion-corrected-kde}{%
-\subsection{Distortion corrected KDE}\label{distortion-corrected-kde}}
+\hypertarget{MetLearn}{%
+\subsection{Riemannian metric estimation}\label{MetLearn}}
+
+To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case there the manifold is not known, but where an output embedding from a dimension reduction algorithm is available requires additional estimates of the Riemannian metric.
+Formally, the Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
+The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).}, can be used to define geometric quantities for example angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
+While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
+This is precisely the situation when data on a manifold are mapped to \(d\)-dimensional Euclidean vectors \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) via a dimension reduction algorithm.
+After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion.
+To alleviate this issue \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\) at each data point.
+These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)\).
+
+Locally isometric embedding to go here\ldots{}
+
+While full details are provided in \textcite{Perrault-Joncas2013-pq} , we briefly describe the Learn Metric algorithm here.
+There are four main steps in the algorithm.
+First a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold.
+Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+Third, a dimension reduction method is applied to obtain the output embedding \(\pmb{y}_1,\dots,\pmb{y}_n\).
+Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at step 2 is a discrete estimator).
+Full details on these four steps are provided in Algorithm \ref{alg:learnmetric}.
+This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
+
+As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(\pmb{H}(p)\) is equal to \(d\); otherwise, the Riemannian metric will be returned.
+To ensure that the estimated Riemannian metric is returned as a \(d\times d\) matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} to derive id estimator from a Poisson point process.
+\textcite{Facco2017-rl} point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter \(1\) and shape parameter \(d\), the intrinsic dimension of the manifold \autocite{Facco2017-rl}.
+By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
+An \emph{R} library named \emph{intRinsic} \autocite{Denti2021-qc} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.
+
+\begin{algorithm}[!htb]
+  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
+  \label{alg:learnmetric}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
+  \OptParameter{ manifold learning parameters EMBED }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
+
+  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
+  $$
+  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
+  $$
+  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
+
+  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
+
+  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
+  $$
+    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+  $$
+  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
+
+  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
+  $$
+    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
+  $$
+  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+
+  \end{algorithmic}
+\end{algorithm}
+
+\hypertarget{DCKDE}{%
+\subsection{Distortion corrected KDE}\label{DCKDE}}
 
@@
 \end{table}
-\begin{table}
-
-\caption{\label{tab:fourdhdrtable}Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
-  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
-\midrule
->99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
-99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
-90 & \cellcolor[HTML]{E6E6E6}{0.919} & \textbf{0.926} & \cellcolor[HTML]{E6E6E6}{\textbf{0.921}} & 0.900 & \cellcolor[HTML]{E6E6E6}{\textbf{0.798}} & 0.440 & \cellcolor[HTML]{E6E6E6}{\textbf{0.779}} & 0.428\\
-50 & \cellcolor[HTML]{E6E6E6}{0.825} & \textbf{0.851} & \cellcolor[HTML]{E6E6E6}{0.831} & \textbf{0.848} & \cellcolor[HTML]{E6E6E6}{\textbf{0.659}} & 0.512 & \cellcolor[HTML]{E6E6E6}{\textbf{0.623}} & 0.448\\
-5 & \cellcolor[HTML]{E6E6E6}{0.508} & \textbf{0.556} & \cellcolor[HTML]{E6E6E6}{0.514} & \textbf{0.562} & \cellcolor[HTML]{E6E6E6}{\textbf{0.431}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.220}} & 0.000\\
-\addlinespace
-1 & \cellcolor[HTML]{E6E6E6}{0.080} & \textbf{0.190} & \cellcolor[HTML]{E6E6E6}{0.090} & \textbf{0.230} & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000\\
-\bottomrule
-\end{tabular}}
-\end{table}
 

### 2022-10-11 paper/paper.Rmd

@@
 
-# Kernel density estimate on Riemannian manifolds - Tas {#kdermtas}
-
-In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm and the Metric Learning algorithm of @Perrault-Joncas2013-pq.
-We will do so by first introducing kernel density estimation for data in Euclidean space, then illustrating how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
-Finally, we describe how our own proposed method generalises the estimator of @Pelletier2005-vu to the case where the underlying manifold is unknown.
-Denoting $\lambda$ as the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
+# Distortion Corrected Kernel density estimate on Riemannian manifolds - Tas {#kdermtas}
+
+In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
+Readers familiar with kernel density estimation may not be familiar with some of the nuances of manifolds.
+Therefore, before introducing our own estimator we first introduce kernel density estimation for data in Euclidean space, then illustrate in Section \@ref(Pellet) how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+In Section \@ref(MetLearn) we describe the Metric Learning algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
+By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq we derive our own novel distortion corrected kernel density estimate in Section \@ref(DCKDE).
+This allows for kernel density estimation on unknown manifolds using any popular dimension reduction algorithm.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers less familiar with differential geometry in Appendix \@ref(riemgeo).
+
+In the following we denote as $M$ the $d$-dimensional manifold from which our data are sampled.
+Points on this manifold are denoted $\pmb{p}$ in general with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
+Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m>>d$.
+However this need not be that case, for instance $\pmb{p}_i$ may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at $\pmb{p}_i$ only require that there be available some sense of 'input' distance $d(\pmb{p}_i,\pmb{p}_j)$.
+We will apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$.
+Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
 \begin{align}
 \label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=1;
-(ii) \int_{\mathbb{R}^{d}} \pmb{x} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})=0;
-(iii) \int_{\mathbb{R}^{d}}\|\pmb{x}\|^{2} K(\|\pmb{x}\|) \mathrm{d} \lambda(\pmb{x})<\infty; \\
+& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
+(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
+(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
 & (iv) \supp K=[0; 1];
-(v) \sup K(||\pmb{x}||)=K(0).
+(v) \sup K(||\pmb{y}||)=K(0).
 \end{align}
@@
 
-Assume the data $\pmb{x}_i\in\mathbb{R}^d$ for $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{x}$ is given by
+For data $\pmb{y}_i\in\mathbb{R}^d$ for $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by
 
@@
 The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point $\pmb{x}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{x}$, times a term that ensures the density integrates to 1.
+The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1.
 In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{x}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
+However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
 For more on kernel density estimation in the Euclidean case see @Scott2015-vl and references therein.
+
+While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors $\pmb{y}_i$ is different to the density on the manifold itself.
+As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
+Furthermore any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
+
+## Kernel Density estimation on manifolds {#Pellet}
 
@@
 Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space $\pmb{v}\in T_{\pmb{p}}M$ to points the manifold $\pmb{q}\in M$ .
-Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and along this geodesic at uniform speed $||\pmb{v}||$ takes place in one unit of time.
+Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travel along this geodesic at uniform speed $||\pmb{v}||$ takes place in one unit of time.
 Now, consider a chart $\varphi$ mapping points in the neighbourhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
@@
 
-## Distortion corrected KDE
+## Riemannian metric estimation {#MetLearn}
+
+To be able to apply the estimator of @Pelletier2005-vu to the case there the manifold is not known, but where an output embedding from a dimension reduction algorithm is available requires additional estimates of the Riemannian metric.
+Formally, the Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
+The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$[^1], can be used to define geometric quantities for example angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
+While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
+This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.
+After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion.
+To alleviate this issue @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$ at each data point.
+These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)$.
+
+Locally isometric embedding to go here...
+
+While full details are provided in @Perrault-Joncas2013-pq , we briefly describe the Learn Metric algorithm here.
+There are four main steps in the algorithm.
+First a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold.
+Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+Third, a dimension reduction method is applied to obtain the output embedding $\pmb{y}_1,\dots,\pmb{y}_n$.
+Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at step 2 is a discrete estimator).
+Full details on these four steps are provided in Algorithm \@ref(alg:learnmetric).
+This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*.
+
+As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned.
+To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process.
+@Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl].
+By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
+An *R* library named *intRinsic* [@Denti2021-qc] has implemented this algorithm and we are going to apply it to when $d$ is unknown.
+
+\begin{algorithm}[!htb]
+  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
+  \label{alg:learnmetric}
+  \DontPrintSemicolon
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
+  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
+  \OptParameter{ manifold learning parameters EMBED }
+  \BlankLine
+  \begin{algorithmic}[1]
+
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
+
+  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
+  $$
+  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
+  $$
+  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
+
+  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
+
+  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
+  $$
+    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+  $$
+  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
+
+  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
+  $$
+    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
+  $$
+  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+
+  \end{algorithmic}
+\end{algorithm}
+
+## Distortion corrected KDE {#DCKDE}
 
@@
 ```
-```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
-load("figures/hdrtable_4d_N10000_4ml_radius10_r1.rda")
+
+```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=F}
+load("figures/HDRTable_4d_N10000_4ml_radius10_r1.rda")
 hdrtable %>% 

### 2022-10-11 paper/paper.Rmd

@@
 In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Readers familiar with kernel density estimation may not be familiar with some of the nuances of manifolds.
-Therefore, before introducing our own estimator we first introduce kernel density estimation for data in Euclidean space, then illustrate in Section \@ref(Pellet) how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in Section \@ref(Pellet) how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
 In Section \@ref(MetLearn) we describe the Metric Learning algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
 By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq we derive our own novel distortion corrected kernel density estimate in Section \@ref(DCKDE).
-This allows for kernel density estimation on unknown manifolds using any popular dimension reduction algorithm.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers less familiar with differential geometry in Appendix \@ref(riemgeo).
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in Appendix \@ref(riemgeo).
 
@@
 Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m>>d$.
-However this need not be that case, for instance $\pmb{p}_i$ may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at $\pmb{p}_i$ only require that there be available some sense of 'input' distance $d(\pmb{p}_i,\pmb{p}_j)$.
-We will apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$.
+However this need not be that case, for instance the $\pmb{p}_i$ may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at each $\pmb{p}_i$ only require that some sense of distance between the 'input' points $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$.
 Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
@@
 
-To extend this to the case where data are on a manifold $\pmb{p}_i\in\mathcal{M}$, for $i=1,\dots,N$, @Pelletier2005-vu propose the following estimator,
+For kernel density estimation on a known manifold, @Pelletier2005-vu propose the following estimator,
 
@@
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
-However the volume of a ball of radius $r$ around $\pmb{p}$ must also be taken into account.
+However the volume of this ball must also be taken into account.
 In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center.
@@
 
-To be able to apply the estimator of @Pelletier2005-vu to the case there the manifold is not known, but where an output embedding from a dimension reduction algorithm is available requires additional estimates of the Riemannian metric.
+To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where an output embedding from a dimension reduction algorithm is available, requires an estimate of the Riemannian metric.
 Formally, the Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
-The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$[^1], can be used to define geometric quantities for example angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
+The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
+For example angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
 While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
@@
 To alleviate this issue @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$ at each data point.
-These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)$.
+These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
 
@@
 
-As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned.
+As pointed out by @Perrault-Joncas2013-pq, if the dimension reduction is carried out such that the dimension of the output vectors is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned.
+
 To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process.
@@
 
-```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=F}
-load("figures/HDRTable_4d_N10000_4ml_radius10_r1.rda")
+```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
+load("figures/hdrtable_4d_N10000_4ml_radius10_r1.rda")
 hdrtable %>% 

### 2022-10-11 paper/paper.tex

@@
 In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Readers familiar with kernel density estimation may not be familiar with some of the nuances of manifolds.
-Therefore, before introducing our own estimator we first introduce kernel density estimation for data in Euclidean space, then illustrate in Section \ref{Pellet} how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in Section \ref{Pellet} how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
 In Section \ref{MetLearn} we describe the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
 By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq} we derive our own novel distortion corrected kernel density estimate in Section \ref{DCKDE}.
-This allows for kernel density estimation on unknown manifolds using any popular dimension reduction algorithm.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers less familiar with differential geometry in Appendix \ref{riemgeo}.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in Appendix \ref{riemgeo}.
 
@@
 Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m>>d\).
-However this need not be that case, for instance \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at \(\pmb{p}_i\) only require that there be available some sense of `input' distance \(d(\pmb{p}_i,\pmb{p}_j)\).
-We will apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\).
+However this need not be that case, for instance the \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at each \(\pmb{p}_i\) only require that some sense of distance between the `input' points \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\).
 Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
@@
 
-To extend this to the case where data are on a manifold \(\pmb{p}_i\in\mathcal{M}\), for \(i=1,\dots,N\), \textcite{Pelletier2005-vu} propose the following estimator,
+For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
 
@@
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
-However the volume of a ball of radius \(r\) around \(\pmb{p}\) must also be taken into account.
+However the volume of this ball must also be taken into account.
 In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center.
@@
 
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case there the manifold is not known, but where an output embedding from a dimension reduction algorithm is available requires additional estimates of the Riemannian metric.
+To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where an output embedding from a dimension reduction algorithm is available, requires an estimate of the Riemannian metric.
 Formally, the Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
-The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).}, can be used to define geometric quantities for example angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
+The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\), can be used to define geometric quantities.
+For example angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
 While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
@@
 To alleviate this issue \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\) at each data point.
-These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}(\pmb{y}_k-\pmb{y}_i)\).
+These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).
 
@@
 
+\begin{table}
+
+\caption{\label{tab:fourdhdrtable}Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
+  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
+\midrule
+>99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
+99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
+90 & \cellcolor[HTML]{E6E6E6}{0.919} & \textbf{0.926} & \cellcolor[HTML]{E6E6E6}{\textbf{0.921}} & 0.900 & \cellcolor[HTML]{E6E6E6}{\textbf{0.798}} & 0.440 & \cellcolor[HTML]{E6E6E6}{\textbf{0.779}} & 0.428\\
+50 & \cellcolor[HTML]{E6E6E6}{0.825} & \textbf{0.851} & \cellcolor[HTML]{E6E6E6}{0.831} & \textbf{0.848} & \cellcolor[HTML]{E6E6E6}{\textbf{0.659}} & 0.512 & \cellcolor[HTML]{E6E6E6}{\textbf{0.623}} & 0.448\\
+5 & \cellcolor[HTML]{E6E6E6}{0.508} & \textbf{0.556} & \cellcolor[HTML]{E6E6E6}{0.514} & \textbf{0.562} & \cellcolor[HTML]{E6E6E6}{\textbf{0.431}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.220}} & 0.000\\
+\addlinespace
+1 & \cellcolor[HTML]{E6E6E6}{0.080} & \textbf{0.190} & \cellcolor[HTML]{E6E6E6}{0.090} & \textbf{0.230} & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
 \hypertarget{application}{%

### 2022-10-24 paper/paper.Rmd
completed section 2 and deleted old section 2

@@
 However this need not be that case, for instance the $\pmb{p}_i$ may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each $\pmb{p}_i$ only require that some sense of distance between the 'input' points $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$.
-Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
+The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$.
+We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$ Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
 \begin{align}
@@
 
-To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where an output embedding from a dimension reduction algorithm is available, requires an estimate of the Riemannian metric.
+To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric in the coordinate system.
 Formally, the Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
@@
 To alleviate this issue @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$ at each data point.
-These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
-
-Locally isometric embedding to go here...
+These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
 
@@
 
-As pointed out by @Perrault-Joncas2013-pq, if the dimension reduction is carried out such that the dimension of the output vectors is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned.
-
-To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process.
-@Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl].
-By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
-An *R* library named *intRinsic* [@Denti2021-qc] has implemented this algorithm and we are going to apply it to when $d$ is unknown.
+As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension $d$.
+In this cae the ranks of the matrices $\pmb{H}_i$ are equal to $d$.
+Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension $d$ globally into $d$-dimensional Euclidean space.
+In our simulated examples we abstract from this issue by constructing examples that can be globally embedded into $d$ dimesional Euclidean space.
+In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used.
+The *R* library *intRinsic* [@Denti2021-qc] implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
 
@@
 
-To come...
-
-# Kernel density estimate on Riemannian manifold embedding {#kderm}
-
-<!-- In this section, we introduce the proposed kernel density estimator on the embedding of Riemannian manifolds.  -->
-For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data.
-@Pelletier2005-vu generalize the multivariate kernel density estimator to Riemannian manifolds.
-However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space.
-Therefore, we propose a distortion-corrected kernel density estimator as in Equation \@ref(eq:denestimator0) based on the low-dimensional manifold learning embedding, $\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, of the high-dimensional data,
-\begin{equation}
-\label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
-\end{equation}
-where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise adaptive Riemannian matrix of dimension $d \times d$.
-The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point.
-These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
-
-<!-- to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings.
-@Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric which is a positive definite square matrix for each data point.
-To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric.
-The outliers could then be defined as the points with the lowest density estimates.
--->
-
-To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds.
-Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from @Perrault-Joncas2013-pq.
-Finally, we show how to use these densities to carry out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
-
-<!-- ## Notations -->
-
-<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathbb{R}^s$.
-The underlying manifold $M$ has the true density $f$.
-Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathbb{R}^d$ and $d \ll s$.
-The Riemannian metric of the embedding is denoted as $H_i, i = 1, 2, \dots, N$.
-In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$.
--->
-
-## Multivariate kernel density estimation {#mvkde}
-
-Kernel density estimation is one of the most popular methods to calculate the probability density function from a sample data.
-@Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
-In general, given multivariate random samples $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
-<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathbb{R}^d$, $i = 1, 2, \dots, N$  -->
-drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d$ is given by
-\begin{equation}
-\label{eq:vkde}
-\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
-\end{equation}
-where $K(\pmb{x})$ is the kernel function, $\pmb{H}_i$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})$.
-When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
-\begin{equation}
-\label{eq:kernel}
-K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
-\end{equation} 
-with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
-Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function.
-Then an average over all data points is taken as the estimated density $\hat{f}$.
-KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point $\pmb{x}$.
-
-The choice of the bandwidth matrix $\pmb{H}_i$ is crucial in the accuracy of $\hat{f}$ because it controls the smoothing across the sample data.
-For a given kernel function $K(\pmb{x})$, the shape of the local smoothing at each point is determined.
-Then the performance of the density estimate $\hat{f}$ depends on the choice of $\pmb{H}_i$.
-When $\pmb{H}_i$ is too small, the density estimates will apprear to be wiggly.
-On the contrary, when $\pmb{H}_i$ is too large, the density estimates might smooth out some local features of $f$.
-So the accuracy of $\hat{f}$ depends strongly on the bandwidth matrix.
-Compared to a global fixed $\pmb{H}$, a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
-
-<!-- Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
-The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
-In this paper, we focus on the multivariate kernel density estimation.
- -->
-
-<!-- Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points.
-However, when the local data structure is not universal for all sample data, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-The bandwidth is varied depending on either the location of the sample points or that of the estimated points [Section 6.6 of @Scott2015-vl].
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point $\pmb{X}_i$ and will refer to this as the *variable/adaptive kernel density estimation(VKDE)* unless otherwise stated.
- -->
-<!-- In multivariate VKDE, the estimator in Equation \@ref(eq:vkde) is similar to that in \@ref(eq:kde) except that the bandwidth is a variable matrix $\pmb{H}_i$ for each $\pmb{X}_i$,
-\begin{equation}
-\label{eq:vkde}
-\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
-\end{equation} -->
-
-For computational purposes, @Duong2007-up implements the kernel density estimates for data from 1 to 6 dimensions in the R package *ks*. 
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the *ks* implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
-More importantly, the estimator in Equation \@ref(eq:vkde) is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the manifold has a differentiable structure, which has to be taken into account when considering the neighborhood of a sample point.
-Therefore, we need to consider the probability density estimation on a Riemannian manifold and adapts the techiniques on Euclidean samples to a non-Euclidean setting.
-This is discussed in \autoref{riemest}.
-
-## Kernel density estimator on Riemannian manifolds {#riemest}
-
-Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold.
-Assume $(M, d_g)$ as a complete metric space, where $d_g$ is the geodesic distance induced by the Riemannian metric $g$ and the strictly positive injectivity radius of the manifold $inj_g(M)$ [@Chavel2006-mp].
-Denote $\pmb{X}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, as i.i.d.
-random objects on $M$ with density $f$.
-To estimate the density of the manifold, @Pelletier2005-vu propose a kernel density estimator for $f$ at point $\pmb{p}\in M$ as
-\begin{equation}
-\label{eq:denriem}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
-\end{equation}
-where $r$ is the global bandwidth satisfying $r \leq r_0$ for fixed $r_0$ such that $0\leq r_0\leq inj_g(M)$, $K: \mathbb{R}_{+} \rightarrow \mathbb{R}$ is a non-negative map, and $\theta_{\pmb{X}_i}(\pmb{p})$ is the volume density function on the manifold for point $\pmb{X}_i$.
-The estimator in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator [@Henry2009-ll].
-
-Denote $\lambda$ as the Lebesque measure of $\mathbb{R}^d$.
-The kernels $K$ in \@ref(eq:denriem) are functions of the geodesic distance of the manifold $d_g$ and they satisfy the following conditions [@Pelletier2005-vu]:
-\begin{align}
-\label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
-(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
-(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
-& (iv) \supp K=[0; 1];
-(v) \sup K(x)=K(0).
-\end{align}
-Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, $r\leq inj_g(M)$.
-Under such condition, for each $\pmb{p} \in M$, there exists a normal coordinate neighborhood of $\pmb{p}$ containing an open ball $B(\pmb{p}, r) \in T_{\pmb{p}}M$ centered at $\pmb{p}$ with radius $r$.
-
-Now consider the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$ in the estimator.
-For any tangent vector $u \in T_{\pmb{p}}M$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ denote a curve passing through $u$, and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$ for all $i=1,\dots,d$ and $\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}$.
-Then $\frac{Dw_i}{dt}(0)$ forms an orthonomal basis of the tangent space $T_{\pmb{p}}M$.
-The volume density function on $M$, $\theta_{\pmb{p}}(u): T_{\pmb{p}}M \rightarrow \mathbb{R}$ is defined in @Le_Brigant2019-lj as 
-\begin{equation}
-\label{eq:volden}
-\theta_{\pmb{p}}(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-\end{equation}
-
-For the exponential map $\exp_{\pmb{p}}: T_{\pmb{p}}M \rightarrow M$ induced by pullback of a volume form $\exp_{\pmb{p}}^*\textit{Vol}$ on $T_{\pmb{p}}M$, $\theta_{\pmb{p}}$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_{\pmb{p}}M$, which gives
-$$
-d\exp_{\pmb{p}}^*\textit{Vo}l(u) = \theta_{\pmb{p}}(u)du,
-$$
-in normal coordinates [@Le_Brigant2019-lj].
-The volume density function is motivated by how a measure $du$ changes under the change of coordinates.
-The volume form is variant to the exponential map except that the manifold is flat.
-Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1.
-It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
-
-@Pelletier2005-vu mention that the volume density function $\theta_{\pmb{p}}(\pmb{q})$ defined for $\pmb{q}$ in a neighborhood of $\pmb{p}$ on the manifold, with the geodesic normal coordinates at $\pmb{p}$, is equal to the determinant of the Riemannian metric $g$ expressed in the coordinates at the logarithm map of $\pmb{q}$, $\exp_{\pmb{p}}^{-1}(\pmb{q})$.
-Note that this is true only when using the geodesic normal coordinates.
-If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise.
-Therefore, the Riemannian manifold density estimator in @Pelletier2005-vu only works when the true manifold $M$ is known.
-Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.
-
-## Distortion-corrected kernel density estimator for manifold embedding {#dckde}
-
-In application, we usually find the observations lying on a $d$-dimensional manifold embedded in a much higher $s$-dimensional space, where $d \ll s$.
-Instead of sampling from the underlying manifold $M$, we could sample from the high-dimensional data only and apply dimension reduction techniques such as manifold learning to find a $d$-dimensional representation of $M$.
-Notice that this representation is not necessarily isometry to $M$, and the underlying manifold tends to be unknown in real data sets.
-We could use the low-dimensional embedding as an approximation of the true manifold $M$.
-However, the pointwise distortion induced by the manifold learning process needs to be taken care of to accurately estimate the density of the manifold.
-Therefore, the kernel density estimator in Equation \@ref(eq:denriem) no longer applies because of the unknown geodesic distance $d_g$ and the volume density function $\theta_{\pmb{X}_i}(\pmb{p})$.
-Further, the bandwidth matrix should be adaptive to each sample point to take the distortion into account.
-
-Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold learning embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
+With all fundamentals introduced we can now give our novel Distortion Corrected KDE (DC-KDE) as
+
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||}{r} \bigg).
 \end{equation}
-where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$
-<!-- , and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1 -->
-.
-We will now state below how to calculate each term in the estimator.
-And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
-
-By the definition of the volume density function in Equation \@ref(eq:volden), we could calculate how the volume change under the change of coordinates.
-When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.
-geodesic normal coordinates.
-In this case, @Pelletier2005-vu gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, $|\det g(\exp^{-1}_{\pmb{p}}(\pmb{q})) |$.
-In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
-Since the logarithm map of the manifold is also unknown, the volume density function in @Pelletier2005-vu is no longer valid.
-Instead, we need a volume density function independent of the chart.
-
-For $\pmb{p} \in M$, denote an open ball centered at $\pmb{p}$ with radius $r$ as $B(\pmb{p},r)$ and the open neighborhood of $0_{\pmb{p}} \in T_{\pmb{p}}M$ can be denoted as $B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))$.
-Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_{\pmb{p}},r)$, with $g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_{\pmb{p}}$ in $B(0_{\pmb{p}},r)$.
-Let $u \in B(0_{\pmb{p}},r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$, we have $u \in \bar{U}$. 
-The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$.
-The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om].
-Therefore, given a point $\pmb{q} \in B(\pmb{p},r)$ and a tangent vector $u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_{\pmb{p}},r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
-\begin{equation}
-\label{eq:volumedensityratio}
-\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }  .
-\end{equation}
-The volume density function is only defined for points satisfying $d_g(\pmb{p},\pmb{q}) < inj_gM$ so that there exists an open ball $B(\pmb{p},r) \in T_{\pmb{p}}M$.
-And $\theta_{\pmb{p}}(\pmb{q})=1$ for all $\pmb{p},\pmb{q} \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.
-
-Now we consider the case with the manifold learning embedding.
-Consider the chart $(U, \psi)$ induced by an orthonormal basis of the tangent space of the embedding $T_{\pmb{y}_i}E$ and a normal neighborhood of $\pmb{y}_i$.
-Given a query point $\pmb{p} \in B(\pmb{y}_i,r)$, the volume of the parallelepiped is spanned by the same basis.
-Then the volume density function is given by 
-\begin{equation}
-\label{eq:volumedensityest}
-\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
-\end{equation} 
-where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in Condition (v) of \@ref(eq:kernelcondition).
-
-As for the geodesic distance $d_g$ in \@ref(eq:denriem), we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
-\begin{equation}
-\label{eq:dgestimate}
-\hat{d}_g(\pmb{p}, \pmb{y}_i) = \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i),
-\end{equation}
-where $\pmb{H}(\pmb{y}_i)$ is the estimated Riemannian metric at embedding point $\pmb{y}_i$.
-
-For two points $\pmb{p},\pmb{q} \in M$, the geodesic distance between $\pmb{p}$ and $\pmb{q}$ is defined as the length of the shortest curve from $\pmb{p}$ to $\pmb{q}$ along the manifold.
-The geodesic distance is intrinsic, which means it is a coordinate-independent quantity.
-The Riemannian metric functions as the correction term for different coordinates of manifold learning embedding.
-By using the first-order distance estimator, $\hat{d}_g(\pmb{p}, \pmb{y}_i)$ serves as a good approximation of the true geodesic distance $d_g$ with the distortion corrected at each embedding point $\pmb{y}_i$.
-Compared with the graph distance, the shortest path between point in the neighborhood graph $G$, our proposed $\hat{d}_g$ is better because we use the same graph $G$ for all manifold learning embeddings but the distances between different nodes in $G$ might not be the same due to the distortion.
-
-<!-- As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|\pmb{p}-\pmb{q}\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality.
-By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric.
-In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1.
-The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension.
-We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and the estimated point $\pmb{p}$ so that for $\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}$, we have -->
-<!-- \begin{equation} -->
-<!-- \label{eq:vbs} -->
-<!-- C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i), -->
-<!-- \end{equation} -->
-<!-- where $\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))$.
-The distortions are corrected for the tangent plane of the embedding points.
-Since we aim to estimate the density of the manifold for the embedding of sample points, $\pmb{p}$ is equal to $\pmb{y}_i$, for $i=1,\dots,N$.
-Therefore, we have -->
-<!-- \begin{equation} -->
-<!-- \label{eq:vbsyi} -->
-<!-- C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i). -->
-<!-- \end{equation} -->
-
-Now we have all the components needed to estimate the density of manifold learning embedding in Equation \@ref(eq:denestimator) except for the Riemannian metric.
-The Riemannian metric is a key to correct the distortion from the manifold learning algorithms and we discuss its estimation in \autoref{riemmetric}.
-
-## Riemannian metric estimation {#riemmetric}
-
-An important goal of manifold learning is to recover the local or global features of the data. However, manifold learning algorithms usually fail to generate an isometric embedding.
-In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimated Riemannian metric.
-Instead of developing a new manifold learning method, they aim to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix variable at each data point.
-The estimated Riemannian metric completes our proposed kernel density estimation in \autoref{dckde} and enables us to take the geometric distortion into consideration when estimating the densities of the manifold.
-
-The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-For each point $\pmb{p} \in M$ in any embedding coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.
-
-The inner product between two vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
-$$
-l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-$$ 
-where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
-$$
-Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
-$$ 
-Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).
-
-[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.
-
-@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any manifold learning embedding with the estimated Riemannian metric.
-There are mainly four steps in the algorithm: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
-We restate these four steps in Algorithm \@ref(alg:learnmetric).
-By applying an existing manifold learning algorithm to the high-dimensional input data $\pmb{X}\in \mathbb{R}^s$ with $N$ observations, a low-dimensional embedding $\pmb{Y} \in \mathbb{R}^d$ can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\hat{\mathcal{L}}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
-The graph Laplacian estimator together with the embedding $\pmb{Y}$ could further be used to achieve the augmented embedding with the Riemannian metric, $(\pmb{Y}, \pmb{H})$.
-This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz].
-It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
-We have re-implemented the algorithm in *R*.
-
-As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned.
-To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process.
-@Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl].
-By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
-An *R* library named *intRinsic* [@Denti2021-qc] has implemented this algorithm and we are going to apply it to when $d$ is unknown.
-
-\begin{algorithm}[!htb]
-  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
-  \label{alg:learnmetric}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters EMBED }
-  \BlankLine
-  \begin{algorithmic}[1]
-
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
-
-  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
-  $$
-  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
-  $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
-
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
-
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
-  $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
-  $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
-  $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
-  $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
-
-  \end{algorithmic}
-\end{algorithm}
-
-## Outlier detection on manifold learning embedding
-
-Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \@ref(eq:denestimator).
-The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection.
-As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \@ref(alg:kdeanomaly).
-There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathbb{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathbb{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation.
-Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels.
-Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator).
-The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f(\pmb{y}_i)$.
-
-\begin{algorithm}[!htb]
-  \caption{Anomaly detection with distortion-corrected kernel density estimates}
-  \label{alg:kdeanomaly}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
-  \Input{ high-dimensional data $\pmb{x}_i$ for all $i=1,\dots,N$ }
-  \Output{ outliers embedding coordinates $\pmb{y}_1, \dots, \pmb{y}_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
-  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
-  \BlankLine
-  \begin{algorithmic}[1]
-  \STATE Estimate the intrinsic dimension $d$ with the TWO-NN algorithm if $d$ is not given;
-  \STATE For all $i=1,\dots,N$, compute the $d$-dimensional embeddings $\pmb{y}_i$ with any existing manifold learning algorithms and the corresponding Riemannian metric $\pmb{H}(\pmb{y}_i)$ using the Learn Metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-  \STATE Compute the kernel density estimates $\hat{f}(\pmb{y}_i)$ for all $i=1,\dots,N$ using the DC\_KDE in \ref{eq:denestimator};
-  \STATE Reorder the embedding coordinates $\pmb{Y}$ from smallest to largest according to the density estimates $\hat{f}$ and subset the top \textit{n\_outliers} points as the outliers.
-  \end{algorithmic}
-\end{algorithm}
-
-If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding plot and find which observations lie in the highest density region of specified coverage, eg.
-1%, 50%, 99%, \>99%.
-
-The proposed schematic is shown in Figure \@ref(fig:vkde).
+
+The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
+To understand these differences it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length is section 6.2 of @Perrault-Joncas2013-pq).
+This is crucial for two reasons.
+First, this implies that the term $|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||$ approximates the geodesic distance between $\pmb{p}_i$ and $\pmb{p}_j$.
+Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\pmb{p}_i$, it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
+For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm) while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
+Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
+
+One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\pmb{H}_j$.
+To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required, since anomalies are identified as the points with lowest density.
+The entire workflow is summarised in Figure \@ref(fig:vkde).
 The last two steps in Figure \@ref(fig:vkde) are the main contributions of Algorithm \@ref(alg:kdeanomaly), generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.

### 2022-10-24 paper/paper.tex
completed section 2 and deleted old section 2

@@
 However this need not be that case, for instance the \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each \(\pmb{p}_i\) only require that some sense of distance between the `input' points \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\).
-Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
+The methods we propose for estimating the density at each \(\pmb{p}_i\) only require some sense of distance between the `input' points \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\).
+We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\) Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
 \begin{align}
@@
 
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where an output embedding from a dimension reduction algorithm is available, requires an estimate of the Riemannian metric.
+To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric in the coordinate system.
 Formally, the Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
@@
 To alleviate this issue \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\) at each data point.
-These matrices estimate the Reimannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).
-
-Locally isometric embedding to go here\ldots{}
+These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends (up to a first order approximation) on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).
 
@@
 
-As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(\pmb{H}(p)\) is equal to \(d\); otherwise, the Riemannian metric will be returned.
-To ensure that the estimated Riemannian metric is returned as a \(d\times d\) matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} to derive id estimator from a Poisson point process.
-\textcite{Facco2017-rl} point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter \(1\) and shape parameter \(d\), the intrinsic dimension of the manifold \autocite{Facco2017-rl}.
-By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
-An \emph{R} library named \emph{intRinsic} \autocite{Denti2021-qc} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.
+As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
+In this cae the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\).
+Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean space.
+In our simulated examples we abstract from this issue by constructing examples that can be globally embedded into \(d\) dimesional Euclidean space.
+In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used.
+The \emph{R} library \emph{intRinsic} \autocite{Denti2021-qc} implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
 
@@
 
-To come\ldots{}
-
-\hypertarget{kderm}{%
-\section{Kernel density estimate on Riemannian manifold embedding}\label{kderm}}
-
-For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data.
-\textcite{Pelletier2005-vu} generalize the multivariate kernel density estimator to Riemannian manifolds.
-However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space.
-Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional manifold learning embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data,
-\begin{equation}
-\label{eq:denestimator0}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
-\end{equation}
-where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
-The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point.
-These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.
-
-To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds.
-Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from \textcite{Perrault-Joncas2013-pq}.
-Finally, we show how to use these densities to carry out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.
-
-\hypertarget{mvkde}{%
-\subsection{Multivariate kernel density estimation}\label{mvkde}}
-
-Kernel density estimation is one of the most popular methods to calculate the probability density function from a sample data.
-\textcite{Scott2015-vl} covers a wide range of techniques for estimating the kernel densities of multivariate data.
-In general, given multivariate random samples \(\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d\)
-drawn from a density \(f\), the variable kernel density estimate \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut} of \(f\) at point \(\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d\) is given by
-\begin{equation}
-\label{eq:vkde}
-\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
-\end{equation}
-where \(K(\pmb{x})\) is the kernel function, \(\pmb{H}_i\) is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension \(d\times d\), and \(K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})\).
-When using the standard gaussian kernel, we have \(K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})\) and hence
-\begin{equation}
-\label{eq:kernel}
-K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
-\end{equation}
-with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).
-Intuitively, KDE smooths out each data point \(\pmb{X}_i\) into a smooth bump with the shape determined by the kernel function.
-Then an average over all data points is taken as the estimated density \(\hat{f}\).
-KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point \(\pmb{x}\).
-
-The choice of the bandwidth matrix \(\pmb{H}_i\) is crucial in the accuracy of \(\hat{f}\) because it controls the smoothing across the sample data.
-For a given kernel function \(K(\pmb{x})\), the shape of the local smoothing at each point is determined.
-Then the performance of the density estimate \(\hat{f}\) depends on the choice of \(\pmb{H}_i\).
-When \(\pmb{H}_i\) is too small, the density estimates will apprear to be wiggly.
-On the contrary, when \(\pmb{H}_i\) is too large, the density estimates might smooth out some local features of \(f\).
-So the accuracy of \(\hat{f}\) depends strongly on the bandwidth matrix.
-Compared to a global fixed \(\pmb{H}\), a variable bandwidth matrix is able to account for the pointwise-adaptive local structure and generate an estimate more applicable to the case where the distortion at each point is different in the low-dimensional manifold learning embedding.
-
-For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
-However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
-More importantly, the estimator in Equation \eqref{eq:vkde} is only applicable when the sample space is Euclidean.
-In the cases of a Riemannian manifold, the manifold has a differentiable structure, which has to be taken into account when considering the neighborhood of a sample point.
-Therefore, we need to consider the probability density estimation on a Riemannian manifold and adapts the techiniques on Euclidean samples to a non-Euclidean setting.
-This is discussed in \autoref{riemest}.
-
-\hypertarget{riemest}{%
-\subsection{Kernel density estimator on Riemannian manifolds}\label{riemest}}
-
-Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold.
-Assume \((M, d_g)\) as a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}.
-Denote \(\pmb{X}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), as i.i.d.
-random objects on \(M\) with density \(f\).
-To estimate the density of the manifold, \textcite{Pelletier2005-vu} propose a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
-\begin{equation}
-\label{eq:denriem}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
-\end{equation}
-where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathbb{R}_{+} \rightarrow \mathbb{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\).
-The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.
-
-Denote \(\lambda\) as the Lebesque measure of \(\mathbb{R}^d\).
-The kernels \(K\) in \eqref{eq:denriem} are functions of the geodesic distance of the manifold \(d_g\) and they satisfy the following conditions \autocite{Pelletier2005-vu}:
-\begin{align}
-\label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
-(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
-(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
-& (iv) \supp K=[0; 1];
-(v) \sup K(x)=K(0).
-\end{align}
-Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, \(r\leq inj_g(M)\).
-Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_{\pmb{p}}M\) centered at \(\pmb{p}\) with radius \(r\).
-
-Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
-For any tangent vector \(u \in T_{\pmb{p}}M\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\).
-Then \(\frac{Dw_i}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_{\pmb{p}}M\).
-The volume density function on \(M\), \(\theta_{\pmb{p}}(u): T_{\pmb{p}}M \rightarrow \mathbb{R}\) is defined in \textcite{Le_Brigant2019-lj} as
-\begin{equation}
-\label{eq:volden}
-\theta_{\pmb{p}}(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
-\end{equation}
-
-For the exponential map \(\exp_{\pmb{p}}: T_{\pmb{p}}M \rightarrow M\) induced by pullback of a volume form \(\exp_{\pmb{p}}^*\textit{Vol}\) on \(T_{\pmb{p}}M\), \(\theta_{\pmb{p}}\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_{\pmb{p}}M\), which gives
-\[
-d\exp_{\pmb{p}}^*\textit{Vo}l(u) = \theta_{\pmb{p}}(u)du,
-\]
-in normal coordinates \autocite{Le_Brigant2019-lj}.
-The volume density function is motivated by how a measure \(du\) changes under the change of coordinates.
-The volume form is variant to the exponential map except that the manifold is flat.
-Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1.
-It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.
-
-\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_{\pmb{p}}(\pmb{q})\) defined for \(\pmb{q}\) in a neighborhood of \(\pmb{p}\) on the manifold, with the geodesic normal coordinates at \(\pmb{p}\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of \(\pmb{q}\), \(\exp_{\pmb{p}}^{-1}(\pmb{q})\).
-Note that this is true only when using the geodesic normal coordinates.
-If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise.
-Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
-Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.
-
-\hypertarget{dckde}{%
-\subsection{Distortion-corrected kernel density estimator for manifold embedding}\label{dckde}}
-
-In application, we usually find the observations lying on a \(d\)-dimensional manifold embedded in a much higher \(s\)-dimensional space, where \(d \ll s\).
-Instead of sampling from the underlying manifold \(M\), we could sample from the high-dimensional data only and apply dimension reduction techniques such as manifold learning to find a \(d\)-dimensional representation of \(M\).
-Notice that this representation is not necessarily isometry to \(M\), and the underlying manifold tends to be unknown in real data sets.
-We could use the low-dimensional embedding as an approximation of the true manifold \(M\).
-However, the pointwise distortion induced by the manifold learning process needs to be taken care of to accurately estimate the density of the manifold.
-Therefore, the kernel density estimator in Equation \eqref{eq:denriem} no longer applies because of the unknown geodesic distance \(d_g\) and the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\).
-Further, the bandwidth matrix should be adaptive to each sample point to take the distortion into account.
-
-Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold learning embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
+With all fundamentals introduced we can now give our novel Distortion Corrected KDE (DC-KDE) as
+
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
+\hat{f}(\pmb{p}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||}{r} \bigg).
 \end{equation}
-where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\)
-.
-We will now state below how to calculate each term in the estimator.
-And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
-
-By the definition of the volume density function in Equation \eqref{eq:volden}, we could calculate how the volume change under the change of coordinates.
-When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates.
-In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_{\pmb{p}}(\pmb{q})) |\).
-In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding.
-Since the logarithm map of the manifold is also unknown, the volume density function in \textcite{Pelletier2005-vu} is no longer valid.
-Instead, we need a volume density function independent of the chart.
-
-For \(\pmb{p} \in M\), denote an open ball centered at \(\pmb{p}\) with radius \(r\) as \(B(\pmb{p},r)\) and the open neighborhood of \(0_{\pmb{p}} \in T_{\pmb{p}}M\) can be denoted as \(B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))\).
-Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_{\pmb{p}},r)\), with \(g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_{\pmb{p}}\) in \(B(0_{\pmb{p}},r)\).
-Let \(u \in B(0_{\pmb{p}},r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_{\pmb{p}},r)\), we have \(u \in \bar{U}\).
-The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\).
-The quotient of these two volumes is then independent of the chart \((\bar{U}, \bar{\psi})\) \autocite{Henry2009-om}.
-Therefore, given a point \(\pmb{q} \in B(\pmb{p},r)\) and a tangent vector \(u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_{\pmb{p}},r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
-\begin{equation}
-\label{eq:volumedensityratio}
-\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }  .
-\end{equation}
-The volume density function is only defined for points satisfying \(d_g(\pmb{p},\pmb{q}) < inj_gM\) so that there exists an open ball \(B(\pmb{p},r) \in T_{\pmb{p}}M\).
-And \(\theta_{\pmb{p}}(\pmb{q})=1\) for all \(\pmb{p},\pmb{q} \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.
-
-Now we consider the case with the manifold learning embedding.
-Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\).
-Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.
-Then the volume density function is given by
-\begin{equation}
-\label{eq:volumedensityest}
-\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
-\end{equation}
-where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
-Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in Condition (v) of \eqref{eq:kernelcondition}.
-
-As for the geodesic distance \(d_g\) in \eqref{eq:denriem}, we propose a distortion-fixed geodesic distance with the use of estimated Riemannian metric as
-\begin{equation}
-\label{eq:dgestimate}
-\hat{d}_g(\pmb{p}, \pmb{y}_i) = \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i),
-\end{equation}
-where \(\pmb{H}(\pmb{y}_i)\) is the estimated Riemannian metric at embedding point \(\pmb{y}_i\).
-
-For two points \(\pmb{p},\pmb{q} \in M\), the geodesic distance between \(\pmb{p}\) and \(\pmb{q}\) is defined as the length of the shortest curve from \(\pmb{p}\) to \(\pmb{q}\) along the manifold.
-The geodesic distance is intrinsic, which means it is a coordinate-independent quantity.
-The Riemannian metric functions as the correction term for different coordinates of manifold learning embedding.
-By using the first-order distance estimator, \(\hat{d}_g(\pmb{p}, \pmb{y}_i)\) serves as a good approximation of the true geodesic distance \(d_g\) with the distortion corrected at each embedding point \(\pmb{y}_i\).
-Compared with the graph distance, the shortest path between point in the neighborhood graph \(G\), our proposed \(\hat{d}_g\) is better because we use the same graph \(G\) for all manifold learning embeddings but the distances between different nodes in \(G\) might not be the same due to the distortion.
-
-Now we have all the components needed to estimate the density of manifold learning embedding in Equation \eqref{eq:denestimator} except for the Riemannian metric.
-The Riemannian metric is a key to correct the distortion from the manifold learning algorithms and we discuss its estimation in \autoref{riemmetric}.
-
-\hypertarget{riemmetric}{%
-\subsection{Riemannian metric estimation}\label{riemmetric}}
-
-An important goal of manifold learning is to recover the local or global features of the data. However, manifold learning algorithms usually fail to generate an isometric embedding.
-In order to recover the geometry of the manifold, \textcite{Perrault-Joncas2013-pq} propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimated Riemannian metric.
-Instead of developing a new manifold learning method, they aim to fix the geometric distortion in the embedding.
-It is worth noticing that the Riemannian metric from \textcite{Perrault-Joncas2013-pq} is a \(d\times d\) positive definite matrix variable at each data point.
-The estimated Riemannian metric completes our proposed kernel density estimation in \autoref{dckde} and enables us to take the geometric distortion into consideration when estimating the densities of the manifold.
-
-The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
-If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
-For each point \(\pmb{p} \in M\) in any embedding coordinate system, the Riemannian metric \(g\) is a \(d\times d\) symmetric positive definite matrix, where \(d\) is the dimension of the manifold.
-
-The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
-The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
-\[
-l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
-\]
-where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function that maps \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
-\[
-Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
-\]
-Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \eqref{eq:volumedensityratio}, \eqref{eq:volumedensityest}, \eqref{eq:dgestimate}, and \eqref{eq:vbs}.
-
-\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any manifold learning embedding with the estimated Riemannian metric.
-There are mainly four steps in the algorithm: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation.
-We restate these four steps in Algorithm \ref{alg:learnmetric}.
-By applying an existing manifold learning algorithm to the high-dimensional input data \(\pmb{X}\in \mathbb{R}^s\) with \(N\) observations, a low-dimensional embedding \(\pmb{Y} \in \mathbb{R}^d\) can be computed.
-Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator \(\Delta_M\) is built.
-The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
-\textcite{Perrault-Joncas2013-pq} have stated one way to compute the approximated \(\Delta_M\) with a discrete consistent estimator, the geometric graph Laplacian \(\hat{\mathcal{L}}_{\varepsilon,n}\) \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-The graph Laplacian estimator together with the embedding \(\pmb{Y}\) could further be used to achieve the augmented embedding with the Riemannian metric, \((\pmb{Y}, \pmb{H})\).
-This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}.
-It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.
-We have re-implemented the algorithm in \emph{R}.
-
-As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(\pmb{H}(p)\) is equal to \(d\); otherwise, the Riemannian metric will be returned.
-To ensure that the estimated Riemannian metric is returned as a \(d\times d\) matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} to derive id estimator from a Poisson point process.
-\textcite{Facco2017-rl} point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter \(1\) and shape parameter \(d\), the intrinsic dimension of the manifold \autocite{Facco2017-rl}.
-By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated.
-An \emph{R} library named \emph{intRinsic} \autocite{Denti2021-qc} has implemented this algorithm and we are going to apply it to when \(d\) is unknown.
-
-\begin{algorithm}[!htb]
-  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
-  \label{alg:learnmetric}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters EMBED }
-  \BlankLine
-  \begin{algorithmic}[1]
-
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
-
-  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
-  $$
-  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
-  $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
-
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
-
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
-  $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
-  $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
-  $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
-  $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
-
-  \end{algorithmic}
-\end{algorithm}
-
-\hypertarget{outlier-detection-on-manifold-learning-embedding}{%
-\subsection{Outlier detection on manifold learning embedding}\label{outlier-detection-on-manifold-learning-embedding}}
-
-Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}.
-The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection.
-As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \ref{alg:kdeanomaly}.
+
+The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
+To understand these differences it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length is section 6.2 of \textcite{Perrault-Joncas2013-pq}).
+This is crucial for two reasons.
+First, this implies that the term \(|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||\) approximates the geodesic distance between \(\pmb{p}_i\) and \(\pmb{p}_j\).
+Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{p}_i\), it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
+For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm) while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
+Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
+
+One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\pmb{H}_j\).
+To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.While density estimation on manifolds may be used in a number of ways, a particular downstream task that we will focus on will be anomaly detection in \ref{alg:kdeanomaly}.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required, since anomalies are identified as the points with lowest density.
+
 There are mainly four steps involved in the algorithm.

### 2022-10-24 R/Electricity2d_01embed_tvd.R
inRrinsic for tvd measure

@@ -12,15 +12,31 @@ library(igraph)
 library(matrixcalc)
 library(intRinsic)
 Jmisc::sourceAll("R/sources")
+scen <- 1
 # scen <- as.numeric(commandArgs()[[6]])
 r <- 1
 
-load("data/spdemand_3639id336tow.rda")
-# nid <- 1
+load("data/half_count_ratio_3639id336tow.rda")
 nid <- 3639
 ntow <- 336
-train <- readRDS(paste0("data/spdemand_", nid, "id", ntow, "tow_train.rds"))
-# intRinsic::twonn(train, method = c("mle", "linfit", "bayes")[scen])
+if(nid == 1) {
+  train <- spdemand %>%
+    lazy_dt() %>%
+    filter(tow <= ntow,
+           # id <= sort(unique(spdemand[,id]))[nid]
+           id == 1003
+    ) %>%
+    dplyr::select(-id, -tow) %>%
+    as.data.table()
+} else if(nid == 3639) {
+  train <- spdemand %>%
+    lazy_dt() %>%
+    # mutate(id_tow = paste0(id, "_", tow)) %>%
+    as.data.table() %>%
+    column_to_rownames("id")
+}
+dim(train)
+rm(spdemand)
 
 paste("Start at:", Sys.time())
 # ----parameters----------------------------------------------------------------
@@ -29,9 +45,9 @@ y <- NULL # without pre-computed embedding
 N <- nrow(x)
 s <- 2 # embedded in 2-D for visualization purpose
 k <- 100 # N / 50
-method <- "annIsomap"
+method <- c("annIsomap", "annLLE", "annLaplacianEigenmaps", "anntSNE", "annUMAP")[scen]
 annmethod <- "kdtree"
-distance <- c("euclidean", "manhattan")[2]
+distance <- c("euclidean", "manhattan")[2] # "manhattan" for all households
 treetype <- "kd"
 searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
 radius <- 10 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
@@ -54,8 +70,8 @@ Rn <- metric_isomap$rmetric
 adj_matrix <- metric_isomap$adj_matrix
 E1 <- fn[,1]; E2 <- fn[,2]
 
-print("True density summary statistics")
-summary(trueden)
+# print("True density summary statistics")
+# summary(trueden)
 
[…]

### 2022-10-24 R/Simulation4d_01datadensity.R

@@ -54,12 +54,12 @@ semisphere <- cbind(X,
 # label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
 # dist2center <- sqrt(radius^2 - semisphere[,5]^2)
 # ## Plot 5-d semi-sphere, run once
-library(tourr)
-colors <- colourvalues::colour_values(- dist2center, palette = "magma") # generate colors for locations
-pchs <- c(rep(16, 0.99*N), rep(17, 0.01*N)) # point shapes for kernels
-animate_xy(semisphere[,1:5], col = colors, pch = pchs, cex = 0.8,
-           axes = "bottomleft", fps=15
-           )
+# library(tourr)
+# colors <- colourvalues::colour_values(- dist2center, palette = "magma") # generate colors for locations
+# pchs <- c(rep(16, 0.99*N), rep(17, 0.01*N)) # point shapes for kernels
+# animate_xy(semisphere[,1:5], col = colors, pch = pchs, cex = 0.8,
+#            axes = "bottomleft", fps=15
+#            )
 # # "figures/tourr_5d_semisphere.png"
 
 y <- matrix(runif(p*p, 0, 1), p, p)
@@ -78,7 +78,7 @@ x <- train
 y <- X # given known embedding with known density
 # y <- NULL
 s <- 2
-k <- N / 20
+k <- 100 # N / 20
 method <- "Isomap" # not used because y is given
 annmethod <- "kdtree"
 distance <- "euclidean"

### 2022-10-24 R/Simulation4d_03plots.R

@@ -107,7 +107,7 @@ hdrtable %>%
 ## 3. Rank comparison plot, highlight outliers
 ###--------------------------------------------------------
 frank <- f %>% 
-  summarise_all(.fun = list(~ rank(x = ., ties.method = "random"))) %>% # random tie method for better visualization (not verticle lines in >99% region)
+  summarise_all(.fun = list(~ rank(x = ., ties.method = "average"))) %>% # random tie method for better visualization (not verticle lines in >99% region)
   mutate(outliers = factor(label, levels = c("TRUE", "FALSE")),
          dist2center = dist2center)
 
@@ -133,6 +133,7 @@ p_rank <- frank_long %>%
   scale_x_log10() + 
   scale_y_log10() + 
   theme(legend.position = "bottom")
+p_rank
 # ggsave(paste0("paper/figures/", "sim4d", N, "_density_comparison_4ml_radius", 10, "_k200_rankdensity_circleoutlier_with1rec.png"), p_rank, width = 8, height = 10, dpi = 300)
 
 

### 2022-10-24 R/electricity_embedding.R
inRrinsic for tvd measure

@@ -1,49 +1,65 @@
+rm(list = ls())
 library(tidyverse)
 library(dimRed)
 library(dtplyr)
+library(data.table)
 library(reticulate)
 library(viridis)
 # library(ks)
 library(hdrcde)
 library(igraph)
-library(seurat)
+# library(seurat)
 library(matrixcalc)
+library(intRinsic)
 Jmisc::sourceAll("R/sources")
+scen <- as.numeric(commandArgs()[[6]])
 
-load("data/spdemand_3639id336tow.rda")
-nid <- 1
+# load("data/spdemand_3639id336tow.rda")
+# nid <- 1
+nid <- 3639
 ntow <- 336
+train <- readRDS(paste0("data/spdemand_", nid, "id", ntow, "tow_train.rds"))
 
-train <- spdemand %>%
-  lazy_dt() %>%
-  filter(tow <= ntow,
-         # id <= sort(unique(spdemand[,id]))[nid],
-         id == 1003) %>%
-  dplyr::select(-id, -tow) %>%
-  data.table::as.data.table()
-
-# saveRDS(train, file = "data/spdemand_1id336tow_train.rds")
+# if(nid == 1) {
+#   train <- spdemand %>%
+#     lazy_dt() %>%
+#     filter(tow <= ntow,
+#            # id <= sort(unique(spdemand[,id]))[nid]
+#            id == 1003
+#     ) %>%
+#     dplyr::select(-id, -tow) %>%
+#     data.table::as.data.table()
+# } else if(nid == 3639) {
+#   train <- spdemand %>% 
+#     lazy_dt() %>%
+#     mutate(id_tow = paste0(id, "_", tow)) %>% 
+#     as.data.table() %>%
+#     column_to_rownames("id_tow") %>% 
+#     select(-c(id, tow))
+# }
+# # saveRDS(train, file = paste0("data/spdemand_", nid, "id", ntow, "tow_train.rds"))
 
 N <- nrow(train)
+# intRinsic::twonn(train, method = c("mle", "linfit", "bayes")[scen]) # 6 for 1ID
 
 # Parameters fixed
 x <- train
 s <- 2
-k <- 20
+k <- 100
 method <- "annIsomap"
 annmethod <- "kdtree"
-distance <- "euclidean"
+distance <- c("euclidean", "manhattan")[2]
 treetype = "kd"
 searchtype = "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
-radius <- .4
+radius <- 1
[…]

### 2022-10-24 R/TwinPeaks.R
Update

@@ -129,9 +129,6 @@ fmeta <- vkde(x = fn, h = h_hdr_meta, vh = Rn, r = r, gridsize = gridsize, eval.
 p_meta <- plot_outlier(x = metric_meta, gridsize = gridsize, prob = prob, riem.scale = riem.scale, f = fmeta, ell.size = 0)
 p_meta$densities %>% summary()
 
-## TODO: save computed results, beautify plots, remove same y axis, scale_x_log to show low density area, edit legend to one line, change color in rank plot to categorized HDR colors
-
-
 # Rank correlation
 f <- tibble(fxy = trueden, # true densities
             fxy_dckde = p_meta$densities, 

### 2022-10-24 R/Electricity_00_intrinsic.R
Rename

@@ -1,0 +1,36 @@
+rm(list = ls())
+library(tidyverse)
+library(dimRed)
+library(dtplyr)
+library(data.table)
+library(intRinsic)
+Jmisc::sourceAll("R/sources")
+scen <- as.numeric(commandArgs()[[6]])
+method <- c("mle", "linfit", "bayes")[scen]
+
+load("data/half_count_ratio_3639id336tow.rda")
+nid <- 3639
+ntow <- 336
+
+if(nid == 1) {
+  train <- spdemand %>%
+    lazy_dt() %>%
+    filter(tow <= ntow,
+           # id <= sort(unique(spdemand[,id]))[nid]
+           id == 1003
+    ) %>%
+    dplyr::select(-id, -tow) %>%
+    as.data.table()
+} else if(nid == 3639) {
+  train <- spdemand %>%
+    lazy_dt() %>%
+    # mutate(id_tow = paste0(id, "_", tow)) %>%
+    as.data.table() %>%
+    column_to_rownames("id")
+}
+dim(train)
+# # saveRDS(train, file = paste0("data/spdemand_", nid, "id", ntow, "tow_train.rds"))
+
+N <- nrow(train)
+d <- intRinsic::twonn(train, method = method) # 6 for 1ID, 7 for all spdemand with 3639*336 rows, 91 for 3639 rows
+print(d)

### 2022-10-24 R/sources/dr_quality.R

@@ -40,7 +40,7 @@ dr_quality <- function(X, Y, K = 20, nn.idx, treetype = "kd", searchtype = "stan
   # Rnx <- dimRed::R_NX(e)
   # mean_Rnx <- dimRed::quality(e, "mean_R_NX")
   # auc_lnK_Rnx <- dimRed::quality(e, "AUC_lnK_R_NX")
-  # suppressMessages(procrustes <- proc.measure(X, Y, K, nn.idx, treetype = treetype, searchtype = searchtype)$R)
+  suppressMessages(procrustes <- proc.measure(X, Y, K, nn.idx, treetype = treetype, searchtype = searchtype)$R)
   
   ## UNCOMMENT IF PROCRUSTES MEASURE IS NEEDED
   # procrustes <- proc.measure(X, Y, K, nn.idx, treetype = treetype, searchtype = searchtype)$R_N # standerdized procrustes measure R_N, unstanderdized $R
@@ -52,7 +52,7 @@ dr_quality <- function(X, Y, K = 20, nn.idx, treetype = "kd", searchtype = "stan
                 # K = K_optim, 
                                  M_T=M_T, M_C=M_C, LCMC=lcmc, Qnx=Qnx, 
                                  W_n=W_n, W_nu=W_nu, 
-                                 # Procrustes=procrustes,
+                                 Procrustes=procrustes,
                                  # Qlocal=Qlocal, Qglobal=Qglobal,
                                  Rnx=Rnx, 
                                  # mean_Rnx=mean_Rnx, auc_lnK_Rnx=auc_lnK_Rnx, 

### 2022-10-24 R/sources/annHLLE.R
Rename

@@ -6,6 +6,7 @@ annHLLE <- setClass(
                    eps = 0, 
                    ndim = 2,
                    get_geod = FALSE,
+                   annmethod = "kdtree", 
                    radius = 1,
                    nt = 50, 
                    nlinks = 16, 

### 2022-10-25 paper/paper.Rmd
writing section 3

@@
 In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
+For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional Euclidean space.
+General stuff here...
+
 We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
@@
 
-## 3-D mapping from a 2-D Gaussian Mixture Model {#twodgaussian}
+## Twin peaks example {#twodgaussian}
 
-We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with different means
+The simulation setup for the twin peaks example is to first generate vector $\pmb{v}_1,\dots,\pmb{v}_N$ for $N=2000$ from a 2-dimensional Gaussian mixture model.
+The mixture has four components with different means
 $\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4$.
 The mixture proportions
-are equally set as $\pi_i=0.25, i=1,2,3,4$.
-<!-- $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$. -->
-Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as 
+are equally set as $\pi_i=0.25, i=1,2,3,4$. The two dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the mapping given by
+
 \begin{equation}
-\label{eq:gmm}
-P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
+\label{eq:twinpeak}
+\begin{array}{lcl}
+x_1 = v_1, \\
+x_2 = v_2, \\
+x_3 = \sin(\pi v_1) \tanh (3 v_2).
+\end{array}
 \end{equation}
 
-```{r mappings, fig.align = 'center', out.width = "80%", fig.cap="3-D Mappings of the meta data with colors and shapes indicating the four kernels.
-Left: swiss roll mapping.
-Right: twin peak mapping.", echo = FALSE}
-knitr::include_graphics("figures/mappings_sr_tp.png")
-```
-
-Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-We use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
-mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
-For simplicity, the simulation results for the swiss roll mapping in te left plot is shown in the Appendix \autoref{swissrollappe}.
-\begin{equation}
-\label{eq:twinpeak}
-\left\{ 
-\begin{array}{lcl}
-X = X_1, \\
-Y = X_2, \\
-Z = \sin(\pi X_1) \tanh (3 X_2).
-\end{array}
-\right.
-\end{equation}
+This is shown on the right panel of \autoref{fig:mappings}.
+We also considered the 'Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in Appendix \autoref{swissrollappe}.
 
@@
 
-We use the multivariate KDE with a fixed plug-in bandwidth in \autoref{mvkde} as the optimal density estimates of the manifold to test the proposed DC-KDE.
-To compare both density estimation methods, it is essential to know the *true density* of the 2-D manifold.
-The Gaussian mixture distribution is known from \@ref(eq:gmm), however, it is not the density of the manifold because of the mapping function in \autoref{eq:twinpeak} distorts the distribution and the manifold of the density is no longer the Gaussian mixture density.
-Therefore, we apply DC-KDE to the 3-D twin peaks data and set the embedding as the 2-D meta data (true manifold) to first calculate the true density of the manifold.
-We use this technique to obtain the true density of the manifold throughout the simulation process.
-
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are *true outliers* to be compared with outliers from the kernel density estimates.
+It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in \autoref{eq:twinpeak} distorts the distribution.
+To recover the true distribution requires the correct change of variables from $\pmb{v}$ to the twin peaks manifold.
+By treating the $\pmb{z}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the learn metric algorithm, the true density on the manifold can be obtained as $p(\pmb{p}_i)=f(\pmb{v}_i)|\Gamma_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals and $|\Gamma_i|^{1/2}$ are the outputs of using the learn metric algorithm on $\pmb{v}$.
+Knowledge of $\pmb{v}$ and $|\Gamma_i|^{1/2}$ will not be used when estimting the density, only to compare the estimated densities to the true densities.
+\autoref{fig:metadensity} shows the simulated $\pmb{v}$ with the colors indicating the true density of data on the manifold.
+Anomalies are defined as points with lowest densities shown in black and 'typical' points are defined as points with the highest densities shown in yellow.
+Based on the true density plot, the anomalies are found around the edges of the plot, but also in the region between the means of the four mixture components.
+The objective is to determine whether we can correctly identify these anomalies without any knowledge of the true density or the $\pmb{v}$.
 

### 2022-10-25 paper/paper.tex
writing section 3

@@
 One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\pmb{H}_j\).
-To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.While density estimation on manifolds may be used in a number of ways, a particular downstream task that we will focus on will be anomaly detection in \ref{alg:kdeanomaly}.
+To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.
 We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required, since anomalies are identified as the points with lowest density.
-
-There are mainly four steps involved in the algorithm.
-The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathbb{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathbb{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation.
-Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels.
-Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}.
-The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f(\pmb{y}_i)\).
-
-\begin{algorithm}[!htb]
-  \caption{Anomaly detection with distortion-corrected kernel density estimates}
-  \label{alg:kdeanomaly}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
-  \Input{ high-dimensional data $\pmb{x}_i$ for all $i=1,\dots,N$ }
-  \Output{ outliers embedding coordinates $\pmb{y}_1, \dots, \pmb{y}_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
-  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
-  \BlankLine
-  \begin{algorithmic}[1]
-  \STATE Estimate the intrinsic dimension $d$ with the TWO-NN algorithm if $d$ is not given;
-  \STATE For all $i=1,\dots,N$, compute the $d$-dimensional embeddings $\pmb{y}_i$ with any existing manifold learning algorithms and the corresponding Riemannian metric $\pmb{H}(\pmb{y}_i)$ using the Learn Metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
-  \STATE Compute the kernel density estimates $\hat{f}(\pmb{y}_i)$ for all $i=1,\dots,N$ using the DC\_KDE in \ref{eq:denestimator};
-  \STATE Reorder the embedding coordinates $\pmb{Y}$ from smallest to largest according to the density estimates $\hat{f}$ and subset the top \textit{n\_outliers} points as the outliers.
-  \end{algorithmic}
-\end{algorithm}
-
-If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding plot and find which observations lie in the highest density region of specified coverage, eg.
-1\%, 50\%, 99\%, \textgreater99\%.
-
-The proposed schematic is shown in Figure \ref{fig:vkde}.
+The entire workflow is summarised in Figure \ref{fig:vkde}.
 The last two steps in Figure \ref{fig:vkde} are the main contributions of Algorithm \ref{alg:kdeanomaly}, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
@@
 In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example.
+For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional Euclidean space.
+General stuff here\ldots{}
+
 We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
@@
 \hypertarget{twodgaussian}{%
-\subsection{3-D mapping from a 2-D Gaussian Mixture Model}\label{twodgaussian}}
-
-We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means
+\subsection{Twin peaks example}\label{twodgaussian}}
+
+The simulation setup for the twin peaks example is to first generate vector \(\pmb{v}_1,\dots,\pmb{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
+The mixture has four components with different means
 \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\).
 The mixture proportions
-are equally set as \(\pi_i=0.25, i=1,2,3,4\).
-Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
-\begin{equation}
-\label{eq:gmm}
-P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
-\end{equation}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/mappings_sr_tp}
-
-}
-
-\caption{3-D Mappings of the meta data with colors and shapes indicating the four kernels. Left: swiss roll mapping. Right: twin peak mapping.}\label{fig:mappings}
-\end{figure}
-
-Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space.
-We use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
-mapping in Equation \eqref{eq:twinpeak}, with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
-For simplicity, the simulation results for the swiss roll mapping in te left plot is shown in the Appendix \autoref{swissrollappe}.
+are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the mapping given by
+
 \begin{equation}
 \label{eq:twinpeak}
-\left\{ 
 \begin{array}{lcl}
-X = X_1, \\
-Y = X_2, \\
-Z = \sin(\pi X_1) \tanh (3 X_2).
+x_1 = v_1, \\
+x_2 = v_2, \\
+x_3 = \sin(\pi v_1) \tanh (3 v_2).
 \end{array}
-\right.
 \end{equation}
+
+This is shown on the right panel of \autoref{fig:mappings}.
+We also considered the `Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in Appendix \autoref{swissrollappe}.
 
@@
 
-We use the multivariate KDE with a fixed plug-in bandwidth in \autoref{mvkde} as the optimal density estimates of the manifold to test the proposed DC-KDE.
-To compare both density estimation methods, it is essential to know the \emph{true density} of the 2-D manifold.
-The Gaussian mixture distribution is known from \eqref{eq:gmm}, however, it is not the density of the manifold because of the mapping function in \autoref{eq:twinpeak} distorts the distribution and the manifold of the density is no longer the Gaussian mixture density.
-Therefore, we apply DC-KDE to the 3-D twin peaks data and set the embedding as the 2-D meta data (true manifold) to first calculate the true density of the manifold.
-We use this technique to obtain the true density of the manifold throughout the simulation process.
-
-\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities.
-We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow.
-Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels.
-These are \emph{true outliers} to be compared with outliers from the kernel density estimates.
+It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in \autoref{eq:twinpeak} distorts the distribution.
+To recover the true distribution requires the correct change of variables from \(\pmb{v}\) to the twin peaks manifold.
+By treating the \(\pmb{z}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the learn metric algorithm, the true density on the manifold can be obtained as \(p(\pmb{p}_i)=f(\pmb{v}_i)|\Gamma_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals and \(|\Gamma_i|^{1/2}\) are the outputs of using the learn metric algorithm on \(\pmb{v}\).
+Knowledge of \(\pmb{v}\) and \(|\Gamma_i|^{1/2}\) will not be used when estimting the density, only to compare the estimated densities to the true densities.
+\autoref{fig:metadensity} shows the simulated \(\pmb{v}\) with the colors indicating the true density of data on the manifold.
+Anomalies are defined as points with lowest densities shown in black and `typical' points are defined as points with the highest densities shown in yellow.
+Based on the true density plot, the anomalies are found around the edges of the plot, but also in the region between the means of the four mixture components.
+The objective is to determine whether we can correctly identify these anomalies without any knowledge of the true density or the \(\pmb{v}\).
 

### 2022-10-27 paper/paper.tex
up to section 3.1

@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional Euclidean space.
-General stuff here\ldots{}
-
-We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points.
-Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as \(d=2\), the same as the meta data dimension.
-This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \ref{alg:kdeanomaly} to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).
+In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion correct KDE.
+For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-dimensional manifold embedded in a 100-dimensional ambient space.
+To estimate the density we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+In general, we aim to highlight two advantages of our proposed distortion corrected KDE compared to KDE applied directly to the output coordinates.
+First, is that the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
+Second, we show how density estimation and anomaly detection is more robust to a different choices of dimension reduction method when distortion correction is used.
 
@@
 
-The simulation setup for the twin peaks example is to first generate vector \(\pmb{v}_1,\dots,\pmb{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
+The simulation setup for the twin peaks example is to first generate vectors \(\pmb{v}_1,\dots,\pmb{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
 The mixture has four components with different means
@@
 The mixture proportions
-are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the mapping given by
+are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the following
 
@@
 This is shown on the right panel of \autoref{fig:mappings}.
-We also considered the `Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in Appendix \autoref{swissrollappe}.
+We also considered the `Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in \autoref{swissrollappe}.
 
@@
 
-\caption{True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
-The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
-The shapes indicate the four kernels.}\label{fig:metadensity}
+\caption{Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
+The colors indicate the true density of the data when they are mapped via the twin peaks function.
+Lower density points in darker colors are scattered both in the outer and center areas.}\label{fig:metadensity}
 \end{figure}
@@
 It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in \autoref{eq:twinpeak} distorts the distribution.
-To recover the true distribution requires the correct change of variables from \(\pmb{v}\) to the twin peaks manifold.
-By treating the \(\pmb{z}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the learn metric algorithm, the true density on the manifold can be obtained as \(p(\pmb{p}_i)=f(\pmb{v}_i)|\Gamma_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals and \(|\Gamma_i|^{1/2}\) are the outputs of using the learn metric algorithm on \(\pmb{v}\).
-Knowledge of \(\pmb{v}\) and \(|\Gamma_i|^{1/2}\) will not be used when estimting the density, only to compare the estimated densities to the true densities.
-\autoref{fig:metadensity} shows the simulated \(\pmb{v}\) with the colors indicating the true density of data on the manifold.
-Anomalies are defined as points with lowest densities shown in black and `typical' points are defined as points with the highest densities shown in yellow.
-Based on the true density plot, the anomalies are found around the edges of the plot, but also in the region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these anomalies without any knowledge of the true density or the \(\pmb{v}\).
+To recover the true distribution requires the correct Jacobian term for the pushforward from \(\pmb{v}\) to the volume form of twin peaks manifold.
+By treating the \(\pmb{v}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\pmb{\Gamma}_i\) for \(i=1,\dots,n\) where \(\pmb{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\pmb{v}\), this notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction.
+The true density on the manifold can be obtained as \(p(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals.
+Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' densities on the manifold.
+\autoref{fig:metadensity} shows the simulated \(\pmb{v}\) with color indicating the true density of data on the manifold.
+Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in yellow.
+The anomalies are found around the edges of the plot, but there are is also a low density region between the means of the four mixture components.
+The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the \(\pmb{v}\).
 
@@
 
-Similar to \autoref{fig:sroutliers}, different manifold learning embeddings are obtained and used to detect outliers with true densities and two bandwidth selection methods shown in \autoref{fig:tpoutliers}.
-In general, the four highest density regions in yellow are identified in almost all manifold learning embeddings except for ISOMAP with fixed bandwidth and t-SNE.
-For ISOMAP, our proposed variable KDE, compared with the true density, gives the most accurate mixture kernel structure with the lowest estimated densities (darker colored points) in the outside and the center of the embedding, and the kernel means (yellow points) with highest densities are also clearly identified.
-In contrast, the fixed bandwidth KDE failed to identify the lowest density area in the center.
-Both variable and fixed bandwidth KDE are quite close with LLE and Laplacian Eigenmaps, but in Laplacian Eigenmaps embedding, the top outliers are indexed in the middle instead of the true outer areas due to the large distortion in the middle.
-For t-SNE and UMAP, there are four clusters in the embedding and UMAP does a better job in finding the HDRs than t-SNE.
-Also due to the clusters in the embedding, the outliers
-found in UMAP are clustered.
+\autoref{fig:tpoutliers} summarises the results.
+Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
+Colors show the different estimated density at each point with anomalies shown in blue, and higher density points shown in yellow.
+For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high density regions.
+On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold in towards the centre.
+The distortion corrected KDE can account for this, KDE without distortion correction on the other hand does not correctly identify the anomalies.
+For t-SNE, the ground truth and distortion corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
+This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
 
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
-Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
-Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.451.
-This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from KDE with and without density correction by \autoref{tab:tpcors}.
+Distortion correction improves the rank correlation for all dimension reduction algorithms.
+In particular while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
 
@@
 
-In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
-The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
-Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to
-fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+In \autoref{fig:tpisomapden}, we plot the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
+Data are presented on a log scale to highlight anomalies.
+The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a 99\% highest density region.
+The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a 95\% and 90\% HDR.
+Points lying outside squares (shown as triangles) will be misidentified as anomalies.
+For example the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99\% HDR), but are not identified as such.
+Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
+\autoref{fig:tptsneden} shows the same plot but for t-SNE.
+The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified.
+However, the difference between KDE with and without distortion correction is stark.
+These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be hogh.
 
@@
 
-\caption{Comparison of outliers found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpisomapvs4ml}
-\end{figure}
-
-Finally, \autoref(fig:tpisomapvs4ml) is used to show the robustness of the density estimates using DC-KDE regardless of the embedding method.
-By plotting the density ranks of LLE, Laplacian Eigenmaps, t-SNE, and UMAP against ISOMAP, we could tell that the rank correlation from DC-KDE is higher that that from KDE, meaning that the rank of DC-KDE are more consistent with the manifold learning methods.
-Compared with KDE where the density ranks varies with the embedding methods even for the same manifold, DC-KDE is more robust when the distortions are fixed.
+\caption{Comparison of ranked densities found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tpisomapvs4ml}
+\end{figure}
+
+Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of dimension reduction algorithm.
+Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
+The left column shoes results when distortion correction is applied, the right column when it is not applied.
+It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
+This is critical since conclusions will be robust to the choice of dimension reduction algorithm.
 

### 2022-10-27 paper/paper.Rmd
up to section 3.1

@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm.
-For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional Euclidean space.
-General stuff here...
-
-We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points.
-Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-The embedded dimension is set as $d=2$, the same as the meta data dimension.
-This enables us to compare the manifold learning embedding with the true meta data.
-We could now apply Algorithm \@ref(alg:kdeanomaly) to get the density estimates of all data points and further detect anomalies.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.
+In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion correct KDE.
+For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-dimensional manifold embedded in a 100-dimensional ambient space.
+To estimate the density we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+In general, we aim to highlight two advantages of our proposed distortion corrected KDE compared to KDE applied directly to the output coordinates.
+First, is that the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
+Second, we show how density estimation and anomaly detection is more robust to a different choices of dimension reduction method when distortion correction is used.
 
@@
 
-The simulation setup for the twin peaks example is to first generate vector $\pmb{v}_1,\dots,\pmb{v}_N$ for $N=2000$ from a 2-dimensional Gaussian mixture model.
+The simulation setup for the twin peaks example is to first generate vectors $\pmb{v}_1,\dots,\pmb{v}_N$ for $N=2000$ from a 2-dimensional Gaussian mixture model.
 The mixture has four components with different means
@@
 The mixture proportions
-are equally set as $\pi_i=0.25, i=1,2,3,4$. The two dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the mapping given by
+are equally set as $\pi_i=0.25, i=1,2,3,4$. The two dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the following
 
@@
 This is shown on the right panel of \autoref{fig:mappings}.
-We also considered the 'Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in Appendix \autoref{swissrollappe}.
+We also considered the 'Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in \autoref{swissrollappe}.
 
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="True density of the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
-The colors indicate the density of the data and lower density points in darker colors are scattered both in the outer and center areas.
-The shapes indicate the four kernels.", echo = FALSE}
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
+The colors indicate the true density of the data when they are mapped via the twin peaks function.
+Lower density points in darker colors are scattered both in the outer and center areas.", echo = FALSE}
 knitr::include_graphics("figures/truedensity_twinpeaks_dc.png")
@@
 It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in \autoref{eq:twinpeak} distorts the distribution.
-To recover the true distribution requires the correct change of variables from $\pmb{v}$ to the twin peaks manifold.
-By treating the $\pmb{z}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the learn metric algorithm, the true density on the manifold can be obtained as $p(\pmb{p}_i)=f(\pmb{v}_i)|\Gamma_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals and $|\Gamma_i|^{1/2}$ are the outputs of using the learn metric algorithm on $\pmb{v}$.
-Knowledge of $\pmb{v}$ and $|\Gamma_i|^{1/2}$ will not be used when estimting the density, only to compare the estimated densities to the true densities.
-\autoref{fig:metadensity} shows the simulated $\pmb{v}$ with the colors indicating the true density of data on the manifold.
-Anomalies are defined as points with lowest densities shown in black and 'typical' points are defined as points with the highest densities shown in yellow.
-Based on the true density plot, the anomalies are found around the edges of the plot, but also in the region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these anomalies without any knowledge of the true density or the $\pmb{v}$.
+To recover the true distribution requires the correct Jacobian term for the pushforward from $\pmb{v}$ to the volume form of twin peaks manifold.
+By treating the $\pmb{v}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\pmb{\Gamma}_i$ for $i=1,\dots,n$ where $\pmb{\Gamma}_i$ is the Riemannian of the coordinate system given by $\pmb{v}$, this notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction.
+The true density on the manifold can be obtained as $p(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals.
+Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' densities on the manifold.
+\autoref{fig:metadensity} shows the simulated $\pmb{v}$ with color indicating the true density of data on the manifold.
+Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in yellow.
+The anomalies are found around the edges of the plot, but there are is also a low density region between the means of the four mixture components.
+The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\pmb{v}$.
 
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
-Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP.
-The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP.
-As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth.
-Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being `r round(min(cors), 3)`.
-This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.
+We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from KDE with and without density correction by \autoref{tab:tpcors}.
+Distortion correction improves the rank correlation for all dimension reduction algorithms.
+In particular while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
 
@@
 
-In \autoref{fig:tpisomapden}, we plot the estimated density against the true density of the ISOMAP embedding for KDE with variable and fixed bandwidth, with colors and shapes showing the four kernels in the meta data.
-The linear positive relationship between the true densities and variable KDEs on the left handside is stronger than that of the fixed bandwidth KDEs.
-Combined with the top-right subplot in \autoref{fig:tpoutliers}, we could tell that most points are underestimated near the true kernel cores, which also suggests that the fixed bandwidth tries to smooth across all the data points and fails to
-fix the local distortions in the manifold learning process like the proposed pointwise variable bandwidth.
+In \autoref{fig:tpisomapden}, we plot the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
+Data are presented on a log scale to highlight anomalies.
+The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a 99% highest density region.
+The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a 95% and 90% HDR.
+Points lying outside squares (shown as triangles) will be misidentified as anomalies.
+For example the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99% HDR), but are not identified as such.
+Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
+\autoref{fig:tptsneden} shows the same plot but for t-SNE.
+The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified.
+However, the difference between KDE with and without distortion correction is stark.
+These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be hogh.
 
-```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
+```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of ranked densities found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
@@
 
-Finally, \autoref(fig:tpisomapvs4ml) is used to show the robustness of the density estimates using DC-KDE regardless of the embedding method.
-By plotting the density ranks of LLE, Laplacian Eigenmaps, t-SNE, and UMAP against ISOMAP, we could tell that the rank correlation from DC-KDE is higher that that from KDE, meaning that the rank of DC-KDE are more consistent with the manifold learning methods.
-Compared with KDE where the density ranks varies with the embedding methods even for the same manifold, DC-KDE is more robust when the distortions are fixed.
+Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of dimension reduction algorithm.
+Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
+The left column shoes results when distortion correction is applied, the right column when it is not applied.
+It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
+This is critical since conclusions will be robust to the choice of dimension reduction algorithm.
 

### 2022-10-28 paper/paper.tex
Fix typo and add commas in Section 2

@@
 \hypersetup{
-            pdftitle={Distortion-corrected kernel density estimate on Riemannian manifolds},
+            pdftitle={Distortion corrected kernel density estimate on Riemannian manifolds},
             pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
@@
 
-\title{Distortion-corrected kernel density estimate on Riemannian manifolds}
+\title{Distortion corrected kernel density estimate on Riemannian manifolds}
 
@@
 
-\hypertarget{kdermtas}{%
-\section{Distortion Corrected Kernel density estimate on Riemannian manifolds - Tas}\label{kdermtas}}
-
-In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in Section \ref{Pellet} how this generalises to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
-In Section \ref{MetLearn} we describe the Metric Learning algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq} we derive our own novel distortion corrected kernel density estimate in Section \ref{DCKDE}.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in Appendix \ref{riemgeo}.
-
-In the following we denote as \(M\) the \(d\)-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted \(\pmb{p}\) in general with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
-Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m>>d\).
-However this need not be that case, for instance the \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each \(\pmb{p}_i\) only require some sense of distance between the `input' points \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding \(\pmb{y}_1,\dots,\pmb{y}_n\) where \(\pmb{y}_i\in\mathbb{R}^d\).
-We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\) Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(||.||\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu} we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\)
+\hypertarget{kderm}{%
+\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}}
+
+In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
+Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
+By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion corrected kernel density estimate in \autoref{DCKDE}.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
+
+In the following, we denote \(M\) as the \(d\)-dimensional manifold from which our data are sampled.
+Points on this manifold are denoted \(\pmb{p}\) in general, with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
+Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\).
+However, it does not need to be the case.
+For instance, \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at each \(\pmb{p}_i\) only require some sense of distance between the `input' points, \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\pmb{y}_1,\dots,\pmb{y}_n\), where \(\pmb{y}_i\in\mathbb{R}^d\).
+We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\).
+Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
 \begin{align}
@@
 & (iv) \supp K=[0; 1];
-(v) \sup K(||\pmb{y}||)=K(0).
+(v) \sup K(\|\pmb{y}\|)=K(0).
 \end{align}
-Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation.
-For instance condition (iv) requires the support of the kernel to be bounded.
+Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
+For instance, condition (iv) requires the support of the kernel to be bounded.
 The reasons for this will become clearer when we discuss the manifold setting in more detail.
 Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise.
-In our empirical section more general kernel functions can be, and are, employed.
-
-For data \(\pmb{y}_i\in\mathbb{R}^d\) for \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{y}\) is given by
-
+In our empirical section, more general kernel functions can be, and are, employed.
+
+For data \(\pmb{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\pmb{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{y}\) is given by
 \begin{equation}
 \label{eq:vkde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{y}-\pmb{y}_i||}{r}\right).
+\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
 \end{equation}
@@
 However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
-For more on kernel density estimation in the Euclidean case see \textcite{Scott2015-vl} and references therein.
-
-While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors \(\pmb{y}_i\) is different to the density on the manifold itself.
+For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein.
+
+While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors \(\pmb{y}_i\) is different from the density on the manifold itself.
 As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-Furthermore any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
+Furthermore, any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
 
@@
 For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
-
 \begin{equation}
@@
 The intuition behind the term \(K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)\) is relatively clear.
-For example, for a uniform kernel the estimator at point \(\pmb{p}\) will still be depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
+For example, for a uniform kernel, the estimator at point \(\pmb{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
 However in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
@@
 
-The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore before providing formal details we will briefly discuss the intuition behind including this term.
+The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind this term.
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
-However the volume of this ball must also be taken into account.
+However, the volume of this ball must also be taken into account.
 In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center.
@@
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space \(\pmb{v}\in T_{\pmb{p}}M\) to points the manifold \(\pmb{q}\in M\) .
-Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and travel along this geodesic at uniform speed \(||\pmb{v}||\) takes place in one unit of time.
-Now, consider a chart \(\varphi\) mapping points in the neighbourhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
+Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space, \(\pmb{v}\in T_{\pmb{p}}M\), to points on the manifold, \(\pmb{q}\in M\).
+Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and travels along this geodesic at uniform speed \(\|\pmb{v}\|\) in one unit of time.
+Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
 The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-For more on the volume density function see \textcite{Le_Brigant2019-lj}.
+For more on the volume density function, see \textcite{Le_Brigant2019-lj}.
 
@@
 The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\), can be used to define geometric quantities.
-For example angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
-While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
+For example, angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
+While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
 This is precisely the situation when data on a manifold are mapped to \(d\)-dimensional Euclidean vectors \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) via a dimension reduction algorithm.
-After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion.
-To alleviate this issue \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\) at each data point.
-These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends (up to a first order approximation) on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).
-
-While full details are provided in \textcite{Perrault-Joncas2013-pq} , we briefly describe the Learn Metric algorithm here.
+After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
+These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
+For example, the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends (up to a first order approximation) on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).
+
+While full details are provided in \textcite{Perrault-Joncas2013-pq}, we briefly describe the Learn Metric algorithm here.
 There are four main steps in the algorithm.
-First a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold.
+First, a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold.
 Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
 Third, a dimension reduction method is applied to obtain the output embedding \(\pmb{y}_1,\dots,\pmb{y}_n\).
-Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at step 2 is a discrete estimator).
-Full details on these four steps are provided in Algorithm \ref{alg:learnmetric}.
+Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
+Full details on these four steps are provided in \autoref{alg:learnmetric}.
 This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
@@
 As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
-In this cae the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\).
+In this case, the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\).
 Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean space.
-In our simulated examples we abstract from this issue by constructing examples that can be globally embedded into \(d\) dimesional Euclidean space.
+In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into \(d\)-dimesional Euclidean space.
 In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used.
@@
 
-With all fundamentals introduced we can now give our novel Distortion Corrected KDE (DC-KDE) as
-
+With all fundamentals introduced, we can now give our novel Distortion Corrected KDE (DC-KDE) as
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||}{r} \bigg).
+\hat{f}(\pmb{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|}{r} \bigg).
 \end{equation}
 
-The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
-To understand these differences it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length is section 6.2 of \textcite{Perrault-Joncas2013-pq}).
+The estimator has a similar structure to \autoref{eq:denriem} with some key differences.
+To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length is Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
 This is crucial for two reasons.
-First, this implies that the term \(|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||\) approximates the geodesic distance between \(\pmb{p}_i\) and \(\pmb{p}_j\).
-Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{p}_i\), it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
-For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm) while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
+First, this implies that the term \(\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|\) approximates the geodesic distance between \(\pmb{y}_i\) and \(\pmb{y}_j\).
+Second, the estimator in \autoref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{y}_i\), and it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
+For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
 Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
@@
 To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required, since anomalies are identified as the points with lowest density.
-The entire workflow is summarised in Figure \ref{fig:vkde}.
-The last two steps in Figure \ref{fig:vkde} are the main contributions of Algorithm \ref{alg:kdeanomaly}, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required because anomalies are identified as the points with lowest density.
+The entire workflow is summarized in \autoref{fig:vkde}.
+The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
 Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-ex}, the changes are also highlighted in blue.
-With this anomaly detection algorithm, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
+With this anomaly detection process, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
 
@@
 This is shown on the right panel of \autoref{fig:mappings}.
-We also considered the `Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in \autoref{swissrollappe}.
+We also considered the `Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarized in \autoref{swissrollappe}.
 
@@
 
-\autoref{fig:tpoutliers} summarises the results.
+\autoref{fig:tpoutliers} summarizes the results.
 Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
@@
 \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
-The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
+The Gaussian mixture densities could be calculated using \autoref{eq:gmm} as the true density of the 5-d meta data.
 \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
@@
 In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \ref{alg:kdeanomaly} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+\autoref{eq:denestimator} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly
@@
 \hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\section{Conclusions}\label{conclusion}}
 
@@
 
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}.
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \autoref{eq:swissroll}.
 The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.

### 2022-10-28 paper/paper.Rmd
Fix typo and add commas in Section 2

@@
 ---
-title: "Distortion-corrected kernel density estimate on Riemannian manifolds"
+title: "Distortion corrected kernel density estimate on Riemannian manifolds"
 author:
@@
 
-# Distortion Corrected Kernel density estimate on Riemannian manifolds - Tas {#kdermtas}
+# Distortion Corrected Kernel density estimate on Riemannian manifolds {#kderm}
 
-In this section we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in Section \@ref(Pellet) how this generalises to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
-In Section \@ref(MetLearn) we describe the Metric Learning algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq we derive our own novel distortion corrected kernel density estimate in Section \@ref(DCKDE).
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in Appendix \@ref(riemgeo).
+In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
+Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
+By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel distortion corrected kernel density estimate in \autoref{DCKDE}.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
 
-In the following we denote as $M$ the $d$-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted $\pmb{p}$ in general with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
-Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m>>d$.
-However this need not be that case, for instance the $\pmb{p}_i$ may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding $\pmb{y}_1,\dots,\pmb{y}_n$ where $\pmb{y}_i\in\mathbb{R}^d$.
-We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$ Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $||.||$ be the usual Euclidean norm and following @Pelletier2005-vu we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$
+In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled.
+Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
+Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$.
+However, it does not need to be the case.
+For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$.
+We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$.
+Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
 \begin{align}
@@
 & (iv) \supp K=[0; 1];
-(v) \sup K(||\pmb{y}||)=K(0).
+(v) \sup K(\|\pmb{y}\|)=K(0).
 \end{align}
-Note that these conditions are different (and in some cases stricter) that those normally used for kernel density estimation.
-For instance condition (iv) requires the support of the kernel to be bounded.
+Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
+For instance, condition (iv) requires the support of the kernel to be bounded.
 The reasons for this will become clearer when we discuss the manifold setting in more detail.
 Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise.
-In our empirical section more general kernel functions can be, and are, employed.
+In our empirical section, more general kernel functions can be, and are, employed.
 
-For data $\pmb{y}_i\in\mathbb{R}^d$ for $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by
-
+For data $\pmb{y}_i\in\mathbb{R}^d$ with $i=1,\dots,N$ and assuming a bandwidth matrix $r\pmb{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by
 \begin{equation}
 \label{eq:vkde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{||\pmb{y}-\pmb{y}_i||}{r}\right).
+\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
 \end{equation}
@@
 However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
-For more on kernel density estimation in the Euclidean case see @Scott2015-vl and references therein.
+For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein.
 
-While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors $\pmb{y}_i$ is different to the density on the manifold itself.
+While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors $\pmb{y}_i$ is different from the density on the manifold itself.
 As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-Furthermore any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
+Furthermore, any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
 
@@
 For kernel density estimation on a known manifold, @Pelletier2005-vu propose the following estimator,
-
 \begin{equation}
@@
 The intuition behind the term $K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)$ is relatively clear.
-For example, for a uniform kernel the estimator at point $\pmb{p}$ will still be depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
+For example, for a uniform kernel, the estimator at point $\pmb{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
 However in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
@@
 
-The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore before providing formal details we will briefly discuss the intuition behind including this term.
+The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind this term.
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
-However the volume of this ball must also be taken into account.
+However, the volume of this ball must also be taken into account.
 In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center.
@@
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space $\pmb{v}\in T_{\pmb{p}}M$ to points the manifold $\pmb{q}\in M$ .
-Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travel along this geodesic at uniform speed $||\pmb{v}||$ takes place in one unit of time.
-Now, consider a chart $\varphi$ mapping points in the neighbourhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
+Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space, $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$.
+Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travels along this geodesic at uniform speed $\|\pmb{v}\|$ in one unit of time.
+Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
 The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-For more on the volume density function see @Le_Brigant2019-lj.
+For more on the volume density function, see @Le_Brigant2019-lj.
 
@@
 The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
-For example angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
-While tangent vectors, the Riemannian metric and the geometric quantities they define are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
+For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
+While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
 This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.
-After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortion.
-To alleviate this issue @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$ at each data point.
-These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm, for example the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
+After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point.
+These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
+For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
 
-While full details are provided in @Perrault-Joncas2013-pq , we briefly describe the Learn Metric algorithm here.
+While full details are provided in @Perrault-Joncas2013-pq, we briefly describe the Learn Metric algorithm here.
 There are four main steps in the algorithm.
-First a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold.
+First, a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold.
 Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
 Third, a dimension reduction method is applied to obtain the output embedding $\pmb{y}_1,\dots,\pmb{y}_n$.
-Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at step 2 is a discrete estimator).
-Full details on these four steps are provided in Algorithm \@ref(alg:learnmetric).
+Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
+Full details on these four steps are provided in \autoref{alg:learnmetric}.
 This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*.
@@
 As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension $d$.
-In this cae the ranks of the matrices $\pmb{H}_i$ are equal to $d$.
+In this case, the ranks of the matrices $\pmb{H}_i$ are equal to $d$.
 Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension $d$ globally into $d$-dimensional Euclidean space.
-In our simulated examples we abstract from this issue by constructing examples that can be globally embedded into $d$ dimesional Euclidean space.
+In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into $d$-dimesional Euclidean space.
 In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used.
@@
 
-With all fundamentals introduced we can now give our novel Distortion Corrected KDE (DC-KDE) as
-
+With all fundamentals introduced, we can now give our novel Distortion Corrected KDE (DC-KDE) as 
 \begin{equation}
 \label{eq:denestimator}
-\hat{f}(\pmb{p}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||}{r} \bigg).
+\hat{f}(\pmb{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|}{r} \bigg).
 \end{equation}
 
-The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
-To understand these differences it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length is section 6.2 of @Perrault-Joncas2013-pq).
+The estimator has a similar structure to \autoref{eq:denriem} with some key differences.
+To understand these differences, it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length is Section 6.2 of @Perrault-Joncas2013-pq).
 This is crucial for two reasons.
-First, this implies that the term $|| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)||$ approximates the geodesic distance between $\pmb{p}_i$ and $\pmb{p}_j$.
-Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\pmb{p}_i$, it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
-For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm) while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
+First, this implies that the term $\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|$ approximates the geodesic distance between $\pmb{y}_i$ and $\pmb{y}_j$.
+Second, the estimator in \autoref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around $\pmb{y}_i$, and it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
+For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
 Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
@@
 To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required, since anomalies are identified as the points with lowest density.
-The entire workflow is summarised in Figure \@ref(fig:vkde).
-The last two steps in Figure \@ref(fig:vkde) are the main contributions of Algorithm \@ref(alg:kdeanomaly), generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required because anomalies are identified as the points with lowest density.
+The entire workflow is summarized in \autoref{fig:vkde}.
+The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
 Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-ex, the changes are also highlighted in blue.
-With this anomaly detection algorithm, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
+With this anomaly detection process, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
 
@@
 In general, we aim to highlight two advantages of our proposed distortion corrected KDE compared to KDE applied directly to the output coordinates.
-First, is that the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
-Second, we show how density estimation and anomaly detection is more robust to a different choices of dimension reduction method when distortion correction is used.
+First, the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
+Second, we show how density estimation and anomaly detection are more robust to a different choice of dimension reduction method when distortion correction is used.
 
@@
 This is shown on the right panel of \autoref{fig:mappings}.
-We also considered the 'Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarised in \autoref{swissrollappe}.
+We also considered the 'Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarized in \autoref{swissrollappe}.
 
@@
 The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation
-$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$.
-The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data.
+$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. The Gaussian mixture densities could be calculated using \autoref{eq:gmm} as the true density of the 5-d meta data.
 \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE].
@@
 In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
-Algorithm \@ref(alg:kdeanomaly) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+\autoref{eq:denestimator} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly
@@
 
-# Conclusion {#conclusion}
+# Conclusions {#conclusion}
 
@@
 
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll).
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \autoref{eq:swissroll}.
 The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.

### 2022-10-29 paper/paper.Rmd
\autoref not working for equations; fig:mappings to be reproduced with twin peaks only

@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds.
+Therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
@@
 
-The estimator has a similar structure to \autoref{eq:denriem} with some key differences.
+The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
 To understand these differences, it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length is Section 6.2 of @Perrault-Joncas2013-pq).
@@
 First, this implies that the term $\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|$ approximates the geodesic distance between $\pmb{y}_i$ and $\pmb{y}_j$.
-Second, the estimator in \autoref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around $\pmb{y}_i$, and it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
+Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\pmb{y}_i$, and it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
 For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
@@
 
-(ref:vkdecaption) The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.
+(ref:vkdecaption) The proposed schematic for anomaly detection with distortion corrected kernel density estimates.
 
@@
 As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-dimensional manifold embedded in a 100-dimensional ambient space.
-To estimate the density we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
 In general, we aim to highlight two advantages of our proposed distortion corrected KDE compared to KDE applied directly to the output coordinates.
@@
 are equally set as $\pi_i=0.25, i=1,2,3,4$. The two dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the following 
-
 \begin{equation}
@@
 x_2 = v_2, \\
-x_3 = \sin(\pi v_1) \tanh (3 v_2).
+x_3 = \sin(\pi v_1) \tan (3 v_2).
 \end{array}
@@
 
-It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in \autoref{eq:twinpeak} distorts the distribution.
+It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in Equation \@ref(eq:twinpeak) distorts the distribution.
 To recover the true distribution requires the correct Jacobian term for the pushforward from $\pmb{v}$ to the volume form of twin peaks manifold.
@@
 The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation
-$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. The Gaussian mixture densities could be calculated using \autoref{eq:gmm} as the true density of the 5-d meta data.
+$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. 
+<!-- The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data. -->
 \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE].
@@
 In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
-\autoref{eq:denestimator} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+Equation \@ref(eq:denestimator) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly
@@
 
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \autoref{eq:swissroll}.
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \@ref(eq:swissroll).
 The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.

### 2022-10-29 paper/paper.tex
\autoref not working for equations; fig:mappings to be reproduced with twin peaks only

@@
 \def\subsectionautorefname{Section}
+%\def\equationautorefname{Equation} % NOT showing Equation 1, but Section 2.1
 \usepackage{bm}
@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds, therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds.
+Therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
@@
 
-The estimator has a similar structure to \autoref{eq:denriem} with some key differences.
+The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
 To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length is Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
@@
 First, this implies that the term \(\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|\) approximates the geodesic distance between \(\pmb{y}_i\) and \(\pmb{y}_j\).
-Second, the estimator in \autoref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{y}_i\), and it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
+Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{y}_i\), and it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
 For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
@@
 
-\caption{The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.}\label{fig:vkde}
+\caption{The proposed schematic for anomaly detection with distortion corrected kernel density estimates.}\label{fig:vkde}
 \end{figure}
@@
 As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-dimensional manifold embedded in a 100-dimensional ambient space.
-To estimate the density we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
+To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
 In general, we aim to highlight two advantages of our proposed distortion corrected KDE compared to KDE applied directly to the output coordinates.
-First, is that the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
-Second, we show how density estimation and anomaly detection is more robust to a different choices of dimension reduction method when distortion correction is used.
+First, the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
+Second, we show how density estimation and anomaly detection are more robust to a different choice of dimension reduction method when distortion correction is used.
 
@@
 are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the following
-
 \begin{equation}
@@
 
-It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in \autoref{eq:twinpeak} distorts the distribution.
+It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in Equation \eqref{eq:twinpeak} distorts the distribution.
 To recover the true distribution requires the correct Jacobian term for the pushforward from \(\pmb{v}\) to the volume form of twin peaks manifold.
@@
 \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
-The Gaussian mixture densities could be calculated using \autoref{eq:gmm} as the true density of the 5-d meta data.
+The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
 \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
@@
 In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
-\autoref{eq:denestimator} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
+Equation \eqref{eq:denestimator} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly
@@
 
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \autoref{eq:swissroll}.
+One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \eqref{eq:swissroll}.
 The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.

### 2022-10-29 paper/paper.tex
Fix figure captions and crossref in Sec 3.1

@@
 \def\subsectionautorefname{Section}
-%\def\equationautorefname{Equation} % NOT showing Equation 1, but Section 2.1
+% \def\equationautorefname~#1\null{Equation~(#1)\null} % NOT showing Equation 1, but Section 2.1
 \usepackage{bm}
@@
 
-This is shown on the right panel of \autoref{fig:mappings}.
-We also considered the `Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarized in \autoref{swissrollappe}.
+The three-dimensional twin peaks mapping is shown in \autoref{fig:twinpeaks3d}.
+The colors in both \autoref{fig:metadensity} and \autoref{fig:twinpeaks3d} indicate the true density of the data via the twin peaks mapping, with lower density points in darker colors scattered in the outer as well as center areas.
 
@@
 Lower density points in darker colors are scattered both in the outer and center areas.}\label{fig:metadensity}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/scatterplot3d_twinpeaks_dc}
+
+}
+
+\caption{Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.}\label{fig:twinpeaks3d}
 \end{figure}
@@
 Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
-Colors show the different estimated density at each point with anomalies shown in blue, and higher density points shown in yellow.
+Colors show the different estimated density at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
 For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high density regions.
 On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold in towards the centre.
-The distortion corrected KDE can account for this, KDE without distortion correction on the other hand does not correctly identify the anomalies.
+The distortion corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
 For t-SNE, the ground truth and distortion corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
@@
 \caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
+Distortion corrected KDE outperfoms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.}
 \centering
@@
 Distortion correction improves the rank correlation for all dimension reduction algorithms.
-In particular while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
+In particular, while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
 
@@
 
-\caption{Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpisomapden}
+\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
+The colors indicate different level of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
+The shading contains all anomalies that are both truely and correctly identified.
+KDE without distortion correction gives more misclassified anomalies.}\label{fig:tpisomapden}
 \end{figure}
@@
 
-\caption{Scatterplot of true density and estimated density ranks of t-SNE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tptsneden}
+\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding.
+KDE without distortion correction gives many more misclassified anomalies.}\label{fig:tptsneden}
 \end{figure}
@@
 Points lying outside squares (shown as triangles) will be misidentified as anomalies.
-For example the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99\% HDR), but are not identified as such.
+For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99\% HDR), but are not identified as such.
 Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
@@
 However, the difference between KDE with and without distortion correction is stark.
-These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be hogh.
+These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.
 
@@
 
-\caption{Comparison of ranked densities found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tpisomapvs4ml}
+\caption{Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
+Distortion corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.}\label{fig:tpisomapvs4ml}
 \end{figure}
@@
 Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
-The left column shoes results when distortion correction is applied, the right column when it is not applied.
+The left column shows results when distortion correction is applied, the right column when it is not applied.
 It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
@@
 \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
-The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data.
 \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
@@
 One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \eqref{eq:swissroll}.
-The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}.
+The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in \autoref{fig:swissroll3d}.
 The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
@@
 \end{equation}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/scatterplot3d_swissroll_dc}
+
+}
+
+\caption{Scatterplot of the 3-d swiss roll data with colors indicating the true density via the swiss roll mapping in \autoref{eq:swissroll}.}\label{fig:swissroll3d}
+\end{figure}
 

### 2022-10-29 paper/paper.Rmd
Fix figure captions and crossref in Sec 3.1

@@
 
-This is shown on the right panel of \autoref{fig:mappings}.
-We also considered the 'Swiss Roll' mapping shown on the left panel of \autoref{fig:mappings}, the results for this manifold are summarized in \autoref{swissrollappe}.
+The three-dimensional twin peaks mapping is shown in \autoref{fig:twinpeaks3d}.
+The colors in both \autoref{fig:metadensity} and \autoref{fig:twinpeaks3d} indicate the true density of the data via the twin peaks mapping, with lower density points in darker colors scattered in the outer as well as center areas.
+<!-- We also considered the 'Swiss Roll' mapping and the results for this manifold are summarized in \autoref{swissrollappe}. -->
 
@@
 knitr::include_graphics("figures/truedensity_twinpeaks_dc.png")
+```
+
+(ref:twinpeaks3dcaption) Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.
+
+```{r twinpeaks3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:twinpeaks3dcaption)", echo = FALSE}
+knitr::include_graphics("figures/scatterplot3d_twinpeaks_dc.png")
 ```
@@
   kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+Distortion corrected KDE outperfoms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>%
@@
 Distortion correction improves the rank correlation for all dimension reduction algorithms.
-In particular while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
+In particular, while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
 
-```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
+The colors indicate different level of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
+The shading contains all anomalies that are both truely and correctly identified.
+KDE without distortion correction gives more misclassified anomalies.", echo = FALSE}
 # knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
@@
 
-```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of t-SNE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
+```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding.
+KDE without distortion correction gives many more misclassified anomalies.", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
@@
 Points lying outside squares (shown as triangles) will be misidentified as anomalies.
-For example the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99% HDR), but are not identified as such.
+For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99% HDR), but are not identified as such.
 Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
@@
 However, the difference between KDE with and without distortion correction is stark.
-These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be hogh.
+These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.
 
-```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of ranked densities found by ISOMAP and other four manifold learning methods for DC-KDE (on the left panel) and KDE (on the right panel).", echo = FALSE}
+```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
+Distortion corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
@@
 One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \@ref(eq:swissroll).
-The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}.
+The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in \autoref{fig:swissroll3d}.
 The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
@@
 \end{equation}
+
+(ref:swissroll3dcaption) Scatterplot of the 3-d swiss roll data with colors indicating the true density via the swiss roll mapping in \autoref{eq:swissroll}.
+
+```{r swissroll3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:swissroll3dcaption)", echo = FALSE}
+knitr::include_graphics("figures/scatterplot3d_swissroll_dc.png")
+```
 

### 2022-10-29 R/Simulation2d_TwinPeaks.R
Update scatterplot3d colors for twin peaks

@@ -20,7 +20,7 @@ p <- 2
 ## 1. Generate 2-D manifold and 3-D twin peaks
 ##--------------------------------------------------------
 set.seed(1234)
-mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3] # only run [1] and [3]
+mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[1] # only run [1] and [3]
 sr <- mldata(N = N, meta = "gaussian", mapping = mapping)
 # sr <- mldata(N = N, meta = "uniform", mapping = mapping)
 swissroll <- sr$data %>% as.data.frame()
@@ -96,12 +96,17 @@ plotmanifold <- preswissroll %>%
              )) +
   geom_point() +
   scale_color_viridis(option = "B") +
-  labs(color = "Density" # , shape = "Kernels"
-       )
+  labs(color = "Density")
 plotmanifold
 # ggsave("paper/figures/truedensity_twinpeaks_dc.png", plotmanifold, width = 8, height = 6, dpi = 300)
 plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = den_2dmanifold, # colored with 3d density
         type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
+graphics.off()
+png("paper/figures/scatterplot3d_swissroll.png", width = 8, height = 6, units = "in", res = 300)
+# par(mfrow = c(1, 2))
+# scatterplot3d::scatterplot3d(swissroll_tp, color = ggplot_build(plotmanifold_tp)$data[1][[1]][,1], pch = 20, xlab = "X1", ylab = "X2", zlab = "X3")
+scatterplot3d::scatterplot3d(swissroll, color = ggplot_build(plotmanifold)$data[1][[1]][,1], pch = 20, xlab = "X1", ylab = "X2", zlab = "X3")
+dev.off()
 trueden <- den_2dmanifold # sr$den
 trueoutliers <- head(order(trueden), noutliers)
 # save(sr, train, trueden, den_twinpeaks, noutliers, trueoutliers, file = paste0("data/simdata_3d_N", N, "_01_trueden_k", k, ".rda"))

### 2022-10-30 R/Simulation2d_TwinPeaks.R
Rename gif

@@ -20,7 +20,7 @@ p <- 2
 ## 1. Generate 2-D manifold and 3-D twin peaks
 ##--------------------------------------------------------
 set.seed(1234)
-mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[1] # only run [1] and [3]
+mapping <- c("Swiss Roll", "semi-sphere", "Twin Peak", "S Curve")[3] # only run [1] and [3]
 sr <- mldata(N = N, meta = "gaussian", mapping = mapping)
 # sr <- mldata(N = N, meta = "uniform", mapping = mapping)
 swissroll <- sr$data %>% as.data.frame()
@@ -96,9 +96,10 @@ plotmanifold <- preswissroll %>%
              )) +
   geom_point() +
   scale_color_viridis(option = "B") +
-  labs(color = "Density")
+  labs(color = "Density",
+       x = "V1", y = "V2")
 plotmanifold
-# ggsave("paper/figures/truedensity_twinpeaks_dc.png", plotmanifold, width = 8, height = 6, dpi = 300)
+# ggsave("paper/figures/truedensity_twinpeaks_dc_labv.png", plotmanifold, width = 8, height = 6, dpi = 300)
 plot_ly(data = swissroll, x = ~ x, y = ~ y, z = ~ z, color = den_2dmanifold, # colored with 3d density
         type = "scatter3d", mode = "markers", size = 1, text = paste("density:", preswissroll$den))
 graphics.off()

### 2022-10-30 R/Simulation4d_01_datadensity.R

@@ -39,7 +39,7 @@ den <- apply(X, 1, gmmdensity)
 
 # semi-sphere radius
 radius <- ceiling(max(sqrt(x1^2 + x2^2 + x3^2 + x4^2)))
-radius # radius = 15 for sigma=5
+radius # radius = 8
 range(X)
 
 semisphere <- cbind(X,
@@ -49,19 +49,37 @@ semisphere <- cbind(X,
 # head(semisphere)
 # semisphere %>% as_tibble() %>% summarise(x1^2 + x2^2 + x3^2 + x4^2 + x5^2)
 
-# # distance to (0,0,0,0,0) is the same, = radius = 6
+# # distance to (0,0,0,0,0) is the same, = radius = 8
 # # calculate distance to (0,0,0,0) and color them
-# label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
-# dist2center <- sqrt(radius^2 - semisphere[,5]^2)
-# ## Plot 5-d semi-sphere, run once
+label <- as.factor(c(rep(1, 0.99*N), rep(2, 0.01*N)))
+dist2center <- sqrt(radius^2 - semisphere[,5]^2)
+## Run once
+# ## Plot 5-d semi-sphere
 # library(tourr)
 # colors <- colourvalues::colour_values(- dist2center, palette = "magma") # generate colors for locations
 # pchs <- c(rep(16, 0.99*N), rep(17, 0.01*N)) # point shapes for kernels
-# animate_xy(semisphere[,1:5], col = colors, pch = pchs, cex = 0.8,
-#            axes = "bottomleft", fps=15
+# animate_xy(semisphere[,1:5], col = colors, pch = pchs, cex = 0.8#,
+#            # axes = "bottomleft", fps = 15
 #            )
 # # "figures/tourr_5d_semisphere.png"
+# ## Render frames as png files in a folder and convert to gif with ImageMagick
+# gifpath <- "paper/figures/tourr_render/"
+# tourr::render(semisphere[,1:5], grand_tour(), display_xy(col = colors, pch = pchs, cex = 0.8), "png",
+#               frames = 200,
+#               file.path(gifpath, "tourr_5d_semisphere-%20d.png")
+#               )
+## Save gif file as online supplementary file
+# system("convert -delay 10 paper/figures/tourr_render/*.png paper/figures/tourr_5d_animation.gif")
+## Not run
[…]

### 2022-10-30 paper/paper.Rmd
Sec 3 done

@@
 This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*.
+Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.
 
@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion correct KDE.
+In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion corrected KDE.
 For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-dimensional manifold embedded in a 100-dimensional ambient space.
+As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 4-dimensional manifold embedded in a 100-dimensional ambient space.
 To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
@@
 Lower density points in darker colors are scattered both in the outer and center areas.", echo = FALSE}
-knitr::include_graphics("figures/truedensity_twinpeaks_dc.png")
+knitr::include_graphics("figures/truedensity_twinpeaks_dc_labv.png")
 ```
@@
 By treating the $\pmb{v}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\pmb{\Gamma}_i$ for $i=1,\dots,n$ where $\pmb{\Gamma}_i$ is the Riemannian of the coordinate system given by $\pmb{v}$, this notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction.
-The true density on the manifold can be obtained as $p(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals.
+The true density on the manifold can be obtained as $f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals.
 Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' densities on the manifold.
@@
 Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
+We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$.
 Colors show the different estimated density at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
@@
 KDE without distortion correction gives more misclassified anomalies.", echo = FALSE}
-# knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_crossfalse.png")
-# knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks_crossfalse.png")
 knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
@@
 Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a 99% highest density region.
-The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a 95% and 90% HDR.
+The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a $>99\%$ highest density region.
+The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a $99\%$ and $90\%$ HDR.
 Points lying outside squares (shown as triangles) will be misidentified as anomalies.
-For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99% HDR), but are not identified as such.
+For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the $99\%$ HDR), but are not identified as such.
 Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
@@
 
-## 100-D mapping from a 5-D semi-hypersphere {#fivedgaussian}
+## Semi-hypersphere example embedded in 100-D space {#fivedgaussian}
 
-```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.", echo = FALSE}
+```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
+Distant points in darker colors could be seen as anomalies.", echo = FALSE}
 knitr::include_graphics("figures/tourr_5d_semisphere.png")
@@
 
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in $d=5$ with manifold learning.
-First, we simulate $N=2,000$ points,
-$(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a 4-dimensional Gaussian mixture model with two mixture components,
-$\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
-In order to manually add anomalies to be distant points from the means, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation
-$X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. 
-<!-- The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data. -->
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE].
+As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, then transform it into a 100-dimensional ambient space.
+To start with, we simulate vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ for $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, with the same means $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$ and different variance-covariance matrices $\pmb{\Sigma}_1 = diag(1,1,1,1)$ and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
+In order to manually add anomalies to be distant points from the core center, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
+The fifth dimension is derived to satisfy the five-dimensional semi-hypersphere surface equation
+$v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path[^1] with the R package *tourr* [@Wickham2011-ir].
 The round and triangular point shapes index the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the distance between the simulated $4-D$ data point from Gaussian mixture model and the kernel means $(0, 0, 0, 0)^\prime$.
-The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
-The dark colors also indicate that they are the true outliers because of their low densities.
+$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the kernel core $(0, 0, 0, 0, 0)^\prime$.
+It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the top outliers are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
 
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the $100$-dimensional data of size $N$(denote the transpose of the data matrix as $\pmb{X}_0$ with dimension $100 \times N$) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$.
+[^1]: See the animation of the 5-D grand tour at https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif.
+
+To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to $\pmb{v}_i$ so that $\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N$.
+Next, we rotate the $100$-dimensional vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ with QR decomposition to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+Denote $\pmb{X}_0 = (\pmb{v}_1, \dots, \pmb{v}_N)$ with dimension $100 \times N$.
+The rotation matrix $Q$ is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$.
+The details are as follows.
 For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix.
@@
 $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
-Therefore, we use matrix $Q$ as the rotation matrix for where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms.
-Again, we set the embedding dimension to be equal to the meta data dimension $d=5$.
+Therefore, we take advantage of the matrix $Q$ as the rotation matrix where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the dimension reduction algorithms.
+It is worth noting that the intrinsic dimension of $X$ is known in this example as $d=4$.
 
@@
 The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
-# knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_riem1k200_rawdensity.png")
 knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec.png")
@@
 
-In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
-Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
-For well-estimated densities, the true
-outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
-This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
-For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and
-the relationship is stronger than the fixed bandwidth.
-This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
-In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
-By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
+Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density in \autoref{fig:fivedisomapden}.
+The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example.
+In \autoref{fig:fivedisomapden}, the log-scale ranks of the estimated density from distortion corrected KDE shown in the left panel and KDE shown in the right panel are compared with the true rank of density for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to embedding dimensions within three.
+Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the kernel cores with outliers shown in darker colors.
+The grey shading area highlights the $>99\%$ highest density region area and points falling into this area are the correctly classified top $1\%$ anomalies.
+Darker points lying outside the shaded squares in the top left area are true anomalies but are misclassified as such.
+By comparing the left and right panel for each row, we could notice that KDE with distortion corrected outperforms KDE in classifying anomalies.
+For UMAP, almost all ground truth top $1\%$ anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected.
 
@@
 
+In \autoref{tab:fivedcors}, the correlations between ranks of true densities and estimated densites from DC-KDE and KDE are presented.
+For ISOMAP and LLE, KDE gives higher rank correlation by less than 0.01, but both density estimators have a very high correlation of more than 0.96.
+However, for Laplacian Eigenmaps and UMAP where there are more distortion in the dimension reduction process, the rank correlation for KDE is close to 0 while our distortion correction methods gives a relatively high correlation of 0.87 and 0.78 respectively.
+
 ```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
@@
 hdrtable %>% 
+  head(n = 2) %>% 
+  `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
   kableExtra::kbl(caption = "Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
@@
   # kable_classic("striped") %>% 
-  kable_styling(latex_options = "scale_down") %>%
+  # kable_styling(latex_options = "scale_down") %>%
   kable_paper(full_width = FALSE) %>%
@@
 ```
+
+Furthermore, when comparing across different dimension reduction algorithms in \autoref{fig:fivedisomapden}, almost all top outliers are correctly classified in the shading areas, which again shows the robustness of the proposed distortion correction methods to the use of dimension reduction algorithms.
+\autoref{tab:fourdhdrtable} further illustrutes the percentage of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions.
+Compared with KDE, the percentages are always higher when distortion correction methods are used.
+The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
+This could be due to the severe distortion usually induced by the UMAP algorithm, which can be seen from the $0\%$ in KDE when no distortion correction is used.
 

### 2022-10-30 paper/paper.tex
Sec 3 done

@@
 This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
+Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.
 
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_twinpeaks_dc}
+{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_twinpeaks_dc_labv}
 
@@
 By treating the \(\pmb{v}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\pmb{\Gamma}_i\) for \(i=1,\dots,n\) where \(\pmb{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\pmb{v}\), this notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction.
-The true density on the manifold can be obtained as \(p(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals.
+The true density on the manifold can be obtained as \(f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals.
 Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' densities on the manifold.
@@
 Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
+We set the bandwidth parameter in the DC-KDE \eqref{eq:denestimator} as \(r=0.5\).
 Colors show the different estimated density at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
@@
 Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a 99\% highest density region.
-The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a 95\% and 90\% HDR.
+The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a \(>99\%\) highest density region.
+The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a \(99\%\) and \(90\%\) HDR.
 Points lying outside squares (shown as triangles) will be misidentified as anomalies.
-For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the 99\% HDR), but are not identified as such.
+For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the \(99\%\) HDR), but are not identified as such.
 Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
@@
 \hypertarget{fivedgaussian}{%
-\subsection{100-D mapping from a 5-D semi-hypersphere}\label{fivedgaussian}}
+\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}}
 
@@
 
-\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
-\end{figure}
-
-As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning.
-First, we simulate \(N=2,000\) points,
-\((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components,
-\(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
-In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation
-\(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
+\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
+Distant points in darker colors could be seen as anomalies.}\label{fig:fivedmeta}
+\end{figure}
+
+As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, then transform it into a 100-dimensional ambient space.
+To start with, we simulate vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), with the same means \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\pmb{\Sigma}_1 = diag(1,1,1,1)\) and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
+In order to manually add anomalies to be distant points from the core center, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
+The fifth dimension is derived to satisfy the five-dimensional semi-hypersphere surface equation
+\(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
+\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} with the R package \emph{tourr} \autocite{Wickham2011-ir}.
 The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
-\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\).
-The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}.
-It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
-The dark colors also indicate that they are the true outliers because of their low densities.
-
-Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\)(denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
+\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the kernel core \((0, 0, 0, 0, 0)^\prime\).
+It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the top outliers are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
+
+To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to \(\pmb{v}_i\) so that \(\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\).
+Next, we rotate the \(100\)-dimensional vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) with QR decomposition to get rid of the zeros so that it could be passed to the manifold learning algorithms.
+Denote \(\pmb{X}_0 = (\pmb{v}_1, \dots, \pmb{v}_N)\) with dimension \(100 \times N\).
+The rotation matrix \(Q\) is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
+The details are as follows.
 For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix.
@@
 \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
-Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms.
-Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).
+Therefore, we take advantage of the matrix \(Q\) as the rotation matrix where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the dimension reduction algorithms.
+It is worth noting that the intrinsic dimension of \(X\) is known in this example as \(d=4\).
 
@@
 
-In \autoref{fig:fivedisomapden}, the estimated densities are compared with the true density on the x-axis for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes, and it is only applicable to embedding dimensions within three.
-Similar to \autoref{fig:fivedmeta}, the point shapes show the two mixture component in the meta data, and the colors represent the distance to the kernel means, with distant outliers shown in darker colors.
-For well-estimated densities, the true
-outliers with low true densities will also have low estimated densities, which suggests that darker-colored points should appear in the bottom-left corner in \autoref{fig:fivedisomapden}.
-This is true for both ISOMAP and LLE, partly true for Laplacian Eigenmaps, but not in UMAP where these outliers have relatively high density estimates.
-For variable bandwidth KDE, there is a strong positive linear relationship with the true densities for ISOMAP, LLE, and Laplacian Eigenmaps, and
-the relationship is stronger than the fixed bandwidth.
-This suggests that our proposed KDE with variable bandwidth is more accurate than the fixed bandwidth in estimating the manifold learning embedding densities.
-In KDE with fixed bandwidth, the bandwidth is often too large to smooth across all data points, especially when there is severe distortion in the embedding data.
-By introducing the pointwise variable Riemannian metric in kernel density estimation, it is reasonable to believe that it could fix the distortion introduced by these three manifold learning algorithms.
+Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density in \autoref{fig:fivedisomapden}.
+The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example.
+In \autoref{fig:fivedisomapden}, the log-scale ranks of the estimated density from distortion corrected KDE shown in the left panel and KDE shown in the right panel are compared with the true rank of density for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to embedding dimensions within three.
+Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the kernel cores with outliers shown in darker colors.
+The grey shading area highlights the \(>99\%\) highest density region area and points falling into this area are the correctly classified top \(1\%\) anomalies.
+Darker points lying outside the shaded squares in the top left area are true anomalies but are misclassified as such.
+By comparing the left and right panel for each row, we could notice that KDE with distortion corrected outperforms KDE in classifying anomalies.
+For UMAP, almost all ground truth top \(1\%\) anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected.
 
@@
 
+In \autoref{tab:fivedcors}, the correlations between ranks of true densities and estimated densites from DC-KDE and KDE are presented.
+For ISOMAP and LLE, KDE gives higher rank correlation by less than 0.01, but both density estimators have a very high correlation of more than 0.96.
+However, for Laplacian Eigenmaps and UMAP where there are more distortion in the dimension reduction process, the rank correlation for KDE is close to 0 while our distortion correction methods gives a relatively high correlation of 0.87 and 0.78 respectively.
+
 \begin{table}
@@
 \centering
-\resizebox{\linewidth}{!}{
 \begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
@@
 \midrule
->99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
-99 & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
-90 & \cellcolor[HTML]{E6E6E6}{0.919} & \textbf{0.926} & \cellcolor[HTML]{E6E6E6}{\textbf{0.921}} & 0.900 & \cellcolor[HTML]{E6E6E6}{\textbf{0.798}} & 0.440 & \cellcolor[HTML]{E6E6E6}{\textbf{0.779}} & 0.428\\
-50 & \cellcolor[HTML]{E6E6E6}{0.825} & \textbf{0.851} & \cellcolor[HTML]{E6E6E6}{0.831} & \textbf{0.848} & \cellcolor[HTML]{E6E6E6}{\textbf{0.659}} & 0.512 & \cellcolor[HTML]{E6E6E6}{\textbf{0.623}} & 0.448\\
-5 & \cellcolor[HTML]{E6E6E6}{0.508} & \textbf{0.556} & \cellcolor[HTML]{E6E6E6}{0.514} & \textbf{0.562} & \cellcolor[HTML]{E6E6E6}{\textbf{0.431}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.220}} & 0.000\\
-\addlinespace
-1 & \cellcolor[HTML]{E6E6E6}{0.080} & \textbf{0.190} & \cellcolor[HTML]{E6E6E6}{0.090} & \textbf{0.230} & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.000}} & 0.000\\
+>99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
+99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
 \bottomrule
-\end{tabular}}
+\end{tabular}
 \end{table}
+
+Furthermore, when comparing across different dimension reduction algorithms in \autoref{fig:fivedisomapden}, almost all top outliers are correctly classified in the shading areas, which again shows the robustness of the proposed distortion correction methods to the use of dimension reduction algorithms.
+\autoref{tab:fourdhdrtable} further illustrutes the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
+Compared with KDE, the percentages are always higher when distortion correction methods are used.
+The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
+This could be due to the severe distortion usually induced by the UMAP algorithm, which can be seen from the \(0\%\) in KDE when no distortion correction is used.
 

### 2022-10-30 paper/paper.Rmd
Fix Figure 9 where top 67 anomalies have the same rank 33.5 and appear in a horizontal line
This is because the true densities are all below 1e-7

@@
 For UMAP, almost all ground truth top $1\%$ anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected.
+It is worth noting that for DC-KDE in the left panel, there are many points having the same rank of estimated density inside the shading area, which is due to the fact that the ground truth density are almost zeros, with values less than $10^{-7}$.
+And these low-density points are all captured by our distortion correction methods regardless of the dimension reduction algorithms, which again shows the robustness of DC-KDE.
 

### 2022-10-30 paper/paper.tex
Fix Figure 9 where top 67 anomalies have the same rank 33.5 and appear in a horizontal line
This is because the true densities are all below 1e-7

@@
 For UMAP, almost all ground truth top \(1\%\) anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected.
+It is worth noting that for DC-KDE in the left panel, there are many points having the same rank of estimated density inside the shading area, which is due to the fact that the ground truth density are almost zeros, with values less than \(10^{-7}\).
+And these low-density points are all captured by our distortion correction methods regardless of the dimension reduction algorithms, which again shows the robustness of DC-KDE.
 

### 2022-10-30 R/Simulation4d_03_plots.R

@@ -2,11 +2,10 @@
 ## 1. Gather and plot the results from 4 scenarios in HPC results
 ###--------------------------------------------------------
 rm(list = ls())
-# first load all .rda files
 library(tidyverse)
 library(viridis)
 N <- 10000
-radius <- 10
+radius <- 8
 r <- 1
 # method <- c("isomap", "lle", "le", "tsne", "umap")[1]
 load("data/simdata_100d_4dmanifold_N10000_trueden_k100.rda")
@@ -17,12 +16,17 @@ methods <- c("isomap", "lle", "le", "umap")
 # load HPC results for four methods
 files <- list.files(path = "data/truedenk100", pattern = "*.rda", full.names = TRUE)
 lapply(files, load, .GlobalEnv)
+
+# getOption("digits")
+# options(digits = 7)
 f <- tibble(fxy = fxy,
             fxy_vkde_isomap = fisomap$estimate, fxy_hdr_isomap = fixden_isomap$estimate,
             fxy_vkde_lle = flle$estimate, fxy_hdr_lle = fixden_lle$estimate,
             fxy_vkde_le = fle$estimate, fxy_hdr_le = fixden_le$estimate,
             fxy_vkde_umap = fumap$estimate, fxy_hdr_umap = fixden_umap$estimate
 )
+# f[which(rank(f$fxy_vkde_isomap) == 33.5),]$fxy_hdr_isomap # There are ties in dc-kde results, leading to rank=33.5 for the top 67 anomalies
+# f[which(rank(f$fxy_vkde_isomap) == 33.5),]$fxy # Their true densities are not exactly the same, but are all below 1e-07, so basically 0
 
 ###--------------------------------------------------------
 ## 2. Table for density correlations
@@ -80,6 +84,8 @@ hdrmax <- (as.matrix(hdrtable[, 1:4]) - as.matrix(hdrtable[, 5:8])) >= 0
 hdrmax <- cbind(hdrmax, !hdrmax)[, c(1, 5, 2, 6, 3, 7, 4, 8)] %>% `rownames<-` (NULL)
 hdrtable <- hdrtable[, c(1, 5, 2, 6, 3, 7, 4, 8)]
 hdrtable %>% 
+  head(n = 2) %>% 
+  `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
[…]

### 2022-10-30 paper/paper.Rmd
Sec 4&5 done

@@
 <!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
-It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
+It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-dh], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
 In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
@@
 This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-dh), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
 
@@
 These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
-Similar to @Cheng2021-ex, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
+Similar to @Cheng2021-dh, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
 
@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-ex, the changes are also highlighted in blue.
+Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue.
 With this anomaly detection process, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 
-## Irish smart meter dataset
+<!-- ## Irish smart meter dataset -->
 
 In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010.
-The CER dataset[^2] records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
-We selected the $3,639$ residential data with no missing values during the data collection period for a total of
-$535$ days.
+The CER dataset[^2] records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
+We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.
 
@@
 For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
-@Cheng2021-ex propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
-@Cheng2021-ex use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
-By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
+@Cheng2021-dh propose two non-Euclidean distance estimators to be used in manifold learning algorithms in case of statistical manifolds with each observation as a distribution.
+The same Irish smart meter data is used for identifying outliers with kernel density estimation, but they fail to consider the distortion and information loss in the 2-dimensional embeddings.
+Therefore, we propose in this paper to apply distortion corrected KDE for anomaly detection as the downstream task of dimension reduction.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to get the 2-D embedding for kernel density estimation and anomaly detection.
+To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms with the use of distortion correction.
+We use the highest density regions plots to visualize the density estimates.
+However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
+Therefore, we will further examine the distributions of the anomalous households to study the typical or anomalous electricity usage patterns.
 
-In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households.
+Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the procedures here.
+First, for each household and each of the $336$ half-hourly periods of the week, we build the empirical distributions and apply the total variation distance estimator for statistical manifold learning, which gives a 2-dimensional embedding for each dimension reduction method.
+Then we apply the kernel density estimator with distortion corrected for each method to the low-dimensional embedding and compare it with the KDE results in the Section 4.4 of @Cheng2021-dh.
+The bandwidth parameter $r$ is set as 1 in Equation \@ref(eq:denestimator).
+
+ ```{r electricityembedding, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
+The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=F}
+# knitr::include_graphics("figures/")
+```
+
+<!-- The results are summarized in \autoref{fig:electricityembedding}.
+TODO -->
+
+<!-- In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-dh in the statistical manifold learning to get the 2-D embedding of all households.
 Equation \@ref(eq:denestimator) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
 The data processing steps have been clearly
-stated in the application section of @Cheng2021-ex and they are skipped here.
+stated in the application section of @Cheng2021-dh and they are skipped here.
 Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
 However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
-
-```{r electricityoutliers, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.", echo = FALSE, eval=T}
+ -->
+```{r electricityoutliers, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/kde_2densities_raw_electricity.png")
@@
 \autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
-As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
+The typical households in the top row are similar except that there are a few spikes for household $1472$ when using KDE.
+As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume.
 It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
-In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
 
-```{r electricityhdrsfixed, fig.align = 'center', out.width = "90%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=T}
+```{r electricityhdrsfixed, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of week for one typical household 1321 and two anomalies, 3243 and 2838.
+The quantile regions displayed in the plot are 99.0\\%, 95.0\\%, 75\\%, and 50\\%.
+Household 1321 has a repeated period-of-the-week electricity usage pattern slightly different for weekdays and weekends.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/hdr_electricity_hdrbox_3id_7dow.png")
@@
 
-```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.", echo = FALSE, eval=T}
+```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of week for one typical household 1472 and two anomalies, 7049 and 5136.
+The day-of-the-week patterns is much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time related usage patterens.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/electricity_hdrbox_3id_7dow.png")
@@
 Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-However, this repeated pattern in a week window is clearly for the typical household ID 1321.
+This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE.
 As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
 The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
-These findings show the difference in finding typical and anomalous households with different bandwidth selections.
+These findings show the difference in finding typical and anomalous households with different kernel density estimation methods.
 
@@
 
-In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
+In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities.
+Compared with KDE, our distortion corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
+The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
+We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension redution algorithm.
 
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
+As an empirical example, we explore the distributions of $3,639$ households and $336$ time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
+Without the ground truth density, we compare both density estimates by looking at the distributions of the most typical households with the highest densities and the anomalous households with the top two lowest densities.
 Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
@@
 There are several open questions to be explored.
-The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of @Cheng2021-ex.
-The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
-The scale of distortion in each point in manifold learning could vary a lot.
-If we could smooth across all data points, eg.
-multiply the Riemannian matrix with the ratio between its determinant
-and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
-The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
-The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related
-to real-life data sets.
+The first involves the selection of bandwidth parameters for kernel density estimation, which has been explored in many KDE-related literatures.
+The second one is related to the quality of the embedding from dimension reduction methods.
+Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitive relationship between embedding quality and density estimation accuracy is not immediate clear.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell wherther the distortion are corrected in the right way.
+The density estimates on the edges of the whole data structure could alos be explored because most outer area points tend to be detected as outliers.
+However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
 
-<!-- Shortcoming and other references for future direction (improvement on density estimation, other kernels, densities on edges, bad embeddings with inaccurate distortions, multi-dimensional kernel density estimate).
--->
-
-<!-- Discuss simulation with non-uniform meta data. -->
-
-<!-- The aim of the project is not to evaluate different manifold learning algorithms.
-Although ISOMAP worked OK here it is still distorted and in real examples, you will never 'know' that the data are uniform on the sphere.
-Instead, the idea of this project is to do kernel density estimation in a way that takes distortion into account.
--->
+<!-- kernel types discussion? -->
 
@@
 
-# Appendix: Simulation with swiss roll mapping {#swissrollappe}
+<!-- # Appendix: Simulation with swiss roll mapping {#swissrollappe}
 
@@
 However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
-
+ -->
 <!-- Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.

### 2022-10-30 paper/paper.tex
Sec 4&5 done

@@
 This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
+For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-dh}), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
 
@@
 These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
-Similar to \textcite{Cheng2021-ex}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
+Similar to \textcite{Cheng2021-dh}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
 
@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-ex}, the changes are also highlighted in blue.
+Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
 With this anomaly detection process, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 
-\hypertarget{irish-smart-meter-dataset}{%
-\subsection{Irish smart meter dataset}\label{irish-smart-meter-dataset}}
-
 In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
-The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems.
-We selected the \(3,639\) residential data with no missing values during the data collection period for a total of
-\(535\) days.
+The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
+We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
 
 For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
-\textcite{Cheng2021-ex} propose two non-Euclidean distance estimators to enable manifold learning algorithms in statistical manifolds with each observation as a distribution.
-\textcite{Cheng2021-ex} use the same smart meter data for identifying outliers with kernel bandwidth estimation but fail to consider the distortion and information loss in the 2-dimensional embeddings given that the input data dimension is much higher.
-By introducing the Riemannian metric as the bandwidth matrix, we could take into account the distortion in the 2-D embedding and further improve the accuracy of the density estimation.
-
-In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households.
-Equation \eqref{eq:denestimator} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps have been clearly
-stated in the application section of \textcite{Cheng2021-ex} and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
-However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
+\textcite{Cheng2021-dh} propose two non-Euclidean distance estimators to be used in manifold learning algorithms in case of statistical manifolds with each observation as a distribution.
+The same Irish smart meter data is used for identifying outliers with kernel density estimation, but they fail to consider the distortion and information loss in the 2-dimensional embeddings.
+Therefore, we propose in this paper to apply distortion corrected KDE for anomaly detection as the downstream task of dimension reduction.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to get the 2-D embedding for kernel density estimation and anomaly detection.
+To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms with the use of distortion correction.
+We use the highest density regions plots to visualize the density estimates.
+However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
+Therefore, we will further examine the distributions of the anomalous households to study the typical or anomalous electricity usage patterns.
+
+Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the procedures here.
+First, for each household and each of the \(336\) half-hourly periods of the week, we build the empirical distributions and apply the total variation distance estimator for statistical manifold learning, which gives a 2-dimensional embedding for each dimension reduction method.
+Then we apply the kernel density estimator with distortion corrected for each method to the low-dimensional embedding and compare it with the KDE results in the Section 4.4 of \textcite{Cheng2021-dh}.
+The bandwidth parameter \(r\) is set as 1 in Equation \eqref{eq:denestimator}.
 
@@
 
-\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and two bandwidth selection methods in columns.}\label{fig:electricityoutliers}
+\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityoutliers}
 \end{figure}
@@
 \autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are close except that there are a few spikes for the one with fixed bandwidth.
-As to the anomalies, variable bandwidth tends to capture the unusual electricity demand volume when the usage is very low or high.
+The typical households in the top row are similar except that there are a few spikes for household \(1472\) when using KDE.
+As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume.
 It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
-In contrast, fixed bandwidth KDE is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
 
@@
 
-\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsfixed}
+\caption{Quantile region plots of electricity demand against the time of week for one typical household 1321 and two anomalies, 3243 and 2838.
+The quantile regions displayed in the plot are 99.0\%, 95.0\%, 75\%, and 50\%.
+Household 1321 has a repeated period-of-the-week electricity usage pattern slightly different for weekdays and weekends.}\label{fig:electricityhdrsfixed}
 \end{figure}
@@
 
-\caption{Two smart-meter demand examples, ID 1003 and ID 1539, from the Irish smart meter data set.}\label{fig:electricityhdrsvkde}
+\caption{Quantile region plots of electricity demand against the time of week for one typical household 1472 and two anomalies, 7049 and 5136.
+The day-of-the-week patterns is much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time related usage patterens.}\label{fig:electricityhdrsvkde}
 \end{figure}
@@
 
-In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities.
-The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms.
-By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions.
-We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.
-
-As an empirical example, we explore the distributions of different households and time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embeddings, and KDE with both variable and fixed bandwidth are used to get the density estimates.
-We compare both density estimates by looking at the distributions of the most typical households with the highest densities and the most anomalous households with the lowest densities.
+In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities.
+Compared with KDE, our distortion corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
+The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
+We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension redution algorithm.
+
+As an empirical example, we explore the distributions of \(3,639\) households and \(336\) time periods of the week in the Irish smart meter data.
+Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
+Without the ground truth density, we compare both density estimates by looking at the distributions of the most typical households with the highest densities and the anomalous households with the top two lowest densities.
 Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
@@
 There are several open questions to be explored.
-The first involves the selection of tuning parameters for the manifold algorithm so that a maximal level of embedding quality is achieved, where embedding quality is measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-ex}.
-The scaling of the Riemannian metric to get the closest range of the true densities is also worth exploring.
-The scale of distortion in each point in manifold learning could vary a lot.
-If we could smooth across all data points, eg.
-multiply the Riemannian matrix with the ratio between its determinant
-and the sum of all Riemannian matrix determinants, the global density estimates could be potentially smoothed.
-The density estimates on the edges of the whole data structure could be improved because most outer area points tend to be detected as outliers.
-The choice of manifold learning algorithms also has a large impact on the embedding accuracy, which we have seen will affect the density estimation and outlier detection.
-However, the outperformance of VKDE with Riemannian bandwidth than the fixed bandwidth has been shown in the higher dimensional simulation data and the electricity usage data, which are more related
-to real-life data sets.
+The first involves the selection of bandwidth parameters for kernel density estimation, which has been explored in many KDE-related literatures.
+The second one is related to the quality of the embedding from dimension reduction methods.
+Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitive relationship between embedding quality and density estimation accuracy is not immediate clear.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell wherther the distortion are corrected in the right way.
+The density estimates on the edges of the whole data structure could alos be explored because most outer area points tend to be detected as outliers.
+However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
 
@@
 
-\hypertarget{swissrollappe}{%
-\section{Appendix: Simulation with swiss roll mapping}\label{swissrollappe}}
-
-In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.
-
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \eqref{eq:swissroll}.
-The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in \autoref{fig:swissroll3d}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
-\begin{equation}
-\label{eq:swissroll}
-\left\{
-\begin{array}{lcl}
-X = X_1 \cos{X_1}, \\
-Y = X_2, \\
-Z = X_1 \sin{X_1}.
-\end{array}
-\right.
-\end{equation}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/scatterplot3d_swissroll_dc}
-
-}
-
-\caption{Scatterplot of the 3-d swiss roll data with colors indicating the true density via the swiss roll mapping in \autoref{eq:swissroll}.}\label{fig:swissroll3d}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08}
-
-}
-
-\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
-\end{figure}
-
-Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1\%, 5\%, 50\%, 99\%, \textgreater99\%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
-the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all
-manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
-
-\begin{table}
-
-\caption{\label{tab:srcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
-\midrule
-Variable bandwidth & 0.0696 & \textbf{0.400} & -0.2357 & 0.023 & \textbf{0.0138}\\
-Fixed bandwidth & \textbf{0.2798} & 0.351 & \textbf{0.0141} & \textbf{0.367} & -0.0110\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
-rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
-correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
-
 \printbibliography

### 2022-10-31 paper/paper.Rmd
polished up sec 2 and 3.2

@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds.
-Therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
@@
 Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
-Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$.
-However, it does not need to be the case.
+Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however this need not be the case.
 For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold.
@@
 
-While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors $\pmb{y}_i$ is different from the density on the manifold itself.
+Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.
 As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-Furthermore, any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
+To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term.
+A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density on the manifold itself.
+Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own.
+This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.
 
@@
 
-The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind this term.
+The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
@@
 Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space, $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$.
-Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travels along this geodesic at uniform speed $\|\pmb{v}\|$ in one unit of time.
+Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travelling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time.
 Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
@@
 For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
-While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
+While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
 This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.
@@
 One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\pmb{H}_j$.
-To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required because anomalies are identified as the points with lowest density.
+To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with lowest density.
 The entire workflow is summarized in \autoref{fig:vkde}.
@@
 To recover the true distribution requires the correct Jacobian term for the pushforward from $\pmb{v}$ to the volume form of twin peaks manifold.
-By treating the $\pmb{v}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\pmb{\Gamma}_i$ for $i=1,\dots,n$ where $\pmb{\Gamma}_i$ is the Riemannian of the coordinate system given by $\pmb{v}$, this notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction.
+By treating the $\pmb{v}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\pmb{\Gamma}_i$ for $i=1,\dots,n$ where $\pmb{\Gamma}_i$ is the Riemannian of the coordinate system given by $\pmb{v}$.
+This notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
 The true density on the manifold can be obtained as $f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals.
-Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' densities on the manifold.
+Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold.
 \autoref{fig:metadensity} shows the simulated $\pmb{v}$ with color indicating the true density of data on the manifold.
@@
 
-In \autoref{fig:tpisomapden}, we plot the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
+In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
 Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a $>99\%$ highest density region.
-The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a $99\%$ and $90\%$ HDR.
-Points lying outside squares (shown as triangles) will be misidentified as anomalies.
-For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the $99\%$ HDR), but are not identified as such.
-Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
+The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a $99\%$ highest density region.
+The middle shaded region contain anomalies that are true positives in the sense of not lying in a $90\%$ HDR.
+Points lying outside squares (shown as triangles) are incorrectly classified.
+For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the $99\%$ HDR, but are not classified as such (although they are true positives if a 90% HDR is used).
+Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
 \autoref{fig:tptsneden} shows the same plot but for t-SNE.
@@
 It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
-This is critical since conclusions will be robust to the choice of dimension reduction algorithm.
+This is critical since conclusions will be more robust to the choice of dimension reduction algorithm.
 
@@
 
-As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, then transform it into a 100-dimensional ambient space.
-To start with, we simulate vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ for $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, with the same means $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$ and different variance-covariance matrices $\pmb{\Sigma}_1 = diag(1,1,1,1)$ and $\pmb{\Sigma}_1 = diag(2,2,2,2)$.
-In order to manually add anomalies to be distant points from the core center, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-The fifth dimension is derived to satisfy the five-dimensional semi-hypersphere surface equation
+As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
+To start with, we simulate vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ for $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, with the same means $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$ and different variance-covariance matrices $\pmb{\Sigma}_1 = diag(1,1,1,1)$ and $\pmb{\Sigma}_2 = diag(2,2,2,2)$.
+The mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
+With this design, the observations from the second component tend to be outlying anomalies.
+The data are mapped to a hemisphere via the equation
 $v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path[^1] with the R package *tourr* [@Wickham2011-ir].
-The round and triangular point shapes index the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the kernel core $(0, 0, 0, 0, 0)^\prime$.
-It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the top outliers are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$.
+\autoref{fig:fivedmeta} shows scatterplot which is a single frame from a 5-D tour path[^1] animation using the R package *tourr* [@Wickham2011-ir].
+The round and triangular point shapes indicate the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
+$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the centre of the distribution.
+It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
 
@@
 To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to $\pmb{v}_i$ so that $\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N$.
-Next, we rotate the $100$-dimensional vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ with QR decomposition to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-Denote $\pmb{X}_0 = (\pmb{v}_1, \dots, \pmb{v}_N)$ with dimension $100 \times N$.
-The rotation matrix $Q$ is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$.
-The details are as follows.
-For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix.
-Matrix $\pmb{Q}$ satisfies
-$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in
-$\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
-Therefore, we take advantage of the matrix $Q$ as the rotation matrix where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the dimension reduction algorithms.
-It is worth noting that the intrinsic dimension of $X$ is known in this example as $d=4$.
+Next, we rotate the $100$-dimensional vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ by multiplying by a randomly generated rotation matrix.
+To generate the rotation matrix we first simulate elements from a uniform $(0,1)$ distribution, stack them into a $100\times 100$ matrix $\pmb{A}$ and then take the R matrix from the QR decomposition of $\pmb{A}$.
+Rotating the vectors results in input vectors that are no longer sparse.
+Nonetheless, the intrinsic dimension of this is still $d=4$.
 
@@
 The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example.
-In \autoref{fig:fivedisomapden}, the log-scale ranks of the estimated density from distortion corrected KDE shown in the left panel and KDE shown in the right panel are compared with the true rank of density for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to embedding dimensions within three.
-Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the kernel cores with outliers shown in darker colors.
-The grey shading area highlights the $>99\%$ highest density region area and points falling into this area are the correctly classified top $1\%$ anomalies.
-Darker points lying outside the shaded squares in the top left area are true anomalies but are misclassified as such.
-By comparing the left and right panel for each row, we could notice that KDE with distortion corrected outperforms KDE in classifying anomalies.
-For UMAP, almost all ground truth top $1\%$ anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected.
-It is worth noting that for DC-KDE in the left panel, there are many points having the same rank of estimated density inside the shading area, which is due to the fact that the ground truth density are almost zeros, with values less than $10^{-7}$.
-And these low-density points are all captured by our distortion correction methods regardless of the dimension reduction algorithms, which again shows the robustness of DC-KDE.
+In \autoref{fig:fivedisomapden}, the ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
+The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
+
+Figure \autoref{fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
+As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors.
+Comparing the left and right panel for each row, we notice that KDE with distortion corrected, has fewer misclassified observations, and therefore outperforms KDE.
+For UMAP, almost all ground truth anomalies (points outside a $99\%$ HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
+For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than $10^{-7}$) which leads to tied ranks.
+These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
 

### 2022-10-31 paper/paper.tex
polished up sec 2 and 3.2

@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Some readers familiar with kernel density estimation may not be as familiar with the nuances of manifolds.
-Therefore, before introducing our own estimator, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
@@
 Points on this manifold are denoted \(\pmb{p}\) in general, with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
-Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\).
-However, it does not need to be the case.
+Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however this need not be the case.
 For instance, \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
@@
 
-While one could in principle use a standard kernel density on the output from a dimension reduction algorithm, it must be noted that the density of the output vectors \(\pmb{y}_i\) is different from the density on the manifold itself.
+Kernel density estimators of this form can and have been applied directly on the output embedding \(\pmb{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
 As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-Furthermore, any analysis based on density estimates of the output embedding will be dependent on the choice of dimension reduction algorithm since each different algorithm will distort the density in a different way.
+To make this clear consider the simpler case computing the density after a change of variables \(\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d\), which involves a Jacobian term.
+A similar notion applies to a manifold embedding so that the density of the output vectors \(\pmb{y}_i\) differs from the density on the manifold itself.
+Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own.
+This motivates a kernel desnity estimate that corrects for the distortion induced by \(\psi\).
 
@@
 
-The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind this term.
+The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
 We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
@@
 Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space, \(\pmb{v}\in T_{\pmb{p}}M\), to points on the manifold, \(\pmb{q}\in M\).
-Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and travels along this geodesic at uniform speed \(\|\pmb{v}\|\) in one unit of time.
+Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and travelling along this geodesic at uniform speed \(\|\pmb{v}\|\) takes place in one unit of time.
 Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
@@
 For example, angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
-While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinates systems.
+While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
 This is precisely the situation when data on a manifold are mapped to \(d\)-dimensional Euclidean vectors \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) via a dimension reduction algorithm.
@@
 One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\pmb{H}_j\).
-To estimate the density at points that do not belong to the sampled points, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required because anomalies are identified as the points with lowest density.
+To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with lowest density.
 The entire workflow is summarized in \autoref{fig:vkde}.
@@
 To recover the true distribution requires the correct Jacobian term for the pushforward from \(\pmb{v}\) to the volume form of twin peaks manifold.
-By treating the \(\pmb{v}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\pmb{\Gamma}_i\) for \(i=1,\dots,n\) where \(\pmb{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\pmb{v}\), this notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction.
+By treating the \(\pmb{v}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\pmb{\Gamma}_i\) for \(i=1,\dots,n\) where \(\pmb{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\pmb{v}\).
+This notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
 The true density on the manifold can be obtained as \(f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals.
-Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' densities on the manifold.
+Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' for densities on the manifold.
 \autoref{fig:metadensity} shows the simulated \(\pmb{v}\) with color indicating the true density of data on the manifold.
@@
 Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identifies as such, where an anomaly is defined as a point not falling within a \(>99\%\) highest density region.
-The middle shaded and top right shaded regions respectively contain points that are anomalies in the sense of not falling within a \(99\%\) and \(90\%\) HDR.
-Points lying outside squares (shown as triangles) will be misidentified as anomalies.
-For example, the three red triangles in the middle left of the left panel are truly anomalies (lie outside the \(99\%\) HDR), but are not identified as such.
-Overall, the right panel contains many more misclassified anomalies showing that failing to apply distortion correction can have a severe impact on anomaly detection.
+The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a \(99\%\) highest density region.
+The middle shaded region contain anomalies that are true positives in the sense of not lying in a \(90\%\) HDR.
+Points lying outside squares (shown as triangles) are incorrectly classified.
+For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used).
+Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
 \autoref{fig:tptsneden} shows the same plot but for t-SNE.
@@
 
-As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, then transform it into a 100-dimensional ambient space.
-To start with, we simulate vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), with the same means \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\pmb{\Sigma}_1 = diag(1,1,1,1)\) and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\).
-In order to manually add anomalies to be distant points from the core center, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-The fifth dimension is derived to satisfy the five-dimensional semi-hypersphere surface equation
+As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
+To start with, we simulate vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), with the same means \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\pmb{\Sigma}_1 = diag(1,1,1,1)\) and \(\pmb{\Sigma}_2 = diag(2,2,2,2)\).
+The mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
+With this design, the observations from the second component tend to be outlying anomalies.
+The data are mapped to a hemisphere via the equation
 \(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
-\autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} with the R package \emph{tourr} \autocite{Wickham2011-ir}.
-The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
-\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the kernel core \((0, 0, 0, 0, 0)^\prime\).
-It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the top outliers are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\).
+\autoref{fig:fivedmeta} shows scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
+The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
+\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the centre of the distribution.
+It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
 
 To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to \(\pmb{v}_i\) so that \(\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\).
-Next, we rotate the \(100\)-dimensional vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) with QR decomposition to get rid of the zeros so that it could be passed to the manifold learning algorithms.
-Denote \(\pmb{X}_0 = (\pmb{v}_1, \dots, \pmb{v}_N)\) with dimension \(100 \times N\).
-The rotation matrix \(Q\) is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
-The details are as follows.
-For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix.
-Matrix \(\pmb{Q}\) satisfies
-\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in
-\(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
-Therefore, we take advantage of the matrix \(Q\) as the rotation matrix where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the dimension reduction algorithms.
-It is worth noting that the intrinsic dimension of \(X\) is known in this example as \(d=4\).
+Next, we rotate the \(100\)-dimensional vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) by multiplying by a randomly generated rotation matrix.
+To generate the rotation matrix we first simulate elements from a uniform \((0,1)\) distribution, stack them into a \(100\times 100\) matrix \(\pmb{A}\) and then take the R matrix from the QR decomposition of \(\pmb{A}\).
+Rotating the vectors results in input vectors that are no longer sparse.
+Nonetheless, the intrinsic dimension of this is still \(d=4\).
 
@@
 The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example.
-In \autoref{fig:fivedisomapden}, the log-scale ranks of the estimated density from distortion corrected KDE shown in the left panel and KDE shown in the right panel are compared with the true rank of density for four manifold learning embeddings, ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to embedding dimensions within three.
-Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the kernel cores with outliers shown in darker colors.
-The grey shading area highlights the \(>99\%\) highest density region area and points falling into this area are the correctly classified top \(1\%\) anomalies.
+In \autoref{fig:fivedisomapden}, the ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
+The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than ot equal to three.
+
+Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the centre of the distribution with outliers shown in darker colors.
+The grey shading area highlights the \(99\%\) highest density region area and points falling into this area are the correctly classified top \(1\%\) anomalies.
 Darker points lying outside the shaded squares in the top left area are true anomalies but are misclassified as such.

### 2022-10-31 R/Electricity2d_04_plots.R
Rename R files, change radius search to priority search

@@ -1,8 +1,6 @@
 # Same as mlann, to be modified
 ## Most Up-to-date script to use!
 # This script contains the code for plotting for smart meter data
-# To generate the loaded data `data/spdemand_3639id336tow.rda`, please run `source("data/electricity_02_quantiles_case1id1003.R")`
-# To generate the loaded data `data/spdemand_3639id_336tow_201length.rda`, please run `source("data/electricity_02_quantiles_case3alladd.R")`
 rm(list=ls())
 library(data.table)
 library(dtplyr)
@@ -18,67 +16,98 @@ library(pryr)
 library(colorspace)
 library(caret)
 library(tictoc)
+library(ggforce)
 Jmisc::sourceAll(here::here("R/sources")) 
 set.seed(1)
 eps <- 20
 nt <- 50
 annmethod <- "kdtree" # for ANN
 ntimes <- 10
-
-
-
-###------------------------------------------------------
-# For electricity_02_quantiles_case1id1003.R for single household ID 1003, run the following section
-nid <- 1
-ntow <- 336
-len <- 100 #201
-filename <- paste0(nid, "id_", ntow, "tow_", len, "quantile")
-# load(paste0('data/spdemand_3639id336tow.rda')) # (3639*336)*203
-load(paste0('data/sqrt_count_ratio_', nid, 'id', ntow, 'tow.rda'))
-train <- spdemand %>%
-  lazy_dt() %>%
-  # filter(id <= sort(unique(spdemand[,id]))[nid],
-  #        tow <= ntow) %>%
-  filter(id == 1003, tow <= ntow) %>%
-  dplyr::select(-c(id, tow)) %>%
-  as.data.table()
-# mem_change(rm(spdemand))
-N <- nrow(train)
-ids <- spdemand %>% dtplyr::lazy_dt() %>% pull(id) %>% unique() %>% sort()
-folder <- "data/1id_countratio/"
-distance <- c("euclidean", "manhattan")[1]
-###------------------------------------------------------
-
-
-# ###------------------------------------------------------
-# # For electricity_02_quantiles_case2allnotow.R for all 3639 households, run this section
-# nid <- 3639
-# ntow <- 336
-# len <- 201
-# filename <- paste0(nid, "id_notow_", len, "length")
-# load(paste0('data/spdemand_', filename, '.rda')) # 3639*(336*201+1)
-# train <- spdemand[, !"id"]
-# N <- nrow(train)
[…]

### 2022-10-31 R/Electricity2d_03_embed_tvd.R
Rename R files, change radius search to priority search

@@ -8,15 +8,15 @@ library(viridis)
 # library(ks)
 library(hdrcde)
 library(igraph)
-# library(seurat)
 library(matrixcalc)
-library(intRinsic)
+library(ggforce)
 Jmisc::sourceAll("R/sources")
 # scen <- 1
 scen <- as.numeric(commandArgs()[[6]])
 r <- 1
 
 load("data/half_count_ratio_3639id336tow_train.rda")
+load("data/ids.rda")
 # nid <- 3639
 # ntow <- 336
 # if(nid == 1) {
@@ -50,47 +50,89 @@ method <- c("annIsomap", "annLLE", "annLaplacianEigenmaps", "anntSNE", "annUMAP"
 annmethod <- "kdtree"
 distance <- c("euclidean", "manhattan")[2] # "manhattan" for all households
 treetype <- "kd"
-searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
+searchtype <- "priority" # "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
 radius <- 20 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
 
 gridsize <- 20
 opt.method <- c("AMISE", "MEAN", "SCALED")[3] # 3 ONLY FOR NOW, no scaling for Rn
 riem.scale <- 1 # tune parameter
+prob <- c(1, 10, 50, 95, 99)
+noutliers <- 20
 
-## ----isomap-------------------------------------------------------------------
+## ----embedding----------------------------------------------------------------
+paste("Embedding began at:", Sys.time())
+pars <- list(knn = k,
+             radius = radius,
+             eps = 0, # for true NN 
+             ndim = s,
+             get_geod = FALSE,
+             annmethod = "kdtree", 
+             nt = 50, 
+             search.k = 50,
+             nlinks = 16, 
+             ef.construction = 500, 
+             distance = distance,
+             treetype = treetype,
+             searchtype = searchtype)
+isomapnn <- embed(train, .method = method, knn = pars$k, radius = pars$radius,
+      annmethod = pars$annmethod,
+      eps = pars$eps,
+      nt = pars$nt,
+      nlinks = pars$nlinks, ef.construction = pars$ef.construction,
[…]

### 2022-10-31 paper/paper.Rmd
polished up to section 4

@@
 
-Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density in \autoref{fig:fivedisomapden}.
+Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density.
 The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example.
-In \autoref{fig:fivedisomapden}, the ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
+In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densites from DC-KDE and KDE are presented.
+For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
+However, for Laplacian Eigenmaps and UMAP there more distortion is induced throughdimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
+In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively,when distortion correction is applied.
+
+Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
+As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors.
+The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
 The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
@@
 
-Figure \autoref{fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
-As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors.
-Comparing the left and right panel for each row, we notice that KDE with distortion corrected, has fewer misclassified observations, and therefore outperforms KDE.
+Comparing the left and right panel for each row, we notice that KDE with distortion correction, has fewer misclassified observations, and therefore outperforms KDE.
 For UMAP, almost all ground truth anomalies (points outside a $99\%$ HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
@@
 ```
-
-In \autoref{tab:fivedcors}, the correlations between ranks of true densities and estimated densites from DC-KDE and KDE are presented.
-For ISOMAP and LLE, KDE gives higher rank correlation by less than 0.01, but both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP where there are more distortion in the dimension reduction process, the rank correlation for KDE is close to 0 while our distortion correction methods gives a relatively high correlation of 0.87 and 0.78 respectively.
 
@@
 
-Furthermore, when comparing across different dimension reduction algorithms in \autoref{fig:fivedisomapden}, almost all top outliers are correctly classified in the shading areas, which again shows the robustness of the proposed distortion correction methods to the use of dimension reduction algorithms.
-\autoref{tab:fourdhdrtable} further illustrutes the percentage of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions.
+\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions.
 Compared with KDE, the percentages are always higher when distortion correction methods are used.
 The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
-This could be due to the severe distortion usually induced by the UMAP algorithm, which can be seen from the $0\%$ in KDE when no distortion correction is used.
+This could be due to the severe distortion usually induced by the UMAP algorithm.
 
@@
 
-For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
-@Cheng2021-dh propose two non-Euclidean distance estimators to be used in manifold learning algorithms in case of statistical manifolds with each observation as a distribution.
-The same Irish smart meter data is used for identifying outliers with kernel density estimation, but they fail to consider the distortion and information loss in the 2-dimensional embeddings.
-Therefore, we propose in this paper to apply distortion corrected KDE for anomaly detection as the downstream task of dimension reduction.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to get the 2-D embedding for kernel density estimation and anomaly detection.
-To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms with the use of distortion correction.
-We use the highest density regions plots to visualize the density estimates.
+For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
+This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
+@Cheng2021-dh propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection.
+To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
+We use the highest density regions plot to visualize the density estimates.
 However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
-Therefore, we will further examine the distributions of the anomalous households to study the typical or anomalous electricity usage patterns.
 
-Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the procedures here.
-First, for each household and each of the $336$ half-hourly periods of the week, we build the empirical distributions and apply the total variation distance estimator for statistical manifold learning, which gives a 2-dimensional embedding for each dimension reduction method.
-Then we apply the kernel density estimator with distortion corrected for each method to the low-dimensional embedding and compare it with the KDE results in the Section 4.4 of @Cheng2021-dh.
+Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the process here.
+For each household, a discrete approximation to the distribution of electricity demand at each one of the $336$ half-hourly periods of the week is found.
+For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm.
 The bandwidth parameter $r$ is set as 1 in Equation \@ref(eq:denestimator).
@@
 
-\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are similar except that there are a few spikes for household $1472$ when using KDE.
-As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume.
-It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
-In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
+The typical households in the top row are similar except that there are a few spikes for household $1472$ (selected as most typical when using KDE without distortion correction).
+As for the anomalies, distortion corrected KDE tends to capture instance of unusual electricity demand where usage is very low with some sudden spikes such as ID 3243 or households with a very high baseline electricity usage only across a sub-period of time such as ID 2838.
+In contrast, KDE without distortion correction can select households that do not exhibit anomalous behavior at all, such as household 5146 and the selection of such houesholds as anomalies may be an artefact of distortion induced by a dimension reduction algorithm.
 
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}.
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde} which are respectively selected when distortion correction is and is not applied.
 Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
 This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE.
-As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week.
-The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh.
-These findings show the difference in finding typical and anomalous households with different kernel density estimation methods.
+As for the households selected when distortion correction is applied, the middle row in \autoref{fig:electricityhdrsfixed} shows a household with small spikes at noon but only on Tuesday and Sunday.
+The anomalies detected without distortion correction show an increasing electricity demand across the day of the week.
+These findings show the difference between anomalies detected with different kernel density estimation methods.
 

### 2022-10-31 paper/paper.tex
polished up to section 4

@@
 
-In \autoref{fig:tpisomapden}, we plot the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
+In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
 Data are presented on a log scale to highlight anomalies.
@@
 
-Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density in \autoref{fig:fivedisomapden}.
+Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density.
 The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example.
-In \autoref{fig:fivedisomapden}, the ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
+In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densites from DC-KDE and KDE are presented.
+For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
+However, for Laplacian Eigenmaps and UMAP there more distortion is induced throughdimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
+In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively,when distortion correction is applied.
+
+Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
+As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors.
+The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
 The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than ot equal to three.
-
-Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the centre of the distribution with outliers shown in darker colors.
-The grey shading area highlights the \(99\%\) highest density region area and points falling into this area are the correctly classified top \(1\%\) anomalies.
-Darker points lying outside the shaded squares in the top left area are true anomalies but are misclassified as such.
-By comparing the left and right panel for each row, we could notice that KDE with distortion corrected outperforms KDE in classifying anomalies.
-For UMAP, almost all ground truth top \(1\%\) anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected.
-It is worth noting that for DC-KDE in the left panel, there are many points having the same rank of estimated density inside the shading area, which is due to the fact that the ground truth density are almost zeros, with values less than \(10^{-7}\).
-And these low-density points are all captured by our distortion correction methods regardless of the dimension reduction algorithms, which again shows the robustness of DC-KDE.
+Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
+
+Comparing the left and right panel for each row, we notice that KDE with distortion correction, has fewer misclassified observations, and therefore outperforms KDE.
+For UMAP, almost all ground truth anomalies (points outside a \(99\%\) HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
+For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than \(10^{-7}\)) which leads to tied ranks.
+These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
 
@@
 \end{table}
-
-In \autoref{tab:fivedcors}, the correlations between ranks of true densities and estimated densites from DC-KDE and KDE are presented.
-For ISOMAP and LLE, KDE gives higher rank correlation by less than 0.01, but both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP where there are more distortion in the dimension reduction process, the rank correlation for KDE is close to 0 while our distortion correction methods gives a relatively high correlation of 0.87 and 0.78 respectively.
 
@@
 
-Furthermore, when comparing across different dimension reduction algorithms in \autoref{fig:fivedisomapden}, almost all top outliers are correctly classified in the shading areas, which again shows the robustness of the proposed distortion correction methods to the use of dimension reduction algorithms.
-\autoref{tab:fourdhdrtable} further illustrutes the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
+\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
 Compared with KDE, the percentages are always higher when distortion correction methods are used.
 The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
-This could be due to the severe distortion usually induced by the UMAP algorithm, which can be seen from the \(0\%\) in KDE when no distortion correction is used.
+This could be due to the severe distortion usually induced by the UMAP algorithm.
 
@@
 
-For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
-\textcite{Cheng2021-dh} propose two non-Euclidean distance estimators to be used in manifold learning algorithms in case of statistical manifolds with each observation as a distribution.
-The same Irish smart meter data is used for identifying outliers with kernel density estimation, but they fail to consider the distortion and information loss in the 2-dimensional embeddings.
-Therefore, we propose in this paper to apply distortion corrected KDE for anomaly detection as the downstream task of dimension reduction.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to get the 2-D embedding for kernel density estimation and anomaly detection.
-To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms with the use of distortion correction.
-We use the highest density regions plots to visualize the density estimates.
+For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
+This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
+\textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection.
+To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
+We use the highest density regions plot to visualize the density estimates.
 However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
-Therefore, we will further examine the distributions of the anomalous households to study the typical or anomalous electricity usage patterns.
-
-Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the procedures here.
-First, for each household and each of the \(336\) half-hourly periods of the week, we build the empirical distributions and apply the total variation distance estimator for statistical manifold learning, which gives a 2-dimensional embedding for each dimension reduction method.
-Then we apply the kernel density estimator with distortion corrected for each method to the low-dimensional embedding and compare it with the KDE results in the Section 4.4 of \textcite{Cheng2021-dh}.
+
+Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here.
+For each household, a discrete approximation to the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
+For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm.
 The bandwidth parameter \(r\) is set as 1 in Equation \eqref{eq:denestimator}.
@@
 
-\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
+\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
 The typical households in the top row are similar except that there are a few spikes for household \(1472\) when using KDE.

### 2022-10-31 paper/paper.tex
post merge commit

@@
 \autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are similar except that there are a few spikes for household \(1472\) when using KDE.
-As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume.
-It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838.
-In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
+The typical households in the top row are similar except that there are a few spikes for household \(1472\) (selected as most typical when using KDE without distortion correction).
+As for the anomalies, distortion corrected KDE tends to capture instance of unusual electricity demand where usage is very low with some sudden spikes such as ID 3243 or households with a very high baseline electricity usage only across a sub-period of time such as ID 2838.
+In contrast, KDE without distortion correction can select households that do not exhibit anomalous behavior at all, such as household 5146 and the selection of such houesholds as anomalies may be an artefact of distortion induced by a dimension reduction algorithm.
 

### 2022-11-06 paper/paper.tex
wrote intro

@@
 
-Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of \(\mathbb{R}^d\).
-Applications of kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on.
-However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than \(\mathbb{R}^d\), kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted.
-\textcite{Pelletier2005-vu} propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known.
-Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold.
-Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than \(\mathbb{R}^d\).
-Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function.
-This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.
-
-For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}.
-The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}.
-In this paper, we focus on the multivariate kernel density estimation.
-
-Note that a fixed bandwidth matrix \(\pmb{H}\) is a global smoothing parameter for all data points.
-However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-The bandwidth is varied depending on either the location of the sample points {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]} or that of the estimated points {[}balloon estimator; \textcite{Terrell1992-ut}{]}.
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} {[}VKDE; Section 6.6 of \textcite{Scott2015-vl}{]} unless otherwise stated.
-However, these kernel density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature \autocite{Henry2009-ll}.
-However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space.
-Therefore, the kernel density estimator from \textcite{Pelletier2005-vu} is not applicable because the geodesic distance and the volume density function are unknown.
-This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-dh}), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
-
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
-One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold \autocite{Goldberg2008-co}.
-To tackle this problem and measure the distortion incurred in manifold learning,
-\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others.
-The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.
-
-By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding.
-Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
-Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information.
-Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
-These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
-Similar to \textcite{Cheng2021-dh}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
+Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
+Applications of KDE include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
+For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well known curse of dimensionality.
+However the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d<<m\).
+The problem of KDE on manifolds has been well studied see \textcite{Pelletier2005-vu} (FIND SOME MORE REFS) for details.
+However, in these setting it is assumed that the underlying manifold in known, which is rarely true in practice.
+Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
+
+When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
+These algorithms which include ISOMAP, LLE, t-SNE and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\).
+Details of these algorithms are summarised (OTHER REF) in \textcite{Cheng2021-dh}.
+One could easily apply KDE to the output vectors \(\pmb{y}\).
+However this would be a misleading representation of the density of points as they lie on the manifold.
+To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
+
+\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M\]
+
+where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\).
+By a change of variables, this would imply the following density for \(\pmb{y}\)
+
+\[Pr(\pmb{y}\in\psi(\mathcal{A}))=\int_{\psi(\mathcal{A})} f(\psi^{-1}(\pmb{y}))|g(\pmb{y})|^{-1/2}d\pmb{y}\]
+where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(g(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
+It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|g(\pmb{y})|^{-1/2}\) which can be considered roughly analogous to the Jacobian determinant.
+Since \(\psi\) will typically be a non-linear function, the value of \(|g(\pmb{y})|^{-1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
+Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
+
+In this paper we propose a new KDE.
+Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm.
+For this reason we refer to it as the distortion corrected KDE (DC-KDE).
+Critical to our estimator, is obtaining an estimate of \(|g(\pmb{y})|^{-1/2}\), for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density.
+In a number of simulation studies we demonstrate two benefits that arise when using our new estimator.
+The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
+The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.
+
+In terms of data visualisation, our approach allows a two or three dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead this is an artefact of the dimension reduction algorithm `squeezing' points together.
+This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
+Furthermore, our estimator can be used for manifolds with elements that are are a variety of objects, for instance rotational frames or images.
+In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
+Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to an extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.
 
@@
 In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds.
-Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection.
-\autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space.
-Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
-\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset.
+Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
+\autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
+\autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
 Conclusions and discussions are presented in \autoref{conclusion}.
-Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.
+Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
 

### 2022-11-06 paper/paper.Rmd
wrote intro

@@
 
+Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Chen2017-dw and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
+Applications of KDE include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq].
+For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well known curse of dimensionality.
+However the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d<<m$.
+The problem of KDE on manifolds has been well studied see @Pelletier2005-vu (FIND SOME MORE REFS) for details.
+However, in these setting it is assumed that the underlying manifold in known, which is rarely true in practice.
+Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
+
+When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
+These algorithms which include ISOMAP, LLE, t-SNE and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$.
+Details of these algorithms are summarised (OTHER REF) in @Cheng2021-dh.
+One could easily apply KDE to the output vectors $\pmb{y}$.
+However this would be a misleading representation of the density of points as they lie on the manifold.
+To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
+
+$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M$$
+
+where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$.
+By a change of variables, this would imply the following density for $\pmb{y}$
+
+$$Pr(\pmb{y}\in\psi(\mathcal{A}))=\int_{\psi(\mathcal{A})} f(\psi^{-1}(\pmb{y}))|g(\pmb{y})|^{-1/2}d\pmb{y}$$
+where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $g(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
+It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|g(\pmb{y})|^{-1/2}$ which can be considered roughly analogous to the Jacobian determinant.
+Since $\psi$ will typically be a non-linear function, the value of $|g(\pmb{y})|^{-1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
+Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
+
+In this paper we propose a new KDE.
+Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm.
+For this reason we refer to it as the distortion corrected KDE (DC-KDE).
+Critical to our estimator, is obtaining an estimate of $|g(\pmb{y})|^{-1/2}$, for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density.
+In a number of simulation studies we demonstrate two benefits that arise when using our new estimator.
+The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
+The second is that when the estimates are found using different dimension reduction algorithms (e.g.
+ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.
+
+In terms of data visualisation, our approach allows a two or three dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high density region or if instead this is an artefact of the dimension reduction algorithm 'squeezing' points together.
+This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
+Furthermore, our estimator can be used for manifolds with elements that are are a variety of objects, for instance rotational frames or images.
+In our application we consider a probabilistic manifold whose elements are  densities, meaning that we estimate a density of densities.
+Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to an extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].
+
+The rest of the paper is organized as follows.
+In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
+Our estimator combines insights from two existing papers, the first is @Pelletier2005-vu who propose a KDE for data on manifolds that are known, the second is @Perrault-Joncas2013-pq who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
+\autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional 'twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
+\autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
+Conclusions and discussions are presented in \autoref{conclusion}.
+Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
+
+<!--
+
+OLD:
+
 Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
@@
 
-<!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
+Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
 It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-dh], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
@@
 To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
--->
 
 For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
-<!-- There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].  -->
+There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
 Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
@@
 Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.
+
+-->
 

### 2022-11-06 paper/paper.Rmd
Sec 4 and 5 done

@@
   `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
-  kableExtra::kbl(caption = "Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
+  kableExtra::kbl(caption = "Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
                   col.names = c(rep(c("DC-KDE", "KDE"), 4))
@@
 
-\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions.
-Compared with KDE, the percentages are always higher when distortion correction methods are used.
-The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
+\autoref{tab:fourdhdrtable} further illustrates the proportions of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions.
+Compared with KDE, the proportions are always higher when distortion correction methods are used.
+The proportions with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
 This could be due to the severe distortion usually induced by the UMAP algorithm.
@@
 @Cheng2021-dh propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a $2$-dimensional embedding for kernel density estimation and anomaly detection.
 To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
@@
 For each household, a discrete approximation to the distribution of electricity demand at each one of the $336$ half-hourly periods of the week is found.
-For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm.
-The bandwidth parameter $r$ is set as 1 in Equation \@ref(eq:denestimator).
+For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
+The number of nearest neighbors in manifold learning $K$ is set as $100$ and the bandwidth paramter $r=180$ to match the distances between the distributions of households.
 
 ```{r electricityembedding, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=F}
-# knitr::include_graphics("figures/")
-```
-
-<!-- The results are summarized in \autoref{fig:electricityembedding}.
-TODO -->
-
-<!-- In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-dh in the statistical manifold learning to get the 2-D embedding of all households.
-Equation \@ref(eq:denestimator) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers.
-The data processing steps have been clearly
-stated in the application section of @Cheng2021-dh and they are skipped here.
-Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}.
-However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
- -->
+The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
+knitr::include_graphics("figures/Electricity_2d_5levels_outliers_comparison_5ml_r180.png")
+```
+```{r electricityoutliers}
+load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
+tabletop20
+```
+
+\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
+There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions.
+The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
+First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
+When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE and UMAP with DC-KDE compared with the KDE.
+As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identifies as anomalies.
+The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
+Again, KDE with distortion correction methods find more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
+Without distortion correction, the density estimates are different not only in scaling, but also shows inconsistency in the rank of density estimates.
+
+```{r electricityhdrpercent}
+load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
+table_outlierpercent
+```
+```{r electricityrank, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
+The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
+knitr::include_graphics("figures/Electricity_2d_rankplot_3639id_umapvs4ml_radius20_r180.png")
+```
+
+Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households.
+Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
+The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}.
+\autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) in $>99\%$ and $99\%$ HDRs that are also identified in different dimension reduction methods.
+Both DC-KDE and KDE are shown in the table.
+Except the lower percentage of ISOMAP in $>99\%$ HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
+The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than $0.5$.
+\autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
+The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}.
+Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.
+
 ```{r electricityoutliers, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.", echo = FALSE, eval=T}
-knitr::include_graphics("figures/kde_2densities_raw_electricity.png")
-```
-
-\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are similar except that there are a few spikes for household $1472$ (selected as most typical when using KDE without distortion correction).
-As for the anomalies, distortion corrected KDE tends to capture instance of unusual electricity demand where usage is very low with some sudden spikes such as ID 3243 or households with a very high baseline electricity usage only across a sub-period of time such as ID 2838.
-In contrast, KDE without distortion correction can select households that do not exhibit anomalous behavior at all, such as household 5146 and the selection of such houesholds as anomalies may be an artefact of distortion induced by a dimension reduction algorithm.
-
-```{r electricityhdrsfixed, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of week for one typical household 1321 and two anomalies, 3243 and 2838.
-The quantile regions displayed in the plot are 99.0\\%, 95.0\\%, 75\\%, and 50\\%.
-Household 1321 has a repeated period-of-the-week electricity usage pattern slightly different for weekdays and weekends.", echo = FALSE, eval=T}
-knitr::include_graphics("figures/hdr_electricity_hdrbox_3id_7dow.png")
-```
+knitr::include_graphics("figures/Electricity_2d_compare4id_1typical_3anomalous.png")
+```
+
+Now let us take a close look at the typical and anomalous distributions of households.
+According to \autore{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration.
+Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityoutliers}.
+Compared with ID 2396, the other three households show different usage patterns.
+ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh.
+As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household.
+The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
+It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.
 
@@
 The day-of-the-week patterns is much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time related usage patterens.", echo = FALSE, eval=T}
-knitr::include_graphics("figures/electricity_hdrbox_3id_7dow.png")
-```
-
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde} which are respectively selected when distortion correction is and is not applied.
-Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE.
-As for the households selected when distortion correction is applied, the middle row in \autoref{fig:electricityhdrsfixed} shows a household with small spikes at noon but only on Tuesday and Sunday.
-The anomalies detected without distortion correction show an increasing electricity demand across the day of the week.
-These findings show the difference between anomalies detected with different kernel density estimation methods.
+knitr::include_graphics("figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous.png")
+```
+
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityoutliers}.
+The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
+This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
+As for the bottom two households, the middle row in \autoref{fig:electricityhdrsfixed} shows a distribution with zero median and a $75\%$ quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
+These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
 
@@
 Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
-Without the ground truth density, we compare both density estimates by looking at the distributions of the most typical households with the highest densities and the anomalous households with the top two lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
+Without the ground truth density, we assume that UMAP gives the true densities and compare both density estimation methods by looking at the distributions of the most typical households with the highest densities and three anomalous households that are identified most times.
+The distortion correction methods could identify the typical households with repeated time-of-the-week usage patterns, while the distributions of the outliers are anomalous in different ways.
 
 There are several open questions to be explored.
-The first involves the selection of bandwidth parameters for kernel density estimation, which has been explored in many KDE-related literatures.
-The second one is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitive relationship between embedding quality and density estimation accuracy is not immediate clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell wherther the distortion are corrected in the right way.
-The density estimates on the edges of the whole data structure could alos be explored because most outer area points tend to be detected as outliers.
+The first involves the selection of bandwidth parameters for kernel density estimation, which has been widely explored in many KDE-related literature.
+The second one is that the proposed density estimation method is only limited to the embedding points.
+Although for the downstream task of anomaly detection, only density estimated on these points are needed, it would be worth exploring the out-of-sample density estimation in the future.
+The last question is related to the quality of the embedding from dimension reduction methods.
+Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediate clear.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion are corrected in the right way.
+The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
 However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
 
-<!-- kernel types discussion? -->
+<!-- kernel types discussion -->
 
@@
 <!-- \clearpage -->
-
 <!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
-
 <!-- %\pagenumbering{arabic} -->
-
 <!-- %\def\thepage{A\arabic{page}} -->

### 2022-11-06 R/Electricity2d_04_plots.R

@@ -316,7 +316,7 @@ table_outlierpercent <-
 hdrtable %>% 
   head(n = 2) %>% 
   `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
-  kableExtra::kbl(caption = "Percentage of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
+  kableExtra::kbl(caption = "Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
                   # col.names = c(rep(c("DC-KDE", "KDE"), 4))
   ) %>%
   kable_classic("striped") %>% 

### 2022-11-06 paper/paper.Rmd
Update PDF, table electricityoutliers hash for color codes to be solved

@@
 knitr::opts_chunk$set(
-  echo = TRUE,
+  echo = FALSE,
   cache = T,
@@
   warning = FALSE,
-  eval = TRUE
-  # include = FALSE
+  eval = TRUE,
+  include = TRUE
 )
@@
 library(here)
-# library(hdrcde)
-# library(car)
 library(patchwork)
@@
 ```
-```{r electricityoutliers}
-load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
-tabletop20
+```{r electricityoutliers, eval=F}
+load("figures/Electricity_2d_table_outlier_idcompare_5ml_radius20_r180.rda")
+# f20cols <- paste0("\\", f20cols)
+# tabletop20 <- 
+  f20 %>% 
+  kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding.
+The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
+The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, col.names = NULL, align=rep('c', 10)) %>%
+  # kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = TRUE) %>%
+  add_header_above(c(" " = 1, rep(c("ISOMAP" = 1, "LLE" = 1, "Laplacian Eigenmaps" = 1, "t-SNE" = 1, "UMAP" = 1), 2))) %>% 
+  add_header_above(c(" " = 1, "DC-KDE" = 5, "KDE" = 5)) %>%
+  # row_spec(1, bold = TRUE) %>% 
+  column_spec(1, bold = FALSE) %>% 
+  column_spec(2, bold = freqmax[,2], color = f20cols[,2]) %>% 
+  column_spec(3, bold = freqmax[,3], color = f20cols[,3]) %>% 
+  column_spec(4, bold = freqmax[,4], color = f20cols[,4]) %>% 
+  column_spec(5, bold = freqmax[,5], color = f20cols[,5]) %>% 
+  column_spec(6, bold = freqmax[,6], color = f20cols[,6]) %>% 
+  column_spec(7, bold = freqmax[,7], color = f20cols[,7], border_left = TRUE) %>% 
+  column_spec(8, bold = freqmax[,8], color = f20cols[,8]) %>% # background = grey(0.9)
+  column_spec(9, bold = freqmax[,9], color = f20cols[,9]) %>% 
+  column_spec(10, bold = freqmax[,10], color = f20cols[,10]) %>% 
+  column_spec(11, bold = freqmax[,11], color = f20cols[,11])
+# tabletop20
 ```
@@
 load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
-table_outlierpercent
+# table_outlierpercent <- 
+hdrtable %>% 
+  head(n = 2) %>% 
+  `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
+  kableExtra::kbl(caption = "Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
+                  # col.names = c(rep(c("DC-KDE", "KDE"), 4))
+  ) %>%
+  # kable_classic("striped") %>%
+  kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = TRUE) %>%
+  add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
+  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "t-SNE" = 2)) %>% 
+  # column_spec(1, bold = TRUE) %>% 
+  # add_indent(1:6) %>% 
+  column_spec(2, bold = hdrmax[,1]) %>% 
+  column_spec(3, bold = hdrmax[,2]) %>% 
+  column_spec(4, bold = hdrmax[,3]) %>% 
+  column_spec(5, bold = hdrmax[,4]) %>% 
+  column_spec(6, bold = hdrmax[,5]) %>% 
+  column_spec(7, bold = hdrmax[,6]) %>% 
+  column_spec(8, bold = hdrmax[,7]) %>% 
+  column_spec(2*(1:4), border_left = T, background = grey(0.9))
+# table_outlierpercent
 ```
@@
 
-```{r electricityoutliers, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.", echo = FALSE, eval=T}
+```{r electricityraw, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/Electricity_2d_compare4id_1typical_3anomalous.png")
@@
 Now let us take a close look at the typical and anomalous distributions of households.
-According to \autore{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration.
-Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityoutliers}.
+According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration.
+Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
 Compared with ID 2396, the other three households show different usage patterns.

### 2022-11-06 paper/paper.tex
Update PDF, table electricityoutliers hash for color codes to be solved

@@
 
-\caption{\label{tab:fourdhdrtable}Percentage comparison of correct highest density regions in density estimation of four manifold learning embeddings.}
+\caption{\label{tab:fourdhdrtable}Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.}
 \centering
@@
 
-\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
-Compared with KDE, the percentages are always higher when distortion correction methods are used.
-The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
+\autoref{tab:fourdhdrtable} further illustrates the proportions of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
+Compared with KDE, the proportions are always higher when distortion correction methods are used.
+The proportions with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
 This could be due to the severe distortion usually induced by the UMAP algorithm.
@@
 \textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection.
 To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
@@
 For each household, a discrete approximation to the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
-For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm.
-The bandwidth parameter \(r\) is set as 1 in Equation \eqref{eq:denestimator}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/kde_2densities_raw_electricity}
-
-}
-
-\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityoutliers}
-\end{figure}
-
-\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
-The typical households in the top row are similar except that there are a few spikes for household \(1472\) (selected as most typical when using KDE without distortion correction).
-As for the anomalies, distortion corrected KDE tends to capture instance of unusual electricity demand where usage is very low with some sudden spikes such as ID 3243 or households with a very high baseline electricity usage only across a sub-period of time such as ID 2838.
-In contrast, KDE without distortion correction can select households that do not exhibit anomalous behavior at all, such as household 5146 and the selection of such houesholds as anomalies may be an artefact of distortion induced by a dimension reduction algorithm.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/hdr_electricity_hdrbox_3id_7dow}
-
-}
-
-\caption{Quantile region plots of electricity demand against the time of week for one typical household 1321 and two anomalies, 3243 and 2838.
-The quantile regions displayed in the plot are 99.0\%, 95.0\%, 75\%, and 50\%.
-Household 1321 has a repeated period-of-the-week electricity usage pattern slightly different for weekdays and weekends.}\label{fig:electricityhdrsfixed}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/electricity_hdrbox_3id_7dow}
+For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
+The number of nearest neighbors in manifold learning \(K\) is set as \(100\) and the bandwidth paramter \(r=180\) to match the distances between the distributions of households.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_5levels_outliers_comparison_5ml_r180}
+
+}
+
+\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
+The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityembedding}
+\end{figure}
+\begin{table}
+
+\caption{\label{tab:electricityoutliers}Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding.
+The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
+The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
+\centering
+\begin{tabu} to \linewidth {>{}c>{}c>{}c>{}c>{}c>{}c|>{}c>{}c>{}c>{}c>{}c}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{5}{c}{DC-KDE} & \multicolumn{5}{c}{KDE} \\
+\cmidrule(l{3pt}r{3pt}){2-6} \cmidrule(l{3pt}r{3pt}){7-11}
+\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ISOMAP} & \multicolumn{1}{c}{LLE} & \multicolumn{1}{c}{Laplacian Eigenmaps} & \multicolumn{1}{c}{t-SNE} & \multicolumn{1}{c}{UMAP} & \multicolumn{1}{c}{ISOMAP} & \multicolumn{1}{c}{LLE} & \multicolumn{1}{c}{Laplacian Eigenmaps} & \multicolumn{1}{c}{t-SNE} & \multicolumn{1}{c}{UMAP} \\
+\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8} \cmidrule(l{3pt}r{3pt}){9-9} \cmidrule(l{3pt}r{3pt}){10-10} \cmidrule(l{3pt}r{3pt}){11-11}
+1 & \textcolor{black}{6317} & \textcolor{black}{7049} & \textcolor{black}{3161} & \textcolor{black}{4826} & \textcolor{black}{4546} & \textcolor{black}{6317} & \textcolor{black}{1881} & \textcolor{black}{6317} & \textcolor{black}{4190} & \textcolor{black}{6533}\\
+2 & \textcolor{0}{1353} & \textcolor[HTML]{08306B}{\textbf{4546}} & \textcolor[HTML]{08306B}{\textbf{5269}} & \textcolor{0}{1069} & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{4884} & \textcolor{0}{6199} & \textcolor{0}{6643} & \textcolor{0}{6817} & \textcolor{0}{5899}\\
+3 & \textcolor{0}{4884} & \textcolor[HTML]{08306B}{\textbf{1881}} & \textcolor{0}{4875} & \textcolor{0}{4546} & \textcolor[HTML]{08306B}{\textbf{1069}} & \textcolor{0}{5092} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor{0}{6764} & \textcolor{0}{6492} & \textcolor{0}{6345}\\
+4 & \textcolor{0}{4546} & \textcolor{0}{6952} & \textcolor{0}{7300} & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{4862} & \textcolor{0}{1353} & \textcolor[HTML]{6BAED6}{\textbf{4233}} & \textcolor{0}{1058} & \textcolor{0}{2309} & \textcolor{0}{1884}\\
+5 & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{6838} & \textcolor{0}{4880} & \textcolor[HTML]{08306B}{\textbf{1533}} & \textcolor{0}{3639} & \textcolor{0}{5580} & \textcolor{0}{7049} & \textcolor{0}{5105} & \textcolor{0}{2546} & \textcolor{0}{3278}\\
+\addlinespace
+6 & \textcolor[HTML]{08306B}{\textbf{1366}} & \textcolor{0}{3874} & \textcolor{0}{5599} & \textcolor[HTML]{6BAED6}{\textbf{4265}} & \textcolor{0}{1964} & \textcolor{0}{5594} & \textcolor[HTML]{08306B}{\textbf{1821}} & \textcolor{0}{6577} & \textcolor{0}{2907} & \textcolor{0}{5463}\\
+7 & \textcolor{0}{3161} & \textcolor{0}{1069} & \textcolor{0}{3222} & \textcolor{0}{4862} & \textcolor[HTML]{6BAED6}{\textbf{2249}} & \textcolor{0}{2576} & \textcolor{0}{4546} & \textcolor{0}{3880} & \textcolor{0}{4536} & \textcolor{0}{5504}\\
+8 & \textcolor[HTML]{08306B}{\textbf{1074}} & \textcolor{0}{2249} & \textcolor{0}{1776} & \textcolor{0}{3639} & \textcolor{0}{6199} & \textcolor{0}{1366} & \textcolor[HTML]{08306B}{\textbf{3431}} & \textcolor{0}{5358} & \textcolor{0}{6898} & \textcolor{0}{1782}\\
+9 & \textcolor{0}{6681} & \textcolor{0}{1964} & \textcolor{0}{1161} & \textcolor{0}{5546} & \textcolor[HTML]{6BAED6}{\textbf{3892}} & \textcolor{0}{6681} & \textcolor{0}{1569} & \textcolor{0}{1366} & \textcolor{0}{4128} & \textcolor{0}{5965}\\
+10 & \textcolor{0}{1701} & \textcolor[HTML]{6BAED6}{\textbf{1569}} & \textcolor{0}{3655} & \textcolor{0}{4896} & \textcolor{0}{6209} & \textcolor{0}{5291} & \textcolor{0}{1964} & \textcolor{0}{5545} & \textcolor{0}{7368} & \textcolor{0}{6081}\\
+\addlinespace
+11 & \textcolor{0}{3136} & \textcolor{0}{1533} & \textcolor{0}{6169} & \textcolor{0}{4555} & \textcolor{0}{6952} & \textcolor{0}{3161} & \textcolor[HTML]{6BAED6}{\textbf{6568}} & \textcolor{0}{2160} & \textcolor{0}{2232} & \textcolor{0}{4894}\\
+12 & \textcolor{0}{1881} & \textcolor[HTML]{6BAED6}{\textbf{4555}} & \textcolor{0}{4830} & \textcolor{0}{6199} & \textcolor{0}{4896} & \textcolor[HTML]{08306B}{\textbf{1701}} & \textcolor{0}{7300} & \textcolor{0}{6585} & \textcolor{0}{6273} & \textcolor{0}{5937}\\
+13 & \textcolor{0}{6785} & \textcolor{0}{6526} & \textcolor{0}{2284} & \textcolor[HTML]{6BAED6}{\textbf{6952}} & \textcolor{0}{6838} & \textcolor{0}{4749} & \textcolor{0}{7438} & \textcolor{0}{5824} & \textcolor{0}{1557} & \textcolor{0}{5522}\\
+14 & \textcolor{0}{5092} & \textcolor{0}{3639} & \textcolor{0}{6526} & \textcolor{0}{6838} & \textcolor{0}{1533} & \textcolor{0}{6699} & \textcolor{0}{4401} & \textcolor{0}{4932} & \textcolor{0}{3500} & \textcolor{0}{5580}\\
+15 & \textcolor{0}{5269} & \textcolor{0}{6199} & \textcolor{0}{1694} & \textcolor{0}{1964} & \textcolor[HTML]{6BAED6}{\textbf{1569}} & \textcolor{0}{5504} & \textcolor{0}{3434} & \textcolor{0}{6410} & \textcolor{0}{1932} & \textcolor{0}{6902}\\
+\addlinespace
+16 & \textcolor{0}{2576} & \textcolor[HTML]{6BAED6}{\textbf{5531}} & \textcolor{0}{3431} & \textcolor[HTML]{6BAED6}{\textbf{7104}} & \textcolor{0}{3213} & \textcolor{0}{3170} & \textcolor{0}{2249} & \textcolor{0}{3023} & \textcolor{0}{3119} & \textcolor{0}{7243}\\
+17 & \textcolor{0}{1262} & \textcolor{0}{4875} & \textcolor{0}{2523} & \textcolor{0}{3892} & \textcolor{0}{4401} & \textcolor{0}{5899} & \textcolor{0}{3213} & \textcolor{0}{1497} & \textcolor{0}{3809} & \textcolor{0}{6699}\\
+18 & \textcolor{0}{5966} & \textcolor{0}{5356} & \textcolor{0}{5356} & \textcolor{0}{3161} & \textcolor{0}{6766} & \textcolor{0}{2658} & \textcolor{0}{3161} & \textcolor{0}{7293} & \textcolor{0}{6108} & \textcolor{0}{5549}\\
+19 & \textcolor{0}{5937} & \textcolor[HTML]{6BAED6}{\textbf{3892}} & \textcolor[HTML]{6BAED6}{\textbf{2233}} & \textcolor[HTML]{08306B}{\textbf{6799}} & \textcolor{0}{4265} & \textcolor{0}{5189} & \textcolor[HTML]{08306B}{\textbf{1533}} & \textcolor{0}{5580} & \textcolor{0}{4516} & \textcolor{0}{5048}\\
+20 & \textcolor{0}{5356} & \textcolor{0}{4896} & \textcolor{0}{5125} & \textcolor{0}{6209} & \textcolor{0}{6799} & \textcolor{0}{2669} & \textcolor[HTML]{6BAED6}{\textbf{4555}} & \textcolor{0}{6224} & \textcolor{0}{5937} & \textcolor{0}{1923}\\
+\bottomrule
+\end{tabu}
+\end{table}
+
+\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
+There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions.
+The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
+First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
+When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE and UMAP with DC-KDE compared with the KDE.
+As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identifies as anomalies.
+The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
+Again, KDE with distortion correction methods find more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
+Without distortion correction, the density estimates are different not only in scaling, but also shows inconsistency in the rank of density estimates.
+
+\begin{table}
+
+\caption{\label{tab:electricityhdrpercent}Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabu} to \linewidth {>{\raggedright}X|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}c>{\centering}X}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{t-SNE} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
+\multicolumn{1}{c}{HDRs} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} \\
+\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8} \cmidrule(l{3pt}r{3pt}){9-9}
+>99\% HDR & \cellcolor[HTML]{E6E6E6}{0.189} & 0.243 & \cellcolor[HTML]{E6E6E6}{0.189} & 0.027 & \cellcolor[HTML]{E6E6E6}{0.622} & 0.000 & \cellcolor[HTML]{E6E6E6}{0.865} & 0.108\\
+99\% HDR & \cellcolor[HTML]{E6E6E6}{0.359} & \textbf{0.248} & \cellcolor[HTML]{E6E6E6}{\textbf{0.503}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.697}} & 0.007 & \cellcolor[HTML]{E6E6E6}{\textbf{0.628}} & 0.083\\
+\bottomrule
+\end{tabu}}
+\end{table}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_rankplot_3639id_umapvs4ml_radius20_r180}
+
+}
+
+\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
+The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityrank}
+\end{figure}
+
+Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households.
+Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
+The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}.
+\autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) in \(>99\%\) and \(99\%\) HDRs that are also identified in different dimension reduction methods.
+Both DC-KDE and KDE are shown in the table.
+Except the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
+The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than \(0.5\).
+\autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
+The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}.
+Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_compare4id_1typical_3anomalous}
+
+}
+
+\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityraw}
+\end{figure}
+
+Now let us take a close look at the typical and anomalous distributions of households.
+According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration.
+Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
+Compared with ID 2396, the other three households show different usage patterns.
+ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh.
+As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household.
+The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
+It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous}
 
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde} which are respectively selected when distortion correction is and is not applied.
-Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE.
-As for the households selected when distortion correction is applied, the middle row in \autoref{fig:electricityhdrsfixed} shows a household with small spikes at noon but only on Tuesday and Sunday.
-The anomalies detected without distortion correction show an increasing electricity demand across the day of the week.
-These findings show the difference between anomalies detected with different kernel density estimation methods.
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityoutliers}.
+The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
+This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
+As for the bottom two households, the middle row in \autoref{fig:electricityhdrsfixed} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
+These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
 
@@
 Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
-Without the ground truth density, we compare both density estimates by looking at the distributions of the most typical households with the highest densities and the anomalous households with the top two lowest densities.
-Both methods could identify the typical households with certain usage patterns, while the outliers are anomalous in different ways.
+Without the ground truth density, we assume that UMAP gives the true densities and compare both density estimation methods by looking at the distributions of the most typical households with the highest densities and three anomalous households that are identified most times.
+The distortion correction methods could identify the typical households with repeated time-of-the-week usage patterns, while the distributions of the outliers are anomalous in different ways.
 
 There are several open questions to be explored.
-The first involves the selection of bandwidth parameters for kernel density estimation, which has been explored in many KDE-related literatures.
-The second one is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitive relationship between embedding quality and density estimation accuracy is not immediate clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell wherther the distortion are corrected in the right way.
-The density estimates on the edges of the whole data structure could alos be explored because most outer area points tend to be detected as outliers.
+The first involves the selection of bandwidth parameters for kernel density estimation, which has been widely explored in many KDE-related literature.
+The second one is that the proposed density estimation method is only limited to the embedding points.
+Although for the downstream task of anomaly detection, only density estimated on these points are needed, it would be worth exploring the out-of-sample density estimation in the future.
+The last question is related to the quality of the embedding from dimension reduction methods.
+Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediate clear.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion are corrected in the right way.
+The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
 However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.

### 2022-11-06 R/Electricity2d_04_plots.R
Update PDF, table electricityoutliers hash for color codes to be solved

@@ -192,6 +192,7 @@ f20cols <- f20cols %>%
   mutate_all(funs(str_replace(., "1", blues9[5]))) %>% 
   mutate_all(funs(str_replace(., "2", blues9[9]))) %>% 
   as.matrix()
+f20cols[f20cols == "0"] <- "\#0000ffff"
 tabletop20 <- f20 %>% 
   kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding. The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE. The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, col.names = NULL, align=rep('c', 10)) %>%
   kable_styling(latex_options = "scale_down") %>%

### 2022-11-06 paper/paper.Rmd
Fix highlighting in Table 4&5

--- 2022-11-06 paper/paper.Rmd
@@
 ```
-```{r electricityoutliers}
-load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
-tabletop20
+```{r electricityoutliers, eval=T}
+load("figures/Electricity_2d_table_outlier_idcompare_5ml_radius20_r180.rda")
+# tabletop20 <- 
+  f20 %>% 
+  kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding.
+The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
+The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
+                  col.names = c(" ", rep(c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"), 2))
+                  ) %>%
+  kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = TRUE) %>%
+  # add_header_above(c(" " = 1, rep(c("ISOMAP" = 1, "LLE" = 1, "Laplacian Eigenmaps" = 1, "t-SNE" = 1, "UMAP" = 1), 2))) %>% 
+  add_header_above(c(" " = 1, "DC-KDE" = 5, "KDE" = 5)) %>%
+  # row_spec(1, bold = TRUE) %>% 
+  column_spec(1, bold = FALSE) %>% 
+  column_spec(2, bold = freqmax[,2], color = f20cols[,2]) %>% 
+  column_spec(3, bold = freqmax[,3], color = f20cols[,3]) %>% 
+  column_spec(4, bold = freqmax[,4], color = f20cols[,4]) %>% 
+  column_spec(5, bold = freqmax[,5], color = f20cols[,5]) %>% 
+  column_spec(6, bold = freqmax[,6], color = f20cols[,6]) %>% 
+  column_spec(7, bold = freqmax[,7], color = f20cols[,7], border_left = TRUE) %>% 
+  column_spec(8, bold = freqmax[,8], color = f20cols[,8]) %>%
+  column_spec(9, bold = freqmax[,9], color = f20cols[,9]) %>% 
+  column_spec(10, bold = freqmax[,10], color = f20cols[,10]) %>% 
+  column_spec(11, bold = freqmax[,11], color = f20cols[,11])
+# tabletop20
 ```
@@
 load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
-table_outlierpercent
+# hdrmax <- hdrmax[1:2,]
+# table_outlierpercent <- 
+hdrtable %>% 
+  head(n = 2) %>% 
+  `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
+  kableExtra::kbl(caption = "Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
+                  col.names = c(rep(c("DC-KDE", "KDE"), 4))
+  ) %>%
+  # kable_classic("striped") %>%
+  # kable_styling(latex_options = "scale_down") %>%
+  kable_paper(full_width = FALSE) %>%
+  # add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
+  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "t-SNE" = 2)) %>% 
+  # column_spec(1, bold = TRUE) %>% 
+  # add_indent(1:6) %>% 
+  column_spec(2, bold = hdrmax[,1]) %>% 
+  column_spec(3, bold = hdrmax[,2]) %>% 
+  column_spec(4, bold = hdrmax[,3]) %>% 
+  column_spec(5, bold = hdrmax[,4]) %>% 
+  column_spec(6, bold = hdrmax[,5]) %>% 
+  column_spec(7, bold = hdrmax[,6]) %>% 
+  column_spec(8, bold = hdrmax[,7]) %>% 
+  column_spec(2*(1:4), border_left = T, background = grey(0.9))
+# table_outlierpercent
 ```
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityoutliers}.
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityraw}.
 The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
 This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsfixed} shows a distribution with zero median and a $75\%$ quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
+As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a $75\%$ quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
 These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.

### 2022-11-06 paper/paper.tex
Fix highlighting in Table 4&5

@@
 \centering
+\resizebox{\linewidth}{!}{
 \begin{tabu} to \linewidth {>{}c>{}c>{}c>{}c>{}c>{}c|>{}c>{}c>{}c>{}c>{}c}
@@
 \cmidrule(l{3pt}r{3pt}){2-6} \cmidrule(l{3pt}r{3pt}){7-11}
-\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ISOMAP} & \multicolumn{1}{c}{LLE} & \multicolumn{1}{c}{Laplacian Eigenmaps} & \multicolumn{1}{c}{t-SNE} & \multicolumn{1}{c}{UMAP} & \multicolumn{1}{c}{ISOMAP} & \multicolumn{1}{c}{LLE} & \multicolumn{1}{c}{Laplacian Eigenmaps} & \multicolumn{1}{c}{t-SNE} & \multicolumn{1}{c}{UMAP} \\
-\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8} \cmidrule(l{3pt}r{3pt}){9-9} \cmidrule(l{3pt}r{3pt}){10-10} \cmidrule(l{3pt}r{3pt}){11-11}
-1 & \textcolor{black}{6317} & \textcolor{black}{7049} & \textcolor{black}{3161} & \textcolor{black}{4826} & \textcolor{black}{4546} & \textcolor{black}{6317} & \textcolor{black}{1881} & \textcolor{black}{6317} & \textcolor{black}{4190} & \textcolor{black}{6533}\\
-2 & \textcolor{0}{1353} & \textcolor[HTML]{08306B}{\textbf{4546}} & \textcolor[HTML]{08306B}{\textbf{5269}} & \textcolor{0}{1069} & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{4884} & \textcolor{0}{6199} & \textcolor{0}{6643} & \textcolor{0}{6817} & \textcolor{0}{5899}\\
-3 & \textcolor{0}{4884} & \textcolor[HTML]{08306B}{\textbf{1881}} & \textcolor{0}{4875} & \textcolor{0}{4546} & \textcolor[HTML]{08306B}{\textbf{1069}} & \textcolor{0}{5092} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor{0}{6764} & \textcolor{0}{6492} & \textcolor{0}{6345}\\
-4 & \textcolor{0}{4546} & \textcolor{0}{6952} & \textcolor{0}{7300} & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{4862} & \textcolor{0}{1353} & \textcolor[HTML]{6BAED6}{\textbf{4233}} & \textcolor{0}{1058} & \textcolor{0}{2309} & \textcolor{0}{1884}\\
-5 & \textcolor[HTML]{08306B}{\textbf{7049}} & \textcolor{0}{6838} & \textcolor{0}{4880} & \textcolor[HTML]{08306B}{\textbf{1533}} & \textcolor{0}{3639} & \textcolor{0}{5580} & \textcolor{0}{7049} & \textcolor{0}{5105} & \textcolor{0}{2546} & \textcolor{0}{3278}\\
+  & ISOMAP & LLE & Laplacian Eigenmaps & t-SNE & UMAP & ISOMAP & LLE & Laplacian Eigenmaps & t-SNE & UMAP\\
+\midrule
+1 & \textcolor[HTML]{000000}{6317} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{4826} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{6317} & \textcolor[HTML]{000000}{1881} & \textcolor[HTML]{000000}{6317} & \textcolor[HTML]{000000}{4190} & \textcolor[HTML]{000000}{6533}\\
+2 & \textcolor[HTML]{000000}{1353} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{5269} & \textcolor[HTML]{000000}{1069} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{4884} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{6643} & \textcolor[HTML]{000000}{6817} & \textcolor[HTML]{000000}{5899}\\
+3 & \textcolor[HTML]{000000}{4884} & \textcolor[HTML]{000000}{1881} & \textcolor[HTML]{000000}{4875} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{1069} & \textcolor[HTML]{000000}{5092} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{000000}{6764} & \textcolor[HTML]{000000}{6492} & \textcolor[HTML]{000000}{6345}\\
+4 & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{6952} & \textcolor[HTML]{000000}{7300} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{4862} & \textcolor[HTML]{000000}{1353} & \textcolor[HTML]{000000}{4233} & \textcolor[HTML]{000000}{1058} & \textcolor[HTML]{000000}{2309} & \textcolor[HTML]{000000}{1884}\\
+5 & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{6838} & \textcolor[HTML]{000000}{4880} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{3639} & \textcolor[HTML]{000000}{5580} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{5105} & \textcolor[HTML]{000000}{2546} & \textcolor[HTML]{000000}{3278}\\
 \addlinespace
-6 & \textcolor[HTML]{08306B}{\textbf{1366}} & \textcolor{0}{3874} & \textcolor{0}{5599} & \textcolor[HTML]{6BAED6}{\textbf{4265}} & \textcolor{0}{1964} & \textcolor{0}{5594} & \textcolor[HTML]{08306B}{\textbf{1821}} & \textcolor{0}{6577} & \textcolor{0}{2907} & \textcolor{0}{5463}\\
-7 & \textcolor{0}{3161} & \textcolor{0}{1069} & \textcolor{0}{3222} & \textcolor{0}{4862} & \textcolor[HTML]{6BAED6}{\textbf{2249}} & \textcolor{0}{2576} & \textcolor{0}{4546} & \textcolor{0}{3880} & \textcolor{0}{4536} & \textcolor{0}{5504}\\
-8 & \textcolor[HTML]{08306B}{\textbf{1074}} & \textcolor{0}{2249} & \textcolor{0}{1776} & \textcolor{0}{3639} & \textcolor{0}{6199} & \textcolor{0}{1366} & \textcolor[HTML]{08306B}{\textbf{3431}} & \textcolor{0}{5358} & \textcolor{0}{6898} & \textcolor{0}{1782}\\
-9 & \textcolor{0}{6681} & \textcolor{0}{1964} & \textcolor{0}{1161} & \textcolor{0}{5546} & \textcolor[HTML]{6BAED6}{\textbf{3892}} & \textcolor{0}{6681} & \textcolor{0}{1569} & \textcolor{0}{1366} & \textcolor{0}{4128} & \textcolor{0}{5965}\\
-10 & \textcolor{0}{1701} & \textcolor[HTML]{6BAED6}{\textbf{1569}} & \textcolor{0}{3655} & \textcolor{0}{4896} & \textcolor{0}{6209} & \textcolor{0}{5291} & \textcolor{0}{1964} & \textcolor{0}{5545} & \textcolor{0}{7368} & \textcolor{0}{6081}\\
+6 & \textcolor[HTML]{000000}{1366} & \textcolor[HTML]{000000}{3874} & \textcolor[HTML]{000000}{5599} & \textcolor[HTML]{000000}{4265} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{5594} & \textcolor[HTML]{000000}{1821} & \textcolor[HTML]{000000}{6577} & \textcolor[HTML]{000000}{2907} & \textcolor[HTML]{000000}{5463}\\
+7 & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{1069} & \textcolor[HTML]{000000}{3222} & \textcolor[HTML]{000000}{4862} & \textcolor[HTML]{000000}{2249} & \textcolor[HTML]{000000}{2576} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{3880} & \textcolor[HTML]{000000}{4536} & \textcolor[HTML]{000000}{5504}\\
+8 & \textcolor[HTML]{000000}{1074} & \textcolor[HTML]{000000}{2249} & \textcolor[HTML]{000000}{1776} & \textcolor[HTML]{000000}{3639} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{1366} & \textcolor[HTML]{000000}{3431} & \textcolor[HTML]{000000}{5358} & \textcolor[HTML]{000000}{6898} & \textcolor[HTML]{000000}{1782}\\
+9 & \textcolor[HTML]{000000}{6681} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{1161} & \textcolor[HTML]{000000}{5546} & \textcolor[HTML]{000000}{3892} & \textcolor[HTML]{000000}{6681} & \textcolor[HTML]{000000}{1569} & \textcolor[HTML]{000000}{1366} & \textcolor[HTML]{000000}{4128} & \textcolor[HTML]{000000}{5965}\\
+10 & \textcolor[HTML]{000000}{1701} & \textcolor[HTML]{000000}{1569} & \textcolor[HTML]{000000}{3655} & \textcolor[HTML]{000000}{4896} & \textcolor[HTML]{000000}{6209} & \textcolor[HTML]{000000}{5291} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{5545} & \textcolor[HTML]{000000}{7368} & \textcolor[HTML]{000000}{6081}\\
 \addlinespace
-11 & \textcolor{0}{3136} & \textcolor{0}{1533} & \textcolor{0}{6169} & \textcolor{0}{4555} & \textcolor{0}{6952} & \textcolor{0}{3161} & \textcolor[HTML]{6BAED6}{\textbf{6568}} & \textcolor{0}{2160} & \textcolor{0}{2232} & \textcolor{0}{4894}\\
-12 & \textcolor{0}{1881} & \textcolor[HTML]{6BAED6}{\textbf{4555}} & \textcolor{0}{4830} & \textcolor{0}{6199} & \textcolor{0}{4896} & \textcolor[HTML]{08306B}{\textbf{1701}} & \textcolor{0}{7300} & \textcolor{0}{6585} & \textcolor{0}{6273} & \textcolor{0}{5937}\\
-13 & \textcolor{0}{6785} & \textcolor{0}{6526} & \textcolor{0}{2284} & \textcolor[HTML]{6BAED6}{\textbf{6952}} & \textcolor{0}{6838} & \textcolor{0}{4749} & \textcolor{0}{7438} & \textcolor{0}{5824} & \textcolor{0}{1557} & \textcolor{0}{5522}\\
-14 & \textcolor{0}{5092} & \textcolor{0}{3639} & \textcolor{0}{6526} & \textcolor{0}{6838} & \textcolor{0}{1533} & \textcolor{0}{6699} & \textcolor{0}{4401} & \textcolor{0}{4932} & \textcolor{0}{3500} & \textcolor{0}{5580}\\
-15 & \textcolor{0}{5269} & \textcolor{0}{6199} & \textcolor{0}{1694} & \textcolor{0}{1964} & \textcolor[HTML]{6BAED6}{\textbf{1569}} & \textcolor{0}{5504} & \textcolor{0}{3434} & \textcolor{0}{6410} & \textcolor{0}{1932} & \textcolor{0}{6902}\\
+11 & \textcolor[HTML]{000000}{3136} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{6169} & \textcolor[HTML]{000000}{4555} & \textcolor[HTML]{000000}{6952} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{6568} & \textcolor[HTML]{000000}{2160} & \textcolor[HTML]{000000}{2232} & \textcolor[HTML]{000000}{4894}\\
+12 & \textcolor[HTML]{000000}{1881} & \textcolor[HTML]{000000}{4555} & \textcolor[HTML]{000000}{4830} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{4896} & \textcolor[HTML]{000000}{1701} & \textcolor[HTML]{000000}{7300} & \textcolor[HTML]{000000}{6585} & \textcolor[HTML]{000000}{6273} & \textcolor[HTML]{000000}{5937}\\
+13 & \textcolor[HTML]{000000}{6785} & \textcolor[HTML]{000000}{6526} & \textcolor[HTML]{000000}{2284} & \textcolor[HTML]{000000}{6952} & \textcolor[HTML]{000000}{6838} & \textcolor[HTML]{000000}{4749} & \textcolor[HTML]{000000}{7438} & \textcolor[HTML]{000000}{5824} & \textcolor[HTML]{000000}{1557} & \textcolor[HTML]{000000}{5522}\\
+14 & \textcolor[HTML]{000000}{5092} & \textcolor[HTML]{000000}{3639} & \textcolor[HTML]{000000}{6526} & \textcolor[HTML]{000000}{6838} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{6699} & \textcolor[HTML]{000000}{4401} & \textcolor[HTML]{000000}{4932} & \textcolor[HTML]{000000}{3500} & \textcolor[HTML]{000000}{5580}\\
+15 & \textcolor[HTML]{000000}{5269} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{1694} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{1569} & \textcolor[HTML]{000000}{5504} & \textcolor[HTML]{000000}{3434} & \textcolor[HTML]{000000}{6410} & \textcolor[HTML]{000000}{1932} & \textcolor[HTML]{000000}{6902}\\
 \addlinespace
-16 & \textcolor{0}{2576} & \textcolor[HTML]{6BAED6}{\textbf{5531}} & \textcolor{0}{3431} & \textcolor[HTML]{6BAED6}{\textbf{7104}} & \textcolor{0}{3213} & \textcolor{0}{3170} & \textcolor{0}{2249} & \textcolor{0}{3023} & \textcolor{0}{3119} & \textcolor{0}{7243}\\
-17 & \textcolor{0}{1262} & \textcolor{0}{4875} & \textcolor{0}{2523} & \textcolor{0}{3892} & \textcolor{0}{4401} & \textcolor{0}{5899} & \textcolor{0}{3213} & \textcolor{0}{1497} & \textcolor{0}{3809} & \textcolor{0}{6699}\\
-18 & \textcolor{0}{5966} & \textcolor{0}{5356} & \textcolor{0}{5356} & \textcolor{0}{3161} & \textcolor{0}{6766} & \textcolor{0}{2658} & \textcolor{0}{3161} & \textcolor{0}{7293} & \textcolor{0}{6108} & \textcolor{0}{5549}\\
-19 & \textcolor{0}{5937} & \textcolor[HTML]{6BAED6}{\textbf{3892}} & \textcolor[HTML]{6BAED6}{\textbf{2233}} & \textcolor[HTML]{08306B}{\textbf{6799}} & \textcolor{0}{4265} & \textcolor{0}{5189} & \textcolor[HTML]{08306B}{\textbf{1533}} & \textcolor{0}{5580} & \textcolor{0}{4516} & \textcolor{0}{5048}\\
-20 & \textcolor{0}{5356} & \textcolor{0}{4896} & \textcolor{0}{5125} & \textcolor{0}{6209} & \textcolor{0}{6799} & \textcolor{0}{2669} & \textcolor[HTML]{6BAED6}{\textbf{4555}} & \textcolor{0}{6224} & \textcolor{0}{5937} & \textcolor{0}{1923}\\
+16 & \textcolor[HTML]{000000}{2576} & \textcolor[HTML]{000000}{5531} & \textcolor[HTML]{000000}{3431} & \textcolor[HTML]{000000}{7104} & \textcolor[HTML]{000000}{3213} & \textcolor[HTML]{000000}{3170} & \textcolor[HTML]{000000}{2249} & \textcolor[HTML]{000000}{3023} & \textcolor[HTML]{000000}{3119} & \textcolor[HTML]{000000}{7243}\\
+17 & \textcolor[HTML]{000000}{1262} & \textcolor[HTML]{000000}{4875} & \textcolor[HTML]{000000}{2523} & \textcolor[HTML]{000000}{3892} & \textcolor[HTML]{000000}{4401} & \textcolor[HTML]{000000}{5899} & \textcolor[HTML]{000000}{3213} & \textcolor[HTML]{000000}{1497} & \textcolor[HTML]{000000}{3809} & \textcolor[HTML]{000000}{6699}\\
+18 & \textcolor[HTML]{000000}{5966} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{6766} & \textcolor[HTML]{000000}{2658} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{7293} & \textcolor[HTML]{000000}{6108} & \textcolor[HTML]{000000}{5549}\\
+19 & \textcolor[HTML]{000000}{5937} & \textcolor[HTML]{000000}{3892} & \textcolor[HTML]{000000}{2233} & \textcolor[HTML]{000000}{6799} & \textcolor[HTML]{000000}{4265} & \textcolor[HTML]{000000}{5189} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{5580} & \textcolor[HTML]{000000}{4516} & \textcolor[HTML]{000000}{5048}\\
+20 & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{000000}{4896} & \textcolor[HTML]{000000}{5125} & \textcolor[HTML]{000000}{6209} & \textcolor[HTML]{000000}{6799} & \textcolor[HTML]{000000}{2669} & \textcolor[HTML]{000000}{4555} & \textcolor[HTML]{000000}{6224} & \textcolor[HTML]{000000}{5937} & \textcolor[HTML]{000000}{1923}\\
 \bottomrule
-\end{tabu}
+\end{tabu}}
 \end{table}
@@
 \centering
-\resizebox{\linewidth}{!}{
-\begin{tabu} to \linewidth {>{\raggedright}X|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}c>{\centering}X}
+\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
 \toprule
@@
 \cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
-\multicolumn{1}{c}{HDRs} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} & \multicolumn{1}{c}{DC-KDE} & \multicolumn{1}{c}{KDE} \\
-\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8} \cmidrule(l{3pt}r{3pt}){9-9}
->99\% HDR & \cellcolor[HTML]{E6E6E6}{0.189} & 0.243 & \cellcolor[HTML]{E6E6E6}{0.189} & 0.027 & \cellcolor[HTML]{E6E6E6}{0.622} & 0.000 & \cellcolor[HTML]{E6E6E6}{0.865} & 0.108\\
-99\% HDR & \cellcolor[HTML]{E6E6E6}{0.359} & \textbf{0.248} & \cellcolor[HTML]{E6E6E6}{\textbf{0.503}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.697}} & 0.007 & \cellcolor[HTML]{E6E6E6}{\textbf{0.628}} & 0.083\\
+  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
+\midrule
+>99\% HDR & \cellcolor[HTML]{E6E6E6}{0.189} & \textbf{0.243} & \cellcolor[HTML]{E6E6E6}{\textbf{0.189}} & 0.027 & \cellcolor[HTML]{E6E6E6}{\textbf{0.622}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.865}} & 0.108\\
+99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.359}} & 0.248 & \cellcolor[HTML]{E6E6E6}{\textbf{0.503}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.697}} & 0.007 & \cellcolor[HTML]{E6E6E6}{\textbf{0.628}} & 0.083\\
 \bottomrule
-\end{tabu}}
+\end{tabular}
 \end{table}
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityoutliers}.
+Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityraw}.
 The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
 This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsfixed} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
+As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
 These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.

### 2022-11-06 R/Electricity2d_04_plots.R
Fix highlighting in Table 4&5

@@ -186,13 +186,9 @@ freqmax <- apply(f20, 2, function(x) x %in% f20rep$.) # highlight all ids that a
 freqmax1 <- apply(f20, 2, function(x) x %in% f20rep$.[f20rep$Freq == 5])
 freqmax2 <- apply(f20, 2, function(x) x %in% f20rep$.[f20rep$Freq == 4])
 f20cols <- freqmax1*2 + freqmax2
-f20cols <- f20cols %>% 
-  as_tibble() %>%
-  mutate_all(as.character) %>% 
-  mutate_all(funs(str_replace(., "1", blues9[5]))) %>% 
-  mutate_all(funs(str_replace(., "2", blues9[9]))) %>% 
-  as.matrix()
-f20cols[f20cols == "0"] <- "\#0000ffff"
+f20cols[f20cols == 0] <- "#000000"
+f20cols[f20cols == 1] <- blues9[5]
+f20cols[f20cols == 2] <- blues9[7]
 tabletop20 <- f20 %>% 
   kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding. The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE. The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, col.names = NULL, align=rep('c', 10)) %>%
   kable_styling(latex_options = "scale_down") %>%
@@ -203,12 +199,12 @@ tabletop20 <- f20 %>%
   column_spec(1, bold = FALSE) %>% 
   column_spec(2, bold = freqmax[,2], color = f20cols[,2]) %>% 
   column_spec(3, bold = freqmax[,3], color = f20cols[,3]) %>% 
-  column_spec(4, bold = freqmax[,4], color = f20cols[,4]) %>% 
+  column_spec(4, bold = freqmax[,4], color = f20cols[,4], width = "1cm") %>% 
   column_spec(5, bold = freqmax[,5], color = f20cols[,5]) %>% 
   column_spec(6, bold = freqmax[,6], color = f20cols[,6]) %>% 
   column_spec(7, bold = freqmax[,7], color = f20cols[,7], border_left = TRUE) %>% 
   column_spec(8, bold = freqmax[,8], color = f20cols[,8]) %>% # background = grey(0.9)
[…]

### 2022-11-07 paper/paper.tex
final polishing of 1 and 2

@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\hypertarget{intro}{%
+\section{Introduction}\label{intro}}
 
@@
 
-\[Pr(\pmb{y}\in\psi(\mathcal{A}))=\int_{\psi(\mathcal{A})} f(\psi^{-1}(\pmb{y}))|g(\pmb{y})|^{-1/2}d\pmb{y}\]
-where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(g(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
-It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|g(\pmb{y})|^{-1/2}\) which can be considered roughly analogous to the Jacobian determinant.
-Since \(\psi\) will typically be a non-linear function, the value of \(|g(\pmb{y})|^{-1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
+\begin{equation}
+\label{eq:changevar}
+Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y}
+\end{equation}
+
+where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
+It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|\pmb{H}(\pmb{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant.
+Since \(\psi\) will typically be a non-linear function, the value of \(|\pmb{H}(\pmb{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
 Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
@@
 For this reason we refer to it as the distortion corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|g(\pmb{y})|^{-1/2}\), for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\), for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
 One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density.
@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, when the data lie on some known manifold.
+Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold.
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
@@
 We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\).
-Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make these assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
+Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
 \begin{align}
@@
 For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein.
-
 Kernel density estimators of this form can and have been applied directly on the output embedding \(\pmb{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
-As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-To make this clear consider the simpler case computing the density after a change of variables \(\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d\), which involves a Jacobian term.
-A similar notion applies to a manifold embedding so that the density of the output vectors \(\pmb{y}_i\) differs from the density on the manifold itself.
-Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own.
-This motivates a kernel desnity estimate that corrects for the distortion induced by \(\psi\).
 
@@
 In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.
+The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.
 
@@
 Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
-The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system.
+The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
+In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\pmb{H}(\pmb{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\pmb{p}_i\).
 For more on the volume density function, see \textcite{Le_Brigant2019-lj}.
@@
 
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric in the coordinate system.
-Formally, the Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(\langle,\rangle_g\) on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
-The inner product between two tangent vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g\), can be used to define geometric quantities.
-For example, angles on a manifold are given by \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\), while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
-While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
-This is precisely the situation when data on a manifold are mapped to \(d\)-dimensional Euclidean vectors \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) via a dimension reduction algorithm.
-After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
+Formally, the Riemannian metric is a symmetric and positive definite tensor field which defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
+Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
 To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
 These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
-For example, the angle between \(\pmb{p}_j\) and \(\pmb{p}_k\) at \(\pmb{p}_i\) depends (up to a first order approximation) on the inner product \((\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)\) rather than the usual Euclidean inner product \((\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)\).
 

### 2022-11-07 paper/paper.Rmd
final polishing of 1 and 2

@@
 
-# Introduction
+# Introduction{#intro}
 
@@
 
-$$Pr(\pmb{y}\in\psi(\mathcal{A}))=\int_{\psi(\mathcal{A})} f(\psi^{-1}(\pmb{y}))|g(\pmb{y})|^{-1/2}d\pmb{y}$$
-where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $g(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
-It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|g(\pmb{y})|^{-1/2}$ which can be considered roughly analogous to the Jacobian determinant.
-Since $\psi$ will typically be a non-linear function, the value of $|g(\pmb{y})|^{-1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
+\begin{equation}
+\label{eq:changevar}
+Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y}
+\end{equation}
+
+where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
+It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant.
+Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
 Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
@@
 For this reason we refer to it as the distortion corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of $|g(\pmb{y})|^{-1/2}$, for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$, for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
 One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density.
@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, when the data lie on some known manifold.
+Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, which can be applied when the data lie on some known manifold.
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
@@
 We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$.
-Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make these assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
+Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
 \begin{align}
@@
 For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein.
-
 Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.
-As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
+
+<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
 To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term.
@@
 Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own.
-This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.
+This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.-->
 
@@
 In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and the volume density function ensures that the density estimate integrates to one.
+The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.
 
@@
 Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
-The volume density function is the square root of the determinant of the Riemannian metric expressed in this coordinate system.
+The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
+In proving that the estimator integrates to 1, the volume density function cancels with the term $|\pmb{H}(\pmb{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\pmb{p}_i$.
 For more on the volume density function, see @Le_Brigant2019-lj.
@@
 
-To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric in the coordinate system.
-Formally, the Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
-The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
+To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
+Formally, the Riemannian metric is a symmetric and positive definite tensor field which defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
+<!--The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
 For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
 While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
-This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.
-After this mapping, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
 To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point.
 These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
-For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
+<!--For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
+-->
 

### 2022-11-07 paper/paper.Rmd
section 4 tidied up

@@
 
-Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households.
-Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
-The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}.
-\autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) in $>99\%$ and $99\%$ HDRs that are also identified in different dimension reduction methods.
+Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households.
+To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
+The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}.
+\autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) where anomalies fall outside the $>99\%$ and $99\%$ HDRs respectively that are also identified as anomalies by different dimension reduction methods.
 Both DC-KDE and KDE are shown in the table.
-Except the lower percentage of ISOMAP in $>99\%$ HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than $0.5$.
+Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
+The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost comletely different anomalies to UMAP.
 \autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
@@
 Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.
+The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}.
+The household indices shown in darker shades of blue are identified as anomalies the most dimension reduction algorithms.
+It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.
 
@@
 
-Now let us take a close look at the typical and anomalous distributions of households.
-According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration.
+We now further examine households 3161, 4546, 7049, which appear as anomalies for at least three dimension reduction algorithms as well as  Households 2396, which is the most typical household with the highest density estimate.
 Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
@@
 The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.
+It is likely that these two households are not occupied all the time during the trial and could be used as temporary accomodation (e.g.
+through a holiday rental service such as AirBnB).
 
@@
 The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
-We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension redution algorithm.
+We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension reduction algorithm.
 

### 2022-11-07 paper/paper.tex
section 4 tidied up

@@
 
-Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households.
-Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
-The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}.
-\autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) in \(>99\%\) and \(99\%\) HDRs that are also identified in different dimension reduction methods.
+Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households.
+To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
+The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}.
+\autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) where anomalies fall outside the \(>99\%\) and \(99\%\) HDRs respectively that are also identified as anomalies by different dimension reduction methods.
 Both DC-KDE and KDE are shown in the table.
-Except the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than \(0.5\).
+Apart from the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
+The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost comletely different anomalies to UMAP.
 \autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
@@
 
-Now let us take a close look at the typical and anomalous distributions of households.
-According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration.
+We now further examine households 3161, 4546, 7049, which appear as anomalies for at least three dimension reduction algorithms as well as Households 2396, which is the most typical household with the highest density estimate.
 Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
@@
 The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.
+It is likely that these two households are not occupied all the time during the trial and could be used as temporary accomodation (e.g.~through a holiday rental service such as AirBnB).
 

### 2022-11-08 paper/paper.Rmd
Add refs and fix typos

@@
 Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Chen2017-dw and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
-Applications of KDE include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq].
-For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well known curse of dimensionality.
-However the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d<<m$.
-The problem of KDE on manifolds has been well studied see @Pelletier2005-vu (FIND SOME MORE REFS) for details.
-However, in these setting it is assumed that the underlying manifold in known, which is rarely true in practice.
+Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq].
+For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality.
+However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$.
+The problem of KDE on manifolds has been well studied in @ @Hendriks1990-bt, @Izenman1991-cj, @@Pelletier2005-vu, @Henry2013-da and @Berenfeld2022-fd.
+However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
 Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
@@
 When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$.
-Details of these algorithms are summarised (OTHER REF) in @Cheng2021-dh.
+These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$.
+Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx and @Cheng2021-dh.
 One could easily apply KDE to the output vectors $\pmb{y}$.
-However this would be a misleading representation of the density of points as they lie on the manifold.
+However, this would be a misleading representation of the density of points as they lie on the manifold.
 To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-
-$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M$$
-
+$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,$$
 where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$.
-By a change of variables, this would imply the following density for $\pmb{y}$
-
+By a change of variables, this would imply the following density for $\pmb{y}$,
 \begin{equation}
 \label{eq:changevar}
-Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y}
+Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
 \end{equation}
-
 where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
@@
 
-In this paper we propose a new KDE.
-Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm.
-For this reason we refer to it as the distortion corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$, for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density.
-In a number of simulation studies we demonstrate two benefits that arise when using our new estimator.
+In this paper, we propose a new KDE.
+Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm.
+For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
+Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
+In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
 The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
@@
 
-In terms of data visualisation, our approach allows a two or three dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high density region or if instead this is an artefact of the dimension reduction algorithm 'squeezing' points together.
+In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
 This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
-Furthermore, our estimator can be used for manifolds with elements that are are a variety of objects, for instance rotational frames or images.
+Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images.
 In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
-Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to an extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].
+Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].
 
@@
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel distortion corrected kernel density estimate in \autoref{DCKDE}.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
+By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
 
@@
 Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
-Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however this need not be the case.
+Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case.
 For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold.
@@
 The reasons for this will become clearer when we discuss the manifold setting in more detail.
-Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise.
+Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise.
 In our empirical section, more general kernel functions can be, and are, employed.
@@
 For example, for a uniform kernel, the estimator at point $\pmb{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
-However in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
+However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
 An additional technical assumption is that $r$ is less than the injectivity radius of the manifold.
@@
 For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\pmb{p}$ is so large that some points `fall inside' the ball more than once.
-For example on a sphere, a ball with radius greater than half the circumference of a great circle will wrap back around the sphere.
+For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
 This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
@@
 Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space,  $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$.
-Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and travelling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time.
+Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and traveling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time.
 Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
@@
 In proving that the estimator integrates to 1, the volume density function cancels with the term $|\pmb{H}(\pmb{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\pmb{p}_i$.
-For more on the volume density function, see @Le_Brigant2019-lj.
+For more on the volume density function, see @Gallot2004-rc and @Le_Brigant2019-lj.
 
@@
 To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field which defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
+Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
 <!--The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
@@
 While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
-This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
 To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point.
@@
 In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into $d$-dimesional Euclidean space.
-In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used.
+In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used (See Chapter 3 of @Lee2007-wq for more about intrinsic dimension estimation).
 The *R* library *intRinsic* [@Denti2021-qc] implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
@@
 The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length is Section 6.2 of @Perrault-Joncas2013-pq).
+To understand these differences, it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq).
 This is crucial for two reasons.
@@
 Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\pmb{y}_i$, and it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
-For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
+For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
 Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
@@
 To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with lowest density.
+We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density.
 The entire workflow is summarized in \autoref{fig:vkde}.
@@
 Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue.
-With this anomaly detection process, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
-
-(ref:vkdecaption) The proposed schematic for anomaly detection with distortion corrected kernel density estimates.
+With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
+
+(ref:vkdecaption) The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.
 
@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion corrected KDE.
+In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion-corrected KDE.
 For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
@@
 To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-In general, we aim to highlight two advantages of our proposed distortion corrected KDE compared to KDE applied directly to the output coordinates.
+In general, we aim to highlight two advantages of our proposed distortion-corrected KDE compared to KDE applied directly to the output coordinates.
 First, the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
@@
 The mixture proportions
-are equally set as $\pi_i=0.25, i=1,2,3,4$. The two dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the following 
+are equally set as $\pi_i=0.25, i=1,2,3,4$. The two-dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the following 
 \begin{equation}
@@
 This notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as $f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is a the density of a four component mixture of normals.
+The true density on the manifold can be obtained as $f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is the density of a four-component mixture of normals.
 Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold.
@@
 Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in yellow.
-The anomalies are found around the edges of the plot, but there are is also a low density region between the means of the four mixture components.
+The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
 The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\pmb{v}$.
@@
 \autoref{fig:tpoutliers} summarizes the results.
-Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
+Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively.
 We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$.
-Colors show the different estimated density at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
-For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high density regions.
-On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold in towards the centre.
-The distortion corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
-For t-SNE, the ground truth and distortion corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
+Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
+For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
+On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.
+The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
+For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
 This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
@@
   kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Distortion corrected KDE outperfoms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>%
@@
 ```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
-The colors indicate different level of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
-The shading contains all anomalies that are both truely and correctly identified.
+The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
+The shading contains all anomalies that are both truly and correctly identified.
 KDE without distortion correction gives more misclassified anomalies.", echo = FALSE}
@@
 
-In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
+In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction.
 Data are presented on a log scale to highlight anomalies.
 The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a $99\%$ highest density region.
-The middle shaded region contain anomalies that are true positives in the sense of not lying in a $90\%$ HDR.
+The middle shaded region contains anomalies that are true positives in the sense of not lying in a $90\%$ HDR.
 Points lying outside squares (shown as triangles) are incorrectly classified.
@@
 ```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
-Distortion corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
+Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
@@
 
-Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of dimension reduction algorithm.
+Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of a dimension reduction algorithm.
 Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
-The left column shows results when distortion correction is applied, the right column when it is not applied.
+The left column shows results when distortion correction is applied, and the right column when it is not applied.
 It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
@@
 $v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
-\autoref{fig:fivedmeta} shows scatterplot which is a single frame from a 5-D tour path[^1] animation using the R package *tourr* [@Wickham2011-ir].
+\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path[^1] animation using the R package *tourr* [@Wickham2011-ir].
 The round and triangular point shapes indicate the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the centre of the distribution.
+$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the center of the distribution.
 It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
@@
 ```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE.
-Four manifold learning methods are used rowwise.
-The point shapes indicates whether they are the true outleirs, and the grey shading highlights the top 1\\% rank region.
+Four manifold learning methods are used row-wise.
+The point shapes indicate whether they are the true outliers, and the grey shading highlights the top 1\\% rank region.
 The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
@@
 The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example.
-In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densites from DC-KDE and KDE are presented.
+In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densities from DC-KDE and KDE are presented.
 For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP there more distortion is induced throughdimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
-In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively,when distortion correction is applied.
+However, for Laplacian Eigenmaps and UMAP there more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
+In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
 
 Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
-As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors.
-The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
-The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
+As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.
+The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion-corrected KDE and panels on the right showing KDE without distortion correction.
+The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
 Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
 
-Comparing the left and right panel for each row, we notice that KDE with distortion correction, has fewer misclassified observations, and therefore outperforms KDE.
+Comparing the left and right panels for each row, we notice that KDE with distortion correction has fewer misclassified observations, and therefore outperforms KDE.
 For UMAP, almost all ground truth anomalies (points outside a $99\%$ HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
-For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than $10^{-7}$) which leads to tied ranks.
+For DC-KDE in the left panel there are several triangles in a horizontal line, this arises, due to many points having an estimated density of close to zero (values less than $10^{-7}$) which leads to tied ranks.
 These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
@@
 Compared with KDE, the proportions are always higher when distortion correction methods are used.
-The proportions with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
+The proportions with distortion correction are very close for ISOMAP, LLE, and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
 This could be due to the severe distortion usually induced by the UMAP algorithm.
@@
 
-For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
+For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week [@Hyndman2018-ia].
 This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
 @Cheng2021-dh propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a $2$-dimensional embedding for kernel density estimation and anomaly detection.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP for dimension reduction to obtain a $2$-dimensional embedding for kernel density estimation and anomaly detection.
 To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
 We use the highest density regions plot to visualize the density estimates.
-However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
+However, for this real data set with an unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
 
 Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the process here.
-For each household, a discrete approximation to the distribution of electricity demand at each one of the $336$ half-hourly periods of the week is found.
-For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
-The number of nearest neighbors in manifold learning $K$ is set as $100$ and the bandwidth paramter $r=180$ to match the distances between the distributions of households.
+For each household, a discrete approximation of the distribution of electricity demand at each one of the $336$ half-hourly periods of the week is found.
+For any pair of households the total variation metric can be found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
+The number of nearest neighbors in manifold learning $K$ is set as $100$ and the bandwidth parameter $r=180$ to match the distances between the distributions of households.
 
@@
 The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
-The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
+The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
                   col.names = c(" ", rep(c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"), 2))
@@
 
-\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
+\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
 There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions.
@@
 First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
-When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE and UMAP with DC-KDE compared with the KDE.
-As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identifies as anomalies.
+When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE.
+As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies.
 The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
-Again, KDE with distortion correction methods find more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
-Without distortion correction, the density estimates are different not only in scaling, but also shows inconsistency in the rank of density estimates.
+Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
+Without distortion correction, the density estimates are different not only in scaling, but also show inconsistency in the rank of density estimates.
 
@@
 Both DC-KDE and KDE are shown in the table.
-Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost comletely different anomalies to UMAP.
+Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods.
+The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP.
 \autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
@@
 The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}.
-The household indices shown in darker shades of blue are identified as anomalies the most dimension reduction algorithms.
-It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.
+The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
+It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE hand UMAP do not identify any of the same anomalies as ISOMAP and LLE.
 
@@
 
-We now further examine households 3161, 4546, 7049, which appear as anomalies for at least three dimension reduction algorithms as well as  Households 2396, which is the most typical household with the highest density estimate.
+We now further examine households 3161, 4546, and 7049, which appear as anomalies for at least three dimension reduction algorithms as well as  Households 2396, which is the most typical household with the highest density estimate.
 Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
@@
 The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as temporary accomodation (e.g.
-through a holiday rental service such as AirBnB).
-
-```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of week for one typical household 1472 and two anomalies, 7049 and 5136.
-The day-of-the-week patterns is much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time related usage patterens.", echo = FALSE, eval=T}
+It is likely that these two households are not occupied all the time during the trial and could be used as temporary accommodation (e.g.
+through a holiday rental service such as Airbnb).
+
+```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of the week for one typical household 1472 and two anomalies, 7049 and 5136.
+The day-of-the-week patterns are much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time-related usage patterns.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous.png")
@@
 The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a $75\%$ quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
+This repeated pattern in a week window is also shown on some days of the week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
+As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a $75\%$ quantile of 1kWh from around 9 pm to 6 am, while the distributions for ID 7049 do not have an obvious pattern across different periods of the week.
 These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
@@
 In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities.
-Compared with KDE, our distortion corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
+Compared with KDE, our distortion-corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
 The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
@@
 There are several open questions to be explored.
-The first involves the selection of bandwidth parameters for kernel density estimation, which has been widely explored in many KDE-related literature.
+The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
+For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition).
 The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points are needed, it would be worth exploring the out-of-sample density estimation in the future.
+Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation in the future.
 The last question is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediate clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion are corrected in the right way.
+Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
 The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
-However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
 
@@
 This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
-The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at the Monash University.
+The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at Monash University.
 
@@
 
-The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold.
+The *geodesics* of $M$ are the smooth curves that locally join the points along the shortest path on the manifold.
 Intuitively, geodesics are the *straightest possible curves* in a Riemannian manifold [Section 7.2.3 of @Nakahara2018-zs].
@@
 
-Denote $B(p, r) \subset M$ as an open ball centered at point $p$ with raidus $r$.
+Denote $B(p, r) \subset M$ as an open ball centered at point $p$ with radius $r$.
 Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$.
@@
 This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
-By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into highest density regions more accurately than KDE.
-By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest density regions, which are usually used to detect anomalies.
+By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest-density regions more accurately than KDE.
+By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest-density regions, which are usually used to detect anomalies.
 
@@
 
-```{r tpleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+```{r tpleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of Laplacian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
@@
 ```{r tpllevs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}

### 2022-11-08 paper/paper.tex
Add refs and fix typos

@@
 Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
-Applications of KDE include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
-For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well known curse of dimensionality.
-However the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d<<m\).
-The problem of KDE on manifolds has been well studied see \textcite{Pelletier2005-vu} (FIND SOME MORE REFS) for details.
-However, in these setting it is assumed that the underlying manifold in known, which is rarely true in practice.
+Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
+For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality.
+However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\).
+The problem of KDE on manifolds has been well studied in @ \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, @\textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}.
+However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
 Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
@@
 When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\).
-Details of these algorithms are summarised (OTHER REF) in \textcite{Cheng2021-dh}.
+These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\).
+Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx} and \textcite{Cheng2021-dh}.
 One could easily apply KDE to the output vectors \(\pmb{y}\).
-However this would be a misleading representation of the density of points as they lie on the manifold.
+However, this would be a misleading representation of the density of points as they lie on the manifold.
 To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-
-\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M\]
-
+\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,\]
 where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\).
-By a change of variables, this would imply the following density for \(\pmb{y}\)
-
+By a change of variables, this would imply the following density for \(\pmb{y}\),
 \begin{equation}
 \label{eq:changevar}
-Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y}
+Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
 \end{equation}
-
 where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
@@
 
-In this paper we propose a new KDE.
-Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm.
-For this reason we refer to it as the distortion corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\), for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density.
-In a number of simulation studies we demonstrate two benefits that arise when using our new estimator.
+In this paper, we propose a new KDE.
+Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm.
+For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
+Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
+In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
 The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
@@
 
-In terms of data visualisation, our approach allows a two or three dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead this is an artefact of the dimension reduction algorithm `squeezing' points together.
+In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
 This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
-Furthermore, our estimator can be used for manifolds with elements that are are a variety of objects, for instance rotational frames or images.
+Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images.
 In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
-Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to an extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.
+Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.
 
@@
 In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion corrected kernel density estimate in \autoref{DCKDE}.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
+By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
+To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
 
@@
 Points on this manifold are denoted \(\pmb{p}\) in general, with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
-Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however this need not be the case.
+Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case.
 For instance, \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
@@
 The reasons for this will become clearer when we discuss the manifold setting in more detail.
-Also, for illustration purposes, in this section we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise.
+Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise.
 In our empirical section, more general kernel functions can be, and are, employed.
@@
 For example, for a uniform kernel, the estimator at point \(\pmb{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
-However in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
+However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
 An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold.
@@
 For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once.
-For example on a sphere, a ball with radius greater than half the circumference of a great circle will wrap back around the sphere.
+For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
 This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
@@
 Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space, \(\pmb{v}\in T_{\pmb{p}}M\), to points on the manifold, \(\pmb{q}\in M\).
-Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and travelling along this geodesic at uniform speed \(\|\pmb{v}\|\) takes place in one unit of time.
+Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and traveling along this geodesic at uniform speed \(\|\pmb{v}\|\) takes place in one unit of time.
 Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
@@
 In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\pmb{H}(\pmb{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\pmb{p}_i\).
-For more on the volume density function, see \textcite{Le_Brigant2019-lj}.
+For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.
 
@@
 To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field which defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
-Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
+Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
 To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
@@
 In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into \(d\)-dimesional Euclidean space.
-In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used.
+In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used (See Chapter 3 of \textcite{Lee2007-wq} for more about intrinsic dimension estimation).
 The \emph{R} library \emph{intRinsic} \autocite{Denti2021-qc} implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
@@
 The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length is Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
+To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
 This is crucial for two reasons.
@@
 Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{y}_i\), and it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
-For this reason there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
+For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
 Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
@@
 Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
-With this anomaly detection process, outliers based on lowest densities could be detected more accurately regardless of the distortion in manifold learning.
+With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
 
@@
 
-\caption{The proposed schematic for anomaly detection with distortion corrected kernel density estimates.}\label{fig:vkde}
+\caption{The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.}\label{fig:vkde}
 \end{figure}
@@
 The mixture proportions
-are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the following
+are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two-dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the following
 \begin{equation}
@@
 This notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as \(f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is a the density of a four component mixture of normals.
+The true density on the manifold can be obtained as \(f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is the density of a four-component mixture of normals.
 Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' for densities on the manifold.
@@
 Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in yellow.
-The anomalies are found around the edges of the plot, but there are is also a low density region between the means of the four mixture components.
+The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
 The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the \(\pmb{v}\).
@@
 \autoref{fig:tpoutliers} summarizes the results.
-Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion corrected KDE and KDE respectively.
+Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively.
 We set the bandwidth parameter in the DC-KDE \eqref{eq:denestimator} as \(r=0.5\).
-Colors show the different estimated density at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
-For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high density regions.
-On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold in towards the centre.
-The distortion corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
-For t-SNE, the ground truth and distortion corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
+Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
+For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
+On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.
+The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
+For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
 This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
@@
 \caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Distortion corrected KDE outperfoms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.}
+Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.}
 \centering
@@
 \caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
-The colors indicate different level of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
-The shading contains all anomalies that are both truely and correctly identified.
+The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
+The shading contains all anomalies that are both truly and correctly identified.
 KDE without distortion correction gives more misclassified anomalies.}\label{fig:tpisomapden}
@@
 
-In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction.
+In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction.
 Data are presented on a log scale to highlight anomalies.
 The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a \(99\%\) highest density region.
-The middle shaded region contain anomalies that are true positives in the sense of not lying in a \(90\%\) HDR.
+The middle shaded region contains anomalies that are true positives in the sense of not lying in a \(90\%\) HDR.
 Points lying outside squares (shown as triangles) are incorrectly classified.
@@
 \caption{Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
-Distortion corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.}\label{fig:tpisomapvs4ml}
-\end{figure}
-
-Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of dimension reduction algorithm.
+Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.}\label{fig:tpisomapvs4ml}
+\end{figure}
+
+Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of a dimension reduction algorithm.
 Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
-The left column shows results when distortion correction is applied, the right column when it is not applied.
+The left column shows results when distortion correction is applied, and the right column when it is not applied.
 It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
@@
 \(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
-\autoref{fig:fivedmeta} shows scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
+\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
 The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
-\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the centre of the distribution.
+\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the center of the distribution.
 It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
@@
 \caption{Rank comparison between the true density and estimated density from both DC-KDE and KDE.
-Four manifold learning methods are used rowwise.
-The point shapes indicates whether they are the true outleirs, and the grey shading highlights the top 1\% rank region.
+Four manifold learning methods are used row-wise.
+The point shapes indicate whether they are the true outliers, and the grey shading highlights the top 1\% rank region.
 The colors show the distance to the center of the semisphere, with darker points being distant from the center.}\label{fig:fivedisomapden}
@@
 The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example.
-In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densites from DC-KDE and KDE are presented.
+In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densities from DC-KDE and KDE are presented.
 For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP there more distortion is induced throughdimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
-In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively,when distortion correction is applied.
+However, for Laplacian Eigenmaps and UMAP there more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
+In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
 
 Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
-As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors.
-The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction.
-The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP.
+As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.
+The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion-corrected KDE and panels on the right showing KDE without distortion correction.
+The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
 Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
 
-Comparing the left and right panel for each row, we notice that KDE with distortion correction, has fewer misclassified observations, and therefore outperforms KDE.
+Comparing the left and right panels for each row, we notice that KDE with distortion correction has fewer misclassified observations, and therefore outperforms KDE.
 For UMAP, almost all ground truth anomalies (points outside a \(99\%\) HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
-For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than \(10^{-7}\)) which leads to tied ranks.
+For DC-KDE in the left panel there are several triangles in a horizontal line, this arises, due to many points having an estimated density of close to zero (values less than \(10^{-7}\)) which leads to tied ranks.
 These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
@@
 Compared with KDE, the proportions are always higher when distortion correction methods are used.
-The proportions with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
+The proportions with distortion correction are very close for ISOMAP, LLE, and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
 This could be due to the severe distortion usually induced by the UMAP algorithm.
@@
 
-For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
+For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week \autocite{Hyndman2018-ia}.
 This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
 \textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection.
+Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection.
 To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
 We use the highest density regions plot to visualize the density estimates.
-However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
+However, for this real data set with an unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
 
 Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here.
-For each household, a discrete approximation to the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
-For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
-The number of nearest neighbors in manifold learning \(K\) is set as \(100\) and the bandwidth paramter \(r=180\) to match the distances between the distributions of households.
+For each household, a discrete approximation of the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
+For any pair of households the total variation metric can be found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
+The number of nearest neighbors in manifold learning \(K\) is set as \(100\) and the bandwidth parameter \(r=180\) to match the distances between the distributions of households.
 
@@
 The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
-The blue colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
+The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
 \centering
@@
 
-\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
+\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
 There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions.
@@
 Both DC-KDE and KDE are shown in the table.
-Apart from the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost comletely different anomalies to UMAP.
+Apart from the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods.
+The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP.
 \autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
@@
 
-We now further examine households 3161, 4546, 7049, which appear as anomalies for at least three dimension reduction algorithms as well as Households 2396, which is the most typical household with the highest density estimate.
+We now further examine households 3161, 4546, and 7049, which appear as anomalies for at least three dimension reduction algorithms as well as Households 2396, which is the most typical household with the highest density estimate.
 Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
@@
 The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as temporary accomodation (e.g.~through a holiday rental service such as AirBnB).
+It is likely that these two households are not occupied all the time during the trial and could be used as temporary accommodation (e.g.~through a holiday rental service such as Airbnb).
 
@@
 
-\caption{Quantile region plots of electricity demand against the time of week for one typical household 1472 and two anomalies, 7049 and 5136.
-The day-of-the-week patterns is much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time related usage patterens.}\label{fig:electricityhdrsvkde}
+\caption{Quantile region plots of electricity demand against the time of the week for one typical household 1472 and two anomalies, 7049 and 5136.
+The day-of-the-week patterns are much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time-related usage patterns.}\label{fig:electricityhdrsvkde}
 \end{figure}
@@
 The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also shown in some days of week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9pm to 6am, while the distributions for ID 7049 does not have an obvious pattern across different periods of the week.
+This repeated pattern in a week window is also shown on some days of the week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
+As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9 pm to 6 am, while the distributions for ID 7049 do not have an obvious pattern across different periods of the week.
 These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
@@
 There are several open questions to be explored.
-The first involves the selection of bandwidth parameters for kernel density estimation, which has been widely explored in many KDE-related literature.
+The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
+For densities on unknown manifolds, \textcite{Berenfeld2021-as} also provide some insights on the assumptions of the kernel function in \eqref{eq:kernelcondition}.
 The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points are needed, it would be worth exploring the out-of-sample density estimation in the future.
+Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation in the future.
 The last question is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediate clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion are corrected in the right way.
+Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
 The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
-However, the outperformance of DC-KDE than KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
 
@@
 
-The \emph{geodesics} of \(M\) are the smooth curves that locally joins the points along the shortest path on the manifold.
+The \emph{geodesics} of \(M\) are the smooth curves that locally join the points along the shortest path on the manifold.
 Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold \autocite[Section 7.2.3 of][]{Nakahara2018-zs}.
@@
 
-Denote \(B(p, r) \subset M\) as an open ball centered at point \(p\) with raidus \(r\).
+Denote \(B(p, r) \subset M\) as an open ball centered at point \(p\) with radius \(r\).
 Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\), where \(exp_p\) is the \emph{exponential map} at point \(p\).
@@
 
-\caption{Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+\caption{Scatterplot of true density and estimated density ranks of Laplacian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpleden}
@@
 \caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-1}
@@
 \caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-2}
@@
 \caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-3}
@@
 \caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represents the four gaussian kernels in the 2-D meta data.
+The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-4}

### 2022-11-08 paper/paper.tex
Density on manifold boundary ref

@@
 The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation in the future.
+Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds \autocite{Berry2017-rz} in the future.
 The last question is related to the quality of the embedding from dimension reduction methods.

### 2022-11-08 paper/paper.Rmd
Density on manifold boundary ref

@@
 The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation in the future.
+Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds [@Berry2017-rz] in the future.
 The last question is related to the quality of the embedding from dimension reduction methods.

### 2022-11-12 paper/paper.tex
Minor

@@
 However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\).
-The problem of KDE on manifolds has been well studied in @ \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, @\textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}.
+The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}.
 However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.

### 2023-06-01 monash/dckde.Rmd
Rename paper/ for monash template

--- 2022-11-08 paper/paper.Rmd
@@
 
-# Introduction{#intro}
+# Introduction {#dckdeintro}
 
@@
 However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$.
-The problem of KDE on manifolds has been well studied in @ @Hendriks1990-bt, @Izenman1991-cj, @@Pelletier2005-vu, @Henry2013-da and @Berenfeld2022-fd.
+The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da and @Berenfeld2022-fd.
 However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
@@
 \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional 'twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
-\autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
-Conclusions and discussions are presented in \autoref{conclusion}.
+\autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
+Conclusions and discussions are presented in \autoref{dckdeconclusion}.
 Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
@@
 $v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path[^1] animation using the R package *tourr* [@Wickham2011-ir].
+\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path^[See the animation of the 5-D grand tour at https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif.] animation using the R package *tourr* [@Wickham2011-ir].
 The round and triangular point shapes indicate the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
@@
 It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
-
-[^1]: See the animation of the 5-D grand tour at https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif.
 
@@
 
-# Application {#application}
+# Application {#dckdeapplication}
 
@@
 In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010.
-The CER dataset[^2] records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
+The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
 We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.
-
-[^2]: accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.
 
@@
 
-# Conclusions {#conclusion}
+# Conclusions {#dckdeconclusion}
 

### 2023-06-01 R/Electricity2d_04_plots.R

@@ -162,6 +162,23 @@ phdr5ml
 ggsave(paste0("paper/figures/Electricity_2d_5levels_outliers_comparison_5ml_r", format(r, decimal.mark = "_"), ".png"), phdr5ml, width = 9, height = 12, dpi = 300)
 
 
+# SLides
+phdriso <- (
+  ((
+    (p_isomap$p + labs(x = "", y = "ISOMAP", title = "DC-KDE")) |
+      (p_hdr_isomap$p + labs(x = "", y = "", title = "KDE") + noytext ) 
+  ) & scale_y_continuous(breaks = c(-250, 0, 250)))
+
+) + 
+  plot_layout(guides = 'collect') &
+  # guides(HDRs = guide_legend(nrow = 1)) &
+  theme(legend.direction = "horizontal", legend.position = "bottom", 
+        legend.box = "horizontal", axis.title.y = element_text(face = "bold"),
+        plot.title = element_text(hjust = 0.5, face = "bold"))
+phdriso
+ggsave(paste0("paper/figures/Electricity_2d_5levels_outliers_comparison_isomap_r", format(r, decimal.mark = "_"), ".png"), phdriso, width = 10, height = 6, dpi = 300)
+
+
 ###-------------------------------------------------
 ## Table for top 20 outlier IDs
 ###-------------------------------------------------

### 2023-06-01 R/Simulation2d_TwinPeaks.R

@@ -566,16 +566,16 @@ p
 
 
 # ISOMAP only for slides
-((p_den_isomap + labs(x = "", y = "ISOMAP", title = "True density") ) |
-    (p_isomap$p + labs(x = "", y = "", title = "DC-KDE") + noytext) |
-    (p_hdr_isomap$p + labs(x = "", y = "", title = "KDE") + noytext)) +
+((p_den_le + labs(x = "", y = "Laplacian Eigenmaps", title = "True density") ) |
+    (p_le$p + labs(x = "", y = "", title = "DC-KDE") + noytext) |
+    (p_hdr_le$p + labs(x = "", y = "", title = "KDE") + noytext)) +
   plot_layout(guides = 'collect') &
   scale_x_continuous(n.breaks = 4) &
   scale_y_continuous(n.breaks = 4) &
   theme(legend.direction = "horizontal", legend.position = "bottom", 
         legend.box = "horizontal",
         plot.title = element_text(hjust = 0.5, face = "bold"))
-# ggsave(paste0("paper/figures/", mapping, N,"_5levels_outliers_comparison_isomap_3cases_r", format(r, decimal.mark = "_"), ".png"), width = 12, height = 5, dpi = 300)
+# ggsave(paste0("paper/figures/", mapping, N,"_5levels_outliers_comparison_le_3cases_r", format(r, decimal.mark = "_"), ".png"), width = 12, height = 5, dpi = 300)
 
 
 

### 2023-06-01 monash/dckde.tex
Rename paper/ for monash template

--- 2022-11-12 paper/paper.tex
@@
 
-\hypertarget{intro}{%
-\section{Introduction}\label{intro}}
+\hypertarget{dckdeintro}{%
+\section{Introduction}\label{dckdeintro}}
 
@@
 \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
-\autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
-Conclusions and discussions are presented in \autoref{conclusion}.
+\autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
+Conclusions and discussions are presented in \autoref{dckdeconclusion}.
 Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
@@
 
-\hypertarget{application}{%
-\section{Application}\label{application}}
+\hypertarget{dckdeapplication}{%
+\section{Application}\label{dckdeapplication}}
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusions}\label{conclusion}}
+\hypertarget{dckdeconclusion}{%
+\section{Conclusions}\label{dckdeconclusion}}
 

### 2023-06-01 jcgs/B-appB-kderm.Rmd
Add jcgs version for kderm, the latest version

--- 2022-11-08 paper/paper.Rmd
@@
----
-title: "Distortion corrected kernel density estimate on Riemannian manifolds"
-author:
-- familyname: Cheng
-  othernames: Fan
-  address: Monash University
-  email: Fan.Cheng@monash.edu
-  # correspondingauthor: true
-- familyname: Panagiotelis
-  othernames: Anastasios
-  address: The University of Sydney
-  email: Anastasios.Panagiotelis@sydney.edu.au
-- familyname: Hyndman
-  othernames: Rob J
-  address: Monash University
-  email: Rob.Hyndman@monash.edu
-abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements.
-The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
-We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
-A shiny app in R is also developed for various simulation scenarios.
-The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
-keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
-# wpnumber: 03/2021
-# jelcodes: C55, C65, C80
-blind: false
-cover: false
-toc: false
-lot: false
-bibliography: references.bib
-biblio-style: authoryear-comp
-output:
-  monash::workingpaper:
-    fig_caption: yes
-    fig_height: 5
-    fig_width: 8
-    includes:
-      in_header: preamble.tex
-      # after_body: appendix.tex
-    keep_tex: yes
-    number_sections: yes
-    citation_package: biblatex
-    latex_engine: pdflatex
-editor_options: 
-  markdown: 
-    wrap: 72
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  cache = T,
-  messages = FALSE,
-  warning = FALSE,
-  eval = TRUE,
-  include = TRUE
-)
-options(tinytex.verbose = TRUE)
-library(tidyverse)
-library(here)
-library(patchwork)
-library(kableExtra)
-# Jmisc::sourceAll(here::here("R/sources"))
-# set.seed(1234)
-```
-
-\newpage
-
-# Introduction{#intro}
-
-Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Chen2017-dw and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
-Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq].
-For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality.
-However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$.
-The problem of KDE on manifolds has been well studied in @ @Hendriks1990-bt, @Izenman1991-cj, @@Pelletier2005-vu, @Henry2013-da and @Berenfeld2022-fd.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
-
-When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$.
-Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx and @Cheng2021-dh.
-One could easily apply KDE to the output vectors $\pmb{y}$.
-However, this would be a misleading representation of the density of points as they lie on the manifold.
-To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,$$
-where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$.
-By a change of variables, this would imply the following density for $\pmb{y}$,
-\begin{equation}
-\label{eq:changevar}
-Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
-\end{equation}
-where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
-It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant.
-Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
-Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
-
-In this paper, we propose a new KDE.
-Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm.
-For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
-In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
-The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
-The second is that when the estimates are found using different dimension reduction algorithms (e.g.
-ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.
-
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
-This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
-Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images.
-In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
-Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].
-
-The rest of the paper is organized as follows.
-In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-Our estimator combines insights from two existing papers, the first is @Pelletier2005-vu who propose a KDE for data on manifolds that are known, the second is @Perrault-Joncas2013-pq who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
-\autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional 'twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
-\autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
-Conclusions and discussions are presented in \autoref{conclusion}.
-Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
-
-<!--
-
-OLD:
-
-Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of $\mathbb{R}^d$.
-Applications of kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac], prime prediction via Twitter messages [@Gerber2014-tq], and so on.
-However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than $\mathbb{R}^d$, kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted.
-@Pelletier2005-vu propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known.
-Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold.
-Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than $\mathbb{R}^d$.
-Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function.
-This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.
-
-Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
-It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-dh], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
-To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
-
-For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
-There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
-The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
-In this paper, we focus on the multivariate kernel density estimation.
-
-Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points.
-However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-The bandwidth is varied depending on either the location of the sample points [sample smoothing estimator; @Terrell1992-ut] or that of the estimated points [balloon estimator; @Terrell1992-ut].
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the *variable/adaptive kernel density estimation* [VKDE; Section 6.6 of @Scott2015-vl] unless otherwise stated.
-However, these kernel density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature [@Henry2009-ll].
-However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space.
-Therefore, the kernel density estimator from @Pelletier2005-vu is not applicable because the geodesic distance and the volume density function are unknown.
-This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-dh), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
-
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
-One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold [@Goldberg2008-co].
-To tackle this problem and measure the distortion incurred in manifold learning,
-@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
-The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.
-
-By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding.
-Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
-Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information.
-Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
-These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
-Similar to @Cheng2021-dh, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
-
-The rest of the paper is organized as follows.
-In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds.
-Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection.
-\autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space.
-Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
-\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset.
-Conclusions and discussions are presented in \autoref{conclusion}.
-Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.
-
--->
-
-# Distortion Corrected Kernel density estimate on Riemannian manifolds {#kderm}
-
-In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of @Pelletier2005-vu, which can be applied when the data lie on some known manifold.
-In \autoref{MetLearn}, we describe the Learn Metric algorithm of @Perrault-Joncas2013-pq, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of @Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
-
-In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
-Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case.
-For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$.
-We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$.
-Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
-\begin{align}
-\label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
-(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
-(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
-& (iv) \supp K=[0; 1];
-(v) \sup K(\|\pmb{y}\|)=K(0).
-\end{align}
-Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
-For instance, condition (iv) requires the support of the kernel to be bounded.
-The reasons for this will become clearer when we discuss the manifold setting in more detail.
-Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise.
-In our empirical section, more general kernel functions can be, and are, employed.
-
-For data $\pmb{y}_i\in\mathbb{R}^d$ with $i=1,\dots,N$ and assuming a bandwidth matrix $r\pmb{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by
-\begin{equation}
-\label{eq:vkde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
-\end{equation}
-
-The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1.
-In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
-For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein.
-Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.
-
-<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term.
-A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density on the manifold itself.
-Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own.
-This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.-->
-
-## Kernel Density estimation on manifolds {#Pellet}
-
-For kernel density estimation on a known manifold, @Pelletier2005-vu propose the following estimator,
-\begin{equation}
-\label{eq:denriem}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
-\end{equation}
-
-where $d_g(\pmb{p}, \pmb{p}_i)$ denotes the geodesic distance between two points on the manifold $\pmb{p}$ and $\pmb{p}_i$ and $\theta_{\pmb{p}_i}(\pmb{p})$ is known as the volume density function.
-The intuition behind the term $K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)$ is relatively clear.
-For example, for a uniform kernel, the estimator at point $\pmb{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
-However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
-An additional technical assumption is that $r$ is less than the injectivity radius of the manifold.
-A definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix.
-For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\pmb{p}$ is so large that some points `fall inside' the ball more than once.
-For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
-This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
-
-The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
-We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
-However, the volume of this ball must also be taken into account.
-In Euclidean space with the usual Lebesgue measure, a radius $r$ ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.
-
-More formally, the volume density function can be explained as follows.
-Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space,  $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$.
-Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and traveling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time.
-Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
-The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-In proving that the estimator integrates to 1, the volume density function cancels with the term $|\pmb{H}(\pmb{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\pmb{p}_i$.
-For more on the volume density function, see @Gallot2004-rc and @Le_Brigant2019-lj.
-
-## Riemannian metric estimation {#MetLearn}
-
-To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
-<!--The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
-For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
-While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
-This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
-To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point.
-These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
-<!--For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
--->
-
-While full details are provided in @Perrault-Joncas2013-pq, we briefly describe the Learn Metric algorithm here.
-There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold.
-Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
-Third, a dimension reduction method is applied to obtain the output embedding $\pmb{y}_1,\dots,\pmb{y}_n$.
-Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
-Full details on these four steps are provided in \autoref{alg:learnmetric}.
-This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*.
-Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.
-
-As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension $d$.
-In this case, the ranks of the matrices $\pmb{H}_i$ are equal to $d$.
-Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension $d$ globally into $d$-dimensional Euclidean space.
-In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into $d$-dimesional Euclidean space.
-In practice, to determine the dimension of the manifold, the *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] can be used (See Chapter 3 of @Lee2007-wq for more about intrinsic dimension estimation).
-The *R* library *intRinsic* [@Denti2021-qc] implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
-
-\begin{algorithm}[!htb]
-  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
-  \label{alg:learnmetric}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters EMBED }
-  \BlankLine
-  \begin{algorithmic}[1]
-
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
-
-  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
-  $$
-  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
-  $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
-
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
-
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
-  $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
-  $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
-  $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
-  $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
-
-  \end{algorithmic}
-\end{algorithm}
-
-## Distortion corrected KDE {#DCKDE}
-
-With all fundamentals introduced, we can now give our novel Distortion Corrected KDE (DC-KDE) as 
-\begin{equation}
-\label{eq:denestimator}
-\hat{f}(\pmb{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|}{r} \bigg).
-\end{equation}
-
-The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq).
-This is crucial for two reasons.
-First, this implies that the term $\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|$ approximates the geodesic distance between $\pmb{y}_i$ and $\pmb{y}_j$.
-Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\pmb{y}_i$, and it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
-For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
-Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
-
-One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\pmb{H}_j$.
-To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density.
-The entire workflow is summarized in \autoref{fig:vkde}.
-The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue.
-With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
-
-(ref:vkdecaption) The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.
-
-```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
-knitr::include_graphics("figures/DC-KDE.png")
-```
-
-# Simulations {#simulation}
-
-In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion-corrected KDE.
-For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 4-dimensional manifold embedded in a 100-dimensional ambient space.
-To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-In general, we aim to highlight two advantages of our proposed distortion-corrected KDE compared to KDE applied directly to the output coordinates.
-First, the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
-Second, we show how density estimation and anomaly detection are more robust to a different choice of dimension reduction method when distortion correction is used.
-
-## Twin peaks example {#twodgaussian}
-
-The simulation setup for the twin peaks example is to first generate vectors $\pmb{v}_1,\dots,\pmb{v}_N$ for $N=2000$ from a 2-dimensional Gaussian mixture model.
-The mixture has four components with different means
-$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4$.
-The mixture proportions
-are equally set as $\pi_i=0.25, i=1,2,3,4$. The two-dimensional data in \autoref{fig:metadensity} is mapped to a 'twin peaks' surface via the following 
-\begin{equation}
-\label{eq:twinpeak}
-\begin{array}{lcl}
-x_1 = v_1, \\
-x_2 = v_2, \\
-x_3 = \sin(\pi v_1) \tan (3 v_2).
-\end{array}
-\end{equation}
-
-The three-dimensional twin peaks mapping is shown in \autoref{fig:twinpeaks3d}.
-The colors in both \autoref{fig:metadensity} and \autoref{fig:twinpeaks3d} indicate the true density of the data via the twin peaks mapping, with lower density points in darker colors scattered in the outer as well as center areas.
-<!-- We also considered the 'Swiss Roll' mapping and the results for this manifold are summarized in \autoref{swissrollappe}. -->
-
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
-The colors indicate the true density of the data when they are mapped via the twin peaks function.
-Lower density points in darker colors are scattered both in the outer and center areas.", echo = FALSE}
-knitr::include_graphics("figures/truedensity_twinpeaks_dc_labv.png")
-```
-
-(ref:twinpeaks3dcaption) Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.
-
-```{r twinpeaks3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:twinpeaks3dcaption)", echo = FALSE}
-knitr::include_graphics("figures/scatterplot3d_twinpeaks_dc.png")
-```
-
-It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in Equation \@ref(eq:twinpeak) distorts the distribution.
-To recover the true distribution requires the correct Jacobian term for the pushforward from $\pmb{v}$ to the volume form of twin peaks manifold.
-By treating the $\pmb{v}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\pmb{\Gamma}_i$ for $i=1,\dots,n$ where $\pmb{\Gamma}_i$ is the Riemannian of the coordinate system given by $\pmb{v}$.
-This notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as $f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is the density of a four-component mixture of normals.
-Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold.
-\autoref{fig:metadensity} shows the simulated $\pmb{v}$ with color indicating the true density of data on the manifold.
-Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in yellow.
-The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\pmb{v}$.
-
-```{r tpoutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
-The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel).
-DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
-```
-
-\autoref{fig:tpoutliers} summarizes the results.
-Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively.
-We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$.
-Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
-For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
-On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.
-The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
-For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
-This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
-
-```{r tpcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_Twin Peak2000_5ml_r0_5.rda")
-col.max <- apply(cors, 2, max)
-max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
-  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  # kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = FALSE) %>%
-  # column_spec(1, width = "6cm") %>% 
-  column_spec(2, bold = max.match[,1]) %>%
-  column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4]) %>%
-  column_spec(6, bold = max.match[,5])
-```
-
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from KDE with and without density correction by \autoref{tab:tpcors}.
-Distortion correction improves the rank correlation for all dimension reduction algorithms.
-In particular, while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
-
-```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
-The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
-The shading contains all anomalies that are both truly and correctly identified.
-KDE without distortion correction gives more misclassified anomalies.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
-```
-
-```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding.
-KDE without distortion correction gives many more misclassified anomalies.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
-```
-
-In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction.
-Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a $99\%$ highest density region.
-The middle shaded region contains anomalies that are true positives in the sense of not lying in a $90\%$ HDR.
-Points lying outside squares (shown as triangles) are incorrectly classified.
-For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the $99\%$ HDR, but are not classified as such (although they are true positives if a 90% HDR is used).
-Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
-\autoref{fig:tptsneden} shows the same plot but for t-SNE.
-The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified.
-However, the difference between KDE with and without distortion correction is stark.
-These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.
-
-```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
-Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
-# knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
-# knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
-# knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
-# knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
-```
-
-Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of a dimension reduction algorithm.
-Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
-The left column shows results when distortion correction is applied, and the right column when it is not applied.
-It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
-This is critical since conclusions will be more robust to the choice of dimension reduction algorithm.
-
-## Semi-hypersphere example embedded in 100-D space {#fivedgaussian}
-
-```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
-Distant points in darker colors could be seen as anomalies.", echo = FALSE}
-knitr::include_graphics("figures/tourr_5d_semisphere.png")
-```
-
-As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
-To start with, we simulate vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ for $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, with the same means $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$ and different variance-covariance matrices $\pmb{\Sigma}_1 = diag(1,1,1,1)$ and $\pmb{\Sigma}_2 = diag(2,2,2,2)$.
-The mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-With this design, the observations from the second component tend to be outlying anomalies.
-The data are mapped to a hemisphere via the equation
-$v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path[^1] animation using the R package *tourr* [@Wickham2011-ir].
-The round and triangular point shapes indicate the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the center of the distribution.
-It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
-
-[^1]: See the animation of the 5-D grand tour at https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif.
-
-To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to $\pmb{v}_i$ so that $\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N$.
-Next, we rotate the $100$-dimensional vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ by multiplying by a randomly generated rotation matrix.
-To generate the rotation matrix we first simulate elements from a uniform $(0,1)$ distribution, stack them into a $100\times 100$ matrix $\pmb{A}$ and then take the R matrix from the QR decomposition of $\pmb{A}$.
-Rotating the vectors results in input vectors that are no longer sparse.
-Nonetheless, the intrinsic dimension of this is still $d=4$.
-
-```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE.
-Four manifold learning methods are used row-wise.
-The point shapes indicate whether they are the true outliers, and the grey shading highlights the top 1\\% rank region.
-The colors show the distance to the center of the semisphere, with darker points being distant from the center.", echo = FALSE}
-knitr::include_graphics("figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec.png")
-```
-
-Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density.
-The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example.
-In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densities from DC-KDE and KDE are presented.
-For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP there more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
-In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
-
-Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
-As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.
-The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion-corrected KDE and panels on the right showing KDE without distortion correction.
-The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
-
-Comparing the left and right panels for each row, we notice that KDE with distortion correction has fewer misclassified observations, and therefore outperforms KDE.
-For UMAP, almost all ground truth anomalies (points outside a $99\%$ HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
-For DC-KDE in the left panel there are several triangles in a horizontal line, this arises, due to many points having an estimated density of close to zero (values less than $10^{-7}$) which leads to tied ranks.
-These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
-
-```{r fivedcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_4d_N10000_4ml_radius10_r1.rda")
-col.max <- apply(cors, 2, max)
-max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
-  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
-  kable_paper(full_width = FALSE) %>%
-  # column_spec(1, width = "6cm") %>% 
-  column_spec(2, bold = max.match[,1]) %>%
-  column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4])
-```
-
-```{r fourdhdrtable, echo=FALSE, message=FALSE, eval=T}
-load("figures/hdrtable_4d_N10000_4ml_radius10_r1.rda")
-hdrtable %>% 
-  head(n = 2) %>% 
-  `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
-  kableExtra::kbl(caption = "Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
-                  col.names = c(rep(c("DC-KDE", "KDE"), 4))
-                  ) %>%
-  # kable_classic("striped") %>% 
-  # kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = FALSE) %>%
-  # add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
-  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "UMAP" = 2)) %>% 
-  # column_spec(1, bold = TRUE) %>% 
-  column_spec(2, bold = hdrmax[,1]) %>% 
-  column_spec(3, bold = hdrmax[,2]) %>% 
-  column_spec(4, bold = hdrmax[,3]) %>% 
-  column_spec(5, bold = hdrmax[,4]) %>% 
-  column_spec(6, bold = hdrmax[,5]) %>% 
-  column_spec(7, bold = hdrmax[,6]) %>% 
-  column_spec(8, bold = hdrmax[,7]) %>% 
-  column_spec(2*(1:4), border_left = T, background = grey(0.9))
-```
-
-\autoref{tab:fourdhdrtable} further illustrates the proportions of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions.
-Compared with KDE, the proportions are always higher when distortion correction methods are used.
-The proportions with distortion correction are very close for ISOMAP, LLE, and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$.
-This could be due to the severe distortion usually induced by the UMAP algorithm.
-
-# Application {#application}
-
-<!-- ## Irish smart meter dataset -->
-
-In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010.
-The CER dataset[^2] records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
-We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.
-
-[^2]: accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.
-
-For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week [@Hyndman2018-ia].
-This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
-@Cheng2021-dh propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP for dimension reduction to obtain a $2$-dimensional embedding for kernel density estimation and anomaly detection.
-To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
-We use the highest density regions plot to visualize the density estimates.
-However, for this real data set with an unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
-
-Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the process here.
-For each household, a discrete approximation of the distribution of electricity demand at each one of the $336$ half-hourly periods of the week is found.
-For any pair of households the total variation metric can be found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
-The number of nearest neighbors in manifold learning $K$ is set as $100$ and the bandwidth parameter $r=180$ to match the distances between the distributions of households.
-
-```{r electricityembedding, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
-knitr::include_graphics("figures/Electricity_2d_5levels_outliers_comparison_5ml_r180.png")
-```
-```{r electricityoutliers, eval=T}
-load("figures/Electricity_2d_table_outlier_idcompare_5ml_radius20_r180.rda")
-# tabletop20 <- 
-  f20 %>% 
-  kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding.
-The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
-The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
-                  col.names = c(" ", rep(c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"), 2))
-                  ) %>%
-  kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = TRUE) %>%
-  # add_header_above(c(" " = 1, rep(c("ISOMAP" = 1, "LLE" = 1, "Laplacian Eigenmaps" = 1, "t-SNE" = 1, "UMAP" = 1), 2))) %>% 
-  add_header_above(c(" " = 1, "DC-KDE" = 5, "KDE" = 5)) %>%
-  # row_spec(1, bold = TRUE) %>% 
-  column_spec(1, bold = FALSE) %>% 
-  column_spec(2, bold = freqmax[,2], color = f20cols[,2]) %>% 
-  column_spec(3, bold = freqmax[,3], color = f20cols[,3]) %>% 
-  column_spec(4, bold = freqmax[,4], color = f20cols[,4]) %>% 
-  column_spec(5, bold = freqmax[,5], color = f20cols[,5]) %>% 
-  column_spec(6, bold = freqmax[,6], color = f20cols[,6]) %>% 
-  column_spec(7, bold = freqmax[,7], color = f20cols[,7], border_left = TRUE) %>% 
-  column_spec(8, bold = freqmax[,8], color = f20cols[,8]) %>%
-  column_spec(9, bold = freqmax[,9], color = f20cols[,9]) %>% 
-  column_spec(10, bold = freqmax[,10], color = f20cols[,10]) %>% 
-  column_spec(11, bold = freqmax[,11], color = f20cols[,11])
-# tabletop20
-```
-
-\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
-There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions.
-The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
-First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
-When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE.
-As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies.
-The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
-Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
-Without distortion correction, the density estimates are different not only in scaling, but also show inconsistency in the rank of density estimates.
-
-```{r electricityhdrpercent}
-load("figures/Electricity_2d_hdrtable_umapvs4ml_radius20_r180.rda")
-# hdrmax <- hdrmax[1:2,]
-# table_outlierpercent <- 
-hdrtable %>% 
-  head(n = 2) %>% 
-  `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
-  kableExtra::kbl(caption = "Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
-                  col.names = c(rep(c("DC-KDE", "KDE"), 4))
-  ) %>%
-  # kable_classic("striped") %>%
-  # kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = FALSE) %>%
-  # add_header_above(c("HDRs" = 1, rep(c("DC-KDE" = 1, "KDE" = 1), 4))) %>%
-  add_header_above(c(" " = 1, "ISOMAP" = 2, "LLE" = 2, "Laplacian Eigenmaps" = 2, "t-SNE" = 2)) %>% 
-  # column_spec(1, bold = TRUE) %>% 
-  # add_indent(1:6) %>% 
-  column_spec(2, bold = hdrmax[,1]) %>% 
-  column_spec(3, bold = hdrmax[,2]) %>% 
-  column_spec(4, bold = hdrmax[,3]) %>% 
-  column_spec(5, bold = hdrmax[,4]) %>% 
-  column_spec(6, bold = hdrmax[,5]) %>% 
-  column_spec(7, bold = hdrmax[,6]) %>% 
-  column_spec(8, bold = hdrmax[,7]) %>% 
-  column_spec(2*(1:4), border_left = T, background = grey(0.9))
-# table_outlierpercent
-```
-```{r electricityrank, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
-knitr::include_graphics("figures/Electricity_2d_rankplot_3639id_umapvs4ml_radius20_r180.png")
-```
-
-Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households.
-To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
-The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}.
-\autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) where anomalies fall outside the $>99\%$ and $99\%$ HDRs respectively that are also identified as anomalies by different dimension reduction methods.
-Both DC-KDE and KDE are shown in the table.
-Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP.
-\autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
-The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}.
-Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.
-The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}.
-The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
-It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE hand UMAP do not identify any of the same anomalies as ISOMAP and LLE.
-
-```{r electricityraw, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.", echo = FALSE, eval=T}
-knitr::include_graphics("figures/Electricity_2d_compare4id_1typical_3anomalous.png")
-```
-
-We now further examine households 3161, 4546, and 7049, which appear as anomalies for at least three dimension reduction algorithms as well as  Households 2396, which is the most typical household with the highest density estimate.
-Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
-Compared with ID 2396, the other three households show different usage patterns.
-ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh.
-As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household.
-The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as temporary accommodation (e.g.
-through a holiday rental service such as Airbnb).
-
-```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of the week for one typical household 1472 and two anomalies, 7049 and 5136.
-The day-of-the-week patterns are much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time-related usage patterns.", echo = FALSE, eval=T}
-knitr::include_graphics("figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous.png")
-```
-
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityraw}.
-The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also shown on some days of the week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a $75\%$ quantile of 1kWh from around 9 pm to 6 am, while the distributions for ID 7049 do not have an obvious pattern across different periods of the week.
-These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
-
-# Conclusions {#conclusion}
-
-In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities.
-Compared with KDE, our distortion-corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
-The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
-We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension reduction algorithm.
-
-As an empirical example, we explore the distributions of $3,639$ households and $336$ time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
-Without the ground truth density, we assume that UMAP gives the true densities and compare both density estimation methods by looking at the distributions of the most typical households with the highest densities and three anomalous households that are identified most times.
-The distortion correction methods could identify the typical households with repeated time-of-the-week usage patterns, while the distributions of the outliers are anomalous in different ways.
-
-There are several open questions to be explored.
-The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
-For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition).
-The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation in the future.
-The last question is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
-The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
-However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
-
-<!-- kernel types discussion -->
-
-# Acknowledgment {-}
-
-This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
-The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at Monash University.
-
-<!-- \clearpage -->
-<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
-<!-- %\pagenumbering{arabic} -->
-<!-- %\def\thepage{A\arabic{page}} -->
-
 \appendix
 
-# Appendix: Notions about Riemannian geometry {#riemgeo}
+# Notions about Riemannian geometry {#riemgeo}
 
@@
 
-<!-- ## Isometry -->
-
 ## Exponential map and logarithmic map
@@
 
-<!-- # Appendix: Simulation of $S^1$ for manifold density estimation {#s1simulation} -->
-
-<!-- To start with, we demonstrate the use of the proposed DC-KDE on a $1$-dimensional example.
-For a $S^1$ curve centered at $O$ with radius $1$, the coordinated of points on the curve could be expressed in polar coordinates $(1, \alpha)$ or Cartesian coordinates $(x, \sqrt{1-x^2})$.
--->
-
-<!-- The first simulation starts with a $1$-dimensional example.  -->
-
-<!-- Manifold: $S^1$ with center 0 and radius $r=1$ -->
-
-<!-- Polar coordinates: $(r, \theta)$ -->
-
-<!-- Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$. -->
-
-<!-- Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$ -->
-
-<!-- Density of $\theta$: $f(\theta) = 2 / \pi$ -->
-
-<!-- Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$ -->
-
-<!-- The CDF of X is given by -->
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!-- P(X \leq x) =& P(\cos(\theta) \leq x) \\ -->
-<!-- =& P(\theta \geq \arccos(x)) \\ -->
-<!-- =& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\ -->
-<!-- =&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\ -->
-<!-- =& 1 - \frac{2}{\pi} \arccos(x). -->
-<!-- \end{aligned} -->
-<!-- $$ -->
-<!-- Then the density of X is the derivative of the CDF w.r.t. $x$  -->
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!-- f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\ -->
-<!-- =& \frac{2}{\pi \sqrt{1 - x^2}}. -->
-<!-- \end{aligned} -->
-<!-- $$ -->
-
-<!-- Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could -->
-<!-- estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ -->
-<!-- and compare it with its true density $f(x)$. -->
-<!-- We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by -->
-<!-- $$ -->
-<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}), -->
-<!-- $$ -->
-<!-- where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
--->
-
-<!-- Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
-Therefore, the geodesic distance is -->
-<!-- $$ -->
-<!-- d_g(p, X_i) = \arccos(p) - \arccos(X_i) -->
-<!-- $$ -->
-<!-- As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
-$|g(\exp_p^{-1}(x_i))|$.
--->
-
-<!-- [When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.] -->
-
-<!-- The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$, -->
-<!-- $$ -->
-<!-- g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R}, -->
-<!-- $$ -->
-<!-- is smooth in a local neighborhood $U$ of $M$. -->
-
-<!-- Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by -->
-<!-- $$ -->
-<!-- \theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}. -->
-<!-- $$ -->
-<!-- Therefore, the estimator for $p \in S^1$ is given by -->
-<!-- $$ -->
-<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg), -->
-<!-- $$ -->
-<!-- where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*. -->
-
-<!-- ### How to derive volume density function $\theta_{X_i}(p)$ -->
-
-<!-- #### Example of $S^2$ in @Henry2009-om. -->
-
-<!-- Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
-Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$.
-Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by -->
-<!-- $$ -->
-<!-- \Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw), -->
-<!-- $$ -->
-<!-- if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. -->
-<!-- Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$. -->
-<!-- Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that -->
-<!-- $$ -->
-<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, -->
-<!-- $$ -->
-<!-- $$ -->
-<!-- B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w. -->
-<!-- $$ -->
-<!-- The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors. -->
-<!-- Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
--->
-<!-- Therefore, we obtain that the volume density on the sphere is -->
-<!-- $$ -->
-<!-- \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2}  -->
-<!-- = \frac{|\sin(d_g(p,q))|}{d_g(p,g)}, -->
-<!-- $$ -->
-<!-- for $q \neq p, -p$, and $\theta_p(p) = 1$. -->
-
-<!-- #### $S^1$ -->
-
-<!-- Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization -->
-<!-- $$ -->
-<!-- \Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv, -->
-<!-- $$ -->
-<!-- and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have -->
-<!-- $$ -->
-<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v. -->
-<!-- $$ -->
-<!-- So the riemannian metric is $g_q = \cos(r)$ and further, -->
-<!-- $$ -->
-<!-- \theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|}, -->
-<!-- $$ -->
-<!-- where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$. -->
-
-# Appendix: Rank comparison plots for twin peaks mapping {#twinpeaksappe}
+# Rank comparison plots for twin peaks mapping {#twinpeaksappe}
 
 This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
-By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest-density regions more accurately than KDE.
-By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest-density regions, which are usually used to detect anomalies.
+By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest density regions more accurately than KDE.
+By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest density regions, which are usually used to detect anomalies.
 
@@
 
-```{r tpleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of Laplacian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+```{r tpleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
@@
 
-```{r tpllevs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+```{r tpllevs4ml, fig.align = 'center', out.width = "90%", fig.cap="Comparison of outliers found by LLE compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
 The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
 Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
-# knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
 knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
+```
+```{r tplevs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by Laplacian Eigemaps compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
+```
+```{r tptsnevs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by t-SNE compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
+```
+```{r tpumapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by UMAP compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).", echo = FALSE}
 knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
 ```
-
-<!-- \clearpage -->
-
-<!-- # Appendix: Simulation with swiss roll mapping {#swissrollappe}
-
-In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.
-
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \@ref(eq:swissroll).
-The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in \autoref{fig:swissroll3d}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
-\begin{equation}
-\label{eq:swissroll}
-\left\{
-\begin{array}{lcl}
-X = X_1 \cos{X_1}, \\
-Y = X_2, \\
-Z = X_1 \sin{X_1}.
-\end{array}
-\right.
-\end{equation}
-
-(ref:swissroll3dcaption) Scatterplot of the 3-d swiss roll data with colors indicating the true density via the swiss roll mapping in \autoref{eq:swissroll}.
-
-```{r swissroll3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:swissroll3dcaption)", echo = FALSE}
-knitr::include_graphics("figures/scatterplot3d_swissroll_dc.png")
-```
-
-```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
-knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
-```
-
-Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1%, 5%, 50%, 99%, \>99%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
-the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all
-manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
-
-```{r srcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_Swiss Roll_4ml_riem0_08.rda")
-col.max <- apply(cors, 2, max)
-max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
-  mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  # kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = FALSE) %>%
-  # column_spec(1, width = "6cm") %>% 
-  column_spec(2, bold = max.match[,1]) %>%
-  column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4]) %>%
-  column_spec(6, bold = max.match[,5])
-```
-
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
-rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
-correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
- -->
-<!-- Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot has indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along with the curved embedding.
--->

### 2023-06-01 jcgs/03-chap3-kderm.Rmd
Add jcgs version for kderm, the latest version

--- 2023-06-01 monash/dckde.Rmd
@@
----
-title: "Distortion corrected kernel density estimate on Riemannian manifolds"
-author:
-- familyname: Cheng
-  othernames: Fan
-  address: Monash University
-  email: Fan.Cheng@monash.edu
-  # correspondingauthor: true
-- familyname: Panagiotelis
-  othernames: Anastasios
-  address: The University of Sydney
-  email: Anastasios.Panagiotelis@sydney.edu.au
-- familyname: Hyndman
-  othernames: Rob J
-  address: Monash University
-  email: Rob.Hyndman@monash.edu
-abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements.
-The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
-We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
-A shiny app in R is also developed for various simulation scenarios.
-The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
-keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
-# wpnumber: 03/2021
-# jelcodes: C55, C65, C80
-blind: false
-cover: false
-toc: false
-lot: false
-bibliography: references.bib
-biblio-style: authoryear-comp
-output:
-  monash::workingpaper:
-    fig_caption: yes
-    fig_height: 5
-    fig_width: 8
-    includes:
-      in_header: preamble.tex
-      # after_body: appendix.tex
-    keep_tex: yes
-    number_sections: yes
-    citation_package: biblatex
-    latex_engine: pdflatex
-editor_options: 
-  markdown: 
-    wrap: 72
----
-
 ```{r setup, include=FALSE}
@@
 
-\newpage
-
 # Introduction {#dckdeintro}
 
-Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Chen2017-dw and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
+Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Scott2015-vl and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
 Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq].
@@
 However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$.
-The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da and @Berenfeld2022-fd.
+The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd.
 However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
@@
 
-When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
+When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
 These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$.
-Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx and @Cheng2021-dh.
+Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh.
 One could easily apply KDE to the output vectors $\pmb{y}$.
@@
 Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
+One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
 In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
@@
 In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
 This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
@@
 Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
-
-<!--
-
-OLD:
-
-Multivariate kernel density estimation has gained lots of attention in exploratory data analysis.
-It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of $\mathbb{R}^d$.
-Applications of kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac], prime prediction via Twitter messages [@Gerber2014-tq], and so on.
-However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than $\mathbb{R}^d$, kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted.
-@Pelletier2005-vu propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known.
-Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold.
-Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than $\mathbb{R}^d$.
-Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function.
-This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.
-
-Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.
-It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-dh], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
-In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies.
-To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.
-
-For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut].
-There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
-Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details].
-For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews].
-The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm.
-In this paper, we focus on the multivariate kernel density estimation.
-
-Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points.
-However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
-The bandwidth is varied depending on either the location of the sample points [sample smoothing estimator; @Terrell1992-ut] or that of the estimated points [balloon estimator; @Terrell1992-ut].
-In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the *variable/adaptive kernel density estimation* [VKDE; Section 6.6 of @Scott2015-vl] unless otherwise stated.
-However, these kernel density estimators are based on random samples in the Euclidean space.
-
-For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points.
-The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature [@Henry2009-ll].
-However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space.
-Therefore, the kernel density estimator from @Pelletier2005-vu is not applicable because the geodesic distance and the volume density function are unknown.
-This is when we introduce manifold learning to reduce the input data dimension.
-For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-dh), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.
-
-In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space.
-One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data.
-These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold [@Goldberg2008-co].
-To tackle this problem and measure the distortion incurred in manifold learning,
-@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself.
-By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
-The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding.
-Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others.
-The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.
-
-By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding.
-Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point.
-Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information.
-Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold.
-These density estimates are useful in many other areas, including classification, clustering and anomaly detection.
-Similar to @Cheng2021-dh, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.
-
-The rest of the paper is organized as follows.
-In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds.
-Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection.
-\autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space.
-Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
-\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset.
-Conclusions and discussions are presented in \autoref{conclusion}.
-Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.
-
--->
 
@@
 Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
-\begin{align}
+\begin{equation}
+\begin{aligned}
 \label{eq:kernelcondition}
@@
 (v) \sup K(\|\pmb{y}\|)=K(0).
-\end{align}
+\end{aligned}
+\end{equation}
+
 Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
@@
 <!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density.
-To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term.
-A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density on the manifold itself.
-Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own.
-This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.-->
+To make this clear consider the simpler case of computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term.
+A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density of the manifold itself.
+Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density on its own.
+This motivates a kernel density estimate that corrects for the distortion induced by $\psi$.-->
 
@@
   \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters EMBED }
+  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}_i$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm EMBED }
+  \OptParameter{ manifold learning parameters in EMBED }
   \BlankLine
@@
 
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
+  \STATE Obtain the $d\times d\times N$ tensor $\pmb{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with elements in $\pmb{\tilde{H}}$ found by
   $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+    \pmb{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^l \odot \pmb{y}^k\right) - \pmb{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^k\right) - \pmb{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^l\right)\right],
   $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
+  where $l,k=1,\dots,d$, $\pmb{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;
 
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
+  \STATE For each embedding data point $\pmb{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\pmb{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}^{\cdot\cdot i}}$ below 
   $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
+    \pmb{H}_i = \pmb{U} \pmb{\Lambda}_d^{-1} \pmb{U}^\prime,
   $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+  where $[\pmb{U}, \pmb{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\pmb{H}^{\cdot\cdot i}}$, $\pmb{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\pmb{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.
 
@@
 
-```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
+```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="Underlying data for the Gaussian mixture model of four kernels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
 The colors indicate the true density of the data when they are mapped via the twin peaks function.
@@
 
-```{r twinpeaks3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:twinpeaks3dcaption)", echo = FALSE}
+```{r twinpeaks3d, fig.align = 'center', out.width = "75%", fig.cap="(ref:twinpeaks3dcaption)", echo = FALSE}
 knitr::include_graphics("figures/scatterplot3d_twinpeaks_dc.png")
@@
 
-```{r tpoutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
-The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel).
+```{r tpoutliers, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
+The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
 DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
@@
   kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+Distortion-corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>%
@@
 
-```{r tpisomapvs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
+```{r tpisomapvs4ml, fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
 Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
@@
 
-```{r fivedisomapden, fig.align = 'center', out.width = "95%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE.
+```{r fivedisomapden, fig.align = 'center', out.width = "90%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE.
 Four manifold learning methods are used row-wise.
@@
 For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP there more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
+However, for Laplacian Eigenmaps and UMAP, more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
 In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
@@
   mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
+  kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data.
+DC-KDE has a slightly lower rank correlation for ISOMAP and LLE above 0.96, but significantly outperfoms KDE for Laplacian Eigenmaps and UMAP when more distortion is induced in dimension reduction.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
   # kable_styling(latex_options = "scale_down") %>% # automatically adjust the table to page width
@@
   `rownames<-` (c(">99\\% HDR", "99\\% HDR")) %>% 
-  kableExtra::kbl(caption = "Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
+  kableExtra::kbl(caption = "Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings for the 100-D data.", booktabs = TRUE, digits = 3, escape = FALSE, align = "c",
                   col.names = c(rep(c("DC-KDE", "KDE"), 4))
@@
   # kable_classic("striped") %>% 
-  # kable_styling(latex_options = "scale_down") %>%
+  kable_styling(latex_options = "scale_down") %>%
   kable_paper(full_width = FALSE) %>%
@@
 
-<!-- ## Irish smart meter dataset -->
+<!-- # Irish smart meter dataset -->
 
@@
 
-```{r electricityembedding, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
+```{r electricityembedding, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
 The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
@@
 Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
-Without distortion correction, the density estimates are different not only in scaling, but also show inconsistency in the rank of density estimates.
+Without distortion correction, the density estimates are different not only in scale, but also show inconsistency in the rank of density estimates.
 
@@
   # kable_classic("striped") %>%
-  # kable_styling(latex_options = "scale_down") %>%
+  kable_styling(latex_options = "scale_down") %>%
   kable_paper(full_width = FALSE) %>%
@@
 ```
-```{r electricityrank, fig.align = 'center', out.width = "90%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
+```{r electricityrank, fig.align = 'center', out.width = "90%", fig.cap="Comparison of ranks of the estimated densities based on UMAP and four dimension reduction algorithms in each row.
+Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows a higher level of robustness to the use of different dimension reduction methods.", echo=FALSE, eval=T}
 knitr::include_graphics("figures/Electricity_2d_rankplot_3639id_umapvs4ml_radius20_r180.png")
@@
 The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
-It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE hand UMAP do not identify any of the same anomalies as ISOMAP and LLE.
+It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.
 
-```{r electricityraw, fig.align = 'center', out.width = "90%", fig.cap="Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.", echo = FALSE, eval=T}
+```{r electricityraw, fig.align = 'center', out.width = "95%", fig.cap="Electricity usage plots of all 535 days for the most typical household (top row) and three anomalies.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/Electricity_2d_compare4id_1typical_3anomalous.png")
@@
 
-```{r electricityhdrsvkde, fig.align = 'center', out.width = "90%", fig.cap="Quantile region plots of electricity demand against the time of the week for one typical household 1472 and two anomalies, 7049 and 5136.
-The day-of-the-week patterns are much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time-related usage patterns.", echo = FALSE, eval=T}
+```{r electricityhdrsvkde, fig.align = 'center', out.width = "95%", fig.cap="Quantile region plots of electricity demand against the time of the week for one typical household 1472 and three anomalies, 3161, 4546, and 7049.
+The day-of-the-week patterns are much more different for Household 4546 and 7049 compared with 3161.", echo = FALSE, eval=T}
 knitr::include_graphics("figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous.png")
@@
 
-<!-- kernel types discussion -->
-
 # Acknowledgment {-}
@@
 The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at Monash University.
-
-<!-- \clearpage -->
-<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
-<!-- %\pagenumbering{arabic} -->
-<!-- %\def\thepage{A\arabic{page}} -->
-
-\appendix
-
-# Appendix: Notions about Riemannian geometry {#riemgeo}
-
-In this appendix, we present some notions about the Riemannian geometry used in this paper.
-
-## Differentiable manifolds
-
-In topology, a *homeomorphism* is a bijective map between two topological spaces that is continuous in both directions.
-A *Hausdorff space* is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-And a $d$-dimensional (topological) *manifold* $M$ is a connected Hausdorff space $(M, \mathcal{T}_M)$ where the neighborhood $U$ for each point $p$ is homeomorphic to an open subset $V$ of the Euclidean space $\mathbb{R}^d$.
-Such a homeomorphism $\varphi: U \rightarrow V$ together with $U$ gives a (coordinate) *chart*, denoted as $(U, \varphi)$, with the corresponding local coordinates $(x^1(p),\dots, x^d(p)) := \varphi(p)$.
-Further, a collection of charts $\{U_\alpha, \varphi_\alpha\}$ ranging over the manifold $M$ is called an *atlas*, denoted as $\mathcal{A}$.
-
-The manifold $M$ is a *differentiable manifold* if there exists an atlas of $M$, $\{U_\alpha, \varphi_\alpha\}$, such that the *transition maps* between any two charts, 
-$$
-\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-$$ are differentiable of class $C^\infty$ (smooth).
-Let $\varphi$ be an injective map: $E \rightarrow \varphi(E)$. Then $\varphi$ is an *embedding* of $E$ into $M$ if and only if
-$\varphi: E \rightarrow \varphi(E)$ is a homeomorphism, and $\varphi(E)$ is called an embedded submanifold of $M$ with the subspace topology.
-
-## Tangent vector and tangent space
-
-The tangent vector at point $p$ can be intuitively viewed as the velocity of a curve passing through point $p$ or as the directional derivatives at $p$.
-Here we define the tangent vector via the velocity of curves.
-
-For any point $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curves passing through $p$, i.e.
-$\gamma_1(0) = \gamma_2(0) = p$.
-We say $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart $(U,\varphi)$ at $p$ such that
-$$
-(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-$$ 
-A *tangent vector* to a manifold $M$ at point $p$, denoted as $v_p$, is any equivalent class of the differentiable curves initialized at $p$.
-The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$.
-The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$.
-The collection of all tangent spaces defines the *tangent bundle*, $TM = \cup_{p \in M}T_pM$.
-
-Tangent vectors can also be seen as the directional derivatives at $p$. For a given coordinate chart $\varphi=(x^1,\dots,x^d)$, the tangent vectors defining partial
-derivatives are denoted as $\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)$, which defines a *basis* of the tangent space.
-The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with the corresponding *cotangent vectors* $z_p: T^\star_pM \rightarrow \mathbb{R}^d$ and a basis denoted as $dx^1(p),\dots,dx^d(p)$.
-
-## Riemannian metric and geodesic distance
-
-A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is a $d\times d$ symmetric positive definite matrix and varies smoothly at $p$.
-Generally, we
-omit the subscript $p$ and refer to $g$ as the Riemannian metric.
-The inner product between two vectors $u, v \in T_pM$ is written as $\langle u, v \rangle_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j} g_{ij}u^iv^j$, is assumed.
-A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.
-
-The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{\langle u,v \rangle_g}$, and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$.
-It could also be used to define the line element $dl^2 = g_{ij}dx^i dx^j$ and the volume element $dV_g = \sqrt{\det(g)}dx^1 \dots dx^d$, where $(x^1,\dots,x^d)$ are the local coordinates of the chart $(U, \varphi)$.
-For a curve $\gamma: I \rightarrow M$, the length of the curve is
-$$
-l(\gamma) = \sqrt{\int_0^1 \|\gamma^\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
-$$
-where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
-$$
-Vol(W) = \int_W \sqrt{\det(g)}dx^1 \cdots dx^d,
-$$
-which is also called the *Riemannian measure* on $M$.
-
-The *geodesics* of $M$ are the smooth curves that locally join the points along the shortest path on the manifold.
-Intuitively, geodesics are the *straightest possible curves* in a Riemannian manifold [Section 7.2.3 of @Nakahara2018-zs].
-A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the second-order ordinary differential equation is satisfied, 
-$$
-\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
-$$ 
-where $\{x^i\}$ are the coordinates of the curve $\gamma$ and $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
-$$
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k} 
-+ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
-$$ 
-The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc
-length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$. 
-The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold.
-For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the distance to the cut locus is the *injectivity radius* at $p \in M$.
-Therefore, the injectivity radius of the Riemannian manifold $(M,g)$, $\textit{inj}_gM$, is the infimum of the injectivity radii over all points on the manifold.
-
-<!-- ## Isometry -->
-
-## Exponential map and logarithmic map
-
-Denote $B(p, r) \subset M$ as an open ball centered at point $p$ with radius $r$.
-Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$.
-The exponential map maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{\prime}(0)=u$, and $\gamma(1)=exp_p(u)$.
-It is a differentiable bijective map of differentiable inverse (i.e.
-*diffeomorphism*).
-Intuitively, the exponential map moves point $p$ to an endpoint at speed $u$ after covering the length of $\|u\|$ along the geodesic in one time unit.
-
-The inverse of the exponential map is called the *logarithm map*, denoted as $\log_p(q):= \exp^{-1}_p(q)$, which gives the tangent vector to get from point $p$ to $q$ in one unit time.
-Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$.
-Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.
-<!-- [TODO: reference in appendix]. -->
-
-## Pushforward and pullback metric
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
-Let $\phi: M \rightarrow E$ be a smooth map between the Riemannian manifold $(M,g)$ to another smooth manifold $E$.
-Then the differential of $\phi$ at point $p$ is a linear map $d\phi_p: T_pM \rightarrow T_{\phi (p)}E$, which pushes the tangent vector $u \in T_pM$ at point $p$ forward to the tangent vector $\phi_*u \in T_{\phi (p)}E$ at the mapping point $\phi(p)$.
-The image of the tangent vector $u \in T_pM$ under the differential $d\phi_p$, denoted as $d\phi_p u$ is called the pushforward of $u$ by the map $\phi$.
-Then pushforward metric $h=\varphi_*g$ of the Riemannian metric $g$ along $\varphi$ is given by the inner product
-$$
-\langle \phi_*u,\phi_*v \rangle_{\varphi*g} = \langle d\phi_p \phi_*u, d\phi_p \phi_*v \rangle_{g}.
-$$
-
-<!--  is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u$ at point $p \in M$ to the tangent vectors $f_*u$ at the mapping point $f(p) \in E$.
-For two tangent vectors $u, v \in T_{f(p)}E$, we have
-$$
-\langle u,v \rangle_{\varphi*g} = \langle df_p^{-1}u, df_p^{-1}v \rangle_{g},
-$$ 
-where $df_p^{-1}$ is the Jacobian of $f^{-1}$.  -->
-
-The tangent vectors $\phi_*u$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{\prime}(0)=u$,
-$$
-d\phi_p(\gamma^{\prime}(0)) = (\phi \circ \gamma)^\prime (0).
-$$
-Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.
-The linear map is called the pullback by $\phi$ and is often denoted as $\phi^*$.
-
-<!-- # Appendix: Simulation of $S^1$ for manifold density estimation {#s1simulation} -->
-
-<!-- To start with, we demonstrate the use of the proposed DC-KDE on a $1$-dimensional example.
-For a $S^1$ curve centered at $O$ with radius $1$, the coordinated of points on the curve could be expressed in polar coordinates $(1, \alpha)$ or Cartesian coordinates $(x, \sqrt{1-x^2})$.
--->
-
-<!-- The first simulation starts with a $1$-dimensional example.  -->
-
-<!-- Manifold: $S^1$ with center 0 and radius $r=1$ -->
-
-<!-- Polar coordinates: $(r, \theta)$ -->
-
-<!-- Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$. -->
-
-<!-- Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$ -->
-
-<!-- Density of $\theta$: $f(\theta) = 2 / \pi$ -->
-
-<!-- Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$ -->
-
-<!-- The CDF of X is given by -->
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!-- P(X \leq x) =& P(\cos(\theta) \leq x) \\ -->
-<!-- =& P(\theta \geq \arccos(x)) \\ -->
-<!-- =& \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\ -->
-<!-- =&  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\ -->
-<!-- =& 1 - \frac{2}{\pi} \arccos(x). -->
-<!-- \end{aligned} -->
-<!-- $$ -->
-<!-- Then the density of X is the derivative of the CDF w.r.t. $x$  -->
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!-- f(x) =& -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\ -->
-<!-- =& \frac{2}{\pi \sqrt{1 - x^2}}. -->
-<!-- \end{aligned} -->
-<!-- $$ -->
-
-<!-- Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could -->
-<!-- estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ -->
-<!-- and compare it with its true density $f(x)$. -->
-<!-- We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by -->
-<!-- $$ -->
-<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}), -->
-<!-- $$ -->
-<!-- where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.
--->
-
-<!-- Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$ with $r=1$.
-Therefore, the geodesic distance is -->
-<!-- $$ -->
-<!-- d_g(p, X_i) = \arccos(p) - \arccos(X_i) -->
-<!-- $$ -->
-<!-- As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e.
-$|g(\exp_p^{-1}(x_i))|$.
--->
-
-<!-- [When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.] -->
-
-<!-- The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$, -->
-<!-- $$ -->
-<!-- g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R}, -->
-<!-- $$ -->
-<!-- is smooth in a local neighborhood $U$ of $M$. -->
-
-<!-- Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by -->
-<!-- $$ -->
-<!-- \theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}. -->
-<!-- $$ -->
-<!-- Therefore, the estimator for $p \in S^1$ is given by -->
-<!-- $$ -->
-<!-- f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg), -->
-<!-- $$ -->
-<!-- where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*. -->
-
-<!-- ### How to derive volume density function $\theta_{X_i}(p)$ -->
-
-<!-- #### Example of $S^2$ in @Henry2009-om. -->
-
-<!-- Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$.
-Let $v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$ is the orthonormal basis of $\mathbb{R}^3$.
-Consider the exponential chart $(U, \Psi)$ induced by the parametrisation $\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by -->
-<!-- $$ -->
-<!-- \Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw), -->
-<!-- $$ -->
-<!-- if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where $r=\sqrt{s^2 + t^2}$. -->
-<!-- Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$. -->
-<!-- Then we take the partial derivatives of $\Psi^{-1}(s,t)$ with respect to the basis of the tangent space, $s$ and $t$, and have that -->
-<!-- $$ -->
-<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w, -->
-<!-- $$ -->
-<!-- $$ -->
-<!-- B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w. -->
-<!-- $$ -->
-<!-- The coefficients for $v$ and $w$ gives the local coordinates of two tangent vectors. -->
-<!-- Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as $g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$.
--->
-<!-- Therefore, we obtain that the volume density on the sphere is -->
-<!-- $$ -->
-<!-- \theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2}  -->
-<!-- = \frac{|\sin(d_g(p,q))|}{d_g(p,g)}, -->
-<!-- $$ -->
-<!-- for $q \neq p, -p$, and $\theta_p(p) = 1$. -->
-
-<!-- #### $S^1$ -->
-
-<!-- Similarly in $S^1$, we could let $\{v, p\}$ be the orthonormal basis of $\mathbb{R}^2$. Then the exponential chart is induced by the parametrization -->
-<!-- $$ -->
-<!-- \Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv, -->
-<!-- $$ -->
-<!-- and $\Psi^{-1}(0) = p$ where $r = s$. Again $r = d_g(p,q)$ if $q = \exp_p(sv)$. Then we have -->
-<!-- $$ -->
-<!-- A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v. -->
-<!-- $$ -->
-<!-- So the riemannian metric is $g_q = \cos(r)$ and further, -->
-<!-- $$ -->
-<!-- \theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|}, -->
-<!-- $$ -->
-<!-- where the geodesic distance is $d_g(p,q) = \arccos(p) - \arccos(q)$, and $\theta_p(p) = 1$. -->
-
-# Appendix: Rank comparison plots for twin peaks mapping {#twinpeaksappe}
-
-This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
-By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest-density regions more accurately than KDE.
-By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest-density regions, which are usually used to detect anomalies.
-
-```{r tplleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of LLE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_lle_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
-```
-
-```{r tpleden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of Laplacian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_le_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
-```
-
-```{r tpumapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of true density and estimated density ranks of UMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.", echo = FALSE}
-knitr::include_graphics("figures/Twin Peak2000_densityrank_comparison_umap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse.png")
-```
-
-```{r tpllevs4ml, fig.align = 'center', out.width = "95%", fig.cap="Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.", echo = FALSE}
-# knitr::include_graphics("figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank.png")
-knitr::include_graphics("figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank.png")
-knitr::include_graphics("figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank.png")
-knitr::include_graphics("figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank.png")
-knitr::include_graphics("figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank.png")
-```
-
-<!-- \clearpage -->
-
-<!-- # Appendix: Simulation with swiss roll mapping {#swissrollappe}
-
-In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.
-
-One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in Equation \@ref(eq:swissroll).
-The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in \autoref{fig:swissroll3d}.
-The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$.
-\begin{equation}
-\label{eq:swissroll}
-\left\{
-\begin{array}{lcl}
-X = X_1 \cos{X_1}, \\
-Y = X_2, \\
-Z = X_1 \sin{X_1}.
-\end{array}
-\right.
-\end{equation}
-
-(ref:swissroll3dcaption) Scatterplot of the 3-d swiss roll data with colors indicating the true density via the swiss roll mapping in \autoref{eq:swissroll}.
-
-```{r swissroll3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:swissroll3dcaption)", echo = FALSE}
-knitr::include_graphics("figures/scatterplot3d_swissroll_dc.png")
-```
-
-```{r sroutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the swiss roll data.
-Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth.
-Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE.
-The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.", echo = FALSE}
-knitr::include_graphics("figures/Swiss Roll_outliers_comparison_4ml_3cases_riem0_08.png")
-```
-
-Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding.
-According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg.
-1%, 5%, 50%, 99%, \>99%.
-For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
-the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth.
-The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
-From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions.
-However, in all
-manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data.
-When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
-In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately.
-For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.
-
-```{r srcors, echo=FALSE, message=FALSE, eval=T}
-load("figures/CorrelationTable_Swiss Roll_4ml_riem0_08.rda")
-col.max <- apply(cors, 2, max)
-max.match <- rep(col.max, each = 2) == cors
-data.frame(cors) %>% 
-  mutate_if(is.numeric, format, nsmall = 3, digits = 3) %>% 
-  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the swiss roll data.
-Variable bandwidth KDE outperfoms for LLE and UMAP, and LLE gives the highest rank correlation.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
-  # kable_styling(latex_options = "scale_down") %>%
-  kable_paper(full_width = FALSE) %>%
-  # column_spec(1, width = "6cm") %>% 
-  column_spec(2, bold = max.match[,1]) %>%
-  column_spec(3, bold = max.match[,2]) %>%
-  column_spec(4, bold = max.match[,3]) %>%
-  column_spec(5, bold = max.match[,4]) %>%
-  column_spec(6, bold = max.match[,5])
-```
-
-To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}.
-It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero.
-The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the
-rectangular structure of the meta data shown in \autoref{fig:metadensity}.
-For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative
-correlation.
-The negative correlation would occur typically when the highest or lowest true density areas are not well estimated.
-As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low.
-This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric.
-However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.
- -->
-<!-- Since there are four kernels in the meta data and darker points are the defined outliers, we could derive that for ISOMAP and LLE, variable kernel density estimates could detect the number of kernels more accurately than the KDE with fixed bandwidth.
-As for t-SNE and UMAP, both embeddings are quite close to each other with points clustered together, but the mixture of kernel structure is not as clear as LLE.
-Instead, we could roughly compare the location of outliers.
-For t-SNE, the middle plot has indexed more true outliers than the bottom plot; while for UMAP, both the middle and bottom plot fail to find the true outliers, but the outliers from fixed bandwidth are more scattered along with the curved embedding.
--->

### 2023-06-01 jcgs/kderm_jcgs.Rmd
Add jcgs version for kderm, the latest version

--- 2022-11-12 paper/paper.tex
@@
-\documentclass[11pt,a4paper,]{article}
-\usepackage{lmodern}
-
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\hypersetup{
-            pdftitle={Distortion corrected kernel density estimate on Riemannian manifolds},
-            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{geometry}
-\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
-\usepackage[style=authoryear-comp,]{biblatex}
-\addbibresource{references.bib}
-\usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\title{Distortion corrected kernel density estimate on Riemannian manifolds}
-
-%% MONASH STUFF
-
-%% CAPTIONS
-\RequirePackage{caption}
-\DeclareCaptionStyle{italic}[justification=centering]
- {labelfont={bf},textfont={it},labelsep=colon}
-\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
-\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
-
-%% FONT
-\RequirePackage{bera}
-\RequirePackage{mathpazo}
-
-%% HEADERS AND FOOTERS
-\RequirePackage{fancyhdr}
-\pagestyle{fancy}
-\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
-\makeatletter
-\lhead{\textsf{\expandafter{\@title}}}
-\makeatother
-\rhead{}
-\cfoot{}
-\setlength{\headheight}{15pt}
-\renewcommand{\headrulewidth}{0.4pt}
-\renewcommand{\footrulewidth}{0.4pt}
-\fancypagestyle{plain}{%
-\fancyhf{} % clear all header and footer fields
-\fancyfoot[C]{\sffamily\thepage} % except the center
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrulewidth}{0pt}}
-
-%% MATHS
-\RequirePackage{bm,amsmath}
-\allowdisplaybreaks
-
-%% GRAPHICS
-\RequirePackage{graphicx}
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.85}
-\renewcommand{\bottomfraction}{0.85}
-\renewcommand{\textfraction}{0.15}
-\renewcommand{\floatpagefraction}{0.8}
-
-%\RequirePackage[section]{placeins}
-
-%% SECTION TITLES
-\RequirePackage[compact,sf,bf]{titlesec}
-\titleformat{\section}[block]
-  {\fontsize{15}{17}\bfseries\sffamily}
-  {\thesection}
-  {0.4em}{}
-\titleformat{\subsection}[block]
-  {\fontsize{12}{14}\bfseries\sffamily}
-  {\thesubsection}
-  {0.4em}{}
-\titlespacing{\section}{0pt}{*5}{*1}
-\titlespacing{\subsection}{0pt}{*2}{*0.2}
-
-%% TITLE PAGE
-\def\Date{\number\day}
-\def\Month{\ifcase\month\or
- January\or February\or March\or April\or May\or June\or
- July\or August\or September\or October\or November\or December\fi}
-\def\Year{\number\year}
-
-\makeatletter
-\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
-\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
-\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
-\def\nojel{\def\showjel{}}
-\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
-\def\cover{{\sffamily\setcounter{page}{0}
-        \thispagestyle{empty}
-        \placefig{2}{1.5}{width=5cm}{monash2}
-        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
-        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
-        \begin{textblock}{7}(12.7,27.9)\hfill
-        \includegraphics[height=0.7cm]{AACSB}~~~
-        \includegraphics[height=0.7cm]{EQUIS}~~~
-        \includegraphics[height=0.7cm]{AMBA}
-        \end{textblock}
-        \vspace*{2cm}
-        \begin{center}\Large
-        Department of Econometrics and Business Statistics\\[.5cm]
-        \footnotesize http://monash.edu/business/ebs/research/publications
-        \end{center}\vspace{2cm}
-        \begin{center}
-        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
-                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
-                \LARGE\@author\end{onehalfspace}
-        }}
-        \end{center}
-        \vfill
-                \begin{center}\Large
-                \Month~\Year\\[1cm]
-                Working Paper \@wp
-        \end{center}\vspace*{2cm}}}
-\def\pageone{{\sffamily\setstretch{1}%
-        \thispagestyle{empty}%
-        \vbox to \textheight{%
-        \raggedright\baselineskip=1.2cm
-     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
-        \vspace{2cm}\par
-        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
-        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
-        \hspace{1cm}\showjel\vss}}}
-\def\blindtitle{{\sffamily
-     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
-     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
-        }}
-\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}
-
-\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
-\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
-\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
-\let\maketitle\titlepage
-\makeatother
-
-%% SPACING
-\RequirePackage{setspace}
-\spacing{1.5}
-
-%% LINE AND PAGE BREAKING
-\sloppy
-\clubpenalty = 10000
-\widowpenalty = 10000
-\brokenpenalty = 10000
-\RequirePackage{microtype}
-
-%% PARAGRAPH BREAKS
-\setlength{\parskip}{1.4ex}
-\setlength{\parindent}{0em}
-
-%% HYPERLINKS
-\RequirePackage{xcolor} % Needed for links
-\definecolor{darkblue}{rgb}{0,0,.6}
-\RequirePackage{url}
-
-\makeatletter
-\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
-\makeatother
-\hypersetup{
-     citecolor=0 0 0,
-     breaklinks=true,
-     bookmarksopen=true,
-     bookmarksnumbered=true,
-     linkcolor=darkblue,
-     urlcolor=blue,
-     citecolor=darkblue,
-     colorlinks=true}
-
-%% KEYWORDS
-\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}
-
-%% ABSTRACT
-\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
-\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
-  {\end{minipage}}
-
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-
-\usepackage[showonlyrefs]{mathtools}
-\usepackage[no-weekday]{eukdate}
-
-%% BIBLIOGRAPHY
-
-\makeatletter
-\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
-\makeatother
-\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}
-
-\DeclareFieldFormat{url}{\texttt{\url{#1}}}
-\DeclareFieldFormat[article]{pages}{#1}
-\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
-\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
-\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
-\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
-\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
-\DeclareFieldFormat[inproceedings]{title}{#1}
-\DeclareFieldFormat{shorthandwidth}{#1}
-% No dot before number of articles
-\usepackage{xpatch}
-\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
-% Remove In: for an article.
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-
-\makeatletter
-\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
-\makeatother
-\renewcommand*{\finalnamedelim}{%
-  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
-  \addspace\&\space}
-
-\nojel
-
-\RequirePackage[absolute,overlay]{textpos}
-\setlength{\TPHorizModule}{1cm}
-\setlength{\TPVertModule}{1cm}
-\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}
-
-\nocover
-
-\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
-\addresses{\textbf{Fan Cheng}\newline
-Monash University
-\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
-\textbf{Anastasios Panagiotelis}\newline
-The University of Sydney
-\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
-\textbf{Rob J Hyndman}\newline
-Monash University
-\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
-}
-
-\date{\sf\Date~\Month~\Year}
-\makeatletter
- \lfoot{\sf Cheng, Panagiotelis, Hyndman: \@date}
-\makeatother
-
-%% Any special functions or other packages can be loaded here.
-\usepackage{tabu,placeins,algorithmic}
-\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
-\urlstyle{tt}  % use monospace font for urls
-\usepackage[cal = txupr]{mathalpha}
-\geometry{margin=2.2cm}
-\mathtoolsset{showonlyrefs}
-\def\sectionautorefname{Section}
-\def\subsectionautorefname{Section}
-% \def\equationautorefname~#1\null{Equation~(#1)\null} % NOT showing Equation 1, but Section 2.1
-\usepackage{bm}
-\DeclareMathOperator\supp{supp}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-
-% Adjust headwidth in case user has changed geometry in header-includes
-\renewcommand{\headwidth}{\textwidth}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
+---
+title: "Distortion corrected kernel density estimate on Riemannian manifolds"
+blinded: 0 # Set to 1 for blinding
+authors:
+  - name: Fan Cheng
+    affiliation: Monash University
+    thanks: Fan.Cheng@monash.edu
+  - name: Rob J Hyndman
+    affiliation: Monash University
+  - name: Anastasios Panagiotelis
+    affiliation: The University of Sydney    
+abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
 However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
@@
 The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D data from the Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
 We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
@@
 The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
-\end{abstract}
-\begin{keywords}
-manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
-\end{keywords}
+This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
+keywords:
+  - manifold learning
+  - variable bandwidth
+  - Riemannian metric
+  - geodesic distance
+  - Gaussian kernels
+bibliography: kderm.bib
+output:
+  bookdown::pdf_document2:
+    template: asa_template.tex
+    fig_caption: yes
+    fig_height: 5
+    fig_width: 8
+    dev: "pdf"
+    keep_tex: yes
+    includes:
+      in_header: preamble.tex
+# after_body: appendix.tex
+link-citations: yes
+---
 
-\newpage
+```{r child="03-chap3-kderm.Rmd"}
 
-\hypertarget{intro}{%
-\section{Introduction}\label{intro}}
+```
 
-Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
-Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
-For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality.
-However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\).
-The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
+```{r child="B-appB-kderm.Rmd"}
 
-When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\).
-Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx} and \textcite{Cheng2021-dh}.
-One could easily apply KDE to the output vectors \(\pmb{y}\).
-However, this would be a misleading representation of the density of points as they lie on the manifold.
-To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,\]
-where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\).
-By a change of variables, this would imply the following density for \(\pmb{y}\),
-\begin{equation}
-\label{eq:changevar}
-Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
-\end{equation}
-where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
-It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|\pmb{H}(\pmb{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant.
-Since \(\psi\) will typically be a non-linear function, the value of \(|\pmb{H}(\pmb{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
-Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
-
-In this paper, we propose a new KDE.
-Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm.
-For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
-In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
-The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
-The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.
-
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
-This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
-Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images.
-In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
-Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.
-
-The rest of the paper is organized as follows.
-In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
-\autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
-\autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
-Conclusions and discussions are presented in \autoref{conclusion}.
-Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
-
-\hypertarget{kderm}{%
-\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}}
-
-In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold.
-In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
-
-In the following, we denote \(M\) as the \(d\)-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted \(\pmb{p}\) in general, with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
-Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case.
-For instance, \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each \(\pmb{p}_i\) only require some sense of distance between the `input' points, \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\pmb{y}_1,\dots,\pmb{y}_n\), where \(\pmb{y}_i\in\mathbb{R}^d\).
-We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\).
-Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
-\begin{align}
-\label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
-(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
-(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
-& (iv) \supp K=[0; 1];
-(v) \sup K(\|\pmb{y}\|)=K(0).
-\end{align}
-Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
-For instance, condition (iv) requires the support of the kernel to be bounded.
-The reasons for this will become clearer when we discuss the manifold setting in more detail.
-Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise.
-In our empirical section, more general kernel functions can be, and are, employed.
-
-For data \(\pmb{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\pmb{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{y}\) is given by
-\begin{equation}
-\label{eq:vkde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
-\end{equation}
-
-The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point \(\pmb{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{y}\), times a term that ensures the density integrates to 1.
-In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
-For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein.
-Kernel density estimators of this form can and have been applied directly on the output embedding \(\pmb{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
-
-\hypertarget{Pellet}{%
-\subsection{Kernel Density estimation on manifolds}\label{Pellet}}
-
-For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
-\begin{equation}
-\label{eq:denriem}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
-\end{equation}
-
-where \(d_g(\pmb{p}, \pmb{p}_i)\) denotes the geodesic distance between two points on the manifold \(\pmb{p}\) and \(\pmb{p}_i\) and \(\theta_{\pmb{p}_i}(\pmb{p})\) is known as the volume density function.
-The intuition behind the term \(K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)\) is relatively clear.
-For example, for a uniform kernel, the estimator at point \(\pmb{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
-However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
-An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold.
-A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
-For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once.
-For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
-This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
-
-The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
-We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
-However, the volume of this ball must also be taken into account.
-In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.
-
-More formally, the volume density function can be explained as follows.
-Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space, \(\pmb{v}\in T_{\pmb{p}}M\), to points on the manifold, \(\pmb{q}\in M\).
-Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and traveling along this geodesic at uniform speed \(\|\pmb{v}\|\) takes place in one unit of time.
-Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
-The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\pmb{H}(\pmb{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\pmb{p}_i\).
-For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.
-
-\hypertarget{MetLearn}{%
-\subsection{Riemannian metric estimation}\label{MetLearn}}
-
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
-Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
-To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
-These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
-
-While full details are provided in \textcite{Perrault-Joncas2013-pq}, we briefly describe the Learn Metric algorithm here.
-There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold.
-Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-Third, a dimension reduction method is applied to obtain the output embedding \(\pmb{y}_1,\dots,\pmb{y}_n\).
-Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
-Full details on these four steps are provided in \autoref{alg:learnmetric}.
-This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
-Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.
-
-As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
-In this case, the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\).
-Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean space.
-In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into \(d\)-dimesional Euclidean space.
-In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used (See Chapter 3 of \textcite{Lee2007-wq} for more about intrinsic dimension estimation).
-The \emph{R} library \emph{intRinsic} \autocite{Denti2021-qc} implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
-
-\begin{algorithm}[!htb]
-  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
-  \label{alg:learnmetric}
-  \DontPrintSemicolon
-  \SetAlgoLined
-  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters EMBED }
-  \BlankLine
-  \begin{algorithmic}[1]
-
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
-
-  \STATE Calculate the $N\times N$ geometric graph Laplacian $\widetilde{\mathcal{L}}_{\varepsilon,N}$ by
-  $$
-  \widetilde{\mathcal{L}}_{\varepsilon,N} = 1/(c\varepsilon)(\widetilde{D}^{-1} \widetilde{W} - I_N),
-  $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
-
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
-
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
-  $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
-  $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
-  $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
-  $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
-
-  \end{algorithmic}
-\end{algorithm}
-
-\hypertarget{DCKDE}{%
-\subsection{Distortion corrected KDE}\label{DCKDE}}
-
-With all fundamentals introduced, we can now give our novel Distortion Corrected KDE (DC-KDE) as
-\begin{equation}
-\label{eq:denestimator}
-\hat{f}(\pmb{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|}{r} \bigg).
-\end{equation}
-
-The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
-This is crucial for two reasons.
-First, this implies that the term \(\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|\) approximates the geodesic distance between \(\pmb{y}_i\) and \(\pmb{y}_j\).
-Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{y}_i\), and it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
-For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
-Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
-
-One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\pmb{H}_j\).
-To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density.
-The entire workflow is summarized in \autoref{fig:vkde}.
-The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
-With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/DC-KDE}
-
-}
-
-\caption{The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.}\label{fig:vkde}
-\end{figure}
-
-\hypertarget{simulation}{%
-\section{Simulations}\label{simulation}}
-
-In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion-corrected KDE.
-For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 4-dimensional manifold embedded in a 100-dimensional ambient space.
-To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-In general, we aim to highlight two advantages of our proposed distortion-corrected KDE compared to KDE applied directly to the output coordinates.
-First, the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
-Second, we show how density estimation and anomaly detection are more robust to a different choice of dimension reduction method when distortion correction is used.
-
-\hypertarget{twodgaussian}{%
-\subsection{Twin peaks example}\label{twodgaussian}}
-
-The simulation setup for the twin peaks example is to first generate vectors \(\pmb{v}_1,\dots,\pmb{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
-The mixture has four components with different means
-\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\).
-The mixture proportions
-are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two-dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the following
-\begin{equation}
-\label{eq:twinpeak}
-\begin{array}{lcl}
-x_1 = v_1, \\
-x_2 = v_2, \\
-x_3 = \sin(\pi v_1) \tan (3 v_2).
-\end{array}
-\end{equation}
-
-The three-dimensional twin peaks mapping is shown in \autoref{fig:twinpeaks3d}.
-The colors in both \autoref{fig:metadensity} and \autoref{fig:twinpeaks3d} indicate the true density of the data via the twin peaks mapping, with lower density points in darker colors scattered in the outer as well as center areas.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/truedensity_twinpeaks_dc_labv}
-
-}
-
-\caption{Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
-The colors indicate the true density of the data when they are mapped via the twin peaks function.
-Lower density points in darker colors are scattered both in the outer and center areas.}\label{fig:metadensity}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{figures/scatterplot3d_twinpeaks_dc}
-
-}
-
-\caption{Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.}\label{fig:twinpeaks3d}
-\end{figure}
-
-It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in Equation \eqref{eq:twinpeak} distorts the distribution.
-To recover the true distribution requires the correct Jacobian term for the pushforward from \(\pmb{v}\) to the volume form of twin peaks manifold.
-By treating the \(\pmb{v}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\pmb{\Gamma}_i\) for \(i=1,\dots,n\) where \(\pmb{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\pmb{v}\).
-This notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as \(f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is the density of a four-component mixture of normals.
-Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' for densities on the manifold.
-\autoref{fig:metadensity} shows the simulated \(\pmb{v}\) with color indicating the true density of data on the manifold.
-Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in yellow.
-The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the \(\pmb{v}\).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
-
-}
-
-\caption{Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
-The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel).
-DC-KDE finds more true outliers than KDE in all five rows.}\label{fig:tpoutliers}
-\end{figure}
-
-\autoref{fig:tpoutliers} summarizes the results.
-Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively.
-We set the bandwidth parameter in the DC-KDE \eqref{eq:denestimator} as \(r=0.5\).
-Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
-For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
-On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.
-The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
-For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
-This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
-
-\begin{table}
-
-\caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & t.SNE & UMAP\\
-\midrule
-DC-KDE & \textbf{0.823} & \textbf{0.673} & \textbf{0.672} & \textbf{0.806} & \textbf{0.794}\\
-KDE & 0.798 & 0.500 & 0.606 & 0.451 & 0.469\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from KDE with and without density correction by \autoref{tab:tpcors}.
-Distortion correction improves the rank correlation for all dimension reduction algorithms.
-In particular, while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_isomap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
-The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
-The shading contains all anomalies that are both truly and correctly identified.
-KDE without distortion correction gives more misclassified anomalies.}\label{fig:tpisomapden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_tsne_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding.
-KDE without distortion correction gives many more misclassified anomalies.}\label{fig:tptsneden}
-\end{figure}
-
-In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction.
-Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a \(99\%\) highest density region.
-The middle shaded region contains anomalies that are true positives in the sense of not lying in a \(90\%\) HDR.
-Points lying outside squares (shown as triangles) are incorrectly classified.
-For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used).
-Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
-\autoref{fig:tptsneden} shows the same plot but for t-SNE.
-The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified.
-However, the difference between KDE with and without distortion correction is stark.
-These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
-Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.}\label{fig:tpisomapvs4ml}
-\end{figure}
-
-Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of a dimension reduction algorithm.
-Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
-The left column shows results when distortion correction is applied, and the right column when it is not applied.
-It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
-This is critical since conclusions will be more robust to the choice of dimension reduction algorithm.
-
-\hypertarget{fivedgaussian}{%
-\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
-
-}
-
-\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
-Distant points in darker colors could be seen as anomalies.}\label{fig:fivedmeta}
-\end{figure}
-
-As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
-To start with, we simulate vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), with the same means \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\pmb{\Sigma}_1 = diag(1,1,1,1)\) and \(\pmb{\Sigma}_2 = diag(2,2,2,2)\).
-The mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-With this design, the observations from the second component tend to be outlying anomalies.
-The data are mapped to a hemisphere via the equation
-\(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
-The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
-\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the center of the distribution.
-It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
-
-To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to \(\pmb{v}_i\) so that \(\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\).
-Next, we rotate the \(100\)-dimensional vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) by multiplying by a randomly generated rotation matrix.
-To generate the rotation matrix we first simulate elements from a uniform \((0,1)\) distribution, stack them into a \(100\times 100\) matrix \(\pmb{A}\) and then take the R matrix from the QR decomposition of \(\pmb{A}\).
-Rotating the vectors results in input vectors that are no longer sparse.
-Nonetheless, the intrinsic dimension of this is still \(d=4\).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec}
-
-}
-
-\caption{Rank comparison between the true density and estimated density from both DC-KDE and KDE.
-Four manifold learning methods are used row-wise.
-The point shapes indicate whether they are the true outliers, and the grey shading highlights the top 1\% rank region.
-The colors show the distance to the center of the semisphere, with darker points being distant from the center.}\label{fig:fivedisomapden}
-\end{figure}
-
-Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density.
-The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example.
-In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densities from DC-KDE and KDE are presented.
-For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP there more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
-In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
-
-Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
-As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.
-The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion-corrected KDE and panels on the right showing KDE without distortion correction.
-The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
-
-Comparing the left and right panels for each row, we notice that KDE with distortion correction has fewer misclassified observations, and therefore outperforms KDE.
-For UMAP, almost all ground truth anomalies (points outside a \(99\%\) HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
-For DC-KDE in the left panel there are several triangles in a horizontal line, this arises, due to many points having an estimated density of close to zero (values less than \(10^{-7}\)) which leads to tied ranks.
-These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
-
-\begin{table}
-
-\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
-\centering
-\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
-\toprule
-  & ISOMAP & LLE & Laplacian.Eigenmaps & UMAP\\
-\midrule
-DC-KDE & 0.968 & 0.970 & \textbf{0.8674} & \textbf{0.782}\\
-KDE & \textbf{0.976} & \textbf{0.971} & 0.0328 & -0.181\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:fourdhdrtable}Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.}
-\centering
-\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{UMAP} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
-  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
-\midrule
->99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.490 & \cellcolor[HTML]{E6E6E6}{\textbf{0.830}} & 0.240 & \cellcolor[HTML]{E6E6E6}{\textbf{0.840}} & 0.840 & \cellcolor[HTML]{E6E6E6}{\textbf{0.770}} & 0.000\\
-99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.818}} & 0.685 & \cellcolor[HTML]{E6E6E6}{\textbf{0.805}} & 0.478 & \cellcolor[HTML]{E6E6E6}{\textbf{0.815}} & 0.595 & \cellcolor[HTML]{E6E6E6}{\textbf{0.632}} & 0.032\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\autoref{tab:fourdhdrtable} further illustrates the proportions of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
-Compared with KDE, the proportions are always higher when distortion correction methods are used.
-The proportions with distortion correction are very close for ISOMAP, LLE, and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
-This could be due to the severe distortion usually induced by the UMAP algorithm.
-
-\hypertarget{application}{%
-\section{Application}\label{application}}
-
-In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
-The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
-We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
-
-For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week \autocite{Hyndman2018-ia}.
-This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
-\textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection.
-To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
-We use the highest density regions plot to visualize the density estimates.
-However, for this real data set with an unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
-
-Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here.
-For each household, a discrete approximation of the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
-For any pair of households the total variation metric can be found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
-The number of nearest neighbors in manifold learning \(K\) is set as \(100\) and the bandwidth parameter \(r=180\) to match the distances between the distributions of households.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_5levels_outliers_comparison_5ml_r180}
-
-}
-
-\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityembedding}
-\end{figure}
-\begin{table}
-
-\caption{\label{tab:electricityoutliers}Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding.
-The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
-The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabu} to \linewidth {>{}c>{}c>{}c>{}c>{}c>{}c|>{}c>{}c>{}c>{}c>{}c}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{5}{c}{DC-KDE} & \multicolumn{5}{c}{KDE} \\
-\cmidrule(l{3pt}r{3pt}){2-6} \cmidrule(l{3pt}r{3pt}){7-11}
-  & ISOMAP & LLE & Laplacian Eigenmaps & t-SNE & UMAP & ISOMAP & LLE & Laplacian Eigenmaps & t-SNE & UMAP\\
-\midrule
-1 & \textcolor[HTML]{000000}{6317} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{4826} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{6317} & \textcolor[HTML]{000000}{1881} & \textcolor[HTML]{000000}{6317} & \textcolor[HTML]{000000}{4190} & \textcolor[HTML]{000000}{6533}\\
-2 & \textcolor[HTML]{000000}{1353} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{5269} & \textcolor[HTML]{000000}{1069} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{4884} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{6643} & \textcolor[HTML]{000000}{6817} & \textcolor[HTML]{000000}{5899}\\
-3 & \textcolor[HTML]{000000}{4884} & \textcolor[HTML]{000000}{1881} & \textcolor[HTML]{000000}{4875} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{1069} & \textcolor[HTML]{000000}{5092} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{000000}{6764} & \textcolor[HTML]{000000}{6492} & \textcolor[HTML]{000000}{6345}\\
-4 & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{6952} & \textcolor[HTML]{000000}{7300} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{4862} & \textcolor[HTML]{000000}{1353} & \textcolor[HTML]{000000}{4233} & \textcolor[HTML]{000000}{1058} & \textcolor[HTML]{000000}{2309} & \textcolor[HTML]{000000}{1884}\\
-5 & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{6838} & \textcolor[HTML]{000000}{4880} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{3639} & \textcolor[HTML]{000000}{5580} & \textcolor[HTML]{2171B5}{\textbf{7049}} & \textcolor[HTML]{000000}{5105} & \textcolor[HTML]{000000}{2546} & \textcolor[HTML]{000000}{3278}\\
-\addlinespace
-6 & \textcolor[HTML]{000000}{1366} & \textcolor[HTML]{000000}{3874} & \textcolor[HTML]{000000}{5599} & \textcolor[HTML]{000000}{4265} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{5594} & \textcolor[HTML]{000000}{1821} & \textcolor[HTML]{000000}{6577} & \textcolor[HTML]{000000}{2907} & \textcolor[HTML]{000000}{5463}\\
-7 & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{1069} & \textcolor[HTML]{000000}{3222} & \textcolor[HTML]{000000}{4862} & \textcolor[HTML]{000000}{2249} & \textcolor[HTML]{000000}{2576} & \textcolor[HTML]{2171B5}{\textbf{4546}} & \textcolor[HTML]{000000}{3880} & \textcolor[HTML]{000000}{4536} & \textcolor[HTML]{000000}{5504}\\
-8 & \textcolor[HTML]{000000}{1074} & \textcolor[HTML]{000000}{2249} & \textcolor[HTML]{000000}{1776} & \textcolor[HTML]{000000}{3639} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{1366} & \textcolor[HTML]{000000}{3431} & \textcolor[HTML]{000000}{5358} & \textcolor[HTML]{000000}{6898} & \textcolor[HTML]{000000}{1782}\\
-9 & \textcolor[HTML]{000000}{6681} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{1161} & \textcolor[HTML]{000000}{5546} & \textcolor[HTML]{000000}{3892} & \textcolor[HTML]{000000}{6681} & \textcolor[HTML]{000000}{1569} & \textcolor[HTML]{000000}{1366} & \textcolor[HTML]{000000}{4128} & \textcolor[HTML]{000000}{5965}\\
-10 & \textcolor[HTML]{000000}{1701} & \textcolor[HTML]{000000}{1569} & \textcolor[HTML]{000000}{3655} & \textcolor[HTML]{000000}{4896} & \textcolor[HTML]{000000}{6209} & \textcolor[HTML]{000000}{5291} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{5545} & \textcolor[HTML]{000000}{7368} & \textcolor[HTML]{000000}{6081}\\
-\addlinespace
-11 & \textcolor[HTML]{000000}{3136} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{6169} & \textcolor[HTML]{000000}{4555} & \textcolor[HTML]{000000}{6952} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{6568} & \textcolor[HTML]{000000}{2160} & \textcolor[HTML]{000000}{2232} & \textcolor[HTML]{000000}{4894}\\
-12 & \textcolor[HTML]{000000}{1881} & \textcolor[HTML]{000000}{4555} & \textcolor[HTML]{000000}{4830} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{4896} & \textcolor[HTML]{000000}{1701} & \textcolor[HTML]{000000}{7300} & \textcolor[HTML]{000000}{6585} & \textcolor[HTML]{000000}{6273} & \textcolor[HTML]{000000}{5937}\\
-13 & \textcolor[HTML]{000000}{6785} & \textcolor[HTML]{000000}{6526} & \textcolor[HTML]{000000}{2284} & \textcolor[HTML]{000000}{6952} & \textcolor[HTML]{000000}{6838} & \textcolor[HTML]{000000}{4749} & \textcolor[HTML]{000000}{7438} & \textcolor[HTML]{000000}{5824} & \textcolor[HTML]{000000}{1557} & \textcolor[HTML]{000000}{5522}\\
-14 & \textcolor[HTML]{000000}{5092} & \textcolor[HTML]{000000}{3639} & \textcolor[HTML]{000000}{6526} & \textcolor[HTML]{000000}{6838} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{6699} & \textcolor[HTML]{000000}{4401} & \textcolor[HTML]{000000}{4932} & \textcolor[HTML]{000000}{3500} & \textcolor[HTML]{000000}{5580}\\
-15 & \textcolor[HTML]{000000}{5269} & \textcolor[HTML]{6BAED6}{\textbf{6199}} & \textcolor[HTML]{000000}{1694} & \textcolor[HTML]{6BAED6}{\textbf{1964}} & \textcolor[HTML]{000000}{1569} & \textcolor[HTML]{000000}{5504} & \textcolor[HTML]{000000}{3434} & \textcolor[HTML]{000000}{6410} & \textcolor[HTML]{000000}{1932} & \textcolor[HTML]{000000}{6902}\\
-\addlinespace
-16 & \textcolor[HTML]{000000}{2576} & \textcolor[HTML]{000000}{5531} & \textcolor[HTML]{000000}{3431} & \textcolor[HTML]{000000}{7104} & \textcolor[HTML]{000000}{3213} & \textcolor[HTML]{000000}{3170} & \textcolor[HTML]{000000}{2249} & \textcolor[HTML]{000000}{3023} & \textcolor[HTML]{000000}{3119} & \textcolor[HTML]{000000}{7243}\\
-17 & \textcolor[HTML]{000000}{1262} & \textcolor[HTML]{000000}{4875} & \textcolor[HTML]{000000}{2523} & \textcolor[HTML]{000000}{3892} & \textcolor[HTML]{000000}{4401} & \textcolor[HTML]{000000}{5899} & \textcolor[HTML]{000000}{3213} & \textcolor[HTML]{000000}{1497} & \textcolor[HTML]{000000}{3809} & \textcolor[HTML]{000000}{6699}\\
-18 & \textcolor[HTML]{000000}{5966} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{6766} & \textcolor[HTML]{000000}{2658} & \textcolor[HTML]{2171B5}{\textbf{3161}} & \textcolor[HTML]{000000}{7293} & \textcolor[HTML]{000000}{6108} & \textcolor[HTML]{000000}{5549}\\
-19 & \textcolor[HTML]{000000}{5937} & \textcolor[HTML]{000000}{3892} & \textcolor[HTML]{000000}{2233} & \textcolor[HTML]{000000}{6799} & \textcolor[HTML]{000000}{4265} & \textcolor[HTML]{000000}{5189} & \textcolor[HTML]{6BAED6}{\textbf{1533}} & \textcolor[HTML]{000000}{5580} & \textcolor[HTML]{000000}{4516} & \textcolor[HTML]{000000}{5048}\\
-20 & \textcolor[HTML]{6BAED6}{\textbf{5356}} & \textcolor[HTML]{000000}{4896} & \textcolor[HTML]{000000}{5125} & \textcolor[HTML]{000000}{6209} & \textcolor[HTML]{000000}{6799} & \textcolor[HTML]{000000}{2669} & \textcolor[HTML]{000000}{4555} & \textcolor[HTML]{000000}{6224} & \textcolor[HTML]{000000}{5937} & \textcolor[HTML]{000000}{1923}\\
-\bottomrule
-\end{tabu}}
-\end{table}
-
-\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
-There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions.
-The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
-First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
-When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE.
-As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies.
-The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
-Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
-Without distortion correction, the density estimates are different not only in scaling, but also show inconsistency in the rank of density estimates.
-
-\begin{table}
-
-\caption{\label{tab:electricityhdrpercent}Proportion of the same outliers as UMAP in two highest density regions when estimating densities for four manifold learning embeddings.}
-\centering
-\begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{ISOMAP} & \multicolumn{2}{c}{LLE} & \multicolumn{2}{c}{Laplacian Eigenmaps} & \multicolumn{2}{c}{t-SNE} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
-  & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE & DC-KDE & KDE\\
-\midrule
->99\% HDR & \cellcolor[HTML]{E6E6E6}{0.189} & \textbf{0.243} & \cellcolor[HTML]{E6E6E6}{\textbf{0.189}} & 0.027 & \cellcolor[HTML]{E6E6E6}{\textbf{0.622}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.865}} & 0.108\\
-99\% HDR & \cellcolor[HTML]{E6E6E6}{\textbf{0.359}} & 0.248 & \cellcolor[HTML]{E6E6E6}{\textbf{0.503}} & 0.000 & \cellcolor[HTML]{E6E6E6}{\textbf{0.697}} & 0.007 & \cellcolor[HTML]{E6E6E6}{\textbf{0.628}} & 0.083\\
-\bottomrule
-\end{tabular}
-\end{table}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_rankplot_3639id_umapvs4ml_radius20_r180}
-
-}
-
-\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityrank}
-\end{figure}
-
-Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households.
-To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
-The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}.
-\autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) where anomalies fall outside the \(>99\%\) and \(99\%\) HDRs respectively that are also identified as anomalies by different dimension reduction methods.
-Both DC-KDE and KDE are shown in the table.
-Apart from the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP.
-\autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
-The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}.
-Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.
-The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}.
-The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
-It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE hand UMAP do not identify any of the same anomalies as ISOMAP and LLE.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_compare4id_1typical_3anomalous}
-
-}
-
-\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityraw}
-\end{figure}
-
-We now further examine households 3161, 4546, and 7049, which appear as anomalies for at least three dimension reduction algorithms as well as Households 2396, which is the most typical household with the highest density estimate.
-Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
-Compared with ID 2396, the other three households show different usage patterns.
-ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh.
-As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household.
-The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as temporary accommodation (e.g.~through a holiday rental service such as Airbnb).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous}
-
-}
-
-\caption{Quantile region plots of electricity demand against the time of the week for one typical household 1472 and two anomalies, 7049 and 5136.
-The day-of-the-week patterns are much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time-related usage patterns.}\label{fig:electricityhdrsvkde}
-\end{figure}
-
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityraw}.
-The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also shown on some days of the week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9 pm to 6 am, while the distributions for ID 7049 do not have an obvious pattern across different periods of the week.
-These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
-
-\hypertarget{conclusion}{%
-\section{Conclusions}\label{conclusion}}
-
-In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities.
-Compared with KDE, our distortion-corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
-The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
-We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension reduction algorithm.
-
-As an empirical example, we explore the distributions of \(3,639\) households and \(336\) time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
-Without the ground truth density, we assume that UMAP gives the true densities and compare both density estimation methods by looking at the distributions of the most typical households with the highest densities and three anomalous households that are identified most times.
-The distortion correction methods could identify the typical households with repeated time-of-the-week usage patterns, while the distributions of the outliers are anomalous in different ways.
-
-There are several open questions to be explored.
-The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
-For densities on unknown manifolds, \textcite{Berenfeld2021-as} also provide some insights on the assumptions of the kernel function in \eqref{eq:kernelcondition}.
-The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds \autocite{Berry2017-rz} in the future.
-The last question is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
-The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
-However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
-
-\hypertarget{acknowledgment}{%
-\section*{Acknowledgment}\label{acknowledgment}}
-\addcontentsline{toc}{section}{Acknowledgment}
-
-This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
-The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at Monash University.
-
-\appendix
-
-\hypertarget{riemgeo}{%
-\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}}
-
-In this appendix, we present some notions about the Riemannian geometry used in this paper.
-
-\hypertarget{differentiable-manifolds}{%
-\subsection{Differentiable manifolds}\label{differentiable-manifolds}}
-
-In topology, a \emph{homeomorphism} is a bijective map between two topological spaces that is continuous in both directions.
-A \emph{Hausdorff space} is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-And a \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
-Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x^1(p),\dots, x^d(p)) := \varphi(p)\).
-Further, a collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
-
-The manifold \(M\) is a \emph{differentiable manifold} if there exists an atlas of \(M\), \(\{U_\alpha, \varphi_\alpha\}\), such that the \emph{transition maps} between any two charts,
-\[
-\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-\] are differentiable of class \(C^\infty\) (smooth).
-Let \(\varphi\) be an injective map: \(E \rightarrow \varphi(E)\). Then \(\varphi\) is an \emph{embedding} of \(E\) into \(M\) if and only if
-\(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
-
-\hypertarget{tangent-vector-and-tangent-space}{%
-\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}}
-
-The tangent vector at point \(p\) can be intuitively viewed as the velocity of a curve passing through point \(p\) or as the directional derivatives at \(p\).
-Here we define the tangent vector via the velocity of curves.
-
-For any point \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curves passing through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
-We say \(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart \((U,\varphi)\) at \(p\) such that
-\[
-(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-\]
-A \emph{tangent vector} to a manifold \(M\) at point \(p\), denoted as \(v_p\), is any equivalent class of the differentiable curves initialized at \(p\).
-The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\).
-The tangent space is a vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the chart \(\varphi\) locally at \(p\).
-The collection of all tangent spaces defines the \emph{tangent bundle}, \(TM = \cup_{p \in M}T_pM\).
-
-Tangent vectors can also be seen as the directional derivatives at \(p\). For a given coordinate chart \(\varphi=(x^1,\dots,x^d)\), the tangent vectors defining partial
-derivatives are denoted as \(\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)\), which defines a \emph{basis} of the tangent space.
-The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with the corresponding \emph{cotangent vectors} \(z_p: T^\star_pM \rightarrow \mathbb{R}^d\) and a basis denoted as \(dx^1(p),\dots,dx^d(p)\).
-
-\hypertarget{riemannian-metric-and-geodesic-distance}{%
-\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}}
-
-A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is a \(d\times d\) symmetric positive definite matrix and varies smoothly at \(p\).
-Generally, we
-omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
-The inner product between two vectors \(u, v \in T_pM\) is written as \(\langle u, v \rangle_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j} g_{ij}u^iv^j\), is assumed.
-A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
-
-The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{\langle u,v \rangle_g}\), and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
-It could also be used to define the line element \(dl^2 = g_{ij}dx^i dx^j\) and the volume element \(dV_g = \sqrt{\det(g)}dx^1 \dots dx^d\), where \((x^1,\dots,x^d)\) are the local coordinates of the chart \((U, \varphi)\).
-For a curve \(\gamma: I \rightarrow M\), the length of the curve is
-\[
-l(\gamma) = \sqrt{\int_0^1 \|\gamma^\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
-\]
-where \(\gamma(I) \subset U\). The volume of \(W \subset U\) is defined as
-\[
-Vol(W) = \int_W \sqrt{\det(g)}dx^1 \cdots dx^d,
-\]
-which is also called the \emph{Riemannian measure} on \(M\).
-
-The \emph{geodesics} of \(M\) are the smooth curves that locally join the points along the shortest path on the manifold.
-Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold \autocite[Section 7.2.3 of][]{Nakahara2018-zs}.
-A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
-\[
-\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
-\]
-where \(\{x^i\}\) are the coordinates of the curve \(\gamma\) and \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
-\[
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k} 
-+ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
-\]
-The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc
-length functional \(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\) when the curves are defined over the interval \([0,1]\).
-The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
-For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the distance to the cut locus is the \emph{injectivity radius} at \(p \in M\).
-Therefore, the injectivity radius of the Riemannian manifold \((M,g)\), \(\textit{inj}_gM\), is the infimum of the injectivity radii over all points on the manifold.
-
-\hypertarget{exponential-map-and-logarithmic-map}{%
-\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}}
-
-Denote \(B(p, r) \subset M\) as an open ball centered at point \(p\) with radius \(r\).
-Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\), where \(exp_p\) is the \emph{exponential map} at point \(p\).
-The exponential map maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{\prime}(0)=u\), and \(\gamma(1)=exp_p(u)\).
-It is a differentiable bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}).
-Intuitively, the exponential map moves point \(p\) to an endpoint at speed \(u\) after covering the length of \(\|u\|\) along the geodesic in one time unit.
-
-The inverse of the exponential map is called the \emph{logarithm map}, denoted as \(\log_p(q):= \exp^{-1}_p(q)\), which gives the tangent vector to get from point \(p\) to \(q\) in one unit time.
-Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
-Then we could interpolate a geodesic \(\gamma\) between two points \(p\) and \(q\) with the exponential map and the logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).
-
-\hypertarget{pushforward-and-pullback-metric}{%
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let \(\phi: M \rightarrow E\) be a smooth map between the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
-Then the differential of \(\phi\) at point \(p\) is a linear map \(d\phi_p: T_pM \rightarrow T_{\phi (p)}E\), which pushes the tangent vector \(u \in T_pM\) at point \(p\) forward to the tangent vector \(\phi_*u \in T_{\phi (p)}E\) at the mapping point \(\phi(p)\).
-The image of the tangent vector \(u \in T_pM\) under the differential \(d\phi_p\), denoted as \(d\phi_p u\) is called the pushforward of \(u\) by the map \(\phi\).
-Then pushforward metric \(h=\varphi_*g\) of the Riemannian metric \(g\) along \(\varphi\) is given by the inner product
-\[
-\langle \phi_*u,\phi_*v \rangle_{\varphi*g} = \langle d\phi_p \phi_*u, d\phi_p \phi_*v \rangle_{g}.
-\]
-
-The tangent vectors \(\phi_*u\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{\prime}(0)=u\),
-\[
-d\phi_p(\gamma^{\prime}(0)) = (\phi \circ \gamma)^\prime (0).
-\]
-Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
-The linear map is called the pullback by \(\phi\) and is often denoted as \(\phi^*\).
-
-\hypertarget{twinpeaksappe}{%
-\section{Appendix: Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
-
-This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
-By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest-density regions more accurately than KDE.
-By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest-density regions, which are usually used to detect anomalies.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_lle_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of LLE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tplleden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_le_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of Laplacian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpleden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_umap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of UMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpumapden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-1}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-2}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-3}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-4}
-\end{figure}
-
-\printbibliography
-
-\end{document}
+```

### 2023-06-01 jcgs/kderm_jcgs.tex
Add jcgs version for kderm, the latest version

--- 2023-06-01 monash/dckde.tex
@@
-\documentclass[11pt,a4paper,]{article}
-\usepackage{lmodern}
-
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\hypersetup{
-            pdftitle={Distortion corrected kernel density estimate on Riemannian manifolds},
-            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{geometry}
-\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
-\usepackage[style=authoryear-comp,]{biblatex}
-\addbibresource{references.bib}
-\usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
+% !TeX program = pdfLaTeX
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx,psfrag,epsf}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{textcomp}
+\usepackage[hyphens]{url} % not crucial - just used below for the URL
+\usepackage{hyperref}
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\title{Distortion corrected kernel density estimate on Riemannian manifolds}
-
-%% MONASH STUFF
-
-%% CAPTIONS
-\RequirePackage{caption}
-\DeclareCaptionStyle{italic}[justification=centering]
- {labelfont={bf},textfont={it},labelsep=colon}
-\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
-\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
-
-%% FONT
-\RequirePackage{bera}
-\RequirePackage{mathpazo}
-
-%% HEADERS AND FOOTERS
-\RequirePackage{fancyhdr}
-\pagestyle{fancy}
-\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
-\makeatletter
-\lhead{\textsf{\expandafter{\@title}}}
-\makeatother
-\rhead{}
-\cfoot{}
-\setlength{\headheight}{15pt}
-\renewcommand{\headrulewidth}{0.4pt}
-\renewcommand{\footrulewidth}{0.4pt}
-\fancypagestyle{plain}{%
-\fancyhf{} % clear all header and footer fields
-\fancyfoot[C]{\sffamily\thepage} % except the center
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrulewidth}{0pt}}
-
-%% MATHS
-\RequirePackage{bm,amsmath}
-\allowdisplaybreaks
-
-%% GRAPHICS
-\RequirePackage{graphicx}
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.85}
-\renewcommand{\bottomfraction}{0.85}
-\renewcommand{\textfraction}{0.15}
-\renewcommand{\floatpagefraction}{0.8}
-
-%\RequirePackage[section]{placeins}
-
-%% SECTION TITLES
-\RequirePackage[compact,sf,bf]{titlesec}
-\titleformat{\section}[block]
-  {\fontsize{15}{17}\bfseries\sffamily}
-  {\thesection}
-  {0.4em}{}
-\titleformat{\subsection}[block]
-  {\fontsize{12}{14}\bfseries\sffamily}
-  {\thesubsection}
-  {0.4em}{}
-\titlespacing{\section}{0pt}{*5}{*1}
-\titlespacing{\subsection}{0pt}{*2}{*0.2}
-
-%% TITLE PAGE
-\def\Date{\number\day}
-\def\Month{\ifcase\month\or
- January\or February\or March\or April\or May\or June\or
- July\or August\or September\or October\or November\or December\fi}
-\def\Year{\number\year}
-
-\makeatletter
-\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
-\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
-\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
-\def\nojel{\def\showjel{}}
-\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
-\def\cover{{\sffamily\setcounter{page}{0}
-        \thispagestyle{empty}
-        \placefig{2}{1.5}{width=5cm}{monash2}
-        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
-        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
-        \begin{textblock}{7}(12.7,27.9)\hfill
-        \includegraphics[height=0.7cm]{AACSB}~~~
-        \includegraphics[height=0.7cm]{EQUIS}~~~
-        \includegraphics[height=0.7cm]{AMBA}
-        \end{textblock}
-        \vspace*{2cm}
-        \begin{center}\Large
-        Department of Econometrics and Business Statistics\\[.5cm]
-        \footnotesize http://monash.edu/business/ebs/research/publications
-        \end{center}\vspace{2cm}
-        \begin{center}
-        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
-                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
-                \LARGE\@author\end{onehalfspace}
-        }}
-        \end{center}
-        \vfill
-                \begin{center}\Large
-                \Month~\Year\\[1cm]
-                Working Paper \@wp
-        \end{center}\vspace*{2cm}}}
-\def\pageone{{\sffamily\setstretch{1}%
-        \thispagestyle{empty}%
-        \vbox to \textheight{%
-        \raggedright\baselineskip=1.2cm
-     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
-        \vspace{2cm}\par
-        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
-        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
-        \hspace{1cm}\showjel\vss}}}
-\def\blindtitle{{\sffamily
-     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
-     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
-        }}
-\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}
-
-\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
-\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
-\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
-\let\maketitle\titlepage
-\makeatother
-
-%% SPACING
-\RequirePackage{setspace}
-\spacing{1.5}
-
-%% LINE AND PAGE BREAKING
-\sloppy
-\clubpenalty = 10000
-\widowpenalty = 10000
-\brokenpenalty = 10000
-\RequirePackage{microtype}
-
-%% PARAGRAPH BREAKS
-\setlength{\parskip}{1.4ex}
-\setlength{\parindent}{0em}
-
-%% HYPERLINKS
-\RequirePackage{xcolor} % Needed for links
-\definecolor{darkblue}{rgb}{0,0,.6}
-\RequirePackage{url}
-
-\makeatletter
-\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
-\makeatother
-\hypersetup{
-     citecolor=0 0 0,
-     breaklinks=true,
-     bookmarksopen=true,
-     bookmarksnumbered=true,
-     linkcolor=darkblue,
-     urlcolor=blue,
-     citecolor=darkblue,
-     colorlinks=true}
-
-%% KEYWORDS
-\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}
-
-%% ABSTRACT
-\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
-\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
-  {\end{minipage}}
-
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-
-\usepackage[showonlyrefs]{mathtools}
-\usepackage[no-weekday]{eukdate}
-
-%% BIBLIOGRAPHY
-
-\makeatletter
-\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
-\makeatother
-\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}
-
-\DeclareFieldFormat{url}{\texttt{\url{#1}}}
-\DeclareFieldFormat[article]{pages}{#1}
-\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
-\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
-\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
-\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
-\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
-\DeclareFieldFormat[inproceedings]{title}{#1}
-\DeclareFieldFormat{shorthandwidth}{#1}
-% No dot before number of articles
-\usepackage{xpatch}
-\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
-% Remove In: for an article.
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-
-\makeatletter
-\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
-\makeatother
-\renewcommand*{\finalnamedelim}{%
-  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
-  \addspace\&\space}
-
-\nojel
-
-\RequirePackage[absolute,overlay]{textpos}
-\setlength{\TPHorizModule}{1cm}
-\setlength{\TPVertModule}{1cm}
-\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}
-
-\nocover
-
-\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
-\addresses{\textbf{Fan Cheng}\newline
-Monash University
-\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
-\textbf{Anastasios Panagiotelis}\newline
-The University of Sydney
-\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
-\textbf{Rob J Hyndman}\newline
-Monash University
-\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
-}
-
-\date{\sf\Date~\Month~\Year}
-\makeatletter
- \lfoot{\sf Cheng, Panagiotelis, Hyndman: \@date}
-\makeatother
+
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
+
+%% load any required packages here
+
+% Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc} % for calculating minipage widths
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage[cal = txupr]{mathalpha}
+\usepackage{geometry}
 \geometry{margin=2.2cm}
+\usepackage{mathtools}
 \mathtoolsset{showonlyrefs}
+\usepackage{amsfonts,amssymb}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
 \def\sectionautorefname{Section}
@@
 
-% Adjust headwidth in case user has changed geometry in header-includes
-\renewcommand{\headwidth}{\textwidth}
-
 \begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+{
+  \title{\bf Distortion corrected kernel density estimate on Riemannian manifolds}
+
+  \author{
+        Fan Cheng \thanks{\href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}} \\
+    Monash University\\
+     and \\     Rob J Hyndman \\
+    Monash University\\
+     and \\     Anastasios Panagiotelis \\
+    The University of Sydney\\
+      }
   \maketitle
+} \fi
+
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Distortion corrected kernel density estimate on Riemannian manifolds}
+  \end{center}
+  \medskip
+} \fi
+
+\bigskip
 \begin{abstract}
@@
 \end{abstract}
-\begin{keywords}
-manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
-\end{keywords}
+
+\noindent%
+{\it Keywords:} manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
+\vfill
 
 \newpage
+\spacingset{1.45} % DON'T change the spacing!
 
@@
 
-Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
-Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
+Multivariate kernel density estimation (KDE see Parzen (\protect\hyperlink{ref-Parzen1962-gt}{1962}); Scott (\protect\hyperlink{ref-Scott2015-vl}{2015}) and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
+Applications of KDE include finding hot spots of traffic networks in the GIS environment (\protect\hyperlink{ref-Xie2008-eb}{Xie and Yan 2008}; \protect\hyperlink{ref-Okabe2009-nb}{Okabe, Satoh, and Sugihara 2009}), automatic detection in visual surveillance systems (\protect\hyperlink{ref-Elgammal2002-cw}{Elgammal et al.
+2002}), wind power density detection (\protect\hyperlink{ref-Jeon2012-ac}{Jeon and Taylor 2012}) and prime prediction via Twitter messages (\protect\hyperlink{ref-Gerber2014-tq}{Gerber 2014}).
 For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality.
 However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\).
-The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}.
+The problem of KDE on manifolds has been well studied in Hendriks (\protect\hyperlink{ref-Hendriks1990-bt}{1990}), Izenman (\protect\hyperlink{ref-Izenman1991-cj}{1991}), Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}), Henry, Muñoz, and Rodríguez (\protect\hyperlink{ref-Henry2013-da}{2013}), and Berenfeld, Rosa, and Rousseau (\protect\hyperlink{ref-Berenfeld2022-fd}{2022}).
 However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
@@
 
-When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
+When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
 These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\).
-Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx} and \textcite{Cheng2021-dh}.
+Details of these algorithms are summarized in Cayton (\protect\hyperlink{ref-Cayton2005-dp}{2005}), Lee and Verleysen (\protect\hyperlink{ref-Lee2007-wq}{2007}), Izenman (\protect\hyperlink{ref-Izenman2012-mx}{2012}), and Cheng, Panagiotelis, and Hyndman (\protect\hyperlink{ref-Cheng2021-dh}{2021}).
 One could easily apply KDE to the output vectors \(\pmb{y}\).
@@
 For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
+Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}) which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
 In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
@@
 In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
 This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
@@
 In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
-Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.
+Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic (see Section 6.6 of \protect\hyperlink{ref-Scott2015-vl}{Scott 2015} for a summary).
 
@@
 In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
+Our estimator combines insights from two existing papers, the first is Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}) who propose a KDE for data on manifolds that are known, the second is Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}) who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
 \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
@@
 In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold.
-In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
+Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}), which can be applied when the data lie on some known manifold.
+In \autoref{MetLearn}, we describe the Learn Metric algorithm of Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}), which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
+By combining elements from the work of Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}) and Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}), we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
 To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
@@
 We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\).
-Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
-\begin{align}
+Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}), we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
+\begin{equation}
+\begin{aligned}
 \label{eq:kernelcondition}
@@
 However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
-For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein.
+For more on kernel density estimation in the Euclidean case, see Scott (\protect\hyperlink{ref-Scott2015-vl}{2015}) and references therein.
 Kernel density estimators of this form can and have been applied directly on the output embedding \(\pmb{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
@@
 
-For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
+For kernel density estimation on a known manifold, Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}) propose the following estimator,
 \begin{equation}
@@
 An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold.
-A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
+A definition of the injectivity radius is given by Chavel (\protect\hyperlink{ref-Chavel2006-mp}{2006}) and also provided in the appendix.
 For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once.
@@
 In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\pmb{H}(\pmb{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\pmb{p}_i\).
-For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.
+For more on the volume density function, see Gallot, Hulin, and Lafontaine (\protect\hyperlink{ref-Gallot2004-rc}{2004}) and Brigant and Puechmorel (\protect\hyperlink{ref-Le_Brigant2019-lj}{2019}).
 
@@
 
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
+To be able to apply the estimator of Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}) to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
 Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
 Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
-To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
+To alleviate this issue, Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}) propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
 These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
 
-While full details are provided in \textcite{Perrault-Joncas2013-pq}, we briefly describe the Learn Metric algorithm here.
+While full details are provided in Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}), we briefly describe the Learn Metric algorithm here.
 There are four main steps in the algorithm.
 First, a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold.
-Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated (\protect\hyperlink{ref-Zhou2011-za}{Zhou and Belkin 2011}), where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
 Third, a dimension reduction method is applied to obtain the output embedding \(\pmb{y}_1,\dots,\pmb{y}_n\).
@@
 Full details on these four steps are provided in \autoref{alg:learnmetric}.
-This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
+This algorithm is implemented in a Python library \emph{megaman} (\protect\hyperlink{ref-McQueen2016-xz}{McQueen et al.
+2016}) although our own results are based on a re-implementation of the algorithm in \emph{R}.
 Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.
 
-As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
+As pointed out by Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}), dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
 In this case, the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\).
@@
 In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into \(d\)-dimesional Euclidean space.
-In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used (See Chapter 3 of \textcite{Lee2007-wq} for more about intrinsic dimension estimation).
-The \emph{R} library \emph{intRinsic} \autocite{Denti2021-qc} implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
+In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} (\protect\hyperlink{ref-Facco2017-rl}{Facco et al.
+2017}; \protect\hyperlink{ref-Denti2021-jl}{Denti et al.
+2021}) can be used (See Chapter 3 of Lee and Verleysen (\protect\hyperlink{ref-Lee2007-wq}{2007}) for more about intrinsic dimension estimation).
+The \emph{R} library \emph{intRinsic} (\protect\hyperlink{ref-Denti2021-qc}{Denti 2021}) implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
 
@@
 
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
+  \STATE Obtain the $d\times d\times N$ tensor $\pmb{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with elements in $\pmb{\tilde{H}}$ found by
   $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+    \pmb{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^l \odot \pmb{y}^k\right) - \pmb{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^k\right) - \pmb{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^l\right)\right],
   $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
+  where $l,k=1,\dots,d$, $\pmb{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;
+
+  \STATE For each embedding data point $\pmb{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\pmb{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}^{\cdot\cdot i}}$ below 
   $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
+    \pmb{H}_i = \pmb{U} \pmb{\Lambda}_d^{-1} \pmb{U}^\prime,
   $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+  where $[\pmb{U}, \pmb{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\pmb{H}^{\cdot\cdot i}}$, $\pmb{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\pmb{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.
 
@@
 The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
+To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013})).
 This is crucial for two reasons.
@@
 For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
-Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
+Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation (\protect\hyperlink{ref-Breiman1977-qc}{Breiman, Meisel, and Purcell 1977}; \protect\hyperlink{ref-Jones1990-oe}{Jones 1990}; \protect\hyperlink{ref-Terrell1992-ut}{Terrell and Scott 1992}).
 
@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
+Compared to the anomaly detection with a general kernel density estimator in Cheng, Panagiotelis, and Hyndman (\protect\hyperlink{ref-Cheng2021-dh}{2021}), the changes are also highlighted in blue.
 With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 
-\caption{Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
+\caption{Underlying data for the Gaussian mixture model of four kernels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$.
 The colors indicate the true density of the data when they are mapped via the twin peaks function.
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{figures/scatterplot3d_twinpeaks_dc}
+{\centering \includegraphics[width=0.75\linewidth]{figures/scatterplot3d_twinpeaks_dc}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
+{\centering \includegraphics[width=0.85\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
 
@@
 \caption{Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
-The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel).
+The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
 DC-KDE finds more true outliers than KDE in all five rows.}\label{fig:tpoutliers}
@@
 \caption{\label{tab:tpcors}Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data.
-Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.}
+Distortion-corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.}
 \centering
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank}
+{\centering \includegraphics[width=0.8\linewidth]{figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank}
 
@@
 \(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
+\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} (\protect\hyperlink{ref-Wickham2011-ir}{Wickham et al.
+2011}).
 The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec}
+{\centering \includegraphics[width=0.9\linewidth]{figures/sim4d10000_density_comparison_4ml_radius10_k200_rankdensity_circleoutlier_with1rec}
 
@@
 
-\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
+\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data.
+DC-KDE has a slightly lower rank correlation for ISOMAP and LLE above 0.96, but significantly outperfoms KDE for Laplacian Eigenmaps and UMAP when more distortion is induced in dimension reduction.}
 \centering
@@
 
-\caption{\label{tab:fourdhdrtable}Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings.}
+\caption{\label{tab:fourdhdrtable}Proportion comparison of outliers in correct highest density regions in density estimation of four manifold learning embeddings for the 100-D data.}
 \centering
+\resizebox{\linewidth}{!}{
 \begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
@@
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
@@
 
-In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
+In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland (\protect\hyperlink{ref-cer2012-data}{Commission for Energy Regulation (CER) 2012}) between 14 July 2009 and 31 December 2010.
 The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
@@
 
-For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week \autocite{Hyndman2018-ia}.
+For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week (\protect\hyperlink{ref-Hyndman2018-ia}{Hyndman, Liu, and Pinson 2018}).
 This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
-\textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
+Cheng, Panagiotelis, and Hyndman (\protect\hyperlink{ref-Cheng2021-dh}{2021}) propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
 Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection.
@@
 
-Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here.
+Although full details for data processing and dimension reduction are provided in Cheng, Panagiotelis, and Hyndman (\protect\hyperlink{ref-Cheng2021-dh}{2021}), we briefly describe the process here.
 For each household, a discrete approximation of the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_5levels_outliers_comparison_5ml_r180}
+{\centering \includegraphics[width=0.85\linewidth]{figures/Electricity_2d_5levels_outliers_comparison_5ml_r180}
 
@@
 \centering
+\resizebox{\linewidth}{!}{
 \begin{tabular}[t]{l|>{}c>{}c|>{}c>{}c|>{}c>{}c|>{}cc}
@@
 
-\caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityrank}
+\caption{Comparison of ranks of the estimated densities based on UMAP and four dimension reduction algorithms in each row.
+Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows a higher level of robustness to the use of different dimension reduction methods.}\label{fig:electricityrank}
 \end{figure}
@@
 The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
-It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE hand UMAP do not identify any of the same anomalies as ISOMAP and LLE.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_compare4id_1typical_3anomalous}
-
-}
-
-\caption{Electricity usage plots of all 535 days for the most typical household and two anomalies in rows and KDE with distortion correction (left panel) and KDE (right panel) in columns.}\label{fig:electricityraw}
+It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.95\linewidth]{figures/Electricity_2d_compare4id_1typical_3anomalous}
+
+}
+
+\caption{Electricity usage plots of all 535 days for the most typical household (top row) and three anomalies.}\label{fig:electricityraw}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous}
-
-}
-
-\caption{Quantile region plots of electricity demand against the time of the week for one typical household 1472 and two anomalies, 7049 and 5136.
-The day-of-the-week patterns are much more different for Household 7049 compared with 5136, the latter of which does not have the obvious meal time-related usage patterns.}\label{fig:electricityhdrsvkde}
+{\centering \includegraphics[width=0.95\linewidth]{figures/Electricity_2d_plot_hdrbox_7dow_4id_1typical_3anomalous}
+
+}
+
+\caption{Quantile region plots of electricity demand against the time of the week for one typical household 1472 and three anomalies, 3161, 4546, and 7049.
+The day-of-the-week patterns are much more different for Household 4546 and 7049 compared with 3161.}\label{fig:electricityhdrsvkde}
 \end{figure}
@@
 The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
-For densities on unknown manifolds, \textcite{Berenfeld2021-as} also provide some insights on the assumptions of the kernel function in \eqref{eq:kernelcondition}.
+For densities on unknown manifolds, Berenfeld and Hoffmann (\protect\hyperlink{ref-Berenfeld2021-as}{2021}) also provide some insights on the assumptions of the kernel function in \eqref{eq:kernelcondition}.
 The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds \autocite{Berry2017-rz} in the future.
+Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds (\protect\hyperlink{ref-Berry2017-rz}{Berry and Sauer 2017}) in the future.
 The last question is related to the quality of the embedding from dimension reduction methods.
 Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
+The embedding quality could be measured using one of the metrics discussed in the online supplementary material of Cheng, Panagiotelis, and Hyndman (\protect\hyperlink{ref-Cheng2021-dh}{2021}), but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
 The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
@@
 \hypertarget{riemgeo}{%
-\section{Appendix: Notions about Riemannian geometry}\label{riemgeo}}
+\section{Notions about Riemannian geometry}\label{riemgeo}}
 
@@
 The \emph{geodesics} of \(M\) are the smooth curves that locally join the points along the shortest path on the manifold.
-Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold \autocite[Section 7.2.3 of][]{Nakahara2018-zs}.
+Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold (Section 7.2.3 of \protect\hyperlink{ref-Nakahara2018-zs}{Nakahara 2018}).
 A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
@@
 \hypertarget{twinpeaksappe}{%
-\section{Appendix: Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
+\section{Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
 
 This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
-By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest-density regions more accurately than KDE.
-By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest-density regions, which are usually used to detect anomalies.
+By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest density regions more accurately than KDE.
+By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest density regions, which are usually used to detect anomalies.
 
@@
 
-\caption{Scatterplot of true density and estimated density ranks of Laplacian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
+\caption{Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
 DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpleden}
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
+{\centering \includegraphics[width=0.9\linewidth]{figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank}
+
+}
+
+\caption{Comparison of outliers found by LLE compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
 The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-1}
+Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml}
 \end{figure}
@@
 
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-2}
+\caption{Comparison of outliers found by Laplacian Eigemaps compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tplevs4ml}
 \end{figure}
@@
 
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-3}
+\caption{Comparison of outliers found by t-SNE compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tptsnevs4ml}
 \end{figure}
@@
 
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-4}
-\end{figure}
-
-\printbibliography
+\caption{Comparison of outliers found by UMAP compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tpumapvs4ml}
+\end{figure}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Berenfeld2021-as}{}}%
+Berenfeld, Clément, and Marc Hoffmann.
+2021.
+{``Density Estimation on an Unknown Submanifold.''} \emph{European Journal of Sport Science: EJSS: Official Journal of the European College of Sport Science} 15 (1): 2179--2223.
+\url{https://doi.org/10.1214/21-EJS1826}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Berenfeld2022-fd}{}}%
+Berenfeld, Clément, Paul Rosa, and Judith Rousseau. 2022. {``Estimating a Density Near an Unknown Manifold: A Bayesian Nonparametric Approach,''} May. \url{http://arxiv.org/abs/2205.15717}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Berry2017-rz}{}}%
+Berry, Tyrus, and Timothy Sauer.
+2017.
+{``Density Estimation on Manifolds with Boundary.''} \emph{Computational Statistics \& Data Analysis} 107 (March): 1--17.
+\url{https://doi.org/10.1016/j.csda.2016.09.011}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Breiman1977-qc}{}}%
+Breiman, Leo, William Meisel, and Edward Purcell.
+1977.
+{``Variable Kernel Estimates of Multivariate Densities.''} \emph{Technometrics: A Journal of Statistics for the Physical, Chemical, and Engineering Sciences} 19 (2): 135--44.
+\url{https://doi.org/10.1080/00401706.1977.10489521}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Le_Brigant2019-lj}{}}%
+Brigant, Alice le, and Stéphane Puechmorel. 2019. {``Approximation of Densities on Riemannian Manifolds.''} \emph{Entropy} 21 (1). \url{https://doi.org/10.3390/e21010043}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Cayton2005-dp}{}}%
+Cayton, Lawrence. 2005. {``Algorithms for Manifold Learning.''} \emph{Univ. Of California at San Diego Tech. Rep} 12 (1-17): 1.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chavel2006-mp}{}}%
+Chavel, Isaac. 2006. \emph{Riemannian Geometry: A Modern Introduction}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Cheng2021-dh}{}}%
+Cheng, Fan, Anastasios Panagiotelis, and Rob J Hyndman.
+2021.
+{``Computationally Efficient Learning of Statistical Manifolds.''} \emph{arXiv e-Prints}, February, arXiv:2103.11773.
+\url{https://ui.adsabs.harvard.edu/abs/2021arXiv210311773C}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cer2012-data}{}}%
+Commission for Energy Regulation (CER). 2012. {``CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010 {[}Dataset{]}.''} \emph{Irish Social Science Data Archive}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Denti2021-qc}{}}%
+Denti, Francesco. 2021. {``intRinsic: An {R} Package for Model-Based Estimation of the Intrinsic Dimension of a Dataset,''} February. \url{https://arxiv.org/abs/2102.11425}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Denti2021-jl}{}}%
+Denti, Francesco, Diego Doimo, Alessandro Laio, and Antonietta Mira. 2021. {``Distributional Results for {Model-Based} Intrinsic Dimension Estimators,''} April. \url{https://arxiv.org/abs/2104.13832}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Elgammal2002-cw}{}}%
+Elgammal, A, R Duraiswami, D Harwood, and L S Davis.
+2002.
+{``Background and Foreground Modeling Using Nonparametric Kernel Density Estimation for Visual Surveillance.''} \emph{Proceedings of the IEEE} 90 (7): 1151--63.
+\url{https://doi.org/10.1109/JPROC.2002.801448}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Facco2017-rl}{}}%
+Facco, Elena, Maria d'Errico, Alex Rodriguez, and Alessandro Laio.
+2017.
+{``Estimating the Intrinsic Dimension of Datasets by a Minimal Neighborhood Information.''} \emph{Scientific Reports} 7 (1): 12140.
+\url{https://doi.org/10.1038/s41598-017-11873-y}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gallot2004-rc}{}}%
+Gallot, Sylvestre, Dominique Hulin, and Jacques Lafontaine.
+2004.
+{``Riemannian Metrics.''} In \emph{Riemannian Geometry}, 51--127.
+Berlin, Heidelberg: Springer Berlin Heidelberg.
+\url{https://doi.org/10.1007/978-3-642-18855-8/_2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gerber2014-tq}{}}%
+Gerber, Matthew S. 2014. {``Predicting Crime Using Twitter and Kernel Density Estimation.''} \emph{Decision Support Systems} 61 (May): 115--25. \url{https://doi.org/10.1016/j.dss.2014.02.003}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hendriks1990-bt}{}}%
+Hendriks, Harrie.
+1990.
+{``Nonparametric Estimation of a Probability Density on a Riemannian Manifold Using Fourier Expansions.''} \emph{Annals of Statistics} 18 (2): 832--49.
+\url{http://www.jstor.org/stable/2242136}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Henry2013-da}{}}%
+Henry, Guillermo, Andrés Muñoz, and Daniela Rodríguez.
+2013.
+{``Locally Adaptive Density Estimation on Riemannian Manifolds.''} \emph{SORT-Statistics and Operations Research Transactions} 37 (2): 111--30.
+\url{https://raco.cat/index.php/SORT/article/view/271165}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hyndman2018-ia}{}}%
+Hyndman, R J, X Liu, and P Pinson. 2018. {``Visualizing Big Energy Data: Solutions for This Crucial Component of Data Analysis.''} \emph{IEEE Power Energ. Mag.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Izenman1991-cj}{}}%
+Izenman, Alan Julian.
+1991.
+{``Review Papers: Recent Developments in Nonparametric Density Estimation.''} \emph{Journal of the American Statistical Association} 86 (413): 205--24.
+\url{https://doi.org/10.1080/01621459.1991.10475021}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Izenman2012-mx}{}}%
+---------. 2012. {``Introduction to Manifold Learning.''} \emph{WIREs Comp Stat} 4 (5): 439--46.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Jeon2012-ac}{}}%
+Jeon, Jooyoung, and James W Taylor.
+2012.
+{``Using Conditional Kernel Density Estimation for Wind Power Density Forecasting.''} \emph{Journal of the American Statistical Association} 107 (497): 66--79.
+\url{https://doi.org/10.1080/01621459.2011.643745}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Jones1990-oe}{}}%
+Jones, M C. 1990. {``Variable Kernel Density Estimates and Variable Kernel Density Estimates.''} \emph{The Australian Journal of Statistics} 32 (3): 361--71.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lee2007-wq}{}}%
+Lee, John A, and Michel Verleysen. 2007. \emph{Nonlinear Dimensionality Reduction}. Springer Science \& Business Media.
+
+\leavevmode\vadjust pre{\hypertarget{ref-McQueen2016-xz}{}}%
+McQueen, James, Marina Meilă, Jacob VanderPlas, and Zhongyue Zhang. 2016. {``Megaman: Scalable Manifold Learning in Python.''} \emph{J. Mach. Learn. Res.} 17 (148): 1--5.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Nakahara2018-zs}{}}%
+Nakahara, M. 2018. \emph{{Geometry, topology and physics}}. {CRC press}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Okabe2009-nb}{}}%
+Okabe, Atsuyuki, Toshiaki Satoh, and Kokichi Sugihara.
+2009.
+{``A Kernel Density Estimation Method for Networks, Its Computational Method and a {GIS‐based} Tool.''} \emph{Geographical Information Systems} 23 (1): 7--32.
+\url{https://doi.org/10.1080/13658810802475491}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Parzen1962-gt}{}}%
+Parzen, Emanuel. 1962. {``On Estimation of a Probability Density Function and Mode.''} \emph{Annals of Mathematical Statistics} 33 (3): 1065--76.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Pelletier2005-vu}{}}%
+Pelletier, Bruno. 2005. {``Kernel Density Estimation on Riemannian Manifolds.''} \emph{Statistics \& Probability Letters} 73 (3): 297--304.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Perrault-Joncas2013-pq}{}}%
+Perrault-Joncas, Dominique, and Marina Meila.
+2013.
+{``Non-Linear Dimensionality Reduction: Riemannian Metric Estimation and the Problem of Geometric Discovery,''} May.
+\url{https://arxiv.org/abs/1305.7255}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Scott2015-vl}{}}%
+Scott, David W. 2015. \emph{{Multivariate Density Estimation: Theory, Practice, and Visualization}}. John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Terrell1992-ut}{}}%
+Terrell, George R, and David W Scott. 1992. {``Variable Kernel Density Estimation.''} \emph{Annals of Statistics} 20 (3): 1236--65.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Wickham2011-ir}{}}%
+Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja.
+2011.
+{``Tourr: An {R} Package for Exploring Multivariate Data with Projections.''} \emph{Journal of Statistical Software} 40 (April): 1--18.
+\url{https://doi.org/10.18637/jss.v040.i02}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Xie2008-eb}{}}%
+Xie, Zhixiao, and Jun Yan.
+2008.
+{``Kernel Density Estimation of Traffic Accidents in a Network Space.''} \emph{Computers, Environment and Urban Systems} 32 (5): 396--406.
+\url{https://doi.org/10.1016/j.compenvurbsys.2008.05.001}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Zhou2011-za}{}}%
+Zhou, Xueyuan, and Mikhail Belkin.
+2011.
+{``Semi-Supervised Learning by Higher Order Regularization.''} In \emph{Proceedings of the Fourteenth International Conference on Artificial Intelligence and Statistics}, 15:892--900.
+Proceedings of Machine Learning Research.
+JMLR Workshop; Conference Proceedings.
+
+\end{CSLReferences}
+
+\bibliographystyle{agsm}
+\bibliography{kderm.bib}
 

### 2023-06-05 jcgs/03-chap3-kderm.Rmd
Formatting and citations

@@
 When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$.
+These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\bm{p}$ to $d$-dimensional vectors $\bm{y}$.
 Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh.
-One could easily apply KDE to the output vectors $\pmb{y}$.
+One could easily apply KDE to the output vectors $\bm{y}$.
 However, this would be a misleading representation of the density of points as they lie on the manifold.
 To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,$$
-where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$.
-By a change of variables, this would imply the following density for $\pmb{y}$,
+$$Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,$$
+where $f$ is the probability density function on the manifold, $\bm{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$.
+By a change of variables, this would imply the following density for $\bm{y}$,
 \begin{equation}
 \label{eq:changevar}
-Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
+Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
 \end{equation}
-where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
-It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant.
-Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
-Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
+where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\bm{H}(\bm{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
+It can be seen that the density in terms of $\bm{y}$, differs from the density on the manifold due to the term $|\bm{H}(\bm{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant.
+Since $\psi$ will typically be a non-linear function, the value of $|\bm{H}(\bm{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
+Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\bm{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
 
@@
 For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+Critical to our estimator, is obtaining an estimate of $|\bm{H}(\bm{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
 One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
@@
 
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
+In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\bm{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space $\bm{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
 This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
@@
 In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample.
-Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case.
-For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$.
-We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$.
+Points on this manifold are denoted $\bm{p}$ in general, with $\bm{p}_1,\dots,\bm{p}_n$ denoting the observed sample.
+Often $\bm{p}_i$ will be high-dimensional vectors such that $\bm{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case.
+For instance, $\bm{p}_i$ may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at each $\bm{p}_i$ only require some sense of distance between the 'input' points, $d(\bm{p}_i,\bm{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\bm{y}_1,\dots,\bm{y}_n$, where $\bm{y}_i\in\mathbb{R}^d$.
+We will denote this embedding as $\bm{y}_i=\psi(\bm{p}_i)$.
 Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
@@
 \label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
-(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
-(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
+& (i) \int_{\mathbb{R}^{d}} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=1;
+(ii) \int_{\mathbb{R}^{d}} \bm{y} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=0;
+(iii) \int_{\mathbb{R}^{d}}\|\bm{y}\|^{2} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})<\infty; \\
 & (iv) \supp K=[0; 1];
-(v) \sup K(\|\pmb{y}\|)=K(0).
+(v) \sup K(\|\bm{y}\|)=K(0).
 \end{aligned}
@@
 
-For data $\pmb{y}_i\in\mathbb{R}^d$ with $i=1,\dots,N$ and assuming a bandwidth matrix $r\pmb{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\pmb{y}$ is given by
+For data $\bm{y}_i\in\mathbb{R}^d$ with $i=1,\dots,N$ and assuming a bandwidth matrix $r\bm{I}$ where $r$ is a global bandwidth, then the usual kernel density estimator at a point $\bm{y}$ is given by
 \begin{equation}
 \label{eq:vkde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
+\hat{f}(\bm{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\bm{y}-\bm{y}_i\|}{r}\right).
 \end{equation}
@@
 The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point $\pmb{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\pmb{y}$, times a term that ensures the density integrates to 1.
+The density at a point $\bm{y}$ is equal to the proportion of sample points that lie within a ball of radius $r$ centered at $\bm{y}$, times a term that ensures the density integrates to 1.
 In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius $r$ centered at $\pmb{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
+However, the intuition remains the same, only that the ball of radius $r$ centered at $\bm{y}$ is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
 For more on kernel density estimation in the Euclidean case, see @Scott2015-vl and references therein.
-Kernel density estimators of this form can and have been applied directly on the output embedding $\pmb{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.
+Kernel density estimators of this form can and have been applied directly on the output embedding $\bm{y}$, and we will consider this approach as a benchmark in \autoref{simulation}.
 
@@
 To make this clear consider the simpler case of computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term.
-A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density of the manifold itself.
+A similar notion applies to a manifold embedding so that the density of the output vectors $\bm{y}_i$ differs from the density of the manifold itself.
 Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density on its own.
@@
 \label{eq:denriem}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
+\hat{f}(\bm{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\bm{p}_i}(\bm{p})} K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right),
 \end{equation}
 
-where $d_g(\pmb{p}, \pmb{p}_i)$ denotes the geodesic distance between two points on the manifold $\pmb{p}$ and $\pmb{p}_i$ and $\theta_{\pmb{p}_i}(\pmb{p})$ is known as the volume density function.
-The intuition behind the term $K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)$ is relatively clear.
-For example, for a uniform kernel, the estimator at point $\pmb{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\pmb{p}$.
+where $d_g(\bm{p}, \bm{p}_i)$ denotes the geodesic distance between two points on the manifold $\bm{p}$ and $\bm{p}_i$ and $\theta_{\bm{p}_i}(\bm{p})$ is known as the volume density function.
+The intuition behind the term $K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)$ is relatively clear.
+For example, for a uniform kernel, the estimator at point $\bm{p}$ will still depend on the proportion of sample points within a ball of radius $r$ centered at $\bm{p}$.
 However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
@@
 A definition of the injectivity radius is given by @Chavel2006-mp and also provided in the appendix.
-For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\pmb{p}$ is so large that some points `fall inside' the ball more than once.
+For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around $\bm{p}$ is so large that some points `fall inside' the ball more than once.
 For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
@@
 
-The inclusion of the volume density function  $\theta_{\pmb{p}_i}(\pmb{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
-We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\pmb{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\pmb{p}$.
+The inclusion of the volume density function  $\theta_{\bm{p}_i}(\bm{p})$ is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
+We have already highlighted that when using a uniform kernel, the kernel density estimate at a point $\bm{p}$ directly depends on the proportion of sample points within a ball of radius $r$ around $\bm{p}$.
 However, the volume of this ball must also be taken into account.
@@
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around $\pmb{p}$, given by $exp_{\pmb{p}}(\pmb{q})$, mapping vectors in the tangent space,  $\pmb{v}\in T_{\pmb{p}}M$, to points on the manifold, $\pmb{q}\in M$.
-Loosely, $\pmb{v}$ `points' in the direction of the geodesic between $\pmb{p}$ and $\pmb{q}$ and traveling along this geodesic at uniform speed $\|\pmb{v}\|$ takes place in one unit of time.
-Now, consider a chart $\varphi$ mapping points in the neighborhood of $\pmb{p}$, via the inverse of the exponential map, to these $\pmb{v}$ vectors, expressed in some local coordinate system.
+Consider the exponential map around $\bm{p}$, given by $exp_{\bm{p}}(\bm{q})$, mapping vectors in the tangent space,  $\bm{v}\in T_{\bm{p}}M$, to points on the manifold, $\bm{q}\in M$.
+Loosely, $\bm{v}$ `points' in the direction of the geodesic between $\bm{p}$ and $\bm{q}$ and traveling along this geodesic at uniform speed $\|\bm{v}\|$ takes place in one unit of time.
+Now, consider a chart $\varphi$ mapping points in the neighborhood of $\bm{p}$, via the inverse of the exponential map, to these $\bm{v}$ vectors, expressed in some local coordinate system.
 The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-In proving that the estimator integrates to 1, the volume density function cancels with the term $|\pmb{H}(\pmb{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\pmb{p}_i$.
+In proving that the estimator integrates to 1, the volume density function cancels with the term $|\bm{H}(\bm{y})|^{1/2}$ in Equation \@ref(eq:changevar), where $\psi$ is the logarithmic map around $\bm{p}_i$.
 For more on the volume density function, see @Gallot2004-rc and @Le_Brigant2019-lj.
@@
 
-To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\pmb{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space $T_{\pmb{p}} M$ for every point $\pmb{p} \in M$.
-<!--The inner product between two tangent vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
+To be able to apply the estimator of @Pelletier2005-vu to the case where the manifold is not known, but where coordinates $\bm{y}_i$ for $i=1,\dots,n$ are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
+Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space $T_{\bm{p}} M$ for every point $\bm{p} \in M$.
+<!--The inner product between two tangent vectors $u,v \in T_{\bm{p}}M$, given by $\langle u,v \rangle_g$, can be used to define geometric quantities.
 For example, angles on a manifold are given by $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$, while distances and volumes on manifolds are also defined with reference to the Riemannian metric.
 While the defined tangent vectors, the Riemannian metric and the geometric quantities are invariant to any specific choice of coordinates, they can still be expressed in terms of local coordinate systems.
-This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\pmb{y}$, angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
-To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n$ with $d\times d$ positive definite matrices, $\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n$, at each data point.
+This is precisely the situation when data on a manifold are mapped to $d$-dimensional Euclidean vectors $\bm{y}_1,\bm{y}_2\dots,\bm{y}_n$ via a dimension reduction algorithm.--> Using the coordinates from a dimension reduction algorithm $\bm{y}$, angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+To alleviate this issue, @Perrault-Joncas2013-pq propose a method to augment $\bm{y}_1,\bm{y}_2\dots,\bm{y}_n$ with $d\times d$ positive definite matrices, $\bm{H}_1,\bm{H}_2\dots,\bm{H}_n$, at each data point.
 These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
-<!--For example, the angle between $\pmb{p}_j$ and $\pmb{p}_k$ at $\pmb{p}_i$ depends (up to a first order approximation) on the inner product $(\pmb{y}_j-\pmb{y}_i)'\pmb{H}^{-1}_i(\pmb{y}_k-\pmb{y}_i)$ rather than the usual Euclidean inner product $(\pmb{y}_j-\pmb{y}_i)'(\pmb{y}_k-\pmb{y}_i)$.
+<!--For example, the angle between $\bm{p}_j$ and $\bm{p}_k$ at $\bm{p}_i$ depends (up to a first order approximation) on the inner product $(\bm{y}_j-\bm{y}_i)'\bm{H}^{-1}_i(\bm{y}_k-\bm{y}_i)$ rather than the usual Euclidean inner product $(\bm{y}_j-\bm{y}_i)'(\bm{y}_k-\bm{y}_i)$.
 -->
@@
 There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between $\pmb{p}_i$ and $\pmb{p}_j$ when $\pmb{p}_i$ is a K-nearest neighbor of $\pmb{p}_j$ or vice versa, and edge weights depending on the distance between $\pmb{p}_i$ and $\pmb{p}_j$ on the manifold.
+First, a weighted neighborhood graph is constructed, with edges between $\bm{p}_i$ and $\bm{p}_j$ when $\bm{p}_i$ is a K-nearest neighbor of $\bm{p}_j$ or vice versa, and edge weights depending on the distance between $\bm{p}_i$ and $\bm{p}_j$ on the manifold.
 Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
-Third, a dimension reduction method is applied to obtain the output embedding $\pmb{y}_1,\dots,\pmb{y}_n$.
+Third, a dimension reduction method is applied to obtain the output embedding $\bm{y}_1,\dots,\bm{y}_n$.
 Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
@@
 As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension $d$.
-In this case, the ranks of the matrices $\pmb{H}_i$ are equal to $d$.
+In this case, the ranks of the matrices $\bm{H}_i$ are equal to $d$.
 Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension $d$ globally into $d$-dimensional Euclidean space.
@@
   \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}_i$ for all $i=1,\ldots,N$ }
+  \Input{ high-dimensional data $\bm{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $\bm{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\bm{H}_i$ for all $i=1,\ldots,N$ }
   \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm EMBED }
@@
 
-  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;
+  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\bm{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\bm{x}_i-\bm{x}_j\|^2)$ for data points $\bm{x}_i,\bm{x}_j \in \mathbb{R}^s$;
 
@@
   $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
+  where $\widetilde{D}=diag{\widetilde{W}\bm{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\bm{1}}$;
 
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
+  \STATE Embed all data point $\bm{X}\in \mathbb{R}^s$ to embedding coordinates $\bm{Y}=(\bm{y}^1,\dots,\bm{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
 
-  \STATE Obtain the $d\times d\times N$ tensor $\pmb{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with elements in $\pmb{\tilde{H}}$ found by
+  \STATE Obtain the $d\times d\times N$ tensor $\bm{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\bm{Y}$ with elements in $\bm{\tilde{H}}$ found by
   $$
-    \pmb{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^l \odot \pmb{y}^k\right) - \pmb{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^k\right) - \pmb{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^l\right)\right],
+    \bm{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\bm{y}^l \odot \bm{y}^k\right) - \bm{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \bm{y}^k\right) - \bm{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \bm{y}^l\right)\right],
   $$
-  where $l,k=1,\dots,d$, $\pmb{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;
+  where $l,k=1,\dots,d$, $\bm{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;
 
-  \STATE For each embedding data point $\pmb{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\pmb{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}^{\cdot\cdot i}}$ below 
+  \STATE For each embedding data point $\bm{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\bm{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\bm{H}^{\cdot\cdot i}}$ below
   $$
-    \pmb{H}_i = \pmb{U} \pmb{\Lambda}_d^{-1} \pmb{U}^\prime,
+    \bm{H}_i = \bm{U} \bm{\Lambda}_d^{-1} \bm{U}^\prime,
   $$
-  where $[\pmb{U}, \pmb{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\pmb{H}^{\cdot\cdot i}}$, $\pmb{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\pmb{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.
+  where $[\bm{U}, \bm{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\bm{H}^{\cdot\cdot i}}$, $\bm{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\bm{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.
 
@@
 \label{eq:denestimator}
-\hat{f}(\pmb{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}_j|}{|\det \pmb{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|}{r} \bigg).
+\hat{f}(\bm{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \bm{H}_j|}{|\det \bm{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|}{r} \bigg).
 \end{equation}
@@
 The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq).
+To understand these differences, it is first critical to appreciate that the coordinates $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq).
 This is crucial for two reasons.
-First, this implies that the term $\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|$ approximates the geodesic distance between $\pmb{y}_i$ and $\pmb{y}_j$.
-Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\pmb{y}_i$, and it is this mapping that is approximated by $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$.
-For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)$ which ensures that the embedding approximates the logarithmic map.
+First, this implies that the term $\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|$ approximates the geodesic distance between $\bm{y}_i$ and $\bm{y}_j$.
+Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\bm{y}_i$, and it is this mapping that is approximated by $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$.
+For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ which ensures that the embedding approximates the logarithmic map.
 Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].
 
-One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\pmb{H}_j$.
+One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$.
 To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
@@
 The entire workflow is summarized in \autoref{fig:vkde}.
-The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\pmb{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
+The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
 Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue.
@@
 
-The simulation setup for the twin peaks example is to first generate vectors $\pmb{v}_1,\dots,\pmb{v}_N$ for $N=2000$ from a 2-dimensional Gaussian mixture model.
+The simulation setup for the twin peaks example is to first generate vectors $\bm{v}_1,\dots,\bm{v}_N$ for $N=2000$ from a 2-dimensional Gaussian mixture model.
 The mixture has four components with different means
-$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4$.
+$\bm{\mu_1}=(0.25, 0.25)^\prime, \bm{\mu_2}=(0.25, 0.75)^\prime, \bm{\mu_3}=(0.75, 0.25)^\prime, \bm{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\bm{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4$.
 The mixture proportions
@@
 It is important to note that the *true density* on the manifold is not simply a Gaussian mixture, since the mapping in Equation \@ref(eq:twinpeak) distorts the distribution.
-To recover the true distribution requires the correct Jacobian term for the pushforward from $\pmb{v}$ to the volume form of twin peaks manifold.
-By treating the $\pmb{v}$ as an 'output' embedding from input points $\pmb{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\pmb{\Gamma}_i$ for $i=1,\dots,n$ where $\pmb{\Gamma}_i$ is the Riemannian of the coordinate system given by $\pmb{v}$.
-This notation is distinct from $\pmb{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as $f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}$, where $f(\pmb{v}_i)$ is the density of a four-component mixture of normals.
-Knowledge of $\pmb{v}$ and $\pmb{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold.
-\autoref{fig:metadensity} shows the simulated $\pmb{v}$ with color indicating the true density of data on the manifold.
+To recover the true distribution requires the correct Jacobian term for the pushforward from $\bm{v}$ to the volume form of twin peaks manifold.
+By treating the $\bm{v}$ as an 'output' embedding from input points $\bm{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\bm{\Gamma}_i$ for $i=1,\dots,n$ where $\bm{\Gamma}_i$ is the Riemannian of the coordinate system given by $\bm{v}$.
+This notation is distinct from $\bm{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
+The true density on the manifold can be obtained as $f(\bm{p}_i)=f(\bm{v}_i)|\bm{\Gamma}_i|^{1/2}$, where $f(\bm{v}_i)$ is the density of a four-component mixture of normals.
+Knowledge of $\bm{v}$ and $\bm{\Gamma_i}$ will not be used when estimating the density but only to establish a 'ground truth' for densities on the manifold.
+\autoref{fig:metadensity} shows the simulated $\bm{v}$ with color indicating the true density of data on the manifold.
 Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in yellow.
 The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\pmb{v}$.
+The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the $\bm{v}$.
 
-```{r tpoutliers, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
+```{r tpoutliers, fig.pos = "!p", fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
 The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
@@
 
-```{r tpisomapvs4ml, fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
+```{r tpisomapvs4ml, fig.pos = "!p", fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
 Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
@@
 ```
+
+\FloatBarrier
 
@@
 
+As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
+To start with, we simulate vectors $(\bm{v}_1, \dots, \bm{v}_N)^\prime$ for $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)$ and $\mathcal{N}(\bm{\mu}_2, \bm{\Sigma}_2)$, with the same means $\bm{\mu}_1 = \bm{\mu}_2 =(0, 0, 0, 0)^\prime$ and different variance-covariance matrices $\bm{\Sigma}_1 = diag(1,1,1,1)$ and $\bm{\Sigma}_2 = diag(2,2,2,2)$.
+The mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
+With this design, the observations from the second component tend to be outlying anomalies.
+The data are mapped to a hemisphere via the equation
+$v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
+\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path^[See the animation of the 5-D grand tour at https://github.com/ffancheng/kderm/blob/master/monash/figures/tourr_5d_animation.gif.] animation using the R package *tourr* [@Wickham2011-ir].
+The round and triangular point shapes indicate the two mixture components $\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)$ and
+$\mathcal{N}(\bm{\mu}_2, \bm{\sigma}_2)$, and the colors indicate the distance between each point and the center of the distribution.
+It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
+
 ```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
@@
 
-As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
-To start with, we simulate vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ for $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, with the same means $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$ and different variance-covariance matrices $\pmb{\Sigma}_1 = diag(1,1,1,1)$ and $\pmb{\Sigma}_2 = diag(2,2,2,2)$.
-The mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
-With this design, the observations from the second component tend to be outlying anomalies.
-The data are mapped to a hemisphere via the equation
-$v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2$ where $v_5>0$ and $r$ is set as $8$.
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path^[See the animation of the 5-D grand tour at https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif.] animation using the R package *tourr* [@Wickham2011-ir].
-The round and triangular point shapes indicate the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and
-$\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors indicate the distance between each point and the center of the distribution.
-It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
-
-To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to $\pmb{v}_i$ so that $\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N$.
-Next, we rotate the $100$-dimensional vectors $(\pmb{v}_1, \dots, \pmb{v}_N)^\prime$ by multiplying by a randomly generated rotation matrix.
-To generate the rotation matrix we first simulate elements from a uniform $(0,1)$ distribution, stack them into a $100\times 100$ matrix $\pmb{A}$ and then take the R matrix from the QR decomposition of $\pmb{A}$.
+To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to $\bm{v}_i$ so that $\bm{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N$.
+Next, we rotate the $100$-dimensional vectors $(\bm{v}_1, \dots, \bm{v}_N)^\prime$ by multiplying by a randomly generated rotation matrix.
+To generate the rotation matrix we first simulate elements from a uniform $(0,1)$ distribution, stack them into a $100\times 100$ matrix $\bm{A}$ and then take the R matrix from the QR decomposition of $\bm{A}$.
 Rotating the vectors results in input vectors that are no longer sparse.
@@
 
-Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
+\autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
 As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.

### 2023-06-05 jcgs/kderm_jcgs.Rmd
Formatting and citations

@@
     template: asa_template.tex
+    citation_package: biblatex
+    bib_engine: biber
     fig_caption: yes
@@
 
-```{r child="03-chap3-kderm.Rmd"}
+```{r child = "03-chap3-kderm.Rmd"}
 
 ```
+
+\clearpage
 
@@
 ```
+
+\clearpage
+
+# References {-}

### 2023-06-05 jcgs/kderm_jcgs.tex
Formatting and citations

--- 2023-06-01 monash/dckde.tex
@@
-\documentclass[11pt,a4paper,]{article}
-\usepackage{lmodern}
-
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\hypersetup{
-            pdftitle={Distortion corrected kernel density estimate on Riemannian manifolds},
-            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{geometry}
-\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
-\usepackage[style=authoryear-comp,]{biblatex}
-\addbibresource{references.bib}
-\usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
+% !TeX program = pdfLaTeX
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx,psfrag,epsf}
+\usepackage{enumerate}
+%\usepackage{natbib}
+\usepackage{textcomp}
+\usepackage[hyphens]{url} % not crucial - just used below for the URL
+\usepackage{hyperref}
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\title{Distortion corrected kernel density estimate on Riemannian manifolds}
-
-%% MONASH STUFF
-
-%% CAPTIONS
-\RequirePackage{caption}
-\DeclareCaptionStyle{italic}[justification=centering]
- {labelfont={bf},textfont={it},labelsep=colon}
-\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
-\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
-
-%% FONT
-\RequirePackage{bera}
-\RequirePackage{mathpazo}
-
-%% HEADERS AND FOOTERS
-\RequirePackage{fancyhdr}
-\pagestyle{fancy}
-\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
-\makeatletter
-\lhead{\textsf{\expandafter{\@title}}}
-\makeatother
-\rhead{}
-\cfoot{}
-\setlength{\headheight}{15pt}
-\renewcommand{\headrulewidth}{0.4pt}
-\renewcommand{\footrulewidth}{0.4pt}
-\fancypagestyle{plain}{%
-\fancyhf{} % clear all header and footer fields
-\fancyfoot[C]{\sffamily\thepage} % except the center
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrulewidth}{0pt}}
-
-%% MATHS
-\RequirePackage{bm,amsmath}
-\allowdisplaybreaks
-
-%% GRAPHICS
-\RequirePackage{graphicx}
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.85}
-\renewcommand{\bottomfraction}{0.85}
-\renewcommand{\textfraction}{0.15}
-\renewcommand{\floatpagefraction}{0.8}
-
-%\RequirePackage[section]{placeins}
-
-%% SECTION TITLES
-\RequirePackage[compact,sf,bf]{titlesec}
-\titleformat{\section}[block]
-  {\fontsize{15}{17}\bfseries\sffamily}
-  {\thesection}
-  {0.4em}{}
-\titleformat{\subsection}[block]
-  {\fontsize{12}{14}\bfseries\sffamily}
-  {\thesubsection}
-  {0.4em}{}
-\titlespacing{\section}{0pt}{*5}{*1}
-\titlespacing{\subsection}{0pt}{*2}{*0.2}
-
-%% TITLE PAGE
-\def\Date{\number\day}
-\def\Month{\ifcase\month\or
- January\or February\or March\or April\or May\or June\or
- July\or August\or September\or October\or November\or December\fi}
-\def\Year{\number\year}
-
-\makeatletter
-\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
-\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
-\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
-\def\nojel{\def\showjel{}}
-\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
-\def\cover{{\sffamily\setcounter{page}{0}
-        \thispagestyle{empty}
-        \placefig{2}{1.5}{width=5cm}{monash2}
-        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
-        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
-        \begin{textblock}{7}(12.7,27.9)\hfill
-        \includegraphics[height=0.7cm]{AACSB}~~~
-        \includegraphics[height=0.7cm]{EQUIS}~~~
-        \includegraphics[height=0.7cm]{AMBA}
-        \end{textblock}
-        \vspace*{2cm}
-        \begin{center}\Large
-        Department of Econometrics and Business Statistics\\[.5cm]
-        \footnotesize http://monash.edu/business/ebs/research/publications
-        \end{center}\vspace{2cm}
-        \begin{center}
-        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
-                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
-                \LARGE\@author\end{onehalfspace}
-        }}
-        \end{center}
-        \vfill
-                \begin{center}\Large
-                \Month~\Year\\[1cm]
-                Working Paper \@wp
-        \end{center}\vspace*{2cm}}}
-\def\pageone{{\sffamily\setstretch{1}%
-        \thispagestyle{empty}%
-        \vbox to \textheight{%
-        \raggedright\baselineskip=1.2cm
-     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
-        \vspace{2cm}\par
-        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
-        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
-        \hspace{1cm}\showjel\vss}}}
-\def\blindtitle{{\sffamily
-     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
-     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
-        }}
-\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}
-
-\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
-\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
-\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
-\let\maketitle\titlepage
-\makeatother
-
-%% SPACING
-\RequirePackage{setspace}
-\spacing{1.5}
-
-%% LINE AND PAGE BREAKING
-\sloppy
-\clubpenalty = 10000
-\widowpenalty = 10000
-\brokenpenalty = 10000
-\RequirePackage{microtype}
-
-%% PARAGRAPH BREAKS
-\setlength{\parskip}{1.4ex}
-\setlength{\parindent}{0em}
-
-%% HYPERLINKS
-\RequirePackage{xcolor} % Needed for links
-\definecolor{darkblue}{rgb}{0,0,.6}
-\RequirePackage{url}
-
-\makeatletter
-\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
-\makeatother
-\hypersetup{
-     citecolor=0 0 0,
-     breaklinks=true,
-     bookmarksopen=true,
-     bookmarksnumbered=true,
-     linkcolor=darkblue,
-     urlcolor=blue,
-     citecolor=darkblue,
-     colorlinks=true}
-
-%% KEYWORDS
-\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}
-
-%% ABSTRACT
-\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
-\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
-  {\end{minipage}}
-
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-
-\usepackage[showonlyrefs]{mathtools}
-\usepackage[no-weekday]{eukdate}
 
@@
 
-\makeatletter
-\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
-\makeatother
-\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}
+\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}
+\ExecuteBibliographyOptions{bibencoding=utf8, minnames=1, maxnames=3, maxbibnames=99,dashed=false, terseinits=true, giveninits=true, uniquename=false, uniquelist=false,doi=false, isbn=false, url=true, sortcites=false, date=year}
+
+\addbibresource{kderm.bib}
 
 \DeclareFieldFormat{url}{\texttt{\url{#1}}}
+\DeclareFieldFormat[article]{url}{}
 \DeclareFieldFormat[article]{pages}{#1}
@@
 
-\nojel
-
-\RequirePackage[absolute,overlay]{textpos}
-\setlength{\TPHorizModule}{1cm}
-\setlength{\TPVertModule}{1cm}
-\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}
-
-\nocover
-
-\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
-\addresses{\textbf{Fan Cheng}\newline
-Monash University
-\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
-\textbf{Anastasios Panagiotelis}\newline
-The University of Sydney
-\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
-\textbf{Rob J Hyndman}\newline
-Monash University
-\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
-}
-
-\date{\sf\Date~\Month~\Year}
-\makeatletter
- \lfoot{\sf Cheng, Panagiotelis, Hyndman: \@date}
-\makeatother
-
-%% Any special functions or other packages can be loaded here.
-\usepackage{tabu,placeins,algorithmic}
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
+
+%% load any required packages here
+
+% Pandoc citation processing
+
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage{tabu,algorithmic}
+\usepackage[section]{placeins}
 \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
@@
 \usepackage[cal = txupr]{mathalpha}
+\usepackage{geometry}
 \geometry{margin=2.2cm}
+\usepackage{mathtools}
 \mathtoolsset{showonlyrefs}
+\usepackage{amsfonts,amssymb,bm}
+\usepackage{mathpazo}
+\usepackage[no-weekday]{eukdate}
 \def\sectionautorefname{Section}
 \def\subsectionautorefname{Section}
-% \def\equationautorefname~#1\null{Equation~(#1)\null} % NOT showing Equation 1, but Section 2.1
-\usepackage{bm}
 \DeclareMathOperator\supp{supp}
+
+\allowdisplaybreaks
+\clubpenalty = 10000
+\widowpenalty = 10000
+\brokenpenalty = 10000
+\setcounter{topnumber}{2}
+\setcounter{bottomnumber}{2}
+\setcounter{totalnumber}{4}
+\renewcommand{\topfraction}{0.9}
+\renewcommand{\bottomfraction}{0.9}
+\renewcommand{\textfraction}{0.1}
+\renewcommand{\floatpagefraction}{0.9}
+
+\usepackage{caption}
+\DeclareCaptionStyle{italic}[justification=centering]
+ {labelfont={bf},textfont={it},labelsep=colon}
+\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
+\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
 \usepackage{booktabs}
@@
 
-% Adjust headwidth in case user has changed geometry in header-includes
-\renewcommand{\headwidth}{\textwidth}
-
 \begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+  {#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+  {
+    \title{\bf Distortion corrected kernel density estimate on Riemannian manifolds}
+
+    \author{
+            Fan Cheng \thanks{\href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}} \\
+      Monash University\\
+       and \\       Rob J Hyndman \\
+      Monash University\\
+       and \\       Anastasios Panagiotelis \\
+      The University of Sydney\\
+          }
     \maketitle
+  } \fi
+
+\if1\blind
+  {
+    \bigskip
+    \bigskip
+    \bigskip
+    \begin{center}
+      {\LARGE\bf Distortion corrected kernel density estimate on Riemannian manifolds}
+    \end{center}
+    \medskip
+  } \fi
+
+\bigskip
 \begin{abstract}
@@
 \end{abstract}
-\begin{keywords}
-manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
-\end{keywords}
+
+\noindent%
+{\it Keywords:} manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
+  \vfill
 
   \newpage
+  \spacingset{1.45} % DON'T change the spacing!
 
@@
 
-Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
+Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Scott2015-vl} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
 Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
@@
 However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\).
-The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}.
+The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da}, and \textcite{Berenfeld2022-fd}.
 However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
@@
 
-When dealing with data that lie on a manifold, it is common to use dimension reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\).
-Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx} and \textcite{Cheng2021-dh}.
-One could easily apply KDE to the output vectors \(\pmb{y}\).
+When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
+These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\).
+Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx}, and \textcite{Cheng2021-dh}.
+One could easily apply KDE to the output vectors \(\bm{y}\).
 However, this would be a misleading representation of the density of points as they lie on the manifold.
 To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,\]
-where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\).
-By a change of variables, this would imply the following density for \(\pmb{y}\),
+\[Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,\]
+where \(f\) is the probability density function on the manifold, \(\bm{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\).
+By a change of variables, this would imply the following density for \(\bm{y}\),
 \begin{equation}
 \label{eq:changevar}
-Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
+Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
 \end{equation}
-where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
-It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|\pmb{H}(\pmb{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant.
-Since \(\psi\) will typically be a non-linear function, the value of \(|\pmb{H}(\pmb{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
-Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
+where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
+It can be seen that the density in terms of \(\bm{y}\), differs from the density on the manifold due to the term \(|\bm{H}(\bm{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant.
+Since \(\psi\) will typically be a non-linear function, the value of \(|\bm{H}(\bm{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
+Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\bm{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
 
@@
 For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density.
+Critical to our estimator, is obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
 In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
@@
 
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
+In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
+In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\bm{y}\) is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
 This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
@@
 In the following, we denote \(M\) as the \(d\)-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted \(\pmb{p}\) in general, with \(\pmb{p}_1,\dots,\pmb{p}_n\) denoting the observed sample.
-Often \(\pmb{p}_i\) will be high-dimensional vectors such that \(\pmb{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case.
-For instance, \(\pmb{p}_i\) may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each \(\pmb{p}_i\) only require some sense of distance between the `input' points, \(d(\pmb{p}_i,\pmb{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\pmb{y}_1,\dots,\pmb{y}_n\), where \(\pmb{y}_i\in\mathbb{R}^d\).
-We will denote this embedding as \(\pmb{y}_i=\psi(\pmb{p}_i)\).
+Points on this manifold are denoted \(\bm{p}\) in general, with \(\bm{p}_1,\dots,\bm{p}_n\) denoting the observed sample.
+Often \(\bm{p}_i\) will be high-dimensional vectors such that \(\bm{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case.
+For instance, \(\bm{p}_i\) may be probability distributions on a statistical manifold.
+The methods we propose for estimating the density at each \(\bm{p}_i\) only require some sense of distance between the `input' points, \(d(\bm{p}_i,\bm{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\bm{y}_1,\dots,\bm{y}_n\), where \(\bm{y}_i\in\mathbb{R}^d\).
+We will denote this embedding as \(\bm{y}_i=\psi(\bm{p}_i)\).
 Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
-\begin{align}
+\begin{equation}
+\begin{aligned}
 \label{eq:kernelcondition}
-& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
-(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
-(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
+& (i) \int_{\mathbb{R}^{d}} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=1;
+(ii) \int_{\mathbb{R}^{d}} \bm{y} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=0;
+(iii) \int_{\mathbb{R}^{d}}\|\bm{y}\|^{2} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})<\infty; \\
 & (iv) \supp K=[0; 1];
-(v) \sup K(\|\pmb{y}\|)=K(0).
-\end{align}
+(v) \sup K(\|\bm{y}\|)=K(0).
+\end{aligned}
+\end{equation}
+
 Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
@@
 
-For data \(\pmb{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\pmb{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\pmb{y}\) is given by
+For data \(\bm{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\bm{y}\) is given by
 \begin{equation}
 \label{eq:vkde}
-\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\pmb{y}-\pmb{y}_i\|}{r}\right).
+\hat{f}(\bm{y})=\frac{1}{N}\sum\limits_{i=1}^N \frac{1}{r^d} K\left(\frac{\|\bm{y}-\bm{y}_i\|}{r}\right).
 \end{equation}
@@
 The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point \(\pmb{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\pmb{y}\), times a term that ensures the density integrates to 1.
+The density at a point \(\bm{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\bm{y}\), times a term that ensures the density integrates to 1.
 In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\pmb{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
+However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\bm{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
 For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein.
-Kernel density estimators of this form can and have been applied directly on the output embedding \(\pmb{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
+Kernel density estimators of this form can and have been applied directly on the output embedding \(\bm{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
 
@@
 \label{eq:denriem}
-\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{p}_i}(\pmb{p})} K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right),
+\hat{f}(\bm{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\bm{p}_i}(\bm{p})} K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right),
 \end{equation}
 
-where \(d_g(\pmb{p}, \pmb{p}_i)\) denotes the geodesic distance between two points on the manifold \(\pmb{p}\) and \(\pmb{p}_i\) and \(\theta_{\pmb{p}_i}(\pmb{p})\) is known as the volume density function.
-The intuition behind the term \(K\left(\frac{d_g(\pmb{p}, \pmb{p}_i)}{r}\right)\) is relatively clear.
-For example, for a uniform kernel, the estimator at point \(\pmb{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\pmb{p}\).
+where \(d_g(\bm{p}, \bm{p}_i)\) denotes the geodesic distance between two points on the manifold \(\bm{p}\) and \(\bm{p}_i\) and \(\theta_{\bm{p}_i}(\bm{p})\) is known as the volume density function.
+The intuition behind the term \(K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)\) is relatively clear.
+For example, for a uniform kernel, the estimator at point \(\bm{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\bm{p}\).
 However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
@@
 A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
-For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\pmb{p}\) is so large that some points `fall inside' the ball more than once.
+For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\bm{p}\) is so large that some points `fall inside' the ball more than once.
 For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
@@
 
-The inclusion of the volume density function \(\theta_{\pmb{p}_i}(\pmb{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
-We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\pmb{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\pmb{p}\).
+The inclusion of the volume density function \(\theta_{\bm{p}_i}(\bm{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
+We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\bm{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\bm{p}\).
 However, the volume of this ball must also be taken into account.
@@
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around \(\pmb{p}\), given by \(exp_{\pmb{p}}(\pmb{q})\), mapping vectors in the tangent space, \(\pmb{v}\in T_{\pmb{p}}M\), to points on the manifold, \(\pmb{q}\in M\).
-Loosely, \(\pmb{v}\) `points' in the direction of the geodesic between \(\pmb{p}\) and \(\pmb{q}\) and traveling along this geodesic at uniform speed \(\|\pmb{v}\|\) takes place in one unit of time.
-Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\pmb{p}\), via the inverse of the exponential map, to these \(\pmb{v}\) vectors, expressed in some local coordinate system.
+Consider the exponential map around \(\bm{p}\), given by \(exp_{\bm{p}}(\bm{q})\), mapping vectors in the tangent space, \(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).
+Loosely, \(\bm{v}\) `points' in the direction of the geodesic between \(\bm{p}\) and \(\bm{q}\) and traveling along this geodesic at uniform speed \(\|\bm{v}\|\) takes place in one unit of time.
+Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\bm{p}\), via the inverse of the exponential map, to these \(\bm{v}\) vectors, expressed in some local coordinate system.
 The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\pmb{H}(\pmb{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\pmb{p}_i\).
+In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\bm{H}(\bm{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\bm{p}_i\).
 For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.
@@
 
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\pmb{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\pmb{p}} M\) for every point \(\pmb{p} \in M\).
-Using the coordinates from a dimension reduction algorithm \(\pmb{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
-To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\pmb{y}_1,\pmb{y}_2\dots,\pmb{y}_n\) with \(d\times d\) positive definite matrices, \(\pmb{H}_1,\pmb{H}_2\dots,\pmb{H}_n\), at each data point.
+To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\bm{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
+Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\bm{p}} M\) for every point \(\bm{p} \in M\).
+Using the coordinates from a dimension reduction algorithm \(\bm{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
+To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\bm{y}_1,\bm{y}_2\dots,\bm{y}_n\) with \(d\times d\) positive definite matrices, \(\bm{H}_1,\bm{H}_2\dots,\bm{H}_n\), at each data point.
 These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
@@
 There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between \(\pmb{p}_i\) and \(\pmb{p}_j\) when \(\pmb{p}_i\) is a K-nearest neighbor of \(\pmb{p}_j\) or vice versa, and edge weights depending on the distance between \(\pmb{p}_i\) and \(\pmb{p}_j\) on the manifold.
+First, a weighted neighborhood graph is constructed, with edges between \(\bm{p}_i\) and \(\bm{p}_j\) when \(\bm{p}_i\) is a K-nearest neighbor of \(\bm{p}_j\) or vice versa, and edge weights depending on the distance between \(\bm{p}_i\) and \(\bm{p}_j\) on the manifold.
 Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-Third, a dimension reduction method is applied to obtain the output embedding \(\pmb{y}_1,\dots,\pmb{y}_n\).
+Third, a dimension reduction method is applied to obtain the output embedding \(\bm{y}_1,\dots,\bm{y}_n\).
 Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
@@
 As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
-In this case, the ranks of the matrices \(\pmb{H}_i\) are equal to \(d\).
+In this case, the ranks of the matrices \(\bm{H}_i\) are equal to \(d\).
 Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean space.
@@
   \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
-  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
-  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
-  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
-  \OptParameter{ manifold learning parameters EMBED }
+  \Input{ high-dimensional data $\bm{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
+  \Output{ low-dimensional data $\bm{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\bm{H}_i$ for all $i=1,\ldots,N$ }
+  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm EMBED }
+  \OptParameter{ manifold learning parameters in EMBED }
   \BlankLine
@@
   $$
-  where $\widetilde{D}=diag{\widetilde{W}\pmb{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\pmb{1}}$;
-
-  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
-
-  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
+  where $\widetilde{D}=diag{\widetilde{W}\bm{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\bm{1}}$;
+
+  \STATE Embed all data point $\bm{X}\in \mathbb{R}^s$ to embedding coordinates $\bm{Y}=(\bm{y}^1,\dots,\bm{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;
+
+  \STATE Obtain the $d\times d\times N$ tensor $\bm{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\bm{Y}$ with elements in $\bm{\tilde{H}}$ found by
   $$
-    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
+    \bm{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\bm{y}^l \odot \bm{y}^k\right) - \bm{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \bm{y}^k\right) - \bm{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \bm{y}^l\right)\right],
   $$
-  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;
-
-  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
+  where $l,k=1,\dots,d$, $\bm{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;
+
+  \STATE For each embedding data point $\bm{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\bm{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\bm{H}^{\cdot\cdot i}}$ below
   $$
-    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
+    \bm{H}_i = \bm{U} \bm{\Lambda}_d^{-1} \bm{U}^\prime,
   $$
-  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.
+  where $[\bm{U}, \bm{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\bm{H}^{\cdot\cdot i}}$, $\bm{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\bm{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.
 
@@
 The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
+To understand these differences, it is first critical to appreciate that the coordinates \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
 This is crucial for two reasons.
-First, this implies that the term \(\| \pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\|\) approximates the geodesic distance between \(\pmb{y}_i\) and \(\pmb{y}_j\).
-Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\pmb{y}_i\), and it is this mapping that is approximated by \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\).
-For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\pmb{H}^{-1/2}_i (\pmb{y}_j - \pmb{y}_i)\) which ensures that the embedding approximates the logarithmic map.
+First, this implies that the term \(\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|\) approximates the geodesic distance between \(\bm{y}_i\) and \(\bm{y}_j\).
+Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\bm{y}_i\), and it is this mapping that is approximated by \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\).
+For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\) which ensures that the embedding approximates the logarithmic map.
 Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
 
-One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\pmb{H}_j\).
+One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\bm{H}_j\).
 To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
@@
 The entire workflow is summarized in \autoref{fig:vkde}.
-The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\pmb{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
+The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\bm{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
 Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
@@
 
-The simulation setup for the twin peaks example is to first generate vectors \(\pmb{v}_1,\dots,\pmb{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
+The simulation setup for the twin peaks example is to first generate vectors \(\bm{v}_1,\dots,\bm{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
 The mixture has four components with different means
-\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\).
+\(\bm{\mu_1}=(0.25, 0.25)^\prime, \bm{\mu_2}=(0.25, 0.75)^\prime, \bm{\mu_3}=(0.75, 0.25)^\prime, \bm{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\bm{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\).
 The mixture proportions
@@
 It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in Equation \eqref{eq:twinpeak} distorts the distribution.
-To recover the true distribution requires the correct Jacobian term for the pushforward from \(\pmb{v}\) to the volume form of twin peaks manifold.
-By treating the \(\pmb{v}\) as an `output' embedding from input points \(\pmb{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\pmb{\Gamma}_i\) for \(i=1,\dots,n\) where \(\pmb{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\pmb{v}\).
-This notation is distinct from \(\pmb{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as \(f(\pmb{p}_i)=f(\pmb{v}_i)|\pmb{\Gamma}_i|^{1/2}\), where \(f(\pmb{v}_i)\) is the density of a four-component mixture of normals.
-Knowledge of \(\pmb{v}\) and \(\pmb{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' for densities on the manifold.
-\autoref{fig:metadensity} shows the simulated \(\pmb{v}\) with color indicating the true density of data on the manifold.
+To recover the true distribution requires the correct Jacobian term for the pushforward from \(\bm{v}\) to the volume form of twin peaks manifold.
+By treating the \(\bm{v}\) as an `output' embedding from input points \(\bm{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\bm{\Gamma}_i\) for \(i=1,\dots,n\) where \(\bm{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\bm{v}\).
+This notation is distinct from \(\bm{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
+The true density on the manifold can be obtained as \(f(\bm{p}_i)=f(\bm{v}_i)|\bm{\Gamma}_i|^{1/2}\), where \(f(\bm{v}_i)\) is the density of a four-component mixture of normals.
+Knowledge of \(\bm{v}\) and \(\bm{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' for densities on the manifold.
+\autoref{fig:metadensity} shows the simulated \(\bm{v}\) with color indicating the true density of data on the manifold.
 Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in yellow.
 The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the \(\pmb{v}\).
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
+The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the \(\bm{v}\).
+
+\begin{figure}[!p]
+
+{\centering \includegraphics[width=0.85\linewidth]{figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5}
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank}
+\begin{figure}[!p]
+
+{\centering \includegraphics[width=0.8\linewidth]{figures/Twin Peak2000_density_compare_isomapvs4ml_radius8_r0_5_rank}
 
@@
 \end{figure}
+
+\FloatBarrier
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
-
-}
-
-\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
-Distant points in darker colors could be seen as anomalies.}\label{fig:fivedmeta}
-\end{figure}
-
 As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
-To start with, we simulate vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), with the same means \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\pmb{\Sigma}_1 = diag(1,1,1,1)\) and \(\pmb{\Sigma}_2 = diag(2,2,2,2)\).
+To start with, we simulate vectors \((\bm{v}_1, \dots, \bm{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)\) and \(\mathcal{N}(\bm{\mu}_2, \bm{\Sigma}_2)\), with the same means \(\bm{\mu}_1 = \bm{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\bm{\Sigma}_1 = diag(1,1,1,1)\) and \(\bm{\Sigma}_2 = diag(2,2,2,2)\).
 The mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
@@
 \(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/paper/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
-The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and
-\(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors indicate the distance between each point and the center of the distribution.
+\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/monash/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
+The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)\) and
+\(\mathcal{N}(\bm{\mu}_2, \bm{\sigma}_2)\), and the colors indicate the distance between each point and the center of the distribution.
 It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
 
-To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to \(\pmb{v}_i\) so that \(\pmb{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\).
-Next, we rotate the \(100\)-dimensional vectors \((\pmb{v}_1, \dots, \pmb{v}_N)^\prime\) by multiplying by a randomly generated rotation matrix.
-To generate the rotation matrix we first simulate elements from a uniform \((0,1)\) distribution, stack them into a \(100\times 100\) matrix \(\pmb{A}\) and then take the R matrix from the QR decomposition of \(\pmb{A}\).
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}
+
+}
+
+\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing two Gaussian mixture components and the colors showing the distance to the kernel cores.
+Distant points in darker colors could be seen as anomalies.}\label{fig:fivedmeta}
+\end{figure}
+
+To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to \(\bm{v}_i\) so that \(\bm{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\).
+Next, we rotate the \(100\)-dimensional vectors \((\bm{v}_1, \dots, \bm{v}_N)^\prime\) by multiplying by a randomly generated rotation matrix.
+To generate the rotation matrix we first simulate elements from a uniform \((0,1)\) distribution, stack them into a \(100\times 100\) matrix \(\bm{A}\) and then take the R matrix from the QR decomposition of \(\bm{A}\).
 Rotating the vectors results in input vectors that are no longer sparse.
@@
 For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP there more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
+However, for Laplacian Eigenmaps and UMAP, more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
 In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
 
-Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
+\autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
 As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.
@@
 
+\clearpage
+
 \appendix
@@
 
-\caption{Comparison of outliers found by one manifold learning method compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml-4}
-\end{figure}
+\caption{Comparison of outliers found by UMAP compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tpumapvs4ml}
+\end{figure}
+
+\clearpage
 

### 2023-07-14 jcgs/kderm_jcgs.Rmd
changed corresponding author and abstract for submission

@@
 ---
-title: "Distortion corrected kernel density estimate on Riemannian manifolds"
+title: "Distortion corrected kernel density estimator on Riemannian manifolds"
 blinded: 0 # Set to 1 for blinding
@@
     affiliation: Monash University
-    thanks: Fan.Cheng@monash.edu
   - name: Rob J Hyndman
     affiliation: Monash University
+    thanks: Rob J Hyndman's research is partially funded by the Australian Government through the Australian Research Council Industrial Transformation Training Centre in Optimisation Technologies, Integrated Methodologies, and Applications (OPTIMA), Project ID IC200100009.
   - name: Anastasios Panagiotelis
     affiliation: The University of Sydney
-abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements.
-The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D data from the Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
-We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
-A shiny app in R is also developed for various simulation scenarios.
+    thanks: Corresponding Author
+abstract: "Manifold learning obtains a low-dimensional representation of an underlying Riemannian manifold supporting high-dimensional data.
+Kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
+We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding, with a bandwidth that depends on the estimated Riemannian metric at each data point.
+Exploiting the geometric information of the manifold leads to more accurate density estimation, which subsequently could be used for anomaly detection.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations including one with data lying in a 100 dimensional ambient space.
+We demonstrate that the proposed DC-KDE improves the density estimates as long as the manifold learning embedding is of sufficient quality, and has higher rank correlations with the true manifold density.
+Further simulation results are provided via a supplementary  R shiny app.
 The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
+"
 keywords:

### 2024-09-03 jcgs/03-chap3-kderm.Rmd
Add reviewers' comments

@@
 The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
+However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice, with the exception of @berenfeld2021-den and @berenfeld2022-est which demonstrate the minimax optimality of certain types of KDE in the context of density estimation on manifolds.
+Our objective in this paper is to propose KDE for data that lie on some unknown manifold after embedding.
 
 When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\bm{p}$ to $d$-dimensional vectors $\bm{y}$.
+These algorithms, which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\bm{p}$ to $d$-dimensional vectors $\bm{y}$, where the dimension $d$ is either assumed to be known a priori or needs to be estimated separately.
 Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh.
@@
 \label{eq:changevar}
-Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
+Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\det \bm{H}(\bm{y})|^{1/2}d\bm{y},
 \end{equation}
-where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\bm{H}(\bm{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$.
+where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\bm{H}(\bm{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$ [@doCarmo1992-rm].
 It can be seen that the density in terms of $\bm{y}$, differs from the density on the manifold due to the term $|\bm{H}(\bm{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant.
@@
 For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of $|\bm{H}(\bm{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+Critical to our estimator, is obtaining an estimate of $|\bm{H}(\bm{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian metric at each data point.
 One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
@@
 
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\bm{y}$ to be augmented by coloring points according to the magnitude of the density estimate.
+In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\bm{y}$ to be augmented by coloring points according to the magnitude of the density estimate, where the dimension of $\bm{y}$, $d$, is set as 2 or 3 in the dimension reduction algorithms.
 In this way, the user is able to tell whether a region that has a cluster of points in the output space $\bm{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together.
@@
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around $\bm{p}$, given by $exp_{\bm{p}}(\bm{q})$, mapping vectors in the tangent space,  $\bm{v}\in T_{\bm{p}}M$, to points on the manifold, $\bm{q}\in M$.
+Consider the exponential map around $\bm{p}$, given by $exp_{\bm{p}}(\bm{v})$, mapping vectors in the tangent space,  $\bm{v}\in T_{\bm{p}}M$, to points on the manifold, $\bm{q}\in M$.
 Loosely, $\bm{v}$ `points' in the direction of the geodesic between $\bm{p}$ and $\bm{q}$ and traveling along this geodesic at uniform speed $\|\bm{v}\|$ takes place in one unit of time.
@@
 There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between $\bm{p}_i$ and $\bm{p}_j$ when $\bm{p}_i$ is a K-nearest neighbor of $\bm{p}_j$ or vice versa, and edge weights depending on the distance between $\bm{p}_i$ and $\bm{p}_j$ on the manifold.
-Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
+First, a weighted neighborhood graph is constructed, with edges between $\bm{p}_i$ and $\bm{p}_j$ when $\bm{p}_i$ is a fixed-radius nearest neighbor of $\bm{p}_j$ or vice versa, and edge weights depending on the distance between $\bm{p}_i$ and $\bm{p}_j$ on the manifold.
+The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively.
+Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\sqrt{\varepsilon}$ is the radius parameter in the previous step and a constant value $c=0.25$ is used for the use of heat kernel in the weighted neighborhood graph.
 Third, a dimension reduction method is applied to obtain the output embedding $\bm{y}_1,\dots,\bm{y}_n$.
@@
 This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*.
+For computational efficiency, approximate nearest neighbor searching methods are implemented to construct the neighborhood graph [@Cheng2021-dh; @Perrault-Joncas2013-pq], which are then used in both steps 1 and 3.
 Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.
@@
   $$
-  where $\widetilde{D}=diag{\widetilde{W}\bm{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\bm{1}}$;
+  where $c=0.25$, $\widetilde{D}=diag{\widetilde{W}\bm{1}}$, $\widetilde{W} = D^{-1}WD^{-1}$, and $D = diag{W\bm{1}}$;
 
@@
 To recover the true distribution requires the correct Jacobian term for the pushforward from $\bm{v}$ to the volume form of twin peaks manifold.
-By treating the $\bm{v}$ as an 'output' embedding from input points $\bm{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\bm{\Gamma}_i$ for $i=1,\dots,n$ where $\bm{\Gamma}_i$ is the Riemannian of the coordinate system given by $\bm{v}$.
+By treating the $\bm{v}$ as an 'output' embedding from input points $\bm{x}$ that lie on the true manifold and applying the Learn Metric algorithm, we can obtain $\bm{\Gamma}_i$ for $i=1,\dots,n$ where $\bm{\Gamma}_i$ is the Riemannian metric of the coordinate system given by $\bm{v}$.
 This notation is distinct from $\bm{H}_i$ which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
@@
 For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
-This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
+This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data [@cai2022-tsne].
 

### 2024-09-03 jcgs/kderm_jcgs.tex
Add reviewers' comments

@@
   {
-    \title{\bf Distortion corrected kernel density estimate on Riemannian manifolds}
+    \title{\bf Distortion corrected kernel density estimator on Riemannian manifolds}
 
     \author{
-            Fan Cheng \thanks{\href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}} \\
+            Fan Cheng \\
       Monash University\\
-       and \\       Rob J Hyndman \\
+       and \\       Rob J Hyndman \thanks{Rob J Hyndman's research is partially funded by the Australian Government through the Australian Research Council Industrial Transformation Training Centre in Optimisation Technologies, Integrated Methodologies, and Applications (OPTIMA), Project ID IC200100009.} \\
       Monash University\\
-       and \\       Anastasios Panagiotelis \\
+       and \\       Anastasios Panagiotelis \thanks{Corresponding Author} \\
       The University of Sydney\\
@@
     \begin{center}
-      {\LARGE\bf Distortion corrected kernel density estimate on Riemannian manifolds}
+      {\LARGE\bf Distortion corrected kernel density estimator on Riemannian manifolds}
     \end{center}
@@
 \begin{abstract}
-  Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data.
-However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements.
-The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D data from the Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space.
-We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density.
-A shiny app in R is also developed for various simulation scenarios.
+  Manifold learning obtains a low-dimensional representation of an underlying Riemannian manifold supporting high-dimensional data.
+Kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
+We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding, with a bandwidth that depends on the estimated Riemannian metric at each data point.
+Exploiting the geometric information of the manifold leads to more accurate density estimation, which subsequently could be used for anomaly detection.
+To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations including one with data lying in a 100 dimensional ambient space.
+We demonstrate that the proposed DC-KDE improves the density estimates as long as the manifold learning embedding is of sufficient quality, and has higher rank correlations with the true manifold density.
+Further simulation results are provided via a supplementary R shiny app.
 The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
 \end{abstract}
@@
 
-\hypertarget{dckdeintro}{%
-\section{Introduction}\label{dckdeintro}}
+\section{Introduction}\label{dckdeintro}
 
@@
 The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da}, and \textcite{Berenfeld2022-fd}.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
+However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice, with the exception of \textcite{berenfeld2021-den} and \textcite{berenfeld2022-est} which demonstrate the minimax optimality of certain types of KDE in the context of density estimation on manifolds.
+Our objective in this paper is to propose KDE for data that lie on some unknown manifold after embedding.
 
 When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\).
+These algorithms, which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\), where the dimension \(d\) is either assumed to be known a priori or needs to be estimated separately.
 Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx}, and \textcite{Cheng2021-dh}.
@@
 \label{eq:changevar}
-Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
+Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\det \bm{H}(\bm{y})|^{1/2}d\bm{y},
 \end{equation}
-where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
+where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\) \autocite{doCarmo1992-rm}.
 It can be seen that the density in terms of \(\bm{y}\), differs from the density on the manifold due to the term \(|\bm{H}(\bm{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant.
@@
 For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
+Critical to our estimator, is obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian metric at each data point.
 One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
@@
 
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
+In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring points according to the magnitude of the density estimate, where the dimension of \(\bm{y}\), \(d\), is set as 2 or 3 in the dimension reduction algorithms.
 In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\bm{y}\) is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
@@
 
-\hypertarget{kderm}{%
-\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}}
+\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}
 
@@
 
-\hypertarget{Pellet}{%
-\subsection{Kernel Density estimation on manifolds}\label{Pellet}}
+\subsection{Kernel Density estimation on manifolds}\label{Pellet}
 
@@
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around \(\bm{p}\), given by \(exp_{\bm{p}}(\bm{q})\), mapping vectors in the tangent space, \(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).
+Consider the exponential map around \(\bm{p}\), given by \(exp_{\bm{p}}(\bm{v})\), mapping vectors in the tangent space, \(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).
 Loosely, \(\bm{v}\) `points' in the direction of the geodesic between \(\bm{p}\) and \(\bm{q}\) and traveling along this geodesic at uniform speed \(\|\bm{v}\|\) takes place in one unit of time.
@@
 
-\hypertarget{MetLearn}{%
-\subsection{Riemannian metric estimation}\label{MetLearn}}
+\subsection{Riemannian metric estimation}\label{MetLearn}
 
@@
 There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between \(\bm{p}_i\) and \(\bm{p}_j\) when \(\bm{p}_i\) is a K-nearest neighbor of \(\bm{p}_j\) or vice versa, and edge weights depending on the distance between \(\bm{p}_i\) and \(\bm{p}_j\) on the manifold.
-Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
+First, a weighted neighborhood graph is constructed, with edges between \(\bm{p}_i\) and \(\bm{p}_j\) when \(\bm{p}_i\) is a fixed-radius nearest neighbor of \(\bm{p}_j\) or vice versa, and edge weights depending on the distance between \(\bm{p}_i\) and \(\bm{p}_j\) on the manifold.
+The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively.
+Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\sqrt{\varepsilon}\) is the radius parameter in the previous step and a constant value \(c=0.25\) is used for the use of heat kernel in the weighted neighborhood graph.
 Third, a dimension reduction method is applied to obtain the output embedding \(\bm{y}_1,\dots,\bm{y}_n\).
@@
 This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
+For computational efficiency, approximate nearest neighbor searching methods are implemented to construct the neighborhood graph \autocite{Cheng2021-dh,Perrault-Joncas2013-pq}, which are then used in both steps 1 and 3.
 Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.
@@
 
-\hypertarget{DCKDE}{%
-\subsection{Distortion corrected KDE}\label{DCKDE}}
+\subsection{Distortion corrected KDE}\label{DCKDE}
 
@@
 
-\hypertarget{simulation}{%
-\section{Simulations}\label{simulation}}
+\section{Simulations}\label{simulation}
 
@@
 
-\hypertarget{twodgaussian}{%
-\subsection{Twin peaks example}\label{twodgaussian}}
+\subsection{Twin peaks example}\label{twodgaussian}
 
@@
 To recover the true distribution requires the correct Jacobian term for the pushforward from \(\bm{v}\) to the volume form of twin peaks manifold.
-By treating the \(\bm{v}\) as an `output' embedding from input points \(\bm{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\bm{\Gamma}_i\) for \(i=1,\dots,n\) where \(\bm{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\bm{v}\).
+By treating the \(\bm{v}\) as an `output' embedding from input points \(\bm{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\bm{\Gamma}_i\) for \(i=1,\dots,n\) where \(\bm{\Gamma}_i\) is the Riemannian metric of the coordinate system given by \(\bm{v}\).
 This notation is distinct from \(\bm{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
@@
 For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
-This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
+This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data \autocite{cai2022-tsne}.
 
@@
 
-\hypertarget{fivedgaussian}{%
-\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}}
+\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}
 
@@
 
-\hypertarget{dckdeapplication}{%
-\section{Application}\label{dckdeapplication}}
+\section{Application}\label{dckdeapplication}
 
@@
 
-\hypertarget{dckdeconclusion}{%
-\section{Conclusions}\label{dckdeconclusion}}
+\section{Conclusions}\label{dckdeconclusion}
 
@@
 
-\hypertarget{acknowledgment}{%
-\section*{Acknowledgment}\label{acknowledgment}}
+\section*{Acknowledgment}\label{acknowledgment}
 \addcontentsline{toc}{section}{Acknowledgment}
@@
 
-\hypertarget{riemgeo}{%
-\section{Notions about Riemannian geometry}\label{riemgeo}}
+\section{Notions about Riemannian geometry}\label{riemgeo}
 
@@
 
-\hypertarget{differentiable-manifolds}{%
-\subsection{Differentiable manifolds}\label{differentiable-manifolds}}
+\subsection{Differentiable manifolds}\label{differentiable-manifolds}
 
@@
 
-\hypertarget{tangent-vector-and-tangent-space}{%
-\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}}
+\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}
 
@@
 
-\hypertarget{riemannian-metric-and-geodesic-distance}{%
-\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}}
+\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}
 
@@
 
-\hypertarget{exponential-map-and-logarithmic-map}{%
-\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}}
+\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}
 
@@
 
-\hypertarget{pushforward-and-pullback-metric}{%
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
+\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}
 
@@
 
-\hypertarget{twinpeaksappe}{%
-\section{Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
+\section{Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}
 
 This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
+The density rank used in the paper is defined as the relative position or order of an item within a set from the lowest to the highest, based on its density value.
 By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest density regions more accurately than KDE.

### 2024-09-08 jcgs/03-chap3-kderm.Rmd
Tas changes

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 ```{r setup, include=FALSE}
@@
 The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice, with the exception of @berenfeld2021-den and @berenfeld2022-est which demonstrate the minimax optimality of certain types of KDE in the context of density estimation on manifolds.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold after embedding.
+However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
+The is only a small literature dealing with the case where the manifold is unknown.
+Recently, @berenfeld2021-den, @berenfeld2022-est and @divol2022measure demonstrated the minimax optimality of certain types of KDE on manifolds.
+In contrast, our objective in this paper is to propose a new KDE for data that lie on some unknown manifold found using embedding techniques.
 
@@
 The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively.
+As an alternative, K nearest neghbors could also be used.
 Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\sqrt{\varepsilon}$ is the radius parameter in the previous step and a constant value $c=0.25$ is used for the use of heat kernel in the weighted neighborhood graph.
@@
 Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.
+Note that in \autoref{alg:learnmetric} $c$ and $\sqrt{\varepsilon} = 0.4$ enter multiplicatively therefore only one of these parameters requires tuning.
+We considered different values of $\sqrt{\varepsilon}$, noting that results were stable for a range of values close to the default value of $\sqrt{\varepsilon} = 0.4$.
 
@@
 
-The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences.
+The estimator has a similar structure to Equation \@ref(eq:denriem) which can be seen as the counterpart to Equation \@ref(eq:denestimator) if the Riemannian is known exactly.
+There are however, some key differences between the two equations.
 To understand these differences, it is first critical to appreciate that the coordinates $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq).
@@
 To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
+To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors.
 We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density.
@@
 The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
+Our method requires the choice of a single scale $r$ that tunes the varying bandwidth matrix.
+In practice this could be selected using a grid search according to some criteria that depends on the downstream task.
+For instance for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation.
 For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition).

### 2024-09-10 R/Electricity2d_03_embed_tvd.R
Code to generate table 2-3 and figure 9 for KDE on 100-D, but use violin plot instead

@@ -16,7 +16,6 @@ Jmisc::sourceAll("R/sources")
 scen <- as.numeric(commandArgs()[[6]])
 r <- 180 #1 
 # quantile(nn2res$nn.dists)
-# 0%      25%      50%      75%     100% 
 # 0.0000 179.6016 187.3877 194.8748 277.0549 
 
 load("data/half_count_ratio_3639id336tow.rda")

### 2024-09-10 R/Simulation4d_03_plots.R
Code to generate table 2-3 and figure 9 for KDE on 100-D, but use violin plot instead

@@ -143,6 +143,382 @@ p_rank
 # ggsave(paste0("paper/figures/", "sim4d", N, "_density_comparison_4ml_radius", 10, "_k200_rankdensity_circleoutlier_with1rec.png"), p_rank, width = 8, height = 10, dpi = 300)
 
 
+
+
+
+
+
+# ----KDE with ks::kde on 100-d data instead of 4-d embedding-------------------
+library(weird)
+library(tictoc)
+
+# From Rob
+n <- 100
+d <- 3
+y <- matrix(rnorm(n*d), ncol = d)
+H <- kde_bandwidth(y)
+fhat <- kde(y, H)
+
+
+# Handcode KDE
+library(mvtnorm)
+multivariate_kde <- function(x, eval_points, H = NULL) {
+  n <- nrow(x)
+  d <- ncol(x)
+  m <- nrow(eval_points)
+  
+  # If H is not provided, use Scott's rule
+  if (is.null(H)) {
+    H <- diag(d) * n^(-1/(d+4))
+  }
+  
+  # Ensure H is symmetric and positive definite
+  H <- (H + t(H)) / 2
+  eigen_values <- eigen(H)$values
+  if (any(eigen_values <= 0)) {
+    stop("Bandwidth matrix H must be positive definite")
+  }
+  
+  # Compute the inverse and determinant of H
+  H_inv <- solve(H)
+  H_det <- det(H)
+  
+  # Compute the density estimate
+  density <- numeric(m)
+  for (i in 1:m) {
+    for (j in 1:n) {
+      diff <- eval_points[i,] - x[j,]
+      density[i] <- density[i] + exp(-0.5 * t(diff) %*% H_inv %*% diff)
+    }
+  }
+  
+  # Normalize the density estimate
+  density <- density / (n * (2*pi)^(d/2) * sqrt(H_det))
+  
+  return(density)
+}
+# # Generate some sample data
+# set.seed(123)
+# n <- 1000
+# d <- 3
+# x <- matrix(rnorm(n*d), ncol=d)
+# # Define points to evaluate the density
+# eval_points <- matrix(seq(-3, 3, length.out=20), ncol=d, nrow=20, byrow=TRUE)
+# # Estimate the density
+# density_estimates <- multivariate_kde(x, eval_points)
+# print(density_estimates)
+
+
+
+# Simplify the function with H=diag(d)
+multivariate_kde_I <- function(x, eval_points) {
+  n <- nrow(x)
+  d <- ncol(x)
[…]

### 2024-09-10 jcgs/03-chap3-kderm.Rmd
Update paper pdf

@@
 However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
-The is only a small literature dealing with the case where the manifold is unknown.
+There is only a small literature dealing with the case where the manifold is unknown.
 Recently, @berenfeld2021-den, @berenfeld2022-est and @divol2022measure demonstrated the minimax optimality of certain types of KDE on manifolds.
@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue.
+Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue shades.
 With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 
-(ref:twinpeaks3dcaption) Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.
+(ref:twinpeaks3dcaption) Scatterplot of the 3-D twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.
 
@@
 \autoref{fig:metadensity} shows the simulated $\bm{v}$ with color indicating the true density of data on the manifold.
-Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in yellow.
+Anomalies are defined as points with the lowest densities shown in darker colors and with 'typical' points having the highest density shown in lighter colors.
 The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
@@
 We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$.
-Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
+Colors show the different estimated densities at each point with anomalies shown in darker colors with blue indexing text, and higher density points shown in lighter colors.
 For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
@@
 
-```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
+```{r tpisomapden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding of the twin peak data.
 The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
@@
 
-```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding.
+```{r tptsneden, fig.align = 'center', out.width = "100%", fig.cap="Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding of the twin peak data.
 KDE without distortion correction gives many more misclassified anomalies.", echo = FALSE}
@@
 
-```{r tpisomapvs4ml, fig.pos = "!p", fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
+```{r tpisomapvs4ml, fig.pos = "!p", fig.align = 'center', out.width = "80%", fig.cap="Comparison of ranks of the estimated densities for the twin peak example based on ISOMAP and four other dimension reduction algorithms in each row.
 Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.", echo = FALSE}
@@
 
-```{r fivedisomapden, fig.align = 'center', out.width = "90%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE.
+```{r fivedisomapden, fig.align = 'center', out.width = "90%", fig.cap="Rank comparison between the true density and estimated density from both DC-KDE and KDE for the semi-hypersphere example.
 Four manifold learning methods are used row-wise.
@@
 data.frame(cors) %>%
-  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>%
+  mutate(across(where(is.numeric), ~format(round(., 3), nsmall = 3))) %>%
   kableExtra::kbl(caption = "Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data.
@@
 There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions.
-The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
+The dark-colored points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
 First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
@@
 Our method requires the choice of a single scale $r$ that tunes the varying bandwidth matrix.
-In practice this could be selected using a grid search according to some criteria that depends on the downstream task.
-For instance for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation.
+In practice, this could be selected using a grid search according to some criteria that depends on the downstream task.
+For instance, for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation.
 For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition).

### 2024-09-10 jcgs/kderm_jcgs.tex
Update paper pdf

@@
 The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da}, and \textcite{Berenfeld2022-fd}.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice, with the exception of \textcite{berenfeld2021-den} and \textcite{berenfeld2022-est} which demonstrate the minimax optimality of certain types of KDE in the context of density estimation on manifolds.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold after embedding.
+However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
+There is only a small literature dealing with the case where the manifold is unknown.
+Recently, \textcite{berenfeld2021-den}, \textcite{berenfeld2022-est} and \textcite{divol2022measure} demonstrated the minimax optimality of certain types of KDE on manifolds.
+In contrast, our objective in this paper is to propose a new KDE for data that lie on some unknown manifold found using embedding techniques.
 
@@
 The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively.
+As an alternative, K nearest neghbors could also be used.
 Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\sqrt{\varepsilon}\) is the radius parameter in the previous step and a constant value \(c=0.25\) is used for the use of heat kernel in the weighted neighborhood graph.
@@
 Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.
+Note that in \autoref{alg:learnmetric} \(c\) and \(\sqrt{\varepsilon} = 0.4\) enter multiplicatively therefore only one of these parameters requires tuning.
+We considered different values of \(\sqrt{\varepsilon}\), noting that results were stable for a range of values close to the default value of \(\sqrt{\varepsilon} = 0.4\).
 
@@
 
-The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
+The estimator has a similar structure to Equation \eqref{eq:denriem} which can be seen as the counterpart to Equation \eqref{eq:denestimator} if the Riemannian is known exactly.
+There are however, some key differences between the two equations.
 To understand these differences, it is first critical to appreciate that the coordinates \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\bm{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
+Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue shades.
 With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 
-\caption{Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.}\label{fig:twinpeaks3d}
+\caption{Scatterplot of the 3-D twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.}\label{fig:twinpeaks3d}
 \end{figure}
@@
 \autoref{fig:metadensity} shows the simulated \(\bm{v}\) with color indicating the true density of data on the manifold.
-Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in yellow.
+Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in lighter colors.
 The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
@@
 We set the bandwidth parameter in the DC-KDE \eqref{eq:denestimator} as \(r=0.5\).
-Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
+Colors show the different estimated densities at each point with anomalies shown in darker colors with blue indexing text, and higher density points shown in lighter colors.
 For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
@@
 
-\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding.
+\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on ISOMAP embedding of the twin peak data.
 The colors indicate different levels of highest density regions and the shapes indicate whether the density estimators correctly classify the true anomalies.
@@
 
-\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding.
+\caption{Scatterplot of log scale ranks of true density and estimated density ranks for DC-KDE (in the left panel) and KDE (in the right panel) based on t-SNE embedding of the twin peak data.
 KDE without distortion correction gives many more misclassified anomalies.}\label{fig:tptsneden}
@@
 
-\caption{Comparison of ranks of the estimated densities based on ISOMAP  and four other dimension reduction algorithms in each row.
+\caption{Comparison of ranks of the estimated densities for the twin peak example based on ISOMAP and four other dimension reduction algorithms in each row.
 Distortion-corrected KDE (on the left panel) and KDE (on the right panel) are compared and DC-KDE shows the robustness to the use of different dimension reduction methods.}\label{fig:tpisomapvs4ml}
@@
 
-\caption{Rank comparison between the true density and estimated density from both DC-KDE and KDE.
+\caption{Rank comparison between the true density and estimated density from both DC-KDE and KDE for the semi-hypersphere example.
 Four manifold learning methods are used row-wise.
@@
 \begin{table}
-
+\centering
 \caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data.
@@
 \midrule
-DC-KDE & 0.968 & 0.970 & \textbf{0.8674} & \textbf{0.782}\\
-KDE & \textbf{0.976} & \textbf{0.971} & 0.0328 & -0.181\\
+DC-KDE & 0.968 & 0.970 & \textbf{0.867} & \textbf{0.782}\\
+KDE & \textbf{0.976} & \textbf{0.971} & 0.033 & -0.181\\
 \bottomrule
@@
 There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions.
-The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
+The dark-colored points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
 First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
@@
 The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
+Our method requires the choice of a single scale \(r\) that tunes the varying bandwidth matrix.
+In practice, this could be selected using a grid search according to some criteria that depends on the downstream task.
+For instance, for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation.
 For densities on unknown manifolds, \textcite{Berenfeld2021-as} also provide some insights on the assumptions of the kernel function in \eqref{eq:kernelcondition}.

### 2024-09-11 jcgs/03-chap3-kderm.tex
minor changes

--- 2024-09-03 jcgs/kderm_jcgs_submission1.tex
@@
-% !TeX program = pdfLaTeX
-\documentclass[12pt]{article}
-\usepackage{amsmath}
-\usepackage{graphicx,psfrag,epsf}
-\usepackage{enumerate}
-%\usepackage{natbib}
-\usepackage{textcomp}
-\usepackage[hyphens]{url} % not crucial - just used below for the URL
-\usepackage{hyperref}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math} % this also loads fontspec
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-%% BIBLIOGRAPHY
-
-\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}
-\ExecuteBibliographyOptions{bibencoding=utf8, minnames=1, maxnames=3, maxbibnames=99,dashed=false, terseinits=true, giveninits=true, uniquename=false, uniquelist=false,doi=false, isbn=false, url=true, sortcites=false, date=year}
-
-\addbibresource{kderm.bib}
-
-\DeclareFieldFormat{url}{\texttt{\url{#1}}}
-\DeclareFieldFormat[article]{url}{}
-\DeclareFieldFormat[article]{pages}{#1}
-\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
-\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
-\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
-\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
-\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
-\DeclareFieldFormat[inproceedings]{title}{#1}
-\DeclareFieldFormat{shorthandwidth}{#1}
-% No dot before number of articles
-\usepackage{xpatch}
-\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
-% Remove In: for an article.
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-
-\makeatletter
-\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
-\makeatother
-\renewcommand*{\finalnamedelim}{%
-  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
-  \addspace\&\space}
-
-%\pdfminorversion=4
-% NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.3in}%
-\addtolength{\topmargin}{-.8in}%
-
-%% load any required packages here
-
-% Pandoc citation processing
-
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{tabu,algorithmic}
-\usepackage[section]{placeins}
-\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
-\urlstyle{tt}  % use monospace font for urls
-\usepackage[cal = txupr]{mathalpha}
-\usepackage{geometry}
-\geometry{margin=2.2cm}
-\usepackage{mathtools}
-\mathtoolsset{showonlyrefs}
-\usepackage{amsfonts,amssymb,bm}
-\usepackage{mathpazo}
-\usepackage[no-weekday]{eukdate}
-\def\sectionautorefname{Section}
-\def\subsectionautorefname{Section}
-\DeclareMathOperator\supp{supp}
-
-\allowdisplaybreaks
-\clubpenalty = 10000
-\widowpenalty = 10000
-\brokenpenalty = 10000
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.9}
-\renewcommand{\bottomfraction}{0.9}
-\renewcommand{\textfraction}{0.1}
-\renewcommand{\floatpagefraction}{0.9}
-
-\usepackage{caption}
-\DeclareCaptionStyle{italic}[justification=centering]
- {labelfont={bf},textfont={it},labelsep=colon}
-\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
-\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+
+\author{}
+\date{\vspace{-2.5em}}
 
 \begin{document}
-
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-  {#1}\small\normalsize} \spacingset{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\if0\blind
-  {
-    \title{\bf Distortion corrected kernel density estimator on Riemannian manifolds}
-
-    \author{
-            Fan Cheng \\
-      Monash University\\
-       and \\       Rob J Hyndman \thanks{Rob J Hyndman's research is partially funded by the Australian Government through the Australian Research Council Industrial Transformation Training Centre in Optimisation Technologies, Integrated Methodologies, and Applications (OPTIMA), Project ID IC200100009.} \\
-      Monash University\\
-       and \\       Anastasios Panagiotelis \thanks{Corresponding Author} \\
-      The University of Sydney\\
-          }
-    \maketitle
-  } \fi
-
-\if1\blind
-  {
-    \bigskip
-    \bigskip
-    \bigskip
-    \begin{center}
-      {\LARGE\bf Distortion corrected kernel density estimator on Riemannian manifolds}
-    \end{center}
-    \medskip
-  } \fi
-
-\bigskip
-\begin{abstract}
-  Manifold learning obtains a low-dimensional representation of an underlying Riemannian manifold supporting high-dimensional data.
-Kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold.
-We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding, with a bandwidth that depends on the estimated Riemannian metric at each data point.
-Exploiting the geometric information of the manifold leads to more accurate density estimation, which subsequently could be used for anomaly detection.
-To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations including one with data lying in a 100 dimensional ambient space.
-We demonstrate that the proposed DC-KDE improves the density estimates as long as the manifold learning embedding is of sufficient quality, and has higher rank correlations with the true manifold density.
-Further simulation results are provided via a supplementary R shiny app.
-The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
-\end{abstract}
-
-\noindent%
-{\it Keywords:} manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
-  \vfill
-
-  \newpage
-  \spacingset{1.45} % DON'T change the spacing!
 
@@
 
-Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Scott2015-vl} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning.
-Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}.
-For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality.
-However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\).
-The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da}, and \textcite{Berenfeld2022-fd}.
-However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice.
-Our objective in this paper is to propose KDE for data that lie on some unknown manifold.
-
-When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques.
-These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\).
-Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx}, and \textcite{Cheng2021-dh}.
-One could easily apply KDE to the output vectors \(\bm{y}\).
-However, this would be a misleading representation of the density of points as they lie on the manifold.
-To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
-\[Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,\]
-where \(f\) is the probability density function on the manifold, \(\bm{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\).
-By a change of variables, this would imply the following density for \(\bm{y}\),
-\begin{equation}
+Multivariate kernel density estimation (KDE see @Parzen1962-gt;
+@Scott2015-vl and references therein) is a fundamental tool for
+exploratory data analysis and unsupervised learning. Applications of KDE
+include finding hot spots of traffic networks in the GIS environment
+{[}@Xie2008-eb; @Okabe2009-nb{]}, automatic detection in visual
+surveillance systems {[}@Elgammal2002-cw{]}, wind power density
+detection {[}@Jeon2012-ac{]} and prime prediction via Twitter messages
+{[}@Gerber2014-tq{]}. For \(m\)-dimensional data where \(m\) is even of
+moderate size, KDE is challenging due to the well-known curse of
+dimensionality. However, the manifold hypothesis suggests that many
+high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\)
+where \(d\ll m\). The problem of KDE on manifolds has been well studied
+in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da,
+and @Berenfeld2022-fd. However, in these settings, it is assumed that
+the underlying manifold is known, which is rarely true in practice. The
+is only a small literature dealing with the case where the manifold is
+unknown. Recently, @berenfeld2021-den, @berenfeld2022-est and
+@divol2022measure demonstrated the minimax optimality of certain types
+of KDE on manifolds. In contrast, our objective in this paper is to
+propose a new KDE for data that lie on some unknown manifold found using
+embedding techniques.
+
+When dealing with data that lie on a manifold, it is common to use
+dimension-reduction techniques. These algorithms, which include ISOMAP,
+LLE, t-SNE, and UMAP among others, can be thought of as a mapping
+\(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold
+\(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\), where the dimension
+\(d\) is either assumed to be known a priori or needs to be estimated
+separately. Details of these algorithms are summarized in
+@Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh. One
+could easily apply KDE to the output vectors \(\bm{y}\). However, this
+would be a misleading representation of the density of points as they
+lie on the manifold. To see how, consider that the density on the
+manifold must be defined with respect to an appropriate volume form,
+that is \[Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,\]
+where \(f\) is the probability density function on the manifold,
+\(\bm{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on
+the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a
+change of variables, this would imply the following density for
+\(\bm{y}\), \begin{equation}
 \label{eq:changevar}
-Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
-\end{equation}
-where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\).
-It can be seen that the density in terms of \(\bm{y}\), differs from the density on the manifold due to the term \(|\bm{H}(\bm{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant.
-Since \(\psi\) will typically be a non-linear function, the value of \(|\bm{H}(\bm{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm.
-Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\bm{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.
-
-In this paper, we propose a new KDE.
-Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm.
-For this reason, we refer to it as the distortion-corrected KDE (DC-KDE).
-Critical to our estimator, is obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point.
-One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density.
-In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator.
-The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used.
-The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.
-
-In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring points according to the magnitude of the density estimate.
-In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\bm{y}\) is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together.
-This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.
-Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images.
-In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities.
-Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.
-
-The rest of the paper is organized as follows.
-In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds.
-Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm.
-\autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space.
-\autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage.
+Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\det \bm{H}(\bm{y})|^{1/2}d\bm{y},
+\end{equation} where \(\psi(\mathcal{A})\) is the image of
+\(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian
+metric expressed in local coordinates given by the mapping \(\psi\)
+{[}@doCarmo1992-rm{]}. It can be seen that the density in terms of
+\(\bm{y}\), differs from the density on the manifold due to the term
+\(|\bm{H}(\bm{y})|^{1/2}\) which can be considered roughly analogous to
+the Jacobian determinant. Since \(\psi\) will typically be a non-linear
+function, the value of \(|\bm{H}(\bm{y})|^{1/2}\) will change across the
+manifold, in a way that measures the distortion induced by the dimension
+reduction algorithm. Notably, different choices of dimension reduction
+algorithm, imply different \(\psi\), meaning that if KDE is applied to
+the output \(\bm{y}\) coming from two different dimension reduction
+algorithms, this could lead to two very different sets of conclusions
+about the density of the same point.
+
+In this paper, we propose a new KDE. Our estimator is also based on the
+output of dimension reduction algorithms but accounts for the distortion
+induced by the algorithm. For this reason, we refer to it as the
+distortion-corrected KDE (DC-KDE). Critical to our estimator, is
+obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we
+use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments
+any dimension reduction algorithm with an estimate of the Riemannian
+metric at each data point. One particular downstream task that we are
+interested in is anomaly detection, where anomalies are identified as
+points of low density. In a number of simulation studies, we demonstrate
+two benefits that arise when using our new estimator. The first is that
+using distortion correction leads to more accurate density estimates and
+better detection of anomalies compared to the case where distortion
+correction is not used. The second is that when the estimates are found
+using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the
+results regarding the density estimates and the detection of anomalies
+are robust only when distortion correction is applied.
+
+In terms of data visualization, our approach allows a two or
+three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring
+points according to the magnitude of the density estimate, where the
+dimension of \(\bm{y}\), \(d\), is set as 2 or 3 in the dimension
+reduction algorithms. In this way, the user is able to tell whether a
+region that has a cluster of points in the output space \(\bm{y}\) is
+truly a high-density region or if instead, this is an artifact of the
+dimension reduction algorithm `squeezing' points together. This is
+particularly pertinent in the case of t-SNE which can form clusters of
+points even when no such clusters exist in the underlying data.
+Furthermore, our estimator can be used for manifolds with elements that
+are a variety of objects, for instance, rotational frames or images. In
+our application we consider a probabilistic manifold whose elements are
+densities, meaning that we estimate a density of densities. Finally, our
+proposed estimator also takes the form of a KDE with variable bandwidth
+meaning that it also contributes to extensive literature on this topic
+{[}see Section 6.6 of @Scott2015-vl for a summary{]}.
+
+The rest of the paper is organized as follows. In \autoref{kderm}, we
+present our distortion-corrected kernel density estimator for Riemannian
+manifolds. Our estimator combines insights from two existing papers, the
+first is @Pelletier2005-vu who propose a KDE for data on manifolds that
+are known, the second is @Perrault-Joncas2013-pq who propose the Learn
+Metric algorithm which we use to estimate the distortion induced by a
+dimension reduction algorithm. \autoref{simulation} is composed of two
+simulations; the first deals with 2-dimensional data from a Gaussian
+mixture model mapped onto a 3-dimensional `twin peaks' manifold and the
+second with a 5-dimensional semi-hypersphere data embedded into a
+100-dimensional space. \autoref{dckdeapplication} contains an
+application to an Irish smart meter dataset that visualizes and
+identifies households with an anomalous distribution of energy usage.
 Conclusions and discussions are presented in \autoref{dckdeconclusion}.
-Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
+Some additional background on important terms in Riemannian geometry
+required to understand our proposed method is provided in
+\autoref{riemgeo}.
 
 \hypertarget{kderm}{%
-\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}}
-
-In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding.
-Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold.
-In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates.
-By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}.
-To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
-
-In the following, we denote \(M\) as the \(d\)-dimensional manifold from which our data are sampled.
-Points on this manifold are denoted \(\bm{p}\) in general, with \(\bm{p}_1,\dots,\bm{p}_n\) denoting the observed sample.
-Often \(\bm{p}_i\) will be high-dimensional vectors such that \(\bm{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case.
-For instance, \(\bm{p}_i\) may be probability distributions on a statistical manifold.
-The methods we propose for estimating the density at each \(\bm{p}_i\) only require some sense of distance between the `input' points, \(d(\bm{p}_i,\bm{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\bm{y}_1,\dots,\bm{y}_n\), where \(\bm{y}_i\in\mathbb{R}^d\).
-We will denote this embedding as \(\bm{y}_i=\psi(\bm{p}_i)\).
-Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
-\begin{equation}
+\section{Distortion Corrected Kernel density estimate on Riemannian
+manifolds}\label{kderm}}
+
+In this section, we introduce our method for kernel density estimation
+on manifolds that uses an embedding from a dimension reduction algorithm
+while correcting for the distortion induced by this embedding. Since
+some readers may be unfamiliar with the nuances of manifolds, we first
+discuss kernel density estimation for data in Euclidean space, then
+illustrate in \autoref{Pellet} how this generalizes to the estimator of
+@Pelletier2005-vu, which can be applied when the data lie on some known
+manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm
+of @Perrault-Joncas2013-pq, which augments an embedding derived from a
+dimension reduction algorithm with an estimate of the Riemannian metric
+expressed in local coordinates. By combining elements from the work of
+@Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel
+distortion-corrected kernel density estimate in \autoref{DCKDE}. To keep
+this section as succinct as possible, we do not define concepts such as
+manifolds, charts, geodesic distance, etc., but provide this information
+for readers unfamiliar with differential geometry in \autoref{riemgeo}.
+
+In the following, we denote \(M\) as the \(d\)-dimensional manifold from
+which our data are sampled. Points on this manifold are denoted
+\(\bm{p}\) in general, with \(\bm{p}_1,\dots,\bm{p}_n\) denoting the
+observed sample. Often \(\bm{p}_i\) will be high-dimensional vectors
+such that \(\bm{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this
+need not be the case. For instance, \(\bm{p}_i\) may be probability
+distributions on a statistical manifold. The methods we propose for
+estimating the density at each \(\bm{p}_i\) only require some sense of
+distance between the `input' points, \(d(\bm{p}_i,\bm{p}_j)\), such that
+we can apply dimension reduction algorithms to obtain an `output'
+embedding, \(\bm{y}_1,\dots,\bm{y}_n\), where
+\(\bm{y}_i\in\mathbb{R}^d\). We will denote this embedding as
+\(\bm{y}_i=\psi(\bm{p}_i)\). Finally, we denote by \(\lambda\) the
+Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual
+Euclidean norm and following @Pelletier2005-vu, we make the following
+assumptions about the kernel function
+\(K:\mathbb{R}_+\rightarrow\mathbb{R}\), \begin{equation}
 \begin{aligned}
@@
 
-Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation.
-For instance, condition (iv) requires the support of the kernel to be bounded.
-The reasons for this will become clearer when we discuss the manifold setting in more detail.
-Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise.
-In our empirical section, more general kernel functions can be, and are, employed.
-
-For data \(\bm{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the usual kernel density estimator at a point \(\bm{y}\) is given by
+Note that these conditions are different from (and in some cases
+stricter than) those normally used for kernel density estimation. For
+instance, condition (iv) requires the support of the kernel to be
+bounded. The reasons for this will become clearer when we discuss the
+manifold setting in more detail. Also, for illustration purposes, in
+this section, we pay particular attention to the uniform kernel for
+which \(K(z)\) equals one if \(0\leq z\leq1\) and zero otherwise. In our
+empirical section, more general kernel functions can be, and are,
+employed.
+
+For data \(\bm{y}_i\in\mathbb{R}^d\) with \(i=1,\dots,N\) and assuming a
+bandwidth matrix \(r\bm{I}\) where \(r\) is a global bandwidth, then the
+usual kernel density estimator at a point \(\bm{y}\) is given by
 \begin{equation}
@@
 The intuition behind this estimator is very clear for a uniform kernel.
-The density at a point \(\bm{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\bm{y}\), times a term that ensures the density integrates to 1.
-In general, the bandwidth matrix need not be proportional to the identity matrix.
-However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\bm{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance.
-For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein.
-Kernel density estimators of this form can and have been applied directly on the output embedding \(\bm{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.
+The density at a point \(\bm{y}\) is equal to the proportion of sample
+points that lie within a ball of radius \(r\) centered at \(\bm{y}\),
+times a term that ensures the density integrates to 1. In general, the
+bandwidth matrix need not be proportional to the identity matrix.
+However, the intuition remains the same, only that the ball of radius
+\(r\) centered at \(\bm{y}\) is found with respect to Mahalanobis
+distance rather than the usual Euclidean distance. For more on kernel
+density estimation in the Euclidean case, see @Scott2015-vl and
+references therein. Kernel density estimators of this form can and have
+been applied directly on the output embedding \(\bm{y}\), and we will
+consider this approach as a benchmark in \autoref{simulation}.
 
@@
 
-For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
-\begin{equation}
+For kernel density estimation on a known manifold, @Pelletier2005-vu
+propose the following estimator, \begin{equation}
 \label{eq:denriem}
@@
 
-where \(d_g(\bm{p}, \bm{p}_i)\) denotes the geodesic distance between two points on the manifold \(\bm{p}\) and \(\bm{p}_i\) and \(\theta_{\bm{p}_i}(\bm{p})\) is known as the volume density function.
-The intuition behind the term \(K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)\) is relatively clear.
-For example, for a uniform kernel, the estimator at point \(\bm{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\bm{p}\).
-However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance.
-An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold.
-A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix.
-For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\bm{p}\) is so large that some points `fall inside' the ball more than once.
-For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere.
-This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.
-
-The inclusion of the volume density function \(\theta_{\bm{p}_i}(\bm{p})\) is perhaps not as immediately clear, therefore, before providing formal details, we will briefly discuss the intuition behind the inclusion of this term.
-We have already highlighted that when using a uniform kernel, the kernel density estimate at a point \(\bm{p}\) directly depends on the proportion of sample points within a ball of radius \(r\) around \(\bm{p}\).
-However, the volume of this ball must also be taken into account.
-In Euclidean space with the usual Lebesgue measure, a radius \(r\) ball will always have the same volume regardless of its center.
-The same does not hold for manifolds and therefore the volume density function must be included in the KDE to ensure that the estimated density integrates to one.
+where \(d_g(\bm{p}, \bm{p}_i)\) denotes the geodesic distance between
+two points on the manifold \(\bm{p}\) and \(\bm{p}_i\) and
+\(\theta_{\bm{p}_i}(\bm{p})\) is known as the volume density function.
+The intuition behind the term
+\(K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)\) is relatively clear.
+For example, for a uniform kernel, the estimator at point \(\bm{p}\)
+will still depend on the proportion of sample points within a ball of
+radius \(r\) centered at \(\bm{p}\). However, in this case, the geodesic
+distance on the manifold is used, rather than Euclidean or Mahalanobis
+distance. An additional technical assumption is that \(r\) is less than
+the injectivity radius of the manifold. A definition of the injectivity
+radius is given by @Chavel2006-mp and also provided in the appendix. For
+our purposes, it is sufficient to note that this assumption precludes
+the possibility that the radius of a ball around \(\bm{p}\) is so large
+that some points `fall inside' the ball more than once. For example on a
+sphere, a ball with a radius greater than half the circumference of a
+great circle will wrap back around the sphere. This phenomenon also
+explains why the kernel function must be bounded for density estimation
+on manifolds.
+
+The inclusion of the volume density function
+\(\theta_{\bm{p}_i}(\bm{p})\) is perhaps not as immediately clear,
+therefore, before providing formal details, we will briefly discuss the
+intuition behind the inclusion of this term. We have already highlighted
+that when using a uniform kernel, the kernel density estimate at a point
+\(\bm{p}\) directly depends on the proportion of sample points within a
+ball of radius \(r\) around \(\bm{p}\). However, the volume of this ball
+must also be taken into account. In Euclidean space with the usual
+Lebesgue measure, a radius \(r\) ball will always have the same volume
+regardless of its center. The same does not hold for manifolds and
+therefore the volume density function must be included in the KDE to
+ensure that the estimated density integrates to one.
 
 More formally, the volume density function can be explained as follows.
-Consider the exponential map around \(\bm{p}\), given by \(exp_{\bm{p}}(\bm{q})\), mapping vectors in the tangent space, \(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).
-Loosely, \(\bm{v}\) `points' in the direction of the geodesic between \(\bm{p}\) and \(\bm{q}\) and traveling along this geodesic at uniform speed \(\|\bm{v}\|\) takes place in one unit of time.
-Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\bm{p}\), via the inverse of the exponential map, to these \(\bm{v}\) vectors, expressed in some local coordinate system.
-The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system.
-In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\bm{H}(\bm{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\bm{p}_i\).
-For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.
+Consider the exponential map around \(\bm{p}\), given by
+\(exp_{\bm{p}}(\bm{v})\), mapping vectors in the tangent space,
+\(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).
+Loosely, \(\bm{v}\) `points' in the direction of the geodesic between
+\(\bm{p}\) and \(\bm{q}\) and traveling along this geodesic at uniform
+speed \(\|\bm{v}\|\) takes place in one unit of time. Now, consider a
+chart \(\varphi\) mapping points in the neighborhood of \(\bm{p}\), via
+the inverse of the exponential map, to these \(\bm{v}\) vectors,
+expressed in some local coordinate system. The volume density function
+is then the square root of the determinant of the Riemannian metric
+expressed in this coordinate system. In proving that the estimator
+integrates to 1, the volume density function cancels with the term
+\(|\bm{H}(\bm{y})|^{1/2}\) in Equation @ref(eq:changevar), where
+\(\psi\) is the logarithmic map around \(\bm{p}_i\). For more on the
+volume density function, see @Gallot2004-rc and @Le\_Brigant2019-lj.
 
@@
 
-To be able to apply the estimator of \textcite{Pelletier2005-vu} to the case where the manifold is not known, but where coordinates \(\bm{y}_i\) for \(i=1,\dots,n\) are obtained from a dimension reduction algorithm, requires an estimate of the Riemannian metric.
-Formally, the Riemannian metric is a symmetric and positive definite tensor field that defines an inner product on the tangent space \(T_{\bm{p}} M\) for every point \(\bm{p} \in M\).
-Using the coordinates from a dimension reduction algorithm \(\bm{y}\), angles, distances, and volumes in this Euclidean `output space' are not the same as on the manifold since dimension reduction algorithms introduce distortions.
-To alleviate this issue, \textcite{Perrault-Joncas2013-pq} propose a method to augment \(\bm{y}_1,\bm{y}_2\dots,\bm{y}_n\) with \(d\times d\) positive definite matrices, \(\bm{H}_1,\bm{H}_2\dots,\bm{H}_n\), at each data point.
-These matrices estimate the Riemannian metric in local coordinates defined by the dimension reduction algorithm.
-
-While full details are provided in \textcite{Perrault-Joncas2013-pq}, we briefly describe the Learn Metric algorithm here.
-There are four main steps in the algorithm.
-First, a weighted neighborhood graph is constructed, with edges between \(\bm{p}_i\) and \(\bm{p}_j\) when \(\bm{p}_i\) is a K-nearest neighbor of \(\bm{p}_j\) or vice versa, and edge weights depending on the distance between \(\bm{p}_i\) and \(\bm{p}_j\) on the manifold.
-Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\varepsilon\) is the radius parameter for the nearest neighbor graph.
-Third, a dimension reduction method is applied to obtain the output embedding \(\bm{y}_1,\dots,\bm{y}_n\).
-Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator).
-Full details on these four steps are provided in \autoref{alg:learnmetric}.
-This algorithm is implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz} although our own results are based on a re-implementation of the algorithm in \emph{R}.
-Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as suggested in the \emph{megaman} library.
-
-As pointed out by \textcite{Perrault-Joncas2013-pq}, dimension reduction can be carried out such that the dimension of the output vectors is larger than the intrinsic manifold dimension \(d\).
-In this case, the ranks of the matrices \(\bm{H}_i\) are equal to \(d\).
-Using a larger embedding dimension is justified since it is in general not possible to embed a manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean space.
-In our simulated examples, we abstract from this issue by constructing examples that can be globally embedded into \(d\)-dimesional Euclidean space.
-In practice, to determine the dimension of the manifold, the \emph{two-nearest neighbor estimator (TWO-NN estimator)} \autocite{Facco2017-rl,Denti2021-jl} can be used (See Chapter 3 of \textcite{Lee2007-wq} for more about intrinsic dimension estimation).
-The \emph{R} library \emph{intRinsic} \autocite{Denti2021-qc} implements this algorithm and is used in all examples involving real data where the intrinsic dimension is unknown.
+To be able to apply the estimator of @Pelletier2005-vu to the case where
+the manifold is not known, but where coordinates \(\bm{y}_i\) for
+\(i=1,\dots,n\) are obtained from a dimension reduction algorithm,
+requires an estimate of the Riemannian metric. Formally, the Riemannian
+metric is a symmetric and positive definite tensor field that defines an
+inner product on the tangent space \(T_{\bm{p}} M\) for every point
+\(\bm{p} \in M\). Using the coordinates from a dimension reduction
+algorithm \(\bm{y}\), angles, distances, and volumes in this Euclidean
+`output space' are not the same as on the manifold since dimension
+reduction algorithms introduce distortions. To alleviate this issue,
+@Perrault-Joncas2013-pq propose a method to augment
+\(\bm{y}_1,\bm{y}_2\dots,\bm{y}_n\) with \(d\times d\) positive definite
+matrices, \(\bm{H}_1,\bm{H}_2\dots,\bm{H}_n\), at each data point. These
+matrices estimate the Riemannian metric in local coordinates defined by
+the dimension reduction algorithm.
+
+While full details are provided in @Perrault-Joncas2013-pq, we briefly
+describe the Learn Metric algorithm here. There are four main steps in
+the algorithm. First, a weighted neighborhood graph is constructed, with
+edges between \(\bm{p}_i\) and \(\bm{p}_j\) when \(\bm{p}_i\) is a
+fixed-radius nearest neighbor of \(\bm{p}_j\) or vice versa, and edge
+weights depending on the distance between \(\bm{p}_i\) and \(\bm{p}_j\)
+on the manifold. The fixed-radius method considers all points within a
+specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to
+varying densities in the data and to capture local structure more
+effectively. Second, the discrete Laplacian on this graph
+\(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated {[}@Zhou2011-za{]},
+where \(\sqrt{\varepsilon}\) is the radius parameter in the previous
+step and a constant value \(c=0.25\) is used for the use of heat kernel
+in the weighted neighborhood graph. Third, a dimension reduction method
+is applied to obtain the output embedding \(\bm{y}_1,\dots,\bm{y}_n\).
+Fourth, the Riemannian metric at each point is estimated by exploiting
+the connection between the Riemannian metric and the Laplace Beltrami
+operator (to which the graph Laplacian at Step 2 is a discrete
+estimator). Full details on these four steps are provided in
+\autoref{alg:learnmetric}. This algorithm is implemented in a Python
+library \emph{megaman} {[}@McQueen2016-xz{]} although our own results
+are based on a re-implementation of the algorithm in \emph{R}. For
+computational efficiency, approximate nearest neighbor searching methods
+are implemented to construct the neighborhood graph {[}@Cheng2021-dh;
+@Perrault-Joncas2013-pq{]}, which are then used in both steps 1 and 3.
+Two parameters, \(c=0.25\) and \(\sqrt{\varepsilon} = 0.4\), are set as
+suggested in the \emph{megaman} library.
+
+As pointed out by @Perrault-Joncas2013-pq, dimension reduction can be
+carried out such that the dimension of the output vectors is larger than
+the intrinsic manifold dimension \(d\). In this case, the ranks of the
+matrices \(\bm{H}_i\) are equal to \(d\). Using a larger embedding
+dimension is justified since it is in general not possible to embed a
+manifold of dimension \(d\) globally into \(d\)-dimensional Euclidean
+space. In our simulated examples, we abstract from this issue by
+constructing examples that can be globally embedded into
+\(d\)-dimesional Euclidean space. In practice, to determine the
+dimension of the manifold, the \emph{two-nearest neighbor estimator
+(TWO-NN estimator)} {[}@Facco2017-rl; @Denti2021-jl{]} can be used (See
+Chapter 3 of @Lee2007-wq for more about intrinsic dimension estimation).
+The \emph{R} library \emph{intRinsic} {[}@Denti2021-qc{]} implements
+this algorithm and is used in all examples involving real data where the
+intrinsic dimension is unknown.
 
@@
 
-With all fundamentals introduced, we can now give our novel Distortion Corrected KDE (DC-KDE) as
-\begin{equation}
+With all fundamentals introduced, we can now give our novel Distortion
+Corrected KDE (DC-KDE) as \begin{equation}
 \label{eq:denestimator}
@@
 
-The estimator has a similar structure to Equation \eqref{eq:denriem} with some key differences.
-To understand these differences, it is first critical to appreciate that the coordinates \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\) give an embedding that is approximately isometric in a small neighborhood around the \(i^{th}\) observed point (this insight is discussed at length in Section 6.2 of \textcite{Perrault-Joncas2013-pq}).
-This is crucial for two reasons.
-First, this implies that the term \(\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|\) approximates the geodesic distance between \(\bm{y}_i\) and \(\bm{y}_j\).
-Second, the estimator in Equation \eqref{eq:denriem} is valid only when the coordinate mapping is the logarithmic map around \(\bm{y}_i\), and it is this mapping that is approximated by \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\).
-For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\) which ensures that the embedding approximates the logarithmic map.
-Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation \autocite{Breiman1977-qc,Jones1990-oe,Terrell1992-ut}.
-
-One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian \(\bm{H}_j\).
-To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead.
-We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density.
-The entire workflow is summarized in \autoref{fig:vkde}.
-The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\bm{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue.
-With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
+The estimator has a similar structure to Equation @ref(eq:denriem) with
+some key differences. To understand these differences, it is first
+critical to appreciate that the coordinates
+\(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\) give an embedding that is
+approximately isometric in a small neighborhood around the \(i^{th}\)
+observed point (this insight is discussed at length in Section 6.2 of
+@Perrault-Joncas2013-pq). This is crucial for two reasons. First, this
+implies that the term \(\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|\)
+approximates the geodesic distance between \(\bm{y}_i\) and
+\(\bm{y}_j\). Second, the estimator in Equation @ref(eq:denriem) is
+valid only when the coordinate mapping is the logarithmic map around
+\(\bm{y}_i\), and it is this mapping that is approximated by
+\(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\). For this reason, there is a
+ratio of two determinants to ensure the density integrates to one, the
+first is a consequence of the mapping from the manifold to the
+coordinate system (from a dimension reduction algorithm), while the
+second is the transformation \(\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\)
+which ensures that the embedding approximates the logarithmic map. Also
+worth noting is the resemblance between the estimator and multivariate
+variable bandwidth estimation {[}@Breiman1977-qc; @Jones1990-oe;
+@Terrell1992-ut{]}.
+
+One limitation of the kernel density estimator is that the density can
+be estimated only at points where data have been observed since the
+estimator requires the Riemannian \(\bm{H}_j\). To estimate the density
+at points that do not correspond to observed data, any smoothed average
+of nearest neighbors can be used instead. We note that the particular
+downstream task that we are interested in is anomaly detection for which
+only the density estimates at observed sample points are required since
+anomalies are identified as the points with the lowest density. The
+entire workflow is summarized in \autoref{fig:vkde}. The last two steps
+in \autoref{fig:vkde} are our main contributions, generating
+distortion-corrected KDE with adaptive Riemannian metric \(\bm{H}_i\) at
+each point and computing the highest density region plots based on the
+density estimates for anomaly detection. Compared to the anomaly
+detection with a general kernel density estimator in @Cheng2021-dh, the
+changes are also highlighted in blue. With this anomaly detection
+process, outliers based on the lowest densities could be detected more
+accurately regardless of the distortion in manifold learning.
+
+(ref:vkdecaption) The proposed schematic for anomaly detection with
+distortion-corrected kernel density estimates.
 
@@
 
-\caption{The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.}\label{fig:vkde}
+\caption{(ref:vkdecaption)}\label{fig:vkde}
 \end{figure}
@@
 
-In this section, we examine two scenarios for both low and high dimensions to test our proposed distortion-corrected KDE.
-For visualization purposes, \autoref{twodgaussian} presents an example of a two-dimensional manifold embedded in 3-dimensional ambient Euclidean space.
-As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 4-dimensional manifold embedded in a 100-dimensional ambient space.
-To estimate the density, we use the dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP.
-In general, we aim to highlight two advantages of our proposed distortion-corrected KDE compared to KDE applied directly to the output coordinates.
-First, the density estimates are closer to the ground truth when distortion correction is used, and as a consequence, distortion correction is more adept at detecting anomalies.
-Second, we show how density estimation and anomaly detection are more robust to a different choice of dimension reduction method when distortion correction is used.
+In this section, we examine two scenarios for both low and high
+dimensions to test our proposed distortion-corrected KDE. For
+visualization purposes, \autoref{twodgaussian} presents an example of a
+two-dimensional manifold embedded in 3-dimensional ambient Euclidean
+space. As a high-dimensional example, the second simulation in
+\autoref{fivedgaussian} is based on a 4-dimensional manifold embedded in
+a 100-dimensional ambient space. To estimate the density, we use the
+dimension reduction algorithms ISOMAP, LLE, Laplacian Eigenmaps, t-SNE,
+and UMAP. In general, we aim to highlight two advantages of our proposed
+distortion-corrected KDE compared to KDE applied directly to the output
+coordinates. First, the density estimates are closer to the ground truth
+when distortion correction is used, and as a consequence, distortion
+correction is more adept at detecting anomalies. Second, we show how
+density estimation and anomaly detection are more robust to a different
+choice of dimension reduction method when distortion correction is used.
 
@@
 
-The simulation setup for the twin peaks example is to first generate vectors \(\bm{v}_1,\dots,\bm{v}_N\) for \(N=2000\) from a 2-dimensional Gaussian mixture model.
-The mixture has four components with different means
-\(\bm{\mu_1}=(0.25, 0.25)^\prime, \bm{\mu_2}=(0.25, 0.75)^\prime, \bm{\mu_3}=(0.75, 0.25)^\prime, \bm{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\bm{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\).
-The mixture proportions
-are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two-dimensional data in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the following
-\begin{equation}
+The simulation setup for the twin peaks example is to first generate
+vectors \(\bm{v}_1,\dots,\bm{v}_N\) for \(N=2000\) from a 2-dimensional
+Gaussian mixture model. The mixture has four components with different
+means
+\(\bm{\mu_1}=(0.25, 0.25)^\prime, \bm{\mu_2}=(0.25, 0.75)^\prime, \bm{\mu_3}=(0.75, 0.25)^\prime, \bm{\mu_4}=(0.75, 0.75)^\prime\)
+and the same variance-covariance matrix
+\(\bm{\Sigma}_i=diag(0.016, 0.016), i=1,2,3,4\). The mixture proportions
+are equally set as \(\pi_i=0.25, i=1,2,3,4\). The two-dimensional data
+in \autoref{fig:metadensity} is mapped to a `twin peaks' surface via the
+following \begin{equation}
 \label{eq:twinpeak}
@@
 
-The three-dimensional twin peaks mapping is shown in \autoref{fig:twinpeaks3d}.
-The colors in both \autoref{fig:metadensity} and \autoref{fig:twinpeaks3d} indicate the true density of the data via the twin peaks mapping, with lower density points in darker colors scattered in the outer as well as center areas.
+The three-dimensional twin peaks mapping is shown in
+\autoref{fig:twinpeaks3d}. The colors in both \autoref{fig:metadensity}
+and \autoref{fig:twinpeaks3d} indicate the true density of the data via
+the twin peaks mapping, with lower density points in darker colors
+scattered in the outer as well as center areas.
 
@@
 
+(ref:twinpeaks3dcaption) Scatterplot of the 3-d twin peaks data with the
+same colors indicating the true density as in \autoref{fig:metadensity}.
+
 \begin{figure}
@@
 
-\caption{Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.}\label{fig:twinpeaks3d}
-\end{figure}
-
-It is important to note that the \emph{true density} on the manifold is not simply a Gaussian mixture, since the mapping in Equation \eqref{eq:twinpeak} distorts the distribution.
-To recover the true distribution requires the correct Jacobian term for the pushforward from \(\bm{v}\) to the volume form of twin peaks manifold.
-By treating the \(\bm{v}\) as an `output' embedding from input points \(\bm{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\bm{\Gamma}_i\) for \(i=1,\dots,n\) where \(\bm{\Gamma}_i\) is the Riemannian of the coordinate system given by \(\bm{v}\).
-This notation is distinct from \(\bm{H}_i\) which is the output of the Learn Metric algorithm for a coordinate system obtained via a dimension reduction algorithm.
-The true density on the manifold can be obtained as \(f(\bm{p}_i)=f(\bm{v}_i)|\bm{\Gamma}_i|^{1/2}\), where \(f(\bm{v}_i)\) is the density of a four-component mixture of normals.
-Knowledge of \(\bm{v}\) and \(\bm{\Gamma_i}\) will not be used when estimating the density but only to establish a `ground truth' for densities on the manifold.
-\autoref{fig:metadensity} shows the simulated \(\bm{v}\) with color indicating the true density of data on the manifold.
-Anomalies are defined as points with the lowest densities shown in darker colors and with `typical' points having the highest density shown in yellow.
-The anomalies are found around the edges of the plot, but there is also a low-density region between the means of the four mixture components.
-The objective is to determine whether we can correctly identify these features without any knowledge of the true density or the \(\bm{v}\).
+\caption{(ref:twinpeaks3dcaption)}\label{fig:twinpeaks3d}
+\end{figure}
+
+It is important to note that the \emph{true density} on the manifold is
+not simply a Gaussian mixture, since the mapping in Equation
+@ref(eq:twinpeak) distorts the distribution. To recover the true
+distribution requires the correct Jacobian term for the pushforward from
+\(\bm{v}\) to the volume form of twin peaks manifold. By treating the
+\(\bm{v}\) as an `output' embedding from input points \(\bm{x}\) that
+lie on the true manifold and applying the Learn Metric algorithm, we can
+obtain \(\bm{\Gamma}_i\) for \(i=1,\dots,n\) where \(\bm{\Gamma}_i\) is
+the Riemannian metric of the coordinate system given by \(\bm{v}\). This
+notation is distinct from \(\bm{H}_i\) which is the output of the Learn
+Metric algorithm for a coordinate system obtained via a dimension
+reduction algorithm. The true density on the manifold can be obtained as
+\(f(\bm{p}_i)=f(\bm{v}_i)|\bm{\Gamma}_i|^{1/2}\), where \(f(\bm{v}_i)\)
+is the density of a four-component mixture of normals. Knowledge of
+\(\bm{v}\) and \(\bm{\Gamma_i}\) will not be used when estimating the
+density but only to establish a `ground truth' for densities on the
+manifold. \autoref{fig:metadensity} shows the simulated \(\bm{v}\) with
+color indicating the true density of data on the manifold. Anomalies are
+defined as points with the lowest densities shown in darker colors and
+with `typical' points having the highest density shown in yellow. The
+anomalies are found around the edges of the plot, but there is also a
+low-density region between the means of the four mixture components. The
+objective is to determine whether we can correctly identify these
+features without any knowledge of the true density or the \(\bm{v}\).
 
@@
 
-\autoref{fig:tpoutliers} summarizes the results.
-Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively.
-We set the bandwidth parameter in the DC-KDE \eqref{eq:denestimator} as \(r=0.5\).
-Colors show the different estimated densities at each point with anomalies shown in black with blue indexing, and higher density points shown in yellow.
-For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
-On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.
-The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies.
-For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes.
-This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data.
+\autoref{fig:tpoutliers} summarizes the results. Each row of panels
+corresponds to a different dimension reduction technique, while the
+left, center and right columns correspond to density estimates for the
+ground truth density, distortion-corrected KDE, and KDE respectively. We
+set the bandwidth parameter in the DC-KDE @ref(eq:denestimator) as
+\(r=0.5\). Colors show the different estimated densities at each point
+with anomalies shown in black with blue indexing, and higher density
+points shown in yellow. For many methods, the salient features of the
+ground truth distribution are clear regardless of whether distortion
+correction is applied, for example for ISOMAP, all three plots, identify
+a similar set of outliers and four high-density regions. On the other
+hand for LLE, the left panel shows that dimension reduction pulls
+outliers on the manifold towards the center. The distortion-corrected
+KDE can account for this, while KDE without distortion correction on the
+other hand does not correctly identify the anomalies. For t-SNE, the
+ground truth and distortion-corrected KDE identify four regions of high
+density, while a KDE estimate without distortion correction seems to
+identify a larger number of modes. This concurs with the common
+observation that t-SNE tends to output clusters even where such clusters
+may not be present in the underlying data {[}@cai2022-tsne{]}.
 
@@
 
-We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from KDE with and without density correction by \autoref{tab:tpcors}.
-Distortion correction improves the rank correlation for all dimension reduction algorithms.
-In particular, while applying KDE to the output of t-SNE and UMAP leads to a moderate correlation below 0.5, applying distortion correction improves these rank correlations to values close to 0.8.
+We can gain further insight by comparing the correlation between ranks
+of true densities and estimated densities from KDE with and without
+density correction by \autoref{tab:tpcors}. Distortion correction
+improves the rank correlation for all dimension reduction algorithms. In
+particular, while applying KDE to the output of t-SNE and UMAP leads to
+a moderate correlation below 0.5, applying distortion correction
+improves these rank correlations to values close to 0.8.
 
@@
 
-In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction.
-Data are presented on a log scale to highlight anomalies.
-The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a \(99\%\) highest density region.
-The middle shaded region contains anomalies that are true positives in the sense of not lying in a \(90\%\) HDR.
-Points lying outside squares (shown as triangles) are incorrectly classified.
-For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used).
-Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
-\autoref{fig:tptsneden} shows the same plot but for t-SNE.
-The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified.
-However, the difference between KDE with and without distortion correction is stark.
-These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.
+In \autoref{fig:tpisomapden}, we plot ranks of the estimated density
+against the true density for the ISOMAP embedding with the left panels
+showing results for distortion correction and the right panel showing
+results without distortion correction. Data are presented on a log scale
+to highlight anomalies. The bottom left shaded region contains all
+points that are truly anomalies and are identified as such (true
+positives), where an anomaly is defined as a point not falling within a
+\(99\%\) highest density region. The middle shaded region contains
+anomalies that are true positives in the sense of not lying in a
+\(90\%\) HDR. Points lying outside squares (shown as triangles) are
+incorrectly classified. For example, the three red triangles in the
+middle left of the left panel are truly anomalies since they lie outside
+the \(99\%\) HDR, but are not classified as such (although they are true
+positives if a 90\% HDR is used). Overall, the right panel contains many
+more misclassified anomalies, which shows that failing to apply
+distortion correction can have a severe impact on anomaly detection.
+\autoref{fig:tptsneden} shows the same plot but for t-SNE. The quality
+of t-SNE is worse than ISOMAP in this example therefore many more
+anomalies are misclassified. However, the difference between KDE with
+and without distortion correction is stark. These results highlight the
+importance of applying distortion correction especially when the quality
+of dimension reduction may not be high.
 
@@
 
-Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of distortion correction methods to the use of a dimension reduction algorithm.
-Each row of panels compares ranks of the estimated densities based on a dimension reduction algorithm to the estimated density based on ISOMAP.
-The left column shows results when distortion correction is applied, and the right column when it is not applied.
-It can be seen that the rank correlation between estimates based on different dimension reduction algorithms is much higher when distortion correction is applied.
-This is critical since conclusions will be more robust to the choice of dimension reduction algorithm.
+Finally, \autoref{fig:tpisomapvs4ml} demonstrates the robustness of
+distortion correction methods to the use of a dimension reduction
+algorithm. Each row of panels compares ranks of the estimated densities
+based on a dimension reduction algorithm to the estimated density based
+on ISOMAP. The left column shows results when distortion correction is
+applied, and the right column when it is not applied. It can be seen
+that the rank correlation between estimates based on different dimension
+reduction algorithms is much higher when distortion correction is
+applied. This is critical since conclusions will be more robust to the
+choice of dimension reduction algorithm.
 
 \hypertarget{fivedgaussian}{%
-\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}}
-
-As a high-dimensional experiment, we generate the underlying data from a 5-dimensional semi-hypersphere, embedded within 100-dimensional ambient space.
-To start with, we simulate vectors \((\bm{v}_1, \dots, \bm{v}_N)^\prime\) for \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)\) and \(\mathcal{N}(\bm{\mu}_2, \bm{\Sigma}_2)\), with the same means \(\bm{\mu}_1 = \bm{\mu}_2 =(0, 0, 0, 0)^\prime\) and different variance-covariance matrices \(\bm{\Sigma}_1 = diag(1,1,1,1)\) and \(\bm{\Sigma}_2 = diag(2,2,2,2)\).
-The mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
-With this design, the observations from the second component tend to be outlying anomalies.
-The data are mapped to a hemisphere via the equation
-\(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and \(r\) is set as \(8\).
-\autoref{fig:fivedmeta} shows a scatterplot which is a single frame from a 5-D tour path\footnote{See the animation of the 5-D grand tour at \url{https://github.com/ffancheng/kderm/blob/master/monash/figures/tourr_5d_animation.gif}.} animation using the R package \emph{tourr} \autocite{Wickham2011-ir}.
-The round and triangular point shapes indicate the two mixture components \(\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)\) and
-\(\mathcal{N}(\bm{\mu}_2, \bm{\sigma}_2)\), and the colors indicate the distance between each point and the center of the distribution.
-It can be seen that the most distant points are in darker colors and triangular shapes, meaning that the most anomalous observations are generated from the second mixture component.
+\subsection{Semi-hypersphere example embedded in 100-D
+space}\label{fivedgaussian}}
+
+As a high-dimensional experiment, we generate the underlying data from a
+5-dimensional semi-hypersphere, embedded within 100-dimensional ambient
+space. To start with, we simulate vectors
+\((\bm{v}_1, \dots, \bm{v}_N)^\prime\) for \(N=10,000\) points from a
+4-dimensional Gaussian mixture model with two mixture components,
+\(\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)\) and
+\(\mathcal{N}(\bm{\mu}_2, \bm{\Sigma}_2)\), with the same means
+\(\bm{\mu}_1 = \bm{\mu}_2 =(0, 0, 0, 0)^\prime\) and different
+variance-covariance matrices \(\bm{\Sigma}_1 = diag(1,1,1,1)\) and
+\(\bm{\Sigma}_2 = diag(2,2,2,2)\). The mixture proportions are set as
+\(\pi_1=0.99\) and \(\pi_2=0.01\). With this design, the observations
+from the second component tend to be outlying anomalies. The data are
+mapped to a hemisphere via the equation
+\(v_1^2 + v_2^2 + v_3^2 + v_4^2 + v_5^2 = r^2\) where \(v_5>0\) and
+\(r\) is set as \(8\). \autoref{fig:fivedmeta} shows a scatterplot which
+is a single frame from a 5-D tour path\footnote{See the animation of the
+  5-D grand tour at
+  \url{https://github.com/ffancheng/kderm/blob/master/monash/figures/tourr_5d_animation.gif}.}
+animation using the R package \emph{tourr} {[}@Wickham2011-ir{]}. The
+round and triangular point shapes indicate the two mixture components
+\(\mathcal{N}(\bm{\mu}_1, \bm{\Sigma}_1)\) and
+\(\mathcal{N}(\bm{\mu}_2, \bm{\sigma}_2)\), and the colors indicate the
+distance between each point and the center of the distribution. It can
+be seen that the most distant points are in darker colors and triangular
+shapes, meaning that the most anomalous observations are generated from
+the second mixture component.
 
@@
 
-To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero columns to \(\bm{v}_i\) so that \(\bm{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\).
-Next, we rotate the \(100\)-dimensional vectors \((\bm{v}_1, \dots, \bm{v}_N)^\prime\) by multiplying by a randomly generated rotation matrix.
-To generate the rotation matrix we first simulate elements from a uniform \((0,1)\) distribution, stack them into a \(100\times 100\) matrix \(\bm{A}\) and then take the R matrix from the QR decomposition of \(\bm{A}\).
-Rotating the vectors results in input vectors that are no longer sparse.
-Nonetheless, the intrinsic dimension of this is still \(d=4\).
+To embed the 5-D hyper-semisphere into 100-D space, we append 95 zero
+columns to \(\bm{v}_i\) so that
+\(\bm{v}_i=(v_1, \dots, v_5, 0, \dots, 0), i=1,\dots,N\). Next, we
+rotate the \(100\)-dimensional vectors
+\((\bm{v}_1, \dots, \bm{v}_N)^\prime\) by multiplying by a randomly
+generated rotation matrix. To generate the rotation matrix we first
+simulate elements from a uniform \((0,1)\) distribution, stack them into
+a \(100\times 100\) matrix \(\bm{A}\) and then take the R matrix from
+the QR decomposition of \(\bm{A}\). Rotating the vectors results in
+input vectors that are no longer sparse. Nonetheless, the intrinsic
+dimension of this is still \(d=4\).
 
@@
 
-Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density.
-The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example.
-In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densities from DC-KDE and KDE are presented.
-For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96.
-However, for Laplacian Eigenmaps and UMAP, more distortion is induced through dimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied.
-In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively, when distortion correction is applied.
-
-\autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies.
-As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the center of the distribution with outliers shown in darker colors.
-The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion-corrected KDE and panels on the right showing KDE without distortion correction.
-The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps, and UMAP.
-Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.
-
-Comparing the left and right panels for each row, we notice that KDE with distortion correction has fewer misclassified observations, and therefore outperforms KDE.
-For UMAP, almost all ground truth anomalies (points outside a \(99\%\) HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion.
-For DC-KDE in the left panel there are several triangles in a horizontal line, this arises, due to many points having an estimated density of close to zero (values less than \(10^{-7}\)) which leads to tied ranks.
-These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.
+Following similar steps to \autoref{twodgaussian}, we estimate the
+densities of the 4-D manifold and compare them with the ground truth
+density. The bandwidth parameter for the DC-KDE estimator is set as
+\(r=1\) for this example. In \autoref{tab:fivedcors}, the rank
+correlations between true densities and estimated densities from DC-KDE
+and KDE are presented. For ISOMAP and LLE, KDE without distortion
+correction has a slightly higher rank correlation with the true density
+than for KDE with distortion correction, but these differences are
+negligible and both density estimators have a very high correlation of
+more than 0.96. However, for Laplacian Eigenmaps and UMAP, more
+distortion is induced through dimension reduction, the rank correlation
+between estimated and true densities is close to 0 when distortion
+correction is not applied. In the same settings, the corresponding rank
+correlations are relatively high at 0.87 and 0.78 for Laplacian
+Eigenmaps and UMAP respectively, when distortion correction is applied.
+
+\autoref{fig:fivedisomapden} is constructed in a similar fashion to
+\autoref{fig:tpisomapden} to show the effect of distortion correction on
+the detection of anomalies. As in \autoref{fig:fivedmeta}, the point
+shapes indicate which of the two mixture Gaussian mixture components an
+observation was generated from and the colors indicate the distance from
+the center of the distribution with outliers shown in darker colors. The
+ranks of the estimated densities are shown against ranks of true
+densities (on the log scale) with panels on the left showing
+distortion-corrected KDE and panels on the right showing KDE without
+distortion correction. The panels from top to bottom show four dimension
+reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps, and UMAP. Note
+that we exclude t-SNE algorithm in this section because it is designed
+mainly for low-dimensional visualization purposes and is only applicable
+to an embedding dimension less than or equal to three.
+
+Comparing the left and right panels for each row, we notice that KDE
+with distortion correction has fewer misclassified observations, and
+therefore outperforms KDE. For UMAP, almost all ground truth anomalies
+(points outside a \(99\%\) HDR) are not correctly detected using KDE,
+while almost all anomalies are correctly detected after correcting for
+distortion. For DC-KDE in the left panel there are several triangles in
+a horizontal line, this arises, due to many points having an estimated
+density of close to zero (values less than \(10^{-7}\)) which leads to
+tied ranks. These low-density points are all detected as anomalies for
+all dimension reduction algorithms, which again shows the robustness of
+DC-KDE.
 
@@
 
-\autoref{tab:fourdhdrtable} further illustrates the proportions of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions.
-Compared with KDE, the proportions are always higher when distortion correction methods are used.
-The proportions with distortion correction are very close for ISOMAP, LLE, and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\).
-This could be due to the severe distortion usually induced by the UMAP algorithm.
+\autoref{tab:fourdhdrtable} further illustrates the proportions of
+correctly classified anomalies in the \(>99\%\) and \(99\%\) highest
+density regions. Compared with KDE, the proportions are always higher
+when distortion correction methods are used. The proportions with
+distortion correction are very close for ISOMAP, LLE, and Laplacian
+Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP
+at around \(77\%\). This could be due to the severe distortion usually
+induced by the UMAP algorithm.
 
@@
 
-In this application, we use the smart meter data from the \emph{CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010} in Ireland \autocite{cer2012-data} between 14 July 2009 and 31 December 2010.
-The CER dataset\footnote{accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.} records the half-hourly electricity consumption of individual residential and commercial properties, but not including energy for cooling or heating systems.
-We selected the \(3,639\) residential data with no missing values during the data collection period for a total of \(535\) days.
-
-For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods of the week \autocite{Hyndman2018-ia}.
-This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions.
-\textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds.
-Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP for dimension reduction to obtain a \(2\)-dimensional embedding for kernel density estimation and anomaly detection.
-To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used.
-We use the highest density regions plot to visualize the density estimates.
-However, for this real data set with an unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.
-
-Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here.
-For each household, a discrete approximation of the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found.
-For any pair of households the total variation metric can be found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn Metric algorithm.
-The number of nearest neighbors in manifold learning \(K\) is set as \(100\) and the bandwidth parameter \(r=180\) to match the distances between the distributions of households.
+In this application, we use the smart meter data from the \emph{CER
+Smart Metering Project - Electricity Customer Behaviour Trial,
+2009-2010} in Ireland {[}@cer2012-data{]} between 14 July 2009 and 31
+December 2010. The CER dataset\footnote{accessed via the Irish Social
+  Science Data Archive - www.ucd.ie/issda.} records the half-hourly
+electricity consumption of individual residential and commercial
+properties, but not including energy for cooling or heating systems. We
+selected the \(3,639\) residential data with no missing values during
+the data collection period for a total of \(535\) days.
+
+For the electricity consumption data of residential individuals, it is
+worthwhile to explore the distribution of electricity demand rather than
+the raw consumption data, so as to study the usage patterns of different
+households or different periods of the week {[}@Hyndman2018-ia{]}. This
+can be considered as a case of dimension reduction on a statistical
+manifold, that is a manifold with elements that are probability
+distributions. @Cheng2021-dh propose estimators of the Total Variation
+Metric and Hellinger distance between distributions that can be used in
+a computationally practical manner for dimension reduction on
+statistical manifolds. Again, we use ISOMAP, LLE, Laplacian Eigenmaps,
+t-SNE, and UMAP for dimension reduction to obtain a \(2\)-dimensional
+embedding for kernel density estimation and anomaly detection. To this
+end, we compare the density estimates from KDE with and without
+distortion correction and show how robust the anomalies are to different
+dimension reduction algorithms, only when distortion correction is used.
+We use the highest density regions plot to visualize the density
+estimates. However, for this real data set with an unknown structure,
+the ground truth densities are unknown and it is not possible to tell
+which anomalies are the true ones.
+
+Although full details for data processing and dimension reduction are
+provided in @Cheng2021-dh, we briefly describe the process here. For
+each household, a discrete approximation of the distribution of
+electricity demand at each one of the \(336\) half-hourly periods of the
+week is found. For any pair of households the total variation metric can
+be found between the distributions corresponding to any half hour of the
+week, and summing over these gives a distance measure between the pair
+of households, subsequently used for dimension reduction algorithms and
+the Learn Metric algorithm. The number of nearest neighbors in manifold
+learning \(K\) is set as \(100\) and the bandwidth parameter \(r=180\)
+to match the distances between the distributions of households.
 
@@
 
-\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE.
-There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions.
-The black points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
-First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
-When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE.
-As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies.
-The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
-Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
-Without distortion correction, the density estimates are different not only in scale, but also show inconsistency in the rank of density estimates.
+\autoref{fig:electricityembedding} presents the highest density regions
+of the 2-dimensional embedding for each dimension reduction method,
+ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels
+showing the densities estimated with the distortion corrected KDE and
+the right panels showing those estimate with KDE. There are six density
+regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\),
+\(99\%\) and \(>99\%\), and darker colors indicate lower density
+regions. The black points indexed with blue household IDs are the top 20
+anomalies with the lowest densities. First of all, for the same
+electricity data, each dimension reduction method generates quite
+different embeddings, meaning that distortions exist in these
+embeddings. When comparing row-wise for the same embedding, it can be
+seen that ISOMAP and LLE give similar anomalies while the outliers are
+more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE
+compared with the KDE. As the anomalous IDs are not easy to distinguish
+in the plots, we list them in \autoref{tab:electricityoutliers} and
+highlight the IDs that are most identified as anomalies. The darker blue
+points are those households that are identified five times out of the
+ten cases shown in \autoref{fig:electricityembedding}, and lighter blue
+points are those identified four times. Again, KDE with distortion
+correction methods finds more anomalies in common regardless of the
+dimension reduction methods, compared with KDE where only LLE could find
+similar anomalies to other cases. Without distortion correction, the
+density estimates are different not only in scale, but also show
+inconsistency in the rank of density estimates.
 
@@
 
-Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households.
-To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE.
-The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}.
-\autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) where anomalies fall outside the \(>99\%\) and \(99\%\) HDRs respectively that are also identified as anomalies by different dimension reduction methods.
-Both DC-KDE and KDE are shown in the table.
-Apart from the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods.
-The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP.
-\autoref{fig:electricityrank} presents the visualization of density ranking in log scale.
-The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}.
-Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.
-The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}.
-The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
-It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.
+Unlike for the simulations where the ground truth densities are known,
+in the applications, we have no knowledge of the truly anomalous
+households. To explore the robustness of anomaly detection both with and
+without distortion correction, we assume, in this case, that the
+anomalies given by UMAP are the `true' ones and evaluate the performance
+of the other four dimension reduction methods, ISOMAP, LLE, Laplacian
+Eigenmaps and t-SNE. The results are summarized in
+\autoref{tab:electricityhdrpercent} which can be interpreted similarly
+to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent}
+presents the proportions of `true' outliers (given by UMAP) where
+anomalies fall outside the \(>99\%\) and \(99\%\) HDRs respectively that
+are also identified as anomalies by different dimension reduction
+methods. Both DC-KDE and KDE are shown in the table. Apart from the
+lower percentage of ISOMAP in \(>99\%\) HDRs, distortion-corrected KDE
+gives higher proportions of true outliers in all other methods. The
+improvement brought by distortion correction is much more pronounced in
+Laplacian Eigenmaps and t-SNE which would otherwise detect almost
+completely different anomalies to UMAP. \autoref{fig:electricityrank}
+presents the visualization of density ranking in log scale. The grey
+shading area covers the density rank from 1 to 20, which are the
+anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots
+show the robustness of density estimated with distortion correction
+regardless of the dimension reduction methods. The robustness of anomaly
+detection to different methods when using distortion correction is also
+shown in \autoref{tab:electricityoutliers}. The household indices shown
+in darker shades of blue are identified as anomalies the most with
+dimension reduction algorithms. It is notable that without applying
+distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not
+identify any of the same anomalies as ISOMAP and LLE.
 
@@
 
-We now further examine households 3161, 4546, and 7049, which appear as anomalies for at least three dimension reduction algorithms as well as Households 2396, which is the most typical household with the highest density estimate.
-Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}.
-Compared with ID 2396, the other three households show different usage patterns.
-ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh.
-As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household.
-The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days.
-It is likely that these two households are not occupied all the time during the trial and could be used as temporary accommodation (e.g.~through a holiday rental service such as Airbnb).
+We now further examine households 3161, 4546, and 7049, which appear as
+anomalies for at least three dimension reduction algorithms as well as
+Households 2396, which is the most typical household with the highest
+density estimate. Their half-hourly electricity demands across the
+entire trial period are plotted in \autoref{fig:electricityraw}.
+Compared with ID 2396, the other three households show different usage
+patterns. ID 3161 has much less electricity usage in summers, mostly
+below 1kWh, than in winters, above 3kWh. As this data set does not
+include the energy used for cooling or heating systems, it is possible
+that there are other large electronic appliances used in winter for this
+household. The other two anomalous IDs, 4546 and 7049, however, have
+nearly zero kWh electricity consumption for many days and some spikes
+for other days. It is likely that these two households are not occupied
+all the time during the trial and could be used as temporary
+accommodation (e.g.~through a holiday rental service such as Airbnb).
 
@@
 
-Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityraw}.
-The distribution of the typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
-This repeated pattern in a week window is also shown on some days of the week for household ID 3161 with small spikes at noon and midnight but not on Fridays and Saturdays.
-As for the bottom two households, the middle row in \autoref{fig:electricityhdrsvkde} shows a distribution with zero median and a \(75\%\) quantile of 1kWh from around 9 pm to 6 am, while the distributions for ID 7049 do not have an obvious pattern across different periods of the week.
-These findings show that distortion correction methods work well in estimating the densities of the electricity usage and detecting anomalous households in the smart meter data despite the various embeddings from different dimension reduction methods.
+Further insights could be gained by comparing the quantile region plots
+of electricity demand against the time of the week for the same typical
+or anomalous households in \autoref{fig:electricityraw}. The
+distribution of the typical households in the top panel has shown a
+repeated period-of-the-week usage pattern, with higher usage during
+mealtime on all seven days of the week and slightly higher usage for
+weekends. This repeated pattern in a week window is also shown on some
+days of the week for household ID 3161 with small spikes at noon and
+midnight but not on Fridays and Saturdays. As for the bottom two
+households, the middle row in \autoref{fig:electricityhdrsvkde} shows a
+distribution with zero median and a \(75\%\) quantile of 1kWh from
+around 9 pm to 6 am, while the distributions for ID 7049 do not have an
+obvious pattern across different periods of the week. These findings
+show that distortion correction methods work well in estimating the
+densities of the electricity usage and detecting anomalous households in
+the smart meter data despite the various embeddings from different
+dimension reduction methods.
 
@@
 
-In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities.
-Compared with KDE, our distortion-corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding.
-The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point.
-We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension reduction algorithm.
-
-As an empirical example, we explore the distributions of \(3,639\) households and \(336\) time periods of the week in the Irish smart meter data.
-Five manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D embedding, followed by density estimation with KDE and DC-KDE.
-Without the ground truth density, we assume that UMAP gives the true densities and compare both density estimation methods by looking at the distributions of the most typical households with the highest densities and three anomalous households that are identified most times.
-The distortion correction methods could identify the typical households with repeated time-of-the-week usage patterns, while the distributions of the outliers are anomalous in different ways.
-
-There are several open questions to be explored.
-The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature.
-For densities on unknown manifolds, \textcite{Berenfeld2021-as} also provide some insights on the assumptions of the kernel function in \eqref{eq:kernelcondition}.
-The second one is that the proposed density estimation method is only limited to the embedding points.
-Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds \autocite{Berry2017-rz} in the future.
-The last question is related to the quality of the embedding from dimension reduction methods.
-Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear.
-The embedding quality could be measured using one of the metrics discussed in the online supplementary material of \textcite{Cheng2021-dh}, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way.
-The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers.
-However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
+In this paper, we propose a novel distortion correction method to
+estimate the density of an embedding from manifold learning algorithms
+and further identify outliers based on the densities. Compared with KDE,
+our distortion-corrected KDE makes use of geometric information for each
+data point to correct the distortion induced by the embedding. The
+Riemannian metric is estimated with the Learn Metric algorithm to
+approximate the geodesic distance and volume density function locally at
+each point. We compare our proposed method with KDE by two simulation
+settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D
+manifold mapped in a 100-D ambient space, and show that DC-KDE could
+generate more accurate kernel density estimation is more robust to the
+choice of dimension reduction algorithm.
+
+As an empirical example, we explore the distributions of \(3,639\)
+households and \(336\) time periods of the week in the Irish smart meter
+data. Five manifold learning algorithms, including ISOMAP, LLE,
+Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get the 2-D
+embedding, followed by density estimation with KDE and DC-KDE. Without
+the ground truth density, we assume that UMAP gives the true densities
+and compare both density estimation methods by looking at the
+distributions of the most typical households with the highest densities
+and three anomalous households that are identified most times. The
+distortion correction methods could identify the typical households with
+repeated time-of-the-week usage patterns, while the distributions of the
+outliers are anomalous in different ways.
+
+There are several open questions to be explored. The first involves the
+selection of bandwidth parameters and kernel types for kernel density
+estimation, which has been widely explored in many KDE-related
+literature. For densities on unknown manifolds, @Berenfeld2021-as also
+provide some insights on the assumptions of the kernel function in
+@ref(eq:kernelcondition). The second one is that the proposed density
+estimation method is only limited to the embedding points. Although for
+the downstream task of anomaly detection, only density estimated on
+these points is needed, it would be worth exploring the out-of-sample
+density estimation and estimation on the boundary of manifolds
+{[}@Berry2017-rz{]} in the future. The last question is related to the
+quality of the embedding from dimension reduction methods. Although our
+distortion correction method is fairly robust to different choices of
+manifold learning methods, in certain cases, when the data structure is
+too complex and the distortion is too severe to correct, the
+quantitative relationship between embedding quality and density
+estimation accuracy is not immediately clear. The embedding quality
+could be measured using one of the metrics discussed in the online
+supplementary material of @Cheng2021-dh, but when the ground truth
+densities are unknown, which is usually the case with real-world data
+set, it is hard to tell whether the distortion is corrected in the right
+way. The density estimates on the edges of the whole data structure
+could also be explored because most outer area points tend to be
+detected as outliers. However, the outperformance of DC-KDE over KDE has
+been shown in the higher dimensional simulation data and the electricity
+usage data, which are more related to real-life data sets.
 
@@
 
-This research was supported in part by the Monash eResearch Centre and eSolutions-Research Support Services through the use of the MonARCH HPC Cluster.
-The first author acknowledges the financial support of the Monash Graduate Scholarship (MGS) and the Monash International Tuition Scholarship (MITS) at Monash University.
-
-\clearpage
-
-\appendix
-
-\hypertarget{riemgeo}{%
-\section{Notions about Riemannian geometry}\label{riemgeo}}
-
-In this appendix, we present some notions about the Riemannian geometry used in this paper.
-
-\hypertarget{differentiable-manifolds}{%
-\subsection{Differentiable manifolds}\label{differentiable-manifolds}}
-
-In topology, a \emph{homeomorphism} is a bijective map between two topological spaces that is continuous in both directions.
-A \emph{Hausdorff space} is a topological space where any two distinct points can be separated by disjoint neighborhoods.
-And a \(d\)-dimensional (topological) \emph{manifold} \(M\) is a connected Hausdorff space \((M, \mathcal{T}_M)\) where the neighborhood \(U\) for each point \(p\) is homeomorphic to an open subset \(V\) of the Euclidean space \(\mathbb{R}^d\).
-Such a homeomorphism \(\varphi: U \rightarrow V\) together with \(U\) gives a (coordinate) \emph{chart}, denoted as \((U, \varphi)\), with the corresponding local coordinates \((x^1(p),\dots, x^d(p)) := \varphi(p)\).
-Further, a collection of charts \(\{U_\alpha, \varphi_\alpha\}\) ranging over the manifold \(M\) is called an \emph{atlas}, denoted as \(\mathcal{A}\).
-
-The manifold \(M\) is a \emph{differentiable manifold} if there exists an atlas of \(M\), \(\{U_\alpha, \varphi_\alpha\}\), such that the \emph{transition maps} between any two charts,
-\[
-\varphi_\beta \circ \varphi_\alpha^{-1}: \varphi_\alpha(U_\alpha \cap U_\beta) \rightarrow \varphi_\beta(U_\alpha \cap U_\beta),
-\] are differentiable of class \(C^\infty\) (smooth).
-Let \(\varphi\) be an injective map: \(E \rightarrow \varphi(E)\). Then \(\varphi\) is an \emph{embedding} of \(E\) into \(M\) if and only if
-\(\varphi: E \rightarrow \varphi(E)\) is a homeomorphism, and \(\varphi(E)\) is called an embedded submanifold of \(M\) with the subspace topology.
-
-\hypertarget{tangent-vector-and-tangent-space}{%
-\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}}
-
-The tangent vector at point \(p\) can be intuitively viewed as the velocity of a curve passing through point \(p\) or as the directional derivatives at \(p\).
-Here we define the tangent vector via the velocity of curves.
-
-For any point \(p \in M\), let \(\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M\) and \(\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M\) be two smooth curves passing through \(p\), i.e.~\(\gamma_1(0) = \gamma_2(0) = p\).
-We say \(\gamma_1\) and \(\gamma_2\) are \emph{equivalent} if and only if there exists a chart \((U,\varphi)\) at \(p\) such that
-\[
-(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
-\]
-A \emph{tangent vector} to a manifold \(M\) at point \(p\), denoted as \(v_p\), is any equivalent class of the differentiable curves initialized at \(p\).
-The set of all tangent vectors at \(p\) defines the \emph{tangent space} of \(M\) at \(p\), denoted as \(T_pM\).
-The tangent space is a vector space of dimension \(d\), equal to the dimension of \(M\), and it does not depend on the chart \(\varphi\) locally at \(p\).
-The collection of all tangent spaces defines the \emph{tangent bundle}, \(TM = \cup_{p \in M}T_pM\).
-
-Tangent vectors can also be seen as the directional derivatives at \(p\). For a given coordinate chart \(\varphi=(x^1,\dots,x^d)\), the tangent vectors defining partial
-derivatives are denoted as \(\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)\), which defines a \emph{basis} of the tangent space.
-The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with the corresponding \emph{cotangent vectors} \(z_p: T^\star_pM \rightarrow \mathbb{R}^d\) and a basis denoted as \(dx^1(p),\dots,dx^d(p)\).
-
-\hypertarget{riemannian-metric-and-geodesic-distance}{%
-\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}}
-
-A Riemannian metric \(g_p\) defined on the tangent space \(T_pM\) at each point \(p\) is a local inner product \(T_pM \times T_pM \rightarrow \mathbb{R}\), where \(g_p\) is a \(d\times d\) symmetric positive definite matrix and varies smoothly at \(p\).
-Generally, we
-omit the subscript \(p\) and refer to \(g\) as the Riemannian metric.
-The inner product between two vectors \(u, v \in T_pM\) is written as \(\langle u, v \rangle_g = g_{ij}u^iv^j\) using the Einstein summation convention where implicit summation over all indices, \(\sum_{i,j} g_{ij}u^iv^j\), is assumed.
-A differentiable manifold \(M\) endowed with the Riemannian metric \(g\) on each tangent space \(T_pM\) is called a \emph{Riemannian manifold} \((M,g)\).
-
-The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{\langle u,v \rangle_g}\), and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
-It could also be used to define the line element \(dl^2 = g_{ij}dx^i dx^j\) and the volume element \(dV_g = \sqrt{\det(g)}dx^1 \dots dx^d\), where \((x^1,\dots,x^d)\) are the local coordinates of the chart \((U, \varphi)\).
-For a curve \(\gamma: I \rightarrow M\), the length of the curve is
-\[
-l(\gamma) = \sqrt{\int_0^1 \|\gamma^\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
-\]
-where \(\gamma(I) \subset U\). The volume of \(W \subset U\) is defined as
-\[
-Vol(W) = \int_W \sqrt{\det(g)}dx^1 \cdots dx^d,
-\]
-which is also called the \emph{Riemannian measure} on \(M\).
-
-The \emph{geodesics} of \(M\) are the smooth curves that locally join the points along the shortest path on the manifold.
-Intuitively, geodesics are the \emph{straightest possible curves} in a Riemannian manifold \autocite[Section 7.2.3 of][]{Nakahara2018-zs}.
-A curve \(\gamma: I \rightarrow M\) is a geodesic if for all indices \(i,j,k\), the second-order ordinary differential equation is satisfied,
-\[
-\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
-\]
-where \(\{x^i\}\) are the coordinates of the curve \(\gamma\) and \(\Gamma^i_{jk}\) is the \emph{Christoffel symbol} defined by
-\[
-\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k}
-+ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
-\]
-The geodesics have a constant speed with norm \(\| \gamma^\prime(t) \|\), and they are the local minimizers of the arc
-length functional \(l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}\) when the curves are defined over the interval \([0,1]\).
-The \emph{geodesic distance} \(d_g\) is the length of the shortest geodesic between two points on the manifold.
-For a point \(p \in M\), when the geodesic distance starting at \(p\) is not minimized, we call such set of points the \emph{cut locus} of \(p\), and the distance to the cut locus is the \emph{injectivity radius} at \(p \in M\).
-Therefore, the injectivity radius of the Riemannian manifold \((M,g)\), \(\textit{inj}_gM\), is the infimum of the injectivity radii over all points on the manifold.
-
-\hypertarget{exponential-map-and-logarithmic-map}{%
-\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}}
-
-Denote \(B(p, r) \subset M\) as an open ball centered at point \(p\) with radius \(r\).
-Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\), where \(exp_p\) is the \emph{exponential map} at point \(p\).
-The exponential map maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{\prime}(0)=u\), and \(\gamma(1)=exp_p(u)\).
-It is a differentiable bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}).
-Intuitively, the exponential map moves point \(p\) to an endpoint at speed \(u\) after covering the length of \(\|u\|\) along the geodesic in one time unit.
-
-The inverse of the exponential map is called the \emph{logarithm map}, denoted as \(\log_p(q):= \exp^{-1}_p(q)\), which gives the tangent vector to get from point \(p\) to \(q\) in one unit time.
-Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
-Then we could interpolate a geodesic \(\gamma\) between two points \(p\) and \(q\) with the exponential map and the logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).
-
-\hypertarget{pushforward-and-pullback-metric}{%
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
-
-Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
-Let \(\phi: M \rightarrow E\) be a smooth map between the Riemannian manifold \((M,g)\) to another smooth manifold \(E\).
-Then the differential of \(\phi\) at point \(p\) is a linear map \(d\phi_p: T_pM \rightarrow T_{\phi (p)}E\), which pushes the tangent vector \(u \in T_pM\) at point \(p\) forward to the tangent vector \(\phi_*u \in T_{\phi (p)}E\) at the mapping point \(\phi(p)\).
-The image of the tangent vector \(u \in T_pM\) under the differential \(d\phi_p\), denoted as \(d\phi_p u\) is called the pushforward of \(u\) by the map \(\phi\).
-Then pushforward metric \(h=\varphi_*g\) of the Riemannian metric \(g\) along \(\varphi\) is given by the inner product
-\[
-\langle \phi_*u,\phi_*v \rangle_{\varphi*g} = \langle d\phi_p \phi_*u, d\phi_p \phi_*v \rangle_{g}.
-\]
-
-The tangent vectors \(\phi_*u\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{\prime}(0)=u\),
-\[
-d\phi_p(\gamma^{\prime}(0)) = (\phi \circ \gamma)^\prime (0).
-\]
-Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).
-The linear map is called the pullback by \(\phi\) and is often denoted as \(\phi^*\).
-
-\hypertarget{twinpeaksappe}{%
-\section{Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
-
-This appendix contains the comparison plots for the density rank between DC-KDE and KDE using different manifold learning algorithms, similar to \autoref{fig:tpisomapden}, \autoref{fig:tptsneden}, and \autoref{fig:tpisomapvs4ml}.
-By comparing these plots, it could be concluded that DC-KDE could categorize the density ranks into the highest density regions more accurately than KDE.
-By correcting the distortion in different manifold learning embeddings, DC-KDE is more robust in identifying the lowest density regions, which are usually used to detect anomalies.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_lle_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of LLE embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tplleden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_le_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of Laplaxian Eigenmaps embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpleden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/Twin Peak2000_densityrank_comparison_umap_radius8_r0_5_logrank_rec_colprob_smallblocks3_crossfalse}
-
-}
-
-\caption{Scatterplot of true density and estimated density ranks of UMAP embedding for DC-KDE and KDE, with colors indicating the absolute rank errors weighted by the sum of true and estimated ranks.
-DC-KDE shows a strong linear positive relationship with a higher rank correlation compared to KDE.}\label{fig:tpumapden}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{figures/Twin Peak2000_density_compare_llevs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by LLE compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).
-The four colors and shapes represent the four gaussian kernels in the 2-D metadata.
-Outliers found by DC-KDE are more consistent regardless of the manifold learning embedding.}\label{fig:tpllevs4ml}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_levs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by Laplacian Eigemaps compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tplevs4ml}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_tsnevs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by t-SNE compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tptsnevs4ml}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{figures/Twin Peak2000_density_compare_umapvs4ml_radius8_r0_5_rank}
-
-}
-
-\caption{Comparison of outliers found by UMAP compared to the other four for DC-KDE (on the left panel) and KDE (on the right panel).}\label{fig:tpumapvs4ml}
-\end{figure}
-
-\clearpage
-
-  \printbibliography
+This research was supported in part by the Monash eResearch Centre and
+eSolutions-Research Support Services through the use of the MonARCH HPC
+Cluster. The first author acknowledges the financial support of the
+Monash Graduate Scholarship (MGS) and the Monash International Tuition
+Scholarship (MITS) at Monash University.
 

### 2024-09-11 jcgs/revision/reviewer1_response.tex
final changes

\documentclass{article}
\usepackage{geometry} % Set margins
\usepackage{setspace} % For double spacing
\usepackage{titlesec} % For section formatting
\usepackage{enumitem} % For list formatting
\usepackage{xcolor} % For highlighting text
\usepackage{amssymb,tabu,placeins,algorithmic} % natbib
\usepackage{mathtools,bm}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{graphicx}
\usepackage[style=authoryear]{biblatex}
\addbibresource{kderm.bib}

\renewcommand*{\finalnamedelim}{\addspace\&\space}
\renewcommand*{\finallistdelim}{\addspace\&\space}
\renewcommand*{\multicitedelim}{\addsemicolon\space}
\renewcommand*{\nameyeardelim}{\addcomma\space}

% Set page margins
\geometry{
    left=1in,
    right=1in,
    top=1in,
    bottom=1in
}

% Set section formatting
\titleformat{\section}
{\bfseries\Large}
{\thesection.}
{0.5em}
{}

% Set list formatting
\setlist{noitemsep}

% Set double spacing
\doublespacing
\setlength{\parskip}{1em}

% Define colors for highlighting
\definecolor{myblue}{RGB}{0,0,200}
\definecolor{myblack}{RGB}{0, 0, 0}

% Color algorithm
\newenvironment{algocolor}{%
   \setlength{\parindent}{0pt}
   \itshape
   \color{myblue}
}{}

\date{\vspace{-8ex}}

\begin{document}

\title{Response to Reviewer 1's \\ \textit{"Report on 'Distortion corrected kernel density estimation on Riemannian manifolds” by F.Cheng, R.J Hyndman \& A.Panagiotelis'} }
% \author{}
% \date{}

\maketitle

We would like to extend our appreciation to the reviewer for their detailed and insightful feedback on our manuscript titled "Distortion corrected kernel density estimator on Riemannian manifolds." Your comments have been invaluable in enhancing the clarity and depth of our work. We have carefully addressed each point raised and provide detailed responses below highlighted in blue text.

\section*{Review}

\textit{However, I believe that the quality of the paper could be improved by comparing in more detail the performances with other density estimation methods that do not rely on a first step being the estimation of an embedding. Indeed, while embedding and then correcting for the distortion is a good idea, if one is only interested in estimating the density then the additional embedding step may potentially introduce unnecessary bias/instability. I have detailed my points below.}

1. \textit{In page 2 lines \(20-27\) it is said that several methods for density estimation on manifolds are limited to the case of known manifolds. While this is true for most of the examples, this is not true for [2. In this example the manifold needs not to be known for the method to be tractable. However it is assumed that the observations are living in a small tubular neighbourhood around the manifold, not exactly on it. Furthermore, at this point in the paper it would be sensible to cite the works of Berenfeld \& Hoffman [1] (which is actually mentioned later on) and Divol [3]. These works show that several types of KDEs are actually minimax optimal in the context of density estimation on manifolds. The goal of the present being slightly different (estimating the density of the sample after embedding), the comparison with the aforementioned works would be relevant and would help the reader to understand the contribution of the present paper.}

   \textcolor{blue}{We have revised the manuscript to clarify the distinction between our approach and those that do not require a known manifold. Additionally, we have included citations to the works of \cite{berenfeld2021-den} and \cite{divol2022measure} to provide context on minimax optimality in density estimation on manifolds. In the introduction we now state:\\
   ``Recently, \cite{berenfeld2021-den}, \cite{berenfeld2022-est} and \cite{divol2022measure} demonstrated the minimax optimality of certain types of KDE on manifolds. In contrast, our objective in this paper is to propose a new KDE for data that lie on some unknown manifold found using embedding techniques.''
   }

2. \textit{Following the previous remark, when the data is embedded in a Euclidean space (such as in the simulated examples: twin peaks and the semihypersphere), it is possible (even if the manifold is unknown) to perform density estimation directly on the data (and not after estimating an embedding) with, for instance, KDEs [1, 3] or special Dirichlet process mixtures [2] while preserving optimal statistical convergence properties (at least theoretically). Would your (empirical) results stay competitive in this case?}

   \textcolor{blue}{While it is possible to perform KDE on the data without using an embedding, this is challenging for the high dimensional examples we consider. The popular \texttt{ks} package [\cite{Duong2007-up}] in R recommends that it not be used for dimensions greater than 6 due to issues with numerical instability. The \texttt{weird} package [\cite{hyndman2024-weird}] in R also implements multivariate kernel density estimation, however its function to find the bandwidth matrix was prohibitively slow even for a dimension of 5. The example we consider is 100-dimensional. In this case, we have coded a KDE by hand with a bandwidth proportional to an identity matrix, i.e. $\mathbf{H} = h^2 \mathbf{I}$. The scale factor $h$ is found using Silverman's rule adapted for multivariate data : $h = \left(\frac{4}{d+2}\right)^{\frac{1}{d+4}} n^{-\frac{1}{d+4}}$. In our semi-hypersphere example, $d=100$ and $n=10,000$. Note that due to the curse of dimensionality the KDE on 100-dimensional data is somewhat flat with the densities of all points close to zero. This can be shown by observing the distribution of the estimated densities evaluated at each observation. This bottom panel (yellow violin plot) of Figure \ref{fig:kde} shows the results for a 100-dimensional KDE without dimension reduction. In contrast to the true densities and the KDEs on emeddings, the 100-dimensional KDE estimate does not detect the high proportion of points with relatively lower densities.}

   \begin{figure}[!h]
   	\begin{center}
    \includegraphics[width=0.4\linewidth]{kde_compare_violin.jpeg}\end{center}
    \caption{Comparison of KDE methods for high-dimensional data.}
    \label{fig:kde}
    \end{figure}

   \textcolor{blue}{
   %Figure \ref{fig:kde} shows the comparison of the density estimates for the 100-D example with different density estimations. We use different violin plots to display the distribution of the density estimates, and they are, from left to right, the true density of the manifold, distortion-corrected KDE, KDE using ISOMAP embedding, and KDE on the 100-D data with $\mathbf{H} = h^2 \mathbf{I}$ as bandwidth matrix. By comparing with the true density distribution, both in the violin shape and the y-axis scale, we could tell that DC-KDE gives the closest density estimates to the true ones, and KDE on 100-D data performs very poorly with almost all 0 estimates. \\
   In high-dimensional spaces, the data sparsity leads to exponentially slower convergence rates and requires an impractically large sample size to achieve reliable density estimates. Computationally, bandwidth selection algorithms are extremely demanding, both in terms of processing time and memory requirements. All these factors combined make traditional KDE methods unsuitable for 100-dimensional data, and dimension reduction is essential as proposed here.
   }

3. \textit{During all of the experiments (both with simulated and real data), the choices of the parameters \(\varepsilon, r\) are not really discussed thoroughly, even though it is known that tuning them is usually quite hard. The problem for \(r\) is even not related to the manifold learning problem as it also appears in classical KDE on flat Euclidean spaces. I believe it would be beneficial to give some guidance on this matter or at least some observations.}

   \textcolor{blue}{We acknowledge the importance of parameter tuning and have added more details on this. Regarding the choice of \(\varepsilon\) we use as stated at the end of Section 2.2 the default choice of 0.4. Experimenting with alternative values did not lead to appreciably better results that the default choice, and we now state this clearly in the paper. Regarding the scale $r$ we similarly experimented with different values. Since there is only a single parameter this could be tuned with reference to the downstream task. For example, for anomaly detection, $r$ could be chosen to maximise the coherence between selecting anomalies both in and out of sample using cross validation. We have revised the manuscript to state this in final paragraph of the conclusion.
   }

\section*{Miscellaneous}

1. \textit{page 2 line 32: it should be made clear at this point that the dimension \(d\) is either considered known or needs to be estimated separately (i.e., the proposed method is not directly dimension agnostic).}

   \textcolor{blue}{
   We have clarified that the dimension \(d\) is assumed to be known as a priori or estimated separately, as our method is not dimension agnostic and $d$ is a crucial parameter. In the second paragraph of the paper we state (with the part in bold now added):\\
  ``These algorithms, which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\), \textbf{where the dimension $d$ is either assumed to be known a priori or needs to be estimated separately}.'' }

2. \textit{Equation (1): a reference to a relevant textbook/source for the formula would be useful for the inexperienced reader.}

   \textcolor{blue}{A book reference, Chapter 3 of "Riemannian Geometry" by \cite{doCarmo1992-rm}, has been added to guide readers to a relevant source for Equation (1). \\
   ``By a change of variables, this would imply the following density for \(\bm{y}\),
\begin{equation}
\label{eq:changevar}
Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\det \bm{H}(\bm{y})|^{1/2}d\bm{y},
\end{equation}
where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\) \autocite{doCarmo1992-rm}.''
   }

3. \textit{page 3 line 24: I think the author meant to write "Riemannian metric."}

   \textcolor{blue}{%
   We have corrected the missing "metric" in the specified line. \\
   ``Critical to our estimator, is obtaining an estimate of $|\bm{H}(\bm{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of \cite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian \textbf{metric} at each data point.''
   }

4. \textit{page 3 line 38: it is not clear how a 2 or 3-dimensional is constructed and is actually confusing in the context: is \(d=2\) or \(3?\) probably not but clarifications would be useful here.}

   \textcolor{blue}{
   Clarifications have been added to explain the construction of 2 or 3-dimensional spaces in order to visualize the output of dimension reduction, where \(d=2\) for a 2-dimensional scatterplot and \(d=3\) for a 3-dimensional scatterplot. We now state: \\
  ``In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring points according to the magnitude of the density estimate, where the dimension of \(\bm{y}\), \(d\), is set as 2 or 3 in the dimension reduction algorithms."
   }

5. \textit{page 7 lines \(36-42\): this is a bit confusing, at first the graph is described as being a k-NN graph but later a radius parameter is mentioned (which is usually a parameter used to define random geometric graphs, not k-NN ones). It would be beneficial to clarify this point/the construction of the graph.}

   \textcolor{blue}{
   Thank you for pointing this out. Either a K-nearest neighbor (KNN) or fixed radius method could be used. In the Learn Metric algorithm, the weighted neighborhood graph is constructed by using a fixed-radius nearest neighbor search instead of a KNN approach. The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\). This radius parameter allows the graph to adapt to varying densities in the data, capturing local structure more effectively. By using a fixed-radius approach, the algorithm ensures that the graph reflects the underlying geometry of the data manifold, which is crucial for the analysis. We now clarify this: \\
   ``First, a weighted neighborhood graph is constructed, with edges between \(\bm{p}_i\) and \(\bm{p}_j\) when \(\bm{p}_i\) is a fixed-radius nearest neighbor of \(\bm{p}_j\) or vice versa, and edge weights depending on the distance between \(\bm{p}_i\) and \(\bm{p}_j\) on the manifold. The fixed-radius method considers all points within a specified radius of \(\sqrt{\varepsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively. Second, the discrete Laplacian on this graph \(\hat{\mathcal{L}}_{\varepsilon,n}\) is estimated \autocite{Zhou2011-za}, where \(\sqrt{\varepsilon}\) is the radius parameter in the previous step and a constant value $c=0.25$ is used for the use of heat kernel in the weighted neighborhood graph.''
   }

6. \textit{page 9 /algorithm 1: at step 2 it looks like you are using the symmetric normalized graph Laplacian. Shouldn't it be \(\tilde{W}=D^{-1 / 2} W D^{-1 / 2}\) and not \(D^{-1} W D^{-1}\)? Also, the constant \(c\) in the definition of \(\tilde{\mathcal{L}}_{n, \varepsilon}\) is not defined, does the value matter?}

   \textcolor{blue}{Here we simply restate the algorithm as defined in Algorithm 1 of \cite{Perrault-Joncas2013-pq} who define \(\tilde{W}=D^{-\lambda} W D^{- \lambda}\), and then $\lambda=1$ in Step 2 of their Algorithm 3. We agree that by setting $\lambda=1/2$, we would be using the symmetric normalized graph Laplacian but for consistency with the original reference, set \(\tilde{W}=D^{-1} W D^{-1)} \) and use this in our computation. \\
   % We have corrected the definition of the symmetric normalized graph Laplacian and provided an explanation for the constant \(c\) in the definition of \(\tilde{\mathcal{L}}_{n, \varepsilon}\). \\
   Setting \(c=0.25\) is recommended when using heat kernel used in the weighted neighborhood graph $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\bm{x}_i-\bm{x}_j\|^2)$. The values for \(c\) is stated in Section 2.2, and we have provided an explanation for this choice now and included the value of \(c\) in Step 2 of Algorithm 1. We also note that since \(c\) and ${\varepsilon}$ only enter Algorithm 1 as a product, it is sufficient to tune only one of these parameters. This is now also discussed in the paper at the end of Section 2.2.
   }

7. \textit{page 9 /algorithm 1: it looks like the steps \(3 \& 4\) are completely independent. In general, in order to get an estimated embedding one can use techniques based on neighborhood/nearest neighbors graphs (like Laplacian Eigenmaps). If one does that and then uses the same graph to perform step 4 would your technique be "improved"? Have you tried that? A few words on this would be helpful.}

   \textcolor{blue}{Thanks for your suggestion. Yes, we did use the same nearest neighbor graph from Step 1 to get the estimated embedding. The computational efficiency is largely improved by doing this as the nearest neighbor searching is the most time-consuming step. For implementation, we used approximate nearest neighbor searching methods proposed in another paper \cite{Cheng2021-dh}. We have now added further explanation in Section 2.2. \\
  ``This algorithm is implemented in a Python library *megaman* (\cite{McQueen2016-xz}) although our own results are based on a re-implementation of the algorithm in \texttt{R}. For computational efficiency, approximate nearest neighbor searching methods are implemented to construct the neighborhood graph (\cite{Cheng2021-dh}, \cite{Perrault-Joncas2013-pq}), which are then used in both steps 1 and 3.''
   }

8. \textit{page 10 line 30: it is stated that for the estimation of the density at an unobserved point one can use "any smoothed average of nearest neighbors." Isn't there the same problem as in the original problem of density estimation at observed points? i.e., isn't there distortion issues as well when doing that?}

   \textcolor{blue}{The referee is correct to point this out. Indeed to prevent this distortion the average of nearest neighbors should be weighted by a term that depends on the the determinant of the Riemannian at each of the nearest neighbors. We now state explicitly in the paper:\\
   ``To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors.''}

9. \textit{page 13 line 53: can you point to a reference for the behavior of t-SNE in this context?}

   \textcolor{blue}{\autocite{cai2022-tsne} has been added here for discussion about the clustering behavior of t-SNE. t-SNE is designed to preserve local structure of the high-dimensional data very well, which means it emphasizes similarities between nearby points, which can lead to the formation of apparent clusters in the low-dimensional visualization. The paper mentions an ``early exaggeration'' stage in t-SNE that allows for easier movement of clusters in the early stages of the algorithm. This can contribute to the formation of spurious clusters.}

10. \textit{page 15 line 48: a definition/short reminder on ranks and their definition would be beneficial (possibly in appendix?).}

   \textcolor{blue}{
   Thanks for your suggestion. We have included a brief definition of density rank in the Appendix B for clarity. \\
   ``The density rank used in the paper is defined as the relative position or order of an item within a set from the lowest to the highest, based on its density value.''
   }

\section*{Conclusion}

\textit{We are enthusiastic about the method and believe this is a valuable contribution; the only major missing point being, in my opinion, a more complete comparison with other methods. I hope my remarks will help the authors to improve the quality of the paper.}

\vspace{0.5cm}

\textcolor{blue}{We are grateful for the reviewer's enthusiasm and constructive remarks. We have made the suggested improvements and believe they have significantly enhanced the quality of the paper. Thank you once again for your valuable feedback.}

% \bibliographystyle{plainnat}
% \bibliography{references}

\printbibliography

\end{document}

### 2024-09-11 jcgs/kderm_jcgs.tex
final changes

@@
 
-\section{Introduction}\label{dckdeintro}
+\hypertarget{dckdeintro}{%
+\section{Introduction}\label{dckdeintro}}
 
@@
 
-\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}
+\hypertarget{kderm}{%
+\section{Distortion Corrected Kernel density estimate on Riemannian manifolds}\label{kderm}}
 
@@
 
-\subsection{Kernel Density estimation on manifolds}\label{Pellet}
+\hypertarget{Pellet}{%
+\subsection{Kernel Density estimation on manifolds}\label{Pellet}}
 
@@
 
-\subsection{Riemannian metric estimation}\label{MetLearn}
+\hypertarget{MetLearn}{%
+\subsection{Riemannian metric estimation}\label{MetLearn}}
 
@@
 
-\subsection{Distortion corrected KDE}\label{DCKDE}
+\hypertarget{DCKDE}{%
+\subsection{Distortion corrected KDE}\label{DCKDE}}
 
@@
 
-\section{Simulations}\label{simulation}
+\hypertarget{simulation}{%
+\section{Simulations}\label{simulation}}
 
@@
 
-\subsection{Twin peaks example}\label{twodgaussian}
+\hypertarget{twodgaussian}{%
+\subsection{Twin peaks example}\label{twodgaussian}}
 
@@
 
-\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}
+\hypertarget{fivedgaussian}{%
+\subsection{Semi-hypersphere example embedded in 100-D space}\label{fivedgaussian}}
 
@@
 \begin{table}
-\centering
+
 \caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings of the 100-D data.
@@
 
-\section{Application}\label{dckdeapplication}
+\hypertarget{dckdeapplication}{%
+\section{Application}\label{dckdeapplication}}
 
@@
 
-\section{Conclusions}\label{dckdeconclusion}
+\hypertarget{dckdeconclusion}{%
+\section{Conclusions}\label{dckdeconclusion}}
 
@@
 
-\section*{Acknowledgment}\label{acknowledgment}
+\hypertarget{acknowledgment}{%
+\section*{Acknowledgment}\label{acknowledgment}}
 \addcontentsline{toc}{section}{Acknowledgment}
@@
 
-\section{Notions about Riemannian geometry}\label{riemgeo}
+\hypertarget{riemgeo}{%
+\section{Notions about Riemannian geometry}\label{riemgeo}}
 
@@
 
-\subsection{Differentiable manifolds}\label{differentiable-manifolds}
+\hypertarget{differentiable-manifolds}{%
+\subsection{Differentiable manifolds}\label{differentiable-manifolds}}
 
@@
 
-\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}
+\hypertarget{tangent-vector-and-tangent-space}{%
+\subsection{Tangent vector and tangent space}\label{tangent-vector-and-tangent-space}}
 
@@
 
-\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}
+\hypertarget{riemannian-metric-and-geodesic-distance}{%
+\subsection{Riemannian metric and geodesic distance}\label{riemannian-metric-and-geodesic-distance}}
 
@@
 
-\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}
+\hypertarget{exponential-map-and-logarithmic-map}{%
+\subsection{Exponential map and logarithmic map}\label{exponential-map-and-logarithmic-map}}
 
@@
 
-\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}
+\hypertarget{pushforward-and-pullback-metric}{%
+\subsection{Pushforward and pullback metric}\label{pushforward-and-pullback-metric}}
 
@@
 
-\section{Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}
+\hypertarget{twinpeaksappe}{%
+\section{Rank comparison plots for twin peaks mapping}\label{twinpeaksappe}}
 

### 2024-09-11 jcgs/kderm_jcgs.tex
Removed colors in text

--- 2024-09-10 jcgs/kderm_jcgs.tex
@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric \(\bm{H}_i\) at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted in blue shades.
+Compared to the anomaly detection with a general kernel density estimator in \textcite{Cheng2021-dh}, the changes are also highlighted with shading.
 With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 \caption{Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
-The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
+The top 20 outliers, highlighted as points and indexed with text labels, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
 DC-KDE finds more true outliers than KDE in all five rows.}\label{fig:tpoutliers}
@@
 We set the bandwidth parameter in the DC-KDE \eqref{eq:denestimator} as \(r=0.5\).
-Colors show the different estimated densities at each point with anomalies shown in darker colors with blue indexing text, and higher density points shown in lighter colors.
+The color shades show the different estimated densities at each point with anomalies shown in darker colors with indexing labels, and higher density points shown in lighter colors.
 For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
@@
 Points lying outside squares (shown as triangles) are incorrectly classified.
-For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used).
+For example, the four colored triangle points in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used).
 Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
@@
 \caption{Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityembedding}
+The highlighted points with ID indexing are the top 20 anomalous households found by each dimension reduction method in each row.}\label{fig:electricityembedding}
 \end{figure}
 \begin{table}
-
+\centering\centering
 \caption{\label{tab:electricityoutliers}Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding.
 The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
-The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
+The colored IDs are household IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.}
 \centering
-\resizebox{\linewidth}{!}{
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
 \begin{tabu} to \linewidth {>{}c>{}c>{}c>{}c>{}c>{}c|>{}c>{}c>{}c>{}c>{}c}
@@
 There are six density regions colored in the plot, \(1\%\), \(5\%\), \(50\%\), \(90\%\), \(99\%\) and \(>99\%\), and darker colors indicate lower density regions.
-The dark-colored points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
+The dark-colored points indexed with household IDs are the top 20 anomalies with the lowest densities.
 First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.
@@
 As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies.
-The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times.
+The darker colored points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter colored points are those identified four times.
 Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases.
@@
 The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}.
-The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms.
+The household indices shown in darker shades are identified as anomalies the most with dimension reduction algorithms.
 It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.

### 2024-09-11 jcgs/03-chap3-kderm.Rmd
Removed colors in text

@@
 The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection.
-Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue shades.
+Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted with shading.
 With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.
@@
 ```{r tpoutliers, fig.pos = "!p", fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row.
-The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
+The top 20 outliers, highlighted as points and indexed with text labels, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel).
 DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
@@
 We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$.
-Colors show the different estimated densities at each point with anomalies shown in darker colors with blue indexing text, and higher density points shown in lighter colors.
+The color shades show the different estimated densities at each point with anomalies shown in darker colors with indexing labels, and higher density points shown in lighter colors.
 For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions.
@@
 Points lying outside squares (shown as triangles) are incorrectly classified.
-For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the $99\%$ HDR, but are not classified as such (although they are true positives if a 90% HDR is used).
+For example, the four colored triangle points in the middle left of the left panel are truly anomalies since they lie outside the $99\%$ HDR, but are not classified as such (although they are true positives if a 90% HDR is used).
 Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection.
@@
 ```{r electricityembedding, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel).
-The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
+The highlighted points with ID indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
 knitr::include_graphics("figures/Electricity_2d_5levels_outliers_comparison_5ml_r180.png")
@@
 The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE.
-The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
+The colored IDs are household IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
                   col.names = c(" ", rep(c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"), 2))
@@
 There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions.
-The dark-colored points indexed with blue household IDs are the top 20 anomalies with the lowest densities.
+The dark-colored points indexed with household IDs are the top 20 anomalies with the lowest densities.
 First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings.

### 2024-09-11 R/Simulation2d_TwinPeaks.R
Final check code comments

@@ -1,3 +1,6 @@
+## -----------------------------------------------------------------------------
+## This script is for the similation for twin peaks data (KDE and DC-KDE)
+## -----------------------------------------------------------------------------
 rm(list = ls())
 library(tidyverse)
 library(dimRed)
@@ -1010,4 +1013,4 @@ p_isovs4ml <- ggplot(p_data, aes(x = !!sym(xname), y = value)) +
 # Spec and sensitivity for correctly categorized outliers, over a small area in the top left (Done using log10 instead)
 # Change plot of true density colors to 5 levels (Done)
 # Table of rank correlation (Done)
-# Table of embedding quality
+# Table of embedding quality (Removed)

### 2024-09-11 R/Simulation4d_03_plots.R

@@ -1,5 +1,5 @@
 ###--------------------------------------------------------
-## 1. Gather and plot the results from 4 scenarios in HPC results
+## 1. Gather and plot the results from 4 scenarios in HPC results: 4-D manifold embedding in 100-D space simulation
 ###--------------------------------------------------------
 rm(list = ls())
 library(tidyverse)
@@ -147,7 +147,7 @@ p_rank
 
 
 
-
+# Not run
 # ----KDE with ks::kde on 100-d data instead of 4-d embedding-------------------
 library(weird)
 library(tictoc)
@@ -473,8 +473,8 @@ frank_long <- frank %>%
 #   theme(legend.position = "none")
 
 
-library(ggplot2)
-library(viridis)  # For a nice color palette
+# Put the violin plot in the review paper
+library(viridis)
 # library(hrbrthemes)  # For a sleek theme
 
 frank_long |> 

### 2024-09-11 R/__Simulation4d_04_kde_vs_dckde.R
Final check code comments

@@ -1,0 +1,59 @@
+## -----------------------------------------------------------------------------
+## This script is for manifold learning and kernel density estimation (KDE and DC-KDE)
+## -----------------------------------------------------------------------------
+# Modify `method` as different manifold learning methods in embed()
+# `radius=10`
+# Tune `r`, the bandwidth parameter, for DC-KDE
+# One ggplot for rank plots, and two .rda files for ML and density estimates
+# This script is modified for HPC jobs for four ML methods
+rm(list = ls())
+library(tidyverse)
+library(dimRed)
+library(hdrcde)
+library(ggforce)
+library(ks)
+library(patchwork)
+Jmisc::sourceAll(here::here("R/sources"))
+set.seed(1234)
+# r <- 1
+# k <- 500
+load(paste0("data/simdata_100d_4dmanifold_N10000_trueden_k", k, ".rda"))
+
+
+# ----KDE with ks::kde on high-dimensional data instead of embedding------------
+library(weird)
+library(tictoc)
+
+# From Rob
+n <- 100
+d <- 3
+y <- matrix(rnorm(n*d), ncol = d)
+H <- kde_bandwidth(y)
+fhat <- kde(y, H)
+
+
+# set H=diag(d)
+# X: true manifold, train: 100d data for dimRed, N, p, den: density of 5d semishepre, trueden: true density of 4d manifold
+
+# KDE on true manifold X
+dim(X) # 10000*4
+dim(train) # 10000*100
+tic()
+H <- kde_bandwidth(X)
+toc() # 0.141s
+fhat <- kde(X, H)
+summary(fhat$estimate)
+summary(trueden)
+
+# set H as Identity matrix for 100d data
+H1 <- diag(ncol(train)) #+ diag(1e-8, ncol(train))
+tic()
+fhat1 <- kde(x = train, H = H1, binned = FALSE, eval.points = train)
+toc() # 20min
+summary(fhat1$estimate)
+summary(trueden)
+
+
+
+# ----Compare with true density and kde, dckde----------------------------------
+

### 2024-09-11 R/app.R

@@ -1,3 +1,4 @@
+# Shiny app for simulation
 # Note: When first deployed, the app will show Error in if: argument is of length zero for the plots location.
 # It comes from ml_outlier(x=sim_data()) where the input is reactive
 # Just ignore it and wait a few seconds and the plots will show up
@@ -5,7 +6,6 @@
 library(shiny)
 library(shinydashboard)
 library(plotly)
-library(maniTools)
 library(tidyverse)
 library(dimRed)
 library(viridis)
@@ -15,6 +15,9 @@ library(matrixcalc)
 library(ggforce)
 library(patchwork)
 # library(copula)
+# devtools::install_github("jefferis/RANN@master-L1")
+library(RANN.L1)
+library(maniTools)
 Jmisc::sourceAll(here::here("R/sources"))
 set.seed(1)
 

### 2024-11-27 jcgs/revision/reviewer2_response.tex
Journal version pdf

\documentclass{article}
\usepackage{geometry} % Set margins
\usepackage{setspace} % For double spacing
\usepackage{titlesec} % For section formatting
\usepackage{enumitem} % For list formatting
\usepackage{xcolor} % For highlighting text
\usepackage{amssymb,tabu,placeins,algorithmic} % natbib
\usepackage{mathtools,bm}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
% \usepackage{natbib}
\usepackage{lipsum}
\usepackage[style=authoryear]{biblatex}
\addbibresource{../kderm.bib}
\usepackage{hyperref}

\renewcommand*{\finalnamedelim}{\addspace\&\space}
\renewcommand*{\finallistdelim}{\addspace\&\space}
\renewcommand*{\multicitedelim}{\addsemicolon\space}

% Set page margins
\geometry{
    left=1in,
    right=1in,
    top=1in,
    bottom=1in
}

% Set section formatting
\titleformat{\section}
{\bfseries\Large}
{\thesection.}
{0.5em}
{}

% Set list formatting
\setlist{noitemsep}

% Set double spacing
\doublespacing
\setlength{\parskip}{1em}

% Define colors for highlighting
\definecolor{myblue}{RGB}{0,0,200}
\definecolor{myblack}{RGB}{0, 0, 0}

% Color algorithm
\newenvironment{algocolor}{%
   \setlength{\parindent}{0pt}
   \itshape
   \color{myblue}
}{}

\date{\vspace{-8ex}}

\begin{document}

\title{Response to Reviewer 2's \\ \textit{"Report on 'Distortion corrected kernel density estimator on Riemannian manifolds' by F. Cheng, R. J. Hyndman and A. Panagiotelis"} }
% \author{}
% \date{}

\maketitle

We would like to thank the reviewer for their thorough evaluation and constructive feedback on our manuscript. We have carefully considered all the comments and have made the necessary revisions to address them. Below, we provide detailed responses to each of the points raised by the reviewer and highlighted the responses in blue text.

\section*{General Comments}

\textit{It would be nice to include some short computation proving that the density estimator defined by formula (4) indeed integrates to one, if the \(H_{i}\) is the true metric matrix at point \(y_{i}\). Or give an "exact" counterpart of (4) that does integrate to one (is it (3)?)}

\textcolor{blue}{We appreciate the reviewer's positive feedback on our manuscript. In response to the suggestion, we clarify that Equation (3) serves as an ``exact'' counterpart of Equation (4). We now state:\\
	``The estimator has a similar structure to Equation (3) which can be seen as the counterpart to Equation (4) if the Riemannian is known exactly.''
}

\section*{Typos and Notational Consistency}

We are grateful for the reviewer's attention to details regarding typographical errors. We have addressed the following points:

1. \textit{p. 3 line 24: missing "metric"}

   \textcolor{blue}{%
   We have corrected the missing "metric" in the specified line. \\
   ``Critical to our estimator, is obtaining an estimate of $|\bm{H}(\bm{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian \textbf{metric} at each data point.''
   }

2. \textit{p. 6 line 50: it should be \(\exp _{p}(v)\) instead of \(\exp _{p}(q)\)}

   \textcolor{blue}{
   The notation has been corrected to \(\exp_{p}(v)\) as suggested. \\
   ``Consider the exponential map around \(\bm{p}\), given by \(exp_{\bm{p}}(\bm{v})\), mapping vectors in the tangent space, \(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).''
   }

3. \textit{p7, in reference "Brigant and Puechmorel", the name is Le Brigant and not Brigant}

   \textcolor{blue}{
   We have updated the reference in the bib file and the inline reference to "Le Brigant and Puechmorel" to reflect the correct name. \\
   ``For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.''}

4. \textit{In formula (1) the determinant seems to be denoted by \(|\cdot|\) while in formula (4) by det. Maybe homogenize the notations.}

   \textcolor{blue}{
   We have homogenized the notations for the determinant across the manuscript. The determinant is now consistently denoted by \(\det\) in all relevant formulas (1) and (4). \\
   \begin{equation}
    \label{eq:changevar}
    Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\det \bm{H}(\bm{y})|^{1/2}d\bm{y},
    \end{equation}
    \begin{equation}
    \label{eq:denestimator}
    \hat{f}(\bm{y}_j) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \bm{H}_j|}{|\det \bm{H}_i|} \bigg)^{1/2} K\bigg( \frac{\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|}{r} \bigg).
    \tag{4}
    \end{equation}
    }

5. \textit{p. 11 line 54: Riemannian metric ?}

   \textcolor{blue}{
   We have reviewed and corrected the use of "Riemannian metric" as suggested. \\
   ``By treating the \(\bm{v}\) as an `output' embedding from input points \(\bm{x}\) that lie on the true manifold and applying the Learn Metric algorithm, we can obtain \(\bm{\Gamma}_i\) for \(i=1,\dots,n\) where \(\bm{\Gamma}_i\) is the Riemannian \textbf{metric} of the coordinate system given by \(\bm{v}\).''
   }

\vspace{0.5cm}

\textcolor{blue}{We believe these changes have strengthened the manuscript, and we are grateful for the reviewer's insights. We hope the revised version meets the standards for publication. Thank you once again for your valuable feedback.}

\printbibliography

% \bibliographystyle{plainnat}
% \bibliography{kderm}

\end{document}

### 2024-11-27 reports/20210318Curves.Rmd
Journal version pdf

---
title: "Differencial Geometry: Curves"
# author:
# - familyname: Cheng
#   othernames: Fan
#   address: Monash University
#   email: Fan.Cheng@monash.edu
blind: true
cover: false
toc: false  
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache=FALSE, messages=FALSE, warning=FALSE, eval=TRUE, fig.pos = 'H')
# options(tinytex.verbose = TRUE)
```

```{r packages, include=FALSE, eval=FALSE}
library(tidyverse)
library(ggplot2)
library(knitr)
```

# Books

## Lee, J. M., Chow, B., Chu, S. C., Glickenstein, D., and Guenther, C. (2009), “Manifolds and differential geometry,” Topology. An International Journal of Mathematics, ams.org.

- Chap 1: Differentiable Manifolds

- Chap 2: The Tangent Structure

- Chap 3: Immersion and Submersion

- Chap 4: Curves and Hypersurfaces in Euclidean Space

- Chap 8: Differential Forms

- Chap 13: Riemannian and Semi-Riemannian Geometry

## Kass, R. E., and Vos, P. W. (2011), Geometrical Foundations of Asymptotic Inference, John Wiley & Sons.

Appendix C: Basic Concepts in Differential Geometry

- Manifolds

- Submanifolds

- The Tangent Space

- Connections

- Riemannian Metric

- Tensors

- Tubular Neighborhoods

- The Measure Determined by a Metric

- Curvature

# Online Tutorials

[YouTube tutorials on differential geometry](https://www.youtube.com/playlist?list=PLdgVBOaXkb9DJjk8V0-RkXTnD4ZXUOFsc)

- Curves

- Arc length and reparameterization

- Arc length as a parameter

- Closed curves and periodic curves

- Curvature: intuition adn derivation

## Curves

1) Level Curves: 2-D and 3-D set, $C$. The intersection of 2 surfaces in 3-D is a curve.

2) Parameterized Curves: a map $\gamma: t \rightarrow \mathbb{R}^n, t \in \mathbb{R}$.

3) Smooth function: infinite differentiable. All components of a parameterized curve $\gamma(t)$ are differentiable.

4) Tangent vector: the tangent vector of $\gamma(t)$ at the parameter value $t$ is $\frac{d\gamma}{dt}$.

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/curves/IMG_1264.PNG")
knitr::include_graphics("figures/curves/IMG_1265.PNG")
knitr::include_graphics("figures/curves/IMG_1267.PNG")
knitr::include_graphics("figures/curves/IMG_1268.PNG")
```

## Arc length and reparameterization

1) Arc length definition: the arc length of a curve $\gamma$ starting at the point $\gamma(\alpha)$ is the function $S(t)$ given by:
$$
S(t) = \int_\alpha^t \|\dot{\gamma}(u)\| du.
$$

2) Reparameterization: smooth bijective map exists such that the inverse map is also smooth.

3) Regular point, singular point and regular curve.

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/curves/IMG_1269.PNG")
knitr::include_graphics("figures/curves/IMG_1270.PNG")
knitr::include_graphics("figures/curves/IMG_1272.PNG")
knitr::include_graphics("figures/curves/IMG_1273.PNG")
```

## Arc length as a parameter

1-3) Proof of the reparameterization with regard to arc length is a valid option.

4) Theorem about unit speed curve.

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/curves/IMG_1274.PNG")
knitr::include_graphics("figures/curves/IMG_1275.PNG")
knitr::include_graphics("figures/curves/IMG_1276.PNG")
knitr::include_graphics("figures/curves/IMG_1277.PNG")
```

## Closed curves and periodic curves

1) T-periodic: $Y(t+T)=Y(t), T\in \mathbb{R}$.

2) Self-intersection of a curve.

3) Theorem: a unit-speed reparameterization of a regular closed curve $\gamma$ is always closed.

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/curves/IMG_1278.PNG")
knitr::include_graphics("figures/curves/IMG_1279.PNG")
knitr::include_graphics("figures/curves/IMG_1280.PNG")
```

## Curvature: intuition adn derivation

1) Curvature: the extend to which $\gamma$ deviates from a straight line. 
$$
K(s) = \|\frac{d^2\gamma}{ds^2}\|, 
$$
which shows how quickly the tangent vector changes.

2-4) Formula for curvature when $\gamma(t)$ is a regular curve with unit speed parameter $s$:
$$
k=\frac{\left\|\frac{d^{2} r}{d t^{2}} \times \frac{d r}{d t}\right\|}{\left\|\frac{d r}{d t}\right\|^{3}}
$$

4) Radius of curvature: $p=1/K$.

```{r , fig.align = 'center', out.width = "100%", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/curves/IMG_1282.PNG")
knitr::include_graphics("figures/curves/IMG_1283.PNG")
knitr::include_graphics("figures/curves/IMG_1284.PNG")
knitr::include_graphics("figures/curves/IMG_1285.PNG")
```

# Discussion

<!-- <!-- # References --> 
<!-- \setlength{\parindent}{-0.2in} -->
<!-- \setlength{\leftskip}{0.2in} -->
<!-- \setlength{\parskip}{8pt} -->
<!-- \vspace*{-0.2in} -->
<!-- \noindent -->

### 2024-11-27 reports/20210520_riem_metric_debug.Rmd
Journal version pdf

---
title: "Riemannian metric from manifold learning"
author: "Fan Cheng"
date: "13/05/2021"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE}
library(tidyverse)
library(dimRed)
# library(dtplyr)
library(reticulate)
library(viridis)
# library(ks)
library(hdrcde)
library(igraph)
# library(seurat)
library(matrixcalc)
library(akima)
Jmisc::sourceAll(here::here("R/sources"))
```

# Smart meter data for one household

```{r data}
# load(here::here("data/spdemand_3639id336tow.rda"))
# nid <- 1
# ntow <- 336
# 
# train <- spdemand %>%
#   lazy_dt() %>%
#   filter(tow <= ntow,
#          # id <= sort(unique(spdemand[,id]))[nid],
#          id == 1003) %>%
#   dplyr::select(-id, -tow) %>%
#   data.table::as.data.table()
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
(N <- nrow(train))
```

# Metric learning

```{r parameters}
# Parameters fixed
x <- train
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4
```

## Radius searching with k-d trees

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype)
summary(metric_isomap)
```
The function `metricML()` returns a list of

- the embedding coordinates `embedding` of \N \times \s
- the embedding metric `rmetric` for each point $p \in x$ as an array
- the weighted neighborhood graph as an `igraph` object `weighted_graph`
- the sparse adjacency matrix from the graph `adj_matrix`

## Embedding plots

```{r}
# Comparison of Isomap embedding plot
par(mfrow = c(1, 2))
plot(metric_isomap$embedding, col = viridis::viridis(24)) # metricML
# plot(e@data@data, col = viridis::viridis(24)) # embedding from dimRed, same as metricML()
emb_isomap <-
  feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather"))
plot(emb_isomap$`0`, emb_isomap$`1`, col = viridis::viridis(24)) # embedding from megaman, radius = 0.4
```

## Riemannian metric

```{r}
# Use adjacency matrix as input for metricML and dimRed, then the embedding should be close, as well as the Riemannian metric
metric_isomap$adj_matrix[1:10, 1:10]# renormalized adjacency matrix
metric_isomap$weighted_graph %>% is.connected()
riem_isomap <- metric_isomap$rmetric
riem_isomap[,,1] %>% 
  # isSymmetric()
  matrixcalc::is.positive.definite() # FALSE
```

**The Riemannian metric from the `metricML()` function is now positive definite for all points. **

Now compare it with the megaman output.

```{r, eval=TRUE}
np = import("numpy")
H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy"))
H_isomap[1,,, drop=TRUE] %>% 
  matrixcalc::is.positive.definite() # TRUE

pd <- rep(NA, N)
for (i in 1:N) {
  pd[i] <- 
    riem_isomap[,,i] %>%  # R
    # H_isomap[i,,, drop=TRUE] %>% # Python
      # isSymmetric()
      matrixcalc::is.positive.definite() # FALSE
}
pd %>% sum

# determinant(riem_isomap[,,i])
# mat <- riem_isomap[,,i]
# eigen(mat)
```

## Debug: riemmanien metric not positive definite (DONE)

**Question:** footnote on P15 of the paper, how to decide the constant `c`?

> In the case of heat kernel (1), c = 1/4, which - crucially - is independent of the dimension of M.

```{r, eval=FALSE}
# Function for graph Laplacian
# Input: W: N*N weight matrix for the neighborhood graph, radius: bandwidth parameter
# Output: L: N*N graph Laplacian matrix
Laplacian <- function(W, radius, lambda = 1){

  D <- Matrix::Diagonal(x = rowSums(W)^(-lambda)) # inverse of a diagonal matrix
  W1 <- D %*% W %*% D
  D1 <- Matrix::Diagonal(x = 1 / rowSums(W1)) # inverse of Tn1
  L <- 1 / (radius^2 / 4) * (D1 %*% W1 - Matrix::Diagonal(nrow(W))) # c=1/4 for heat kernel, depending on the choice of weights, P15 footnote

  return(L)
}
```

```{r, eval=FALSE}
# Function for Riemannian metric for each point
# The Riemannian metric and its dual associated with an embedding Y. 
# The intrinsic dimension d
# The Riemannian metric is currently denoted by G, its dual metric by H, and the Laplacian by L. 
# G at each point is the matrix inverse of H.
riemann_metric <- function(Y, laplacian, d){

  # TODO: add dimension check for all inputs

  H <- array(NA, dim = c(d, d, nrow(Y)))

  for (i in 1:d) {
    for (j in i:d) {
      yij <- Y[, i] * Y[, j]
      H[i, j, ] <- as.vector(0.5 * (laplacian %*% yij - Y[, i] * (laplacian %*% Y[, j]) - Y[, j] * (laplacian %*% Y[, i])))
      H[j, i, ] <- H[i, j, ] # symmetric matrix
    }
  }

  ## Pseudo inverse of H gives the final embedding metric h
  ## Array H corresponds to \tilde{H} in Step 4(a)
  ## The embedding metric H is the pseudo inverse of \tilde{H}
  for (i in 1:nrow(Y)) {
    Hsvals <- eigen(H[ , ,i])
    Huu <- Hsvals$vectors
    Hvv <- Hsvals$values[1:d] # top d largest eigenvalues, already sorted in decreasing order
    Hvv1 <- diag(x = 1 / Hvv)
    H[ , ,i] <- Huu %*% Hvv1 %*% t(Huu)
    H[, , i] <- 0.5 * (H[, , i] + t(H[, , i])) # fix not symmetric issue
  }

  return(H)
}
```

<!-- # Plotting -->

<!-- # Plot ellipse to show distortion -->

<!-- Using the Riemannian metric for each point as the covariance and the 2-d embeddings as the mean to plot an ellipse for each point.  -->

<!-- The underlying requirement is that the Riemannian metric needs to be a square positive definite matrix, i.e. `matrixcalc::is.positive.definite()` returns `TRUE`. -->

<!-- ```{r} -->
<!-- emb <- metric_isomap$embedding -->
<!-- i = 1 -->
<!-- mat <- riem_isomap[,,i] # not positive definite for metricML() -->
<!-- center <- c(emb[i,1], emb[i,2]) -->

<!-- mat <- H_isomap[i,,, drop=TRUE] # is.positive.definite for megaman -->

<!-- # mat <- matrix(c(2.2, 0.4, 0.4, 2.8), nrow=2) # covariance matrix -> cov(dataMatrix) -->

<!-- library(ellipse) -->
<!-- # plotcorr(mat) -->
<!-- # plotcorr(solve(mat)) -->
<!-- plot(ellipse::ellipse(mat, center = center, scale = c(1,1)), type = "l", col = "blue", asp = 1) -->

<!-- car::ellipse(center, shape=mat, radius=1, col="red", lty=2) -->
<!-- # library(mixtools) -->
<!-- mixtools::ellipse(mu = center, sigma = mat, type = "l") -->
<!-- ``` -->

# Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by
$$\hat{f}(x) = n^{-1} \sum_i K_H (x - X_i).$$
```{r}
# 2-d kde
mkde <- function (x, h) {
  # Data is a matrix of n*d
  # H is an array of dimension (d,d,n)
  start <- Sys.time()

  n <- nrow(x)
  if (dim(x)[2] < 2)
    stop("Data matrix has only one variable.")
  if (any(!is.finite(x))) 
    stop("Missing or infinite values in the data are not allowed! ")
  if (!all.equal(nrow(h), ncol(h), dim(x)[2]))
    stop("The bandwidth matrix is of wrong dimension. ")

  s <- rep(0, n)
  y <- rep(0, n)
  for (i in 1:n) {
    for (j in 1:n){
      s[i] <- s[i] + abs(det(h[,,i]))^(-1/2) * exp(- 1/2 * t(x[i,] - x[j,]) %*% solve(h[,,i]) %*% (x[i,] - x[j,]))
    }
    y[i] <- s[i] / (n * 2 * pi)
  }
  print(Sys.time() - start)

  return(y)
}
```

# Test the variable KDE function with megaman output

If we use the embedding `x` and the Riemannian metric `h` from `megaman`, we could produce contour plots.

```{r pycontour, eval=T}
x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
pyriem_isomap <- array(NA, dim = c(s, s, N))
for(i in 1:N){
  pyriem_isomap[,,i] <- H_isomap[i,,]
}
f <- mkde(x = x, h = pyriem_isomap)
summary(f)
# Top 20 index with the lowest densities
head(order(f), n=20)
```

## Contour plot for megaman

```{r}
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
# library(akima)
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax), lenbreak)
# hdrcde
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99))
plot.hdr2d(hdrinfo)
# Contour
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))})

```

# Contour plot after interpolation for R metricML()

Now apply the `mkde()` function to the output of `metricML()` written in R.

```{r rdensity}
x <- metric_isomap$embedding
riem_isomap <- metric_isomap$rmetric
f <- mkde(x = x, h = riem_isomap)
summary(f)
head(order(f), n=20)
```

We can check the outliers based on density `f`. In R, the outliers are different from `megaman` output. 
```{r, echo=FALSE}
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
# library(akima)
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax), lenbreak)
# hdrcde
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99))
plot.hdr2d(hdrinfo)
# Contour
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))})

```

### 2024-11-27 reports/20210603_fix_embedding.Rmd
Journal version pdf

---
title: "Riemannian metric from manifold learning"
author: "Fan Cheng"
date: "03/06/2021"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(dimRed)
library(reticulate)
library(viridis)
library(hdrcde)
library(igraph)
library(matrixcalc)
library(akima)
library(car)
Jmisc::sourceAll(here::here("R/sources"))
set.seed(123)
```

# Smart meter data for one household

```{r data}
# load(here::here("data/spdemand_3639id336tow.rda"))
# nid <- 1
# ntow <- 336
# 
# train <- spdemand %>%
#   lazy_dt() %>%
#   filter(tow <= ntow,
#          # id <= sort(unique(spdemand[,id]))[nid],
#          id == 1003) %>%
#   dplyr::select(-id, -tow) %>%
#   data.table::as.data.table()
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
```

# Metric learning

## Radius searching with k-d trees

```{r parameters}
# Parameters fixed
x <- train
N <- nrow(train)
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm
```

## The metric learning algorithm

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, invert.h = TRUE)
summary(metric_isomap)
```

The function `metricML()` returns a list of

-   the embedding coordinates `embedding` of $N \times s$
-   the embedding metric `rmetric` for each point $p \in x$ as an array
-   the weighted neighborhood graph as an `igraph` object `weighted_graph`
-   the sparse adjacency matrix from the graph `adj_matrix`
-   the graph laplacian matrix `laplacian`

## Check the output for each of the four steps

The metric learning algorithm consists of four steps:

1. construct a weighted neighborhood graph `weighted_graph`
2. calculate the graph laplacian `laplacian`
3. map the data $p \in x$ to the embedding coordinates `embedding`
4. apply the graph laplacian to the embedding corrdinates to obtain the embedding metric `rmetric`

Based on the output from megaman package, I compared the R function `metricML()` output for each of the four steps.

### Step1: weighted graph

We compare the graph by using the weighted adjacency matrix.

```{r, eval=T, include=T}
metric_isomap$adj_matrix[1:6, 1:6]
aff <-
  feather::read_feather(here::here("data/affinity_matrix_1id336tow.feather"))
aff[1:6, 1:6]
all.equal(as.matrix(aff), as.matrix(metric_isomap$adj_matrix), tolerance = 1e-5, check.attributes = F)
```

### Step2: Laplacian matrix

```{r, eval=T, include=T}
Ln <- metric_isomap$laplacian
Ln[1:6, 1:6]
lap <-
  feather::read_feather(here::here("data/laplacian_matrix_1id336tow.feather"))
lap[1:6, 1:6]
all.equal(as.matrix(lap), as.matrix(Ln), tolerance = 1e-5, check.attributes = F)
```

### Step3: Embedding coordinates

```{r, eval=T, include=T}
fn <- metric_isomap$embedding
fn %>% head()
emb_isomap <- feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather"))
emb_isomap %>% head()
all.equal(as.matrix(emb_isomap), as.matrix(fn), tolerance = 1e-5, check.attributes = F)
par(mfrow = c(1, 2))
plot(fn, col = viridis::viridis(24), asp=1) 
plot(emb_isomap$`0`, emb_isomap$`1`, col = viridis::viridis(24), asp=1)
```

### Step4: Riemannian metric

In `metricML()`, the `invert.h` argument controls whether the inverse of Riemannian metric should be returned. FALSE by default.

The megaman package returns the inverse of Riemannian matrix.

```{r, eval=T, include=T}
# Rn <- riemann_metric(Y = fn, laplacian = Ln, d = s, invert.h = T)
Rn <- metric_isomap$rmetric
Rn[,,1:3]

np = import("numpy")
H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy"))
H_isomap[1,,, drop=TRUE] %>% 
  matrixcalc::is.positive.definite()
pyh_isomap <- array(NA, dim = c(s, s, N))
for(i in 1:N){
  pyh_isomap[,,i] <- H_isomap[i,,]
}
pyh_isomap[,,1:3] # inverted

all.equal(pyh_isomap, Rn, tolerance = 1e-5, check.attributes = F)

# If we use the embedding from Python, the inverse of Riem metric is the same. 
hn <- riemann_metric(Y = as.matrix(emb_isomap), laplacian = Ln, d = s, invert.h = T)
hn[,,1:3]
all.equal(pyh_isomap, hn, tolerance = 1e-5, check.attributes = F) # use python embedding
```
Finally, the R function `metricML()` gives the same/close outputs as the Python megaman package.

<!-- ## Embedding plots -->

<!-- ```{r} -->
<!-- # Comparison of Isomap embedding plot -->
<!-- par(mfrow = c(1, 2)) -->
<!-- plot(metric_isomap$embedding, col = viridis::viridis(24), asp=1) # metricML -->
<!-- emb_isomap <- -->
<!--   feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather")) -->
<!-- plot(emb_isomap$`0`, emb_isomap$`1`, col = viridis::viridis(24), asp=1) # embedding from megaman, radius = 0.4 -->
<!-- ``` -->

<!-- ## Riemannian metric -->

<!-- ```{r} -->
<!-- # Use adjacency matrix as input for metricML and dimRed, then the embedding should be close, as well as the Riemannian metric -->
<!-- metric_isomap$adj_matrix[1:6, 1:6]# renormalized adjacency matrix -->
<!-- metric_isomap$weighted_graph %>% is.connected() -->
<!-- riem_isomap <- metric_isomap$rmetric -->
<!-- riem_isomap[,,1] %>%  -->
<!--   # isSymmetric() -->
<!--   matrixcalc::is.positive.definite() # FALSE -->
<!-- ``` -->

<!-- **The Riemannian metric from the `metricML()` function is now positive definite for all points.** -->

<!-- Now compare it with the megaman output. -->

<!-- ```{r, eval=TRUE} -->
<!-- np = import("numpy") -->
<!-- H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy")) -->
<!-- H_isomap[1,,, drop=TRUE] %>%  -->
<!--   matrixcalc::is.positive.definite() # TRUE -->

<!-- pd <- rep(NA, N) -->
<!-- for (i in 1:N) { -->
<!--   pd[i] <-  -->
<!--     riem_isomap[,,i] %>%  # R -->
<!--     # H_isomap[i,,, drop=TRUE] %>% # Python -->
<!--       # isSymmetric() -->
<!--       matrixcalc::is.positive.definite() # FALSE -->
<!-- } -->
<!-- pd %>% sum -->

<!-- # determinant(riem_isomap[,,i]) -->
<!-- # mat <- riem_isomap[,,i] -->
<!-- # eigen(mat) -->
<!-- ``` -->

# Ellipse plot

Using the Riemannian metric for each point as the 2-d covariance matrix and the 2-d embeddings as the centroid to plot an ellipse for each point.

The underlying requirement is that the Riemannian metric needs to be a square positive definite matrix, i.e. `matrixcalc::is.positive.definite()` returns `TRUE`.

```{r}
# TODO: if add = FALSE, fix xlim, ylim, radius for generating ellipse
# TODO: add argument `n.plot` to specify how many ellipse to be added
x <- metric_isomap$embedding
cols <- viridis::viridis(24) 
plot(metric_isomap$embedding, col = cols, asp=1, pch = 20)
for(i in 1:N){
  mat <- Rn[,,i] # pyh_isomap[,,I]
  center <- c(x[i,1], x[i,2])
  # add <- ifelse(i == 1, F, T)
  add <- T
  car::ellipse(center, mat, 
               radius = .1, # controls the size of ellipse, currently set manually
               # col = cols[i], 
               col = blues9[5],
               asp = 1, pty = "s", lwd = 1, center.pch = 19, center.cex = 0,
               fill = T, fill.alpha = 0.2, add = add, grid = T,
               # xlim = c(-.2, .25), ylim = c(-.2, .2)
               )
}
```

# Contour plots

## Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by $$\hat{f}(x) = n^{-1} \sum_i K_H (x - X_i).$$

### Inverse of Riemannian metric as bandwidth matrix

If we use the embedding `x` and the inverse of Riemannian metric `h`, we could produce contour plots based on the density estimates.

```{r pydensity, eval=T}
x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
f <- mkde(x = x, h = pyh_isomap)
summary(f)
# Top 20 index with the lowest densities
head(order(f), n=20)
```

```{r, echo=FALSE, message=FALSE}
p_pyh <- plot.contour(x, f, plot.hdr = F)
```

Now apply the `mkde()` function to the output of `metricML()` written in R.

```{r rdensity}
x <- metric_isomap$embedding
riem_isomap <- metric_isomap$rmetric
f <- mkde(x = x, h = riem_isomap)
summary(f)
head(order(f), n=20)
```

```{r, echo=FALSE, message=FALSE}
p_rh <- plot.contour(x, f, plot.hdr = F)
```

We can check the outliers based on density `f`.
In R, the outliers are similar to the `megaman` output.

### Riemannian metric itself as bandwidth matrix

```{r pydensity1, eval=T}
R_isomap <- np$load(here::here("data/rmatrix_isomap_1id336tow.npy"))
# R_isomap[1,,, drop=TRUE] %>% 
#   matrixcalc::is.positive.definite() # not necessarily symmetric
pyr_isomap <- array(NA, dim = c(s, s, N))
for(i in 1:N){
  pyr_isomap[,,i] <- R_isomap[i,,]
}
pyr_isomap[,,1:3]

x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
f <- mkde(x = x, h = pyr_isomap)
summary(f)
# Top 20 index with the lowest densities
head(order(f), n=20)
p_pyr <- plot.contour(x, f, plot.hdr = F)
```

Now apply the `mkde()` function to the output of `metricML()` written in R.

```{r rdensity1}
hn <- riemann_metric(Y = as.matrix(emb_isomap), laplacian = Ln, d = s, invert.h = T)
f <- mkde(x = metric_isomap$embedding, h = hn)
summary(f)
head(order(f), n=20)
p_rr <- plot.contour(x, f, plot.hdr = F)
```

# t-SNE

```{r tsne, message=FALSE}
metric_tsne <- metricML(x, s = s, k = k, radius = radius, method = "tSNE", annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, invert.h = TRUE)
summary(metric_tsne)
```

```{r}
f <- mkde(x = metric_tsne$embedding, h = metric_tsne$rmetric)
summary(f)
head(order(f), n=20)
plot.contour(x = metric_tsne$embedding, f, plot.hdr = F)
```

```{r, echo=FALSE, eval=FALSE}
# Countour plot for megaman, dual metric
# x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))})
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
# library(akima)
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax), lenbreak)
# hdrcde
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99))
plot.hdr2d(hdrinfo)
# Contour
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))},
               # plot.title = {par(cex.main=1);title(main = "The Topography of Maunga Whau", xlab = "Meters North", ylab = "Meters West")},
               # plot.axes = { axis(1, seq(100, 800, by = 100))
                 # axis(2, seq(100, 600, by = 100)) },
               # key.title = {par(cex.main=0.3);title(main="Height\n(meters)")},
               # key.axes = axis(4, seq(-1, 1.5, by = 0.5), asp=1)
)

```

### 2024-11-27 reports/20210610_contourplot.Rmd
Journal version pdf

---
title: "Plotting for metric learning"
author: "Fan Cheng"
date: "2021-06-10"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(dimRed)
library(reticulate)
library(here)
library(viridis)
library(hdrcde)
library(igraph)
library(matrixcalc)
library(akima)
library(car)
Jmisc::sourceAll(here::here("R/sources"))
set.seed(123)
```

# Metric learning

## Smart meter data for one household

```{r data}
# load(here::here("data/spdemand_3639id336tow.rda"))
# nid <- 1
# ntow <- 336
# 
# train <- spdemand %>%
#   lazy_dt() %>%
#   filter(tow <= ntow,
#          # id <= sort(unique(spdemand[,id]))[nid],
#          id == 1003) %>%
#   dplyr::select(-id, -tow) %>%
#   data.table::as.data.table()
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
```

## Radius searching with k-d trees

```{r parameters}
# Parameters fixed
x <- train
N <- nrow(train)
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm
```

## The metric learning algorithm

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, invert.h = TRUE)
summary(metric_isomap)
```

The function `metricML()` returns a list of

-   the embedding coordinates `embedding` of $N \times s$
-   the embedding metric `rmetric` for each point $p \in x$ as an array. By default, `invert.h = TRUE` returns the inverse of Riemannian metric. 
-   the weighted neighborhood graph as an `igraph` object `weighted_graph`
-   the sparse adjacency matrix from the graph `adj_matrix`
-   the graph laplacian matrix `laplacian`

# Embedding plot

```{r embedplot, eval=FALSE}
# Embedding plot of metricML() output for electricity data
plot.embedding <- function(x) {
  fn <- x$embedding

  todcolor <- colorspace::scale_color_continuous_sequential(
    palette = "viridis",
    breaks = c(0, 12, 24, 36, 48),
    labels=c("00:00", "06:00", "12:00", "18:00", "24:00"),
    name="Time of day",
    guide=guide_colorbar(barwidth = 10))

  p <- fn %>% 
    cbind(tod = rep(1:48, N / 48)) %>% 
    as_tibble() %>% 
    ggplot(aes(x = E1, y = E2, col = tod)) + 
    geom_point() + 
    # labs(x = "ISO1", y = "ISO2") +
    coord_fixed(ratio = 1) + 
    todcolor + 
    theme(legend.position = 'bottom')

  return(p)
}
```

```{r, eval=T, include=T}
# fn <- metric_isomap$embedding
# head(fn)
# plot(fn, col = viridis::viridis(24), asp = 1)
plot.embedding(metric_isomap) +
  labs(x = "ISO1", y = "ISO2")
```

# Ellipse plot

Using the Riemannian metric for each point as the 2-d covariance matrix and the 2-d embeddings as the centroid to plot an ellipse for each point.

The underlying requirement is that the Riemannian metric needs to be a square positive definite matrix.

```{r}
# Done: add argument `n.plot` to specify how many ellipse to be added
# If add = FALSE, fix xlim, ylim, radius for generating ellipse
plot.ellipse <- function(x, n.plot = 50) {
  fn <- x$embedding
  Rn <- x$rmetric
  cols <- viridis::viridis(24) 
  n.plot <- 50
  plot(fn, col = cols, asp = 1, pch = 20)
  for (i in sample(1:N, n.plot)) {
    mat <- Rn[,,i] 
    center <- c(fn[i,1], fn[i,2])
    add <- T 
    car::ellipse(center, mat, 
                 radius = .1, # controls the size of ellipse, currently set manually
                 # xlim = c(-.2, .25), ylim = c(-.2, .2),
                 # col = cols[i],
                 col = blues9[5],
                 asp = 1, pty = "s", lwd = 1, center.pch = 19, center.cex = 0,
                 fill = T, fill.alpha = 0.2, grid = T, add = add, 
                 )
  }
}
plot.ellipse(metric_isomap)
```

To plot as an ggplot2 object, the exsiting `stat_ellipse` does not work for variable covariance matrix. I tried to modify the code and create `stat_ellipse_cov()` which takes another aesthetic `covmat`. But all points are connected as one path.

Also, `purrr:::pmap_dfc()` does not work correctly, so for loop is currently in use.

```{r}
fn <- metric_isomap$embedding
Rn <- metric_isomap$rmetric
a <- fn %>% 
  as_tibble() %>% 
  add_column(covmat = purrr::array_branch(Rn, 3))
a
# a$covmat %>% head()

a %>% 
  cbind(tod = rep(1:48, N / 48)) %>% 
  ggplot(aes(E1, E2)) + 
  geom_point() +
  stat_ellipse()
  # stat_ellipse(aes(group = tod, color = tod))

# a %>% 
#   cbind(tod = rep(1:48, N / 48)) %>%
#   # group_by(tod) %>%
#   ggplot(aes(x = E1, y = E2, color = tod)) +
#   geom_point() +
#   scale_color_viridis() + 
#   stat_ellipse_cov(aes(covmat = covmat))

plot.embedding(x = fn, embedding = T) + 
  stat_ellipse_cov(data = cbind(a, tod = rep(1:48, N / 48)), aes(covmat = covmat))
```

# Contour plot

## Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by $$\hat{f}(x) = n^{-1} \sum_i K_{H_i} (x - X_i).$$

### Inverse of Riemannian metric as bandwidth matrix

If we use the embedding `x` and the inverse of Riemannian metric `h`, we could produce contour plots based on the density estimates.

```{r pydensity, eval=T}
emb_isomap <- feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather"))
np <- reticulate::import("numpy")
H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy"))
pyh_isomap <- array(NA, dim = c(s, s, N))
for(i in 1:N){
  pyh_isomap[,,i] <- H_isomap[i,,]
}

x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
f <- mkde(x = x, h = pyh_isomap)
summary(f)
# Top 20 index with the lowest densities
head(order(f), n = 20)
```

```{r, echo=FALSE, message=FALSE}
p_pyh <- plot.contour(x, f, plot.hdr = F)
```

Try linear interpolation using `approx()`.

```{r, eval=FALSE, echo=TRUE}
x_interp <- approx(x[,1], x[,2], n = 100, yleft = 0, yright = 1, ties = mean)
x_int <- cbind(x_interp$x, x_interp$y)
dim(x_int)
f_linerinterp <- mkde(x = x_int, h = pyh_isomap)
f_linerinterp
plot.contour(x_interp, f_linerinterp)
hdrscatterplot(x_interp$x, x_interp$y, noutliers = 10)
p <- filled.contour(akima.spl, color.palette = viridis,
                 plot.axes = { axis(1); axis(2);
                   # title("smooth interp(*, linear = FALSE)");
                   points(embed_den, pch = 3, col= hcl(c=20, l = 10))})
```

Now apply the `mkde()` function to the output of `metricML()` written in R.

```{r rdensity}
x <- metric_isomap$embedding
riem_isomap <- metric_isomap$rmetric
f <- mkde(x = x, h = riem_isomap)
summary(f)
head(order(f), n=20)
```

```{r, echo=FALSE, message=FALSE}
p_rh <- plot.contour(x, f, plot.hdr = F)
```

We can check the outliers based on density `f`.
In R, the outliers are similar to the `megaman` output.

<!-- ### Riemannian metric itself as bandwidth matrix -->

<!-- ```{r pydensity1, eval=T} -->
<!-- R_isomap <- np$load(here::here("data/rmatrix_isomap_1id336tow.npy")) -->
<!-- # R_isomap[1,,, drop=TRUE] %>%  -->
<!-- #   matrixcalc::is.positive.definite() # not necessarily symmetric -->
<!-- pyr_isomap <- array(NA, dim = c(s, s, N)) -->
<!-- for(i in 1:N){ -->
<!--   pyr_isomap[,,i] <- R_isomap[i,,] -->
<!-- } -->
<!-- pyr_isomap[,,1:3] -->

<!-- x <- cbind(emb_isomap$`0`, emb_isomap$`1`) -->
<!-- f <- mkde(x = x, h = pyr_isomap) -->
<!-- summary(f) -->
<!-- # Top 20 index with the lowest densities -->
<!-- head(order(f), n=20) -->
<!-- p_pyr <- plot.contour(x, f, plot.hdr = F) -->
<!-- ``` -->

<!-- Now apply the `mkde()` function to the output of `metricML()` written in R. -->

<!-- ```{r rdensity1} -->
<!-- hn <- riemann_metric(Y = as.matrix(emb_isomap), laplacian = Ln, d = s, invert.h = T) -->
<!-- f <- mkde(x = metric_isomap$embedding, h = hn) -->
<!-- summary(f) -->
<!-- head(order(f), n=20) -->
<!-- p_rr <- plot.contour(x, f, plot.hdr = F) -->
<!-- ``` -->

# t-SNE

```{r tsne, message=FALSE, warning=FALSE, eval=TRUE}
method <- "tSNE"
perplexity <- 30
theta <- 0.5
metric_tsne <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = perplexity, theta = theta, invert.h = TRUE)
summary(metric_tsne)
```

```{r, include=FALSE, eval=FALSE}
perplexity <- 30
theta <- 0.5
method <- "tSNE"
e <- dimRed::embed(x,
                     .method = method,
                     knn = k,
                     ndim = s,
                     annmethod = annmethod,
                     radius = radius,
                     eps = eps,
                     nt = nt,
                     nlinks = nlinks,
                     ef.construction = ef.construction,
                     distance = distance,
                     treetype = treetype,
                     searchtype = searchtype,
                     perplexity = perplexity,
                     theta = theta,
                     .mute = c("output"),
                     # ...

  )
  fn <- e@data@data
plot(metric_tsne$embedding)
```

```{r, message=FALSE, eval=TRUE}
f <- mkde(x = metric_tsne$embedding, h = metric_tsne$rmetric)
summary(f)
head(order(f), n=20)
plot.contour(x = metric_tsne$embedding, f, plot.hdr = F)
```

```{r anntsne, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
x <- train
nn2res <- dplyr::case_when(distance=="euclidean" ~ 
                                 RANN::nn2(data = x, query = x, k = k + 1, 
                                           treetype = treetype, searchtype = searchtype, eps = eps,
                                           radius = radius),
                               distance=="manhattan" ~ 
                                 RANN.L1::nn2(data = x, query = x, k = k + 1, 
                                              treetype = treetype, searchtype = searchtype, eps = eps,
                                              radius = radius),
    )
    names(nn2res) <- c("nn.idx", "nn.dists")

metric_tsne1 <- metricML(x, nn.idx = nn2res$nn.idx, nn.dists = nn2res$nn.dists, s = s, k = k, radius = radius, method = "anntSNE", annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = k-1, theta = 0.5, invert.h = TRUE)
summary(metric_tsne1)

f <- mkde(x = metric_tsne1$embedding, h = metric_tsne1$rmetric)
summary(f)
head(order(f), n=20)
plot.contour(x = metric_tsne1$embedding, f, plot.hdr = F)
```

```{r, echo=FALSE, eval=FALSE}
# Countour plot for megaman, dual metric
# x <- cbind(emb_isomap$`0`, emb_isomap$`1`)
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))})
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
# library(akima)
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax), lenbreak)
# hdrcde
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99))
plot.hdr2d(hdrinfo)
# Contour
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))},
               # plot.title = {par(cex.main=1);title(main = "The Topography of Maunga Whau", xlab = "Meters North", ylab = "Meters West")},
               # plot.axes = { axis(1, seq(100, 800, by = 100))
                 # axis(2, seq(100, 600, by = 100)) },
               # key.title = {par(cex.main=0.3);title(main="Height\n(meters)")},
               # key.axes = axis(4, seq(-1, 1.5, by = 0.5), asp=1)
)

```

### 2024-11-27 reports/20210617_plots.Rmd
Journal version pdf

---
title: "Plotting for metric learning output"
author: "Fan Cheng"
date: "2021-06-17"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(dimRed)
library(reticulate)
library(here)
library(viridis)
library(hdrcde)
library(igraph)
library(matrixcalc)
library(akima)
library(car)
library(ggforce)
library(ks)
Jmisc::sourceAll(here::here("R/sources"))
set.seed(123)
```

# Metric learning

## Smart meter data for one household

```{r data}
# load(here::here("data/spdemand_3639id336tow.rda"))
# nid <- 1
# ntow <- 336
# 
# train <- spdemand %>%
#   lazy_dt() %>%
#   filter(tow <= ntow,
#          # id <= sort(unique(spdemand[,id]))[nid],
#          id == 1003) %>%
#   dplyr::select(-id, -tow) %>%
#   data.table::as.data.table()
train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
```

## Radius searching with k-d trees

```{r parameters}
# Parameters fixed
x <- train
N <- nrow(train)
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- .4 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm
```

## The metric learning algorithm

```{r, message=FALSE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, invert.h = TRUE)
summary(metric_isomap)
```

The function `metricML()` returns a list of

-   the embedding coordinates `embedding` of $N \times s$
-   the embedding metric `rmetric` for each point $p \in x$ as an array. By default, `invert.h = TRUE` returns the inverse of Riemannian metric. 
-   the weighted neighborhood graph as an `igraph` object `weighted_graph`
-   the sparse adjacency matrix from the graph `adj_matrix`
-   the graph laplacian matrix `laplacian`

# Embedding plot

```{r embedplot, eval=FALSE, include=FALSE}
# Embedding plot of metricML() output for electricity data
plot_embedding <- function(x) {
  fn <- x$embedding

  todcolor <- colorspace::scale_color_continuous_sequential(
    palette = "viridis",
    breaks = c(0, 12, 24, 36, 48),
    labels=c("00:00", "06:00", "12:00", "18:00", "24:00"),
    name="Time of day",
    guide=guide_colorbar(barwidth = 10))

  p <- fn %>% 
    cbind(tod = rep(1:48, N / 48)) %>% 
    as_tibble() %>% 
    ggplot(aes(x = E1, y = E2, col = tod)) + 
    geom_point() + 
    # labs(x = "ISO1", y = "ISO2") +
    coord_fixed(ratio = 1) + 
    todcolor + 
    theme(legend.position = 'bottom')

  return(p)
}
```

```{r, eval=T, include=T}
# fn <- metric_isomap$embedding
# head(fn)
# plot(fn, col = viridis::viridis(24), asp = 1)
plot_embedding(metric_isomap) +
  labs(x = "ISO1", y = "ISO2")
```

# Ellipse plot

Using the Riemannian metric for each point as the 2-d covariance matrix and the 2-d embeddings as the centroid to plot an ellipse for each point.

The underlying requirement is that the Riemannian metric needs to be a square positive definite matrix.

```{r car, include=FALSE, eval=FALSE}
# Done: add argument `n.plot` to specify how many ellipse to be added
# If add = FALSE, fix xlim, ylim, radius for generating ellipse
plot.ellipse <- function(x, n.plot = 50) {
  fn <- x$embedding
  Rn <- x$rmetric
  cols <- viridis::viridis(24) 
  n.plot <- 50
  plot(fn, col = cols, asp = 1, pch = 20)
  for (i in sample(1:N, n.plot)) {
    mat <- Rn[,,i] 
    center <- c(fn[i,1], fn[i,2])
    add <- T 
    car::ellipse(center, mat, 
                 radius = .1, # controls the size of ellipse, currently set manually
                 # xlim = c(-.2, .25), ylim = c(-.2, .2),
                 # col = cols[i],
                 col = blues9[5],
                 asp = 1, pty = "s", lwd = 1, center.pch = 19, center.cex = 0,
                 fill = T, fill.alpha = 0.2, grid = T, add = add, 
                 )
  }
}
plot.ellipse(metric_isomap)
```

```{r ggellipse}
plot_ellipse(metric_isomap, add = F, n.plot = 50, scale = 20, 
             color = blues9[5], fill = blues9[5], alpha = 0.2)
# fn <- metric_isomap$embedding
# Rn <- metric_isomap$rmetric # array
# n.plot <- 50
# e <- riem2ellipse(Rn, scale = 20) %>% 
#   cbind(fn) %>% 
#   as_tibble()
# OR
# plot_embedding(metric_isomap, embedding = F) +
#   plot_ellipse(metric_isomap, add = T)
# OR
# # plot_embedding(fn, embedding = T) + 
#   ggplot(data = e, aes(E1, E2)) +
#   geom_point() +
#   geom_ellipse(data = slice_sample(e, n = n.plot), 
#                aes(x0 = E1, y0 = E2, a = a, b = b, angle = angle, group = -1), 
#                color = blues9[5], fill = blues9[5], alpha = 0.2)
```

# Variable kernel density estimate

For multivariate data, the variable kernel density estimate is given by $$\hat{f}(x) = n^{-1} \sum_i K_{H_i} (x - X_i).$$

## Outlier plot based on density estimates

### Fixed bandwidth

```{r}
# fixed bandwidth
fn <- metric_isomap$embedding
E1 <- fn[,1] # rename as Ed to match the aesthetics in plot_ellipse()
E2 <- fn[,2]
prob <- c(1, 50, 99)
p_outlier <- hdrscatterplot(E1, E2, levels = prob, noutliers = 20, label = NULL) + 
  plot_ellipse(metric_isomap, add = T, n.plot = 50, scale = 20, 
             color = blues9[5], fill = blues9[5], alpha = 0.2)
p_outlier
```

### Pointwise adaptive bandwidth

```{r outliers}
Rn <- metric_isomap$rmetric # array
n.grid <- 10
f <- vkde2d(x = fn[,1], y = fn[,2], h = Rn, n = n.grid)
plot_contour(metric_isomap, n.grid = n.grid, scale = 0.1)
```

```{r hdroutliers}
source(here::here("R/sources/hdrplotting.R"))
p_outlier_vkde <- plot_outlier(x = metric_isomap, n.grid = 50, prob = prob)
# den <- f
# x <- E1; y <- E2
#  <-  = c(50, 95, 99)
# # prob = (1:10)*10
# # Convert prob to coverage percentage if necessary
# if(max(prob) > 50) {# Assume prob is coverage percentage
#   alpha <- (100-prob)/100
# } else {# prob is tail probability (for backward compatibility)
#   alpha <- prob}
# alpha <- sort(alpha)
# # Calculates falpha needed to compute HDR of bivariate density den.
# # Also finds approximate mode.
# fxy <- hdrcde:::interp.2d(den$x,den$y,den$z,x,y)
# falpha <- quantile(fxy, alpha)
# index <- which.max(fxy)
# mode <- c(x[index],y[index])
# hdr2d_info <- structure(list(mode=mode,falpha=falpha,fxy=fxy, den=den, alpha=alpha, x=x, y=y), class="hdr2d") # list for hdr.2d() output
# # plot.hdr2d(hdr2d_info, show.points = T, outside.points = T, pointcol = grey(0.5), xlim = round(range(x)), ylim = round(range(y)))
# 
# p_outlier_vkde <- hdrscatterplot(E1, E2, levels = prob, noutliers = 20, label = NULL, den = hdr2d_info)  + 
#   plot_ellipse(metric_isomap, add = T, n.plot = 50, scale = 20, 
#              color = blues9[5], fill = blues9[5], alpha = 0.2)
```

```{r compoutlier}
library(patchwork)
(p_outlier + p_outlier_vkde ) + coord_fixed()
```

```{r, include=FALSE}
# todcolor <- colorspace::scale_color_continuous_sequential(
#   palette = "viridis",
#   # breaks = c(0, 12, 24, 36, 48),
#   # labels=c("00:00", "06:00", "12:00", "18:00", "24:00"),
#   name="Density",
#   guide=guide_colorbar(barwidth = 10))
# 
# p <- fn %>% 
#   # cbind(tod = rep(1:48, N / 48)) %>%
#   as_tibble() %>% 
#   ggplot(aes(x = E1, y = E2)) + 
#   geom_point() + 
#   # labs(x = "ISO1", y = "ISO2") +
#   coord_fixed(ratio = 1) + 
#   # todcolor + 
#   theme(legend.position = 'bottom')
#   
# p + 
#   plot_ellipse(metric_isomap, add = T, n.plot = 50, scale = 20, 
#              color = blues9[5], fill = blues9[5], alpha = 0.2)
```

```{r, include=FALSE, eval=FALSE}
fn <- metric_isomap$embedding
hdrscatterplot(fn[,1], fn[,2], noutliers = 10)
hdrinfo <- hdr.2d(fn[,1], fn[,2], prob = prob)
plot.hdr2d(hdrinfo)

# f <- vkde2d(x = fn[,1], y = fn[,2], h = Rn, n = 20)
# # modified
# f1 <- den.estimate.2d(x = fn[,1], y = fn[,2], kde.package = "ks", h = Rn[,,1], xextend =0.15, yextend=0.15)  # only one bandwidth at a time
# hdrinfo_h <- hdr.2d(fn[,1], fn[,2], prob = (1:20)*5, den = f1)
# plot.hdr2d(hdrinfo_h)

Rh <- apply(Rn, 3, det)
vfhat.sp <- ks:::kde.sp.2d(x=fn, H=Rh)
plot(vfhat.sp, display="persp", box=TRUE, phi=20, zlim=c(0,max(vfhat.sp$estimate)))

# den <- den.estimate.2d(x=fn[,1], y=fn[,2], kde.package="ks", h=Rh)
x=fn[,1]; y=fn[,2]
den <- vfhat.sp
den <- list(x=den$eval.points[[1]],y=den$eval.points[[2]],z=den$estimate)

prob = c(50, 95, 99)
prob = (1:10)*10
# Convert prob to coverage percentage if necessary
if(max(prob) > 50) {# Assume prob is coverage percentage
  alpha <- (100-prob)/100
} else {# prob is tail probability (for backward compatibility)
  alpha <- prob}
alpha <- sort(alpha)
# Calculates falpha needed to compute HDR of bivariate density den.
# Also finds approximate mode.
fxy <- hdrcde:::interp.2d(den$x,den$y,den$z,x,y)
falpha <- quantile(fxy, alpha)
index <- which.max(fxy)
mode <- c(x[index],y[index])
hdr2d_info <- structure(list(mode=mode,falpha=falpha,fxy=fxy, den=den, alpha=alpha, x=x, y=y), class="hdr2d") # list for hdr.2d() output
plot.hdr2d(hdr2d_info, show.points = F, outside.points = T, pointcol = grey(0.5), xlim = round(range(x)), ylim = round(range(y)))

filled.contour(den, color.palette = viridis,
                      plot.axes = { axis(1); axis(2);
                        points(fn, pch = 3, col= hcl(c=20, l = 8))})
```

```{r, eval=FALSE, include=FALSE}
fn <- metric_isomap$embedding
Rn <- metric_isomap$rmetric
# Variable KDE
data(worldbank)
wb <- as.matrix(na.omit(worldbank[,4:5]))
xmin <- c(-70,-25); xmax <- c(35,70)
fhat <- kde(x=wb, xmin=xmin, xmax=xmax)
fhat.sp <- kde.sp(x=wb, xmin=xmin, xmax=xmax)
vfhat.sp <- ks:::kde.sp.2d(x=wb, xmin=xmin, xmax=xmax, H = fhat.sp$H)
plot(fhat, display="persp", box=TRUE, phi=20, zlim=c(0,max(fhat$estimate)))
plot(fhat.sp, display="persp", box=TRUE, phi=20, zlim=c(0,max(fhat.sp$estimate)))
plot(vfhat.sp, display="persp", box=TRUE, phi=20, zlim=c(0,max(vfhat.sp$estimate)))
## Not run:
fhat.ball <- kde.balloon(x=wb, xmin=xmin, xmax=xmax)
plot(fhat.ball, display="persp", box=TRUE, phi=20, zlim=c(0,max(fhat.sp$estimate)))

ks::Hpi.diag(fn,binned=TRUE)
Hns(fn, deriv.order=4)

fhat.sp$H # vector of grid point length, not matrices
```

<!-- # Contour plot -->

<!-- ### Inverse of Riemannian metric as bandwidth matrix -->

<!-- If we use the embedding `x` and the inverse of Riemannian metric `h`, we could produce contour plots based on the density estimates. -->

<!-- ```{r pydensity, eval=F, include=FALSE} -->
<!-- emb_isomap <- feather::read_feather(here::here("data/embedding_isomap_1id336tow.feather")) -->
<!-- np <- reticulate::import("numpy") -->
<!-- H_isomap <- np$load(here::here("data/hmatrix_isomap_1id336tow.npy")) -->
<!-- pyh_isomap <- array(NA, dim = c(s, s, N)) -->
<!-- for(i in 1:N){ -->
<!--   pyh_isomap[,,i] <- H_isomap[i,,] -->
<!-- } -->
<!-- x <- cbind(emb_isomap$`0`, emb_isomap$`1`) -->
<!-- # f <- mkde(x = x, h = pyh_isomap) -->
<!-- # Top 20 index with the lowest densities -->
<!-- # head(order(f), n = 20) -->
<!-- plot_contour(list(embedding = x, rmetric = pyh_isomap), n.grid = 10) -->
<!-- ``` -->

<!-- ```{r contour} -->
<!-- plot_contour(metric_isomap, n.grid = 20) -->
<!-- # fn <- metric_isomap$embedding -->
<!-- # Rn <- metric_isomap$rmetric # array -->
<!-- # #h <- t(apply(Rn, 3, diag)) -->
<!-- # f <- vkde2d(x = fn[,1], y = fn[,2], h = Rn, n = 100) -->
<!-- # str(f) -->
<!-- # # image(f) -->
<!-- # p <- filled.contour(f, color.palette = viridis, -->
<!-- #                  plot.axes = { axis(1); axis(2); -->
<!-- #                    points(fn, pch = 3, col= hcl(c=20, l = 8))}) -->
<!-- ``` -->

<!-- ### Riemannian metric itself as bandwidth matrix -->

<!-- ```{r rdensity1} -->
<!-- fn <- metric_isomap$embedding -->
<!-- s <- 2 -->
<!-- hn <- riemann_metric(Y = fn, laplacian = metric_isomap$laplacian, d = s, invert.h = F) -->
<!-- # f <- vkde2d(x = fn[,1], y = fn[,2], h = hn, n = 20) -->
<!-- # str(f) -->
<!-- # # image(f) -->
<!-- # p <- filled.contour(f, color.palette = viridis, -->
<!-- #                  plot.axes = { axis(1); axis(2); -->
<!-- #                    points(fn, pch = 3, col= hcl(c=20, l = 8))}) -->
<!-- plot_contour(list(embedding = fn, rmetric = hn), n.grid = 100) -->
<!-- ``` -->

# t-SNE

```{r tsne, message=FALSE, warning=FALSE, eval=TRUE}
x <- train
method <- "tSNE"
perplexity <- 30
theta <- 0.5
metric_tsne <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = perplexity, theta = theta, invert.h = TRUE)
summary(metric_tsne)
```

```{r, message=FALSE, eval=TRUE}
ftsne <- metric_tsne$embedding
E1 <- ftsne[,1]; E2 <- ftsne[,2]
# plot_embedding(metric_tsne)
plot_ellipse(metric_tsne, n.plot = 50)
plot_contour(metric_tsne, n.grid = 20, scale = 1/8)
p_tsne <- plot_outlier(x = metric_tsne, n.grid = 20, prob = prob, noutliers = 20, scale = 1/8)
p_tsne
hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20) + 
  plot_ellipse(metric_tsne, n.plot = 50, add = T)
p_outlier_vkde + p_tsne
```

```{r, include=FALSE, eval=FALSE}
perplexity <- 30
theta <- 0.5
method <- "tSNE"
e <- dimRed::embed(x,
                     .method = method,
                     knn = k,
                     ndim = s,
                     annmethod = annmethod,
                     radius = radius,
                     eps = eps,
                     nt = nt,
                     nlinks = nlinks,
                     ef.construction = ef.construction,
                     distance = distance,
                     treetype = treetype,
                     searchtype = searchtype,
                     perplexity = perplexity,
                     theta = theta,
                     .mute = c("output"),
                     # ...

  )
  fn <- e@data@data
plot(metric_tsne$embedding)
```

```{r anntsne, message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
x <- train
nn2res <- dplyr::case_when(distance=="euclidean" ~ 
                                 RANN::nn2(data = x, query = x, k = k + 1, 
                                           treetype = treetype, searchtype = searchtype, eps = eps,
                                           radius = radius),
                               distance=="manhattan" ~ 
                                 RANN.L1::nn2(data = x, query = x, k = k + 1, 
                                              treetype = treetype, searchtype = searchtype, eps = eps,
                                              radius = radius),
    )
    names(nn2res) <- c("nn.idx", "nn.dists")

metric_tsne1 <- metricML(x, nn.idx = nn2res$nn.idx, nn.dists = nn2res$nn.dists, s = s, k = k, radius = radius, method = "anntSNE", annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = k-1, theta = 0.5, invert.h = TRUE)
summary(metric_tsne1)

plot.contour(metric_tsne1, n.grid = 10)
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
embed_den <- as_tibble(cbind(x = x[,1], y = x[,2], z = f)) %>%
  drop_na()
# library(akima)
pixel <- 100
lenbreak <- 5
akima.spl <- interp(embed_den$x, embed_den$y, embed_den$z, nx=pixel, ny=pixel, linear=FALSE)
p.zmin <- min(embed_den$z,na.rm=TRUE)
p.zmax <- max(embed_den$z,na.rm=TRUE)
breaks <- pretty(c(p.zmin,p.zmax), lenbreak)
# hdrcde
hdrscatterplot(embed_den$x, embed_den$y, noutliers = 10)
hdrinfo <- hdr.2d(embed_den$x, embed_den$y, prob = c(50, 95, 99))
plot.hdr2d(hdrinfo)
# Contour
contour(akima.spl, main = "smooth interp(*, linear = FALSE)", col = viridis(length(breaks)-1), levels=breaks, add=TRUE)
points(embed_den, pch = 20)

filled.contour(akima.spl, color.palette = viridis,
               plot.axes = { axis(1); axis(2);
                 title("smooth interp(*, linear = FALSE)");
                 points(embed_den, pch = 3, col= hcl(c=20, l = 10))},
               # plot.title = {par(cex.main=1);title(main = "The Topography of Maunga Whau", xlab = "Meters North", ylab = "Meters West")},
               # plot.axes = { axis(1, seq(100, 800, by = 100))
                 # axis(2, seq(100, 600, by = 100)) },
               # key.title = {par(cex.main=0.3);title(main="Height\n(meters)")},
               # key.axes = axis(4, seq(-1, 1.5, by = 0.5), asp=1)
)

```

### 2024-11-27 reports/20210729_review.Rmd
Journal version pdf

---
title: "July Summary"
author: "Fan Cheng"
date: "2021-07-28"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Supervision form

- External supervisor form needed

- Second Monash supervisor

# July summary

Below is a summary of some events I attended with some feedback.

## ANZSC 2021

Machine learning session; 10min talk + 5min Q&A.

- How to define anomalies?

- Anomalies sensitivity to manifold learning methods.

- Examine distributions of nearby typical instead of anomalous households.

## AMSI winter school

### Lectures

- More insights about dimension reduction from Tas.

- A lot of Bayesian statistics.

### Runner-up for the participants talks

- Two rounds: 8 people group, 13 finalist; vote for top 3 talks; 
- 10min talk, no live questions;

- skipped manifold learning and quality measures after the lectures

### Possible collaborations

ABS: Victoria smart meter data project, detailed to be discussed.

> Brian Chan: Ohh interesting, is the Vic data publicly available? The reason why I mention it is because I know we are doing some very interesting work with Smart Meter Data (not my team so I don't know the details) - perhaps there might be some collaboration opportunities here, I can definitely follow-up if you are interested

> Owen Forbes: how I might use manifold learning with my EEG/neuroscience data for my PhD. 
One approach I am considering for my future work is representing each EEG frequency spectrogram as a curve - could potentially use functional PCA where each channel’s spectrogram is a curve. But seeing your application to electrical usage patterns, it seems comparable not necessarily to the EEG time series but to the power as a function of frequency for each electrode

<!-- , kind of like this https://www.google.com/url?sa=i&url=https%3A%2F%2Fwww.researchgate.net%2Ffigure%2FPS[…]e=images&cd=vfe&ved=0CAsQjRxqFwoTCODswZzE7vECFQAAAAAdAAAAABAF -->

## ECSSC 2021

A 20min talk scheduled at Friday 11:20 am.

More time on manifold learning, quality measures and the result plots.

## MLANN paper revision

- Added t-SNE and UMAP implementation.

- Introduction: add references for t-SNE and UMAP; more explanations on why embedding quality matters more than recall rate; add references on t-SNE variations and UMAP using NN-Descent for fast NN and their limitation [?].

- Manifold learning: explain parameter settings for $K$ and $d$; t-SNE and UMAP; UMAP algorithm pointed to the paper [?].

- Quality measures: recall rate definition and why high recall rate is not enough for preserving accurate local structure [any insights?].

- Unify embedding dimension as $d=2$ for a better-visualization reason. Plus, t-SNE only works for $d=1,2,3$ since it's designed mainly for high-dimensional data visualization with scatterplots.

- MNIST experiments: list ML parameter settings; Update $d=2$ result.

- Smart meter data application: clarify dimension $p=200$ and N for different scenarios.

- Conclusions: extension is possible for other methods (ML & ANN) to avoid adding other methods later; add intrinsic dimension reference from Chris [?].

### 2024-11-27 reports/20210805_tsne_umap.Rmd
Journal version pdf

---
title: "Comparison between t-SNE and UMAP"
author: "Fan Cheng"
date: "2021-08-05"
bibliography: references.bib
biblio-style: authoryear-comp
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    # rmdformats::readthedown:
    # toc: true
    # float: true
  bookdown::pdf_document2:
    keep_tex: false
    toc: false
    includes:
      in_header: preamble.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- 
## MLANN paper revision

- Added t-SNE and UMAP implementation.

- Introduction: add references for t-SNE and UMAP; more explanations on why embedding quality matters more than recall rate; add references on t-SNE variations and UMAP using NN-Descent for fast NN and their limitation [?].

- Manifold learning: explain parameter settings for $K$ and $d$; t-SNE and UMAP; UMAP algorithm pointed to the paper [?].

- Quality measures: recall rate definition and why high recall rate is not enough for preserving accurate local structure [any insights?].

- Unify embedding dimension as $d=2$ for a better-visualization reason. Plus, t-SNE only works for $d=1,2,3$ since it's designed mainly for high-dimensional data visualization with scatterplots.

- MNIST experiments: list ML parameter settings; Update $d=2$ result.

- Smart meter data application: clarify dimension $p=200$ and N for different scenarios.

- Conclusions: extension is possible for other methods (ML & ANN) to avoid adding other methods later; add intrinsic dimension reference from Chris [?]. 
-->

# t-SNE

t-Distributed Stochastic Neighbor Embedding [t-SNE; @Van_der_Maaten2008-dv] is a dimension reduction technique well suited for the visualization of high-dimensional data in scatterplots. It works by minimizing the Kullback-Leibler divergence $KL(P||Q)$ between two joint distributions, $p_{ij}$ and $q_{ij}$, measuring the pairwise similarities of the high-dimensional input data $x_i$ and the low-dimensional embedded points $y_i$ respectively. 
To better capture the local data structure and solve the crowding problem, a normalized Student-t kernel is chosen for $q_{ij}$ since it is heavy-tailed, while a Gaussian kernel is used for $p_{ij}$. The bandwidth of the Gaussian kernel, $\sigma_i$ is set to control the *perplexity* of the conditional distribution, relating to the density of the input data. Considering the asymmetry of the Kullback-Leibler divergence, the object function aims at modelling similar points in the input space by nearby points in the embedding space. Analytical form of the gradient is used in the Gradient Descent optimization of Kullback-Leibler divergence. Details can be found in Algorithm \@ref(alg:tsne). Variants of the Barnes-Hut algorithm can be used to approximate the gradient of the object function [@Van_Der_Maaten2014-in].

\begin{algorithm}[!htb]
  \caption{t-SNE}
  \label{alg:tsne}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional embedding $y_i$ for all $i=1,\ldots,N$ }
  \Parameter{ \textit{perplexity} }
  \OptParameter{ number of iterations $T$, learning rate $\eta$, momentum $\alpha(t)$ }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Compute the Gaussian conditional probability of distances for $x_i$ and $x_j$ as $p_{j \mid i}$ using
  $$
    p_{j \mid i}=\frac{\exp \left(-\left\|x_{i}-x_{j}\right\|^{2} / 2 \sigma_{i}^{2}\right)}{\sum_{k \neq i} \exp \left(-\left\|x_{i}-x_{k}\right\|^{2} / 2 \sigma_{i}^{2}\right)}
  $$
  where the variance $\sigma_{i}$ for each $x_i$ optimized with the perplexity parameter by $ 2 ^ {-\sum_j{p_{j \mid i} \log_2{p_{j \mid i}}}} = \textit{perplexity} $, and set $p_{i \mid i}=0$;

  \STATE Set the probability as $p_{i j}=(p_{j \mid i}+p_{i \mid j}) / (2 N)$;

  \STATE Initialize the optimization solution as $\mathcal{Y}^{(0)} = \{ y_1, \dots, y_N \}$ and $\mathcal{Y}^{(1)}$ sampled from normal distribution $\mathcal{N}(0, 10^{-4} \boldsymbol{I})$;

  \FOR{$t = 1,\dots,T$}

  \STATE Compute the Student t-distribution of distances between $y_i$ and $y_j$ as 
  $$
    q_{i j}=\frac{\left(1+\left\|y_{i}-y_{j}\right\|^{2}\right)^{-1}}{\sum_{k \neq l}\left(1+\left\|y_{k}-y_{l}\right\|^{2}\right)^{-1}};
  $$

  \STATE Compute the gradient of the Kullback-Leibler divergence loss function for all $i=1,\ldots,N$ as
  $$
    \frac{\delta KL}{\delta y_{i}}=4 \sum_{j}\left(p_{i j}-q_{i j}\right)\left(y_{i}-y_{j}\right)\left(1+\left\|y_{i}-y_{j}\right\|^{2}\right)^{-1} ;
  $$

  \STATE Update $ \mathcal{Y}^{(t)}=\mathcal{Y}^{(t-1)} + \eta \frac{\delta KL}{\delta \mathcal{Y}} + \alpha(t)\left(\mathcal{Y}^{(t-1)}-\mathcal{Y}^{(t-2)}\right). $

  \ENDFOR

  \end{algorithmic}
\end{algorithm}

Perplexity: to balance attention between local and global aspects of your data.

- The parameter is a guess about the number of close neighbors each point has. The original paper says, *“The performance of SNE is fairly robust to changes in the perplexity, and typical values are between 5 and 50.”*

Other hyperparameters: these choices interact with the amount of momentum and the step size are employed in the gradient descent. It is therefore common to run the optimization several times on a data set to find appropriate values for the parameters.

## Illustration

Below are some figures to illustrate the steps in t-SNE from a book named *Machine Learning with R, the tidyverse, and mlr* by Hefin I. Rhys.

```{r tsne, fig.align = 'center', out.width = "80%", echo = FALSE, fig.cap = "", eval=TRUE}
knitr::include_graphics(here::here("reports/figures/tsne1.png"))
knitr::include_graphics(here::here("reports/figures/tsne2.png"))
knitr::include_graphics(here::here("reports/figures/tsne3.png"))
knitr::include_graphics(here::here("reports/figures/tsne4.png"))
```

\clearpage

# UMAP

UMAP [Uniform Manifold Approximation and Projection; @McInnes2018-xo] is a competitive dimension reduction methods with t-SNE in increasing speed and preseving global structure. It is based on Laplacian Eigenmaps, considering Riemannian geometry and algebraic topology. Instead of optimizing the similarities in t-SNE, UMAP works by minimizing the error between two topological representations from the input space and embedding space. The error is measured by the cross entropy $C$ of two local fuzzy simplical sets built from the nearest neighborhood graph, with edge weights representing the likelihood that two points are connected.
UMAP extends a raidus outwards at each point until overlap with the nearest neighbors, with a \textit{min\_dist} as a desired separation between nearby embedded points to control global structure. Finally, similar to t-SNE, stochastic gradient descent is used for the optimization of object function. The algorithm and hyperparameters are detailed in Section 4 of @McInnes2018-xo.

\begin{algorithm}[!htb]
  \caption{UMAP}
  \label{alg:umap}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional embedding $y_i$ for all $i=1,\ldots,N$ }
  \Parameter{ number of nearest neighbors $K$, minimum distance between embedded points \textit{min\_dist}, amount of optimization work \textit{n\_epochs}}
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Construct the weighted $K$-nearest neighborhood graph with the distance to the nearest neighbor as $\rho_i$ for each $x_i$;

  \STATE Compute the exponential probability of distances between $x_i$ and $x_j$ as $p_{j \mid i}$ using
  $$
    p_{i \mid j}=e^{-\frac{d\left(x_{i}, x_{j}\right)-\rho_{i}}{\sigma_{i}}},
  $$
  where $d$ can be any distance metric, such that $\log_2{K} = \sum_{i} p_{i j}$;

  \STATE Construct a local fuzzy simplicial set by setting $p_{i j}=p_{j \mid i}+p_{i \mid j} - p_{i \mid j} p_{j \mid i}$;

  \STATE Compute the distribution of distances between $y_i$ and $y_j$ as
  $$
    q_{i j}=\left(1+a\left(y_{i}-y_{j}\right)^{2 b}\right)^{-1},
  $$
  where $a$ and $b$ are found using \textit{min\_dist} parameter such that
  $$
      q_{i j} \approx\left\{\begin{array}{ll}
  1 & \text { if } y_{i}-y_{j} \leq \textit{min\_dist} \\
  e^{-\left(y_{i}-y_{j}\right)-\textit{min\_dist}} & \text { if } y_{i}-y_{j} > \textit{min\_dist};
  \end{array}\right.
  $$

  \STATE The cost function is constructed using a binary fuzzy set cross entropy (CE) as
  $$
    CE(X, Y)=\sum_{i} \sum_{j}\left[p_{i j}(X) \log \left(\frac{p_{i j}(X)}{q_{i j}(Y)}\right)+\left(1-p_{i j}(X)\right) \log \left(\frac{1-p_{i j}(X)}{1-q_{i j}(Y)}\right)\right]
  $$

  \STATE Initialize the embedding coordinates $y_i$ using the graph Laplacian similar to Laplacian Eigenmaps in Algorithm \@ref(alg:le);

  \STATE Optimize $y_i$ by minimizing the cross entropy using stochastic gradient descent with \textit{n\_epochs}.

  \end{algorithmic}
\end{algorithm}

## Parameters

- $K$: number of nearest neighbors

- \textit{min\_dist}: the minimum distance between points in low-dimensional space.

    This parameter controls how tightly UMAP clumps points together, with low values leading to more tightly packed embeddings. Larger values of \textit{min\_dist} will make UMAP pack points together more loosely, focusing instead on the preservation of the broad topological structure.

## Illustration

Below are some figures to illustrate the steps in UMAP from the same book.

```{r umap, fig.align = 'center', out.width = "75%", echo = FALSE, fig.cap = "", eval=TRUE}
knitr::include_graphics(here::here("reports/figures/umap.png"))
```

# Compare

The biggest difference between the the output of UMAP when compared with t-SNE is this balance between local and global structure - UMAP is often better at preserving global structure.

## Visualization examples

The following blog contains visualization examples for comparing t-SNE and UMAP, as well as the effect of hyperparameters on the embedding.

https://pair-code.github.io/understanding-umap/

## t-SNE disadvantages

- tSNE does not scale well for rapidly increasing sample sizes.

- tSNE does not preserve global data structure. Only within cluster distances are meaningful while between cluster similarities are not guaranteed, therefore it is widely acknowledged that clustering on tSNE is not a very good idea.

- tSNE can practically only embed into 2 or 3 dimensions.

- tSNE can not work with high-dimensional data directly, Autoencoder or PCA are often used for performing a pre-dimensionality reduction before plugging it into the tSNE.

- tSNE consumes too much memory for its computations which becomes especially obvious when using large perplexity.

## Takeaways from the blog

- Hyperparameters really matter

- Cluster sizes in a UMAP plot mean nothing

- Distances between clusters might not mean anything

- Random noise doesn’t always look random

  Especially at low values of n_neighbors, spurious clustering can be observed.

- You may need more than one plot

# Things related to our paper

t-SNE Variants: implemented via Barnes-Hut approximations for gradient descent process, which also limits the implementation to two or three dimensions. @Tang2016-ho extended t-SNE with the LargeVis algorithm where random projection trees are used to reduce time cost.

L.J.P. van der Maaten. Accelerating t-SNE using Tree-Based Algorithms. Journal of Machine Learning Research 15(Oct):3221-3245, 2014.

UMAP: allow any distance metric for constructing high-dimensional distance probabilities using NN-Descent [@Dong2011-ee] algorithm; used procrustes measure to measure stability of the embedding.

# Estimating intrinsic dimensions

There are multiple methods to estimate the intrinsic dimensions, with 17 of them implemented in R package *Rdimtools*.

https://kisungyou.com/Rdimtools/reference/index.html#section--dimension-estimation

- Running 5 of them on MNIST dataset gives around 12 dimensions as an output.

- How to deal with the implementation of t-SNE then?

- How to unify this in the paper?

# TODO

- Run simulations for both dimension 2 and 12 on MNIST and check if the conclusion holds.

- For electricity dataset, since the goal is to find anomalies, we could choose dimension 2 for simplicity. (otherwise compute densities for high-dimensional embedding)

# References

### 2024-11-27 reports/20210819_otherdata.Rmd
Journal version pdf

--- 2021-09-02 reports/20210709_outliers.Rmd
@@
 ---
-title: "Plotting for metric learning output"
+title: "More data and methods for metric learning"
 author: "Fan Cheng"
-date: "2021-06-17"
+date: "2021-08-19"
 output: 
@@
 train <- readRDS(here::here("data/spdemand_1id336tow_train.rds"))
+# train <- readRDS(here::here("data/spdemand_3639id_notow_201length.rds"))
 # Parameters fixed
@@
 f <- vkde2d(x = fn[,1], y = fn[,2], h = Rn, n = n.grid)
-plot_contour(metric_isomap, n.grid = n.grid, scale = 1/10)
+plot_contour(metric_isomap, n.grid = n.grid, scale = 1/20)
 ```
@@
 
-```{r compoutlier}
+```{r compoutlier, eval = FALSE}
 (p_hdr_isomap_p + p_isomap$p ) + coord_fixed()
@@
 E1 <- ftsne[,1]; E2 <- ftsne[,2]
-plot_embedding(metric_tsne)
-plot_ellipse(metric_tsne, n.plot = 50)
-plot_contour(metric_tsne, n.grid = 50, scale = 1/10)
+# plot_embedding(metric_tsne)
+# plot_ellipse(metric_tsne, n.plot = 50)
+plot_contour(metric_tsne, n.grid = 50, scale = 1/20)
 ```
 
-```{r}
-p_tsne <- plot_outlier(x = metric_tsne, n.grid = 50, prob = prob, noutliers = 20, scale = 1/10)
+```{r, echo = F}
+p_tsne <- plot_outlier(x = metric_tsne, n.grid = 50, prob = prob, noutliers = 20, scale = 1/20)
 p_hdr_tsne <- hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20)
@@
 (p_hdr_tsne_p + p_tsne$p) + coord_fixed()
+```
+
+# UMAP
+
+```{r umap, message=FALSE, warning=FALSE, eval=TRUE}
+x <- train
+method <- "annUMAP"
+# perplexity <- round(k / 3) # 30 by default
+theta <- 0 # for exact tSNE in the C++ tSNE Barnes-Hut implementation
+# tictoc::tic()
+metric_umap <- metricML(x, s = s, k = k, radius = radius, method = method, annmethod = annmethod, eps = 0, distance = distance, treetype = treetype, searchtype = searchtype, perplexity = perplexity, theta = theta, invert.h = TRUE)
+# summary(metric_umap)
+# tictoc::toc()
+```
+
+```{r, message=FALSE, eval=TRUE}
+fumap <- metric_umap$embedding
+E1 <- fumap[,1]; E2 <- fumap[,2]
+# plot_embedding(metric_umap)
+# plot_ellipse(metric_umap, n.plot = 50)
+plot_contour(metric_umap, n.grid = 50, scale = 1/20)
+```
+
+```{r, echo = F}
+p_umap <- plot_outlier(x = metric_umap, n.grid = 50, prob = prob, noutliers = 20, scale = 1/20)
+p_hdr_umap <- hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20)
+p_hdr_umap_p <- p_hdr_umap$p + 
+  plot_ellipse(metric_umap, n.plot = 50, add = T)
+(p_hdr_umap_p + p_umap$p) + coord_fixed()
 ```
@@
 ```{r}
-(p_isomap$p + p_tsne$p) + coord_fixed()
+(p_isomap$p + p_tsne$p + p_umap$p) + coord_fixed()
 ```
@@
 ```{r}
-p_isomap_all <- plot_outlier(x = metric_isomap, n.grid = 50, prob = prob, noutliers = N, scale = 1/8)
-p_tsne_all <- plot_outlier(x = metric_tsne, n.grid = 50, prob = prob, noutliers =N, scale = 1/10)
+p_isomap_all <- plot_outlier(x = metric_isomap, n.grid = 50, prob = prob, noutliers = N, scale = 1/20)
+p_tsne_all <- plot_outlier(x = metric_tsne, n.grid = 50, prob = prob, noutliers =N, scale = 1/20)
+p_umap_all <- plot_outlier(x = metric_umap, n.grid = 50, prob = prob, noutliers =N, scale = 1/20)
 
@@
 head(p_tsne_all$outlier, n = 20)
+head(p_umap_all$outlier, n = 20)
 outlier_isomap <- order(as.numeric(p_isomap_all$outlier))
 outlier_tsne <- order(as.numeric(p_tsne_all$outlier))
+# outlier_umap <- order(as.numeric(p_umap_all$outlier))
+
 # outliers <- cbind(outlier_isomap, outlier_tsne)
-plot(outlier_isomap, outlier_tsne)
-
+cor(outlier_isomap, outlier_tsne)
 ```
 
-```{r, include=FALSE, eval=FALSE}
+```{r, include=T, eval=TRUE}
 # ggcorrplot::ggcorrplot(r, hc.order = TRUE, type = "lower", lab = TRUE)
-vec1 <- p_isomap$outlier
-vec2 <- p_tsne$outlier
-topconfects::rank_rank_plot(vec1, vec2, n=20)
+topconfects::rank_rank_plot(outlier_isomap, outlier_tsne, n=50)
 
@@
 ```{r}
-head(p_hdr_isomap$outlier, n = 20)
+head(p_isomap$outlier, n = 20)
 head(p_hdr_tsne$outlier, n = 20)
-hdroutlier_isomap <- order(as.numeric(p_hdr_isomap$outlier))
+hdroutlier_isomap <- order(as.numeric(p_isomap$outlier))
 hdroutlier_tsne <- order(as.numeric(p_hdr_tsne$outlier))
@@
 cor(hdroutlier_isomap %>% head(50), hdroutlier_tsne %>% head(50))
+topconfects::rank_rank_plot(hdroutlier_isomap, hdroutlier_tsne, n=50)
 ```
@@
 
-Similarly, we compare the highest density regions from both methods and expect a similar output in detecing most typical observations.
+Similarly, we compare the highest density regions from both methods and expect a similar output in detecting most typical observations.
 
@@
 ```{r}
-tail(p_hdr_isomap$outlier, n = 20)
+tail(p_isomap$outlier, n = 20)
 tail(p_hdr_tsne$outlier, n = 20)

### 2024-11-27 reports/20210826_swissroll.Rmd
Journal version pdf

---
title: "More data and methods for metric learning"
author: "Fan Cheng"
date: "2021-08-19"
output: 
  # prettydoc::html_pretty:
    # theme: cayman
    # theme: flatly
    rmdformats::readthedown:
    toc: true
    float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r libraries, message=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(dimRed)
library(reticulate)
library(here)
library(viridis)
library(hdrcde)
library(igraph)
library(matrixcalc)
library(akima)
library(car)
library(ggforce)
library(ks)
library(patchwork)
library(copula)
library(plotly)
Jmisc::sourceAll(here::here("R/sources"))
set.seed(123)
```

```{r}
# data size and meta data dimension
N <- 2000
p <- 2
```

# Swiss roll from copula

```{r simu, echo=TRUE}
# ----Swiss roll from copula-----------------------------------------
# ## a 3-dimensional normal copula
# norm.cop <- normalCopula(0.5, dim = p)
# u <- rCopula(N, norm.cop)
# den <- dCopula(u, norm.cop)
# # 3-d t copula
# t.cop <- tCopula(c(0.5, 0.3), dim = p, dispstr = "toep",
#                  df = 2, df.fixed = TRUE)
# u <- rCopula(N, t.cop)
# den <- dCopula(u, t.cop)
# 
## a 2-dimensional Clayton copula
cl3 <- claytonCopula(2, dim = p)
co <- rCopula(N, cl3)
# pairs(u)
claytonden <- dCopula(co, cl3) # true density

# swiss roll mapping
swissRollMapping <- function (x, y) { # U(1.5pi, 4.5pi)
    cbind(x = x * cos(x),
          y = y,
          z = x * sin(x))
}
u_sr <- (1.5 * pi) * (1 + 2 * co) 
claytonsr <- swissRollMapping(u_sr[,1], u_sr[,2])

# semi-sphere mapping
sphereMapping <- function (phi, psi) { # x~U(0,2pi); y~U(-1,1)
    cbind(x = cos(phi) * sin(psi),
          y = sin(phi) * sin(psi),
          z = cos(psi))
}
u_sp <- tibble(x = co[,1], y = co[,2]) %>% 
  mutate(x = 2 * pi * x, y = 2 * y - 1)
claytonsp <- sphereMapping(u_sp$x, u_sp$y)

# twin peak mapping
twinPeaksMapping <- function (x, y) { # x,y~U(-1,1)
    cbind(x = x,
          y = y,
          z = sin(pi * x) * tanh(3 * y))
}
u_tp <- 2 * co - 1
claytontp <- twinPeaksMapping(u_tp[,1], u_tp[,2])

# S curve mapping
sCurveMapping <- function (t, y) { # x~U(-1.5pi, 1.5pi); y~U(0,2)
    cbind(x = sin(t),
          y = y,
          z = sign(t) * (cos(t) - 1))
}
u_sc <- cbind(1.5 * pi * (2 * co[,1] - 1), 2 * co[,2])
claytonsc <- sCurveMapping(u_sc[,1], u_sc[,2])
```

```{r copulaplot}
claytondata <- as_tibble(claytonsr)
u <- u_sp
# highlight true density of u: red for high densities and black for low ones
col <- rep("grey", N)
col[head(order(claytonden), n=20)] <- "black"
col[tail(order(claytonden), n=20)] <- "gold"
par(mfrow=c(1,2))
plot(u[,1:2], xlab = expression(italic(u)[1]*"'"), ylab = expression(italic(u)[2]*"'"), 
     col = col, cex = 0.4)
scatterplot3d::scatterplot3d(claytondata, color = col)
plot_ly(data = claytondata, x = ~ x, y = ~ y, z = ~ z, color = col,
        type = "scatter3d", mode = "markers", size = 1, text = paste("density:", round(claytonden, 4)))
```

# Swiss roll from a Gaussian Mixture Model

```{r gaussian}
# randomly sampling from a Gaussian Mixture Model with centers/means at (7.5,7.5), (7.5,12.5), (12.5,7.5) and (12.5,12.5). The covariance for each gaussian was the 2x2 identity matrix
# 400 points in each of the four clusters
## Option 1. read data from online resource http://people.cs.uchicago.edu/~dinoj/manifold/swissroll.html
# preswissroll <- read_table("http://people.cs.uchicago.edu/~dinoj/manifold/preswissroll.dat", col_names = FALSE)
# preswissroll_label <- read_table("http://people.cs.uchicago.edu/~dinoj/manifold/preswissroll_labels.dat", col_names = FALSE)$X1 %>% as.factor()
# swissroll <- read_table("http://people.cs.uchicago.edu/~dinoj/manifold/swissroll.dat", col_names = c("x", "y", "z"))
# preswissroll %>% 
#   add_column(label = preswissroll_label) %>% 
#   ggplot(aes(X1, X2, col = label)) + 
#   geom_point()

## Option 2. manually generate multivariate normal random numbers
n <- round(N/4)
R <- matrix(c(1, 0,
              0, 1), 
            nrow = 2, ncol = 2)
# mu1 <- c(7.5, 7.5)
# mu2 <- c(7.5, 12.5)
# mu3 <- c(12.5, 7.5)
# mu4 <- c(12.5, 12.5)
mu <- tribble(
  ~mu1, ~mu2,
  7.5, 7.5,
  7.5, 12.5,
  12.5, 7.5,
  12.5, 12.5)

# mvtnorm::rmvnorm(n, mean = mu, sigma = R)
# MASS::mvrnorm(n, mu = mu1, Sigma = R)
preswissroll <- NULL
for(i in 1:4) {
  # mui <- get(paste0("mu", i))
  mui <- as.matrix(mu[i,])
  a <- mvtnorm::rmvnorm(n, mean = mui, sigma = R)
  den <- mvtnorm::dmvnorm(a, mean = mui, sigma = R)
  a <- cbind(a, i, den)
  preswissroll <- rbind(preswissroll, a)
}
colnames(preswissroll) <- c("x", "y", "label", "den")
preswissroll <- preswissroll %>%
  as_tibble() %>% 
  mutate(label = as.factor(label),
         x = (x - 5)/10, y = (y - 5)/10,
         den = den * 10 # y = h(x) = (x-5)/10; x = 10*y + 5; dx/dy = 10; f_Y(y) = f_X(x) * 10
         ) 
preswissroll %>%
  ggplot(aes(x, y, col = den)) + 
  geom_point() + 
  scale_color_viridis()
```

```{r gaussianmap, echo=FALSE, eval=TRUE}
# meta data co \in [0,1]
co <- as.matrix(preswissroll[,1:2])
# Swiss Roll mapping
u_sr <- (1.5 * pi) * (1 + 2 * co) 
gaussiansr <- swissRollMapping(u_sr[,1], u_sr[,2])
# semi-sphere mapping
u_sp <- tibble(x = co[,1], y = co[,2]) %>% 
  mutate(x = 2 * pi * x, y = 2 * y - 1)
gaussiansp <- sphereMapping(u_sp$x, u_sp$y)
# twin peak mapping
u_tp <- 2 * co - 1
gaussiantp <- twinPeaksMapping(u_tp[,1], u_tp[,2])
# S curve mapping
u_sc <- cbind(1.5 * pi * (2 * co[,1] - 1), 2 * co[,2])
gaussiansc <- sCurveMapping(u_sc[,1], u_sc[,2])
```

```{r gaussinplot}
gaussiandata <- as_tibble(gaussiansr)
u <- u_sr
plot_ly(data = gaussiandata, x = ~ x, y = ~ y, z = ~ z, color = preswissroll$den,
        type = "scatter3d", mode = "markers", size = 1, 
        text = paste("density:", round(preswissroll$den, 5))
        )
```

# Isomap

## Metric learning

```{r, eval = FALSE}
load("swissroll/gaussiansr_metric4methods.rda")
```

```{r parameters}
# train <- claytondata
train <- gaussiandata
# Parameters fixed
x <- train
N <- nrow(x)
s <- 2
k <- 20
method <- "annIsomap"
annmethod <- "kdtree"
distance <- "euclidean"
treetype <- "kd"
searchtype <- "radius" # change searchtype for radius search based on `radius`, or KNN search based on `k`
radius <- 8 # the bandwidth parameter, \sqrt(\elsilon), as in algorithm. Note that the radius need to be changed for different datasets, not to increase k
```

```{r, message=FALSE, eval = TRUE}
metric_isomap <- metricML(x, s = s, k = k, radius = radius, method = method, invert.h = TRUE, 
                          eps = 0,
                          annmethod = annmethod, distance = distance, treetype = treetype,
                          searchtype = searchtype
                          )
summary(metric_isomap)
```

```{r ggellipse, include=FALSE, eval=FALSE}
plot_embedding(metric_isomap) +
  labs(x = "ISO1", y = "ISO2") +
  plot_ellipse(metric_isomap, add = T, n.plot = 50,
             color = blues9[5], fill = blues9[5], alpha = 0.1,
             ell_size = .3)
```

## Outlier plot based on density estimates

### Fixed bandwidth

```{r}
# fixed bandwidth
fn <- metric_isomap$embedding
E1 <- fn[,1] # rename as Ed to match the aesthetics in plot_ellipse()
E2 <- fn[,2]
prob <- c(1, 50, 99)
ell_size <- 0 # ellipse size for plotting
p_hdr_isomap <- hdrscatterplot_new(E1, E2, levels = prob, noutliers = 20, label = NULL)
p_hdr_isomap_p <- p_hdr_isomap$p + 
  plot_ellipse(metric_isomap, add = T, n.plot = 50, 
             color = blues9[5], fill = blues9[5], alpha = 0.2, ell_size = ell_size)
```

### Pointwise adaptive bandwidth

```{r outliers}
Rn <- metric_isomap$rmetric # array
n.grid <- 10
fisomap <- vkde2d(x = fn[,1], y = fn[,2], h = Rn, n = n.grid)
fxy_isomap <- hdrcde:::interp.2d(fisomap$x, fisomap$y, fisomap$z, x0 = E1, y0 = E2)
# plot_contour(metric_isomap, n.grid = n.grid, scales = 1/20)
```

```{r hdroutliers}
# source(here::here("R/sources/hdrplotting.R"))
p_isomap <- plot_outlier(x = metric_isomap, n.grid = n.grid, prob = prob, scales = 1, ell_size = ell_size, 
                         f = fisomap
                         )
# p_isomap$p
# p_isomap$outlier # top outliers first
# p_isomap$densities # ascending densities matching $outlier order
(p_hdr_isomap_p / p_isomap$p ) + # coord_fixed()
  plot_layout(guides = 'collect')
```

# LLE

```{r lle, eval = TRUE}
x <- train
method <- "annLLE"
metric_lle <- metricML(x, s = s, k = k, radius = radius, method = method, invert.h = TRUE, 
                       eps = 0,
                      annmethod = annmethod, distance = distance, treetype = treetype,
                      searchtype = searchtype
)
```

```{r lleplot}
fn <- metric_lle$embedding
Rn <- metric_lle$rmetric
E1 <- fn[,1]; E2 <- fn[,2]
flle <- vkde2d(x = E1, y = E2, h = Rn, n = n.grid)
fxy_lle <- hdrcde:::interp.2d(flle$x, flle$y, flle$z, x0 = E1, y0 = E2)
# plot_embedding(metric_lle)
# plot_ellipse(metric_lle, n.plot = 50)
# plot_contour(metric_lle, n.grid = 20, scales = 1/20)

p_lle <- plot_outlier(x = metric_lle, n.grid = 20, prob = prob, noutliers = 20, scales = 1, ell_size = ell_size,
                      f = flle)
p_hdr_lle <- hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20)
p_hdr_lle_p <- p_hdr_lle$p + 
  plot_ellipse(metric_lle, n.plot = 50, add = T, ell_size = ell_size)
(p_hdr_lle_p + p_lle$p) + coord_fixed()
```

# t-SNE

```{r tsne, message=FALSE, warning=FALSE, eval=TRUE}
## ----tsne, message=FALSE, warning=FALSE, eval=TRUE------------------------------
x <- train
method <- "anntSNE"
perplexity <- 30 # round(k / 3) # 30 by default
theta <- 0 # for exact tSNE in the C++ tSNE Barnes-Hut implementation
# tictoc::tic()
metric_tsne <- metricML(x, s = s, k = k, radius = radius, method = method, 
                        annmethod = annmethod, eps = 0, distance = distance, treetype = treetype,
                        searchtype = searchtype, perplexity = perplexity, theta = theta, invert.h = TRUE)
```

```{r tsneplot, eval = TRUE}
fn <- metric_tsne$embedding
Rn <- metric_tsne$rmetric
E1 <- fn[,1]; E2 <- fn[,2]
ftsne <- vkde2d(x = E1, y = E2, h = Rn, n = n.grid)
fxy_tsne <- hdrcde:::interp.2d(ftsne$x, ftsne$y, ftsne$z, x0 = E1, y0 = E2)
# plot_embedding(metric_tsne)
# plot_ellipse(metric_tsne, n.plot = 50)
# plot_contour(metric_tsne, n.grid = 20, scales = 1/20)
```

```{r, echo = F}
# source(here::here("R/sources/hdrplotting.R"))
p_tsne <- plot_outlier(x = metric_tsne, n.grid = 20, prob = prob, noutliers = 20, scales = 1, f = ftsne, ell_size = ell_size)
p_hdr_tsne <- hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20)
p_hdr_tsne_p <- p_hdr_tsne$p + 
  plot_ellipse(metric_tsne, n.plot = 50, add = T, ell_size = ell_size)
(p_hdr_tsne_p + p_tsne$p) + coord_fixed()
```

# UMAP

```{r , eval = TRUE}
x <- train
method <- "annUMAP"
metric_umap <- metricML(x, s = s, k = k, radius = radius, method = method, 
                        annmethod = annmethod, eps = 0, distance = distance, treetype = treetype,
                        searchtype = searchtype, 
                        invert.h = TRUE)
```

```{r}
fumap <- metric_umap$embedding
E1 <- fumap[,1]; E2 <- fumap[,2]
fumap <- vkde2d(x = E1, y = E2, h = Rn, n = n.grid)
fxy_umap <- hdrcde:::interp.2d(fumap$x, fumap$y, fumap$z, x0 = E1, y0 = E2)
# plot_embedding(metric_umap)
# plot_ellipse(metric_umap, n.plot = 50)
# plot_contour(metric_umap, n.grid = 20, scales = 1/20)

p_umap <- plot_outlier(x = metric_umap, n.grid = 20, prob = prob, noutliers = 20, scales = 1, ell_size = ell_size)
p_hdr_umap <- hdrscatterplot(E1, E2, kde.package = "ks", noutliers = 20)
p_hdr_umap_p <- p_hdr_umap$p +
  plot_ellipse(metric_umap, n.plot = 50, add = T)
(p_hdr_umap_p + p_umap$p) + coord_fixed()
```

# Outliers with KDE compared with true densities

```{r trueden}
fxy <- claytonden # True densities
fxy <- preswissroll$den
# Calculates falpha needed to compute HDR of bivariate density den.
# Also finds approximate mode.
if(max(prob) > 50) {# Assume prob is coverage percentage
  alpha <- (100-prob)/100
} else {# prob is tail probability (for backward compatibility)
  alpha <- prob}
alpha <- sort(alpha)
falpha <- quantile(fxy, alpha)
# # Vector region of length N
# region <- rep(100, N)
# levels <- prob
# for (i in seq_along(levels))
#   region[fxy > falpha[i]] <- 100 - levels[i]
# # Construct region factor
# levels <- sort(unique(region[region < 100]), decreasing = TRUE)
# levels <- c(levels, 100)
# x$Region <- factor(region,
#                    levels = levels,
#                    labels = c(paste(head(levels, -1)), ">99")
#                   )
# # Sort data so the larger regions go first (other than outliers)
# r <- region
# r[region == 100] <- 0
# ord <- order(r, decreasing = TRUE)
# x <- x[ord, ]

# p_den_isomap <- plot_embedding(metric_isomap$embedding[ord,], embedding = TRUE, color = x$Region) 
# p_den_lle <- plot_embedding(metric_lle) + 
#   geom_point(aes(col = fxy)) 
# p_den_tsne <- plot_embedding(metric_tsne) + 
#   geom_point(aes(col = fxy)) 
# p_den_umap <- plot_embedding(metric_umap) + 
#   geom_point(aes(col = pfxy))

p_den_isomap <- hdrscatterplot(x = metric_isomap$embedding[,1], y = metric_isomap$embedding[,2], levels = prob, noutliers = 20, den = list(fxy = fxy, falpha = falpha))
p_den_lle <- hdrscatterplot(x = metric_lle$embedding[,1], y = metric_lle$embedding[,2], levels = prob, noutliers = 20, den = list(fxy = fxy, falpha = falpha))
p_den_tsne <- hdrscatterplot(x = metric_tsne$embedding[,1], y = metric_tsne$embedding[,2], levels = prob, noutliers = 20, den = list(fxy = fxy, falpha = falpha))
p_den_umap <- hdrscatterplot(x = metric_umap$embedding[,1], y = metric_umap$embedding[,2], levels = prob, noutliers = 20, den = list(fxy = fxy, falpha = falpha))

p1 <- (((p_den_isomap$p + ggtitle("Isomap")) | (p_den_lle$p + ggtitle("LLE")) | (p_den_tsne$p + ggtitle("t-SNE")) | (p_den_umap$p + ggtitle("UMAP"))) /
    ((p_isomap$p) | 
    (p_lle$p) | 
    (p_tsne$p) | 
    (p_umap$p )) /
(p_hdr_isomap$p | p_hdr_lle$p | p_hdr_tsne | p_hdr_umap) +
  coord_fixed() +
  plot_layout(guides = 'collect') +
  plot_annotation(subtitle = "
Top: Outliers with true densities;                  
Middle: Outliers with variable bandwidth;
Bottom: Outliers with fixed bandwidth;
",
  title = "Twin peak mapping data from Clayton copula")) &
  labs(x = "", y = "") 
p1
```

```{r, eval = FALSE}
save(metric_isomap, metric_lle, metric_tsne, metric_umap, claytonden, x, file = "swissroll/copulatp_metric4methods.rda")
```

<!-- # Compare the density estimates with the true density -->
<!-- NOT WORKING DUE TO DIFFERENT SCALES -->

```{r, eval=FALSE}
# fxy_METHOD is scaled to a different level when applying interp.2d
# can't plot the true densities directly for comparison
par(mfrow=c(2,2))
for(i in 1:4) {
  methods <- c("isomap", "lle", "tsne", "umap")[i]
  fxy_ml <- get(paste0("fxy_", methods))
  plot(fxy, fxy_ml, xlab = "True density", ylab = "Density estimates", asp = 1)
  abline(0, 1, col = "red")
}
```

```{r}
i <- 1
for(i in 1:4){
  methods <- c("isomap", "lle", "tsne", "umap")[i]
  fxy_ml <- get(paste0("fxy_", methods))
  fxy_hdr <- get(paste0("p_hdr_", methods))
  fxy_hdr <- fxy_hdr$densities
  cor(fxy, fxy_ml, method = "spearman") %>% print() # change hdrscatterplot() output
  cor(fxy, fxy_hdr, method = "spearman") %>% print() # hdr densities

  p1 <- cbind(fxy, fxy_ml) %>% 
    as_tibble() %>% 
    ggplot(aes(fxy, fxy_ml)) + 
    geom_point(col = col) +
    scale_x_log10() + 
    scale_y_log10() + 
    ggtitle("Variable bandwidth")
  p2 <- cbind(fxy, fxy_hdr) %>% 
    as_tibble() %>% 
    ggplot(aes(fxy, fxy_hdr)) + 
    geom_point(col = col) +
    scale_x_log10() + 
    scale_y_log10() + 
    ggtitle("Fixed bandwidth")

  print(p1 + p2)
}
```

# Compare the rank of outliers with the true density

```{r}
den_rank <- order(fxy) # true density ranks in increasing order, anomalies first
p_isomap_all <- plot_outlier(x = metric_isomap, n.grid = 20, prob = prob, noutliers = N, scales = 1, f = fisomap)
p_lle_all <- plot_outlier(x = metric_lle, n.grid = 20, prob = prob, noutliers = N, scales = 1, f = flle)
p_tsne_all <- plot_outlier(x = metric_tsne, n.grid = 20, prob = prob, noutliers =N, scales = 1, f = ftsne)
p_umap_all <- plot_outlier(x = metric_umap, n.grid = 20, prob = prob, noutliers =N, scales = 1, f = fumap)
```

```{r}
par(mfrow=c(2,2))
for(i in 1:4) {
  methods <- c("isomap", "lle", "tsne", "umap")[i]
  outlier <- get(paste0("p_", methods, "_all"))
  plot(den_rank, outlier$outlier, xlab = "True outlier rankinng", ylab = "Estimated outlier ranking", col = col, main = toupper(methods))
  abline(0, 1, col = "red")
}
```

<!-- # Compare outliers among different methods -->

```{r, eval=FALSE}
par(mfrow=c(2,3))
for(i in 1:4) {
  for (j in (i+1):4) {
    methods <- c("isomap", "lle", "tsne", "umap")
    outlier1 <- get(paste0("p_", methods[i], "_all"))
    outlier2 <- get(paste0("p_", methods[j], "_all"))
    plot(outlier1$outlier, outlier2$outlier, xlab = paste("Outlier rankinng from", methods[i]), ylab = paste("Outlier rankinng from", methods[j]), col = col)
    abline(0, 1, col = "red")
  }
}
```

<!-- To compare two manifold learning methods, we calculate the overlap among outliers. To do so, we first rank the outliers from highest to lowest densities. Then calculate the correlation of ranks for both methods.  -->

<!-- To show our methods performs better, we expect a higher correlation of ranks. This shows robustness in finding outliers using different manifold learning methods. -->

<!-- ```{r} -->
<!-- p_isomap_all <- plot_outlier(x = metric_isomap, n.grid = 20, prob = prob, noutliers = N, scales = 1/20, f = fisomap) -->
<!-- p_lle_all <- plot_outlier(x = metric_lle, n.grid = 20, prob = prob, noutliers = N, scales = 1/20, f = flle) -->
<!-- p_tsne_all <- plot_outlier(x = metric_tsne, n.grid = 20, prob = prob, noutliers =N, scales = 1/20, f = ftsne) -->
<!-- p_umap_all <- plot_outlier(x = metric_umap, n.grid = 20, prob = prob, noutliers =N, scales = 1/20) -->

<!-- head(p_isomap_all$outlier, n = 20) -->
<!-- head(p_tsne_all$outlier, n = 20) -->
<!-- # head(p_umap_all$outlier, n = 20) -->
<!-- outlier_isomap <- order(as.numeric(p_isomap_all$outlier)) -->
<!-- outlier_tsne <- order(as.numeric(p_tsne_all$outlier)) -->
<!-- # outlier_umap <- order(as.numeric(p_umap_all$outlier)) -->
<!-- cor(outlier_isomap, outlier_tsne) -->
<!-- plot(outlier_isomap, outlier_tsne, asp = 1) -->
<!-- abline(0, 1) -->
<!-- ``` -->

<!-- ## Outlier ranking comaprison using fixed bandwidth -->

<!-- ```{r} -->
<!-- head(p_isomap$outlier, n = 20) -->
<!-- head(p_hdr_tsne$outlier, n = 20) -->
<!-- hdroutlier_isomap <- order(as.numeric(p_isomap$outlier)) -->
<!-- hdroutlier_tsne <- order(as.numeric(p_hdr_tsne$outlier)) -->
<!-- cor(hdroutlier_isomap, hdroutlier_tsne) -->
<!-- plot(hdroutlier_isomap, hdroutlier_tsne, asp = 1) -->
<!-- abline(0, 1) -->
<!-- # cor(hdroutlier_isomap %>% head(50), hdroutlier_tsne %>% head(50)) -->
<!-- # topconfects::rank_rank_plot(hdroutlier_isomap, hdroutlier_tsne, n=50) -->
<!-- ``` -->

<!-- # Compare high density regions -->

<!-- Similarly, we compare the highest density regions from both methods and expect a similar output in detecting most typical observations.  -->

<!-- ## Variable bandwidth -->

<!-- ```{r} -->
<!-- tail(p_isomap_all$outlier, n = 20) -->
<!-- tail(p_tsne_all$outlier, n = 20) -->
<!-- cor(outlier_isomap %>% tail(50), outlier_tsne %>% tail(50)) -->
<!-- ``` -->

<!-- ## Fixed bandwidth -->

<!-- ```{r} -->
<!-- tail(p_isomap$outlier, n = 20) -->
<!-- tail(p_hdr_tsne$outlier, n = 20) -->
<!-- cor(hdroutlier_isomap %>% tail(50), hdroutlier_tsne %>% tail(50)) -->
<!-- ``` -->
